---
title: "00_importantObjects"
author: "Caitlin Page"
date: "2025-11-20"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

## Description
* I have a lot of objects I use across analysis files
* And I am bad at keeping track of them and where they start
* Which is a problem for building websites and having them work
* So all the file processing put here and save the RDS objects to load at the start of each website

## Packages

```{r}
library(BSgenome.Hsapiens.UCSC.hg38)
library(data.table)
library(plyranges)
library(tidyr)
library(stringr)
library(dplyr)
library(ggplot2)
```


## Objects


# CpG sites
* just make it every time
```{r}
cg_sites <- cbind(data.frame(Biostrings::matchPattern("CG", BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38[["chr22"]])), seqnames = "chr22") %>% 
  mutate(pos = paste0(seqnames, "-", start)) %>%
  relocate(pos, seqnames)
cg_sites$index <- 1:nrow(cg_sites)
```

```{r}
seqs <- seqnames(BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38)[1:22]

cg_sites <- lapply(seqs, function(x) {
        cbind(data.frame(Biostrings::matchPattern("CG", BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38[[x]])),
            seqnames = x
        )}) %>% bind_rows()
```





# Nanopore processed file


```{r}
indiv_reads <- readRDS("/researchers/caitlin.page/cf_nano/r_output/indiv_reads.rds")
bed_betas <- readRDS("/researchers/caitlin.page/cf_nano/r_output/bed_betas.rds")
bed_betas$pos <- paste0(bed_betas$seqnames, "-", bed_betas$start)
bed_betas_hu5.10 <- bed_betas %>% filter(sample == "hu5.10")
```

```{r}
indiv_reads$ref_position <- indiv_reads$ref_position + 1 # convert to 1 based
indiv_reads <- indiv_reads %>% mutate(ref_position = ifelse(ref_strand == "-", ref_position - 1, ref_position))
indiv_reads$read_and_pos <- paste0(indiv_reads$read_id, ":", indiv_reads$ref_position)
indiv_reads$start <- indiv_reads$ref_position
indiv_reads$end <- indiv_reads$ref_position
alpha_and_beta2 <- find_overlaps(as_granges(indiv_reads), as_granges(bed_betas_hu5.10)) %>% data.frame() %>% filter(call_prob > 0.66) %>%
  group_by(read_id) %>% .[order(.$alignment_start),] %>% 
  mutate(call_code = ifelse(call_code == "m", "M", "U")) %>%
  mutate(meth_pattern = paste(call_code, collapse = "")) %>%
  mutate(num_meth = str_count(meth_pattern, pattern = "M"),
         total = str_length(meth_pattern),
         alpha = num_meth/total) %>%
  mutate(genom_positions = paste0(ref_position, collapse = ",")) %>%
  ungroup()
alpha_and_beta2 <- alpha_and_beta2 %>% data.frame()
head(alpha_and_beta2)
```

```{r}
saveRDS(alpha_and_beta2, "/researchers/caitlin.page/cf_nano/r_output/nano_alpha_beta.rds")
```

```{r}
indiv_reads <- nano_reads
indiv_reads$ref_position <- indiv_reads$ref_position + 1 # convert to 1 based
indiv_reads <- indiv_reads %>% mutate(ref_position = ifelse(ref_strand == "-", ref_position - 1, ref_position))
indiv_reads$read_and_pos <- paste0(indiv_reads$read_id, ":", indiv_reads$ref_position)
indiv_reads$seqnames <- indiv_reads$chrom
indiv_reads$start <- indiv_reads$ref_position
indiv_reads$end <- indiv_reads$ref_position
indiv_reads <- indiv_reads %>% filter(call_prob > 0.66) %>%
  group_by(read_id) %>% .[order(.$alignment_start),] %>% 
  mutate(call_code = ifelse(call_code == "m", "M", "U")) %>%
  mutate(meth_pattern = paste(call_code, collapse = "")) %>%
  mutate(num_meth = str_count(meth_pattern, pattern = "M"),
         total = str_length(meth_pattern),
         alpha = num_meth/total) %>%
  mutate(genom_positions = paste0(ref_position, collapse = ",")) %>%
  ungroup() %>%
  group_by(seqnames, start) %>% 
  mutate(beta = ) %>% data.frame()
saveRDS(indiv_reads, "/researchers/caitlin.page/cf_nano/r_output/nano_alpha_beta_all.rds")
```
```{r}
nano_reads
```


```{r}
rrbs_alpha_beta
```


```{r}
alpha_and_beta2_pat <- alpha_and_beta2 %>% 
  mutate(start_pos = sapply(str_split(genom_positions, ","), head, 1),
         end_pos = sapply(str_split(genom_positions, ","), tail, 1)) %>% 
  filter(ref_position == start_pos) %>%
  distinct(read_id, seqnames, start_pos, meth_pattern, alpha, end_pos, beta) %>% 
  group_by(seqnames, start_pos, meth_pattern, alpha, end_pos, beta) %>% 
  summarise(n=n()) %>%
  ungroup()
nrow(alpha_and_beta2_pat)
alpha_and_beta2_pat %>% group_by(n) %>% summarise(n())

alpha_and_beta2_pat <- alpha_and_beta2_pat %>%
  mutate(start_pos = as.integer(start_pos),
         end_pos = as.integer(end_pos))

alpha_and_beta2_pat[1:10,]
```
```{r}
saveRDS(alpha_and_beta2_pat, "/researchers/caitlin.page/cf_nano/r_output/nano_pat_alpha_beta.rds")
```


# WGBS data

```{r}
files <- list.files("/researchers/caitlin.page/pancreatic_cancer_wgbs")
files[grep("CpG", files)]
```

```{r}
files <- list.files("/researchers/caitlin.page/pancreatic_cancer_wgbs/bismark_res")
files[grep("CpG", files)]
```
```{r}
source("../processData.R")

bis <- fread("/researchers/caitlin.page/pancreatic_cancer_wgbs/bismark_res/CpG_OT_SRR1646792_GSM1541788_38-Lg_rep1_BS-Seq_Homo_sapiens_Bisulfite-Seq_1_bismark_bt2_pe.txt")
pancreatic_a38_lg1 <- process_wgbs(bis)

saveRDS(pancreatic_a38_lg1, "/researchers/caitlin.page/cf_nano/r_output/pancreatic_a38_lg1.rds")
```

# RRBS processed file

FastQ to CpG text file with bismark described [here](../code/seqFastqToReadMeth.txt)

* ENCODE cell line
```{r}
files <- list.files("/researchers/caitlin.page/rrbs_encode_cell_line/")
files[grep("CpG", files)]
```


```{r}
source("../processData.R")

bis <- fread("/researchers/caitlin.page/rrbs_encode_cell_line/CpG_OT_ENCFF000LYO_bismark_bt2.txt")
for_beta <- process_wgbs(bis)
for_beta <- for_beta %>% filter(seqnames == "chr22")
rrbs_for_beta_pat <- pat_for_wgbs(for_beta)
just_beta <- for_beta %>% distinct(start, beta) %>% .[order(.$start),]
```
```{r}
saveRDS(for_beta, "/researchers/caitlin.page/cf_nano/r_output/rrbs_alpha_beta.rds")
```

```{r}
readRDS("../output/cluster_files/rrbs_alpha_beta.rds")
```


```{r}
source("../processData.R")

bis <- fread("/researchers/caitlin.page/rrbs_encode_cell_line/CpG_OT_ENCFF000LYP_bismark_bt2.txt")
gm12878_2 <- process_wgbs(bis)
gm12878_2_pat <- pat_for_wgbs(gm12878_2)
gm12878_2_beta <- gm12878_2 %>% distinct(start, beta) %>% .[order(.$start),]

saveRDS(gm12878_2, "/researchers/caitlin.page/cf_nano/r_output/rrbs_alpha_beta_rep2.rds")
```

* ENCODE Cell line 562

```{r}
source("../processData.R")

bis <- fread("/researchers/caitlin.page/rrbs_encode_cell_line/CpG_OT_k562_cml_line1_bismark_bt2.txt")
k562 <- process_wgbs(bis)
k562_pat <- pat_for_wgbs(k562)
k562_beta <- k562 %>% distinct(start, beta) %>% .[order(.$start),]
```

```{r}
saveRDS(k562, "/researchers/caitlin.page/cf_nano/r_output/k562_rep1.rds")
```

```{r}
source("../processData.R")

bis <- fread("/researchers/caitlin.page/rrbs_encode_cell_line/CpG_OT_k562_cml_line2_bismark_bt2.txt")
k562_2 <- process_wgbs(bis)
#k562_2_pat <- pat_for_wgbs(k562_2)
#k562_2_beta <- k562_2 %>% distinct(start, beta) %>% .[order(.$start),]
```

```{r}
saveRDS(k562_2, "/researchers/caitlin.page/cf_nano/r_output/k562_rep2.rds")
```

```{r}
cols <- c("read_id", "seqnames", "start", "end", "strand", "meth_pattern", "meth_status", "num_meth", "total", "alpha", "read_start", "read_end", "beta", "sample")

all_samples <- rbind(rrbs_alpha_beta %>% mutate(sample = "b1") %>% .[,cols], gm12878_2 %>% mutate(sample = "b2") %>% .[,cols], 
                     k562 %>% mutate(sample = "cml1") %>% .[,cols], k562_2 %>% mutate(sample = "cml2") %>% .[,cols])
all_samples_simp <- all_samples %>% distinct(seqnames, read_start, read_end, strand, sample, read_id, alpha, meth_pattern, num_meth, total)
colnames(all_samples_simp)[2:3] <- c("start", "end")
```

```{r}
saveRDS(all_samples, "/researchers/caitlin.page/cf_nano/r_output/all_rrbs_samples.rds")
```


# genome tiles
```{r, eval=FALSE}
for_op <- rbind(just_beta, just_beta_gm2, just_beta_k1, just_beta_k2) %>% distinct()
for_op <- for_op %>% mutate(end = start, strand = "*", numCs = beta * coverage, numTs = coverage - numCs, chr = seqnames) %>% .[,c(10,2,6,7,4,8,9)]
mr <- new(
  "methylRaw",
  for_op,
  sample.id = "sample1",
  assembly = "hg38",
  context = "CpG",
  resolution = "base"
)

length_region <- 300
tiles <- methylKit::tileMethylCounts(mr,win.size=length_region,step.size=length_region,cov.bases = 0)

tilesdf_all <- methylKit::getData(tiles)

optiles_all <- merging_consecutive_regions(mr, tiles, da_thr = length_region, cb_thr = length_region)

tilesdf_all <- tilesdf_all[,1:4] %>% mutate(width = end - start + 1) 
colnames(tilesdf_all)[1] <- "seqnames"
tilesdf_all$id <- 1:nrow(tilesdf_all)

optiles_all <- optiles_all %>% mutate(width = end - start + 1)
colnames(optiles_all)[1] <- "seqnames"
optiles_all$index <- 1:nrow(optiles_all)
saveRDS(optiles_all, "/researchers/caitlin.page/cf_nano/r_output/optiles_all.rds")
```



```{r}
optiles_all <- readRDS("/researchers/caitlin.page/cf_nano/r_output/optiles_all.rds")
```

## results

# dmrs
```{r}
dmrcate_seq_dmr <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dmr_encode_gm12878_k562.rds")
```


# dtu tiles
```{r, eval=FALSE}
dtu_optile <- find_overlaps(as_granges(all_samples), as_granges(optiles_all)) %>% data.frame() %>%
  distinct(sample, read_id, read_start, read_end, alpha, region_id) %>% 
  mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8"))

dtu_optile <- dtu_optile %>% 
  mutate(transcript = paste0(region_id, "_", alpha_group)) %>%
  group_by(sample, transcript) %>%
 summarise(transcript_count = n()) %>% ungroup() %>% data.frame()

# build matrix of "transcript" counts - transcript rows, sample as columns
dtu_optile <- dtu_optile %>% pivot_wider(names_from = sample, values_from = transcript_count)
dtu_optile[is.na(dtu_optile)] <- 0
dtu_optile$gene_id <- sub("_.*", "", dtu_optile$transcript)

# DTU analysis with edgeR
## Alicia prefers gene method over Simes method (documentation slightly confusing about difference)
y <- DGEList(as.matrix(dtu_optile[,2:5]), group = c(1,1,2,2), genes = dtu_optile[,c("gene_id", "transcript")])
design <- model.matrix(~ c(1,1,2,2), data = y$samples)
fit <- glmQLFit(y, design, robust=TRUE)
ds <- diffSpliceDGE(fit, coef = 2, geneid = "gene_id", exonid = "transcript")
ts.gene_optiles <- topSpliceDGE(ds, test = "gene", number = Inf)


# add info to results: get proportion counts
gene_cg <- find_overlaps(as_granges(optiles_all), as_granges(cg_sites)) %>% data.frame() %>%
  group_by(region_id) %>%
  summarise(n_cg = n()) %>%
  ungroup() %>% data.frame()
fit$genes <- fit$genes %>% mutate(n_cg = gene_cg[match(.$gene_id, gene_cg$region_id), "n_cg"])
rpk <- y$counts/rep(fit$genes$n_cg/1000, times = ncol(y$counts))
tpm_mat <- sweep(rpk, 2, colSums(rpk)/1e6, "/")
tpm_mat <- tpm_mat %>% data.frame() %>% mutate(id = dtu_optile$transcript, gene_id = dtu_optile$gene_id)

tpm_mat <- tpm_mat %>%
  group_by(gene_id) %>%
  mutate(prop_S1_gm = b1/sum(b1),
         prop_S2_gm = b2/sum(b2),
         prop_S3_k = cml1/sum(cml1),
         prop_S4_k = cml2/sum(cml2)) %>%
  ungroup() %>% data.frame()


# add locations back in so compatible with plyranges stuff
## this bit only applicable to OPTILES because of the way it structures the ID
ts.gene_optiles <- ts.gene_optiles %>%
  mutate(region_id = gene_id) %>%
  separate_wider_delim(gene_id, delim = ":", names = c("seqnames", "gene_id")) %>%
  mutate(seqnames = gsub("chrchr", "chr", seqnames)) %>%
  separate_wider_delim(gene_id, delim = "-", names = c("start", "end"))
ts.gene_optiles$start <- as.double(ts.gene_optiles$start)
ts.gene_optiles$end <- as.double(ts.gene_optiles$end)
ts.gene_optiles$dtu_rank <- 1:nrow(ts.gene_optiles)
saveRDS(dtu.optiles, "/researchers/caitlin.page/cf_nano/r_output/ts.gene_optiles.rds")
```


```{r}
dtu.optiles <- readRDS("/researchers/caitlin.page/cf_nano/r_output/ts.gene_optiles.rds")
```

# dtu genes

```{r}
dtu.gene <- readRDS("/researchers/caitlin.page/cf_nano/r_output/ts.gene.gene.rds")
```



# more dtu objects

```{r}
saveRDS(dtu_tiles_5_beta, "/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_5_beta.rds")
saveRDS(dtu_tiles_5_exon, "/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_5_exon.rds")
saveRDS(dtu_tiles_5_mean_alpha, "/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_5_mean_alpha.rds")
```

