---
title: "analyseEntropy"
author: "caitlinpage"
date: "2026-05-28"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

## Introduction

```{r}
library(annotatr)
library(data.table)
library(plyranges)
library(tidyr)
library(stringr)
library(dplyr)
library(ggplot2)
library(edgeR)
library(OpTiles)
library(ggVennDiagram)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
library(missMethyl)
```


* metheor ME [github](https://github.com/dohlee/metheor)
* method they used is from Xie (2011), [paper](https://academic.oup.com/nar/article/39/10/4099/1303103?login=false&guestAccessKey=)

```{r}
seqs <- seqnames(BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38)[1:22]

cg_sites <- lapply(seqs, function(x) {
        cbind(data.frame(Biostrings::matchPattern("CG", BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38[[x]])),
            seqnames = x
        )}) %>% bind_rows()
```

```{r}
all_samples <- readRDS("/researchers/caitlin.page/cf_nano/r_output/all_rrbs_samples.rds")
optiles_all <- readRDS("/researchers/caitlin.page/cf_nano/r_output/optiles_all.rds")
dtu.optiles <- readRDS("/researchers/caitlin.page/cf_nano/r_output/ts.gene_optiles.rds")
dmrcate_seq_dmr <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dmr_encode_gm12878_k562.rds")
```

```{r}
all_samples_simp <- all_samples %>% distinct(seqnames, read_start, read_end, strand, sample, read_id, alpha, meth_pattern, num_meth, total)
colnames(all_samples_simp)[2:3] <- c("start", "end")
```

```{r}
entropy_gm1 <- read.table("/researchers/caitlin.page/cf_nano/run_metheor/me_ENCFF000LYO.tsv")
colnames(entropy_gm1) <- c("seqnames", "start", "start_2", "start_3", "start_4", "ME")
entropy_gm1$start <- entropy_gm1$start + 1
entropy_gm1$start_2 <- entropy_gm1$start_2 + 1
entropy_gm1$start_3 <- entropy_gm1$start_3 + 1
entropy_gm1$start_4 <- entropy_gm1$start_4 + 1

entropy_gm1 <- entropy_gm1 %>% filter(seqnames == "chr1") %>% .[order(.$start),]
entropy_gm1
```

```{r}
entropy_gm1 %>%
  ggplot(aes(x = ME)) +
  geom_density()
```


# calculate entropy

* ooh there's a Bioconductor package that uses same definiton by Xie that metheor implements
[MethTargetedNGS](https://www.bioconductor.org/packages/release/bioc/html/MethTargetedNGS.html)
* can I just steal the code and figure out how to make it work in read based way?

* potential read lengths
```{r}
all_samples_simp %>%
  ggplot(aes(x = as.factor(total))) +
  geom_bar()
```

```{r}
all_samples_simp %>% filter(total > 2) %>%
  ggplot(aes(x = as.factor(total))) +
  geom_bar()
```

* methtargetedngs
```{r}
methEntropy <- function(x)
{
  t1 <- Sys.time()
  entropy <- list()
  ni_list <- list()
  for (i in 1:(ncol(x)-3))
  {
    sum=0
    entropy_list <- unlist(1:nrow(x))
    for (j in 1:(nrow(x)-1))
    {
      ni_new=0
      for (k in (j+1):length(entropy_list))
      {
        if (identical(x[j,i:(i+3)],x[entropy_list[k],i:(i+3)]))
        { 
          entropy_list <- entropy_list[-k]
          ni_new=ni_new + 1
        }
      }
      
      if (ni_new > 0)
      {
        ni_list[j] <- ni_new
        ni_for_sum <- (-((ni_new/nrow(x))*(log(ni_new/nrow(x)))))
        sum= sum + ni_for_sum
      }
    }
    entropy[i] <- sum/4
  }
  t2 <- Sys.time()
  print (round(t2-t1,digits=2))
  return(round(unlist(entropy),digits=2))
}
```
** this method uses sliding windows
** input is the hAlign object but not quite sure what that looks like


* perplexity approach

```{r}
methylation_entropy <- function(patterns, base = 2) {
  patterns <- patterns[!is.na(patterns)]
  if (length(patterns) == 0) return(NA_real_)
  
  p <- prop.table(table(patterns))
  -sum(p * log(p, base = base))
}
```

```{r}
patterns <- c("0000", "0000", "0010", "0010", "1111", "1010")
methylation_entropy(patterns)
```

** with cpg matrix

```{r}
window_entropy <- function(mat) {
  patterns <- apply(mat, 1, paste0, collapse = "")
  methylation_entropy(patterns)
}
```

```{r}
mat <- matrix(
  c(1,0,0,1,
    1,0,0,1,
    0,0,1,0,
    1,1,1,1),
  nrow = 4, byrow = TRUE
)

window_entropy(mat)
```


# [dnmtools (command lines)](https://dnmtools.readthedocs.io/en/latest/entropy/)

* Can do entropy on different window sizes
* but uses basically pat format (just not with the counts), chrom, start cpg index, meth pattern


# the math

entropy = -Sum[Prob(x) log2(Prob(x))].  Prob(x)

meth entropy = e/numcpgs Sum[(- occ pattern i /total reads) Log(occ patt i/total reads)]

(I think then it's about doing the sum for each meth pattern)

```{r}
entropy_gm1
```


```{r}
all_samples %>% filter(sample == "b1", seqnames == "chr1", start >= 827517, end <= 827537) %>% distinct(start)
```

```{r}
try_entropy <- all_samples %>% filter(sample == "b1", seqnames == "chr1", start >= 827517, end <= 827537)
try_entropy %>% distinct(read_start)
```
```{r}
format_entropy <- try_entropy %>% mutate(ent_pat = substr(meth_pattern, 1, 4)) %>% distinct(read_id, ent_pat) %>% 
  mutate(ent_pat = gsub("U", 0, ent_pat)) %>% separate_wider_position(ent_pat, widths = c(cpg1 = 1, cpg2 = 1, cpg3 = 1, cpg4 = 1)) %>%
  mutate(cpg1 = as.double(cpg1), cpg2 = as.double(cpg2), cpg3 = as.double(cpg3), cpg4 = as.double(cpg4))
```

meth entropy = e/numcpgs Sum[(- occ pattern i /total reads) Log(occ patt i/total reads)]
```{r}
format2 <- try_entropy %>% mutate(ent_pat = substr(meth_pattern, 1, 4)) %>% distinct(read_id, ent_pat) %>% 
  mutate(ent_pat = gsub("U", 0, ent_pat)) %>% mutate(ent_pat = as.double(ent_pat))
```
```{r}
?prop.table()
```

```{r}
prop.table(table(format_entropy), margin = 2)
prop.table(table(format_entropy[,2:5]))
prop.table(table(format2))
```


```{r}
methylation_entropy <- function(patterns, base = 2) {
  patterns <- patterns[!is.na(patterns)]
  if (length(patterns) == 0) return(NA_real_)
  
  p <- prop.table(table(patterns))
  -sum(p * log(p, base = base))
}
```

```{r}
methylation_entropy(format2)
```

```{r}
methEntropy(format_entropy)
```
* well the stolen code worked - but also why did it output 2 things

```{r}
methEntropy(format2)
```
```{r}
entropy_gm1
```

```{r}
try_entropy <- all_samples %>% filter(sample == "b1", seqnames == "chr1", start >= 827413, end <= 827445)
try_entropy %>% distinct(read_start)

format_entropy <- try_entropy %>% mutate(ent_pat = substr(meth_pattern, 1, 4)) %>% distinct(read_id, ent_pat) %>% mutate(ent_pat = str_replace_all(ent_pat, c("U" = "0", "M" = "1"))) %>% separate_wider_position(ent_pat, widths = c(cpg1 = 1, cpg2 = 1, cpg3 = 1, cpg4 = 1)) %>%
  mutate(cpg1 = as.double(cpg1), cpg2 = as.double(cpg2), cpg3 = as.double(cpg3), cpg4 = as.double(cpg4))
methEntropy(format_entropy)
```

* that is not correct
