---
title: "basicTilingAnalysis"
author: "Caitlin Page"
date: "2026-08-14"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

## Introduction

Tile the hg38 genome into 500 bp non-overlapping windows across all standard chromosomes (chr1–22, chrX, chrY, chrM).

## Packages

```{r}
library(BSgenome.Hsapiens.UCSC.hg38)
library(data.table)
library(plyranges)
library(tidyr)
library(stringr)
library(dplyr)
library(ggplot2)
```

## Tile genome

```{r}
# Get chromosome lengths for standard chromosomes only
seqinfo_hg38 <- seqlengths(BSgenome.Hsapiens.UCSC.hg38)
std_chroms <- grepl("^chr([1-9]|1[0-9]|2[0-2]|X|Y|M)$", names(seqinfo_hg38))
seqinfo_hg38 <- seqinfo_hg38[std_chroms]
seqinfo_hg38
```

```{r}
# Tile into 500 bp non-overlapping windows
tiles_500bp <- GenomicRanges::tileGenome(seqinfo_hg38, tilewidth = 500, cut.last.tile.in.chrom = TRUE)

# Add a tile number column
tiles_500bp <- tiles_500bp %>%
  data.frame() %>%
  mutate(tile_num = 1:n()) %>%
  as_granges()

tiles_500bp
```

```{r}
# How many tiles?
length(tiles_500bp)
```

```{r, eval=FALSE}
saveRDS(tiles_500bp, "/researchers/caitlin.page/cf_nano/r_output/tiles_500bp_hg38.rds")
```


# overlap tiles and CpGs

```{r}
# cpgs
seqs <- seqnames(BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38)[1:22]

cg_sites <- lapply(seqs, function(x) {
  cbind(data.frame(Biostrings::matchPattern("CG", BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38[[x]])),
        seqnames = x
  )}) %>% bind_rows()
```

```{r}
tiles_500bp_cg <- find_overlaps(tiles_500bp, as_granges(cg_sites)) %>% data.frame() %>%
  group_by(tile_num) %>% mutate(n_cg = n())
tiles_500bp_cg <- tiles_500bp_cg %>% ungroup() %>% distinct()
```

## how many tiles were removed?

```{r}
length(tiles_500bp)
nrow(tiles_500bp_cg)
```


```{r}
tiles_500bp %>% data.frame() %>% mutate(has_cg = ifelse(tile_num %in% tiles_500bp_cg$tile_num, TRUE, FALSE)) %>%
  ggplot(aes(x = has_cg, fill = has_cg)) +
  geom_bar() +
  theme(legend.position = "none")
```

# where are they along the chromosome?

* guessing probably starts and ends

```{r}
tiles_500bp %>% data.frame() %>% mutate(has_cg = ifelse(tile_num %in% tiles_500bp_cg$tile_num, 2, 1)) %>% 
  filter(seqnames == "chr22") %>%
  ggplot(aes(x = tile_num, y = has_cg)) +
  geom_point(alpha = 0.2)
```
* ok so it's tricky to see
* not sure how to visualise - and don't think it matters too much


# overlap to OpTiles tiles

```{r}
optiles_all <- readRDS("/researchers/caitlin.page/cf_nano/r_output/optiles_all_emseq.rds")
```


```{r}
tiles_500bp_cg$tile_id <- paste0(tiles_500bp_cg$seqnames, ":", tiles_500bp_cg$start, "-", tiles_500bp_cg$end)
```

```{r}
tile_overlap <- find_overlaps(as_granges(tiles_500bp_cg), as_granges(optiles_all)) %>% data.frame()
```

```{r, eval=FALSE}
optiles_all %>% mutate(overlap_tile = ifelse(region_id %in% tile_overlap$region_id, TRUE, FALSE)) %>%
  ggplot(aes(x = region_id, fill = region_id)) +
  geom_bar() +
  theme(legend.position = "none")
```

```{r,eval=FALSE}
tiles_500bp_cg %>% mutate(overlap_region = ifelse(tile_id %in% tile_overlap$tile_id, TRUE, FALSE)) %>%
  ggplot(aes(x = tile_id, fill = tile_id)) +
  geom_bar() +
  theme(legend.position = "none")
```

```{r}
tile_overlap_p <- tile_overlap %>%
  mutate(
    # extract start from each ID  e.g. "chr1:13283-13782" → 13283
    start_fixed = start,   # already have this
    start_op    = as.integer(str_extract(region_id, "(?<=:)\\d+"))
  )
```

## offset distrib

```{r}
tile_overlap_p %>%
  mutate(offset = abs(start_fixed - start_op)) %>%
  ggplot(aes(x = offset)) +
  geom_histogram(bins = 100) +
  labs(x = "start position offset between methods (bp)", y = "count")
```
* spike at 0 means most tiles share a starting position (good)
* bump at ~500 - adjacent tiles

* follow up
```{r}
tile_overlap_p %>%
  mutate(offset = abs(start - start_op)) %>%
  filter(offset > 10, offset < 490) %>%          # exclude the two main spikes
  count(seqnames, sort = TRUE)                     # are they concentrated on specific chromosomes?

tile_overlap_p %>%
  mutate(offset = abs(start - start_op)) %>%
  filter(offset > 10, offset < 490) %>%
  ggplot(aes(x = seqnames)) +
  geom_bar()
```



```{r}
# Or: are they at chromosome ends?
tile_overlap_p %>%
  mutate(offset = abs(start - start_op)) %>%
  filter(offset > 10, offset < 490) %>%
  ggplot(aes(x = start)) +
  geom_histogram(bins = 100) +
  facet_wrap(~seqnames, scales = "free_x")
```

## how many basic tiles share same OPTile

```{r}
tile_overlap_p %>%
  count(region_id, name = "n_fixed_tiles") %>%
  count(n_fixed_tiles) %>%
  ggplot(aes(x = factor(n_fixed_tiles), y = n)) +
  geom_col() +
  labs(x = "number of fixed tiles per operational tile", y = "count of operational tiles")
```

## look at a region

```{r}
tile_overlap_p %>%
  filter(seqnames == "chr1", start < 70000) %>%
  mutate(end_op = start_op + 500) %>%
  ggplot() +
  geom_rect(aes(xmin = start, xmax = end, ymin = 1.1, ymax = 1.9),
            fill = "#4477AA", alpha = 0.7) +
  geom_rect(aes(xmin = start_op, xmax = end_op, ymin = 0.1, ymax = 0.9),
            fill = "#EE6677", alpha = 0.7) +
  scale_y_continuous(breaks = c(0.5, 1.5), labels = c("operational", "fixed")) +
  labs(x = "genomic position", y = NULL) +
  theme_bw()
```


## Summary
* The tiling methods appear to be very similar
* This is good

# DAU analysis with Tiles

```{r}
healthy_samples <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_all_healthy.rds")
cancer_samples <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.rds")

healthy_samples$type <- "healthy"
cancer_samples$type <- "cancer"

all_samples <- rbind(healthy_samples, cancer_samples) %>% data.frame()
```

```{r}
rm(healthy_samples)
rm(cancer_samples)
```


```{r}
# overlap tiles and reads
overlap_pos <- find_overlaps(as_granges(tiles_500bp_cg), as_granges(all_samples)) %>% data.frame()

# dtu

dtu_optile <- overlap_pos %>%
  distinct(sample, read_id, read_start, read_end, alpha, tile_id, type) %>%
  mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                 alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                 alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                 alpha <= 0.6 & alpha <= 0.8 ~ "0.6 <= a <= 0.8",
                                 TRUE ~ "a > 0.8",), 
         transcript = paste0(tile_id, "_", alpha_group)) %>%
  group_by(sample, transcript) %>%
  summarise(transcript_count = n()) %>% ungroup() %>% data.frame()

# build matrix of "transcript" counts - transcript rows, sample as columns
dtu_optile <- dtu_optile %>% pivot_wider(names_from = sample, values_from = transcript_count, values_fill = 0)
dtu_optile$gene_id <- sub("_.*", "", dtu_optile$transcript)

# DTU analysis with edgeR
## Alicia prefers gene method over Simes method (documentation slightly confusing about difference)
y <- DGEList(as.matrix(dtu_optile[,2:13]), group = c(1,1,1,1,1,1,2,2,2,2,2,2), genes = dtu_optile[,c("gene_id", "transcript")])
design <- model.matrix(~ c(1,1,1,1,1,1,2,2,2,2,2,2), data = y$samples)
fit <- glmQLFit(y, design, robust=TRUE)
ds <- diffSpliceDGE(fit, coef = 2, geneid = "gene_id", exonid = "transcript")
dau_basictiles <- topSpliceDGE(ds, test = "gene", number = Inf)


dau_basictiles <- dau_basictiles %>%
  mutate(region_id = gene_id) %>%
  separate_wider_delim(gene_id, delim = ":", names = c("seqnames", "gene_id")) %>%
  separate_wider_delim(gene_id, delim = "-", names = c("start", "end"))
dau_basictiles$start <- as.double(dau_basictiles$start)
dau_basictiles$end <- as.double(dau_basictiles$end)
dau_basictiles$dtu_rank <- 1:nrow(dau_basictiles)
saveRDS(dau_basictiles, "/researchers/caitlin.page/cf_nano/r_output/dtu_basictiles_emseq_cancer_healthy.rds")

```


# Explore results

```{r}
dau_basictiles %>% filter(FDR < 0.05) %>% nrow()
```


```{r}
dau_basictiles %>% mutate(signif = ifelse(FDR < 0.05, TRUE, FALSE), tile = "tile") %>%
  ggplot(aes(x = tile, fill = signif)) +
  geom_bar(position = "fill")
```


## plot reads

```{r}
rrbs_plot5 <- function(read_data, read_ids, cpgs = cg_sites, jit_amount = 0.1) {
  set.seed(42) # For reproducibility
  # Calculate a single jitter amount per row (segment)
  read_data <- read_data %>% dplyr::filter(read_id %in% read_ids)
  jitter_amount <- runif(nrow(read_data), -jit_amount, jit_amount)
  read_data$alpha_jit <- read_data$alpha + jitter_amount
  read_data <- data.frame(read_data)

  # Add jittered coordinates to new columns in the data frame
  long_data <- read_data %>%
    mutate(meth_status = strsplit(meth_pattern, ""), start = strsplit(genom_positions, ",")) %>%
    unnest(meth_status, start) %>%
    mutate(start = as.double(start))

  chr <- unique(read_data$seqnames)
  min_start <- min(read_data$read_start)
  max_end <- max(read_data$read_end)

  read_data <- read_data %>%
    mutate(alpha_group =
             factor(alpha_group,
                    levels = c("a < 0.2", "0.2 <= a <= 0.4",
                               "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8")))

  plot <- read_data %>%
    ggplot() +
    geom_segment(aes(x = read_start, xend = read_end,
                     y = alpha_jit, yend = alpha_jit, colour = alpha_group)) +
    geom_point(data = long_data,
               aes(x = start, y = alpha_jit,
                   fill = meth_status, colour = alpha_group), shape = 21) +
    scale_fill_manual(values = c("U" = "white", "M" = "black")) +
    facet_wrap(~ type, nrow=2) +
    labs(title = paste0(chr, ":", min_start, "-", max_end))
  plot
}
```


```{r}
reads <- find_overlaps(dau_basictiles %>% filter(FDR < 0.05) %>% .[3,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(dau_basictiles %>% filter(FDR < 0.05) %>% .[4,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(dau_basictiles %>% filter(FDR < 0.05) %>% .[5,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(dau_basictiles %>% filter(FDR < 0.05) %>% .[6,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(dau_basictiles %>% filter(FDR < 0.05) %>% .[7,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(dau_basictiles %>% filter(FDR < 0.05) %>% .[8,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(dau_basictiles %>% filter(FDR < 0.05) %>% .[9,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(dau_basictiles %>% filter(FDR < 0.05) %>% .[10,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(dau_basictiles %>% filter(FDR < 0.05) %>% .[11,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)
```


# Compare to results from OpTiles

```{r}
dau_optiles <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_emseq_cancer_healthy.rds")
```

```{r}
dau_basictiles %>% filter(FDR < 0.05) %>% nrow()
dau_optiles %>% filter(FDR < 0.05) %>% nrow()
```


```{r}
tile_overlap %>% filter(tile_id %in% filter(dau_basictiles, FDR < 0.05)$tile_id) %>% distinct(tile_id) %>% nrow()
tile_overlap %>% filter(tile_id %in% filter(dau_basictiles, FDR < 0.05)$tile_id) %>% distinct(region_id) %>% nrow()
```

```{r}
tile_overlap %>% filter(region_id %in% filter(dau_optiles, FDR < 0.05)$region_id) %>% distinct(region_id) %>% nrow()
tile_overlap %>% filter(region_id %in% filter(dau_optiles, FDR < 0.05)$region_id) %>% distinct(tile_id) %>% nrow()
```


```{r}
tile_overlap %>% filter(region_id %in% filter(dau_optiles, FDR < 0.05)$region_id,
                        tile_id %in% filter(dau_basictiles, FDR < 0.05)$tile_id) %>% distinct(region_id, tile_id) %>% nrow()
```


