---
title: "buildFunctions"
author: "caitlinpage"
date: "2025-10-29"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

## Introduction

```{r}
library(plyranges)
library(stringr)
library(BSgenome.Hsapiens.UCSC.hg38)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(tidyr)
library(stringr)
library(dplyr)
library(ggplot2)
library(data.table)
library(duckdb)
library(arrow)

```

* starting a new file because outputs from modkit (nanopore) and bismark (wgbs + rrbs) are very very different
* and I want to get them in the same place

* then this also stays centralised as a way to make sure if I modify pieces other functions still run
* and I love making new files for new ideas
* so use this as an end to end workflow as well


* going from read+cpg to context with beta and alpha

# files
* pat file (with extra info) : gives alpha
* beta file
* og read level: has everything

* technically can get beta from alpha
* nanopore issue was figuring out the filtering metrics from modkit for that to be accurate

* will have to go through the same process for bismark files

# processing files
* from practical standpoint: use datatable and fread because files are LARGE

## helpers
```{r}
cg_sites <- cbind(data.frame(Biostrings::matchPattern("CG", BSgenome.Hsapiens.UCSC.hg38[["chr22"]])), seqnames = "chr22") %>% 
  mutate(pos = paste0(seqnames, "-", start)) %>%
  relocate(pos, seqnames)
cg_sites$index <- 1:nrow(cg_sites)
c(number_of_sites = nrow(cg_sites))
head(cg_sites)
```


## nanopore

* need modkit read file from modkit_extract_full
* bed file from downloaded data (bed file from modkit was weird)

* fread is a nightmare
```{r}
#fread(path_to_modkit_output)
process_modkit <- function(modkit_fread, path_to_bed_file) {
  reads <- modkit_fread
  reads <- reads$ref_position + 1 # convert to 1 based
  reads <- reads[, ref_position := ifelse(strand == "-", ref_position - 1, ref_position)] # convert to 1 based
  reads <- reads[, read_and_pos := paste0(reads$read_id, ":", reads$ref_position)]
  reads$start <- reads$ref_position
  reads$end <- reads$ref_position
  
  bed <- fread(path_to_bed_file)
  
  alpha_and_beta2 <- find_overlaps(as_granges(reads), as_granges(bed)) %>% data.frame() %>% 
    filter(call_prob > 0.66) %>% # solution for alpha and beta match
    group_by(read_id) %>% 
    .[order(.$alignment_start),] %>% 
    mutate(call_code = ifelse(call_code == "m", "M", "U")) %>% 
    mutate(meth_pattern = paste(call_code, collapse = "")) %>% 
    mutate(num_meth = str_count(meth_pattern, pattern = "M"),
         total = str_length(meth_pattern),
         alpha = num_meth/total) %>% 
    mutate(genom_positions = paste0(ref_position, collapse = ",")) %>% 
    ungroup()
  alpha_and_beta2 <- alpha_and_beta2 %>% data.frame()
  alpha_and_beta2
}

modkit_to_pat <- function(alpha_and_beta) {
  alpha_and_beta2_pat <- alpha_and_beta2 %>% 
    mutate(start_pos = sapply(str_split(genom_positions, ","), head, 1),
           end_pos = sapply(str_split(genom_positions, ","), tail, 1)) %>% 
    filter(ref_position == start_pos) %>% 
    distinct(read_id, seqnames, start_pos, meth_pattern, alpha, end_pos, beta) %>% 
    group_by(seqnames, start_pos, meth_pattern, alpha, end_pos, beta) %>% 
    summarise(n=n()) %>% 
    ungroup()

  alpha_and_beta2_pat <- alpha_and_beta2_pat %>% 
    mutate(start_pos = as.integer(start_pos), 
           end_pos = as.integer(end_pos))
  alpha_and_beta2_pat
}
```


## wgbs/rrbs

* fread is a nightmare
* it won't run within a fn - doesn't trust the path
* but it's way faster than any other alternative
* so for now - read in the file with fread and then put that into the fn
```{r}
#read <- fread(file = path_to_bismark_res)
#bed <- fread(file = path_to_bed) # for betas - otherwise will make them myself
process_wgbs <- function(bismark, bed = NULL) {
  read <- bismark
  names(read) <- c("read_id", "strand", "seqnames", "start", "meth_status")
  read$end <- read$start
  
 # if (is.null(bed) == TRUE) {
#    read_overlap <- read
 # } else {
#  read_overlap <- find_overlaps(as_granges(read), as_granges(bed)) %>% data.frame()
 # }
  read_overlap <- read
  read_overlap[, meth_status := ifelse(meth_status == "Z", "M", "U")]
  read_overlap[, meth_pattern := paste0(meth_status, collapse = ""), by = read_id]
  read_overlap[, num_meth := str_count(meth_pattern, pattern = "M")]
  read_overlap[, total := str_length(meth_pattern)]
  read_overlap[, alpha := num_meth/total]
  read_overlap[, genom_positions := paste0(start, collapse = ","), by = read_id]

  read_overlap <- read_overlap %>% .[order(.$read_id),]
  
  betas <- betas_for_wgbs(read_overlap)
  betas <- data.frame(betas)
  read_overlap <- read_overlap %>% group_by(read_id) %>% 
    mutate(read_start = min(start), read_end = max(end), 
           genom_positions = paste0(start, collapse = ",")) %>% 
    ungroup()
  read_overlap <- read_overlap %>% data.frame() %>% 
    mutate(beta = betas[match(.$start, betas$start), "beta"])
  
  read_overlap
}


```
```{r}
betas_for_wgbs <- function(bismark_processed) {
  res <- bismark_processed %>% group_by(start, meth_status) %>% 
  .[order(.$start),] %>%
  mutate(meth_count_pos = n(), unmeth_count_pos = n()) %>%
  mutate(meth_count_pos = ifelse(meth_status == "M", meth_count_pos, NA),
         unmeth_count_pos = ifelse(meth_status == "U", unmeth_count_pos, NA)) %>% 
    ungroup() %>% 
    distinct(seqnames, start, end, meth_count_pos, unmeth_count_pos) %>%  
 group_by(start) %>% 
    .[order(.$meth_count_pos),] %>%
  fill(meth_count_pos, .direction = "down") %>%
  fill(unmeth_count_pos, .direction = "up") %>% 
  replace(is.na(.), 0) %>%
  mutate(beta = meth_count_pos/(meth_count_pos + unmeth_count_pos)) %>%
  ungroup() %>%
  distinct(seqnames, start, end, meth_count_pos, unmeth_count_pos, beta)
  
  res
}
```
```{r}
pat_for_wgbs <- function(process_res) {
  pat <- process_res %>% 
  mutate(start_pos = sapply(str_split(genom_positions, ","), head, 1),
         end_pos = sapply(str_split(genom_positions, ","), tail, 1)) %>% 
  filter(start == start_pos) %>%
  distinct(read_id, seqnames, start_pos, meth_pattern, alpha, end_pos, beta) %>% 
  group_by(seqnames, start_pos, meth_pattern, alpha, end_pos, beta) %>% 
  summarise(n=n()) %>%
  ungroup()  %>%
  mutate(start_pos = as.integer(start_pos),
         end_pos = as.integer(end_pos))
  
  pat
}
```

```{r}
bis <- fread("/researchers/caitlin.page/rrbs_encode_cell_line/CpG_OT_ENCFF000LYO_bismark_bt2.txt")
for_beta <- process_wgbs(bis)
for_beta <- for_beta %>% filter(seqnames == "chr22")
rrbs_for_beta_pat <- pat_for_wgbs(for_beta)
```

# dau analysis
```{r}
# overlap tiles and reads
overlap_pos <- find_overlaps(as_granges(tiles_500bp_cg), as_granges(all_samples)) %>% data.frame()

# dtu

dtu_optile <- overlap_pos %>%
  distinct(sample, read_id, read_start, read_end, alpha, tile_id, type) %>%
  mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                 alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                 alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                 alpha <= 0.6 & alpha <= 0.8 ~ "0.6 <= a <= 0.8",
                                 TRUE ~ "a > 0.8"), 
         transcript = paste0(tile_id, "_", alpha_group)) %>%
  group_by(sample, transcript) %>%
  summarise(transcript_count = n()) %>% ungroup() %>% data.frame()

# build matrix of "transcript" counts - transcript rows, sample as columns
dtu_optile <- dtu_optile %>% pivot_wider(names_from = sample, values_from = transcript_count, values_fill = 0)
dtu_optile$gene_id <- sub("_.*", "", dtu_optile$transcript)

# DTU analysis with edgeR
## Alicia prefers gene method over Simes method (documentation slightly confusing about difference)
y <- DGEList(as.matrix(dtu_optile[,2:13]), group = c(1,1,1,1,1,1,2,2,2,2,2,2), genes = dtu_optile[,c("gene_id", "transcript")])
design <- model.matrix(~ c(1,1,1,1,1,1,2,2,2,2,2,2), data = y$samples)
fit <- glmQLFit(y, design, robust=TRUE)
ds <- diffSpliceDGE(fit, coef = 2, geneid = "gene_id", exonid = "transcript")
dau_basictiles <- topSpliceDGE(ds, test = "gene", number = Inf)


dau_basictiles <- dau_basictiles %>%
  mutate(region_id = gene_id) %>%
  separate_wider_delim(gene_id, delim = ":", names = c("seqnames", "gene_id")) %>%
  separate_wider_delim(gene_id, delim = "-", names = c("start", "end"))
dau_basictiles$start <- as.double(dau_basictiles$start)
dau_basictiles$end <- as.double(dau_basictiles$end)
dau_basictiles$dtu_rank <- 1:nrow(dau_basictiles)
```



# plotting

```{r}
plot_fn <- function(chrom = chrom, region = c(start, end), data_big = data_big, cg_sites = cg_sites) {
  chr <- chrom
  start <- region[1]
  end <- region[2]
  
  data <- data_big
  cpgs <- cg_sites
  
  data <- data %>% 
    filter(seqnames == chr,
           start >= start,
           end <= end) 
  data2 <- data %>%
    distinct(read_id, alignment_start, alignment_end, alpha)
  
  plot <- data2 %>% 
    ggplot() +
    geom_segment(aes(x = alignment_start, xend = alignment_end, #read length
                     y = alpha, yend = alpha), alpha = 0.2) +
    geom_point(data = data, 
             aes(x = ref_position, y = alpha, colour = call_code)) + # alpha dots, 1 per read
    geom_point(data = distinct(data, ref_position, beta),            # beta as X
            aes(x = ref_position, y = beta), shape = 4, size = 3) +
    geom_segment(data = filter(cpgs, start >= start, end <= end),    # green lines showing cpg positions
                 aes(x = start, xend = start, y = -0.15, yend = -0.05), colour = "green")
  plot
}
```

```{r}
# provide vector of read ids
rrbs_plot <- function(long_data, read_ids, cpgs = cg_sites) {
  long_data %>% filter(read_id %in% read_ids) %>% 
    distinct(read_id, read_start, read_end, alpha) %>%
  ggplot() +
  geom_segment(aes(x = read_start, xend = read_end, y = alpha, yend = alpha), alpha = 0.2) +
  geom_point(data = filter(long_data, read_id %in% read_ids), 
             aes(x = start, y = alpha, colour = meth_status)) +
  geom_point(data = distinct(filter(long_data, read_id %in% read_ids), start, beta), 
            aes(x = start, y = beta), shape = 4, size = 3) +
  geom_segment(data = filter(cg_sites, start >= min(filter(long_data, read_id %in% read_ids)$start), 
                             end <= max(filter(long_data, read_id %in% read_ids)$end)), 
               aes(x = start, xend = start, y = -0.15, yend = -0.05), colour = "green")
}
```


# plotting layers
* I'm borrowing from my previous work with Damsel
* create ggplot2 layers that I can then pile up together
* probably also want a function that just plots them all together - given that I'm having some trouble with layers

## reads
* coloured by alpha band

```{r}
rrbs_plot5 <- function(read_data, read_ids, cpgs = cg_sites, jit_amount = 0.1) {
  set.seed(42) # For reproducibility
  # Calculate a single jitter amount per row (segment)
  read_data <- read_data %>% dplyr::filter(read_id %in% read_ids)
  jitter_amount <- runif(nrow(read_data), -jit_amount, jit_amount)
  read_data$alpha_jit <- read_data$alpha + jitter_amount
  read_data <- data.frame(read_data)

  # Add jittered coordinates to new columns in the data frame
  long_data <- read_data %>%
    mutate(meth_status = strsplit(meth_pattern, ""), start = strsplit(genom_positions, ",")) %>%
    unnest(meth_status, start) %>%
    mutate(start = as.double(start))
  
  chr <- unique(read_data$seqnames)
  min_start <- min(read_data$read_start)
  max_end <- max(read_data$read_end)
  
  read_data <- read_data %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8")))
  
  plot <- read_data %>%
    ggplot() +
    geom_segment(aes(x = read_start, xend = read_end, y = alpha_jit, yend = alpha_jit, colour = alpha_group)) +
    geom_point(data = long_data,
               aes(x = start, y = alpha_jit, fill = meth_status, colour = alpha_group), shape = 21) +
    scale_fill_manual(values = c("U" = "white", "M" = "black")) +
    facet_wrap(~ type, nrow=2) +
    labs(title = paste0(chr, ":", min_start, "-", max_end)) 
  plot
}
```

```{r}
rrbs_plot5 <- function(read_data, read_ids, cpgs = cg_sites, jit_amount = 0.1) {
  set.seed(42) # For reproducibility
  # Calculate a single jitter amount per row (segment)
  read_data <- read_data %>% dplyr::filter(read_id %in% read_ids)
  jitter_amount <- runif(nrow(read_data), -jit_amount, jit_amount)
  read_data$alpha_jit <- read_data$alpha + jitter_amount
  read_data <- data.frame(read_data)

  # Add jittered coordinates to new columns in the data frame
  long_data <- read_data %>%
    mutate(meth_status = strsplit(meth_pattern, ""), start = strsplit(genom_positions, ",")) %>%
    unnest(meth_status, start) %>%
    mutate(start = as.double(start))

  chr <- unique(read_data$seqnames)
  min_start <- min(read_data$read_start)
  max_end <- max(read_data$read_end)

  plot <- read_data %>%
    ggplot() +
    geom_segment(aes(x = read_start, xend = read_end,
                     y = alpha_jit, yend = alpha_jit, colour = alpha_group)) +
    geom_point(data = long_data,
               aes(x = start, y = alpha_jit,
                   fill = meth_status, colour = alpha_group), shape = 21) +
    scale_fill_manual(values = c("U" = "white", "M" = "black")) +
    facet_wrap(~ type, nrow=2) +
    labs(title = paste0(chr, ":", min_start, "-", max_end))
  plot
}
```

*rewritten efficient function (hopefully)
```{r}
rrbs_plot5 <- function(read_data, read_ids, cpgs = cg_sites, jit_amount = 0.1) {
  set.seed(42)

  # ── 1. Fast filter ────────────────────────────────────────────────────────
  # data.table key lookup is O(log n) vs dplyr's O(n) scan on 700M rows.
  # Call setkey(read_data, read_id) ONCE after loading your data, outside
  # this function, so the sort cost isn't paid on every plot call.
  if (!inherits(read_data, "data.table")) {
    data.table::setDT(read_data)          # in-place: no 700M-row copy
  }

  if (!is.null(data.table::key(read_data)) &&
      data.table::key(read_data)[1] == "read_id") {
    filtered <- read_data[.(read_ids), nomatch = NULL]   # binary-search join
  } else {
    filtered <- read_data[read_id %chin% read_ids]       # %chin% = fast string %in%
  }

  # ── 2. Jitter (in-place, no copy) ─────────────────────────────────────────
  n <- nrow(filtered)
  data.table::set(filtered, j = "alpha_jit",
                  value = filtered$alpha + runif(n, -jit_amount, jit_amount))

  # ── 3. Summary stats before expansion (cheap on filtered rows) ────────────
  chr       <- filtered[, unique(seqnames)]
  min_start <- filtered[, min(read_start)]
  max_end   <- filtered[, max(read_end)]

  # ── 4. Vectorised expansion — replaces strsplit+mutate+unnest ─────────────
  # strsplit returns a list; unlist + rep lets us do it in one vectorised pass
  # instead of unnest()'s slow row-by-row expansion.
  ms_list <- strsplit(filtered$meth_pattern,  "")
  st_list <- strsplit(filtered$genom_positions, ",")
  lens    <- lengths(ms_list)              # number of CpGs per read
  idx     <- rep(seq_len(n), lens)         # row index for each expanded position

  long_data <- data.table::data.table(
    meth_status = unlist(ms_list, use.names = FALSE),
    start       = as.double(unlist(st_list, use.names = FALSE)),
    alpha_jit   = filtered$alpha_jit[idx],
    alpha_group = filtered$alpha_group[idx],
    type        = filtered$type[idx]
  )

  # ── 5. Plot (ggplot works directly with data.table) ───────────────────────
  ggplot(filtered) +
    geom_segment(aes(x = read_start, xend = read_end,
                     y = alpha_jit,  yend = alpha_jit,
                     colour = alpha_group)) +
    geom_point(data = long_data,
               aes(x = start, y = alpha_jit,
                   fill = meth_status, colour = alpha_group),
               shape = 21) +
    scale_fill_manual(values = c("U" = "white", "M" = "black")) +
    facet_wrap(~ type, nrow = 2) +
    labs(title = paste0(chr, ":", min_start, "-", max_end))
}
data.table::setDT(read_data)
data.table::setkey(read_data, read_id)
```

* using paraquet for the read file because it's big
* meaning that the read file won't be in memory - paraquet calls the bits that are needed each time
* oops except can't be using the read ids - don't have them without the file which I don't want to read in
* could give region id though
```{r}
rrbs_plot5 <- function(read_data_path, chrom, region_start, region_end, cpgs = cg_sites,
                       jit_amount = 0.1) {
  set.seed(42)

  # ── 1. Read only the matching rows — never loads unmatched data into RAM ───
  con <- dbConnect(duckdb())
  on.exit(dbDisconnect(con))
  
  filtered <- dbGetQuery(con, glue::glue("
    SELECT *
    FROM read_parquet('{read_data_path}')
    WHERE seqnames = '{chrom}'
      AND \"start\" >= {region_start}
      AND \"end\"   <= {region_end}
  "))
  
  healthy_samples <- c("a1", "a2", "a3", "b1", "b2", "b3")
  filtered$type   <- ifelse(filtered$sample %in% healthy_samples, "healthy", "cancer")
  
  # ── 2. Jitter ─────────────────────────────────────────────────────────────
  n <- nrow(filtered)
  filtered$alpha_jit <- filtered$alpha + runif(n, -jit_amount, jit_amount)

  # ── 3. Summary stats ──────────────────────────────────────────────────────
  chr       <- unique(filtered$seqnames)
  min_start <- min(filtered$read_start)
  max_end   <- max(filtered$read_end)

  # ── 4. Vectorised expansion ───────────────────────────────────────────────
  ms_list <- strsplit(filtered$meth_pattern, "")
  st_list <- strsplit(gsub(",+$", "", filtered$genom_positions), ",")  # strip trailing commas
  
  # Drop any rows where lengths still don't match
  lens_ms <- lengths(ms_list)
  lens_st <- lengths(st_list)
  keep    <- lens_ms == lens_st
  
  if (!all(keep)) {
    message(sprintf("Dropping %d rows with mismatched meth_pattern/genom_positions lengths",
                    sum(!keep)))
    filtered  <- filtered[keep, ]
    ms_list   <- ms_list[keep]
    st_list   <- st_list[keep]
    lens      <- lens_ms[keep]
  } else {
    lens <- lens_ms
  }
  
  n   <- nrow(filtered)
  idx <- rep(seq_len(n), lens)

  long_data <- data.frame(
    meth_status = unlist(ms_list, use.names = FALSE),
    start       = as.double(unlist(st_list, use.names = FALSE)),
    alpha_jit   = filtered$alpha_jit[idx],
    alpha_group = filtered$alpha_group[idx],
    type        = filtered$type[idx]
  )

  # ── 5. Plot ───────────────────────────────────────────────────────────────
  ggplot(filtered) +
    geom_segment(aes(x = read_start, xend = read_end,
                     y = alpha_jit,  yend = alpha_jit,
                     colour = alpha_group)) +
    geom_point(data = long_data,
               aes(x = start, y = alpha_jit,
                   fill = meth_status, colour = alpha_group),
               shape = 21) +
    scale_fill_manual(values = c("U" = "white", "M" = "black")) +
    facet_wrap(~ type, nrow = 2) +
    labs(title = paste0(chr, ":", min_start, "-", max_end))
}
```

* mix of claude to get the reads in - and then back to my og because more "efficient" is not running
```{r}
rrbs_plot5 <- function(read_data_path_a, read_data_path_b, chrom, region_start, region_end, cpgs = cg_sites,
                       jit_amount = 0.1) {
  set.seed(42)

  # ── 1. Read only the matching rows — never loads unmatched data into RAM ───
  con <- dbConnect(duckdb())
  on.exit(dbDisconnect(con))
  
  read_data_a <- dbGetQuery(con, glue::glue("
    SELECT *
    FROM read_parquet('{read_data_path_a}')
    WHERE seqnames = '{chrom}'
      AND \"start\" >= {region_start}
      AND \"end\"   <= {region_end}
  "))
  
  read_data_b <- dbGetQuery(con, glue::glue("
    SELECT *
    FROM read_parquet('{read_data_path_b}')
    WHERE seqnames = '{chrom}'
      AND \"start\" >= {region_start}
      AND \"end\"   <= {region_end}
  "))
  
  read_data_a$type <- "healthy"
  read_data_b$type <- "cancer"
  
  read_data <- rbind(read_data_a, read_data_b)
  
  jitter_amount <- runif(nrow(read_data), -jit_amount, jit_amount)
  read_data$alpha_jit <- read_data$alpha + jitter_amount
  read_data <- data.frame(read_data)
  
  read_data <- read_data %>% mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                 alpha > 0.8 ~ "a > 0.8",
                                 alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                 alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                 TRUE ~ "0.6 <= a <= 0.8"))
  read_data <- read_data %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8")))
  
  # Add jittered coordinates to new columns in the data frame
  long_data <- read_data %>%
    mutate(meth_status = strsplit(meth_pattern, ""), start = strsplit(genom_positions, ",")) %>%
    unnest(meth_status, start) %>%
    mutate(start = as.double(start))

  chr <- unique(read_data$seqnames)
  min_start <- min(read_data$read_start)
  max_end <- max(read_data$read_end)
  
  plot <- read_data %>%
    ggplot() +
    geom_segment(aes(x = read_start, xend = read_end,
                     y = alpha_jit, yend = alpha_jit, colour = alpha_group)) +
    geom_point(data = long_data,
               aes(x = start, y = alpha_jit,
                   fill = meth_status, colour = alpha_group), shape = 21) +
    scale_fill_manual(values = c("U" = "white", "M" = "black")) +
    facet_wrap(~ type, nrow=2) +
    labs(title = paste0(chr, ":", min_start, "-", max_end)) +
    theme_bw() +
    labs(x = "genomic position")
  plot
}
```

```{r}
#library(ggrastr)
rrbs_plot5_poster_svg <- function(read_data_path_a, read_data_path_b, chrom, region_start, region_end, cpgs = cg_sites,
                       jit_amount = 0.1) {
  set.seed(42)

  # ── 1. Read only the matching rows — never loads unmatched data into RAM ───
  con <- dbConnect(duckdb())
  on.exit(dbDisconnect(con))
  
  read_data_a <- dbGetQuery(con, glue::glue("
    SELECT *
    FROM read_parquet('{read_data_path_a}')
    WHERE seqnames = '{chrom}'
      AND \"start\" >= {region_start}
      AND \"end\"   <= {region_end}
  "))
  
  read_data_b <- dbGetQuery(con, glue::glue("
    SELECT *
    FROM read_parquet('{read_data_path_b}')
    WHERE seqnames = '{chrom}'
      AND \"start\" >= {region_start}
      AND \"end\"   <= {region_end}
  "))
  
  read_data_a$type <- "healthy"
  read_data_b$type <- "cancer"
  
  read_data <- rbind(read_data_a, read_data_b)
  
  jitter_amount <- runif(nrow(read_data), -jit_amount, jit_amount)
  read_data$alpha_jit <- read_data$alpha + jitter_amount
  read_data <- data.frame(read_data)
  
  read_data <- read_data %>% mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                 alpha > 0.8 ~ "a > 0.8",
                                 alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                 alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                 TRUE ~ "0.6 <= a <= 0.8"))
  read_data <- read_data %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8")))
  
  # Add jittered coordinates to new columns in the data frame
  long_data <- read_data %>%
    mutate(meth_status = strsplit(meth_pattern, ""), start = strsplit(genom_positions, ",")) %>%
    unnest(meth_status, start) %>%
    mutate(start = as.double(start))

  chr <- unique(read_data$seqnames)
  min_start <- min(read_data$read_start)
  max_end <- max(read_data$read_end)
  
  plot <- read_data %>%
    ggplot() +
    rasterise(geom_segment(aes(x = read_start, xend = read_end,
                     y = alpha_jit, yend = alpha_jit, colour = alpha_group)), dpi = 300) +
    rasterise(geom_point(data = long_data,
               aes(x = start, y = alpha_jit,
                   fill = meth_status, colour = alpha_group), shape = 21), dpi = 300) +
    scale_fill_manual(values = c("U" = "white", "M" = "black")) +
    facet_wrap(~ type, nrow=2) +
    labs(title = paste0(chr, ":", min_start, "-", max_end)) +
    theme_bw() +
    labs(x = "genomic position")
  plot
}
```


## beta

```{r}
geom_beta <- function(all_b, style = "line", plot.space=1, plot.height=1) {
    structure(list(all_b = all_b, style = style, plot.space = plot.space,
        plot.height = plot.height),
        class = "beta"
    )
}

ggplot_add.beta <- function(object, plot, object_name) {
plot2 <- plot
    while (inherits(plot2, "patchwork")) {
        plot2 <- plot2[[1]]
    }
    plot.data <- plot2$labels$title
    plot.data <- stringr::str_split_1(plot.data, ":")
    # prepare plot range
    # the plot region are not normal, so start is minimum value
    plot.chr <- plot.data[1]
    plot.data <- stringr::str_split_1(plot.data[2], "-")
    plot.region.start <- plot.data[1] %>% as.numeric()
    plot.region.end <- plot.data[2] %>% as.numeric()
    
    all_b <- object$all_b
    plot.space <- object$plot.space
    plot.height <- object$plot.height
    style <- object$style
    
 #   all_b <- all_b %>%
#  tidyr::pivot_longer(cols = c(cancer_0223, cancer_0224, cancer_0225, cancer_0454, cancer_0456, cancer_0457, healthy_036, healthy_037, healthy_785, healthy_786, healthy_787, healthy_789), names_to = "sample", values_to = "beta") %>% mutate(type = substr(sample, 1, 1)) %>% filter(seqnames.x %in% plot.chr,
       #                         start.x >= plot.region.start,
      #                          start.x <= plot.region.end)
    all_b <- all_b %>%
      mutate(type = substr(sample, 1, 1)) %>%
      filter(seqnames %in% plot.chr,
                                start >= plot.region.start,
                                start <= plot.region.end)
    
    if(style == "line") {
      b_plot <- all_b %>%
  ggplot(aes(x = start, y = beta, colour = type)) +
  geom_line() +
  theme_bw() +
        scale_color_manual(values = c("c" = "#C0622A", "h" = "#1B3A6B"),
                           labels = c("c" = "cancer", "h" = "healthy")) +
        labs(x = "genomic position")
    } else if (style == "smooth") {
      b_plot <- all_b %>%
  ggplot(aes(x = start, y = beta, colour = type)) +
        geom_smooth() +
        theme_bw() +
        scale_color_manual(values = c("c" = "#C0622A", "h" = "#1B3A6B"),
                           labels = c("c" = "cancer", "h" = "healthy")) +
        labs(x = "genomic position")
    }
    else {
    b_plot <- all_b %>%
  ggplot(aes(x = start, y = beta, colour = type)) +
  geom_point(alpha = 0.3) +
      theme_bw() +
      scale_color_manual(values = c("c" = "#C0622A", "h" = "#1B3A6B"),
                         labels = c("c" = "cancer", "h" = "healthy")) +
      labs(x = "genomic position")
    }
    
    patchwork::wrap_plots(plot + ggplot2::theme(plot.margin = ggplot2::margin(t = plot.space, b = plot.space)),
        b_plot,
        ncol = 1, heights = c(1, plot.height)
    )
}
```


## theme for dau/dmr
```{r}
..peakGatcTheme <- function(margin.len, x.start, x.end) {
    list(
        ggplot2::scale_x_continuous(limits = c(x.start, x.end)),
        ggplot2::theme_classic(),
        ggplot2::theme(
            axis.line.y = ggplot2::element_blank(),
            axis.text.y = ggplot2::element_blank(),
            axis.title.y.right = ggplot2::element_text(color = "black", angle = 90, vjust = 0.5),
            axis.ticks.y = ggplot2::element_blank(),
            axis.text.x = ggplot2::element_blank(),
            axis.title.x = ggplot2::element_blank(),
            axis.ticks.x = ggplot2::element_blank(),
            panel.border = ggplot2::element_rect(colour = "black", fill = NA, linewidth = 1),
            plot.margin = ggplot2::margin(t = margin.len, b = margin.len)
        ),
        ggplot2::scale_y_continuous(
            limits = c(1 - 0.1, 1 + 0.1),
            expand = c(0, 0), position = "right"
        )
        #ggplot2::scale_x_continuous(expand = c(0, 0)),
        #ggplot2::coord_cartesian(xlim = x.range)
    )
}
```

## dau

```{r}
geom_dau <- function(
    dau.df=NULL, dau.label=FALSE, dau.color="black", dau.size=5,
    plot.space=0.1, plot.height=0.05) {
    structure(
        list(
            dau.df = dau.df, dau.label = dau.label, dau.color = dau.color, dau.size = dau.size,
            plot.space = plot.space, plot.height = plot.height
        ),
        class = "dau"
    )
}


#' @export
ggplot_add.dau <- function(object, plot, object_name) {
    if (!is.data.frame(object$dau.df)) {
        stop("data.frame of daus is required")
    }
    plot2 <- plot
    while (inherits(plot2, "patchwork")) {
        plot2 <- plot2[[1]]
    }
    plot.data <- plot2$labels$title
    plot.data <- stringr::str_split_1(plot.data, ":")

    plot.chr <- plot.data[1]
    plot.data <- stringr::str_split_1(plot.data[2], "-")
    plot.region.start <- plot.data[1] %>% as.numeric()
    plot.region.end <- plot.data[2] %>% as.numeric()

    # get parameters
    dau.df <- object$dau.df
    dau.label <- object$dau.label
    dau.color <- object$dau.color
    dau.size <- object$dau.size
    plot.space <- object$plot.space
    plot.height <- object$plot.height

    dau.df <- dau.df %>% filter(seqnames %in% plot.chr,
                                start >= plot.region.start,
                                end <= plot.region.end, 
                                FDR < 0.05) 
  
    dau.plot <- dau.df %>%
            ggplot2::ggplot(ggplot2::aes(x = .data$start, y = 1)) +
            ggplot2::geom_segment(
                mapping = ggplot2::aes(
                    x = .data$start, y = 1,
                    xend = .data$end, yend = 1
                ),
                linewidth = dau.size, color = dau.color
            ) +
      labs(y = "dau") +
      scale_x_continuous(limits = c(plot.region.start, plot.region.end)) +
      ..peakGatcTheme(margin.len = plot.space, x.start = plot.region.start, x.end = plot.region.end)

    # assemble plot
    patchwork::wrap_plots(plot + ggplot2::theme(plot.margin = ggplot2::margin(t = plot.space, b = plot.space)),
        dau.plot,
        ncol = 1, heights = c(1, plot.height)
    )
}
```


## dmr

```{r}
geom_dmr <- function(
    dmr.df=NULL, dmr.label=FALSE, dmr.color="black", dmr.size=5,
    plot.space=0.1, plot.height=0.05) {
    structure(
        list(
            dmr.df = dmr.df, dmr.label = dmr.label, dmr.color = dmr.color, dmr.size = dmr.size,
            plot.space = plot.space, plot.height = plot.height
        ),
        class = "dmr"
    )
}


#' @export
ggplot_add.dmr <- function(object, plot, object_name) {
    if (!is.data.frame(object$dmr.df)) {
        stop("data.frame of dmr is required")
    }
    plot2 <- plot
    while (inherits(plot2, "patchwork")) {
        plot2 <- plot2[[1]]
    }
    plot.data <- plot2$labels$title
    plot.data <- stringr::str_split_1(plot.data, ":")

    plot.chr <- plot.data[1]
    plot.data <- stringr::str_split_1(plot.data[2], "-")
    plot.region.start <- plot.data[1] %>% as.numeric()
    plot.region.end <- plot.data[2] %>% as.numeric()

    # get parameters
    dmr.df <- object$dmr.df
    dmr.label <- object$dmr.label
    dmr.color <- object$dmr.color
    dmr.size <- object$dmr.size
    plot.space <- object$plot.space
    plot.height <- object$plot.height
    
    dmr.df <- dmr.df %>% filter(seqnames %in% plot.chr,
                                start >= plot.region.start,
                                end <= plot.region.end)
    
   # dmr.df <-  dmr.df %>% mutate(score = c(1,1,1)) 
    dmr.plot <- dmr.df %>%
            ggplot2::ggplot(ggplot2::aes(x = .data$start, y = 1)) +
            ggplot2::geom_segment(
                mapping = ggplot2::aes(
                    x = .data$start, y = 1,
                    xend = .data$end, yend = 1
                ),
                linewidth = dmr.size, color = dmr.color
            ) +
      labs(y = "dmr") +
      ..peakGatcTheme(margin.len = plot.space, x.start = plot.region.start, x.end = plot.region.end)

    # assemble plot
    patchwork::wrap_plots(plot + ggplot2::theme(plot.margin = ggplot2::margin(t = plot.space, b = plot.space)),
        dmr.plot,
        ncol = 1, heights = c(1, plot.height)
    )
}
```


## reads 

