---
title: "compareNoveltyPPGL"
author: "caitlinpage"
date: "2026-09-03"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

## Introduction

```{r}
dau_df_out
```

```{r}
dmr_df
```

```{r}
dau_gr %>% data.frame()
```


```{r}
# ── Compare novelty results: DMR vs DAU (Known PPGL-Specific Loci) ─────────────

library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(ggvenn)
library(GenomicRanges)
```


```{r}
# ── 1. Novelty class summary ───────────────────────────────────────────────────
novelty_summary <- bind_rows(
  dmr_df     |> dplyr::count(novelty_class) |> mutate(method = "DMR"),
  dau_df_out |> dplyr::count(novelty_class) |> mutate(method = "DAU")
) |>
  group_by(method) |>
  mutate(pct = n / sum(n) * 100) |>
  ungroup() |>
  mutate(
    novelty_class = factor(novelty_class),
    method = factor(method, levels = c("DMR", "DAU"))
  )

print(novelty_summary)
```


```{r}
# ── 2. Bar chart: novelty tier breakdown ──────────────────────────────────────
tier_colours <- c(
  "Tier 1: Novel"  = "#E07A5F",
  "Tier 2: Novel"  = "#F2A65A",
  "Tier 3: Known"  = "#4E8098",
  "Tier 4: Known"  = "#81B29A"
  # extend as needed — levels will auto-match from your data
)

p_bar <- novelty_summary |>
  ggplot(aes(x = method, y = pct, fill = novelty_class)) +
  geom_col(width = 0.6, colour = "white", linewidth = 0.4) +
  geom_text(aes(label = sprintf("%d\n(%.1f%%)", n, pct)),
            position = position_stack(vjust = 0.5),
            size = 3.2, colour = "white", fontface = "bold") +
  scale_fill_manual(values = tier_colours, name = NULL) +
  labs(title = "Known PPGL loci: novelty tier distribution",
       x = NULL, y = "% of significant hits") +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 9))
p_bar
```


```{r}
# ── 3. Gene-level comparison (DMR genes vs DAU-overlapping genes) ─────────────
# Split comma-separated genes in DMR
dmr_genes_long <- dmr_df |>
  select(dmr_id, overlapping.genes, novelty_class) |>
  separate_rows(overlapping.genes, sep = ",\\s*") |>
  filter(overlapping.genes != "", !is.na(overlapping.genes)) |>
  rename(gene = overlapping.genes)

# Annotate DAU windows with overlapping DMR genes via genomic ranges
dmr_gr <- makeGRangesFromDataFrame(dmr_df, keep.extra.columns = TRUE)
dau_gr <- makeGRangesFromDataFrame(dau_df_out, keep.extra.columns = TRUE)

hits <- findOverlaps(dau_gr, dmr_gr)

dau_with_genes <- dau_df_out[queryHits(hits), c("dau_id", "novelty_class")] |>
  bind_cols(
    dmr_df[subjectHits(hits), "overlapping.genes", drop = FALSE] |>
      rename(genes_from_dmr = overlapping.genes)
  ) |>
  separate_rows(genes_from_dmr, sep = ",\\s*") |>
  filter(genes_from_dmr != "", !is.na(genes_from_dmr)) |>
  rename(gene = genes_from_dmr) |>
  distinct()

# Novel genes per method
novel_dmr_genes <- dmr_genes_long |>
  filter(grepl("Novel", novelty_class)) |>
  pull(gene) |> unique() |> sort()

novel_dau_genes <- dau_with_genes |>
  filter(grepl("Novel", novelty_class)) |>
  pull(gene) |> unique() |> sort()

known_dmr_genes <- dmr_genes_long |>
  filter(grepl("Known", novelty_class)) |>
  pull(gene) |> unique() |> sort()

known_dau_genes <- dau_with_genes |>
  filter(grepl("Known", novelty_class)) |>
  pull(gene) |> unique() |> sort()
```


```{r}
# ── 4. Venn diagrams ───────────────────────────────────────────────────────────
p_venn_novel <- ggvenn(
  list(DMR = novel_dmr_genes, DAU = novel_dau_genes),
  fill_color = c("#E07A5F", "#81B29A"),
  stroke_size = 0.5, set_name_size = 4.5, text_size = 4
) + labs(title = "Novel genes")

p_venn_known <- ggvenn(
  list(DMR = known_dmr_genes, DAU = known_dau_genes),
  fill_color = c("#4E8098", "#F2CC8F"),
  stroke_size = 0.5, set_name_size = 4.5, text_size = 4
) + labs(title = "Known PPGL genes recovered")
p_venn_novel
p_venn_known
```


```{r}
# ── 5. Genomic overlap: novel DMR regions vs novel DAU windows ─────────────────
novel_dmr_gr <- dmr_gr[grepl("Novel", dmr_gr$novelty_class)]
novel_dau_gr <- dau_gr[grepl("Novel", dau_gr$novelty_class)]

novel_overlaps <- findOverlaps(novel_dmr_gr, novel_dau_gr)

cat(sprintf(
  "\nNovel DMR regions:     %d\nNovel DAU windows:     %d\nDMR regions with ≥1 novel DAU window: %d (%.1f%%)\n",
  length(novel_dmr_gr),
  length(novel_dau_gr),
  length(unique(queryHits(novel_overlaps))),
  100 * length(unique(queryHits(novel_overlaps))) / length(novel_dmr_gr)
))
```


```{r}
# ── 6. Print gene lists ────────────────────────────────────────────────────────
novel_both  <- intersect(novel_dmr_genes, novel_dau_genes)
novel_dmr_only <- setdiff(novel_dmr_genes, novel_dau_genes)
novel_dau_only <- setdiff(novel_dau_genes, novel_dmr_genes)

cat("\n── Novel in BOTH (high-confidence) n =", length(novel_both), "──\n")
print(novel_both)

cat("\n── Novel in DMR only n =", length(novel_dmr_only), "──\n")
print(novel_dmr_only)

cat("\n── Novel in DAU only n =", length(novel_dau_only), "──\n")
print(novel_dau_only)
```


```{r}
# ── 7. Assemble figure ─────────────────────────────────────────────────────────
p_final <- p_bar | (p_venn_novel / p_venn_known)
p_final + plot_annotation(
  title = "Known PPGL-Specific Loci: DMR vs DAU novelty comparison",
  theme = theme(plot.title = element_text(size = 14, face = "bold"))
)
```


# repeat on the filtered dau results - was wrong before
```{r}
dau_df_out
dau_df_out2
```


```{r}
# ── 1. Novelty class summary ───────────────────────────────────────────────────
novelty_summary <- bind_rows(
  dmr_df     |> dplyr::count(novelty_class) |> mutate(method = "DMR"),
  dau_df_out2 |> dplyr::count(novelty_class) |> mutate(method = "DAU")
) |>
  group_by(method) |>
  mutate(pct = n / sum(n) * 100) |>
  ungroup() |>
  mutate(
    novelty_class = factor(novelty_class),
    method = factor(method, levels = c("DMR", "DAU"))
  )

print(novelty_summary)
```


```{r}
# ── 2. Bar chart: novelty tier breakdown ──────────────────────────────────────
tier_colours <- c(
  "Tier 1: Novel"  = "#E07A5F",
  "Tier 2: Novel"  = "#F2A65A",
  "Tier 3: Known"  = "#4E8098",
  "Tier 4: Known"  = "#81B29A"
  # extend as needed — levels will auto-match from your data
)

p_bar <- novelty_summary |>
  ggplot(aes(x = method, y = pct, fill = novelty_class)) +
  geom_col(width = 0.6, colour = "white", linewidth = 0.4) +
  geom_text(aes(label = sprintf("%d\n(%.1f%%)", n, pct)),
            position = position_stack(vjust = 0.5),
            size = 3.2, colour = "white", fontface = "bold") +
  scale_fill_manual(values = tier_colours, name = NULL) +
  labs(title = "Known PPGL loci: novelty tier distribution",
       x = NULL, y = "% of significant hits") +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 9))
p_bar
```


```{r}
# ── 3. Gene-level comparison (DMR genes vs DAU-overlapping genes) ─────────────
# Split comma-separated genes in DMR
dmr_genes_long <- dmr_df |>
  select(dmr_id, overlapping.genes, novelty_class) |>
  separate_rows(overlapping.genes, sep = ",\\s*") |>
  filter(overlapping.genes != "", !is.na(overlapping.genes)) |>
  rename(gene = overlapping.genes)

# Annotate DAU windows with overlapping DMR genes via genomic ranges
dmr_gr <- makeGRangesFromDataFrame(dmr_df, keep.extra.columns = TRUE)
dau_gr <- makeGRangesFromDataFrame(dau_df_out2, keep.extra.columns = TRUE)

hits <- findOverlaps(dau_gr, dmr_gr)

dau_with_genes <- dau_df_out2[queryHits(hits), c("dau_id", "novelty_class")] |>
  bind_cols(
    dmr_df[subjectHits(hits), "overlapping.genes", drop = FALSE] |>
      rename(genes_from_dmr = overlapping.genes)
  ) |>
  separate_rows(genes_from_dmr, sep = ",\\s*") |>
  filter(genes_from_dmr != "", !is.na(genes_from_dmr)) |>
  rename(gene = genes_from_dmr) |>
  distinct()

# Novel genes per method
novel_dmr_genes <- dmr_genes_long |>
  filter(grepl("Novel", novelty_class)) |>
  pull(gene) |> unique() |> sort()

novel_dau_genes <- dau_with_genes |>
  filter(grepl("Novel", novelty_class)) |>
  pull(gene) |> unique() |> sort()

known_dmr_genes <- dmr_genes_long |>
  filter(grepl("Known", novelty_class)) |>
  pull(gene) |> unique() |> sort()

known_dau_genes <- dau_with_genes |>
  filter(grepl("Known", novelty_class)) |>
  pull(gene) |> unique() |> sort()
```


```{r}
# ── 4. Venn diagrams ───────────────────────────────────────────────────────────
p_venn_novel <- ggvenn(
  list(DMR = novel_dmr_genes, DAU = novel_dau_genes),
  fill_color = c("#E07A5F", "#81B29A"),
  stroke_size = 0.5, set_name_size = 4.5, text_size = 4
) + labs(title = "Novel genes")

p_venn_known <- ggvenn(
  list(DMR = known_dmr_genes, DAU = known_dau_genes),
  fill_color = c("#4E8098", "#F2CC8F"),
  stroke_size = 0.5, set_name_size = 4.5, text_size = 4
) + labs(title = "Known PPGL genes recovered")
p_venn_novel
p_venn_known
```


```{r}
# ── 5. Genomic overlap: novel DMR regions vs novel DAU windows ─────────────────
novel_dmr_gr <- dmr_gr[grepl("Novel", dmr_gr$novelty_class)]
novel_dau_gr <- dau_gr[grepl("Novel", dau_gr$novelty_class)]

novel_overlaps <- findOverlaps(novel_dmr_gr, novel_dau_gr)

cat(sprintf(
  "\nNovel DMR regions:     %d\nNovel DAU windows:     %d\nDMR regions with ≥1 novel DAU window: %d (%.1f%%)\n",
  length(novel_dmr_gr),
  length(novel_dau_gr),
  length(unique(queryHits(novel_overlaps))),
  100 * length(unique(queryHits(novel_overlaps))) / length(novel_dmr_gr)
))
```


```{r}
# ── 6. Print gene lists ────────────────────────────────────────────────────────
novel_both  <- intersect(novel_dmr_genes, novel_dau_genes)
novel_dmr_only <- setdiff(novel_dmr_genes, novel_dau_genes)
novel_dau_only <- setdiff(novel_dau_genes, novel_dmr_genes)

cat("\n── Novel in BOTH (high-confidence) n =", length(novel_both), "──\n")
print(novel_both)

cat("\n── Novel in DMR only n =", length(novel_dmr_only), "──\n")
print(novel_dmr_only)

cat("\n── Novel in DAU only n =", length(novel_dau_only), "──\n")
print(novel_dau_only)
```


```{r}
# ── 7. Assemble figure ─────────────────────────────────────────────────────────
p_final <- p_bar | (p_venn_novel / p_venn_known)
p_final + plot_annotation(
  title = "Known PPGL-Specific Loci: DMR vs DAU novelty comparison",
  theme = theme(plot.title = element_text(size = 14, face = "bold"))
)
```


# compare novel only (use filtered dau though)

```{r}
# ── Novel-only comparison: DMR vs DAU ─────────────────────────────────────────

# DMR: split overlapping.genes
dmr_genes_long <- dmr_df |>
  select(dmr_id, overlapping.genes, direction, meandiff) |>
  separate_rows(overlapping.genes, sep = ",\\s*") |>
  filter(overlapping.genes != "", !is.na(overlapping.genes)) |>
  rename(gene = overlapping.genes)

# DAU: inherit genes from overlapping DMR regions
dmr_gr <- makeGRangesFromDataFrame(dmr_df, keep.extra.columns = TRUE)
dau_gr <- makeGRangesFromDataFrame(dau_df_out2, keep.extra.columns = TRUE)
hits   <- findOverlaps(dau_gr, dmr_gr)

dau_with_genes <- dau_df_out2[queryHits(hits), c("dau_id", "FDR", "gene.F")] |>
  bind_cols(dmr_df[subjectHits(hits), "overlapping.genes", drop = FALSE]) |>
  separate_rows(overlapping.genes, sep = ",\\s*") |>
  filter(overlapping.genes != "", !is.na(overlapping.genes)) |>
  rename(gene = overlapping.genes) |>
  distinct()

# DAU windows with NO overlapping DMR region
dau_no_dmr <- dau_df_out2[-unique(queryHits(hits)), ] |>
  select(dau_id, FDR, gene.F)

# ── Gene sets ─────────────────────────────────────────────────────────────────
genes_dmr <- unique(dmr_genes_long$gene)
genes_dau <- unique(dau_with_genes$gene)

genes_both     <- intersect(genes_dmr, genes_dau)
genes_dmr_only <- setdiff(genes_dmr, genes_dau)
genes_dau_only <- setdiff(genes_dau, genes_dmr)

cat("Novel genes in DMR:           ", length(genes_dmr), "\n")
cat("Novel genes in DAU (via DMR): ", length(genes_dau), "\n")
cat("Shared:                       ", length(genes_both), "\n")
cat("DMR only:                     ", length(genes_dmr_only), "\n")
cat("DAU only (in a DMR region):   ", length(genes_dau_only), "\n")
cat("DAU windows outside any DMR:  ", nrow(dau_no_dmr), "\n\n")

cat("── Shared genes ──\n"); print(sort(genes_both))
cat("\n── DMR only ──\n");   print(sort(genes_dmr_only))
cat("\n── DAU only ──\n");   print(sort(genes_dau_only))

# ── Venn ──────────────────────────────────────────────────────────────────────
ggvenn(
  list(DMR = genes_dmr, DAU = genes_dau),
  fill_color = c("#E07A5F", "#81B29A"),
  stroke_size = 0.5, set_name_size = 5, text_size = 4.5
) +
  labs(title = "Novel genes: DMR vs DAU overlap",
       subtitle = sprintf("DMR=%d  DAU=%d  shared=%d",
                          length(genes_dmr), length(genes_dau), length(genes_both)))
```



```{r}
# ── Rank concordance for shared genes ─────────────────────────────────────────
# (do the genes found by both methods rank similarly in each?)
shared_ranks <- dmr_genes_long |>
  filter(gene %in% genes_both) |>
  left_join(dmr_df |> select(dmr_id, dmr_rank), by = "dmr_id") |>
  group_by(gene) |>
  slice_min(dmr_rank, n = 1) |>
  ungroup() |>
  select(gene, dmr_rank, direction, meandiff) |>
  left_join(
    dau_with_genes |>
      filter(gene %in% genes_both) |>
      left_join(dau_df_out |> select(dau_id, dtu_rank), by = "dau_id") |>
      group_by(gene) |>
      slice_min(dtu_rank, n = 1) |>
      ungroup() |>
      select(gene, dtu_rank, FDR),   # FDR already in dau_with_genes
    by = "gene"
  ) |>
  arrange(dmr_rank)

print(shared_ranks, n = Inf)

ggplot(shared_ranks, aes(x = dmr_rank, y = dtu_rank, label = gene)) +
  geom_point(aes(colour = direction), size = 3) +
  ggrepel::geom_text_repel(size = 3, max.overlaps = 20) +
  scale_colour_manual(values = c(hypo = "#4E8098", hyper = "#E07A5F")) +
  labs(title = "Rank concordance for shared novel genes",
       x = "DMR rank", y = "DAU rank") +
  theme_bw(base_size = 12)
```




```{r}
dau_no_dmr
```






```{r}
dau_optiles %>% filter(FDR < 0.05)
dau_df_out2
```

```{r}
dau_df_out %>%
  ggplot(aes(x = FDR)) +
  geom_histogram(bins = 75)
```

# gene ontology 


```{r}
# ── Gene Ontology enrichment ───────────────────────────────────────────────────

library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)

# ── Helper: symbol → Entrez, run GO, return result ────────────────────────────
run_go <- function(gene_symbols, label,
                   ont = "BP",        # BP / MF / CC or "ALL"
                   pval = 0.05,
                   universe_symbols = NULL) {

  ids <- bitr(gene_symbols,
              fromType = "SYMBOL", toType = "ENTREZID",
              OrgDb = org.Hs.eg.db)

  cat(sprintf("\n── %s: %d symbols → %d mapped ──\n",
              label, length(gene_symbols), nrow(ids)))

  uni <- if (!is.null(universe_symbols)) {
    bitr(universe_symbols,
         fromType = "SYMBOL", toType = "ENTREZID",
         OrgDb = org.Hs.eg.db)$ENTREZID
  } else NULL

  enrichGO(
    gene          = ids$ENTREZID,
    universe      = uni,
    OrgDb         = org.Hs.eg.db,
    ont           = ont,
    pAdjustMethod = "BH",
    pvalueCutoff  = pval,
    qvalueCutoff  = 0.2,
    readable      = TRUE   # converts Entrez back to symbols in output
  )
}

# Background = all genes tested by DMR (all overlapping.genes)
universe_genes <- dmr_genes_long$gene |> unique()

# ── Run GO ────────────────────────────────────────────────────────────────────
go_shared   <- run_go(genes_both,     "Shared (DMR ∩ DAU)", universe_symbols = universe_genes)
go_dmr_only <- run_go(genes_dmr_only, "DMR only",           universe_symbols = universe_genes)

# Quick look at top terms
as.data.frame(go_shared)   |> select(Description, GeneRatio, p.adjust, geneID) |> head(10)
as.data.frame(go_dmr_only) |> select(Description, GeneRatio, p.adjust, geneID) |> head(10)

# ── Dotplots ──────────────────────────────────────────────────────────────────
p_go_shared <- dotplot(go_shared, showCategory = 15, font.size = 10) +
  labs(title = "GO:BP — Shared genes (DMR ∩ DAU)")

p_go_dmr <- dotplot(go_dmr_only, showCategory = 15, font.size = 10) +
  labs(title = "GO:BP — DMR-only genes")

p_go_shared
p_go_dmr

# ── Side-by-side comparison if both have results ──────────────────────────────
if (nrow(as.data.frame(go_shared)) > 0 & nrow(as.data.frame(go_dmr_only)) > 0) {

  go_compare <- compareCluster(
    geneClusters = list(
      "DMR ∩ DAU" = bitr(genes_both,     "SYMBOL", "ENTREZID", org.Hs.eg.db)$ENTREZID,
      "DMR only"  = bitr(genes_dmr_only, "SYMBOL", "ENTREZID", org.Hs.eg.db)$ENTREZID
    ),
    fun           = "enrichGO",
    universe      = bitr(universe_genes, "SYMBOL", "ENTREZID", org.Hs.eg.db)$ENTREZID,
    OrgDb         = org.Hs.eg.db,
    ont           = "BP",
    pAdjustMethod = "BH",
    pvalueCutoff  = 0.05,
    readable      = TRUE
  )

  dotplot(go_compare, showCategory = 15, font.size = 10) +
    labs(title = "GO:BP comparison: shared vs DMR-only novel genes")
}
```


```{r}
dmr_df %>% dplyr::filter(seqnames == "chr1")
```


# look at indiv genes / regions etc

```{r}
shared_ranks
```

```{r}
gene_group_df
dau_with_genes
dmr_genes
genes_txdb
```
```{r}
basic_tiles_genes
```

```{r}
gene_ranges
```

```{r}
shared_ranks$gene
```

```{r}
ppgl_relevant_genes <- c(
  "PTPRN2", "NTRK1", "NR4A2", "PDE2A", "PDE2A-AS2",
  "PCDHGA1", "PCDHGA2", "PCDHGA3", "PCDHGA4", "PCDHGA5", "PCDHGA6",
  "PCDHGA7", "PCDHGA8", "PCDHGA9", "PCDHGA10", "PCDHGA11", "PCDHGA12",
  "PCDHGB1", "PCDHGB2", "PCDHGB3", "PCDHGB4", "PCDHGB5", "PCDHGB6", "PCDHGB7",
  "PCDHGC3", "PCDHGC4", "PCDHGC5",
  "NR2F1-AS1", "TBR1", "EBF3", "NKX6-2", "MIR9-3HG",
  "SLC32A1", "SLC12A5", "SLC4A10", "CACNA1H",
  "BCL2L11", "APC2", "RUNX1", "TBX2", "TBX2-AS1",
  "WNT6", "EYA1", "IRX4", "IRX5", "ISLR2",
  "SARM1", "PROX1", "TRPM5", "INSRR"
)

shared_ranks_f <- shared_ranks %>% filter(gene %in% ppgl_relevant_genes)
```

```{r}
shared_ranks %>% filter(gene %in% ppgl_relevant_genes) %>% distinct(dtu_rank)
shared_ranks %>% filter(gene %in% ppgl_relevant_genes) %>% distinct(dmr_rank)
```


```{r}
basic_tiles_genes %>% filter(dtu_rank %in% filter(shared_ranks, gene %in% ppgl_relevant_genes)$dtu_rank) 
```

```{r}
dmr_genes %>% filter(dmr_rank %in% filter(shared_ranks, gene %in% ppgl_relevant_genes)$dmr_rank) 
```

```{r}
entrez_ids <- mapIds(
  org.Hs.eg.db,
  keys = ppgl_relevant_genes,
  column = "ENTREZID",
  keytype = "SYMBOL",
  multiVals = "first"
)

# Remove NAs (genes not mapped, e.g. some lncRNAs)
entrez_ids <- entrez_ids[!is.na(entrez_ids)]

# As a plain vector
entrez_vector <- unname(entrez_ids)
```
```{r}
genes_txdb %>% data.frame() %>% filter(gene_id %in% entrez_ids) %>%
  ggplot(aes(x = seqnames, fill = seqnames)) +
  geom_bar()
```


```{r}
shared_ranks_f
```

```{r}
dmr_genes %>% filter(dmr_rank == 4)
```


```{r}
library(duckdb)
library(arrow)

# Convert RDS to parquet first
#df_rds <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_all_healthy.rds")
write_parquet(df_rds, "/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet")

con <- dbConnect(duckdb())

dbExecute(con, "
  COPY (
    SELECT *, 'cancer' AS type
    FROM read_parquet('/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet')
    UNION ALL
    SELECT *, 'healthy' AS type
    FROM read_parquet('/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet')
  ) TO '/researchers/caitlin.page/cf_nano/r_output/emseq_combined.parquet' (FORMAT PARQUET)
")

dbDisconnect(con)
```

```{r}
genes_txdb %>% data.frame() %>% filter(gene_id %in% entrez_ids)
```


```{r}
test_plot <- rrbs_plotdata("/researchers/caitlin.page/cf_nano/r_output/emseq_combined.parquet", "chr2", 111119378, 111168444)
distinct(test_plot) %>% distinct(sample, type)
```

```{r}
con <- dbConnect(duckdb())
on.exit(dbDisconnect(con))

s <- dbGetQuery(con, "SELECT DISTINCT sample FROM read_parquet('/researchers/caitlin.page/cf_nano/r_output/emseq_combined.parquet') ORDER BY sample")
s$sample
```

```{r}
con <- dbConnect(duckdb())
on.exit(dbDisconnect(con))

dbGetQuery(con, "SELECT DISTINCT sample FROM read_parquet('/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet') ORDER BY sample")
```

```{r}
con <- dbConnect(duckdb())
on.exit(dbDisconnect(con))

dbGetQuery(con, "SELECT DISTINCT sample FROM read_parquet('/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet') ORDER BY sample")
```

```{r}
con <- dbConnect(duckdb())
on.exit(dbDisconnect(con))
dbExecute(con, "
  COPY (
    SELECT *, 'cancer'  AS type FROM read_parquet('/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet')
    UNION ALL
    SELECT *, 'healthy' AS type FROM read_parquet('/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet')
  ) TO '/researchers/caitlin.page/cf_nano/r_output/emseq_combined_NEW.parquet' (FORMAT PARQUET)
")
```


```{r}
dbGetQuery(con, "SELECT DISTINCT sample, type FROM read_parquet('/researchers/caitlin.page/cf_nano/r_output/emseq_combined_NEW.parquet') ORDER BY type, sample")
```


```{r}
dbGetQuery(con, "DESCRIBE SELECT * FROM read_parquet('/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet') LIMIT 0")
dbGetQuery(con, "DESCRIBE SELECT * FROM read_parquet('/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet') LIMIT 0")
```


```{r}
dbGetQuery(con, "SELECT DISTINCT sample FROM read_parquet('/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet') ORDER BY sample")
```


```{r}
con <- dbConnect(duckdb())
on.exit(dbDisconnect(con))

dbGetQuery(con, "SELECT DISTINCT sample FROM read_parquet('/researchers/caitlin.page/cf_nano/r_output/emseq_combined_fixed.parquet') ORDER BY sample")
```
```{r}
head(df_rds)
```


```{r}
genes_interest <- genes_txdb %>% data.frame() %>% filter(gene_id %in% entrez_ids)
```

```{r}
rrbs_plot5("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", "chr16", 54930865, 54934485) +
  geom_dau(dau_basictiles) +
  geom_dmr(dmr_emseq)
```

* finally
```{r}
genes_interest
```




```{r}
p_vec <- c()
for(i in 1:50) {
  p <- rrbs_plot5("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", as.character(genes_interest[i,]$seqnames), genes_interest[i,]$start, genes_interest[i,]$end) +
  geom_dau(dau_basictiles) +
  geom_dmr(dmr_emseq)
  
  print(p)
}
```





* change in proportions delta

```{r}
dau_basictiles
```




"/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", 


```{r}
tile_prop <- function(read_data_path_a, read_data_path_b, chrom, region_start, region_end) {

  # ── 1. Read only the matching rows — never loads unmatched data into RAM ───
  con <- dbConnect(duckdb())
  on.exit(dbDisconnect(con))
  
  read_data_a <- dbGetQuery(con, glue::glue("
    SELECT *
    FROM read_parquet('{read_data_path_a}')
    WHERE seqnames = '{chrom}'
      AND \"start\" >= {region_start}
      AND \"end\"   <= {region_end}
  "))
  
  read_data_b <- dbGetQuery(con, glue::glue("
    SELECT *
    FROM read_parquet('{read_data_path_b}')
    WHERE seqnames = '{chrom}'
      AND \"start\" >= {region_start}
      AND \"end\"   <= {region_end}
  "))
  
  read_data_a$type <- "healthy"
  read_data_b$type <- "cancer"
  
  read_data <- rbind(read_data_a, read_data_b)
  read_data
}
test_tile <- tile_prop("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", 
          "chr1", 147079001, 147079500)
```

```{r}
test_tile %>% mutate(region_id = "example") %>%
  mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                     alpha > 0.8 ~ "a > 0.8",
                                     alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                     alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                     TRUE ~ "0.6 <= a <= 0.8")) %>%
  mutate(transcript = paste0(region_id, "_", alpha_group)) %>%
    group_by(transcript, type, region_id, alpha_group) %>%
    summarise(tile_band_count = n()) %>% ungroup() %>%
  group_by(region_id, type) %>% 
  mutate(tile_count = sum(tile_band_count), 
         prop_alpha_in_tile = tile_band_count/sum(tile_band_count))
```

```{r}
test_tile %>% mutate(region_id = "example") %>%
  mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                     alpha > 0.8 ~ "a > 0.8",
                                     alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                     alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                     TRUE ~ "0.6 <= a <= 0.8")) %>%
  mutate(transcript = paste0(region_id, "_", alpha_group)) %>%
    group_by(transcript, type, region_id, alpha_group) %>%
    summarise(tile_band_count = n()) %>% ungroup() %>%
  group_by(region_id, type) %>% 
  mutate(tile_count = sum(tile_band_count), 
         prop_alpha_in_tile = tile_band_count/sum(tile_band_count)) %>%
  select(alpha_group, type, prop_alpha_in_tile) %>%
  pivot_wider(names_from = type, values_from = prop_alpha_in_tile) %>%
  mutate(prop_change = cancer - healthy)
```

```{r}
test_tile %>%
  mutate(
    region_id = "example",
    alpha_group = case_when(
      alpha < 0.2  ~ "a < 0.2",
      alpha > 0.8  ~ "a > 0.8",
      alpha <= 0.4 ~ "0.2 <= a <= 0.4",
      alpha < 0.6  ~ "0.4 < a < 0.6",
      TRUE         ~ "0.6 <= a <= 0.8"
    )
  ) %>%
  count(type, region_id, alpha_group, name = "tile_band_count") %>%
  group_by(region_id, type) %>%
  mutate(prop_alpha_in_tile = tile_band_count / sum(tile_band_count)) %>%
  ungroup() %>%
  select(alpha_group, type, prop_alpha_in_tile) %>%
  pivot_wider(names_from = type, values_from = prop_alpha_in_tile) %>%
  mutate(prop_change = cancer - healthy)
```

```{r}
dau_basictiles |>
  group_by(seqnames) |>
  summarise(min_start = min(start), max_end = max(end)) %>% mutate(width = max_end - min_start)

dau_basictiles[1:10,] |>
  group_by(seqnames) |>
  summarise(min_start = min(start), max_end = max(end)) %>% mutate(width = max_end - min_start)
```



```{r}
tile_prop <- function(read_data_path_a, read_data_path_b, tiles, cols_from_tiles = NULL) {

  # ── 1. Read only the matching rows — never loads unmatched data into RAM ───
  con <- dbConnect(duckdb())
  duckdb::dbWriteTable(con, "tiles_temp", tiles, overwrite = TRUE)
  on.exit(duckdb::dbRemoveTable(con, "tiles_temp"), add = TRUE)
  
  
  # Build SELECT clause for tile columns
  if (is.null(cols_from_tiles)) {
    # Exclude coordinate cols to avoid clashes with parquet
    tile_cols <- setdiff(colnames(tiles), c("seqnames", "start", "end"))
  } else {
    tile_cols <- cols_from_tiles
  }
  
  tile_select <- if (length(tile_cols) > 0) {
    paste(glue::glue(', t."{tile_cols}"'), collapse = "")
  } else ""
  
  read_data_a <- dbGetQuery(con, glue::glue("
    SELECT b.*{tile_select}
    FROM read_parquet('{read_data_path_a}') b
    JOIN tiles_temp t
      ON b.seqnames = t.seqnames
     AND b.start <= t.end
     AND b.end   >= t.start
  "))
  
  read_data_b <- dbGetQuery(con, glue::glue("
    SELECT b.*{tile_select}
    FROM read_parquet('{read_data_path_b}') b
    JOIN tiles_temp t
      ON b.seqnames = t.seqnames
     AND b.start <= t.end
     AND b.end   >= t.start
  "))
  
  read_data_a$type <- "healthy"
  read_data_b$type <- "cancer"
    read_data <- rbind(read_data_a, read_data_b)
  read_data
}
test_multip_tile_more <- tile_prop("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", 
          dau_basictiles[1:20,])
```

```{r}
test_multip_tile %>%
  mutate(
    alpha_group = case_when(
      alpha < 0.2  ~ "a < 0.2",
      alpha > 0.8  ~ "a > 0.8",
      alpha <= 0.4 ~ "0.2 <= a <= 0.4",
      alpha < 0.6  ~ "0.4 < a < 0.6",
      TRUE         ~ "0.6 <= a <= 0.8"
    )
  ) %>%
  count(type, region_id, alpha_group, name = "tile_band_count") %>%
  group_by(region_id, type) %>%
  mutate(prop_alpha_in_tile = tile_band_count / sum(tile_band_count)) %>%
  ungroup() %>%
  select(region_id, alpha_group, type, prop_alpha_in_tile) %>%
  pivot_wider(names_from = type, values_from = prop_alpha_in_tile) %>%
  mutate(prop_change = cancer - healthy)
```

```{r}
props_from_test <- test_multip_tile_more %>%
  mutate(
    alpha_group = case_when(
      alpha < 0.2  ~ "a < 0.2",
      alpha > 0.8  ~ "a > 0.8",
      alpha <= 0.4 ~ "0.2 <= a <= 0.4",
      alpha < 0.6  ~ "0.4 < a < 0.6",
      TRUE         ~ "0.6 <= a <= 0.8"
    )
  ) %>%
  count(type, dtu_rank, region_id, alpha_group, name = "tile_band_count") %>%
  group_by(region_id, dtu_rank, type) %>%
  mutate(prop_alpha_in_tile = tile_band_count / sum(tile_band_count)) %>%
  ungroup() %>%
  select(region_id, dtu_rank, alpha_group, type, prop_alpha_in_tile) %>%
  pivot_wider(names_from = type, values_from = prop_alpha_in_tile) %>%
  mutate(prop_change = cancer - healthy)
```

```{r}
props_from_test %>% filter(abs(prop_change) >= 0.2)
```



```{r}
props_from_test %>%
  ggplot(aes(x = alpha_group, y = prop_change)) +
  geom_boxplot()
```

```{r}
props_from_test %>% filter(dtu_rank == 17)
```
* this region has a medium -> low swithc
* plot it

```{r}
rrbs_plot5("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", 
           "chr11", 126851501, 126852000) +
  geom_dau(dau_basictiles) +
  geom_dmr(dmr_emseq)
```
chr2:113387501-113388000
```{r}
props_from_test %>% filter(dtu_rank == 19)
rrbs_plot5("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", 
           "chr2", 113387501, 113388000) +
  geom_dau(dau_basictiles) +
  geom_dmr(dmr_emseq)
```

```{r}
test_multip_tile_more %>%
  mutate(
    alpha_group = case_when(
      alpha < 0.2  ~ "a < 0.2",
      alpha > 0.8  ~ "a > 0.8",
      alpha <= 0.4 ~ "0.2 <= a <= 0.4",
      alpha < 0.6  ~ "0.4 < a < 0.6",
      TRUE         ~ "0.6 <= a <= 0.8"
    )
  ) %>%
  count(type, dtu_rank, region_id, alpha_group, name = "tile_band_count") %>%
  group_by(region_id, dtu_rank, type) %>%
  mutate(prop_alpha_in_tile = tile_band_count / sum(tile_band_count), tile_count = sum(tile_band_count)) %>% ungroup() %>% distinct(dtu_rank, region_id, type, tile_count)
```


```{r}
dau_basictiles
```


dtu_optile <- find_overlaps(as_granges(all_samples), as_granges(optiles_all)) %>% data.frame() %>%
    distinct(sample, read_id, read_start, read_end, alpha, region_id, type) %>%
    mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                     alpha > 0.8 ~ "a > 0.8",
                                     alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                     alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                     TRUE ~ "0.6 <= a <= 0.8")) %>%
  mutate(transcript = paste0(region_id, "_", alpha_group)) %>%
    group_by(transcript, type, region_id, alpha_group) %>%
    summarise(tile_band_count = n()) %>% ungroup() %>%
  group_by(region_id, type) %>% 
  mutate(tile_count = sum(tile_band_count), 
         prop_alpha_in_tile = tile_band_count/sum(tile_band_count)) %>%
  m
  
