---
title: "compareTilingMethods"
author: "caitlinpage"
date: "2026-09-02"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r}
library(plyranges)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
```


## compare the tile positions
```{r}
dau_basictiles <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dtu_basictiles_emseq_cancer_healthy.rds")

dau_optiles <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_emseq_cancer_healthy.rds")
```

```{r}
dau_basictiles %>% filter(FDR < 0.05) %>% nrow()
dau_optiles %>% filter(FDR < 0.05) %>% nrow()
```



```{r}
tile_overlap <- find_overlaps(as_granges(mutate(dau_basictiles, tiling = "basic")), as_granges(mutate(dau_optiles, tiling = "optiles"))) %>% data.frame()
```

```{r}
bas_gr <- as_granges(mutate(dau_basictiles, tiling = "basic"))
op_gr <- as_granges(mutate(dau_optiles, tiling = "optiles"))
```

```{r}
hits <- findOverlaps(as_granges(mutate(dau_basictiles, tiling = "basic")), as_granges(mutate(dau_optiles, tiling = "optiles")))

cat("basic regions:", length(bas_gr), "\n")
cat("optiles regions:", length(op_gr), "\n")
cat("basic with ≥1 overlap:", length(unique(queryHits(hits))), "\n")
cat("optiles with ≥1 overlap:", length(unique(subjectHits(hits))), "\n")
```

```{r}
# Create a consensus locus set from the union of both
consensus <- reduce(c(bas_gr, op_gr))  # merges all overlapping regions into one

# For each consensus locus, find best FDR from method 1
hits1 <- findOverlaps(consensus, bas_gr)
best1 <- as.data.frame(hits1) %>%
  mutate(fdr = bas_gr$FDR[subjectHits]) %>%
  group_by(queryHits) %>%
  slice_min(fdr, n = 1) %>%           # best (lowest) FDR per locus
  rename(locus_idx = queryHits)

# Same for method 2
hits2 <- findOverlaps(consensus, op_gr)
best2 <- as.data.frame(hits2) %>%
  mutate(fdr = op_gr$FDR[subjectHits]) %>%
  group_by(queryHits) %>%
  slice_min(fdr, n = 1) %>%
  rename(locus_idx = queryHits)

# Join — NA where a locus has no coverage in one method
comparison <- full_join(
  ungroup(best1) %>% select(locus_idx, fdr_m1 = fdr),
  ungroup(best2) %>% select(locus_idx, fdr_m2 = fdr),
  by = "locus_idx"
)

# Loci covered by both methods only
both_covered <- comparison %>% filter(!is.na(fdr_m1) & !is.na(fdr_m2))
```

```{r}
# Rank correlation (robust to the heavy skew in FDR values)
cor(-log10(both_covered$fdr_m1), -log10(both_covered$fdr_m2), method = "spearman")

# Agreement on significant calls
table(
  sig_m1 = both_covered$fdr_m1 < 0.05,
  sig_m2 = both_covered$fdr_m2 < 0.05
)

# Scatter
ggplot(both_covered, aes(-log10(fdr_m1), -log10(fdr_m2))) +
  geom_point(alpha = 0.2, size = 0.8) +
  geom_abline(colour = "red", linetype = "dashed") +
  labs(x = "Method 1  –log10(FDR)", y = "Method 2  –log10(FDR)",
       title = sprintf("n = %d shared loci", nrow(both_covered)))
```

```{r}
ggplot(both_covered, aes(-log10(fdr_m1), -log10(fdr_m2))) +
  geom_point(alpha = 0.2, size = 0.8) +
  geom_abline(colour = "red", linetype = "dashed") +
  labs(x = "Basic  –log10(FDR)", y = "OPtiles  –log10(FDR)",
       title = sprintf("n = %d shared loci", nrow(both_covered)))
```


mmmmmmmmmmmmmmmmm
idk what to do

```{r}
dau_basictiles
dau_optiles
```


```{r}
genes_txdb <- genes(TxDb.Hsapiens.UCSC.hg38.knownGene)
```

```{r}
basic_tiles_genes <- find_overlaps(as_granges(dau_basictiles), genes_txdb) %>% data.frame()
op_tiles_genes <- find_overlaps(as_granges(dau_optiles), genes_txdb) %>% data.frame()
```


```{r}
ggVennDiagram::ggVennDiagram(list(basic = basic_tiles_genes$gene_id, optiles = op_tiles_genes$gene_id))
```
* so position wise there is a high overlap - because this while it filters a few out, is what was tested
```{r}
basic_tiles_genes %>% filter(!gene_id %in% op_tiles_genes$gene_id) %>% distinct(gene_id)
op_tiles_genes %>% filter(!gene_id %in% basic_tiles_genes$gene_id) %>% distinct(gene_id)
```



* just the significant tiles
```{r}
ggVennDiagram::ggVennDiagram(list(basic = filter(basic_tiles_genes, FDR < 0.05)$gene_id, optiles = filter(op_tiles_genes, FDR < 0.05)$gene_id))
```
* sense check
```{r}
ggVennDiagram::ggVennDiagram(list(basic_filt = filter(basic_tiles_genes, FDR < 0.05)$gene_id, basic = basic_tiles_genes$gene_id))
```

```{r libraries}
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(org.Hs.eg.db)
library(GenomicFeatures)
library(AnnotationDbi)
library(clusterProfiler)
library(enrichplot)
library(ggplot2)
library(dplyr)
library(tidyr)
library(stringr)
library(knitr)
library(DT)
 
txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene
```
 
---
 
# Gene list
 
Replace this vector with your own Entrez IDs.
 
```{r gene-list}
gene_ids <- basic_tiles_genes %>% filter(!gene_id %in% op_tiles_genes$gene_id) %>% distinct(gene_id) %>% .$gene_id
```
 
---
 
# 1  Chromosome locations
 
```{r genomic-locations}
# Pull gene-level ranges from TxDb
gene_ranges <- genes(txdb, filter = list(gene_id = gene_ids))
 
# Retrieve gene symbols and descriptions from org.Hs.eg.db
anno <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys    = gene_ids,
  columns = c("SYMBOL", "GENENAME", "ENTREZID"),
  keytype = "ENTREZID"
)
 
# Build a tidy location table
location_df <- as.data.frame(gene_ranges) %>%
  rename(entrezid = gene_id,
         chr      = seqnames,
         start_bp = start,
         end_bp   = end,
         width_bp = width) %>%
  left_join(anno, by = c("entrezid" = "ENTREZID")) %>%
  select(SYMBOL, entrezid, chr, start_bp, end_bp, width_bp, strand, GENENAME) %>%
  arrange(chr, start_bp)
 
datatable(
  location_df,
  caption   = "Genomic coordinates (hg38)",
  rownames  = FALSE,
  filter    = "top",
  options   = list(pageLength = 20, scrollX = TRUE)
)
```
 
## Chromosome distribution
 
```{r chr-plot, fig.height=4}
chr_order <- paste0("chr", c(1:22, "X", "Y", "M"))
 
location_df %>%
  mutate(chr = factor(chr, levels = chr_order)) %>%
  dplyr::count(chr) %>%
  ggplot(aes(x = chr, y = n, fill = chr)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = n), vjust = -0.3, size = 3) +
  scale_fill_viridis_d() +
  labs(title = "Number of genes per chromosome",
       x = "Chromosome", y = "Gene count") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
 
---
 
# 2  Gene functions
 
```{r go-anno}
go_anno <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys    = gene_ids,
  columns = c("SYMBOL", "GO", "ONTOLOGY", "GOALL"),
  keytype = "ENTREZID"
) %>%
  filter(!is.na(GO)) %>%
  rename(entrezid = ENTREZID)
 
# Summarise GO term counts per gene
go_counts <- go_anno %>%
  group_by(entrezid, SYMBOL, ONTOLOGY) %>%
  summarise(n_terms = n_distinct(GO), .groups = "drop") %>%
  pivot_wider(names_from = ONTOLOGY, values_from = n_terms,
              values_fill = 0, names_prefix = "GO_") %>%
  left_join(anno %>% select(ENTREZID, GENENAME),
            by = c("entrezid" = "ENTREZID"))
 
datatable(
  go_counts,
  caption  = "GO term counts per gene (BP = biological process, CC = cellular component, MF = molecular function)",
  rownames = FALSE,
  filter   = "top",
  options  = list(pageLength = 20, scrollX = TRUE)
)
```
 
## GO enrichment (over-representation analysis)
 
```{r go-ora}
ego_bp <- enrichGO(
  gene          = gene_ids,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENTREZID",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.20,
  readable      = TRUE
)
 
if (nrow(as.data.frame(ego_bp)) > 0) {
  dotplot(ego_bp, showCategory = 20, title = "GO Biological Process enrichment") +
    theme(axis.text.y = element_text(size = 9))
} else {
  message("No significant GO BP terms found — consider relaxing thresholds or expanding the gene list.")
}
```
 
```{r go-table}
if (nrow(as.data.frame(ego_bp)) > 0) {
  as.data.frame(ego_bp) %>%
    select(Description, GeneRatio, BgRatio, pvalue, p.adjust, geneID) %>%
    datatable(
      caption  = "Enriched GO Biological Process terms",
      rownames = FALSE,
      filter   = "top",
      options  = list(pageLength = 15, scrollX = TRUE)
    )
}
```
 
## KEGG pathway enrichment
 
```{r kegg-ora}
ekegg <- enrichKEGG(
  gene          = gene_ids,
  organism      = "hsa",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.20
)
 
# Translate Entrez IDs to symbols for readability
if (!is.null(ekegg) && nrow(as.data.frame(ekegg)) > 0) {
  ekegg <- setReadable(ekegg, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")
  dotplot(ekegg, showCategory = 20, title = "KEGG pathway enrichment") +
    theme(axis.text.y = element_text(size = 9))
} else {
  message("No significant KEGG pathways found — consider relaxing thresholds or expanding the gene list.")
}
```
 
```{r kegg-table}
if (!is.null(ekegg) && nrow(as.data.frame(ekegg)) > 0) {
  as.data.frame(ekegg) %>%
    select(Description, GeneRatio, BgRatio, pvalue, p.adjust, geneID) %>%
    datatable(
      caption  = "Enriched KEGG pathways",
      rownames = FALSE,
      filter   = "top",
      options  = list(pageLength = 15, scrollX = TRUE)
    )
}
```
 
---
 
# 3  Relationships between genes
 
## 3a  Chromosomal proximity
 
Genes on the same chromosome sorted by position, with distance (bp) to the nearest neighbour in the list.
 
```{r proximity}
prox_df <- location_df %>%
  arrange(chr, start_bp) %>%
  group_by(chr) %>%
  mutate(
    next_gene     = lead(SYMBOL),
    dist_to_next  = lead(start_bp) - end_bp
  ) %>%
  ungroup() %>%
  filter(!is.na(next_gene)) %>%
  select(chr, SYMBOL, start_bp, end_bp, next_gene, dist_to_next) %>%
  arrange(dist_to_next)
 
if (nrow(prox_df) > 0) {
  datatable(
    prox_df,
    caption  = "Adjacent gene pairs on the same chromosome (sorted by gap size)",
    rownames = FALSE,
    filter   = "top",
    options  = list(pageLength = 20, scrollX = TRUE)
  )
} else {
  message("No gene pairs share a chromosome in this list.")
}
```
 
```{r proximity-plot, fig.height=5}
# Ideogram-style strip: one row per chromosome containing ≥1 gene
if (nrow(location_df) > 0) {
  chr_order <- paste0("chr", c(1:22, "X", "Y"))
  plot_df <- location_df %>%
    filter(chr %in% chr_order) %>%
    mutate(chr = factor(chr, levels = rev(chr_order)))
 
  ggplot(plot_df, aes(x = start_bp / 1e6, y = chr, colour = chr, label = SYMBOL)) +
    geom_point(size = 3, show.legend = FALSE) +
    ggrepel::geom_text_repel(size = 3, show.legend = FALSE,
                             max.overlaps = 20) +
    labs(title = "Gene positions along chromosomes (hg38)",
         x = "Position (Mb)", y = NULL) +
    theme_minimal(base_size = 11)
}
```
 
> **Note:** `ggrepel` is used for label placement. Install with `install.packages("ggrepel")` if needed.
 
## 3b  Shared GO terms (gene–gene overlap)
 
Pairwise count of how many GO Biological Process terms any two genes share.
 
```{r shared-go}
# Per-gene BP term sets
bp_sets <- go_anno %>%
  filter(ONTOLOGY == "BP") %>%
  group_by(SYMBOL) %>%
  summarise(terms = list(unique(GO)), .groups = "drop")
 
# All pairs
symbols <- bp_sets$SYMBOL
pairs <- combn(symbols, 2, simplify = FALSE)
 
shared_go <- lapply(pairs, function(pair) {
  t1 <- bp_sets$terms[bp_sets$SYMBOL == pair[1]][[1]]
  t2 <- bp_sets$terms[bp_sets$SYMBOL == pair[2]][[1]]
  tibble(
    gene_A       = pair[1],
    gene_B       = pair[2],
    shared_terms = length(intersect(t1, t2)),
    jaccard      = length(intersect(t1, t2)) / length(union(t1, t2))
  )
}) %>%
  bind_rows() %>%
  arrange(desc(shared_terms))
 
datatable(
  shared_go,
  caption  = "Pairwise GO BP term overlap between genes",
  rownames = FALSE,
  filter   = "top",
  options  = list(pageLength = 20, scrollX = TRUE)
)
```
 
```{r shared-go-heatmap, fig.height=6}
# Symmetric matrix for heatmap
mat_df <- bind_rows(
  shared_go %>% rename(from = gene_A, to = gene_B),
  shared_go %>% rename(from = gene_B, to = gene_A),
  tibble(from = symbols, to = symbols,
         shared_terms = NA_integer_, jaccard = NA_real_)
)
 
ggplot(mat_df, aes(x = to, y = from, fill = shared_terms)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  scale_fill_gradient(low = "#f0f0f0", high = "#2166ac",
                      na.value = "grey80", name = "Shared\nGO BP terms") +
  labs(title = "Pairwise shared GO Biological Process terms",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
 
## 3c  Shared KEGG pathways
 
```{r shared-kegg}
if (!is.null(ekegg) && nrow(as.data.frame(ekegg)) > 0) {
  kegg_df <- as.data.frame(ekegg) %>%
    select(Description, geneID) %>%
    mutate(geneID = str_split(geneID, "/")) %>%
    unnest(geneID) %>%
    rename(SYMBOL = geneID, pathway = Description)
 
  # Pairs sharing a pathway
  kegg_pairs <- kegg_df %>%
    inner_join(kegg_df, by = "pathway", relationship = "many-to-many") %>%
    filter(SYMBOL.x < SYMBOL.y) %>%
    group_by(gene_A = SYMBOL.x, gene_B = SYMBOL.y) %>%
    summarise(shared_pathways = n_distinct(pathway),
              pathways        = paste(unique(pathway), collapse = "; "),
              .groups = "drop") %>%
    arrange(desc(shared_pathways))
 
  datatable(
    kegg_pairs,
    caption  = "Gene pairs sharing enriched KEGG pathways",
    rownames = FALSE,
    filter   = "top",
    options  = list(pageLength = 20, scrollX = TRUE)
  )
} else {
  message("No enriched KEGG pathways — shared-pathway table skipped.")
}
```
 
---
 
# 4  Full annotation table
 
A single table joining all information above for easy export.
 
```{r full-table}
full_anno <- location_df %>%
  left_join(go_counts %>% select(-GENENAME), by = c("SYMBOL", "entrezid")) %>%
  select(SYMBOL, entrezid, chr, start_bp, end_bp, strand,
         GO_BP, GO_CC, GO_MF, GENENAME) %>%
  arrange(chr, start_bp)
 
datatable(
  full_anno,
  caption  = "Full annotation summary",
  rownames = FALSE,
  filter   = "top",
  options  = list(pageLength = 25, scrollX = TRUE)
)
```

## with the ones only found by optiles
```{r gene-list}
gene_ids <- op_tiles_genes %>% filter(!gene_id %in% basic_tiles_genes$gene_id) %>% distinct(gene_id) %>% .$gene_id
```
 
---
 
# 1  Chromosome locations
 
```{r genomic-locations}
# Pull gene-level ranges from TxDb
gene_ranges <- genes(txdb, filter = list(gene_id = gene_ids))
 
# Retrieve gene symbols and descriptions from org.Hs.eg.db
anno <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys    = gene_ids,
  columns = c("SYMBOL", "GENENAME", "ENTREZID"),
  keytype = "ENTREZID"
)
 
# Build a tidy location table
location_df <- as.data.frame(gene_ranges) %>%
  rename(entrezid = gene_id,
         chr      = seqnames,
         start_bp = start,
         end_bp   = end,
         width_bp = width) %>%
  left_join(anno, by = c("entrezid" = "ENTREZID")) %>%
  select(SYMBOL, entrezid, chr, start_bp, end_bp, width_bp, strand, GENENAME) %>%
  arrange(chr, start_bp)
 
datatable(
  location_df,
  caption   = "Genomic coordinates (hg38)",
  rownames  = FALSE,
  filter    = "top",
  options   = list(pageLength = 20, scrollX = TRUE)
)
```
 
## Chromosome distribution
 
```{r chr-plot, fig.height=4}
chr_order <- paste0("chr", c(1:22, "X", "Y", "M"))
 
location_df %>%
  mutate(chr = factor(chr, levels = chr_order)) %>%
  dplyr::count(chr) %>%
  ggplot(aes(x = chr, y = n, fill = chr)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = n), vjust = -0.3, size = 3) +
  scale_fill_viridis_d() +
  labs(title = "Number of genes per chromosome",
       x = "Chromosome", y = "Gene count") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
 
---
 
# 2  Gene functions
 
```{r go-anno}
go_anno <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys    = gene_ids,
  columns = c("SYMBOL", "GO", "ONTOLOGY", "GOALL"),
  keytype = "ENTREZID"
) %>%
  filter(!is.na(GO)) %>%
  rename(entrezid = ENTREZID)
 
# Summarise GO term counts per gene
go_counts <- go_anno %>%
  group_by(entrezid, SYMBOL, ONTOLOGY) %>%
  summarise(n_terms = n_distinct(GO), .groups = "drop") %>%
  pivot_wider(names_from = ONTOLOGY, values_from = n_terms,
              values_fill = 0, names_prefix = "GO_") %>%
  left_join(anno %>% select(ENTREZID, GENENAME),
            by = c("entrezid" = "ENTREZID"))
 
datatable(
  go_counts,
  caption  = "GO term counts per gene (BP = biological process, CC = cellular component, MF = molecular function)",
  rownames = FALSE,
  filter   = "top",
  options  = list(pageLength = 20, scrollX = TRUE)
)
```
 
## GO enrichment (over-representation analysis)
 
```{r go-ora}
ego_bp <- enrichGO(
  gene          = gene_ids,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENTREZID",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.20,
  readable      = TRUE
)
 
if (nrow(as.data.frame(ego_bp)) > 0) {
  dotplot(ego_bp, showCategory = 20, title = "GO Biological Process enrichment") +
    theme(axis.text.y = element_text(size = 9))
} else {
  message("No significant GO BP terms found — consider relaxing thresholds or expanding the gene list.")
}
```
 
```{r go-table}
if (nrow(as.data.frame(ego_bp)) > 0) {
  as.data.frame(ego_bp) %>%
    select(Description, GeneRatio, BgRatio, pvalue, p.adjust, geneID) %>%
    datatable(
      caption  = "Enriched GO Biological Process terms",
      rownames = FALSE,
      filter   = "top",
      options  = list(pageLength = 15, scrollX = TRUE)
    )
}
```
 
## KEGG pathway enrichment
 
```{r kegg-ora}
ekegg <- enrichKEGG(
  gene          = gene_ids,
  organism      = "hsa",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.20
)
 
# Translate Entrez IDs to symbols for readability
if (!is.null(ekegg) && nrow(as.data.frame(ekegg)) > 0) {
  ekegg <- setReadable(ekegg, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")
  dotplot(ekegg, showCategory = 20, title = "KEGG pathway enrichment") +
    theme(axis.text.y = element_text(size = 9))
} else {
  message("No significant KEGG pathways found — consider relaxing thresholds or expanding the gene list.")
}
```
 
```{r kegg-table}
if (!is.null(ekegg) && nrow(as.data.frame(ekegg)) > 0) {
  as.data.frame(ekegg) %>%
    select(Description, GeneRatio, BgRatio, pvalue, p.adjust, geneID) %>%
    datatable(
      caption  = "Enriched KEGG pathways",
      rownames = FALSE,
      filter   = "top",
      options  = list(pageLength = 15, scrollX = TRUE)
    )
}
```
 
---
 
# 3  Relationships between genes
 
## 3a  Chromosomal proximity
 
Genes on the same chromosome sorted by position, with distance (bp) to the nearest neighbour in the list.
 
```{r proximity}
prox_df <- location_df %>%
  arrange(chr, start_bp) %>%
  group_by(chr) %>%
  mutate(
    next_gene     = lead(SYMBOL),
    dist_to_next  = lead(start_bp) - end_bp
  ) %>%
  ungroup() %>%
  filter(!is.na(next_gene)) %>%
  select(chr, SYMBOL, start_bp, end_bp, next_gene, dist_to_next) %>%
  arrange(dist_to_next)
 
if (nrow(prox_df) > 0) {
  datatable(
    prox_df,
    caption  = "Adjacent gene pairs on the same chromosome (sorted by gap size)",
    rownames = FALSE,
    filter   = "top",
    options  = list(pageLength = 20, scrollX = TRUE)
  )
} else {
  message("No gene pairs share a chromosome in this list.")
}
```
 
```{r proximity-plot, fig.height=5}
# Ideogram-style strip: one row per chromosome containing ≥1 gene
if (nrow(location_df) > 0) {
  chr_order <- paste0("chr", c(1:22, "X", "Y"))
  plot_df <- location_df %>%
    filter(chr %in% chr_order) %>%
    mutate(chr = factor(chr, levels = rev(chr_order)))
 
  ggplot(plot_df, aes(x = start_bp / 1e6, y = chr, colour = chr, label = SYMBOL)) +
    geom_point(size = 3, show.legend = FALSE) +
    ggrepel::geom_text_repel(size = 3, show.legend = FALSE,
                             max.overlaps = 20) +
    labs(title = "Gene positions along chromosomes (hg38)",
         x = "Position (Mb)", y = NULL) +
    theme_minimal(base_size = 11)
}
```
 
> **Note:** `ggrepel` is used for label placement. Install with `install.packages("ggrepel")` if needed.
 
## 3b  Shared GO terms (gene–gene overlap)
 
Pairwise count of how many GO Biological Process terms any two genes share.
 
```{r shared-go}
# Per-gene BP term sets
bp_sets <- go_anno %>%
  filter(ONTOLOGY == "BP") %>%
  group_by(SYMBOL) %>%
  summarise(terms = list(unique(GO)), .groups = "drop")
 
# All pairs
symbols <- bp_sets$SYMBOL
pairs <- combn(symbols, 2, simplify = FALSE)
 
shared_go <- lapply(pairs, function(pair) {
  t1 <- bp_sets$terms[bp_sets$SYMBOL == pair[1]][[1]]
  t2 <- bp_sets$terms[bp_sets$SYMBOL == pair[2]][[1]]
  tibble(
    gene_A       = pair[1],
    gene_B       = pair[2],
    shared_terms = length(intersect(t1, t2)),
    jaccard      = length(intersect(t1, t2)) / length(union(t1, t2))
  )
}) %>%
  bind_rows() %>%
  arrange(desc(shared_terms))
 
datatable(
  shared_go,
  caption  = "Pairwise GO BP term overlap between genes",
  rownames = FALSE,
  filter   = "top",
  options  = list(pageLength = 20, scrollX = TRUE)
)
```
 
```{r shared-go-heatmap, fig.height=6}
# Symmetric matrix for heatmap
mat_df <- bind_rows(
  shared_go %>% rename(from = gene_A, to = gene_B),
  shared_go %>% rename(from = gene_B, to = gene_A),
  tibble(from = symbols, to = symbols,
         shared_terms = NA_integer_, jaccard = NA_real_)
)
 
ggplot(mat_df, aes(x = to, y = from, fill = shared_terms)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  scale_fill_gradient(low = "#f0f0f0", high = "#2166ac",
                      na.value = "grey80", name = "Shared\nGO BP terms") +
  labs(title = "Pairwise shared GO Biological Process terms",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
 
## 3c  Shared KEGG pathways
 
```{r shared-kegg}
if (!is.null(ekegg) && nrow(as.data.frame(ekegg)) > 0) {
  kegg_df <- as.data.frame(ekegg) %>%
    select(Description, geneID) %>%
    mutate(geneID = str_split(geneID, "/")) %>%
    unnest(geneID) %>%
    rename(SYMBOL = geneID, pathway = Description)
 
  # Pairs sharing a pathway
  kegg_pairs <- kegg_df %>%
    inner_join(kegg_df, by = "pathway", relationship = "many-to-many") %>%
    filter(SYMBOL.x < SYMBOL.y) %>%
    group_by(gene_A = SYMBOL.x, gene_B = SYMBOL.y) %>%
    summarise(shared_pathways = n_distinct(pathway),
              pathways        = paste(unique(pathway), collapse = "; "),
              .groups = "drop") %>%
    arrange(desc(shared_pathways))
 
  datatable(
    kegg_pairs,
    caption  = "Gene pairs sharing enriched KEGG pathways",
    rownames = FALSE,
    filter   = "top",
    options  = list(pageLength = 20, scrollX = TRUE)
  )
} else {
  message("No enriched KEGG pathways — shared-pathway table skipped.")
}
```
 
---
 
# 4  Full annotation table
 
A single table joining all information above for easy export.
 
```{r full-table}
full_anno <- location_df %>%
  left_join(go_counts %>% select(-GENENAME), by = c("SYMBOL", "entrezid")) %>%
  select(SYMBOL, entrezid, chr, start_bp, end_bp, strand,
         GO_BP, GO_CC, GO_MF, GENENAME) %>%
  arrange(chr, start_bp)
 
datatable(
  full_anno,
  caption  = "Full annotation summary",
  rownames = FALSE,
  filter   = "top",
  options  = list(pageLength = 25, scrollX = TRUE)
)
```





## compare lists

```{r libraries}
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(org.Hs.eg.db)
library(GenomicFeatures)
library(AnnotationDbi)
library(clusterProfiler)
library(enrichplot)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(DT)
library(pheatmap)
library(ggrepel)
library(ComplexUpset)
library(RColorBrewer)
 
txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene
```
 
---
 
# Gene lists  ← edit here
 
Add or remove lists freely. Names become labels throughout the report.
All IDs are **Entrez** (character strings).
 
```{r gene-lists}
gene_lists <- list(
  unique_op = op_tiles_genes %>% filter(!gene_id %in% basic_tiles_genes$gene_id) %>% distinct(gene_id) %>% .$gene_id,  
  unique_basic = basic_tiles_genes %>% filter(!gene_id %in% op_tiles_genes$gene_id) %>% distinct(gene_id) %>% .$gene_id,
  unique_op_filt = filter(op_tiles_genes, FDR < 0.05) %>% filter(!gene_id %in% filter(basic_tiles_genes, FDR < 0.05)$gene_id) %>% distinct(gene_id) %>% .$gene_id,
  unique_basic_filt = filter(basic_tiles_genes, FDR < 0.05) %>% filter(!gene_id %in% filter(op_tiles_genes, FDR < 0.05)$gene_id) %>% distinct(gene_id) %>% .$gene_id)
 
# Palette — one colour per list (extend if you have more than 8 lists)
list_colours <- setNames(
  brewer.pal(max(4, length(gene_lists)), "Set2")[seq_along(gene_lists)],
  names(gene_lists)
)
gene_lists
```
 
```{r}
ggVennDiagram::ggVennDiagram(gene_lists)
```
 
 
```{r helpers, include=FALSE}
# Retrieve gene symbols for a vector of Entrez IDs
entrez_to_symbol <- function(ids) {
  AnnotationDbi::select(
    org.Hs.eg.db,
    keys    = ids,
    columns = c("SYMBOL", "GENENAME"),
    keytype = "ENTREZID"
  )
}
 
# All unique IDs across lists
all_ids <- unique(unlist(gene_lists))
all_anno <- entrez_to_symbol(all_ids)
```
 
---
 
# 1  List overview
 
```{r list-overview}
overview <- purrr::imap_dfr(gene_lists, function(ids, name) {
  tibble(
    List        = name,
    n_genes     = length(ids),
    genes       = paste(
      all_anno$SYMBOL[match(ids, all_anno$ENTREZID)],
      collapse = ", "
    )
  )
})
 
knitr::kable(overview, caption = "Gene list contents")
```
 
---
 
# 2  Pairwise similarity
 
Two complementary metrics are shown:
 
- **Jaccard index** — intersection / union (penalises size differences)
- **Overlap coefficient** — intersection / min(|A|, |B|) (asks "is one list a subset of the other?")
 
```{r similarity}
list_names <- names(gene_lists)
n <- length(list_names)
 
jaccard_mat  <- matrix(NA_real_, n, n, dimnames = list(list_names, list_names))
overlap_mat  <- matrix(NA_real_, n, n, dimnames = list(list_names, list_names))
 
for (i in seq_len(n)) {
  for (j in seq_len(n)) {
    a <- gene_lists[[i]]
    b <- gene_lists[[j]]
    isect <- length(intersect(a, b))
    jaccard_mat[i, j] <- isect / length(union(a, b))
    overlap_mat[i, j] <- isect / min(length(a), length(b))
  }
}
 
# Long form for ggplot
to_long <- function(mat, metric) {
  as.data.frame(mat) %>%
    tibble::rownames_to_column("List_A") %>%
    pivot_longer(-List_A, names_to = "List_B", values_to = metric)
}
 
sim_long <- left_join(
  to_long(jaccard_mat,  "Jaccard"),
  to_long(overlap_mat,  "Overlap"),
  by = c("List_A", "List_B")
)
 
datatable(
  sim_long %>% mutate(across(c(Jaccard, Overlap), \(x) round(x, 3))),
  caption  = "Pairwise similarity metrics",
  rownames = FALSE,
  filter   = "top",
  options  = list(pageLength = 20)
)
```
 
## Jaccard heatmap
 
```{r jaccard-heatmap, fig.height=4}
pheatmap(
  jaccard_mat,
  display_numbers = TRUE,
  number_format   = "%.2f",
  color           = colorRampPalette(c("white", "#2166ac"))(50),
  main            = "Jaccard similarity",
  fontsize         = 12,
  angle_col        = 45
)
```
 
## Overlap coefficient heatmap
 
```{r overlap-heatmap, fig.height=4}
pheatmap(
  overlap_mat,
  display_numbers = TRUE,
  number_format   = "%.2f",
  color           = colorRampPalette(c("white", "#b2182b"))(50),
  main            = "Overlap coefficient (intersection / min set size)",
  fontsize         = 12,
  angle_col        = 45
)
```
 
---
 
# 3  Gene membership (UpSet plot)
 
UpSet plots scale well beyond what Venn diagrams can show cleanly.
Each bar represents a specific combination of lists.
 
```{r upset, fig.height=5}
# Binary membership matrix
membership <- purrr::map_dfc(gene_lists, function(ids) {
  as.integer(all_ids %in% ids)
}) %>%
  mutate(gene = all_ids) %>%
  left_join(all_anno %>% select(ENTREZID, SYMBOL),
            by = c("gene" = "ENTREZID")) %>%
  select(SYMBOL, everything(), -gene)
 
upset_mat <- membership %>%
  select(all_of(list_names)) %>%
  as.data.frame()
 
ComplexUpset::upset(
  upset_mat,
  intersect        = list_names,
  name             = "Gene list combination",
  min_size         = 1,
  width_ratio      = 0.2,
  base_annotations = list(
    "Intersection size" = intersection_size(text = list(size = 3.5))
  ),
  themes = upset_modify_themes(list("default" = theme_minimal(base_size = 11)))
)
```
 
## Genes per intersection
 
```{r intersection-table}
membership %>%
  mutate(
    combination = apply(select(., all_of(list_names)), 1,
                        function(r) paste(list_names[as.logical(r)], collapse = " ∩ "))
  ) %>%
  group_by(combination) %>%
  summarise(
    n_genes = dplyr::n(),
    genes   = paste(SYMBOL, collapse = ", "),
    .groups = "drop"
  ) %>%
  arrange(desc(n_genes)) %>%
  datatable(
    caption  = "Genes in each list combination",
    rownames = FALSE,
    filter   = "top",
    options  = list(pageLength = 20, scrollX = TRUE)
  )
```
 
---
 
# 4  Comparative enrichment
 
`clusterProfiler::compareCluster()` runs ORA on each list simultaneously,
enabling direct visual comparison.
 
## GO Biological Process
 
```{r compare-go-bp, fig.height=8}
# Build a data frame of gene + group for compareCluster
gene_group_df <- purrr::imap_dfr(gene_lists, function(ids, name) {
  tibble(ENTREZID = ids, group = name)
})
 
cmp_go <- compareCluster(
  ENTREZID ~ group,
  data          = gene_group_df,
  fun           = "enrichGO",
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENTREZID",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.20,
  readable      = TRUE
)
 
if (!is.null(cmp_go) && nrow(as.data.frame(cmp_go)) > 0) {
  dotplot(cmp_go, showCategory = 15, title = "GO BP — comparison across lists") +
    theme(axis.text.y = element_text(size = 9),
          legend.position = "right")
} else {
  message("No significant GO BP terms found across any list.")
}
```
 
```{r compare-go-table}
if (!is.null(cmp_go) && nrow(as.data.frame(cmp_go)) > 0) {
  as.data.frame(cmp_go) %>%
    select(Cluster, Description, GeneRatio, BgRatio, p.adjust, geneID) %>%
    datatable(
      caption  = "GO BP enrichment per list",
      rownames = FALSE,
      filter   = "top",
      options  = list(pageLength = 20, scrollX = TRUE)
    )
}
```
 
## KEGG pathways
 
```{r compare-kegg, fig.height=7}
cmp_kegg <- compareCluster(
  ENTREZID ~ group,
  data          = gene_group_df,
  fun           = "enrichKEGG",
  organism      = "hsa",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.20
)
 
if (!is.null(cmp_kegg) && nrow(as.data.frame(cmp_kegg)) > 0) {
  cmp_kegg <- setReadable(cmp_kegg, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")
  dotplot(cmp_kegg, showCategory = 15, title = "KEGG — comparison across lists") +
    theme(axis.text.y = element_text(size = 9),
          legend.position = "right")
} else {
  message("No significant KEGG pathways found across any list.")
}
```
 
## Enrichment heatmap (shared terms)
 
Shows which GO BP terms are significant in which lists.
 
```{r enrich-heatmap, fig.height=7}
if (!is.null(cmp_go) && nrow(as.data.frame(cmp_go)) > 0) {
  go_wide <- as.data.frame(cmp_go) %>%
    select(Cluster, Description, p.adjust) %>%
    mutate(neg_log10_padj = -log10(p.adjust)) %>%
    pivot_wider(
      id_cols     = Description,
      names_from  = Cluster,
      values_from = neg_log10_padj,
      values_fill = 0
    ) %>%
    tibble::column_to_rownames("Description")
 
  # Keep top 30 terms by max score across lists
  top_terms <- names(sort(apply(go_wide, 1, max), decreasing = TRUE))[1:min(30, nrow(go_wide))]
  go_wide   <- go_wide[top_terms, , drop = FALSE]
 
  pheatmap(
    go_wide,
    color        = colorRampPalette(c("white", "#4dac26", "#b8e186"))(50) |>
                   rev() |>
                   colorRampPalette(c("white", "#d01c8b"))(50) |>
                   c(colorRampPalette(c("white", "#4dac26"))(50)),
    # simpler: single gradient
    color        = colorRampPalette(c("#f7f7f7", "#4393c3", "#053061"))(50),
    main         = "GO BP enrichment across lists\n(-log10 adjusted p-value)",
    angle_col    = 45,
    fontsize_row = 8,
    fontsize_col = 11,
    border_color = NA,
    na_col       = "grey95"
  )
}
```
 
---
 
# 5  Chromosomal distribution
 
```{r genomic-locs}
# Pull locations for all genes
all_ranges <- genes(txdb, filter = list(gene_id = all_ids))
 
loc_all <- as.data.frame(all_ranges) %>%
  rename(entrezid = gene_id, chr = seqnames,
         start_bp = start, end_bp = end) %>%
  left_join(all_anno, by = c("entrezid" = "ENTREZID")) %>%
  left_join(
    purrr::imap_dfr(gene_lists, function(ids, nm) tibble(entrezid = ids, list = nm)),
    by = "entrezid"
  ) %>%
  filter(!is.na(list))
 
chr_order <- paste0("chr", c(1:22, "X", "Y"))
loc_all   <- loc_all %>%
  filter(chr %in% chr_order) %>%
  mutate(chr = factor(chr, levels = rev(chr_order)),
         list = factor(list, levels = names(gene_lists)))
```
 
## Strip plot — all lists together
 
```{r strip-plot, fig.height=7}
ggplot(loc_all, aes(x = start_bp / 1e6, y = chr,
                    colour = list, shape = list)) +
  geom_point(size = 3, alpha = 0.8,
             position = position_jitter(height = 0.15, seed = 42)) +
  geom_text_repel(aes(label = SYMBOL), size = 2.5, max.overlaps = 15,
                  show.legend = FALSE) +
  scale_colour_manual(values = list_colours) +
  scale_shape_manual(values = c(16, 17, 15, 18, 8, 9)[seq_along(gene_lists)]) +
  labs(title    = "Gene positions across chromosomes (hg38)",
       x        = "Position (Mb)",
       y        = NULL,
       colour   = "List",
       shape    = "List") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```
 
## Per-list faceted view
 
```{r facet-strip, fig.height=7}
ggplot(loc_all, aes(x = start_bp / 1e6, y = chr, colour = list)) +
  geom_point(size = 2.5, alpha = 0.9) +
  geom_text_repel(aes(label = SYMBOL), size = 2.2, max.overlaps = 10,
                  show.legend = FALSE) +
  scale_colour_manual(values = list_colours) +
  facet_wrap(~list, ncol = 1) +
  labs(title  = "Gene positions — faceted by list",
       x      = "Position (Mb)",
       y      = NULL,
       colour = "List") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none",
        strip.text      = element_text(face = "bold", size = 11))
```
 
## Chromosome enrichment — which chromosomes are over-represented?
 
```{r chr-bar, fig.height=5}
loc_all %>%
  dplyr::count(list, chr) %>%
  ggplot(aes(x = chr, y = n, fill = list)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = list_colours) +
  labs(title = "Gene count per chromosome by list",
       x = NULL, y = "Gene count", fill = "List") +
  coord_flip() +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```
 
---
 
# 6  Full membership table
 
Every gene, which lists it belongs to, its genomic location and description.
 
```{r full-table}
full_df <- membership %>%
  left_join(
    loc_all %>%
      select(entrezid, chr, start_bp, end_bp, strand) %>%
      distinct(),
    by = c("gene" = "entrezid")
  ) %>%
  rename(entrezid = gene) %>%
  left_join(all_anno, by = c("entrezid" = "ENTREZID")) %>%
  select(SYMBOL, entrezid, all_of(list_names),
         chr, start_bp, end_bp, strand, GENENAME) %>%
  arrange(chr, start_bp)
 
datatable(
  full_df,
  caption  = "All genes with list membership and genomic location",
  rownames = FALSE,
  filter   = "top",
  options  = list(pageLength = 30, scrollX = TRUE)
)
```
 



```{r}
basic_tiles_genes
op_tiles_genes
```

```{r}
b <- basic_tiles_genes %>% group_by(gene_id) %>% summarise(basic_tiles = n()) %>% data.frame()
o <- op_tiles_genes %>% group_by(gene_id) %>% summarise(op_tiles = n()) %>% data.frame()
```

```{r}
data.frame(gene_id = unique(c(basic_tiles_genes$gene_id, op_tiles_genes$gene_id))) %>%
  mutate(basic_tiles = b[match(.$gene_id, b$gene_id), "basic_tiles"],
         op_tiles = o[match(.$gene_id, o$gene_id), "op_tiles"]) %>%
  ggplot(aes(x = op_tiles, y = basic_tiles)) +
  geom_point(alpha = 0.3)
```

```{r}
dau_basictiles
dau_optiles
```

```{r}
dau_basictiles %>% filter(FDR < 0.05)
```

#################################

# basic comparisons between tiling approaches

* how many cpgs
* how many left out 
etc from there

```{r}
nrow(optiles_all)
```
```{r}
tiles_500bp_cg
```


```{r}
optiles <- find_overlaps(as_granges(optiles_all), as_granges(cg_sites)) %>% data.frame() %>% group_by(seqnames, start, end, region_id) %>% summarise(n_cg = n()) %>% ungroup()
```

```{r}
optiles %>%
  ggplot(aes(x = n_cg)) +
  geom_histogram(bins = 50)

tiles_500bp_cg %>%
  ggplot(aes(x = n_cg)) +
  geom_histogram(bins = 50)
```
* cpgs not in tiles
```{r}
as_granges(cg_sites) %>% plyranges::filter_by_non_overlaps(as_granges(optiles))
```
* how close/far?
```{r}
cgs_missed <- as_granges(cg_sites) %>% plyranges::filter_by_non_overlaps(as_granges(optiles))
```

```{r}
nn_dist <- distanceToNearest(cgs_missed)

# Summarise
summary(mcols(nn_dist)$distance)
hist(mcols(nn_dist)$distance, breaks = 50,
     main = "Nearest-neighbour distances", xlab = "Distance (bp)")
```
* appear to be clustered
```{r}
nn_dist %>% data.frame() %>% 
  ggplot(aes(x = distance)) +
  geom_histogram(bins = 50) +
  scale_x_continuous(limits = c(0,100))
```
*it's missing a lot

```{r}
#library(regioneR)

mean_nn_dist <- function(A, ...) {
  if (length(A) < 2) return(NA)
  mean(mcols(distanceToNearest(A))$distance)
}

pt <- permTest(
  A = cgs_missed,
  ntimes = 1000,
  randomize.function = resampleRegions,
  universe = as_granges(cg_sites),
  evaluate.function = mean_nn_dist,
  mc.cores = 1
)

summary(pt)
plot(pt)
```
```{r}
system.time({
  A_rand <- resampleRegions(A = cgs_missed, universe = as_granges(cg_sites))
  mean_nn_dist(A_rand)
})
```


```{r}
as_granges(cg_sites) %>% plyranges::filter_by_non_overlaps(as_granges(tiles_500bp_cg))
```


```{r}
as_granges(cg_sites) %>% plyranges::filter_by_non_overlaps(as_granges(dau_optiles))
```

```{r}
as_granges(cg_sites) %>% plyranges::filter_by_non_overlaps(as_granges(dau_basictiles))
```


* so optiles is missing a lot of cpgs
* but are these cpgs interesting?

```{r}
optiles %>% mutate(tile_signif = ifelse(region_id %in% filter(dau_optiles, FDR < 0.05)$region_id, TRUE, FALSE), nrow_tile = 1:nrow(.))
```

```{r}
optiles %>% mutate(tile_signif = ifelse(region_id %in% filter(dau_optiles, FDR < 0.05)$region_id, TRUE, FALSE), nrow_tile = 1:nrow(.)) %>% 
  filter(tile_signif == TRUE)
```

* this is 3 tiles in a row that are signif
* so now look for cpgs in between
```{r}
optiles[611:613,]
```

```{r}
gaps(as_granges(optiles[611:613,]))[2:3]
as_granges(cg_sites) %>% plyranges::filter_by_overlaps(gaps(as_granges(optiles[611:613,]))[2])
as_granges(cg_sites) %>% plyranges::filter_by_overlaps(gaps(as_granges(optiles[611:613,]))[3])
```
* so gap between 611 and 612 has 1 cpg
* and gap between 612 and 613 has 27 cpgs

* but this is based on the reference - my data may not have any coverage there - in which case it's ok to filter them out
* need to do both

```{r}
signif_runs <- optiles %>% mutate(tile_signif = ifelse(region_id %in% filter(dau_optiles, FDR < 0.05)$region_id, TRUE, FALSE), nrow_tile = 1:nrow(.)) %>%
  group_by(seqnames) %>%
  arrange(nrow_tile) %>%
  filter(
    tile_signif,
    # check previous tile is both adjacent and significant
    (lag(tile_signif, default = FALSE) & lag(nrow_tile, default = 0) == nrow_tile - 1) |
    # OR next tile is both adjacent and significant  
    (lead(tile_signif, default = FALSE) & lead(nrow_tile, default = 0) == nrow_tile + 1)
  ) %>%
  ungroup() %>%
  group_by(seqnames) %>%
  mutate(
    run_id = cumsum(nrow_tile - lag(nrow_tile, default = first(nrow_tile) - 1) != 1)
  ) %>%
  ungroup()
```
```{r}
signif_runs %>% filter(seqnames == "chr5")
```

```{r}
# Start from tile-level data with run_id assigned
tiles_with_runs <- signif_runs %>%
  group_by(seqnames) %>%
  arrange(nrow_tile) %>%
  mutate(run_id = cumsum(nrow_tile - lag(nrow_tile, default = first(nrow_tile) - 1) != 1)) %>%
  ungroup()

# Find genomic gaps between tiles within each run
intra_run_gaps <- tiles_with_runs %>%
  group_by(seqnames, run_id) %>%
  arrange(start) %>%
  mutate(
    gap_start = lag(end) + 1,
    gap_end   = start - 1
  ) %>%
  filter(!is.na(gap_start), gap_end >= gap_start) %>%
  ungroup()

# Count CpGs in each gap
intra_run_gaps_gr <- intra_run_gaps %>%
  select(seqnames, start = gap_start, end = gap_end, run_id, nrow_tile) %>%
  as_granges() %>%
  mutate(n_cgs = count_overlaps(., as_granges(cg_sites)))

as_tibble(intra_run_gaps_gr)
```


```{r}
intra_run_gaps_gr %>% data.frame() %>%
  ggplot(aes(x = n_cgs)) +
  geom_histogram(bins = 50)
```

```{r}
intra_run_gaps_gr %>% data.frame() %>%
  ggplot(aes(x = width)) +
  geom_histogram(bins = 50)
```

```{r}
intra_run_gaps_gr %>% data.frame() %>% mutate(cg_density = n_cgs/width) %>%
  ggplot(aes(x = cg_density)) +
  geom_histogram(bins = 50) +
  scale_y_continuous(limits = c(0,100))

optiles %>%
  ggplot(aes(x = n_cg/500)) +
  geom_histogram(bins = 50) +
  scale_y_continuous(limits = c(0,100))

tiles_500bp_cg %>%
  ggplot(aes(x = n_cg/500)) +
  geom_histogram(bins = 50) +
  scale_y_continuous(limits = c(0,100))
```

```{r}
optiles %>%
  ggplot(aes(x = n_cg/500)) +
  geom_histogram(bins = 50)

tiles_500bp_cg %>%
  ggplot(aes(x = n_cg/500)) +
  geom_histogram(bins = 50)
```

* ok so established that in between the significant optiles - there are variable amounts of cpgs being missed

* important next question - are there gaps between significant tiles in the basic tiles?
** we know that as the tiles themselves - there are no cpgs missing
** but - if like a tile gets skipped over significance - how do we feel about that?

```{r}
run_signif_check <- tiles_500bp_cg %>% mutate(tile_signif = ifelse(tile_id %in% filter(dau_basictiles, FDR < 0.05)$region_id, TRUE, FALSE), nrow_tile = 1:nrow(.)) %>%
  group_by(seqnames) %>%
  arrange(nrow_tile) %>%
  filter(
    tile_signif,
    # check previous tile is both adjacent and significant
    (lag(tile_signif, default = FALSE) & lag(nrow_tile, default = 0) == nrow_tile - 1) |
    # OR next tile is both adjacent and significant  
    (lead(tile_signif, default = FALSE) & lead(nrow_tile, default = 0) == nrow_tile + 1)
  ) %>%
  ungroup() %>%
  group_by(seqnames) %>%
  mutate(
    run_id = cumsum(nrow_tile - lag(nrow_tile, default = first(nrow_tile) - 1) != 1)
  ) %>%
  ungroup()

# Start from tile-level data with run_id assigned
tiles_with_runs_check <- run_signif_check %>%
  group_by(seqnames) %>%
  arrange(nrow_tile) %>%
  mutate(run_id = cumsum(nrow_tile - lag(nrow_tile, default = first(nrow_tile) - 1) != 1)) %>%
  ungroup()

# Find genomic gaps between tiles within each run
intra_run_gaps_check <- tiles_with_runs_check %>%
  group_by(seqnames, run_id) %>%
  arrange(start) %>%
  mutate(
    gap_start = lag(end) + 1,
    gap_end   = start - 1
  ) %>%
  filter(!is.na(gap_start), gap_end >= gap_start) %>%
  ungroup()

# Count CpGs in each gap
intra_run_gaps_gr_check <- intra_run_gaps_check %>%
  select(seqnames, start = gap_start, end = gap_end, run_id, nrow_tile) %>%
  as_granges() %>%
  mutate(n_cgs = count_overlaps(., as_granges(cg_sites)))

as_tibble(intra_run_gaps_gr_check)
```

* nothing - that's good

# now on the actual cpgs from the data

* because different to reference
* next version is doing it with the cpgs I have data for
* and seeing if the numbers of cpgs change - because more important if things got missed that actually have coverage

```{r}
betas_h <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_betas_healthy.rds")
betas_c <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_betas_cancer.rds")
```

```{r}
betas_h %>% filter(meth_count_pos == 0, unmeth_count_pos == 0)
```

```{r}
cg_data <- rbind(betas_c[,2:3], betas_h[,2:3]) %>% distinct()
cg_data$end <- cg_data$start + 1
```

```{r}
optiles <- find_overlaps(as_granges(optiles), as_granges(cg_data)) %>% data.frame() %>% group_by(seqnames, start, end, region_id) %>% summarise(n_cg_data = n()) %>% ungroup()
```

```{r}
as_granges(cg_data) %>% plyranges::filter_by_non_overlaps(as_granges(optiles))
```

```{r}
as_granges(cg_data) %>% plyranges::filter_by_overlaps(as_granges(optiles))
```
```{r}
as_granges(cg_data) %>% plyranges::filter_by_non_overlaps(as_granges(tiles_500bp_cg))
as_granges(cg_sites) %>% plyranges::filter_by_non_overlaps(as_granges(tiles_500bp_cg))
```
* well oops 
* there should be nothing there
* could be more though

*check cg sites vs cg data
```{r}
as_granges(cg_data) %>% plyranges::filter_by_non_overlaps(as_granges(cg_sites))
as_granges(cg_sites) %>% plyranges::filter_by_non_overlaps(as_granges(cg_data))
```


```{r}
run_signif_check <- optiles %>% mutate(tile_signif = ifelse(region_id %in% filter(dau_optiles, FDR < 0.05)$region_id, TRUE, FALSE), nrow_tile = 1:nrow(.)) %>%
  group_by(seqnames) %>%
  arrange(nrow_tile) %>%
  filter(
    tile_signif,
    # check previous tile is both adjacent and significant
    (lag(tile_signif, default = FALSE) & lag(nrow_tile, default = 0) == nrow_tile - 1) |
    # OR next tile is both adjacent and significant  
    (lead(tile_signif, default = FALSE) & lead(nrow_tile, default = 0) == nrow_tile + 1)
  ) %>%
  ungroup() %>%
  group_by(seqnames) %>%
  mutate(
    run_id = cumsum(nrow_tile - lag(nrow_tile, default = first(nrow_tile) - 1) != 1)
  ) %>%
  ungroup()

# Start from tile-level data with run_id assigned
tiles_with_runs_check <- run_signif_check %>%
  group_by(seqnames) %>%
  arrange(nrow_tile) %>%
  mutate(run_id = cumsum(nrow_tile - lag(nrow_tile, default = first(nrow_tile) - 1) != 1)) %>%
  ungroup()

# Find genomic gaps between tiles within each run
intra_run_gaps_check <- tiles_with_runs_check %>%
  group_by(seqnames, run_id) %>%
  arrange(start) %>%
  mutate(
    gap_start = lag(end) + 1,
    gap_end   = start - 1
  ) %>%
  filter(!is.na(gap_start), gap_end >= gap_start) %>%
  ungroup()

# Count CpGs in each gap
intra_run_gaps_gr_check <- intra_run_gaps_check %>%
  select(seqnames, start = gap_start, end = gap_end, run_id, nrow_tile) %>%
  as_granges() %>%
  mutate(n_cgs = count_overlaps(., as_granges(cg_data)))

as_tibble(intra_run_gaps_gr_check)
```

```{r}
intra_run_gaps_gr %>% data.frame() %>%
  ggplot(aes(x = n_cgs)) +
  geom_histogram(bins = 50)

intra_run_gaps_gr %>% data.frame() %>%
  ggplot(aes(x = width)) +
  geom_histogram(bins = 50)
```
* plot still looks similar which is good, given that positions have done weird things
```{r}
cg_sites
```


* also sense check
```{r}
tiles_500bp_cg %>% mutate(tile_signif = ifelse(tile_id %in% filter(dau_basictiles, FDR < 0.05)$region_id, TRUE, FALSE)) %>%
  ggplot(aes(x = n_cg, colour = tile_signif)) +
  geom_density()
```
```{r}
optiles %>% mutate(tile_signif = ifelse(region_id %in% filter(dau_optiles, FDR < 0.05)$region_id, TRUE, FALSE), nrow_tile = 1:nrow(.)) %>%
  ggplot(aes(x = n_cg, colour = tile_signif)) +
  geom_density()
```
* yeah - more cpgs, more likely significant

```{r}
cg_sites 
```

```{r}
optiles
```





* tested tiles - dudde just filter it 

```{r}
dau_basictiles <- find_overlaps(as_granges(dau_basictiles), as_granges(cg_sites)) %>% data.frame() 
dau_optiles <- find_overlaps(as_granges(dau_optiles), as_granges(cg_sites)) %>% data.frame() 
```

```{r}
dau_basictiles %>% group_by(region_id) %>% count()
```

