---
title: "coverageInResults"
author: "caitlinpage"
date: "2026-06-25"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

## Introduction
```{r}
library(annotatr)
library(data.table)
library(plyranges)
library(tidyr)
library(stringr)
library(dplyr)
library(ggplot2)
library(edgeR)
library(OpTiles)
library(ggVennDiagram)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
library(missMethyl)
```

```{r}
seqs <- seqnames(BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38)[1:22]

cg_sites <- lapply(seqs, function(x) {
        cbind(data.frame(Biostrings::matchPattern("CG", BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38[[x]])),
            seqnames = x
        )}) %>% bind_rows()
```


```{r}
all_samples <- readRDS("/researchers/caitlin.page/cf_nano/r_output/all_rrbs_samples.rds")
optiles_all <- readRDS("/researchers/caitlin.page/cf_nano/r_output/optiles_all.rds")
```

```{r}
dtu.optiles <- readRDS("/researchers/caitlin.page/cf_nano/r_output/ts.gene_optiles.rds")
dtu_tiles_5_exon <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_5_exon.rds")
dmrcate_seq_dmr <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dmr_encode_gm12878_k562.rds")
```

```{r}
all_samples_simp <- all_samples %>% distinct(seqnames, read_start, read_end, strand, sample, read_id, alpha, meth_pattern, num_meth, total)
colnames(all_samples_simp)[2:3] <- c("start", "end")
```


```{r}
dtu_function <- function(all_samples, optiles_all, num_groups = 5) {
  dtu_optile <- find_overlaps(as_granges(all_samples), as_granges(optiles_all)) %>% data.frame() %>%
  distinct(sample, read_id, read_start, read_end, alpha, region_id) 
  
  if(num_groups == 5) {
  dtu_optile <- dtu_optile %>% 
    mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8"))
  } else if(num_groups == 3) {
  dtu_optile <- dtu_optile %>% 
    mutate(alpha_group = case_when(alpha < 0.34 ~ "a < 0.34",
                                alpha > 0.65 ~ "a > 0.65",
                                TRUE ~ "0.34 <= a <= 0.65"))
  } else if(num_groups == 4) {
  dtu_optile <- dtu_optile %>% 
    mutate(alpha_group = case_when(alpha <= 0.25 ~ "a <= 0.25",
                                alpha >= 0.8 ~ "a >= 0.75",
                                alpha > 0.25 & alpha <= 0.5 ~ "0.25 <= a <= 0.5",
                                TRUE ~ "0.5 < a < 0.75"))
}

dtu_optile <- dtu_optile %>% 
  mutate(transcript = paste0(region_id, "_", alpha_group)) %>%
  group_by(sample, transcript) %>%
 summarise(transcript_count = n()) %>% ungroup() %>% data.frame()

# build matrix of "transcript" counts - transcript rows, sample as columns
dtu_optile <- dtu_optile %>% pivot_wider(names_from = sample, values_from = transcript_count)
dtu_optile[is.na(dtu_optile)] <- 0
dtu_optile$gene_id <- sub("_.*", "", dtu_optile$transcript)

# DTU analysis with edgeR
## Alicia prefers gene method over Simes method (documentation slightly confusing about difference)
y <- DGEList(as.matrix(dtu_optile[,2:5]), group = c(1,1,2,2), genes = dtu_optile[,c("gene_id", "transcript")])
design <- model.matrix(~ c(1,1,2,2), data = y$samples)
fit <- glmQLFit(y, design, robust=TRUE)
ds <- diffSpliceDGE(fit, coef = 2, geneid = "gene_id", exonid = "transcript")
ts.gene_optiles <- topSpliceDGE(ds, test = "gene", number = Inf)

# add info to results: get proportion counts
gene_cg <- find_overlaps(as_granges(optiles_all), as_granges(cg_sites)) %>% data.frame() %>%
  group_by(region_id) %>%
  summarise(n_cg = n()) %>%
  ungroup() %>% data.frame()
fit$genes <- fit$genes %>% mutate(n_cg = gene_cg[match(.$gene_id, gene_cg$region_id), "n_cg"])
rpk <- y$counts/rep(fit$genes$n_cg/1000, times = ncol(y$counts))
tpm_mat <- sweep(rpk, 2, colSums(rpk)/1e6, "/")
tpm_mat <- tpm_mat %>% data.frame() %>% mutate(id = dtu_optile$transcript, gene_id = dtu_optile$gene_id)

tpm_mat <- tpm_mat %>%
  group_by(gene_id) %>%
  mutate(prop_S1_gm = b1/sum(b1),
         prop_S2_gm = b2/sum(b2),
         prop_S3_k = cml1/sum(cml1),
         prop_S4_k = cml2/sum(cml2)) %>%
  ungroup() %>% data.frame()


# add locations back in so compatible with plyranges stuff
## this bit only applicable to OPTILES because of the way it structures the ID
ts.gene_optiles <- ts.gene_optiles %>%
  mutate(region_id = gene_id) %>%
  separate_wider_delim(gene_id, delim = ":", names = c("seqnames", "gene_id")) %>%
  mutate(seqnames = gsub("chrchr", "chr", seqnames)) %>%
  separate_wider_delim(gene_id, delim = "-", names = c("start", "end"))
ts.gene_optiles$start <- as.double(ts.gene_optiles$start)
ts.gene_optiles$end <- as.double(ts.gene_optiles$end)
ts.gene_optiles$dtu_rank <- 1:nrow(ts.gene_optiles)
ts.gene_optiles
}

```

```{r}
dtu_tiles_3 <- dtu_function(all_samples, optiles_all, num_groups = 3)
```


* coverage in tiles

```{r}
optiles_all <- optiles_all %>% group_by(seqnames) %>% mutate(num = 1:n()) %>% ungroup() %>% mutate(t_num = 1:n())
```

```{r}
all_samples_simp[1:5,]
```


```{r}
tiles_reads <- find_overlaps(as_granges(optiles_all), as_granges(all_samples_simp)) %>% data.frame()
```


```{r}
tiles_reads %>% filter(t_num < 6671) %>% group_by(region_id, t_num, sample) %>% summarise(coverage = n()) %>% ungroup() %>% 
  ggplot(aes(x = t_num, y = coverage)) +
  geom_line()
```
* tile coverage in chrom1

```{r}
tiles_reads %>% filter(t_num < 6671) %>% group_by(region_id, t_num, sample) %>% summarise(coverage = n()) %>% ungroup() %>% 
  ggplot(aes(x = t_num, y = coverage)) +
  geom_line() +
  facet_wrap(~ sample)
```


```{r}
tiles_reads_summary <- tiles_reads %>% group_by(region_id, t_num, sample) %>% summarise(coverage = n()) %>% ungroup()
tiles_reads_summary <- data.frame(tiles_reads_summary)
```

```{r}
dtu.optiles %>% filter(FDR < 0.05) %>% nrow()
```

```{r}
tiles_reads_summary %>% mutate(sig_fdr = ifelse(region_id %in% filter(dtu.optiles, FDR < 0.05)$region_id, TRUE, FALSE)) %>%
  group_by(region_id, t_num, sig_fdr) %>% summarise(cov = sum(coverage)) %>%
  ggplot(aes(x = log2(cov), colour = sig_fdr)) +
  geom_density()
```

```{r}
tiles_reads_summary %>% mutate(sig_fdr = ifelse(region_id %in% filter(dtu.optiles, FDR < 0.05)$region_id, TRUE, FALSE)) %>%
  group_by(region_id, t_num, sig_fdr) %>% summarise(cov = sum(coverage)) %>%
  ggplot(aes(x = log2(cov), colour = sig_fdr)) +
  geom_histogram(bins = 50) +
  facet_wrap(~sig_fdr)
```
* so significant tiles have more coverage
* makes sense

* but is it higher coverage or lower coverage comparative to dmr??
```{r}
dmrcate_seq_dmr$rank <- 1:nrow(dmrcate_seq_dmr)
```


```{r}
tiles_dmr <- find_overlaps(as_granges(optiles_all), as_granges(dmrcate_seq_dmr)) %>% data.frame()
```

```{r}
tiles_reads_summary %>% mutate(sig_fdr = ifelse(region_id %in% filter(dtu.optiles, FDR < 0.05)$region_id, TRUE, FALSE),
                               sig_dmr = ifelse(region_id %in% tiles_dmr$region_id, TRUE, FALSE)) %>%
  group_by(region_id, t_num, sig_fdr, sig_dmr) %>% summarise(cov = sum(coverage)) %>%
  ggplot(aes(x = sig_fdr, y = log2(cov), fill = sig_fdr)) +
  geom_boxplot()
```

```{r}
tiles_reads_summary %>% mutate(sig_fdr = ifelse(region_id %in% filter(dtu.optiles, FDR < 0.05)$region_id, TRUE, FALSE),
                               sig_dmr = ifelse(region_id %in% tiles_dmr$region_id, TRUE, FALSE)) %>%
  group_by(region_id, t_num, sig_fdr, sig_dmr) %>% summarise(cov = sum(coverage)) %>%
  ggplot(aes(x = sig_dmr, y = log2(cov), fill = sig_dmr)) +
  geom_boxplot()
```
* damn this does not look promising

```{r}
tiles_reads_summary %>% mutate(sig_fdr = ifelse(region_id %in% filter(dtu.optiles, FDR < 0.05)$region_id, TRUE, FALSE),
                               sig_dmr = ifelse(region_id %in% tiles_dmr$region_id, TRUE, FALSE)) %>%
  group_by(region_id, t_num, sig_fdr, sig_dmr) %>% summarise(cov = sum(coverage)) %>%
  pivot_longer(cols = c(sig_fdr, sig_dmr), names_to = "method", values_to = "signif") %>%
  ggplot(aes(x = method, y = log2(cov), fill = signif)) +
  geom_boxplot()
```

* a lot smaller of a difference than expected
* yes mean is slightly smaller - but tail is bigger in dmr
* maybe would look better combining filtering with the delta???


* try with exon as well first though

```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% distinct(region_id) %>% nrow()
```

```{r}
tiles_reads_summary %>% mutate(sig_fdr = ifelse(region_id %in% filter(dtu.optiles, FDR < 0.05)$region_id, TRUE, FALSE),
                               sig_fdr_exon = ifelse(region_id %in% distinct(filter(dtu_tiles_5_exon, FDR < 0.05), region_id)$region_id, TRUE, FALSE),
                               sig_dmr = ifelse(region_id %in% tiles_dmr$region_id, TRUE, FALSE)) %>%
  group_by(region_id, t_num, sig_fdr, sig_fdr_exon, sig_dmr) %>% summarise(cov = sum(coverage)) %>%
  pivot_longer(cols = c(sig_fdr, sig_fdr_exon, sig_dmr), names_to = "method", values_to = "signif") %>%
  ggplot(aes(x = method, y = log2(cov), fill = signif)) +
  geom_boxplot()
```

```{r}
bin_delta <- find_overlaps(as_granges(all_samples_simp), as_granges(optiles_all)) %>% data.frame() %>%
  distinct(sample, read_id, start, end, alpha, region_id) %>% 
    mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8")) %>%
  mutate(transcript = paste0(region_id, "_", alpha_group)) %>% mutate(sample_type = ifelse(sample %in% c("b1", "b2"), "B", "CML")) %>%
  group_by(region_id, sample_type) %>% mutate(mean_alpha = mean(alpha)) %>% ungroup() %>% distinct(region_id, sample_type, mean_alpha) %>% .[order(.$region_id),]

bin_delta <- bin_delta %>% pivot_wider(names_from = sample_type, values_from = mean_alpha) %>% mutate(delta_b_minus_cml = B - CML)
bin_delta <- data.frame(bin_delta)
```

```{r}
dtu.optiles <- dtu.optiles %>% mutate(delta_b_minus_cml = bin_delta[match(.$region_id, bin_delta$region_id), "delta_b_minus_cml"])
dtu.optiles %>%
  ggplot(aes(x = delta_b_minus_cml, y = -log10(FDR))) +
  geom_point(alpha = 0.2)
```
* let's say abs(delta) > 0.2

```{r}
dtu.optiles %>% filter(FDR < 0.05) %>%
  ggplot(aes(x = abs(delta_b_minus_cml))) +
  geom_density()

dtu.optiles %>% filter(FDR < 0.05) %>%
  ggplot(aes(x = abs(delta_b_minus_cml))) +
  geom_histogram(bins = 50) +
  geom_vline(xintercept = 0.2, colour = "red")
```
* as a prop
```{r}
dtu.optiles %>% mutate(fdr_sig = ifelse(FDR < 0.05, TRUE, FALSE), delta_gr_0.2 = ifelse(abs(delta_b_minus_cml) > 0.2, TRUE, FALSE)) %>%
  ggplot(aes(x = fdr_sig, fill = delta_gr_0.2)) +
  geom_bar()

dtu.optiles %>% mutate(fdr_sig = ifelse(FDR < 0.05, TRUE, FALSE), delta_gr_0.2 = ifelse(abs(delta_b_minus_cml) > 0.2, TRUE, FALSE)) %>% filter(fdr_sig == TRUE) %>%
  ggplot(aes(x = fdr_sig, fill = delta_gr_0.2)) +
  geom_bar(position = "fill")
```


```{r}
dtu.optiles %>% filter(FDR < 0.05, abs(delta_b_minus_cml) > 0.2) %>% nrow()
```


```{r}
tiles_reads_summary %>% mutate(delta = bin_delta[match(.$region_id, bin_delta$region_id), "delta_b_minus_cml"]) %>% 
  mutate(sig_fdr = ifelse(region_id %in% filter(dtu.optiles, FDR < 0.05)$region_id, TRUE, FALSE),
        sig_fdr_exon = ifelse(region_id %in% distinct(filter(dtu_tiles_5_exon, FDR < 0.05), region_id)$region_id, TRUE, FALSE),
        sig_fdr_delta = ifelse(region_id %in% filter(dtu.optiles, FDR < 0.05, abs(delta_b_minus_cml) > 0.2)$region_id, TRUE, FALSE),
        sig_dmr = ifelse(region_id %in% tiles_dmr$region_id, TRUE, FALSE)) %>%
  group_by(region_id, t_num, sig_fdr, sig_fdr_delta, sig_fdr_exon, sig_dmr) %>% summarise(cov = sum(coverage)) %>%
  pivot_longer(cols = c(sig_fdr, sig_fdr_delta, sig_fdr_exon, sig_dmr), names_to = "method", values_to = "signif") %>%
  ggplot(aes(x = method, y = log2(cov), fill = signif)) +
  geom_boxplot()
```
```{r}
tiles_reads_summary %>% mutate(delta = bin_delta[match(.$region_id, bin_delta$region_id), "delta_b_minus_cml"]) %>% 
  mutate(sig_fdr = ifelse(region_id %in% filter(dtu.optiles, FDR < 0.05)$region_id, TRUE, FALSE),
        sig_fdr_exon = ifelse(region_id %in% distinct(filter(dtu_tiles_5_exon, FDR < 0.05), region_id)$region_id, TRUE, FALSE),
        sig_fdr_delta = ifelse(region_id %in% filter(dtu.optiles, FDR < 0.05, abs(delta_b_minus_cml) > 0.2)$region_id, TRUE, FALSE),
        sig_dmr = ifelse(region_id %in% tiles_dmr$region_id, TRUE, FALSE)) %>%
  group_by(region_id, t_num, sig_fdr, sig_fdr_delta, sig_fdr_exon, sig_dmr) %>% summarise(cov = sum(coverage)) %>%
  pivot_longer(cols = c(sig_fdr, sig_fdr_delta, sig_fdr_exon, sig_dmr), names_to = "method", values_to = "signif") %>%
  ggplot(aes(x = method, y = log2(cov), fill = signif)) +
  geom_violin()
```

* but yeah - damn

* ok let's look at the tail dmrs
* cov < 32
```{r}
log2(32)
2^5
```



```{r}
tiles_reads_summary <- tiles_reads_summary %>% mutate(sig_fdr = ifelse(region_id %in% filter(dtu.optiles, FDR < 0.05)$region_id, TRUE, FALSE),
        sig_fdr_exon = ifelse(region_id %in% distinct(filter(dtu_tiles_5_exon, FDR < 0.05), region_id)$region_id, TRUE, FALSE),
        sig_fdr_delta = ifelse(region_id %in% filter(dtu.optiles, FDR < 0.05, abs(delta_b_minus_cml) > 0.2)$region_id, TRUE, FALSE),
        sig_dmr = ifelse(region_id %in% tiles_dmr$region_id, TRUE, FALSE))
```

```{r}
tiles_reads_summary %>%
  group_by(region_id, t_num, sig_fdr, sig_fdr_delta, sig_fdr_exon, sig_dmr) %>% summarise(cov = sum(coverage)) %>%
  filter(sig_dmr == TRUE, cov < 32) %>% nrow() # %>%
 # mutate(delta = bin_delta[match(.$region_id, bin_delta$region_id), "delta_b_minus_cml"])
```
* at least it's not many 
* hopefully they aren't ranked very high - like if delta is tiny that would be good
```{r}
reg <- tiles_reads_summary %>%
  group_by(region_id, t_num, sig_fdr, sig_fdr_delta, sig_fdr_exon, sig_dmr) %>% summarise(cov = sum(coverage)) %>%
  filter(sig_dmr == TRUE, cov < 32) %>% .$region_id
```

```{r}
tiles_dmr %>% filter(region_id %in% reg) %>% distinct(rank, no.cpgs, min_smoothed_fdr, maxdiff, meandiff) %>% nrow()
tiles_dmr %>% filter(region_id %in% reg) %>% distinct(rank, no.cpgs, min_smoothed_fdr, maxdiff, meandiff) %>% .[1:5,]
```

```{r}
tiles_dmr %>% mutate(low_cov_signif = ifelse(region_id %in% reg, TRUE, FALSE)) %>%
group_by(rank, low_cov_signif) %>% summarise(n=n()) %>% filter(n > 1) %>% nrow()
```

# will the coverage thing look better with 3 alpha categories??

* Logic: 5 groups is a lot - unclear if the method requires counts in each etc
* Using 3 may mean it doesn't have to work as hard

```{r}
dtu_tiles_3 <- dtu_tiles_3 %>% mutate(delta_b_minus_cml = bin_delta[match(.$region_id, bin_delta$region_id), "delta_b_minus_cml"])
```


```{r}
ggVennDiagram(list(alpha_5 = filter(dtu.optiles, FDR < 0.05, abs(delta_b_minus_cml) > 0.2)$region_id, 
                   alpha_3 = filter(dtu_tiles_3, FDR < 0.05, abs(delta_b_minus_cml) > 0.2)$region_id))
```
* of course 3 groups makes even more significance
* but what does it do to coverage??


```{r}
tiles_reads_summary %>% mutate(delta = bin_delta[match(.$region_id, bin_delta$region_id), "delta_b_minus_cml"]) %>% 
  mutate(sig_dtu_3 = ifelse(region_id %in% filter(dtu_tiles_3, FDR < 0.05, abs(delta_b_minus_cml) > 0.2)$region_id, TRUE, FALSE),
        sig_dtu = ifelse(region_id %in% filter(dtu.optiles, FDR < 0.05, abs(delta_b_minus_cml) > 0.2)$region_id, TRUE, FALSE),
        sig_dmr = ifelse(region_id %in% tiles_dmr$region_id, TRUE, FALSE)) %>%
  group_by(region_id, t_num, sig_dtu, sig_dtu_3, sig_dmr) %>% summarise(cov = sum(coverage)) %>%
  pivot_longer(cols = c(sig_dtu, sig_dtu_3, sig_dmr), names_to = "method", values_to = "signif") %>%
  ggplot(aes(x = method, y = log2(cov), fill = signif)) +
  geom_boxplot()
```
* huh no change in dtu vs dtu with 3 groups - ok
* again this was actually coverage in the tile - not the dmr itself

```{r}
find_overlaps(as_granges(dmrcate_seq_dmr), as_granges(all_samples_simp)) %>% data.frame() %>% 
  group_by(rank, sample) %>% summarise(coverage = n()) %>% ungroup() %>% data.frame() %>%
  group_by(rank) %>% summarise(cov = sum(coverage)) %>% ungroup() %>% mutate(a = "a") %>%
  ggplot(aes(x = a, y = log2(cov))) +
  geom_boxplot()
```
* ok this does make the mean a bit higher, tail is still longer though

```{r}
tiles_reads <- find_overlaps(as_granges(optiles_all), as_granges(all_samples_simp)) %>% data.frame()
tiles_reads_summary <- tiles_reads %>% group_by(region_id, t_num, sample) %>% summarise(coverage = n()) %>% ungroup()
tiles_reads_summary <- data.frame(tiles_reads_summary)
```

* ok is it the least significant dmrs with low coverage???? -please let me have something

* oh wait it's same as before where mid beta was for tile instead of dmr - not the actual length
(fixing it other time didn't actually change the plot but we can only hope)

```{r}
dmr_cov <- find_overlaps(as_granges(dmrcate_seq_dmr), as_granges(all_samples_simp)) %>% data.frame() %>% 
  group_by(rank, sample) %>% summarise(coverage = n()) %>% ungroup() %>% data.frame()
dmr_cov %>% group_by(rank) %>% summarise(cov = sum(coverage)) %>% ungroup() %>% mutate(dmr = "dmr") %>%
  ggplot(aes(x = dmr, y = cov)) +
  geom_boxplot()

dmr_cov %>% group_by(rank) %>% summarise(cov = sum(coverage)) %>% ungroup() %>% mutate(dmr = "dmr") %>%
  ggplot(aes(x = dmr, y = log2(cov))) +
  geom_boxplot()
```
* and again looks exact same boo
```{r}
dmr_cov %>% group_by(rank) %>% summarise(cov = sum(coverage)) %>% ungroup() %>%
  ggplot(aes(x = rank, y = cov)) +
  geom_point(alpha = 0.3)
```

* kinda - least significant have less coverage

### building website screwed up around here - 


Error: ! in callr subprocess.
Caused by error in `data[[var]][scale_index[[i]]]`:
! Can't subset columns past the end.
ℹ Locations 2, 3, 4, …, 187739, and 187740 don't exist.
ℹ There is only 1 column.

CHUNK 44, 45, 46

```{r, eval=FALSE}
dmr_cov %>% group_by(rank) %>% summarise(cov = sum(coverage)) %>% ungroup() %>%
  ggplot(aes(x = rank, y = cov)) +
  geom_point(alpha = 0.1) +
  scale_y_continuous(limits = c(0,5000))
```

```{r, eval=FALSE}
dmr_cov %>% group_by(rank) %>% summarise(cov = sum(coverage)) %>% ungroup() %>%
  ggplot(aes(x = rank, y = log2(cov))) +
  geom_point(alpha = 0.1)
```
* ok this does show it

* except same thing with dtu

```{r, eval=FALSE}
tiles_reads_summary %>% mutate(rank = dtu.optiles[match(.$region_id, dtu.optiles$region_id), "dtu_rank"]) %>% 
  group_by(rank) %>% summarise(cov = sum(coverage)) %>% ungroup() %>%
  ggplot(aes(x = rank, y = cov)) +
  geom_point(alpha = 0.1)

tiles_reads_summary %>% mutate(rank = dtu.optiles[match(.$region_id, dtu.optiles$region_id), "dtu_rank"]) %>% 
  group_by(rank) %>% summarise(cov = sum(coverage)) %>% ungroup() %>%
  ggplot(aes(x = rank, y = log2(cov))) +
  geom_point(alpha = 0.1)
```
* well huh
* this is weird and unnatural looking and wrong
* but at least it worked in dmrs
```{r, eval=FALSE}
tiles_reads_summary %>% mutate(rank = dtu.optiles[match(.$region_id, dtu.optiles$region_id), "dtu_rank"]) %>% 
  group_by(rank, region_id) %>% summarise(cov = sum(coverage)) %>% ungroup() %>% .[order(.$cov, decreasing = TRUE),] %>% .[1:5,]
```

```{r, eval=FALSE}
tiles_reads_summary %>% mutate(rank = dtu.optiles[match(.$region_id, dtu.optiles$region_id), "dtu_rank"]) %>% 
  group_by(rank) %>% summarise(cov = sum(coverage)) %>% ungroup() %>%
  ggplot(aes(x = rank, y = cov)) +
  geom_point(alpha = 0.1) +
  scale_y_continuous(limits = c(0, 25000))
```




tiles_reads_summary %>% mutate(delta = bin_delta[match(.$region_id, bin_delta$region_id), "delta_b_minus_cml"]) %>% 
  mutate(sig_dtu_3 = ifelse(region_id %in% filter(dtu_tiles_3, FDR < 0.05, abs(delta_b_minus_cml) > 0.2)$region_id, TRUE, FALSE),
        sig_dtu = ifelse(region_id %in% filter(dtu.optiles, FDR < 0.05, abs(delta_b_minus_cml) > 0.2)$region_id, TRUE, FALSE),
        sig_dmr = ifelse(region_id %in% tiles_dmr$region_id, TRUE, FALSE)) %>%





