---
title: "differentialAlphaUsage"
author: "caitlinpage"
date: "2026-02-20"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r}
library(annotatr)
library(data.table)
library(plyranges)
library(tidyr)
library(stringr)
library(dplyr)
library(ggplot2)
library(edgeR)
```



```{r}
find_overlaps(as_granges(all_samples), as_granges(dmrcate_seq_dmr[1,])) %>% data.frame()
```




```{r}
find_overlaps(as_granges(all_samples), as_granges(dmrcate_seq_dmr[1,])) %>% data.frame() %>%
  distinct(sample, read_id, read_start, read_end, alpha) %>%
  ggplot() +
  geom_segment(aes(x = read_start, xend = read_end, y = alpha, colour = read_id)) +
  theme(legend.position = "none") +
  facet_wrap(~ sample)
```

```{r}
find_overlaps(as_granges(all_samples), as_granges(dmrcate_seq_dmr[1,])) %>% data.frame() %>%
  group_by(sample, read_id, read_start, read_end, alpha) %>%
  summarise(count = n()) %>% ungroup() %>% filter(count > 1)
```


```{r}
find_overlaps(as_granges(all_samples), as_granges(dmrcate_seq_dmr[1,])) %>% data.frame() %>%
  group_by(sample, read_id, read_start, read_end, alpha) %>%
  summarise(count = n()) %>% ungroup() %>%
  ggplot() +
  geom_segment(aes(x = read_start, xend = read_end, y = alpha, colour = read_id, size = count)) +
  theme(legend.position = "none") +
  facet_wrap(~ sample)
```


```{r}
find_overlaps(as_granges(all_samples), as_granges(dmrcate_seq_dmr[1,])) %>% data.frame() %>%
  group_by(sample, read_id, read_start, read_end, alpha) %>%
  summarise(count = n()) %>% ungroup() %>%
  ggplot() +
  geom_segment(aes(x = read_start, xend = read_end, y = alpha, colour = read_id, linewidth = count)) +
  theme(legend.position = "none") +
  facet_wrap(~ sample)
```



rrbs_plot <- function(long_data, read_ids, cpgs = cg_sites, count_instead = FALSE) {
  if(count_instead == FALSE) {
  plot <- long_data %>% dplyr::filter(read_id %in% read_ids) %>%
    distinct(read_id, read_start, read_end, alpha) %>%
    ggplot() +
    geom_segment(aes(x = read_start, xend = read_end, y = alpha, yend = alpha), alpha = 0.2) +
    geom_point(data = dplyr::filter(long_data, read_id %in% read_ids),
               aes(x = start, y = alpha, colour = meth_status)) +
    geom_point(data = distinct(dplyr::filter(long_data, read_id %in% read_ids), start, beta),
               aes(x = start, y = beta), shape = 4, size = 3) +
    geom_segment(data = filter(cg_sites, start >= min(dplyr::filter(long_data, read_id %in% read_ids)$start),
                               end <= max(dplyr::filter(long_data, read_id %in% read_ids)$end)),
                 aes(x = start, xend = start, y = -0.15, yend = -0.05), colour = "green")
  } else {
    plot <- long_data %>% dplyr::filter(read_id %in% read_ids) %>%
      distinct(read_id, read_start, read_end, alpha) %>%
      ggplot() +
      geom_segment(aes(x = read_start, xend = read_end, y = alpha, yend = alpha), alpha = 0.2) +
      geom_count(data = dplyr::filter(long_data, read_id %in% read_ids),
                 aes(x = start, y = alpha, colour = meth_status)) +
      geom_point(data = distinct(dplyr::filter(long_data, read_id %in% read_ids), start, beta),
                 aes(x = start, y = beta), shape = 4, size = 3) +
      geom_segment(data = filter(cg_sites, start >= min(dplyr::filter(long_data, read_id %in% read_ids)$start),
                                 end <= max(dplyr::filter(long_data, read_id %in% read_ids)$end)),
                   aes(x = start, xend = start, y = -0.15, yend = -0.05), colour = "green")
  }
  plot
}
## Data Objects
```{r}
cg_sites <- cbind(data.frame(Biostrings::matchPattern("CG", BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38[["chr22"]])), seqnames = "chr22") %>% 
  mutate(pos = paste0(seqnames, "-", start)) %>%
  relocate(pos, seqnames)
cg_sites$index <- 1:nrow(cg_sites)
```

* GM1268 cell line
```{r}
source("../code/processData.R")
for_beta_all <- readRDS("/researchers/caitlin.page/cf_nano/r_output/rrbs_alpha_beta.rds")
#for_beta <- readRDS("../output/cluster_files/rrbs_alpha_beta.rds")
for_beta <- for_beta_all %>% filter(seqnames == "chr22")
rrbs_for_beta_pat <- pat_for_wgbs(for_beta)

for_beta <- for_beta %>% group_by(start) %>% mutate(coverage = n(), mean_alpha = mean(alpha)) %>% ungroup() %>% data.frame()
just_beta <- for_beta %>% distinct(start, beta, coverage, mean_alpha) %>% .[order(.$start),]

```

* REP 2 GM12878 cell line
```{r}
source("../code/processData.R")
gm12878_2 <- readRDS("/researchers/caitlin.page/cf_nano/r_output/rrbs_alpha_beta_rep2.rds")
#for_beta <- readRDS("../output/cluster_files/rrbs_alpha_beta.rds")
gm12878_2_chr22 <- gm12878_2 %>% filter(seqnames == "chr22")

gm12878_2_chr22 <- gm12878_2_chr22 %>% group_by(start) %>% mutate(coverage = n(), mean_alpha = mean(alpha)) %>% ungroup() %>% data.frame()
```

* K562 cml cell line
```{r}
source("../code/processData.R")
k562 <- readRDS("/researchers/caitlin.page/cf_nano/r_output/k562_rep1.rds")
k562 <- k562 %>% as_granges() %>% data.frame()
k562_chr22 <- k562 %>% filter(seqnames == "chr22")

k562_chr22 <- k562_chr22 %>% group_by(start) %>% mutate(coverage = n(), mean_alpha = mean(alpha)) %>% ungroup() %>% data.frame()
```

* REP 2 K562
```{r}
source("../code/processData.R")
k562_2 <- readRDS("/researchers/caitlin.page/cf_nano/r_output/k562_rep2.rds")

k562_2 <- k562_2 %>% as_granges() %>% data.frame()
k562_2_chr22 <- k562_2 %>% filter(seqnames == "chr22")

k562_2_chr22 <- k562_2_chr22 %>% group_by(start) %>% mutate(coverage = n(), mean_alpha = mean(alpha)) %>% ungroup() %>% data.frame()
```

* add cpg islands and gene annotations

annots <- c("hg38_cpgs")

annotations <- annotatr::build_annotations(genome = 'hg38', annotations = annots)
seqs <- annotations %>% data.frame() %>% distinct(seqnames) %>% .[1:22,]
annotations <- annotations %>% data.frame() %>% filter(seqnames %in% seqs)

for_beta <- find_overlaps(as_granges(for_beta), as_granges(annotations)) %>% data.frame() %>%
  mutate(cgi = ifelse(type == "hg38_cpg_islands", "island", "not_island"))
gm12878_2_chr22 <- find_overlaps(as_granges(gm12878_2_chr22), as_granges(annotations)) %>% data.frame() %>%
  mutate(cgi = ifelse(type == "hg38_cpg_islands", "island", "not_island"))
k562_chr22 <- find_overlaps(as_granges(k562_chr22), as_granges(annotations)) %>% data.frame() %>%
  mutate(cgi = ifelse(type == "hg38_cpg_islands", "island", "not_island"))
k562_2_chr22 <- find_overlaps(as_granges(k562_2_chr22), as_granges(annotations)) %>% data.frame() %>%
  mutate(cgi = ifelse(type == "hg38_cpg_islands", "island", "not_island"))


```{r}
annots <- c("hg38_basicgenes", "hg38_genes_intergenic")

annotations <- annotatr::build_annotations(genome = 'hg38', annotations = annots)
seqs <- annotations %>% data.frame() %>% distinct(seqnames) %>% .[1:22,]
annotations <- annotations %>% data.frame() %>% filter(seqnames %in% seqs)
annotations %>% group_by(type) %>% summarise(n=n())

#for_beta_genes <- find_overlaps(as_granges(for_beta), as_granges(annotations)) %>% data.frame()
gm12878_2_chr22_genes <- find_overlaps(as_granges(gm12878_2_chr22), as_granges(annotations)) %>% data.frame()
k562_chr22_genes <- find_overlaps(as_granges(k562_chr22), as_granges(annotations)) %>% data.frame()
k562_2_chr22_genes <- find_overlaps(as_granges(k562_2_chr22), as_granges(annotations)) %>% data.frame()
k562_2_chr22_genes[1:10,]
```



# plot check

```{r}
for_beta %>%
  mutate(colour_0.5 = ifelse(beta == 0.5, "blue", "black")) %>%
  distinct(seqnames, start, mean_alpha, beta, colour_0.5) %>%
  ggplot(aes(x = beta, y = mean_alpha, colour = colour_0.5)) +
  geom_point(alpha = 0.4) +
  scale_colour_manual(values = c("black", "blue")) +
  geom_abline(colour = "red", linetype = "dashed") +
  theme_bw()
```

```{r}
k562_chr22 %>%
  mutate(colour_0.5 = ifelse(beta == 0.5, "blue", "black")) %>%
  distinct(seqnames, start, mean_alpha, beta, colour_0.5) %>%
  ggplot(aes(x = beta, y = mean_alpha, colour = colour_0.5)) +
  geom_point(alpha = 0.4) +
  scale_colour_manual(values = c("black", "blue")) +
  geom_abline(colour = "red", linetype = "dashed") +
  theme_bw()
```


# differential transcript usage

CONDITIONS: GM12878 X K562 cell lines

Lymphoblastoid (B cell) X lymphoblast from CML
(not a great comparison to pick because different cell types and cancer/non cancer)
but differential methylation should be associated with cancer at least

* apply it to alpha/meth pattern

* realistically we care about methylation in a genomic context

* so overlap reads to gene promoters
* DTU is change in relative abundance of transcripts from same gene between experimental conditions


```{r}
for_beta_genes %>% distinct(read_id, alpha, type.y) %>%
  mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8")) %>%
  ggplot(aes(fill = type.y, x = alpha_group)) +
  geom_bar(position = "fill") 
```
* k562
```{r}
k562_chr22_genes %>% distinct(read_id, alpha, type.y) %>%
  mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8")) %>%
  ggplot(aes(fill = type.y, x = alpha_group)) +
  geom_bar(position = "fill") 
```


```{r}
for_beta_genes %>% distinct(start, beta, type.y) %>%
  mutate(beta_group = case_when(beta < 0.2 ~ "b < 0.2",
                                beta > 0.8 ~ "b > 0.8",
                                beta >= 0.2 & beta <= 0.4 ~ "0.2 <= b <= 0.4",
                                beta > 0.4 & beta < 0.6 ~ "0.4 < b < 0.6",
                                TRUE ~ "0.6 <= b <= 0.8")) %>%
  ggplot(aes(x = type.y, fill = beta_group)) +
  geom_bar(position = "fill") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

* k562

```{r}
k562_chr22_genes %>% distinct(start, beta, type.y) %>%
  mutate(beta_group = case_when(beta < 0.2 ~ "b < 0.2",
                                beta > 0.8 ~ "b > 0.8",
                                beta >= 0.2 & beta <= 0.4 ~ "0.2 <= b <= 0.4",
                                beta > 0.4 & beta < 0.6 ~ "0.4 < b < 0.6",
                                TRUE ~ "0.6 <= b <= 0.8")) %>%
  ggplot(aes(x = type.y, fill = beta_group)) +
  geom_bar(position = "fill") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# conducting DTU analysis

* group alphas into ranges
[0,0.2)
[0.2, 0.4]
(0.4, 0.6)
[0.6, 0.8]
(0.8, 1]

* these serve as 5 possible "transcripts"
* only using reads that overlap to promoters as identified by `annotatr`


[2025 benchmark paper](https://academic.oup.com/nargab/article/7/3/lqaf117/8251390#531509445)
* edgeR is listed as one of the top performers

[edgeR documentation](https://bioconductor.org/packages/release/bioc/vignettes/edgeR/inst/doc/edgeRUsersGuide.pdf)
* Section 2.20 for description (page 30)
* Section 4.6 for example (page 90)

```{r}
dtu_for_beta <- for_beta_genes %>% distinct(read_id, alpha, gene_id.y, symbol.y, read_start, read_end) %>%
  mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8"))
dtu_beta_2 <- gm12878_2_chr22_genes %>% distinct(read_id, alpha, gene_id, symbol, read_start, read_end) %>%
  mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8"))

dtu_k562 <- k562_chr22_genes %>% distinct(read_id, alpha, gene_id, symbol, read_start, read_end) %>%
  mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8"))

dtu_k562_2 <- k562_2_chr22_genes %>% distinct(read_id, alpha, gene_id, symbol, read_start, read_end) %>%
  mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8"))
dtu_k562_2[1:10,]
```

* ok to fudge this I need the counts of the "transcripts"

```{r}
trans_count_beta <- dtu_for_beta %>% 
  mutate(transcript = paste0(symbol.y, "_", alpha_group)) %>%
  group_by(transcript) %>%
  summarise(transcript_count = n()) %>% data.frame()

trans_count_beta2 <- dtu_beta_2 %>% 
  mutate(transcript = paste0(symbol, "_", alpha_group)) %>%
  group_by(transcript) %>%
  summarise(transcript_count = n()) %>% data.frame()

trans_count_k <- dtu_k562 %>% 
  mutate(transcript = paste0(symbol, "_", alpha_group)) %>%
  group_by(transcript) %>%
  summarise(transcript_count = n()) %>% data.frame()

trans_count_k_2 <- dtu_k562_2 %>% 
  mutate(transcript = paste0(symbol, "_", alpha_group)) %>%
  group_by(transcript) %>%
  summarise(transcript_count = n()) %>% data.frame()
trans_count_k_2[1:10,]
```

```{r}
edger_counts <- c(dtu_for_beta %>% 
  mutate(transcript = paste0(symbol.y, "_", alpha_group)) %>% .$transcript,
  dtu_k562 %>% 
  mutate(transcript = paste0(symbol, "_", alpha_group)) %>% .$transcript,
  dtu_beta_2 %>% 
  mutate(transcript = paste0(symbol, "_", alpha_group)) %>% .$transcript,
  dtu_k562_2 %>% 
  mutate(transcript = paste0(symbol, "_", alpha_group)) %>% .$transcript) %>% data.frame() %>% distinct()
names(edger_counts) <- "transcript"
edger_counts <- edger_counts %>% mutate(Sample1_gm = trans_count_beta[match(.$transcript, trans_count_beta$transcript), "transcript_count"],
                                        Sample2_gm = trans_count_beta2[match(.$transcript, trans_count_beta2$transcript), "transcript_count"],
                        Sample3_k = trans_count_k[match(.$transcript, trans_count_k$transcript), "transcript_count"],
                        Sample4_k = trans_count_k_2[match(.$transcript, trans_count_k_2$transcript), "transcript_count"])
edger_counts[is.na(edger_counts)] <- 0
edger_counts$gene_id <- sub("_.*", "", edger_counts$transcript)
edger_counts <- edger_counts %>% filter(!gene_id == "NA") 
rownames(edger_counts) <- edger_counts$transcript
edger_counts[1:10,]
```

```{r}
y <- edgeR::DGEList(as.matrix(edger_counts[,2:5]), group = c(1,1,2,2), genes = edger_counts[,c("gene_id", "transcript")])
y
```


* main steps
- dge list
- design
- fit
- test
- top

```{r}
design <- model.matrix(~ c(1,1,2,2), data = y$samples)
```

```{r}
fit <- edgeR::glmQLFit(y, design, robust=TRUE)
fit
```
```{r}
colnames(fit$genes)
mean(table(fit$genes$gene_id))
```

```{r}
ds <- edgeR::diffSpliceDGE(fit, coef = 2, geneid = "gene_id", exonid = "transcript")
```
```{r}
ds
```
```{r}
ts.gene.simes <- edgeR::topSpliceDGE(ds, test = "Simes", number = Inf)
ts.gene.simes[1:10,]
```

```{r}
ts.gene.gene <- edgeR::topSpliceDGE(ds, test = "gene", number = Inf)
ts.gene.gene[1:10,]
```

* odd because in documentation, the F column appears in the global test, not in the genes test. And in the documentation, the gene test is Simes


```{r}
#exon not recc but doing it anyway
ts.gene.exon <- topSpliceDGE(ds, test = "exon", number = Inf)
ts.gene.exon[1:10,]
```



Ranks genes or exons by evidence for differential splicing. The exon-level tests test for differences between each exon and all the exons for the same gene. The gene-level tests test for any differences in exon usage between experimental conditions.

The Simes method processes the exon-level p-values to give an overall call of differential splicing for each gene. It returns the minimum Simes-adjusted p-values for each gene.

The gene-level tests are likely to be powerful for genes in which several exons are differentially splices. The Simes p-values is likely to be more powerful when only a minority of the exons for a gene are differentially spliced. The exon-level tests are not recommended for formal error rate control.


# results as expression proportions

* except of course the functions listed in the example are not found anywhere 
 TPM <- tpm(fit, fit$genes$EffectiveLength, fit$genes$Overdispersion)
> TPMProp <- tpmProp(TPM, y$genes$SYMBOL)
> g <- which(fit$genes$SYMBOL=="HLA-A")
> TPM[g,]

* asked perplexity and it has provided code to calculate it
* oops need a length
* or maybe that's the number of cpgs?

```{r}
gene_cg <- find_overlaps(as_granges(annotations), as_granges(cg_sites)) %>% data.frame() %>%
  group_by(gene_id, symbol) %>%
  summarise(n_cg = n()) %>%
  ungroup() %>% data.frame()

fit$genes <- fit$genes %>% mutate(n_cg = gene_cg[match(.$gene_id, gene_cg$symbol), "n_cg"])
```


```{r}
rpk <- y$counts/rep(fit$genes$n_cg/1000, times = ncol(y$counts))
tpm_mat <- sweep(rpk, 2, colSums(rpk)/1e6, "/")
```


```{r}
tpm_mat <- tpm_mat %>% data.frame() %>% mutate(id = edger_counts$transcript, gene_id = edger_counts$gene_id)
tpm_mat[1:10,]
```

* now add in proportions
```{r}
tpm_mat <- tpm_mat %>% 
  group_by(gene_id) %>% 
  mutate(prop_S1_gm = Sample1_gm/sum(Sample1_gm),
         prop_S2_gm = Sample2_gm/sum(Sample2_gm),
         prop_S3_k = Sample3_k/sum(Sample3_k),
         prop_S4_k = Sample4_k/sum(Sample4_k)) %>%
  ungroup() %>% data.frame()
tpm_mat %>% .[order(.$gene_id),] %>% .[1:10,]
```
* nice can see even in these 2 genes, difference in proportions between the groups

```{r}
ts.gene.simes[1:10,]
tail(ts.gene.simes)
c(signif = ts.gene.simes %>% filter(FDR < 0.05) %>% nrow(),
  total = nrow(ts.gene.simes))
```

```{r}
tpm_mat
ts.gene.gene
ts.gene.simes
```
```{r}
saveRDS(tpm_mat, "../output/cluster_files/dtu_prop.rds")
saveRDS(ts.gene.gene, "../output/cluster_files/dtu_res_gene.rds")
saveRDS(ts.gene.simes, "../output/cluster_files/dtu_res_simes.rds")
```


```{r}
tpm_mat %>% filter(gene_id == ts.gene.simes$gene_id[1])
```

```{r}
tpm_mat %>% filter(gene_id == ts.gene.simes$gene_id[643])
```


* picking most and least significant can see it worked
* in least, proportions are same across the cell lines
* in most, proportions are different across the cell lines

```{r}
tpm_mat %>% mutate(alpha_group = sub(".*_", "", id)) %>% .[,6:11] %>%
  filter(gene_id %in% ts.gene.simes$gene_id[1:10]) %>%
  reshape2::melt(id = c("alpha_group", "gene_id")) %>%
  ggplot(aes(x = variable, y = value, fill = alpha_group)) +
  geom_col() +
  scale_x_discrete(labels = c("B_cell_1", "B_cell_2", "CML_1", "CML_2")) +
  facet_wrap(~ gene_id)
```

```{r}
tpm_mat %>% mutate(alpha_group = sub(".*_", "", id)) %>% .[,6:11] %>%
  filter(gene_id %in% ts.gene.simes$gene_id[1:3]) %>%
  reshape2::melt(id = c("alpha_group", "gene_id")) %>%
  ggplot(aes(x = variable, y = value, fill = alpha_group)) +
  geom_col() +
  scale_x_discrete(labels = c("B_cell_1", "B_cell_2", "CML_1", "CML_2")) +
  facet_wrap(~ gene_id) +
  theme(legend.position = "none")
```

```{r}
tpm_mat %>% mutate(alpha_group = sub(".*_", "", id)) %>% .[,6:11] %>%
  filter(gene_id %in% ts.gene.simes$gene_id[1:3])
```

```{r}
rbind(for_beta_genes %>% filter(symbol.y %in% ts.gene.simes[1:3,"gene_id"]) %>% mutate(symbol = symbol.y) %>% distinct(symbol, read_id, alpha) %>% mutate(sample = "b1"),
      gm12878_2_chr22_genes %>% filter(symbol %in% ts.gene.simes[1:3,"gene_id"]) %>% distinct(symbol, read_id, alpha) %>% mutate(sample = "b2"),
      k562_chr22_genes %>% filter(symbol %in% ts.gene.simes[1:3,"gene_id"]) %>% distinct(symbol, read_id, alpha) %>% mutate(sample = "cml1"),
      k562_2_chr22_genes %>% filter(symbol %in% ts.gene.simes[1:3,"gene_id"]) %>% distinct(symbol, read_id, alpha) %>% mutate(sample = "cml2")) %>% data.frame() %>%
  ggplot(aes(x = sample, y = alpha, fill = sample)) +
  geom_violin() +
  theme_bw() +
  theme(legend.position = "none")

rbind(for_beta_genes %>% filter(symbol.y %in% ts.gene.simes[1:3,"gene_id"]) %>% mutate(symbol = symbol.y) %>% distinct(symbol, read_id, alpha) %>% mutate(sample = "b1"),
      gm12878_2_chr22_genes %>% filter(symbol %in% ts.gene.simes[1:3,"gene_id"]) %>% distinct(symbol, read_id, alpha) %>% mutate(sample = "b2"),
      k562_chr22_genes %>% filter(symbol %in% ts.gene.simes[1:3,"gene_id"]) %>% distinct(symbol, read_id, alpha) %>% mutate(sample = "cml1"),
      k562_2_chr22_genes %>% filter(symbol %in% ts.gene.simes[1:3,"gene_id"]) %>% distinct(symbol, read_id, alpha) %>% mutate(sample = "cml2")) %>% data.frame() %>%
  ggplot(aes(x = sample, y = alpha, fill = sample)) +
  geom_violin() +
  theme_bw() +
  theme(legend.position = "none") +
  facet_wrap(~ symbol)

rbind(for_beta_genes %>% filter(symbol.y %in% ts.gene.simes[1:3,"gene_id"]) %>% distinct(read_id, alpha) %>% mutate(sample = "b1"),
      gm12878_2_chr22_genes %>% filter(symbol %in% ts.gene.simes[1:3,"gene_id"]) %>% distinct(read_id, alpha) %>% mutate(sample = "b2"),
      k562_chr22_genes %>% filter(symbol %in% ts.gene.simes[1:3,"gene_id"]) %>% distinct(read_id, alpha) %>% mutate(sample = "cml1"),
      k562_2_chr22_genes %>% filter(symbol %in% ts.gene.simes[1:3,"gene_id"]) %>% distinct(read_id, alpha) %>% mutate(sample = "cml2")) %>% data.frame() %>%
  ggplot(aes(x = alpha, colour = sample)) +
  geom_density() +
  theme_bw() 
```




* bottom 10 genes

```{r}
tpm_mat %>% mutate(alpha_group = sub(".*_", "", id)) %>% .[,6:11] %>%
  filter(gene_id %in% ts.gene.simes$gene_id[c(640,641,643)]) %>%
  reshape2::melt(id = c("alpha_group", "gene_id")) %>%
  ggplot(aes(x = variable, y = value, fill = alpha_group)) +
  geom_col() +
  scale_x_discrete(labels = c("B_cell_1", "B_cell_2", "CML_1", "CML_2")) +
  facet_wrap(~ gene_id)

tpm_mat %>% mutate(alpha_group = sub(".*_", "", id)) %>% .[,6:11] %>%
  filter(gene_id %in% ts.gene.simes$gene_id[c(640,641,643)]) %>%
  reshape2::melt(id = c("alpha_group", "gene_id")) %>%
  ggplot(aes(x = variable, y = value, fill = alpha_group)) +
  geom_col() +
  scale_x_discrete(labels = c("B_cell_1", "B_cell_2", "CML_1", "CML_2")) +
  facet_wrap(~ gene_id) +
  theme(legend.position = "none")
```

```{r}
rbind(for_beta_genes %>% filter(symbol.y %in% ts.gene.simes[c(640,641,643),"gene_id"]) %>% mutate(symbol = symbol.y) %>% distinct(symbol, read_id, alpha) %>% mutate(sample = "b1"),
      gm12878_2_chr22_genes %>% filter(symbol %in% ts.gene.simes[c(640,641,643),"gene_id"]) %>% distinct(symbol, read_id, alpha) %>% mutate(sample = "b2"),
      k562_chr22_genes %>% filter(symbol %in% ts.gene.simes[c(640,641,643),"gene_id"]) %>% distinct(symbol, read_id, alpha) %>% mutate(sample = "cml1"),
      k562_2_chr22_genes %>% filter(symbol %in% ts.gene.simes[c(640,641,643),"gene_id"]) %>% distinct(symbol, read_id, alpha) %>% mutate(sample = "cml2")) %>% data.frame() %>%
  ggplot(aes(x = sample, y = alpha, fill = sample)) +
  geom_violin() +
  theme_bw() +
  theme(legend.position = "none") +
  facet_wrap(~ symbol)
```


```{r}
tpm_mat %>% mutate(alpha_group = sub(".*_", "", id)) %>% .[,6:11] %>%
  filter(gene_id %in% ts.gene.simes$gene_id[634:643]) %>%
  reshape2::melt(id = c("alpha_group", "gene_id")) %>%
  ggplot(aes(x = variable, y = value, fill = alpha_group)) +
  geom_col() +
  scale_x_discrete(labels = c("B_cell_1", "B_cell_2", "CML_1", "CML_2")) +
  facet_wrap(~ gene_id)
```

* gene level: differences in exon usage between conditions
* Simes:

* gene: better when many exons differentially splices
* Simes: better when minority of exons are differentially spliced

* look up the alphas in the top genes for the different samples


# repeat for all chr
```{r}
annots <- c("hg38_basicgenes", "hg38_genes_intergenic")

annotations <- annotatr::build_annotations(genome = 'hg38', annotations = annots)
seqs <- annotations %>% data.frame() %>% distinct(seqnames) %>% .[1:22,]
annotations <- annotations %>% data.frame() %>% filter(seqnames %in% seqs)


for_beta_all <- for_beta_all %>% filter(seqnames %in% seqs)
gm12878_2 <- gm12878_2 %>% filter(seqnames %in% seqs)
k562 <- k562 %>% filter(seqnames %in% seqs)
k562_2 <- k562_2 %>% filter(seqnames %in% seqs)

for_beta_all_genes <- find_overlaps(as_granges(for_beta_all), as_granges(annotations)) %>% data.frame()
gm12878_2_genes <- find_overlaps(as_granges(gm12878_2), as_granges(annotations)) %>% data.frame()
k562_genes <- find_overlaps(as_granges(k562), as_granges(annotations)) %>% data.frame()
k562_2_genes <- find_overlaps(as_granges(k562_2), as_granges(annotations)) %>% data.frame()
```

```{r}
distinct(for_beta_all, seqnames) %>% filter(seqnames %in% unique(annotations$seqnames))
distinct(for_beta_all, seqnames) %>% filter(seqnames %in% seqs)
```

```{r}
# set up
dtu_for_beta <- for_beta_all_genes %>% distinct(read_id, alpha, gene_id.y, symbol.y, seqnames, read_start, read_end) %>%
  mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8"))
dtu_beta_2 <- gm12878_2_genes %>% distinct(read_id, alpha, gene_id, symbol, seqnames, read_start, read_end) %>%
  mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8"))

dtu_k562 <- k562_genes %>% distinct(read_id, alpha, gene_id, symbol, seqnames, read_start, read_end) %>%
  mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8"))

dtu_k562_2 <- k562_2_genes %>% distinct(read_id, alpha, gene_id, symbol, seqnames, read_start, read_end) %>%
  mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8"))

trans_count_beta <- dtu_for_beta %>% 
  mutate(transcript = paste0(symbol.y, "_", alpha_group)) %>%
  group_by(transcript) %>%
  summarise(transcript_count = n()) %>% data.frame()

trans_count_beta2 <- dtu_beta_2 %>% 
  mutate(transcript = paste0(symbol, "_", alpha_group)) %>%
  group_by(transcript) %>%
  summarise(transcript_count = n()) %>% data.frame()

trans_count_k <- dtu_k562 %>% 
  mutate(transcript = paste0(symbol, "_", alpha_group)) %>%
  group_by(transcript) %>%
  summarise(transcript_count = n()) %>% data.frame()

trans_count_k_2 <- dtu_k562_2 %>% 
  mutate(transcript = paste0(symbol, "_", alpha_group)) %>%
  group_by(transcript) %>%
  summarise(transcript_count = n()) %>% data.frame()

edger_counts <- c(dtu_for_beta %>% 
  mutate(transcript = paste0(symbol.y, "_", alpha_group)) %>% .$transcript,
  dtu_k562 %>% 
  mutate(transcript = paste0(symbol, "_", alpha_group)) %>% .$transcript,
  dtu_beta_2 %>% 
  mutate(transcript = paste0(symbol, "_", alpha_group)) %>% .$transcript,
  dtu_k562_2 %>% 
  mutate(transcript = paste0(symbol, "_", alpha_group)) %>% .$transcript) %>% data.frame() %>% distinct()
names(edger_counts) <- "transcript"
edger_counts <- edger_counts %>% mutate(Sample1_gm = trans_count_beta[match(.$transcript, trans_count_beta$transcript), "transcript_count"],
                                        Sample2_gm = trans_count_beta2[match(.$transcript, trans_count_beta2$transcript), "transcript_count"],
                        Sample3_k = trans_count_k[match(.$transcript, trans_count_k$transcript), "transcript_count"],
                        Sample4_k = trans_count_k_2[match(.$transcript, trans_count_k_2$transcript), "transcript_count"])
edger_counts[is.na(edger_counts)] <- 0
edger_counts$gene_id <- sub("_.*", "", edger_counts$transcript)
edger_counts <- edger_counts %>% filter(!gene_id == "NA") 
rownames(edger_counts) <- edger_counts$transcript

#testing
y <- edgeR::DGEList(as.matrix(edger_counts[,2:5]), group = c(1,1,2,2), genes = edger_counts[,c("gene_id", "transcript")])

design <- model.matrix(~ c(1,1,2,2), data = y$samples)

fit <- edgeR::glmQLFit(y, design, robust=TRUE)

ds <- edgeR::diffSpliceDGE(fit, coef = 2, geneid = "gene_id", exonid = "transcript")

ts.gene.simes <- edgeR::topSpliceDGE(ds, test = "Simes", number = Inf)

ts.gene.gene <- edgeR::topSpliceDGE(ds, test = "gene", number = Inf)

gene_cg <- find_overlaps(as_granges(annotations), as_granges(cg_sites)) %>% data.frame() %>%
  group_by(gene_id, symbol) %>%
  summarise(n_cg = n()) %>%
  ungroup() %>% data.frame()

# proportions to visualise
fit$genes <- fit$genes %>% mutate(n_cg = gene_cg[match(.$gene_id, gene_cg$symbol), "n_cg"])

rpk <- y$counts/rep(fit$genes$n_cg/1000, times = ncol(y$counts))
tpm_mat <- sweep(rpk, 2, colSums(rpk)/1e6, "/")

tpm_mat <- tpm_mat %>% data.frame() %>% mutate(id = edger_counts$transcript, gene_id = edger_counts$gene_id)

tpm_mat <- tpm_mat %>% 
  group_by(gene_id) %>% 
  mutate(prop_S1_gm = Sample1_gm/sum(Sample1_gm),
         prop_S2_gm = Sample2_gm/sum(Sample2_gm),
         prop_S3_k = Sample3_k/sum(Sample3_k),
         prop_S4_k = Sample4_k/sum(Sample4_k)) %>%
  ungroup() %>% data.frame()
# save results
saveRDS(tpm_mat, "../output/cluster_files/dtu_all_prop.rds")
saveRDS(ts.gene.gene, "../output/cluster_files/dtu_all_res_gene.rds")
saveRDS(ts.gene.simes, "../output/cluster_files/dtu_all_res_simes.rds")
```



# jovana single cell version:

* big table
* biological feature: try promoters

Columns:    Sample1Read1. Sample1Read2 ... Sample2Read1...
Rows: Gene1
      Gene2
```{r}
k562 <- k562[,colnames(rrbs_alpha_beta)]
k562_2 <- k562_2[,colnames(rrbs_alpha_beta)]
```

```{r}
all_samples <- rbind(rrbs_alpha_beta %>% mutate(sample = "b1"), gm12878_2 %>% mutate(sample = "b2"), 
                     k562 %>% mutate(sample = "cml1"), k562_2 %>% mutate(sample = "cml2"))
```

```{r}
annots <- c("hg38_basicgenes", "hg38_genes_intergenic")

annotations <- annotatr::build_annotations(genome = 'hg38', annotations = annots)
seqs <- annotations %>% data.frame() %>% distinct(seqnames) %>% .[1:22,]
annotations <- annotations %>% data.frame() %>% filter(seqnames %in% seqs)
annotations <- annotations[annotations$type == "hg38_genes_promoters",]

all_samples_simple <- all_samples %>% distinct(sample, read_id, seqnames, read_start, read_end, alpha, meth_pattern, num_meth, total)
all_samples_simple$start <- all_samples_simple$read_start
all_samples_simple$end <- all_samples_simple$read_end
all_samples_genes <- find_overlaps(as_granges(all_samples_simple), as_granges(annotations)) %>% data.frame()
all_samples_genes <- all_samples_genes[!is.na(all_samples_genes$symbol),]
```

* so there's multiple promoters per gene here because of multiple transcripts
* but going to ignore that for now
* just use the promoter id
* forgot TxDb was an option


```{r}
all_samples_genes %>%
  distinct(sample, read_id, alpha, gene_id, symbol) %>%
  group_by(sample, symbol) %>%
  mutate(alpha_id = paste0(sample, "read", 1:n())) %>% .[order(.$symbol),] %>% ungroup() %>% .[,c(3,5,6)]
```

```{r}
single_cell <- all_samples_genes %>%
  distinct(sample, read_id, alpha, gene_id, symbol) %>%
  group_by(sample, symbol) %>%
  mutate(alpha_id = paste0(sample, "read", 1:n())) %>% .[order(.$symbol),] %>% ungroup() %>% .[,c(3,5,6)] %>%
  pivot_wider(names_from = alpha_id, values_from = alpha)

nrow(single_cell)
ncol(single_cell)
single_cell[1:5,1:5]
```

* instead have alpha as the columns
* and the value is the number of reads with that alpha
* think of a region: has 4 cpgs
* alpha has limited options, of if read has 1,2,3,4 cpgs
* each sample: how many reads in each cat
* if imprinted: 20 reads, 10 0 10 1
* if random: in between
* samples columns, rows alpha

* probably binomial distribution
* library size link factor - how many reads in distrib
* need replicates
* do a test between conditions for a region - difference in distribution

* are we thinking of coming up with framework of differential alpha?
* likely would largely overlap with differential beta
* others that wouldn't: could be occuring when beta = 0.5 - beta doesn't have the information
* could be from different cell(types) or allele specific

* if dmr from beta = 1, beta = 0.5
* is alpha going to half 0 or about otherwise

* a little package exploring differnces in alpha and betas
* there are differences, and this is ways we can see them
* imprinted regions would be good examples
* if can get one region: how data should look and test - then can repeat
* may be able to use differential beta framework: proportion, instead of counts its the distrib
* 2 samples as columns, test to compare distributions
* ask a statistician



* putting a pause on this
* because there are NAs
* and in single cells they put 0s
* but we can't put a 0 here because there is no read
* so it's unclear if we can find a workaround for this

[belinda's analysis](https://bphipson.github.io/Human_Development_snRNAseq/14-MarkerAnalysisBCT.html)





```{r}
promoters(TxDb.Hsapiens.UCSC.hg19.knownGene::TxDb.Hsapiens.UCSC.hg19.knownGene)
```

```{r}
all_samples_genes_txdb <- find_overlaps(as_granges(all_samples_simple), promoters(TxDb.Hsapiens.UCSC.hg19.knownGene::TxDb.Hsapiens.UCSC.hg19.knownGene)) %>% data.frame()

```




