---
title: "diffAlphaUsageCompare"
author: "Caitlin Page"
date: "2026-03-20"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---



## Introduction

* Q: do the DTUs overlap the DMRs??

* except I did dmrs on everything, and did dtu on chr22
* and memory limit keeps expiring uh

```{r}
tpm_mat <- readRDS("../output/cluster_files/dtu_prop.rds")
ts.gene.gene <- readRDS("../output/cluster_files/dtu_res_gene.rds")
ts.gene.simes <- readRDS("../output/cluster_files/dtu_res_simes.rds")
```

```{r}
tpm_mat <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dtu_prop.rds")
ts.gene.gene <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dtu_res_gene.rds")
ts.gene.simes <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dtu_res_simes.rds")
```


```{r}
dmrcate_seq_dmr <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dmr_encode_gm12878_k562.rds")
#dmrcate_seq_dmr <- readRDS("../output/cluster_files/dmr_encode_gm12878_k562.rds")
```

```{r}
dtu.gene <- readRDS("/researchers/caitlin.page/cf_nano/r_output/ts.gene.gene.rds")
dtu.optiles <- readRDS("/researchers/caitlin.page/cf_nano/r_output/ts.gene_optiles.rds")
```

# compare dmrs and dtu_tiles
setdiff: ranges in first set that don't overlap to 2nd set
```{r}
saveRDS(all_samples, "/researchers/caitlin.page/cf_nano/r_output/all_rrbs_samples.rds")
```

* Q is: what alphas/betas are there in the dtu tiles not overlapping dmrs?
* compared to the alphas in the overlapping ones
* because dream is that betas are more 0.5 in the non-overlapping ones
* and then alpha is picking up more nuanced signal
```{r}
dtu.optiles <- dtu.optiles %>% filter(FDR < 0.05)
dmrcate_seq_dmr %>% tail()
dmrcate_seq_dmr$rank <- 1:nrow(dmrcate_seq_dmr)
```

```{r}
overlap <- find_overlaps(as_granges(dmrcate_seq_dmr), as_granges(dtu.optiles)) %>% data.frame()
```

```{r}
all_samples_simp
```

```{r}
all_samples_beta <- all_samples %>% distinct(seqnames, start, sample, beta)

all_samples_beta <- all_samples_beta %>% pivot_wider(names_from = sample, values_from = beta)
all_samples_beta$end <- all_samples_beta$start + 1
```

```{r}
overlap_beta <- find_overlaps(as_granges(overlap), as_granges(all_samples_beta)) %>% data.frame()
```

```{r}
overlap_beta %>% pivot_longer(cols = c(b1, b2, cml1, cml2), names_to = "sample", values_to = "beta") %>%
  ggplot(aes(x = beta, colour = sample)) +
  geom_density()

overlap_beta %>% pivot_longer(cols = c(b1, b2, cml1, cml2), names_to = "sample", values_to = "beta") %>%
  ggplot(aes(x = sample, y= beta, fill = sample)) +
  geom_boxplot() +
  theme(legend.position = "none")
```
```{r}
c(prop_tiles_overlap = overlap_beta %>% distinct(region_id) %>% nrow()/nrow(dtu.optiles), 
  prop_dmrs_overlap = overlap_beta %>% distinct(rank) %>% nrow()/nrow(dmrcate_seq_dmr))
```



* all the reads
```{r}
all_samples_beta %>% pivot_longer(cols = c(b1, b2, cml1, cml2), names_to = "sample", values_to = "beta") %>%
  ggplot(aes(x = beta, colour = sample)) +
  geom_density()

all_samples_beta %>% pivot_longer(cols = c(b1, b2, cml1, cml2), names_to = "sample", values_to = "beta") %>%
  ggplot(aes(x = sample, y= beta, fill = sample)) +
  geom_boxplot() +
  theme(legend.position = "none")
```

* all dtu
```{r}
find_overlaps(as_granges(dtu.optiles), as_granges(all_samples_beta)) %>% data.frame() %>%
  pivot_longer(cols = c(b1, b2, cml1, cml2), names_to = "sample", values_to = "beta") %>%
  ggplot(aes(x = beta, colour = sample)) +
  geom_density()

find_overlaps(as_granges(dtu.optiles), as_granges(all_samples_beta)) %>% data.frame() %>% 
  pivot_longer(cols = c(b1, b2, cml1, cml2), names_to = "sample", values_to = "beta") %>%
  ggplot(aes(x = sample, y= beta, fill = sample)) +
  geom_boxplot() +
  theme(legend.position = "none")
```

* only dtu (no overlap)
```{r}
find_overlaps(filter_by_non_overlaps(as_granges(dtu.optiles), as_granges(dmrcate_seq_dmr)), as_granges(all_samples_beta)) %>% data.frame() %>%
  pivot_longer(cols = c(b1, b2, cml1, cml2), names_to = "sample", values_to = "beta") %>%
  ggplot(aes(x = beta, colour = sample)) +
  geom_density()

find_overlaps(filter_by_non_overlaps(as_granges(dtu.optiles), as_granges(dmrcate_seq_dmr)), as_granges(all_samples_beta)) %>% data.frame() %>%
  pivot_longer(cols = c(b1, b2, cml1, cml2), names_to = "sample", values_to = "beta") %>%
  ggplot(aes(x = sample, y= beta, fill = sample)) +
  geom_boxplot() +
  theme(legend.position = "none")
```


* all dmr
```{r}
find_overlaps(as_granges(dmrcate_seq_dmr), as_granges(all_samples_beta)) %>% data.frame() %>%
  pivot_longer(cols = c(b1, b2, cml1, cml2), names_to = "sample", values_to = "beta") %>%
  ggplot(aes(x = beta, colour = sample)) +
  geom_density()

find_overlaps(as_granges(dmrcate_seq_dmr), as_granges(all_samples_beta)) %>% data.frame() %>% 
  pivot_longer(cols = c(b1, b2, cml1, cml2), names_to = "sample", values_to = "beta") %>%
  ggplot(aes(x = sample, y= beta, fill = sample)) +
  geom_boxplot() +
  theme(legend.position = "none")
```

* only in dmr

```{r}
find_overlaps(filter_by_non_overlaps(as_granges(dmrcate_seq_dmr), as_granges(dtu.optiles)), as_granges(all_samples_beta)) %>% data.frame() %>%
  pivot_longer(cols = c(b1, b2, cml1, cml2), names_to = "sample", values_to = "beta") %>%
  ggplot(aes(x = beta, colour = sample)) +
  geom_density()

find_overlaps(filter_by_non_overlaps(as_granges(dmrcate_seq_dmr), as_granges(dtu.optiles)), as_granges(all_samples_beta)) %>% data.frame() %>%
  pivot_longer(cols = c(b1, b2, cml1, cml2), names_to = "sample", values_to = "beta") %>%
  ggplot(aes(x = sample, y= beta, fill = sample)) +
  geom_boxplot() +
  theme(legend.position = "none")
```

# thoughts
* not sure if this is showing what I want
* think need to plot some of these regions


* dtu region overlapping dmr
```{r}
overlap_beta %>% .[order(.$dtu_rank),] %>% distinct(region_id, dtu_rank)
```

```{r}
find_overlaps(as_granges(all_samples_simp), as_granges(dtu.optiles[1,])) %>% data.frame()
```

```{r}
source("../code/processData.R")
reads <- find_overlaps(as_granges(all_samples_simp), as_granges(dtu.optiles[1,])) %>% data.frame() %>% distinct(read_id) %>% .$read_id

rrbs_plot4(all_samples, reads, cpgs = cg_sites, jit_amount = 0.05) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(x = "coordinate")
```

```{r}
source("../code/processData.R")
reads <- find_overlaps(as_granges(all_samples_simp), as_granges(dtu.optiles[1,])) %>% data.frame() %>% distinct(read_id) %>% .$read_id

rrbs_plot4(all_samples, reads, cpgs = cg_sites, jit_amount = 0.05) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(x = "coordinate")
```


####### old

```{r}
dmrcate_seq_dmr
```

```{r}
ts.gene.simes %>% group_by(gene_id) %>% summarise(n=n()) %>% filter(n>1)
ts.gene.gene %>% group_by(gene_id) %>% summarise(n=n()) %>% filter(n>1)
```

* need gene positions/read positions


```{r}
colnames(dtu_for_beta)[3] <- "gene_id"
colnames(dtu_for_beta)[4] <- "symbol"
```

```{r}
dtu_all <- rbind(dtu_for_beta %>% mutate(sample = "b1"),
      dtu_beta_2 %>% mutate(sample = "b2"),
      dtu_k562 %>% mutate(sample = "cml1"),
      dtu_k562_2 %>% mutate(sample = "cml2"))
```
```{r}
ts.gene.simes %>% filter(FDR < 0.05)
```

```{r}
dtu_all <- dtu_all %>% filter(symbol %in% filter(ts.gene.simes, FDR < 0.05)$gene_id) %>% mutate(seqnames = "chr22", start = read_start, end = read_end)
```
```{r}
dmrcate_seq_dmr$rank <- 1:nrow(dmrcate_seq_dmr)
```

```{r}
overlap <- find_overlaps(as_granges(dmrcate_seq_dmr), as_granges(dtu_all)) %>% data.frame()
```

```{r}
ts.gene.simes %>% filter(FDR < 0.05) %>% mutate(rank = 1:n(), overlap_dmr = ifelse(gene_id %in% overlap$symbol, TRUE, FALSE)) %>% group_by(overlap_dmr) %>% summarise(n=n())
```

```{r}
dmrcate_seq_dmr %>% filter(seqnames == "chr22") %>% mutate(overlap_dtu = ifelse(rank %in% overlap$rank, TRUE, FALSE)) %>% group_by(overlap_dtu) %>% summarise(n=n())
```
```{r}
dmrcate_seq_dmr %>% filter(seqnames == "chr22")
```

* ok so not all overlap
* but would more overlap if I did it right?

```{r}
ts.gene.simes %>% filter(FDR < 0.05) %>% mutate(rank = 1:n(), overlap_dmr = ifelse(gene_id %in% overlap$symbol, TRUE, FALSE)) %>%
  ggplot(aes(x = rank, colour = overlap_dmr)) +
  geom_density()

ts.gene.simes %>% filter(FDR < 0.05) %>% mutate(rank = 1:n(), overlap_dmr = ifelse(gene_id %in% overlap$symbol, TRUE, FALSE)) %>%
  ggplot(aes(x = rank, fill = overlap_dmr)) +
  geom_histogram()

```

```{r}
dmrcate_seq_dmr %>% mutate(overlap_dtu = ifelse(rank %in% overlap$rank, TRUE, FALSE)) %>%
  ggplot(aes(x = rank, colour = overlap_dtu)) +
  geom_density()

dmrcate_seq_dmr %>% mutate(overlap_dtu = ifelse(rank %in% overlap$rank, TRUE, FALSE)) %>%
  ggplot(aes(x = rank, fill = overlap_dtu)) +
  geom_histogram()
```


# what are the genes that didn't overlap?
```{r}
ts.gene.simes %>% filter(FDR < 0.05) %>% mutate(rank = 1:n(), overlap_dmr = ifelse(gene_id %in% overlap$symbol, TRUE, FALSE)) %>% filter(overlap_dmr == FALSE)
```

```{r}
tpm_mat
```

####

```{r}
all
```


```{r}
source("../code/processData.R")
reads <- all_samples_simp %>% filter(seqnames == "chr22", start >= 10571810, start <= 10571840) %>% distinct(read_id) %>% .$read_id

rrbs_plot4(all_samples, reads, cpgs = cg_sites, jit_amount = 0.05) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(x = "coordinate")
```

```{r}
source("../code/processData.R")
reads <- rrbs_alpha_beta %>% filter(seqnames == "chr22", start >= 10571810, start <= 10571840) %>% distinct(read_id) %>% .$read_id

rrbs_plot4(mutate(rrbs_alpha_beta, sample = "b1"), reads, cpgs = cg_sites, jit_amount = 0.05) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(x = "coordinate")
```

```{r}
overlap[1,] # top dmr - overlaps multiple dtu regions
overlap %>% filter(rank == 1)
```


```{r}
source("../code/processData.R")
reads <- all_samples_simp %>% filter(seqnames == "chr10", start >= 132784117, start <= 132788846) %>% distinct(read_id) %>% .$read_id

rrbs_plot4(all_samples, reads, cpgs = cg_sites, jit_amount = 0.05) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(x = "coordinate")

all_samples_simp %>% filter(read_id %in% reads) %>%
  ggplot(aes(x = alpha, colour = sample)) +
  geom_density()
```

```{r}
find_overlaps(genes_txdb, as_granges(overlap[1,])) %>% data.frame()
```


#

```{r}
only_in_dtu <- find_overlaps(filter_by_non_overlaps(as_granges(dtu.optiles), as_granges(dmrcate_seq_dmr)), as_granges(all_samples_simp)) %>% data.frame()
only_in_dtu %>% distinct(dtu_rank) %>% .[order(.$dtu_rank),]
```

```{r}
source("../code/processData.R")
reads <- only_in_dtu %>% filter(dtu_rank == 17) %>% distinct(read_id) %>% .$read_id

rrbs_plot4(all_samples, reads, cpgs = cg_sites, jit_amount = 0.05) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(x = "coordinate")
```

```{r}
all_samples_simp %>% filter(read_id %in% reads) %>%
  ggplot(aes(x = alpha, colour = sample)) +
  geom_density()
```


```{r}
find_overlaps(genes_txdb, as_granges(only_in_dtu[1,])) %>% data.frame()
```

# top dtu

```{r}
source("../code/processData.R")
reads <- only_in_dtu %>% filter(dtu_rank == 17) %>% distinct(read_id) %>% .$read_id

rrbs_plot4(all_samples, reads, cpgs = cg_sites, jit_amount = 0.05) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(x = "coordinate")
```

```{r}
all_samples_simp %>% filter(read_id %in% reads) %>%
  ggplot(aes(x = alpha, colour = sample)) +
  geom_density()
```


```{r}
only_in_dmr <- find_overlaps(filter_by_non_overlaps(as_granges(dmrcate_seq_dmr), as_granges(dtu.optiles)), as_granges(all_samples_simp)) %>% data.frame()
only_in_dmr %>% distinct(dtu_rank) %>% .[order(.$dtu_rank),]
```


