---
title: "diffAlphaUsageDiffMeth"
author: "caitlinpage"
date: "2026-02-25"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

## Introduction

Continuation of developing [differential alpha usage](differentialAlphaUsage.html)
Conduct differential methylation analysis and conduct DTU on those regions
Compare back to others


```{r}
library(bsseq)
library(DMRcate)
library(annotatr)
library(data.table)
library(plyranges)
library(tidyr)
library(stringr)
library(dplyr)
library(ggplot2)
library(edgeR)
```


## Data Objects
```{r}
cg_sites <- cbind(data.frame(Biostrings::matchPattern("CG", BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38[["chr22"]])), seqnames = "chr22") %>% 
  mutate(pos = paste0(seqnames, "-", start)) %>%
  relocate(pos, seqnames)
cg_sites$index <- 1:nrow(cg_sites)
```

* GM1268 cell line

```{r}
source("../code/processData.R")
rrbs_alpha_beta <- readRDS("/researchers/caitlin.page/cf_nano/r_output/rrbs_alpha_beta.rds")
```

* REP 2 GM12878 cell line
```{r}
gm12878_2 <- readRDS("/researchers/caitlin.page/cf_nano/r_output/rrbs_alpha_beta_rep2.rds")

```

* K562 cml cell line
```{r}
source("../code/processData.R")
k562 <- readRDS("/researchers/caitlin.page/cf_nano/r_output/k562_rep1.rds")
k562 <- k562 %>% as_granges() %>% data.frame()
```

* REP 2 K562
```{r}
source("../code/processData.R")
k562_2 <- readRDS("/researchers/caitlin.page/cf_nano/r_output/k562_rep2.rds")
```





# Differential Methylation

* b 1 and 2.  (gm12878)
* cml 1 and 2 (k562)


```{r}
b_1 <- rrbs_alpha_beta %>% 
  group_by(seqnames, start) %>%
  mutate(cov = n()) %>%
  filter(meth_status == "M") %>%
mutate(meth = n()) %>%
    ungroup() %>% distinct(seqnames, start, meth, cov) %>% .[order(.$seqnames),] %>% .[order(.$start),]
b_2 <- gm12878_2 %>% 
  group_by(seqnames, start) %>%
  mutate(cov = n()) %>%
  filter(meth_status == "M") %>%
mutate(meth = n()) %>%
    ungroup() %>% distinct(seqnames, start, meth, cov) %>% .[order(.$seqnames),] %>% .[order(.$start),]

cml_1 <- k562 %>% 
  group_by(seqnames, start) %>%
  mutate(cov = n()) %>%
  filter(meth_status == "M") %>%
mutate(meth = n()) %>%
    ungroup() %>% distinct(seqnames, start, meth, cov) %>% .[order(.$seqnames),] %>% .[order(.$start),]

cml_2 <- k562_2 %>% 
  group_by(seqnames, start) %>%
  mutate(cov = n()) %>%
  filter(meth_status == "M") %>%
mutate(meth = n()) %>%
    ungroup() %>% distinct(seqnames, start, meth, cov) %>% .[order(.$seqnames),] %>% .[order(.$start),]
```

```{r}
b_1 <- b_1 %>% mutate(pos = paste0(seqnames, "-", start)) %>% data.frame()
b_2 <- b_2 %>% mutate(pos = paste0(seqnames, "-", start)) %>% data.frame()
cml_1 <- cml_1 %>% mutate(pos = paste0(seqnames, "-", start)) %>% data.frame()
cml_2 <- cml_2 %>% mutate(pos = paste0(seqnames, "-", start)) %>% data.frame()
```

```{r}
seqs <- data.frame(s = 1:22) %>% mutate(seqs = paste0("chr", s))
b_1 <- b_1 %>% filter(seqnames %in% seqs$seqs)
b_2 <- b_2 %>% filter(seqnames %in% seqs$seqs)
cml_1 <- cml_1 %>% filter(seqnames %in% seqs$seqs)
cml_2 <- cml_2 %>% filter(seqnames %in% seqs$seqs)
```


```{r}
dm_table <- rbind(b_1[,1:2], b_2[,1:2,], cml_1[,1:2], cml_2[,1:2]) %>%
  distinct() %>%
  mutate(pos = paste0(seqnames, "-", start))
dm_meth <- dm_table %>%
  mutate(b1 = b_1[match(.$pos, b_1$pos), "meth"],
         b2 = b_2[match(.$pos, b_2$pos), "meth"],
         cml1 = cml_1[match(.$pos, cml_1$pos), "meth"],
         cml2 = cml_2[match(.$pos, cml_2$pos), "meth"])
dm_cov <- dm_table %>%
  mutate(b1 = b_1[match(.$pos, b_1$pos), "cov"],
         b2 = b_2[match(.$pos, b_2$pos), "cov"],
         cml1 = cml_1[match(.$pos, cml_1$pos), "cov"],
         cml2 = cml_2[match(.$pos, cml_2$pos), "cov"])

dm_meth[is.na(dm_meth)] <- 0
dm_cov[is.na(dm_cov)] <- 0

dm_meth <- dm_meth %>% .[order(.$seqnames),]
dm_cov <- dm_cov %>% .[order(.$seqnames),]
```


```{r}
#library(bsseq)
```

```{r}
sample_stuff <- c(Type = "b", "b", "cml", "cml") %>% data.frame()
rownames(sample_stuff) <- colnames(dm_meth)[4:7]

## make BSseq object
bseq <- bsseq::BSseq(M = as.matrix(dm_meth[,4:7]),
                             Cov = as.matrix(dm_cov[,4:7]), pData = sample_stuff,
                             chr = dm_meth$seqnames, pos = as.numeric(dm_meth$start),
                             sampleNames = colnames(dm_meth)[4:7])
bseq <- sort(bseq)
```


```{r}
saveRDS(bseq, "/researchers/caitlin.page/cf_nano/r_output/bseq.rds")
```



## BSmooth by sample - so it doesn't crash



```{r}
bsmooth1 <- BSmooth(BSseq = bseq[,1], BPPARAM = BiocParallel::MulticoreParam(workers = 4), verbose = TRUE)

bsmooth2 <- BSmooth(BSseq = bseq[,2], BPPARAM = BiocParallel::MulticoreParam(workers = 4), verbose = TRUE)
bsmooth3 <- BSmooth(BSseq = bseq[,3], BPPARAM = BiocParallel::MulticoreParam(workers = 4), verbose = TRUE)
bsmooth4 <- BSmooth(BSseq = bseq[,4], BPPARAM = BiocParallel::MulticoreParam(workers = 4), verbose = TRUE)
```


```{r}
bsmooth_obj <- vctrs::vec_c(bsmooth1, bsmooth2, bsmooth3, bsmooth4)
```
```{r}
vctrs::vec_c
```

bsmooth_obj <- combine(bsmooth1, bsmooth2, bsmooth3, bsmooth4)
```{r}
list_bs <- list(bsmooth1, bsmooth2, bsmooth3, bsmooth4)
bsmooth_obj <- combineList(list_bs)
```

```{r}
bsmooth_obj <- bsseq::combine(bsmooth1, bsmooth2, bsmooth3, bsmooth4)
```

```{r}
saveRDS(bsmooth1, "/researchers/caitlin.page/cf_nano/r_output/bsmooth_1.rds")

saveRDS(bsmooth2, "/researchers/caitlin.page/cf_nano/r_output/bsmooth_2.rds")
saveRDS(bsmooth3, "/researchers/caitlin.page/cf_nano/r_output/bsmooth_3.rds")
saveRDS(bsmooth4, "/researchers/caitlin.page/cf_nano/r_output/bsmooth_4.rds")
saveRDS(rrbs_alpha_beta, "/researchers/caitlin.page/cf_nano/r_output/rrbs_alpha_beta.rds")
```

```{r}
bseq@assays@data$M
```

```{r}
bsmooth_obj <- BSmooth(BSseq = bseq, BPPARAM = BiocParallel::MulticoreParam(workers = 4), verbose = TRUE)
```
```{r}
bsmooth_obj
```

```{r}
bsmooth_obj$.
```

```{r}
## filter
bseq.cov <- getCoverage(bsmooth_obj)
keep <- which(rowSums(bseq.cov[, bsmooth_obj$. == "b"] >= 2) >= 2 &
                rowSums(bseq.cov[, bsmooth_obj$. == "cml"] >= 2) >= 2)
```
```{r}
length(keep)
```

```{r}
colnames(bsmooth_obj)
pData(bsmooth_obj) <- data.frame(replicate = c("1", "2", "1", "2"), 
                                 tissue = c("b", "b", "cml", "cml"))
                                 row.names(colnames(bsmooth_obj)))
```
```{r}
data.frame(replicate = c("1", "2", "1", "2"), 
                                 tissue = c("b", "b", "cml", "cml"))
```

```{r}
#bsmooth_obj <- bsmooth_obj[keep,]
saveRDS(bsmooth_obj, "../researchers/caitlin.page/cf_nano/r_output/bsmooth_obj_encode.rds")

pData(bsmooth_obj) <- data.frame(replicate = c("1", "2", "1", "2"), 
                                 tissue = c("b", "b", "cml", "cml"), row.names = colnames(bsmooth_obj))

colData(bsmooth_obj)

tissue <- factor(pData(bsmooth_obj)$tissue)
tissue
#Regular matrix design
design <- model.matrix(~tissue)
rownames(design) <- colnames(bsmooth_obj)
design

methdesign <- modelMatrixMeth(design)
methdesign
```

```{r}
bsmooth_obj
```

* this does some filtering


```{r}
seq_annot <- sequencing.annotate(bsmooth_obj, methdesign, all.cov = TRUE,
                                 coef = 6, fdr=0.05)

seq_dmrcate.res <- dmrcate(seq_annot, C=2, min.cpgs = 5)

dmrcate_seq_dmr <- extractRanges(seq_dmrcate.res, genome = "hg19")
dmrcate_seq_dmr <- dmrcate_seq_dmr %>% data.frame()
dmrcate_seq_anno <- seq_annot@ranges %>% data.frame()
```


```{r}
saveRDS(dmrcate_seq_dmr, "/researchers/caitlin.page/cf_nano/r_output/dmr_encode_gm12878_k562.rds")
saveRDS(dmrcate_seq_anno, "/researchers/caitlin.page/cf_nano/r_output/dm_encode_gm12878_k562.rds")
```


####

## DMR analysis

```{r}
dmrcate_seq_dmr
```

```{r}
k562 <- k562[,colnames(rrbs_alpha_beta)]
k562_2 <- k562_2[,colnames(rrbs_alpha_beta)]
```


```{r}
all_samples <- rbind(rrbs_alpha_beta %>% mutate(sample = "b1"), gm12878_2 %>% mutate(sample = "b2"), 
                     k562 %>% mutate(sample = "cml1"), k562_2 %>% mutate(sample = "cml2"))
```

```{r}
rrbs_alpha_beta
```

```{r}
find_overlaps(as_granges(all_samples), as_granges(dmrcate_seq_dmr[1,])) %>% data.frame()
```

```{r}
find_overlaps(as_granges(all_samples), as_granges(dmrcate_seq_dmr[1,])) %>% data.frame()
```




```{r}
find_overlaps(as_granges(all_samples), as_granges(dmrcate_seq_dmr[1,])) %>% data.frame() %>%
  distinct(sample, read_id, read_start, read_end, alpha) %>%
  ggplot() +
  geom_segment(aes(x = read_start, xend = read_end, y = alpha, colour = read_id)) +
  theme(legend.position = "none") +
  facet_wrap(~ sample)
```

```{r}
find_overlaps(as_granges(all_samples), as_granges(dmrcate_seq_dmr[1,])) %>% data.frame() %>%
  group_by(sample, read_id, read_start, read_end, alpha) %>%
  summarise(count = n()) %>% ungroup() %>% filter(count > 1)
```


```{r}
find_overlaps(as_granges(all_samples), as_granges(dmrcate_seq_dmr[1,])) %>% data.frame() %>%
  group_by(sample, read_id, read_start, read_end, alpha) %>%
  summarise(count = n()) %>% ungroup() %>%
  ggplot() +
  geom_segment(aes(x = read_start, xend = read_end, y = alpha, colour = read_id, linewidth = count)) +
  theme(legend.position = "none") +
  facet_wrap(~ sample)
```


```{r}
find_overlaps(as_granges(all_samples), as_granges(dmrcate_seq_dmr[2,])) %>% data.frame() %>%
  group_by(sample, read_id, read_start, read_end, alpha) %>%
  summarise(count = n()) %>% ungroup() %>%
  ggplot() +
  geom_segment(aes(x = read_start, xend = read_end, y = alpha, colour = read_id, linewidth = count)) +
  theme(legend.position = "none") +
  facet_wrap(~ sample)
```



rrbs_plot <- function(long_data, read_ids, cpgs = cg_sites, count_instead = FALSE) {
  if(count_instead == FALSE) {
  plot <- long_data %>% dplyr::filter(read_id %in% read_ids) %>%
    distinct(read_id, read_start, read_end, alpha) %>%
    ggplot() +
    geom_segment(aes(x = read_start, xend = read_end, y = alpha, yend = alpha), alpha = 0.2) +
    geom_point(data = dplyr::filter(long_data, read_id %in% read_ids),
               aes(x = start, y = alpha, colour = meth_status)) +
    geom_point(data = distinct(dplyr::filter(long_data, read_id %in% read_ids), start, beta),
               aes(x = start, y = beta), shape = 4, size = 3) +
    geom_segment(data = filter(cg_sites, start >= min(dplyr::filter(long_data, read_id %in% read_ids)$start),
                               end <= max(dplyr::filter(long_data, read_id %in% read_ids)$end)),
                 aes(x = start, xend = start, y = -0.15, yend = -0.05), colour = "green")
  } else {
    plot <- long_data %>% dplyr::filter(read_id %in% read_ids) %>%
      distinct(read_id, read_start, read_end, alpha) %>%
      ggplot() +
      geom_segment(aes(x = read_start, xend = read_end, y = alpha, yend = alpha), alpha = 0.2) +
      geom_count(data = dplyr::filter(long_data, read_id %in% read_ids),
                 aes(x = start, y = alpha, colour = meth_status)) +
      geom_point(data = distinct(dplyr::filter(long_data, read_id %in% read_ids), start, beta),
                 aes(x = start, y = beta), shape = 4, size = 3) +
      geom_segment(data = filter(cg_sites, start >= min(dplyr::filter(long_data, read_id %in% read_ids)$start),
                                 end <= max(dplyr::filter(long_data, read_id %in% read_ids)$end)),
                   aes(x = start, xend = start, y = -0.15, yend = -0.05), colour = "green")
  }
  plot
}
