---
title: "dmrVsDAU"
author: "caitlinpage"
date: "2026-09-02"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r}
library(plyranges)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
```


## compare the tile positions
```{r}
dau_basictiles <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dtu_basictiles_emseq_cancer_healthy.rds")

dau_optiles <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_emseq_cancer_healthy.rds")

dmr_emseq <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dmr_emseq_strict.rds")
dmr_emseq <- dmr_emseq %>% .[order(.$min_smoothed_fdr),] %>% mutate(dmr_rank = 1:n())
```

```{r}
dau_basictiles[1:5,]
```


```{r}
dmr_emseq[1:5,]
```

# ppgl validity analysis (claude)
```{r}
library(GenomicRanges)
library(annotatr)
library(regioneR)
library(AnnotationHub)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(org.Hs.eg.db)
library(missMethyl)
library(TCGAbiolinks)
library(SummarizedExperiment)
library(BiocParallel)
library(ggplot2)
library(dplyr)
library(patchwork)
```

```{r}
# ============================================================
# DMR Validity Analysis — Normal vs Cell-Free PPGL
# ============================================================
# Sections:
#   0. Packages
#   1. Load DMRs
#   2. Functional annotation (annotatr)
#   3. Chromatin state enrichment (regioneR + Roadmap Epigenomics)
#   4. TCGA PCPG replication
#   5. Pathway enrichment (missMethyl)
# ============================================================
```

## DMR

```{r}
# Set genome build — change to "hg19" if your data is GRCh37
GENOME <- "hg38"

# ── 1. Load DMRs ────────────────────────────────────────────

dmrs_gr <- makeGRangesFromDataFrame(
  dmr_emseq %>%
    mutate(
      direction = ifelse(meandiff > 0, "hyper", "hypo"),
      dmr_id    = paste0("DMR_", dmr_rank)
    ),
  seqnames.field     = "seqnames",
  start.field        = "start",
  end.field          = "end",
  strand.field       = "strand",
  keep.extra.columns = TRUE
)
seqlevelsStyle(dmrs_gr) <- "UCSC"
# Separate hyper and hypo DMRs for direction-aware analyses
dmrs_hyper <- dmrs_gr[dmrs_gr$direction == "hyper"]
dmrs_hypo  <- dmrs_gr[dmrs_gr$direction == "hypo"]

cat("Total DMRs:", length(dmrs_gr), "\n")
cat("  Hypermethylated:", length(dmrs_hyper), "\n")
cat("  Hypomethylated: ", length(dmrs_hypo), "\n")


# ── 2. Functional Annotation (annotatr) ─────────────────────
# Annotates DMRs to CpG features (islands/shores/shelves/open sea)
# and genic features (promoters/exons/introns/intergenic).
# For PPGL, enrichment at CGI promoters supports the known
# SDHx-driven hypermethylator phenotype.

annot_types <- c(
  # CpG features
  paste0(GENOME, "_cpgs"),
  # Genic features
  paste0(GENOME, "_genes_promoters"),
  paste0(GENOME, "_genes_5UTRs"),
  paste0(GENOME, "_genes_exons"),
  paste0(GENOME, "_genes_introns"),
  paste0(GENOME, "_genes_3UTRs"),
  paste0(GENOME, "_genes_intergenic")
)

annotations <- build_annotations(genome = GENOME, annotations = annot_types)

dmr_annotated <- annotate_regions(
  regions    = dmrs_gr,
  annotations = annotations,
  ignore.strand = TRUE,
  quiet = FALSE
)

dmr_annotated_df <- as.data.frame(dmr_annotated)
```


```{r}
## 2a. CpG context summary
dmr_annotated_df <- dmr_annotated_df %>%
  mutate(across(everything(), function(x) {
    if (is.list(x) || inherits(x, "AsIs"))
      sapply(x, function(v) if (length(v) == 0) NA_character_ else paste(as.character(v), collapse = ","))
    else x
  }))
cpg_summary <- dmr_annotated_df %>%
  mutate(
    annot.type = as.character(annot.type),
    direction  = as.character(direction)
  ) %>%
  filter(grepl("cpg", annot.type)) %>%
  mutate(
    cpg_context = case_when(
      grepl("islands",  annot.type) ~ "CpG Island",
      grepl("shores",   annot.type) ~ "Shore",
      grepl("shelves",  annot.type) ~ "Shelf",
      grepl("inter",    annot.type) ~ "Open Sea"
    )
  ) %>%
  dplyr::count(cpg_context, direction) %>%
  group_by(direction) %>%
  mutate(pct = n / sum(n) * 100)

p_cpg <- ggplot(cpg_summary, aes(x = cpg_context, y = pct, fill = direction)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(hyper = "#d62728", hypo = "#1f77b4")) +
  labs(
    title = "DMR distribution across CpG features",
    subtitle = "Hyper = hypermethylated in PPGL cfDNA vs normal",
    x = "CpG context", y = "% of DMRs", fill = "Direction"
  ) +
  theme_bw(base_size = 12)

## 2b. Genic context summary
gene_summary <- dmr_annotated_df %>%
  filter(grepl("genes", annot.type)) %>%
  mutate(
    gene_context = case_when(
      grepl("promoter",    annot.type) ~ "Promoter",
      grepl("5UTR",        annot.type) ~ "5'UTR",
      grepl("exon",        annot.type) ~ "Exon",
      grepl("intron",      annot.type) ~ "Intron",
      grepl("3UTR",        annot.type) ~ "3'UTR",
      grepl("intergenic",  annot.type) ~ "Intergenic"
    ),
    direction = direction
  ) %>%
  dplyr::count(gene_context, direction) %>%
  group_by(direction) %>%
  mutate(pct = n / sum(n) * 100)

p_gene <- ggplot(gene_summary, aes(x = gene_context, y = pct, fill = direction)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(hyper = "#d62728", hypo = "#1f77b4")) +
  labs(title = "DMR distribution across genic features",
       x = "Genic context", y = "% of DMRs", fill = "Direction") +
  theme_bw(base_size = 12)

print(p_cpg / p_gene)

## 2c. Meandiff distribution & size vs effect
p_meandiff <- ggplot(as.data.frame(dmrs_gr), aes(x = meandiff, fill = direction)) +
  geom_histogram(bins = 50, colour = "white", alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c(hyper = "#d62728", hypo = "#1f77b4")) +
  labs(title = "Distribution of meandiff across DMRs",
       subtitle = "Positive = hypermethylated in PPGL cfDNA vs normal",
       x = "meandiff", y = "Count", fill = "Direction") +
  theme_bw(base_size = 12)

p_size_effect <- ggplot(as.data.frame(dmrs_gr),
                        aes(x = no.cpgs, y = abs(meandiff), colour = direction)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_colour_manual(values = c(hyper = "#d62728", hypo = "#1f77b4")) +
  scale_x_log10() +
  labs(title = "DMR size vs effect magnitude",
       x = "Number of CpGs (log scale)", y = "|meandiff|", colour = "Direction") +
  theme_bw(base_size = 12)

print(p_meandiff | p_size_effect)

## 2c. Number/proportion of DMRs at CGI promoters (PPGL key check)
cgi_promoter_dmrs <- dmr_annotated_df %>%
  filter(grepl("island", annot.type) | grepl("promoter", annot.type)) %>%
  pull(annot.symbol) %>%
  unique()

cat("\nDMRs overlapping CGI or promoter regions:",
    length(unique(dmr_annotated_df$annot.name[
      grepl("island|promoter", dmr_annotated_df$annot.type)])), "\n")
```


```{r}
# ── 3. Chromatin State Enrichment (regioneR + Roadmap) ──────
# Tests whether DMRs are enriched in regulatory regions of
# PPGL-relevant cell types vs random expectation.
# E080 = fetal adrenal gland (best available proxy for chromaffin cells)
# E053/E054 = neural crest-derived neurospheres
# E029 = monocytes (as a negative control — should NOT be enriched)

ah <- AnnotationHub()

## Helper to fetch Roadmap 15-state chromatin state BED for one sample
fetch_roadmap_states <- function(ah, eid, active_states = c("1_TssA", "2_TssAFlnk",
                                                             "6_EnhG", "7_Enh")) {
  # Query AnnotationHub for the chromatin state segmentation
  q <- query(ah, c("EpigenomeRoadMap", eid, "15_coreMarks", "bed"))
  if (length(q) == 0) {
    message("No Roadmap data found for ", eid)
    return(NULL)
  }
  states_gr <- ah[[names(q)[1]]]
  # Keep only active regulatory states
  active_gr <- states_gr[states_gr$name %in% active_states]
  seqlevelsStyle(active_gr) <- "UCSC"
  active_gr
}

## Fetch reference epigenomes
cat("Fetching Roadmap chromatin states (this caches locally after first run)...\n")
roadmap_refs <- list(
  fetal_adrenal  = fetch_roadmap_states(ah, "E080"),  # fetal adrenal gland
  neurosphere_dr = fetch_roadmap_states(ah, "E053"),  # neural crest-derived
  neurosphere_sp = fetch_roadmap_states(ah, "E054"),  # neural crest-derived
  monocyte       = fetch_roadmap_states(ah, "E029")   # negative control
)
roadmap_refs <- Filter(Negate(is.null), roadmap_refs)

## Genome mask for regioneR permutations (exclude gaps/unassembled regions)
# Using a simplified mask — replace with BSgenome mask for production runs
genome_mask <- filterChromosomes(getGenome(GENOME))

library(rtracklayer)

# Download the hg19 → hg38 chain file (once)
chain_file <- "hg19ToHg38.over.chain"
if (!file.exists(chain_file)) {
  download.file(
    "https://hgdownload.soe.ucsc.edu/goldenPath/hg19/liftOver/hg19ToHg38.over.chain.gz",
    paste0(chain_file, ".gz")
  )
  R.utils::gunzip(paste0(chain_file, ".gz"))  # or: system(paste("gunzip", paste0(chain_file, ".gz")))
}
chain <- import.chain(chain_file)

# LiftOver all Roadmap references from hg19 → hg38
roadmap_refs_hg38 <- lapply(roadmap_refs, function(gr) {
  gr_lifted <- unlist(liftOver(gr, chain))
  seqlevelsStyle(gr_lifted) <- "UCSC"
  gr_lifted
})

# Verify the liftover worked
sapply(roadmap_refs_hg38, length)

## Run permutation test for each reference
set.seed(123)
perm_results <- lapply(names(roadmap_refs), function(ref_name) {
  cat("  Testing enrichment in:", ref_name, "\n")
  tryCatch({
    pt <- permTest(
      A                  = dmrs_gr,
      B                  = roadmap_refs_hg38[[ref_name]],
      genome             = GENOME,           # "hg38" or "hg19" string
      randomize.function = randomizeRegions, # no universe required
      evaluate.function  = numOverlaps,
      ntimes             = 1000,
      mc.set.seed        = TRUE,
      mc.cores           = 1
    )
    list(
      ref      = ref_name,
      observed = pt$numOverlaps$observed,
      expected = mean(pt$numOverlaps$permuted),
      pval     = pt$numOverlaps$pval,
      zscore   = pt$numOverlaps$zscore,
      result   = pt
    )
  }, error = function(e) {
    message("  Failed for ", ref_name, ": ", conditionMessage(e))
    NULL
  })
})
perm_results <- Filter(Negate(is.null), perm_results)
names(perm_results) <- sapply(perm_results, `[[`, "ref")

## Summarise enrichment
enrich_df <- bind_rows(lapply(perm_results, function(x) {
  data.frame(
    reference = x$ref,
    observed  = x$observed,
    expected  = round(x$expected, 1),
    fold_enrich = round(x$observed / x$expected, 2),
    zscore    = round(x$zscore, 2),
    pval      = x$pval
  )
}))

cat("\n── Chromatin State Enrichment Summary ──\n")
print(enrich_df)

p_enrich <- ggplot(enrich_df,
                   aes(x = reorder(reference, fold_enrich),
                       y = fold_enrich,
                       fill = pval < 0.05)) +
  geom_col() +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey40") +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#2ca02c", "FALSE" = "#aec7e8"),
                    labels = c("TRUE" = "p < 0.05", "FALSE" = "n.s.")) +
  labs(title = "DMR enrichment in active regulatory regions",
       subtitle = "Roadmap Epigenomics chromatin states (active TSS + enhancers)",
       x = NULL, y = "Fold enrichment over random", fill = "") +
  theme_bw(base_size = 12)

print(p_enrich)
```


```{r}
# ── 4. TCGA PCPG Replication ────────────────────────────────
# Downloads TCGA pheochromocytoma/paraganglioma 450K methylation data
# and checks whether CpGs within your DMRs are differentially
# methylated in TCGA tissue data — tissue replication of cfDNA signal.
# Note: TCGA PCPG uses hg38; adjust liftOver if your DMRs are hg19.

cat("\nDownloading TCGA PCPG 450K methylation data...\n")

tcga_query <- GDCquery(
  project            = "TCGA-PCPG",
  data.category      = "DNA Methylation",
  data.type          = "Methylation Beta Value",
  platform           = "Illumina Human Methylation 450"
)

GDCdownload(tcga_query, method = "api", files.per.chunk = 10)

tcga_data <- GDCprepare(tcga_query)   # SummarizedExperiment

# Extract beta matrix and sample metadata
beta_mat  <- assay(tcga_data)         # rows = CpGs, cols = samples
col_data  <- as.data.frame(colData(tcga_data))
row_data  <- as.data.frame(rowRanges(tcga_data))

# TCGA PCPG includes both tumour and normal tissue
# Classify samples
col_data$sample_type_clean <- ifelse(
  grepl("Normal", col_data$sample_type, ignore.case = TRUE),
  "Normal", "Tumour"
)

cat("TCGA PCPG samples:\n")
print(table(col_data$sample_type_clean))

# Build GRanges for TCGA CpGs
tcga_cpg_gr <- makeGRangesFromDataFrame(
  row_data,
  seqnames.field = "seqnames",
  start.field    = "start",
  end.field      = "end",
  keep.extra.columns = FALSE
)
names(tcga_cpg_gr) <- rownames(row_data)
seqlevelsStyle(tcga_cpg_gr) <- "UCSC"

## Find CpGs that fall within your DMRs
hits <- findOverlaps(tcga_cpg_gr, dmrs_gr)
cpgs_in_dmrs <- names(tcga_cpg_gr)[queryHits(hits)]

cat("\nTCGA CpGs overlapping your DMRs:", length(unique(cpgs_in_dmrs)), "\n")

## Compare mean methylation for CpGs inside vs outside DMRs
tumour_idx <- which(col_data$sample_type_clean == "Tumour")
normal_idx <- which(col_data$sample_type_clean == "Normal")

# Mean beta per CpG in tumour vs normal
mean_tumour <- rowMeans(beta_mat[, tumour_idx], na.rm = TRUE)
mean_normal <- rowMeans(beta_mat[, normal_idx], na.rm = TRUE)
delta_beta   <- mean_tumour - mean_normal

## Summarise delta beta for DMR-overlapping CpGs
dmr_cpg_delta  <- delta_beta[cpgs_in_dmrs]
bg_cpg_delta   <- delta_beta[setdiff(names(delta_beta), cpgs_in_dmrs)]

replication_df <- data.frame(
  group     = c(rep("DMR CpGs", length(dmr_cpg_delta)),
                rep("Background CpGs", length(bg_cpg_delta))),
  delta_beta = c(dmr_cpg_delta, bg_cpg_delta)
)

# Wilcoxon test: are DMR CpGs more differentially methylated in TCGA?
wt <- wilcox.test(dmr_cpg_delta, bg_cpg_delta)
cat(sprintf("\nWilcoxon test (DMR vs background CpGs in TCGA): p = %.3e\n", wt$p.value))

p_replication <- ggplot(replication_df, aes(x = group, y = delta_beta, fill = group)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.1, outlier.shape = NA, fill = "white") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("DMR CpGs" = "#d62728", "Background CpGs" = "#aec7e8")) +
  annotate("text", x = 1.5, y = max(replication_df$delta_beta, na.rm = TRUE) * 0.9,
           label = sprintf("Wilcoxon p = %.2e", wt$p.value), size = 4) +
  labs(
    title = "TCGA PCPG replication of cfDNA DMR signal",
    subtitle = "Delta beta (Tumour – Normal) for CpGs inside vs outside DMRs",
    x = NULL, y = "Delta beta (TCGA tissue)", fill = NULL
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "none")

print(p_replication)

## Scatter: cfDNA delta methylation vs TCGA delta beta (requires per-CpG cfDNA stats)
# If you have per-CpG statistics from your own data (e.g., from DSS DMLtest),
# uncomment and adapt:
# cfdna_stats <- read.csv("your_dml_stats.csv")   # must have: chr, pos, diff (cfDNA delta)
# cfdna_gr    <- makeGRangesFromDataFrame(cfdna_stats, seqnames="chr", start="pos", end="pos")
# hits2       <- findOverlaps(tcga_cpg_gr, cfdna_gr)
# concordance_df <- data.frame(
#   tcga_delta  = delta_beta[queryHits(hits2)],
#   cfdna_delta = cfdna_stats$diff[subjectHits(hits2)]
# )
# cor_val <- cor(concordance_df$tcga_delta, concordance_df$cfdna_delta,
#                method = "spearman", use = "complete.obs")
# ggplot(concordance_df, aes(x = tcga_delta, y = cfdna_delta)) +
#   geom_point(alpha = 0.3, size = 0.8) +
#   geom_smooth(method = "lm") +
#   annotate("text", x = -0.3, y = 0.4,
#            label = sprintf("Spearman r = %.2f", cor_val)) +
#   labs(title = "cfDNA vs TCGA tissue methylation concordance",
#        x = "TCGA delta beta", y = "cfDNA delta methylation") +
#   theme_bw()
```


```{r}
# ── 5. Pathway Enrichment (missMethyl) ──────────────────────
# Tests whether genes near your DMRs are enriched in GO/KEGG terms.
# missMethyl corrects for the number of CpGs per gene (critical for
# 450K/EPIC arrays). For PPGL, expect enrichment in catecholamine
# synthesis, neural crest development, HIF signalling (SDHx/VHL).

# missMethyl works on array probe IDs. If you have WGBS DMRs, map
# them to the nearest 450K/EPIC probe first (see commented code below).

## If you have array data — provide the universe of tested probe IDs
# all_cpgs <- rownames(beta_mat)   # all probes tested
# sig_cpgs  <- cpgs_in_dmrs        # probes within DMRs

## WGBS → array probe mapping (if needed)
# library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
# anno450 <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
# anno_gr  <- makeGRangesFromDataFrame(
#   anno450[, c("chr","pos","pos")],
#   seqnames.field = "chr", start.field = "pos", end.field = "pos"
# )
# names(anno_gr) <- rownames(anno450)
# hits_wgbs <- findOverlaps(anno_gr, dmrs_gr)
# sig_cpgs  <- names(anno_gr)[queryHits(hits_wgbs)]
# all_cpgs  <- names(anno_gr)

## Gene Ontology enrichment
go_results <- gometh(
  sig.cpg   = sig_cpgs,
  all.cpg   = all_cpgs,
  collection = "GO",
  array.type = "450K",   # change to "EPIC" if applicable
  prior.prob = TRUE
)

go_sig <- go_results %>%
  filter(FDR < 0.05) %>%
  arrange(FDR)

cat("\nSignificant GO terms (FDR < 0.05):", nrow(go_sig), "\n")
print(head(go_sig, 20))

## KEGG pathway enrichment
kegg_results <- gometh(
  sig.cpg   = sig_cpgs,
  all.cpg   = all_cpgs,
  collection = "KEGG",
  array.type = "450K",
  prior.prob = TRUE
)

kegg_sig <- kegg_results %>%
  filter(FDR < 0.05) %>%
  arrange(FDR)

cat("\nSignificant KEGG pathways (FDR < 0.05):", nrow(kegg_sig), "\n")
print(head(kegg_sig, 20))

## Visualise top GO terms
top_go <- head(go_sig, 20)
p_go <- ggplot(top_go,
               aes(x = reorder(TERM, -log10(FDR)),
                   y = -log10(FDR),
                   size = N,
                   colour = -log10(FDR))) +
  geom_point() +
  coord_flip() +
  scale_colour_gradient(low = "#fdae6b", high = "#d62728") +
  scale_size_continuous(range = c(3, 10)) +
  labs(title = "GO enrichment — genes near PPGL cfDNA DMRs",
       subtitle = "missMethyl (CpG density corrected)",
       x = NULL, y = "-log10(FDR)", size = "N genes", colour = "-log10(FDR)") +
  theme_bw(base_size = 11)

print(p_go)
```


```{r}
# ── Save outputs ─────────────────────────────────────────────

dir.create("/researchers/caitlin.page/cf_nano/r_output/dmr_validity_results", showWarnings = FALSE)

write.csv(dmr_annotated_df,     "/researchers/caitlin.page/cf_nano/r_output/dmr_validity_results/dmr_annotations.csv",     row.names = FALSE)
write.csv(enrich_df,            "/researchers/caitlin.page/cf_nano/r_output/dmr_validity_results/chromatin_enrichment.csv", row.names = FALSE)
write.csv(go_sig,               "/researchers/caitlin.page/cf_nano/r_output/dmr_validity_results/GO_enrichment.csv",        row.names = FALSE)
write.csv(kegg_sig,             "/researchers/caitlin.page/cf_nano/r_output/dmr_validity_results/KEGG_enrichment.csv",      row.names = FALSE)

ggsave("/researchers/caitlin.page/cf_nano/r_output/dmr_validity_results/01_cpg_context.pdf",       p_cpg,         width = 8, height = 4)
ggsave("/researchers/caitlin.page/cf_nano/r_output/dmr_validity_results/02_gene_context.pdf",      p_gene,        width = 8, height = 4)
ggsave("/researchers/caitlin.page/cf_nano/r_output/dmr_validity_results/03_chromatin_enrich.pdf",  p_enrich,      width = 7, height = 4)
ggsave("/researchers/caitlin.page/cf_nano/r_output/dmr_validity_results/04_tcga_replication.pdf",  p_replication, width = 6, height = 5)
ggsave("/researchers/caitlin.page/cf_nano/r_output/dmr_validity_results/05_go_enrichment.pdf",     p_go,          width = 10, height = 6)

cat("\n✓ All outputs saved to dmr_validity_results/\n")
```

## DAU

# ── 2. Functional Annotation (annotatr) ─────────────────────

annot_types <- c(
  paste0(GENOME, "_cpgs"),
  paste0(GENOME, "_genes_promoters"),
  paste0(GENOME, "_genes_5UTRs"),
  paste0(GENOME, "_genes_exons"),
  paste0(GENOME, "_genes_introns"),
  paste0(GENOME, "_genes_3UTRs"),
  paste0(GENOME, "_genes_intergenic")
)

annotations   <- build_annotations(genome = GENOME, annotations = annot_types)
dmr_annotated <- annotate_regions(regions = dmrs_gr, annotations = annotations,
                                  ignore.strand = TRUE, quiet = FALSE)
dmr_annotated_df <- as.data.frame(dmr_annotated)

# Force plain types — annotatr can produce list/AsIs columns
dmr_annotated_df <- dmr_annotated_df %>%
  mutate(across(everything(), function(x) {
    if (is.list(x) || inherits(x, "AsIs"))
      sapply(x, function(v) if (length(v) == 0) NA_character_
                            else paste(as.character(v), collapse = ","))
    else x
  })) %>%
  mutate(
    annot.type = as.character(annot.type),
    gene.F     = as.numeric(gene.F),
    FDR        = as.numeric(FDR)
  )

## 2a. CpG context distribution
cpg_summary <- dmr_annotated_df %>%
  filter(grepl("cpg", annot.type)) %>%
  mutate(cpg_context = case_when(
    grepl("islands", annot.type) ~ "CpG Island",
    grepl("shores",  annot.type) ~ "Shore",
    grepl("shelves", annot.type) ~ "Shelf",
    grepl("inter",   annot.type) ~ "Open Sea"
  )) %>%
  dplyr::count(cpg_context) %>%
  mutate(pct = n / sum(n) * 100)

p_cpg <- ggplot(cpg_summary, aes(x = reorder(cpg_context, pct), y = pct)) +
  geom_col(fill = "#7b2d8b") +
  coord_flip() +
  labs(title = "Region distribution across CpG features",
       x = NULL, y = "% of regions") +
  theme_bw(base_size = 12)

## 2b. Genic context distribution
gene_summary <- dmr_annotated_df %>%
  filter(grepl("genes", annot.type)) %>%
  mutate(gene_context = case_when(
    grepl("promoter",   annot.type) ~ "Promoter",
    grepl("5UTR",       annot.type) ~ "5'UTR",
    grepl("exon",       annot.type) ~ "Exon",
    grepl("intron",     annot.type) ~ "Intron",
    grepl("3UTR",       annot.type) ~ "3'UTR",
    grepl("intergenic", annot.type) ~ "Intergenic"
  )) %>%
  dplyr::count(gene_context) %>%
  mutate(pct = n / sum(n) * 100)

p_gene <- ggplot(gene_summary, aes(x = reorder(gene_context, pct), y = pct)) +
  geom_col(fill = "#2c7bb6") +
  coord_flip() +
  labs(title = "Region distribution across genic features",
       x = NULL, y = "% of regions") +
  theme_bw(base_size = 12)

print(p_cpg / p_gene)

## 2c. Effect size (gene.F) and significance (FDR) diagnostics
p_fstat <- ggplot(as.data.frame(dmrs_gr),
                  aes(x = gene.F, y = -log10(FDR))) +
  geom_point(alpha = 0.5, size = 1.5, colour = "#7b2d8b") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey40") +
  labs(title = "F-statistic vs significance",
       subtitle = "Dashed line = FDR 0.05",
       x = "gene.F", y = "-log10(FDR)") +
  theme_bw(base_size = 12)

p_nexons <- ggplot(as.data.frame(dmrs_gr),
                   aes(x = NExons, y = gene.F)) +
  geom_jitter(alpha = 0.4, width = 0.2, colour = "#2c7bb6") +
  geom_smooth(method = "lm", se = FALSE, colour = "grey30") +
  labs(title = "Region complexity vs effect size",
       x = "NExons", y = "gene.F") +
  theme_bw(base_size = 12)

print(p_fstat | p_nexons)

## 2d. CGI promoter overlap — PPGL key check
n_cgi_promoter <- dmr_annotated_df %>%
  filter(grepl("island|promoter", annot.type)) %>%
  pull(dmr_id) %>% unique() %>% length()

cat(sprintf("\nRegions overlapping CGI or promoter: %d (%.1f%%)\n",
            n_cgi_promoter, n_cgi_promoter / length(dmrs_gr) * 100))


# ── 3. Chromatin State Enrichment (regioneR + Roadmap) ──────
# E080 = fetal adrenal gland, E053/E054 = neurospheres (neural crest),
# E029 = monocytes (negative control)

ah <- AnnotationHub()

fetch_roadmap_states <- function(ah, eid,
                                 active_states = c("1_TssA", "2_TssAFlnk",
                                                   "6_EnhG", "7_Enh")) {
  q <- query(ah, c("EpigenomeRoadMap", eid, "15_coreMarks", "bed"))
  if (length(q) == 0) { message("No data for ", eid); return(NULL) }
  gr <- ah[[names(q)[1]]]
  gr <- gr[gr$name %in% active_states]
  seqlevelsStyle(gr) <- "UCSC"
  gr
}

roadmap_refs <- Filter(Negate(is.null), list(
  fetal_adrenal  = fetch_roadmap_states(ah, "E080"),
  neurosphere_dr = fetch_roadmap_states(ah, "E053"),
  neurosphere_sp = fetch_roadmap_states(ah, "E054"),
  monocyte       = fetch_roadmap_states(ah, "E029")
))

genome_mask <- filterChromosomes(getGenome(GENOME))

set.seed(123)
perm_results <- lapply(names(roadmap_refs), function(ref_name) {
  cat("  Testing:", ref_name, "\n")
  pt <- permTest(
    A = dmrs_gr, B = roadmap_refs[[ref_name]],
    genome = genome_mask,
    randomize.function = resampleRegions,
    evaluate.function  = numOverlaps,
    ntimes   = 1000, mc.set.seed = TRUE,
    mc.cores = max(1, parallel::detectCores() - 1)
  )
  data.frame(reference   = ref_name,
             observed    = pt$numOverlaps$observed,
             expected    = round(mean(pt$numOverlaps$permuted), 1),
             fold_enrich = round(pt$numOverlaps$observed / mean(pt$numOverlaps$permuted), 2),
             zscore      = round(pt$numOverlaps$zscore, 2),
             pval        = pt$numOverlaps$pval)
})
enrich_df <- bind_rows(perm_results)

cat("\n── Chromatin State Enrichment ──\n")
print(enrich_df)

p_enrich <- ggplot(enrich_df,
                   aes(x = reorder(reference, fold_enrich), y = fold_enrich,
                       fill = pval < 0.05)) +
  geom_col() +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey40") +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#2ca02c", "FALSE" = "#aec7e8"),
                    labels = c("TRUE" = "p < 0.05", "FALSE" = "n.s.")) +
  labs(title = "Region enrichment in active regulatory states",
       subtitle = "Roadmap Epigenomics (active TSS + enhancers)",
       x = NULL, y = "Fold enrichment over random", fill = "") +
  theme_bw(base_size = 12)

print(p_enrich)


# ── 4. TCGA PCPG Replication ────────────────────────────────
# Tests whether CpGs within your regions are more differentially
# methylated in TCGA PPGL tissue data than background CpGs.

cat("\nDownloading TCGA PCPG 450K data...\n")

tcga_query <- GDCquery(
  project       = "TCGA-PCPG",
  data.category = "DNA Methylation",
  data.type     = "Methylation Beta Value",
  platform      = "Illumina Human Methylation 450"
)
GDCdownload(tcga_query, method = "api", files.per.chunk = 10)
tcga_data <- GDCprepare(tcga_query)

beta_mat <- assay(tcga_data)
col_data <- as.data.frame(colData(tcga_data))
row_data <- as.data.frame(rowRanges(tcga_data))

col_data$sample_type_clean <- ifelse(
  grepl("Normal", col_data$sample_type, ignore.case = TRUE), "Normal", "Tumour"
)
cat("TCGA PCPG samples:\n"); print(table(col_data$sample_type_clean))

tcga_cpg_gr <- makeGRangesFromDataFrame(
  row_data, seqnames.field = "seqnames",
  start.field = "start", end.field = "end"
)
names(tcga_cpg_gr) <- rownames(row_data)
seqlevelsStyle(tcga_cpg_gr) <- "UCSC"

hits         <- findOverlaps(tcga_cpg_gr, dmrs_gr)
cpgs_in_dmrs <- names(tcga_cpg_gr)[queryHits(hits)]
cat("TCGA CpGs overlapping your regions:", length(unique(cpgs_in_dmrs)), "\n")

tumour_idx  <- which(col_data$sample_type_clean == "Tumour")
normal_idx  <- which(col_data$sample_type_clean == "Normal")
delta_beta  <- rowMeans(beta_mat[, tumour_idx], na.rm = TRUE) -
               rowMeans(beta_mat[, normal_idx], na.rm = TRUE)

dmr_cpg_delta <- delta_beta[cpgs_in_dmrs]
bg_cpg_delta  <- delta_beta[setdiff(names(delta_beta), cpgs_in_dmrs)]

wt <- wilcox.test(dmr_cpg_delta, bg_cpg_delta)
cat(sprintf("Wilcoxon test (region CpGs vs background): p = %.3e\n", wt$p.value))

replication_df <- data.frame(
  group      = c(rep("Region CpGs",     length(dmr_cpg_delta)),
                 rep("Background CpGs", length(bg_cpg_delta))),
  delta_beta = c(dmr_cpg_delta, bg_cpg_delta)
)

p_replication <- ggplot(replication_df,
                        aes(x = group, y = delta_beta, fill = group)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.1, outlier.shape = NA, fill = "white") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("Region CpGs" = "#7b2d8b",
                               "Background CpGs" = "#aec7e8")) +
  annotate("text", x = 1.5,
           y = max(replication_df$delta_beta, na.rm = TRUE) * 0.9,
           label = sprintf("Wilcoxon p = %.2e", wt$p.value), size = 4) +
  labs(title = "TCGA PCPG replication",
       subtitle = "Delta beta (Tumour – Normal) for CpGs inside vs outside regions",
       x = NULL, y = "Delta beta (TCGA tissue)") +
  theme_bw(base_size = 12) + theme(legend.position = "none")

print(p_replication)

# Also test: do higher-F regions have larger TCGA delta beta?
region_tcga_df <- as.data.frame(dmrs_gr) %>%
  mutate(dmr_idx = seq_len(n())) %>%
  rowwise() %>%
  mutate(
    mean_delta = {
      idx <- subjectHits(hits)[subjectHits(hits) == dmr_idx]
      cpg_ids <- names(tcga_cpg_gr)[queryHits(hits)[subjectHits(hits) == dmr_idx]]
      if (length(cpg_ids) > 0) mean(abs(delta_beta[cpg_ids]), na.rm = TRUE) else NA_real_
    }
  ) %>% ungroup()

p_fstat_tcga <- ggplot(region_tcga_df %>% filter(!is.na(mean_delta)),
                       aes(x = gene.F, y = mean_delta)) +
  geom_point(alpha = 0.5, size = 1.5, colour = "#7b2d8b") +
  geom_smooth(method = "lm", se = TRUE, colour = "grey30") +
  labs(title = "F-statistic vs TCGA methylation change",
       subtitle = "Concordance between your effect size and tissue signal",
       x = "gene.F", y = "Mean |delta beta| in TCGA") +
  theme_bw(base_size = 12)

print(p_fstat_tcga)


# ── 5. Pathway Enrichment (missMethyl) ──────────────────────
# Map your regions to 450K/EPIC probe space for CpG-corrected
# GO/KEGG enrichment.

library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
anno450 <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
anno_gr <- GRanges(
  seqnames = anno450$chr,
  ranges   = IRanges(start = anno450$pos, width = 1)
)
names(anno_gr) <- rownames(anno450)
seqlevelsStyle(anno_gr) <- "UCSC"

hits_450 <- findOverlaps(anno_gr, dmrs_gr)
sig_cpgs <- names(anno_gr)[queryHits(hits_450)]
all_cpgs <- names(anno_gr)

cat("\n450K probes overlapping regions:", length(unique(sig_cpgs)), "\n")

go_results <- gometh(sig.cpg = sig_cpgs, all.cpg = all_cpgs,
                     collection = "GO", array.type = "450K", prior.prob = TRUE)
go_sig <- go_results %>% filter(FDR < 0.05) %>% arrange(FDR)
cat("Significant GO terms (FDR < 0.05):", nrow(go_sig), "\n")
print(head(go_sig, 20))

kegg_results <- gometh(sig.cpg = sig_cpgs, all.cpg = all_cpgs,
                       collection = "KEGG", array.type = "450K", prior.prob = TRUE)
kegg_sig <- kegg_results %>% filter(FDR < 0.05) %>% arrange(FDR)
cat("Significant KEGG pathways (FDR < 0.05):", nrow(kegg_sig), "\n")
print(head(kegg_sig, 20))

p_go <- ggplot(head(go_sig, 20),
               aes(x = reorder(TERM, -log10(FDR)),
                   y = -log10(FDR), size = N, colour = -log10(FDR))) +
  geom_point() + coord_flip() +
  scale_colour_gradient(low = "#fdae6b", high = "#d62728") +
  scale_size_continuous(range = c(3, 10)) +
  labs(title = "GO enrichment — genes near PPGL cfDNA regions",
       subtitle = "missMethyl (CpG density corrected)",
       x = NULL, y = "-log10(FDR)", size = "N genes") +
  theme_bw(base_size = 11)

print(p_go)



```{r}
# Set genome build — change to "hg19" if your data is GRCh37
GENOME <- "hg38"

# ── 1. Load daus ────────────────────────────────────────────
daus_gr <- makeGRangesFromDataFrame(
  dau_basictiles %>% mutate(dau_id = region_id),
  seqnames.field     = "seqnames",
  start.field        = "start",
  end.field          = "end",
  keep.extra.columns = TRUE
)
seqlevelsStyle(daus_gr) <- "UCSC"

# Separate hyper and hypo daus for direction-aware analyses
daus_hyper <- daus_gr[daus_gr$direction == "hyper"]
daus_hypo  <- daus_gr[daus_gr$direction == "hypo"]

cat("Total daus:", length(daus_gr), "\n")
cat("  Hypermethylated:", length(daus_hyper), "\n")
cat("  Hypomethylated: ", length(daus_hypo), "\n")


# ── 2. Functional Annotation (annotatr) ─────────────────────
# Annotates daus to CpG features (islands/shores/shelves/open sea)
# and genic features (promoters/exons/introns/intergenic).
# For PPGL, enrichment at CGI promoters supports the known
# SDHx-driven hypermethylator phenotype.

annot_types <- c(
  # CpG features
  paste0(GENOME, "_cpgs"),
  # Genic features
  paste0(GENOME, "_genes_promoters"),
  paste0(GENOME, "_genes_5UTRs"),
  paste0(GENOME, "_genes_exons"),
  paste0(GENOME, "_genes_introns"),
  paste0(GENOME, "_genes_3UTRs"),
  paste0(GENOME, "_genes_intergenic")
)

annotations <- build_annotations(genome = GENOME, annotations = annot_types)

dau_annotated <- annotate_regions(
  regions    = daus_gr,
  annotations = annotations,
  ignore.strand = TRUE,
  quiet = FALSE
)

dau_annotated_df <- as.data.frame(dau_annotated)
```


```{r}
## 2a. CpG context distribution
cpg_summary <- dau_annotated_df %>%
  filter(grepl("cpg", annot.type)) %>%
  mutate(cpg_context = case_when(
    grepl("islands", annot.type) ~ "CpG Island",
    grepl("shores",  annot.type) ~ "Shore",
    grepl("shelves", annot.type) ~ "Shelf",
    grepl("inter",   annot.type) ~ "Open Sea"
  )) %>%
  dplyr::count(cpg_context) %>%
  mutate(pct = n / sum(n) * 100)

p_cpg <- ggplot(cpg_summary, aes(x = reorder(cpg_context, pct), y = pct)) +
  geom_col(fill = "#7b2d8b") +
  coord_flip() +
  labs(title = "Region distribution across CpG features",
       x = NULL, y = "% of regions") +
  theme_bw(base_size = 12)

## 2b. Genic context distribution
gene_summary <- dau_annotated_df %>%
  filter(grepl("genes", annot.type)) %>%
  mutate(gene_context = case_when(
    grepl("promoter",   annot.type) ~ "Promoter",
    grepl("5UTR",       annot.type) ~ "5'UTR",
    grepl("exon",       annot.type) ~ "Exon",
    grepl("intron",     annot.type) ~ "Intron",
    grepl("3UTR",       annot.type) ~ "3'UTR",
    grepl("intergenic", annot.type) ~ "Intergenic"
  )) %>%
  dplyr::count(gene_context) %>%
  mutate(pct = n / sum(n) * 100)

p_gene <- ggplot(gene_summary, aes(x = reorder(gene_context, pct), y = pct)) +
  geom_col(fill = "#2c7bb6") +
  coord_flip() +
  labs(title = "Region distribution across genic features",
       x = NULL, y = "% of regions") +
  theme_bw(base_size = 12)

print(p_cpg / p_gene)

## 2c. Effect size (gene.F) and significance (FDR) diagnostics
p_fstat <- ggplot(as.data.frame(daus_gr),
                  aes(x = gene.F, y = -log10(FDR))) +
  geom_point(alpha = 0.5, size = 1.5, colour = "#7b2d8b") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey40") +
  labs(title = "F-statistic vs significance",
       subtitle = "Dashed line = FDR 0.05",
       x = "gene.F", y = "-log10(FDR)") +
  theme_bw(base_size = 12)

p_nexons <- ggplot(as.data.frame(daus_gr),
                   aes(x = NExons, y = gene.F)) +
  geom_jitter(alpha = 0.4, width = 0.2, colour = "#2c7bb6") +
  geom_smooth(method = "lm", se = FALSE, colour = "grey30") +
  labs(title = "Region complexity vs effect size",
       x = "NExons", y = "gene.F") +
  theme_bw(base_size = 12)

print(p_fstat | p_nexons)

## 2d. CGI promoter overlap — PPGL key check
n_cgi_promoter <- dau_annotated_df %>%
  filter(grepl("island|promoter", annot.type)) %>%
  pull(dau_id) %>% unique() %>% length()

cat(sprintf("\nRegions overlapping CGI or promoter: %d (%.1f%%)\n",
            n_cgi_promoter, n_cgi_promoter / length(daus_gr) * 100))
```


```{r}
# ── 3. Chromatin State Enrichment (regioneR + Roadmap) ──────
# Tests whether daus are enriched in regulatory regions of
# PPGL-relevant cell types vs random expectation.
# E080 = fetal adrenal gland (best available proxy for chromaffin cells)
# E053/E054 = neural crest-derived neurospheres
# E029 = monocytes (as a negative control — should NOT be enriched)

ah <- AnnotationHub()

## Helper to fetch Roadmap 15-state chromatin state BED for one sample
fetch_roadmap_states <- function(ah, eid, active_states = c("1_TssA", "2_TssAFlnk",
                                                             "6_EnhG", "7_Enh")) {
  # Query AnnotationHub for the chromatin state segmentation
  q <- query(ah, c("EpigenomeRoadMap", eid, "15_coreMarks", "bed"))
  if (length(q) == 0) {
    message("No Roadmap data found for ", eid)
    return(NULL)
  }
  states_gr <- ah[[names(q)[1]]]
  # Keep only active regulatory states
  active_gr <- states_gr[states_gr$name %in% active_states]
  seqlevelsStyle(active_gr) <- "UCSC"
  active_gr
}

## Fetch reference epigenomes
cat("Fetching Roadmap chromatin states (this caches locally after first run)...\n")
roadmap_refs <- list(
  fetal_adrenal  = fetch_roadmap_states(ah, "E080"),  # fetal adrenal gland
  neurosphere_dr = fetch_roadmap_states(ah, "E053"),  # neural crest-derived
  neurosphere_sp = fetch_roadmap_states(ah, "E054"),  # neural crest-derived
  monocyte       = fetch_roadmap_states(ah, "E029")   # negative control
)
roadmap_refs <- Filter(Negate(is.null), roadmap_refs)

## Genome mask for regioneR permutations (exclude gaps/unassembled regions)
# Using a simplified mask — replace with BSgenome mask for production runs
genome_mask <- filterChromosomes(getGenome(GENOME))

## Run permutation test for each reference
set.seed(123)
perm_results <- lapply(names(roadmap_refs), function(ref_name) {
  cat("  Testing enrichment in:", ref_name, "\n")
  pt <- permTest(
    A                = daus_gr,
    B                = roadmap_refs[[ref_name]],
    genome           = genome_mask,
    randomize.function = resampleRegions,
    evaluate.function  = numOverlaps,
    ntimes           = 1000,
    mc.set.seed      = TRUE,
    mc.cores         = 1
  )
  list(
    ref      = ref_name,
    observed = pt$numOverlaps$observed,
    expected = mean(pt$numOverlaps$permuted),
    pval     = pt$numOverlaps$pval,
    zscore   = pt$numOverlaps$zscore,
    result   = pt
  )
})
names(perm_results) <- names(roadmap_refs)

## Summarise enrichment
enrich_df <- bind_rows(lapply(perm_results, function(x) {
  data.frame(
    reference = x$ref,
    observed  = x$observed,
    expected  = round(x$expected, 1),
    fold_enrich = round(x$observed / x$expected, 2),
    zscore    = round(x$zscore, 2),
    pval      = x$pval
  )
}))

cat("\n── Chromatin State Enrichment Summary ──\n")
print(enrich_df)

p_enrich <- ggplot(enrich_df,
                   aes(x = reorder(reference, fold_enrich),
                       y = fold_enrich,
                       fill = pval < 0.05)) +
  geom_col() +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey40") +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#2ca02c", "FALSE" = "#aec7e8"),
                    labels = c("TRUE" = "p < 0.05", "FALSE" = "n.s.")) +
  labs(title = "dau enrichment in active regulatory regions",
       subtitle = "Roadmap Epigenomics chromatin states (active TSS + enhancers)",
       x = NULL, y = "Fold enrichment over random", fill = "") +
  theme_bw(base_size = 12)

print(p_enrich)
```


```{r}
# ── 4. TCGA PCPG Replication ────────────────────────────────
# Downloads TCGA pheochromocytoma/paraganglioma 450K methylation data
# and checks whether CpGs within your daus are differentially
# methylated in TCGA tissue data — tissue replication of cfDNA signal.
# Note: TCGA PCPG uses hg38; adjust liftOver if your daus are hg19.

cat("\nDownloading TCGA PCPG 450K methylation data...\n")

tcga_query <- GDCquery(
  project            = "TCGA-PCPG",
  data.category      = "DNA Methylation",
  data.type          = "Methylation Beta Value",
  platform           = "Illumina Human Methylation 450"
)

GDCdownload(tcga_query, method = "api", files.per.chunk = 10)

tcga_data <- GDCprepare(tcga_query)   # SummarizedExperiment

# Extract beta matrix and sample metadata
beta_mat  <- assay(tcga_data)         # rows = CpGs, cols = samples
col_data  <- as.data.frame(colData(tcga_data))
row_data  <- as.data.frame(rowRanges(tcga_data))

# TCGA PCPG includes both tumour and normal tissue
# Classify samples
col_data$sample_type_clean <- ifelse(
  grepl("Normal", col_data$sample_type, ignore.case = TRUE),
  "Normal", "Tumour"
)

cat("TCGA PCPG samples:\n")
print(table(col_data$sample_type_clean))

# Build GRanges for TCGA CpGs
tcga_cpg_gr <- makeGRangesFromDataFrame(
  row_data,
  seqnames.field = "seqnames",
  start.field    = "start",
  end.field      = "end",
  keep.extra.columns = FALSE
)
names(tcga_cpg_gr) <- rownames(row_data)
seqlevelsStyle(tcga_cpg_gr) <- "UCSC"

## Find CpGs that fall within your daus
hits <- findOverlaps(tcga_cpg_gr, daus_gr)
cpgs_in_daus <- names(tcga_cpg_gr)[queryHits(hits)]

cat("\nTCGA CpGs overlapping your daus:", length(unique(cpgs_in_daus)), "\n")

## Compare mean methylation for CpGs inside vs outside daus
tumour_idx <- which(col_data$sample_type_clean == "Tumour")
normal_idx <- which(col_data$sample_type_clean == "Normal")

# Mean beta per CpG in tumour vs normal
mean_tumour <- rowMeans(beta_mat[, tumour_idx], na.rm = TRUE)
mean_normal <- rowMeans(beta_mat[, normal_idx], na.rm = TRUE)
delta_beta   <- mean_tumour - mean_normal

## Summarise delta beta for dau-overlapping CpGs
dau_cpg_delta  <- delta_beta[cpgs_in_daus]
bg_cpg_delta   <- delta_beta[setdiff(names(delta_beta), cpgs_in_daus)]

replication_df <- data.frame(
  group     = c(rep("dau CpGs", length(dau_cpg_delta)),
                rep("Background CpGs", length(bg_cpg_delta))),
  delta_beta = c(dau_cpg_delta, bg_cpg_delta)
)

# Wilcoxon test: are dau CpGs more differentially methylated in TCGA?
wt <- wilcox.test(dau_cpg_delta, bg_cpg_delta)
cat(sprintf("\nWilcoxon test (dau vs background CpGs in TCGA): p = %.3e\n", wt$p.value))

p_replication <- ggplot(replication_df, aes(x = group, y = delta_beta, fill = group)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.1, outlier.shape = NA, fill = "white") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("dau CpGs" = "#d62728", "Background CpGs" = "#aec7e8")) +
  annotate("text", x = 1.5, y = max(replication_df$delta_beta, na.rm = TRUE) * 0.9,
           label = sprintf("Wilcoxon p = %.2e", wt$p.value), size = 4) +
  labs(
    title = "TCGA PCPG replication of cfDNA dau signal",
    subtitle = "Delta beta (Tumour – Normal) for CpGs inside vs outside daus",
    x = NULL, y = "Delta beta (TCGA tissue)", fill = NULL
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "none")

print(p_replication)

## Scatter: cfDNA delta methylation vs TCGA delta beta (requires per-CpG cfDNA stats)
# If you have per-CpG statistics from your own data (e.g., from DSS DMLtest),
# uncomment and adapt:
# cfdna_stats <- read.csv("your_dml_stats.csv")   # must have: chr, pos, diff (cfDNA delta)
# cfdna_gr    <- makeGRangesFromDataFrame(cfdna_stats, seqnames="chr", start="pos", end="pos")
# hits2       <- findOverlaps(tcga_cpg_gr, cfdna_gr)
# concordance_df <- data.frame(
#   tcga_delta  = delta_beta[queryHits(hits2)],
#   cfdna_delta = cfdna_stats$diff[subjectHits(hits2)]
# )
# cor_val <- cor(concordance_df$tcga_delta, concordance_df$cfdna_delta,
#                method = "spearman", use = "complete.obs")
# ggplot(concordance_df, aes(x = tcga_delta, y = cfdna_delta)) +
#   geom_point(alpha = 0.3, size = 0.8) +
#   geom_smooth(method = "lm") +
#   annotate("text", x = -0.3, y = 0.4,
#            label = sprintf("Spearman r = %.2f", cor_val)) +
#   labs(title = "cfDNA vs TCGA tissue methylation concordance",
#        x = "TCGA delta beta", y = "cfDNA delta methylation") +
#   theme_bw()
```


```{r}
# ── 5. Pathway Enrichment (missMethyl) ──────────────────────
# Tests whether genes near your daus are enriched in GO/KEGG terms.
# missMethyl corrects for the number of CpGs per gene (critical for
# 450K/EPIC arrays). For PPGL, expect enrichment in catecholamine
# synthesis, neural crest development, HIF signalling (SDHx/VHL).

# missMethyl works on array probe IDs. If you have WGBS daus, map
# them to the nearest 450K/EPIC probe first (see commented code below).

## If you have array data — provide the universe of tested probe IDs
# all_cpgs <- rownames(beta_mat)   # all probes tested
# sig_cpgs  <- cpgs_in_daus        # probes within daus

## WGBS → array probe mapping (if needed)
# library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
# anno450 <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
# anno_gr  <- makeGRangesFromDataFrame(
#   anno450[, c("chr","pos","pos")],
#   seqnames.field = "chr", start.field = "pos", end.field = "pos"
# )
# names(anno_gr) <- rownames(anno450)
# hits_wgbs <- findOverlaps(anno_gr, daus_gr)
# sig_cpgs  <- names(anno_gr)[queryHits(hits_wgbs)]
# all_cpgs  <- names(anno_gr)

## Gene Ontology enrichment
go_results <- gometh(
  sig.cpg   = sig_cpgs,
  all.cpg   = all_cpgs,
  collection = "GO",
  array.type = "450K",   # change to "EPIC" if applicable
  prior.prob = TRUE
)

go_sig <- go_results %>%
  filter(FDR < 0.05) %>%
  arrange(FDR)

cat("\nSignificant GO terms (FDR < 0.05):", nrow(go_sig), "\n")
print(head(go_sig, 20))

## KEGG pathway enrichment
kegg_results <- gometh(
  sig.cpg   = sig_cpgs,
  all.cpg   = all_cpgs,
  collection = "KEGG",
  array.type = "450K",
  prior.prob = TRUE
)

kegg_sig <- kegg_results %>%
  filter(FDR < 0.05) %>%
  arrange(FDR)

cat("\nSignificant KEGG pathways (FDR < 0.05):", nrow(kegg_sig), "\n")
print(head(kegg_sig, 20))

## Visualise top GO terms
top_go <- head(go_sig, 20)
p_go <- ggplot(top_go,
               aes(x = reorder(TERM, -log10(FDR)),
                   y = -log10(FDR),
                   size = N,
                   colour = -log10(FDR))) +
  geom_point() +
  coord_flip() +
  scale_colour_gradient(low = "#fdae6b", high = "#d62728") +
  scale_size_continuous(range = c(3, 10)) +
  labs(title = "GO enrichment — genes near PPGL cfDNA daus",
       subtitle = "missMethyl (CpG density corrected)",
       x = NULL, y = "-log10(FDR)", size = "N genes", colour = "-log10(FDR)") +
  theme_bw(base_size = 11)

print(p_go)


# ── Save outputs ─────────────────────────────────────────────

dir.create("/researchers/caitlin.page/cf_nano/r_output/dau_validity_results", showWarnings = FALSE)

write.csv(dau_annotated_df,     "/researchers/caitlin.page/cf_nano/r_output/dau_validity_results/dau_annotations.csv",     row.names = FALSE)
write.csv(enrich_df,            "/researchers/caitlin.page/cf_nano/r_output/dau_validity_results/chromatin_enrichment.csv", row.names = FALSE)
write.csv(go_sig,               "/researchers/caitlin.page/cf_nano/r_output/dau_validity_results/GO_enrichment.csv",        row.names = FALSE)
write.csv(kegg_sig,             "/researchers/caitlin.page/cf_nano/r_output/dau_validity_results/KEGG_enrichment.csv",      row.names = FALSE)

ggsave("/researchers/caitlin.page/cf_nano/r_output/dau_validity_results/01_cpg_context.pdf",       p_cpg,         width = 8, height = 4)
ggsave("/researchers/caitlin.page/cf_nano/r_output/dau_validity_results/02_gene_context.pdf",      p_gene,        width = 8, height = 4)
ggsave("/researchers/caitlin.page/cf_nano/r_output/dau_validity_results/03_chromatin_enrich.pdf",  p_enrich,      width = 7, height = 4)
ggsave("/researchers/caitlin.page/cf_nano/r_output/dau_validity_results/04_tcga_replication.pdf",  p_replication, width = 6, height = 5)
ggsave("/researchers/caitlin.page/cf_nano/r_output/dau_validity_results/05_go_enrichment.pdf",     p_go,          width = 10, height = 6)

cat("\n✓ All outputs saved to dau_validity_results/\n")
```

