---
title: "dtuDiffGroupsAndGst"
author: "caitlinpage"
date: "2026-05-27"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---


# things to run through gst

* traditional beta value dmr regions

* dtu with alpha
** 3, 4, 5 groups

* dtu with beta

* de dtu alpha (each gene:transcript own thing)


See [workflow](workflow.html) for code showing how these files were made

# Packages

```{r}
library(annotatr)
library(data.table)
library(plyranges)
library(tidyr)
library(stringr)
library(dplyr)
library(ggplot2)
library(edgeR)
library(OpTiles)
library(ggVennDiagram)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
library(missMethyl)
```

```{r}
seqs <- seqnames(BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38)[1:22]

cg_sites <- lapply(seqs, function(x) {
        cbind(data.frame(Biostrings::matchPattern("CG", BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38[[x]])),
            seqnames = x
        )}) %>% bind_rows()
```

```{r}
all_samples <- readRDS("/researchers/caitlin.page/cf_nano/r_output/all_rrbs_samples.rds")
optiles_all <- readRDS("/researchers/caitlin.page/cf_nano/r_output/optiles_all.rds")
dtu.optiles <- readRDS("/researchers/caitlin.page/cf_nano/r_output/ts.gene_optiles.rds")
dmrcate_seq_dmr <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dmr_encode_gm12878_k562.rds")
```

```{r}
all_samples_simp <- all_samples %>% distinct(seqnames, read_start, read_end, strand, sample, read_id, alpha, meth_pattern, num_meth, total)
colnames(all_samples_simp)[2:3] <- c("start", "end")
```


# run tiles method on 3 alpha groups

```{r}
dtu_function <- function(all_samples, optiles_all, num_groups = 5) {
  dtu_optile <- find_overlaps(as_granges(all_samples), as_granges(optiles_all)) %>% data.frame() %>%
  distinct(sample, read_id, read_start, read_end, alpha, region_id) 
  
  if(num_groups == 5) {
  dtu_optile <- dtu_optile %>% 
    mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8"))
  } else if(num_groups == 3) {
  dtu_optile <- dtu_optile %>% 
    mutate(alpha_group = case_when(alpha < 0.34 ~ "a < 0.34",
                                alpha > 0.65 ~ "a > 0.65",
                                TRUE ~ "0.34 <= a <= 0.65"))
  } else if(num_groups == 4) {
  dtu_optile <- dtu_optile %>% 
    mutate(alpha_group = case_when(alpha <= 0.25 ~ "a <= 0.25",
                                alpha >= 0.8 ~ "a >= 0.75",
                                alpha > 0.25 & alpha <= 0.5 ~ "0.25 <= a <= 0.5",
                                TRUE ~ "0.5 < a < 0.75"))
}

dtu_optile <- dtu_optile %>% 
  mutate(transcript = paste0(region_id, "_", alpha_group)) %>%
  group_by(sample, transcript) %>%
 summarise(transcript_count = n()) %>% ungroup() %>% data.frame()

# build matrix of "transcript" counts - transcript rows, sample as columns
dtu_optile <- dtu_optile %>% pivot_wider(names_from = sample, values_from = transcript_count)
dtu_optile[is.na(dtu_optile)] <- 0
dtu_optile$gene_id <- sub("_.*", "", dtu_optile$transcript)

# DTU analysis with edgeR
## Alicia prefers gene method over Simes method (documentation slightly confusing about difference)
y <- DGEList(as.matrix(dtu_optile[,2:5]), group = c(1,1,2,2), genes = dtu_optile[,c("gene_id", "transcript")])
design <- model.matrix(~ c(1,1,2,2), data = y$samples)
fit <- glmQLFit(y, design, robust=TRUE)
ds <- diffSpliceDGE(fit, coef = 2, geneid = "gene_id", exonid = "transcript")
ts.gene_optiles <- topSpliceDGE(ds, test = "gene", number = Inf)

# add info to results: get proportion counts
gene_cg <- find_overlaps(as_granges(optiles_all), as_granges(cg_sites)) %>% data.frame() %>%
  group_by(region_id) %>%
  summarise(n_cg = n()) %>%
  ungroup() %>% data.frame()
fit$genes <- fit$genes %>% mutate(n_cg = gene_cg[match(.$gene_id, gene_cg$region_id), "n_cg"])
rpk <- y$counts/rep(fit$genes$n_cg/1000, times = ncol(y$counts))
tpm_mat <- sweep(rpk, 2, colSums(rpk)/1e6, "/")
tpm_mat <- tpm_mat %>% data.frame() %>% mutate(id = dtu_optile$transcript, gene_id = dtu_optile$gene_id)

tpm_mat <- tpm_mat %>%
  group_by(gene_id) %>%
  mutate(prop_S1_gm = b1/sum(b1),
         prop_S2_gm = b2/sum(b2),
         prop_S3_k = cml1/sum(cml1),
         prop_S4_k = cml2/sum(cml2)) %>%
  ungroup() %>% data.frame()


# add locations back in so compatible with plyranges stuff
## this bit only applicable to OPTILES because of the way it structures the ID
ts.gene_optiles <- ts.gene_optiles %>%
  mutate(region_id = gene_id) %>%
  separate_wider_delim(gene_id, delim = ":", names = c("seqnames", "gene_id")) %>%
  mutate(seqnames = gsub("chrchr", "chr", seqnames)) %>%
  separate_wider_delim(gene_id, delim = "-", names = c("start", "end"))
ts.gene_optiles$start <- as.double(ts.gene_optiles$start)
ts.gene_optiles$end <- as.double(ts.gene_optiles$end)
ts.gene_optiles$dtu_rank <- 1:nrow(ts.gene_optiles)
ts.gene_optiles
}
```

```{r}
dtu_tiles_3 <- dtu_function(all_samples, optiles_all, num_groups = 3)
dtu_tiles_4 <- dtu_function(all_samples, optiles_all, num_groups = 4)
```


```{r}
dtu_tiles_4 %>% tail()
dtu.optiles %>% tail()
```


# compare the 3 dtu

```{r}
ggVennDiagram(list(alpha_5 = filter(dtu.optiles, FDR < 0.05)$region_id, 
                   alpha_3 = filter(dtu_tiles_3, FDR < 0.05)$region_id,
                   alpha_4 = filter(dtu_tiles_4, FDR < 0.05)$region_id))
```


```{r}
dmr_tiles <- find_overlaps(as_granges(dmrcate_seq_dmr), as_granges(optiles_all)) %>% data.frame() %>% distinct(region_id)
```

```{r}
ggVennDiagram(list(alpha_5 = filter(dtu.optiles, FDR < 0.05)$region_id, 
                   alpha_3 = filter(dtu_tiles_3, FDR < 0.05)$region_id,
                   alpha_4 = filter(dtu_tiles_4, FDR < 0.05)$region_id,
                   dmr_beta = dmr_tiles$region_id))
```
```{r}
library(UpSetR)
```
[documentation](https://github.com/hms-dbmi/UpSetR)

```{r}
upset(fromList(list(alpha_5 = filter(dtu.optiles, FDR < 0.05)$region_id, 
                   alpha_3 = filter(dtu_tiles_3, FDR < 0.05)$region_id,
                   alpha_4 = filter(dtu_tiles_4, FDR < 0.05)$region_id,
                   dmr_beta = dmr_tiles$region_id)), order.by = "freq")
```

# overlap genes to regions

```{r}
txdb_genes <- genes(TxDb.Hsapiens.UCSC.hg19.knownGene)
```


```{r}
genes_5 <- find_overlaps(txdb_genes, as_granges(dtu.optiles)) %>% data.frame()
genes_4 <- find_overlaps(txdb_genes, as_granges(dtu_tiles_4)) %>% data.frame()
genes_3 <- find_overlaps(txdb_genes, as_granges(dtu_tiles_3)) %>% data.frame()
genes_dmr <- find_overlaps(txdb_genes, as_granges(dmrcate_seq_dmr)) %>% data.frame()
```


```{r}
ggVennDiagram(list(alpha_5 = filter(genes_5, FDR < 0.05)$gene_id, 
                   alpha_3 = filter(genes_3, FDR < 0.05)$gene_id,
                   alpha_4 = filter(genes_4, FDR < 0.05)$gene_id))

```

```{r}
upset(fromList(list(alpha_5 = filter(genes_5, FDR < 0.05)$gene_id, 
                   alpha_3 = filter(genes_3, FDR < 0.05)$gene_id,
                   alpha_4 = filter(genes_4, FDR < 0.05)$gene_id,
                  dmr_beta = genes_dmr$gene_id)), order.by = "freq")
```

# visualise tiles and compare counts in tiles
```{r}
dtu_tiles_3
```


```{r}
dtu_groups_count <- function(all_samples, optiles_all, num_groups = 5) {
  dtu_optile <- find_overlaps(as_granges(all_samples), as_granges(optiles_all)) %>% data.frame() %>%
  distinct(sample, read_id, read_start, read_end, alpha, region_id) 
  
  if(num_groups == 5) {
  dtu_optile <- dtu_optile %>% 
    mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8"))
  } else if(num_groups == 3) {
  dtu_optile <- dtu_optile %>% 
    mutate(alpha_group = case_when(alpha < 0.34 ~ "a < 0.34",
                                alpha > 0.65 ~ "a > 0.65",
                                TRUE ~ "0.34 <= a <= 0.65"))
  } else if(num_groups == 4) {
  dtu_optile <- dtu_optile %>% 
    mutate(alpha_group = case_when(alpha <= 0.25 ~ "a <= 0.25",
                                alpha >= 0.8 ~ "a >= 0.75",
                                alpha > 0.25 & alpha <= 0.5 ~ "0.25 <= a <= 0.5",
                                TRUE ~ "0.5 < a < 0.75"))
}

dtu_optile <- dtu_optile %>% 
  mutate(transcript = paste0(region_id, "_", alpha_group)) %>%
  group_by(sample, transcript) %>%
 summarise(transcript_count = n()) %>% ungroup() %>% data.frame()

# build matrix of "transcript" counts - transcript rows, sample as columns
dtu_optile <- dtu_optile %>% pivot_wider(names_from = sample, values_from = transcript_count)
dtu_optile[is.na(dtu_optile)] <- 0
dtu_optile$gene_id <- sub("_.*", "", dtu_optile$transcript)
dtu_optile$group <- sub(".*_", "", dtu_optile$transcript)
  dtu_optile
}
counts_3 <- dtu_groups_count(all_samples, optiles_all, num_groups = 3)
counts_4 <- dtu_groups_count(all_samples, optiles_all, num_groups = 4)
counts_5 <- dtu_groups_count(all_samples, optiles_all, num_groups = 5)
```


```{r}
counts_3 %>% pivot_longer(cols = c(b1, b2, cml1, cml2)) %>% .[1:500,] %>%
  ggplot(aes(x = group, y = value, fill = name)) +
  geom_boxplot()

counts_3 %>% pivot_longer(cols = c(b1, b2, cml1, cml2)) %>% 
  ggplot(aes(x = group, y = value, fill = name)) +
  geom_boxplot()
```

```{r}
counts_3 %>% pivot_longer(cols = c(b1, b2, cml1, cml2)) %>% filter(value > 4000)

all_samples %>% filter(seqnames == "chr21", start >= 8260801, start <= 8261100) %>% distinct(read_id, meth_pattern, alpha) %>% nrow()
```

```{r}
counts_3 %>% pivot_longer(cols = c(b1, b2, cml1, cml2)) %>% 
  ggplot(aes(x = group, y = log2(value + 1), fill = name)) +
  geom_boxplot()
```


```{r}
counts_4 %>% pivot_longer(cols = c(b1, b2, cml1, cml2)) %>% .[1:500,] %>%
  ggplot(aes(x = group, y = value, fill = name)) +
  geom_boxplot()

counts_4 %>% pivot_longer(cols = c(b1, b2, cml1, cml2)) %>% 
  ggplot(aes(x = group, y = value, fill = name)) +
  geom_boxplot()

counts_4 %>% pivot_longer(cols = c(b1, b2, cml1, cml2)) %>% 
  ggplot(aes(x = group, y = log2(value + 1), fill = name)) +
  geom_boxplot()
```

```{r}
counts_5 %>% pivot_longer(cols = c(b1, b2, cml1, cml2)) %>% .[1:500,] %>%
  ggplot(aes(x = group, y = value, fill = name)) +
  geom_boxplot()

counts_5 %>% pivot_longer(cols = c(b1, b2, cml1, cml2)) %>% 
  ggplot(aes(x = group, y = value, fill = name)) +
  geom_boxplot()

counts_5 %>% pivot_longer(cols = c(b1, b2, cml1, cml2)) %>% 
  ggplot(aes(x = group, y = log2(value + 1), fill = name)) +
  geom_boxplot()
```


```{r}
counts_3$group_size <- "Alpha_3"
counts_4$group_size <- "Alpha_4"
counts_5$group_size <- "Alpha_5"
```

```{r}
rbind(counts_3, counts_4, counts_5) %>% pivot_longer(cols = c(b1, b2, cml1, cml2)) %>%
  ggplot(aes(x = log2(value + 1), fill = group)) +
  geom_density() +
  facet_wrap(~name)
```

```{r}
counts_3 %>% pivot_longer(cols = c(b1, b2, cml1, cml2)) %>% 
  ggplot(aes(x = group, y = log2(value + 1), fill = name)) +
  geom_violin()
counts_4 %>% pivot_longer(cols = c(b1, b2, cml1, cml2)) %>% 
  ggplot(aes(x = group, y = log2(value + 1), fill = name)) +
  geom_violin()
counts_5 %>% pivot_longer(cols = c(b1, b2, cml1, cml2)) %>% 
  ggplot(aes(x = group, y = log2(value + 1), fill = name)) +
  geom_violin()
```

* redo the plots but remove the 0s?

* just remove the group from the tile with a 0
```{r}
counts_3 %>% pivot_longer(cols = c(b1, b2, cml1, cml2)) %>% filter(value != 0) %>%
  ggplot(aes(x = group, y = log2(value + 1), fill = name)) +
  geom_violin()
counts_4 %>% pivot_longer(cols = c(b1, b2, cml1, cml2)) %>% filter(value != 0) %>%
  ggplot(aes(x = group, y = log2(value + 1), fill = name)) +
  geom_violin()
counts_5 %>% pivot_longer(cols = c(b1, b2, cml1, cml2)) %>% filter(value != 0) %>%
  ggplot(aes(x = group, y = log2(value + 1), fill = name)) +
  geom_violin()

counts_3 %>% pivot_longer(cols = c(b1, b2, cml1, cml2)) %>% filter(value != 0) %>%
  ggplot(aes(x = group, y = log2(value + 1), fill = name)) +
  geom_boxplot()
counts_4 %>% pivot_longer(cols = c(b1, b2, cml1, cml2)) %>% filter(value != 0) %>%
  ggplot(aes(x = group, y = log2(value + 1), fill = name)) +
  geom_boxplot()
counts_5 %>% pivot_longer(cols = c(b1, b2, cml1, cml2)) %>% filter(value != 0) %>%
  ggplot(aes(x = group, y = log2(value + 1), fill = name)) +
  geom_boxplot()
```
* remove the tile that has any 0s
```{r}
counts_3 %>% filter(b1 != 0, b2 != 0, cml1 != 0, cml2 != 0) %>% 
  pivot_longer(cols = c(b1, b2, cml1, cml2)) %>%
  ggplot(aes(x = group, y = log2(value + 1), fill = name)) +
  geom_violin()
counts_4 %>% filter(b1 != 0, b2 != 0, cml1 != 0, cml2 != 0) %>% 
  pivot_longer(cols = c(b1, b2, cml1, cml2)) %>%
  ggplot(aes(x = group, y = log2(value + 1), fill = name)) +
  geom_violin()
counts_5 %>% filter(b1 != 0, b2 != 0, cml1 != 0, cml2 != 0) %>% 
  pivot_longer(cols = c(b1, b2, cml1, cml2)) %>%
  ggplot(aes(x = group, y = log2(value + 1), fill = name)) +
  geom_violin()

counts_3 %>% filter(b1 != 0, b2 != 0, cml1 != 0, cml2 != 0) %>% 
  pivot_longer(cols = c(b1, b2, cml1, cml2)) %>%
  ggplot(aes(x = group, y = log2(value + 1), fill = name)) +
  geom_boxplot()
counts_4 %>% filter(b1 != 0, b2 != 0, cml1 != 0, cml2 != 0) %>% 
  pivot_longer(cols = c(b1, b2, cml1, cml2)) %>%
  ggplot(aes(x = group, y = log2(value + 1), fill = name)) +
  geom_boxplot()
counts_5 %>% filter(b1 != 0, b2 != 0, cml1 != 0, cml2 != 0) %>% 
  pivot_longer(cols = c(b1, b2, cml1, cml2)) %>%
  ggplot(aes(x = group, y = log2(value + 1), fill = name)) +
  geom_boxplot()
```


# gene set testing time
* been avoiding this coz gross

* need the genes - txdb object
* which ones de - the ones overlapping
* and the bias - cpgs on genes

```{r}
txdb_genes <- find_overlaps(txdb_genes, as_granges(cg_sites)) %>% data.frame() %>%
  group_by(gene_id) %>% mutate(num_cg = n()) %>% ungroup() %>% distinct()
```

```{r}
genes_3
```


```{r}
source("../code/runGST.R")
gene_try <- genes_3 %>% filter(FDR < 0.05)
gene_info <- txdb_genes %>% mutate(is_de = ifelse(gene_id %in% gene_try$gene_id, 1, 0)) %>% distinct(gene_id, num_cg, is_de, width)

run_test <- list(
  sig.eg = gene_info %>% filter(is_de == 1) %>% .$gene_id,
  universe = gene_info$gene_id,
  freq = gene_info$num_cg,
  de = gene_info$is_de,
  equiv = gene_info$num_cg
)

gst_3 <- run_miss_methyl_b(run_test, method = "Wallenius")
```


```{r}
source("../code/runGST.R")
gene_try <- genes_4 %>% filter(FDR < 0.05)
gene_info <- txdb_genes %>% mutate(is_de = ifelse(gene_id %in% gene_try$gene_id, 1, 0)) %>% distinct(gene_id, num_cg, is_de, width)

run_test <- list(
  sig.eg = gene_info %>% filter(is_de == 1) %>% .$gene_id,
  universe = gene_info$gene_id,
  freq = gene_info$num_cg,
  de = gene_info$is_de,
  equiv = gene_info$num_cg
)

gst_4 <- run_miss_methyl_b(run_test, method = "Wallenius")
```
* shouldn't have tried to run with everything - dumb old mistake
```{r}
gst_4
```

```{r}
source("../code/runGST.R")
gene_try <- genes_5 %>% filter(FDR < 0.05)
gene_info <- txdb_genes %>% mutate(is_de = ifelse(gene_id %in% gene_try$gene_id, 1, 0)) %>% distinct(gene_id, num_cg, is_de, width)

run_test <- list(
  sig.eg = gene_info %>% filter(is_de == 1) %>% .$gene_id,
  universe = gene_info$gene_id,
  freq = gene_info$num_cg,
  de = gene_info$is_de,
  equiv = gene_info$num_cg
)

gst_5 <- run_miss_methyl_b(run_test, method = "Wallenius")
```


```{r}
gene_info %>% distinct(is_de)
```

* try grab top 1k genes: order by dtu then get top 1k 
```{r}
gene_info %>% filter(is_de == 1)
genes_5 %>% filter(FDR < 0.05) %>% distinct(gene_id)
```

```{r}
source("../code/runGST.R")
gene_try <- genes_3 %>% filter(FDR < 0.05) %>% .[order(.$dtu_rank),] %>% distinct(gene_id)
gene_info <- txdb_genes %>% mutate(is_de = ifelse(gene_id %in% gene_try$gene_id, 1, 0)) %>% distinct(gene_id, num_cg, is_de, width)
run_test <- list(
  sig.eg = gene_info %>% filter(is_de == 1) %>% .$gene_id,
  universe = gene_info$gene_id,
  freq = gene_info$num_cg,
  de = gene_info$is_de,
  equiv = gene_info$num_cg
)
gst_3 <- run_miss_methyl_b(run_test, method = "Wallenius")

gene_try <- genes_4 %>% filter(FDR < 0.05) %>% .[order(.$dtu_rank),] %>% distinct(gene_id)
gene_info <- txdb_genes %>% mutate(is_de = ifelse(gene_id %in% gene_try$gene_id, 1, 0)) %>% distinct(gene_id, num_cg, is_de, width)
run_test <- list(
  sig.eg = gene_info %>% filter(is_de == 1) %>% .$gene_id,
  universe = gene_info$gene_id,
  freq = gene_info$num_cg,
  de = gene_info$is_de,
  equiv = gene_info$num_cg
)
gst_4 <- run_miss_methyl_b(run_test, method = "Wallenius")


gene_try <- genes_5 %>% filter(FDR < 0.05) %>% .[order(.$dtu_rank),] %>% distinct(gene_id) 
gene_info <- txdb_genes %>% mutate(is_de = ifelse(gene_id %in% gene_try$gene_id, 1, 0)) %>% distinct(gene_id, num_cg, is_de, width)
run_test <- list(
  sig.eg = gene_info %>% filter(is_de == 1) %>% .$gene_id,
  universe = gene_info$gene_id,
  freq = gene_info$num_cg,
  de = gene_info$is_de,
  equiv = gene_info$num_cg
)
gst_5 <- run_miss_methyl_b(run_test, method = "Wallenius")
```

```{r}
gene_info %>% distinct(is_de)
```


```{r}
source("../code/runGST.R")
gene_try <- genes_3 %>% filter(FDR < 0.05) %>% .[order(.$dtu_rank),] %>% distinct(gene_id) %>% .[1:1000,]
gene_info <- txdb_genes %>% mutate(is_de = ifelse(gene_id %in% gene_try, 1, 0)) %>% distinct(gene_id, num_cg, is_de, width)
run_test <- list(
  sig.eg = gene_info %>% filter(is_de == 1) %>% .$gene_id,
  universe = gene_info$gene_id,
  freq = gene_info$num_cg,
  de = gene_info$is_de,
  equiv = gene_info$num_cg
)
gst_3_2k <- run_miss_methyl_b(run_test, method = "Wallenius")

gene_try <- genes_4 %>% filter(FDR < 0.05) %>% .[order(.$dtu_rank),] %>% distinct(gene_id) %>% .[1:1000,]
gene_info <- txdb_genes %>% mutate(is_de = ifelse(gene_id %in% gene_try, 1, 0)) %>% distinct(gene_id, num_cg, is_de, width)
run_test <- list(
  sig.eg = gene_info %>% filter(is_de == 1) %>% .$gene_id,
  universe = gene_info$gene_id,
  freq = gene_info$num_cg,
  de = gene_info$is_de,
  equiv = gene_info$num_cg
)
gst_4_2k <- run_miss_methyl_b(run_test, method = "Wallenius")


gene_try <- genes_5 %>% filter(FDR < 0.05) %>% .[order(.$dtu_rank),] %>% distinct(gene_id) %>% .[1:1000,]
gene_info <- txdb_genes %>% mutate(is_de = ifelse(gene_id %in% gene_try, 1, 0)) %>% distinct(gene_id, num_cg, is_de, width)
run_test <- list(
  sig.eg = gene_info %>% filter(is_de == 1) %>% .$gene_id,
  universe = gene_info$gene_id,
  freq = gene_info$num_cg,
  de = gene_info$is_de,
  equiv = gene_info$num_cg
)
gst_5_2k <- run_miss_methyl_b(run_test, method = "Wallenius")
```

```{r}
gst_3_2k
```


* gst for dmr
```{r}
source("../code/runGST.R")
gene_try <- genes_dmr %>% filter(FDR < 0.05) %>% .[order(.$dtu_rank),] %>% distinct(gene_id) %>% .[1:1000,]
gene_info <- txdb_genes %>% mutate(is_de = ifelse(gene_id %in% gene_try, 1, 0)) %>% distinct(gene_id, num_cg, is_de, width)
run_test <- list(
  sig.eg = gene_info %>% filter(is_de == 1) %>% .$gene_id,
  universe = gene_info$gene_id,
  freq = gene_info$num_cg,
  de = gene_info$is_de,
  equiv = gene_info$num_cg
)
gst_3_2k <- run_miss_methyl_b(run_test, method = "Wallenius")
```



```{r}
gst_3_2k[[1]]
```


```{r}
ggVennDiagram(list(gst_5 = gst_5_2k[[1]][1:100,]$GOID, 
                   gst_3 = gst_3_2k[[1]][1:100,]$GOID,
                   gst_4 = gst_4_2k[[1]][1:100,]$GOID))
```


## individual group level testing for dtu


```{r}
dtu_function_exon <- function(all_samples, optiles_all, num_groups = 5) {
  dtu_optile <- find_overlaps(as_granges(all_samples), as_granges(optiles_all)) %>% data.frame() %>%
  distinct(sample, read_id, read_start, read_end, alpha, region_id) 
  
  if(num_groups == 5) {
  dtu_optile <- dtu_optile %>% 
    mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8"))
  } else if(num_groups == 3) {
  dtu_optile <- dtu_optile %>% 
    mutate(alpha_group = case_when(alpha < 0.34 ~ "a < 0.34",
                                alpha > 0.65 ~ "a > 0.65",
                                TRUE ~ "0.34 <= a <= 0.65"))
  } else if(num_groups == 4) {
  dtu_optile <- dtu_optile %>% 
    mutate(alpha_group = case_when(alpha <= 0.25 ~ "a <= 0.25",
                                alpha >= 0.8 ~ "a >= 0.75",
                                alpha > 0.25 & alpha <= 0.5 ~ "0.25 <= a <= 0.5",
                                TRUE ~ "0.5 < a < 0.75"))
}

dtu_optile <- dtu_optile %>% 
  mutate(transcript = paste0(region_id, "_", alpha_group)) %>%
  group_by(sample, transcript) %>%
 summarise(transcript_count = n()) %>% ungroup() %>% data.frame()

# build matrix of "transcript" counts - transcript rows, sample as columns
dtu_optile <- dtu_optile %>% pivot_wider(names_from = sample, values_from = transcript_count)
dtu_optile[is.na(dtu_optile)] <- 0
dtu_optile$gene_id <- sub("_.*", "", dtu_optile$transcript)

# DTU analysis with edgeR
## Alicia prefers gene method over Simes method (documentation slightly confusing about difference)
y <- DGEList(as.matrix(dtu_optile[,2:5]), group = c(1,1,2,2), genes = dtu_optile[,c("gene_id", "transcript")])
design <- model.matrix(~ c(1,1,2,2), data = y$samples)
fit <- glmQLFit(y, design, robust=TRUE)
ds <- diffSpliceDGE(fit, coef = 2, geneid = "gene_id", exonid = "transcript")
ts.gene_optiles <- topSpliceDGE(ds, test = "exon", number = Inf)

# add locations back in so compatible with plyranges stuff
## this bit only applicable to OPTILES because of the way it structures the ID
ts.gene_optiles <- ts.gene_optiles %>%
  mutate(region_id = gene_id) %>%
  separate_wider_delim(gene_id, delim = ":", names = c("seqnames", "gene_id")) %>%
  mutate(seqnames = gsub("chrchr", "chr", seqnames)) %>%
  separate_wider_delim(gene_id, delim = "-", names = c("start", "end")) %>% 
  mutate(region_and_transcript = transcript, transcript = sub(".*_", "", transcript))
  
ts.gene_optiles$start <- as.double(ts.gene_optiles$start)
ts.gene_optiles$end <- as.double(ts.gene_optiles$end)
ts.gene_optiles$dtu_rank <- 1:nrow(ts.gene_optiles)
ts.gene_optiles <- data.frame(ts.gene_optiles)
ts.gene_optiles
}
```

```{r}
dtu_tiles_3_exon <- dtu_function_exon(all_samples, optiles_all, num_groups = 3)
dtu_tiles_4_exon <- dtu_function_exon(all_samples, optiles_all, num_groups = 4)
dtu_tiles_5_exon <- dtu_function_exon(all_samples, optiles_all, num_groups = 5)
```

```{r}
dtu_tiles_5_exon
```

```{r}
ggVennDiagram(list(alpha_5 = filter(dtu_tiles_5_exon, FDR < 0.05)$region_and_transcript, 
                   alpha_3 = filter(dtu_tiles_3_exon, FDR < 0.05)$region_and_transcript,
                   alpha_4 = filter(dtu_tiles_4_exon, FDR < 0.05)$region_and_transcript))
```

```{r}
ggVennDiagram(list(alpha_5 = filter(dtu_tiles_5_exon, FDR < 0.05)$region_id, 
                   alpha_3 = filter(dtu_tiles_3_exon, FDR < 0.05)$region_id,
                   alpha_4 = filter(dtu_tiles_4_exon, FDR < 0.05)$region_id))
```

```{r}
dtu_tiles_3_exon %>% filter(FDR < 0.05) %>% 
  ggplot(aes(x = transcript, fill = transcript)) +
  geom_bar() +
  theme(legend.position = "none")

dtu_tiles_4_exon %>% filter(FDR < 0.05) %>% 
  ggplot(aes(x = transcript, fill = transcript)) +
  geom_bar() +
  theme(legend.position = "none")

dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% 
  ggplot(aes(x = transcript, fill = transcript)) +
  geom_bar() +
  theme(legend.position = "none")
```

```{r}
dtu_tiles_3_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% summarise(n_groups = n()) %>%
  ggplot(aes(x = n_groups, fill = as.factor(n_groups))) +
  geom_bar() +
  theme(legend.position = "none")

dtu_tiles_4_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% summarise(n_groups = n()) %>%
  ggplot(aes(x = n_groups, fill = as.factor(n_groups))) +
  geom_bar() +
  theme(legend.position = "none")

dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% summarise(n_groups = n()) %>%
  ggplot(aes(x = n_groups, fill = as.factor(n_groups))) +
  geom_bar() +
  theme(legend.position = "none")
```
```{r}
dtu_tiles_3_exon <- data.frame(dtu_tiles_3_exon)
```


```{r}
genes_5_exon <- find_overlaps(as_granges(txdb_genes), as_granges(dtu_tiles_5_exon)) %>% data.frame()
genes_4_exon <- find_overlaps(as_granges(txdb_genes), as_granges(dtu_tiles_4_exon)) %>% data.frame()
genes_3_exon <- find_overlaps(as_granges(txdb_genes), as_granges(dtu_tiles_3_exon)) %>% data.frame()
```

```{r}
genes_3_exon %>% filter(FDR < 0.05) %>% 
  ggplot(aes(x = transcript, fill = transcript)) +
  geom_bar() +
  theme(legend.position = "none")

genes_4_exon %>% filter(FDR < 0.05) %>% 
  ggplot(aes(x = transcript, fill = transcript)) +
  geom_bar() +
  theme(legend.position = "none")

genes_5_exon %>% filter(FDR < 0.05) %>% 
  ggplot(aes(x = transcript, fill = transcript)) +
  geom_bar() +
  theme(legend.position = "none")

```

```{r}
dtu_tiles_3_exon %>% mutate(overlap_gene = ifelse(region_and_transcript %in% genes_3_exon$region_and_transcript, TRUE, FALSE)) %>% 
  filter(FDR < 0.05) %>% 
  ggplot(aes(x = transcript, fill = overlap_gene)) +
  geom_bar() 

dtu_tiles_4_exon %>% mutate(overlap_gene = ifelse(region_and_transcript %in% genes_4_exon$region_and_transcript, TRUE, FALSE)) %>% 
  filter(FDR < 0.05) %>% 
  ggplot(aes(x = transcript, fill = overlap_gene)) +
  geom_bar() 

dtu_tiles_5_exon %>% mutate(overlap_gene = ifelse(region_and_transcript %in% genes_5_exon$region_and_transcript, TRUE, FALSE)) %>% 
  filter(FDR < 0.05) %>% 
  ggplot(aes(x = transcript, fill = overlap_gene)) +
  geom_bar() 
```

```{r}
dtu_tiles_3_exon %>% mutate(overlap_gene = ifelse(region_and_transcript %in% genes_3_exon$region_and_transcript, TRUE, FALSE)) %>% 
  filter(FDR < 0.05) %>% 
  ggplot(aes(x = transcript, fill = overlap_gene)) +
  geom_bar(position = "fill") 

dtu_tiles_4_exon %>% mutate(overlap_gene = ifelse(region_and_transcript %in% genes_4_exon$region_and_transcript, TRUE, FALSE)) %>% 
  filter(FDR < 0.05) %>% 
  ggplot(aes(x = transcript, fill = overlap_gene)) +
  geom_bar(position = "fill") 

dtu_tiles_5_exon %>% mutate(overlap_gene = ifelse(region_and_transcript %in% genes_5_exon$region_and_transcript, TRUE, FALSE)) %>% 
  filter(FDR < 0.05) %>% 
  ggplot(aes(x = transcript, fill = overlap_gene)) +
  geom_bar(position = "fill") 
```

* no groups are more likely to be overlapping gene than others


* are the big/small more likely to overlap dmr??
```{r}
dmrcate_seq_dmr$dmr_rank <- 1:nrow(dmrcate_seq_dmr)
```

```{r}
dmr_5_exon <- find_overlaps(as_granges(dmrcate_seq_dmr), as_granges(dtu_tiles_5_exon)) %>% data.frame()
dmr_4_exon <- find_overlaps(as_granges(dmrcate_seq_dmr), as_granges(dtu_tiles_4_exon)) %>% data.frame()
dmr_3_exon <- find_overlaps(as_granges(dmrcate_seq_dmr), as_granges(dtu_tiles_3_exon)) %>% data.frame()
```


```{r}
dtu_tiles_3_exon %>% mutate(overlap_dmr = ifelse(region_and_transcript %in% dmr_3_exon$region_and_transcript, TRUE, FALSE)) %>% 
  filter(FDR < 0.05) %>% 
  ggplot(aes(x = transcript, fill = overlap_dmr)) +
  geom_bar() 

dtu_tiles_4_exon %>% mutate(overlap_dmr = ifelse(region_and_transcript %in% dmr_4_exon$region_and_transcript, TRUE, FALSE)) %>% 
  filter(FDR < 0.05) %>% 
  ggplot(aes(x = transcript, fill = overlap_dmr)) +
  geom_bar() 

dtu_tiles_5_exon %>% mutate(overlap_dmr = ifelse(region_and_transcript %in% dmr_5_exon$region_and_transcript, TRUE, FALSE)) %>% 
  filter(FDR < 0.05) %>% 
  ggplot(aes(x = transcript, fill = overlap_dmr)) +
  geom_bar() 
```

```{r}
dtu_tiles_3_exon %>% mutate(overlap_dmr = ifelse(region_and_transcript %in% dmr_3_exon$region_and_transcript, TRUE, FALSE)) %>% 
  filter(FDR < 0.05) %>% 
  ggplot(aes(x = transcript, fill = overlap_dmr)) +
  geom_bar(position = "fill") 

dtu_tiles_4_exon %>% mutate(overlap_dmr = ifelse(region_and_transcript %in% dmr_4_exon$region_and_transcript, TRUE, FALSE)) %>% 
  filter(FDR < 0.05) %>% 
  ggplot(aes(x = transcript, fill = overlap_dmr)) +
  geom_bar(position = "fill") 

dtu_tiles_5_exon %>% mutate(overlap_dmr = ifelse(region_and_transcript %in% dmr_5_exon$region_and_transcript, TRUE, FALSE)) %>% 
  filter(FDR < 0.05) %>% 
  ggplot(aes(x = transcript, fill = overlap_dmr)) +
  geom_bar(position = "fill") 
```

* interesting, the smaller changes have a higher proportion of overlap to dmrs than the bigger changes
* expected the opposite
* but also it's not that much of a signal




# dmrs that don't overlap dtu

```{r}
dmrcate_seq_dmr %>% filter(!dmr_rank %in% dmr_5_exon$dmr_rank)
```


# add info to results: get proportion counts
gene_cg <- find_overlaps(as_granges(optiles_all), as_granges(cg_sites)) %>% data.frame() %>%
  group_by(region_id) %>%
  summarise(n_cg = n()) %>%
  ungroup() %>% data.frame()
fit$genes <- fit$genes %>% mutate(n_cg = gene_cg[match(.$gene_id, gene_cg$region_id), "n_cg"])
rpk <- y$counts/rep(fit$genes$n_cg/1000, times = ncol(y$counts))
tpm_mat <- sweep(rpk, 2, colSums(rpk)/1e6, "/")
tpm_mat <- tpm_mat %>% data.frame() %>% mutate(id = dtu_optile$transcript, gene_id = dtu_optile$gene_id)

tpm_mat <- tpm_mat %>%
  group_by(gene_id) %>%
  mutate(prop_S1_gm = b1/sum(b1),
         prop_S2_gm = b2/sum(b2),
         prop_S3_k = cml1/sum(cml1),
         prop_S4_k = cml2/sum(cml2)) %>%
  ungroup() %>% data.frame()
