---
title: "dtuResultsChangeInSignifInterpret"
author: "caitlinpage"
date: "2026-07-01"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

## Introduction
```{r}
library(annotatr)
library(data.table)
library(plyranges)
library(tidyr)
library(stringr)
library(dplyr)
library(ggplot2)
library(edgeR)
library(OpTiles)
library(ggVennDiagram)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
library(missMethyl)
library(UpSetR)
```

```{r}
seqs <- seqnames(BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38)[1:22]

cg_sites <- lapply(seqs, function(x) {
        cbind(data.frame(Biostrings::matchPattern("CG", BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38[[x]])),
            seqnames = x
        )}) %>% bind_rows()
```


```{r}
all_samples <- readRDS("/researchers/caitlin.page/cf_nano/r_output/all_rrbs_samples.rds")
optiles_all <- readRDS("/researchers/caitlin.page/cf_nano/r_output/optiles_all.rds")
dtu.optiles <- readRDS("/researchers/caitlin.page/cf_nano/r_output/ts.gene_optiles.rds")
dmrcate_seq_dmr <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dmr_encode_gm12878_k562.rds")
```

```{r}
all_samples_simp <- all_samples %>% distinct(seqnames, read_start, read_end, strand, sample, read_id, alpha, meth_pattern, num_meth, total)
colnames(all_samples_simp)[2:3] <- c("start", "end")
```

```{r}
dtu_tiles_5_exon <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_5_exon.rds")
```


```{r}
bin_delta <- find_overlaps(as_granges(all_samples_simp), as_granges(optiles_all)) %>% data.frame() %>%
  distinct(sample, read_id, start, end, alpha, region_id) %>% 
    mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8")) %>%
  mutate(transcript = paste0(region_id, "_", alpha_group)) %>% mutate(sample_type = ifelse(sample %in% c("b1", "b2"), "B", "CML")) %>%
  group_by(region_id, sample_type) %>% mutate(mean_alpha = mean(alpha)) %>% ungroup() %>% distinct(region_id, sample_type, mean_alpha) %>% .[order(.$region_id),]

bin_delta <- bin_delta %>% pivot_wider(names_from = sample_type, values_from = mean_alpha) %>% mutate(delta_b_minus_cml = B - CML)
bin_delta <- data.frame(bin_delta)
```


####

# Analysis


```{r}
overlap2 <- find_overlaps(as_granges(filter(dtu_tiles_5_exon, FDR < 0.05)), as_granges(dmrcate_seq_dmr)) %>% data.frame()
```

```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>%
  ggplot(aes(x = overlap_dmr, fill = transcript)) +
  geom_bar()

dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>%
  ggplot(aes(x = overlap_dmr, fill = transcript)) +
  geom_bar(position = "fill")
```

```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>%
  ggplot(aes(x = transcript, fill = overlap_dmr)) +
  geom_bar()

dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>%
  ggplot(aes(x = transcript, fill = overlap_dmr)) +
  geom_bar(position = "fill")
```

```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>%
  filter(transcript == "0.4 < a < 0.6", overlap_dmr == FALSE)  
```

## dmrs and alpha significance

```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>%
  ggplot(aes(x = dtu_rank, y = transcript, colour = overlap_dmr)) +
  geom_jitter(alpha = 0.3)
```

```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>%
  ggplot(aes(x = dtu_rank, y = transcript, colour = overlap_dmr)) +
  geom_jitter(alpha = 0.3) +
  scale_x_continuous(limits = c(0, 20000))
```

```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>%
  filter(transcript == "0.4 < a < 0.6", overlap_dmr == FALSE, dtu_rank <= 20000) %>% nrow()
```

```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>%
  filter(transcript == "0.4 < a < 0.6", overlap_dmr == FALSE, dtu_rank <= 20000) %>% .[1:5,]
```


* ok what are the betas for these regions

```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>%
  filter(transcript == "0.4 < a < 0.6", overlap_dmr == FALSE) %>% .[1:5,]
```


```{r}
betas <- all_samples %>% distinct(seqnames, start, sample, beta) %>% pivot_wider(names_from = sample, values_from = beta)

betas <- betas %>% mutate(mean_b = rowMeans(across(c(b1, b2)), na.rm = TRUE), 
                 mean_cml = rowMeans(across(c(cml1, cml2)), na.rm = TRUE),
                 beta_delta_b_minus_cml = mean_b - mean_cml) %>% data.frame()
betas$end <- betas$start

betas$pos <- paste0(betas$seqnames, "-", betas$start)
```

```{r}
find_overlaps(dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>%
  filter(transcript == "0.4 < a < 0.6", overlap_dmr == FALSE) %>% as_granges(), as_granges(betas)) %>% data.frame() %>%
  ggplot(aes(x = mean_b)) +
  geom_density() +
  geom_density(aes(x = mean_cml), colour = "red")
```

```{r}
find_overlaps(dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>%
  filter(overlap_dmr == FALSE) %>% as_granges(), as_granges(betas)) %>% data.frame() %>%
  ggplot(aes(x = mean_b)) +
  geom_density() +
  geom_density(aes(x = mean_cml), colour = "red")

# beta distrib of tiles that are signif +  overlap dmr 
find_overlaps(dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>%
  filter(overlap_dmr == TRUE) %>% as_granges(), as_granges(betas)) %>% data.frame() %>%
  ggplot(aes(x = mean_b)) +
  geom_density() +
  geom_density(aes(x = mean_cml), colour = "red")
```



* not sure what this says

```{r}
overlap_beta <- find_overlaps(dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>% as_granges(), as_granges(betas)) %>% data.frame()

betas %>% mutate(signif = ifelse(pos %in% overlap_beta$pos, TRUE, FALSE)) %>%
  ggplot(aes(x = mean_b, colour = signif)) +
  geom_density() +
  geom_density(aes(x = mean_cml, colour = signif), linetype = "dashed")
```

```{r}
find_overlaps(dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>% as_granges(), as_granges(betas)) %>% data.frame() %>% mutate(mid_alpha = ifelse(transcript == "0.4 < a < 0.6", TRUE, FALSE)) %>%
  ggplot(aes(x = mean_b, colour = mid_alpha)) +
  geom_density() +
  facet_wrap(~overlap_dmr)
```

```{r}
find_overlaps(dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>% as_granges(), as_granges(betas)) %>% data.frame() %>% mutate(mid_alpha = ifelse(transcript == "0.4 < a < 0.6", TRUE, FALSE)) %>%
  ggplot(aes(x = mean_b, y = mean_cml, colour = transcript)) +
  geom_point(alpha = 0.1)
```

```{r}
find_overlaps(dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>% as_granges(), as_granges(betas)) %>% data.frame() %>% mutate(mid_alpha = ifelse(transcript == "0.4 < a < 0.6", TRUE, FALSE)) %>%
  ggplot(aes(x = mean_b, y = mean_cml, colour = transcript)) +
  geom_point(alpha = 0.1) +
  facet_wrap(~ transcript)
```

```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% mutate(n_trans = n()) %>% ungroup() %>%
  ggplot(aes(x = transcript, y = n_trans)) +
  geom_boxplot() +
  geom_violin()
```

```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% mutate(n_trans = n(), prop = 1/n()) %>% ungroup() %>% .[order(.$region_id),] %>% .[1:5,]
```

* I want how common is each like transcript combo - like if low is sig - what prop of that has the others as sig - in same 
* ok not quite there - but got the proportion in a region of the thing

```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% mutate(n_trans = n(), prop = 1/n()) %>% ungroup() %>% .[order(.$region_id),] %>%
  ggplot(aes(x = transcript, fill = as.factor(prop))) +
  geom_bar()

dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% mutate(n_trans = n(), prop = 1/n()) %>% ungroup() %>% .[order(.$region_id),] %>%
  ggplot(aes(x = transcript, fill = as.factor(prop))) +
  geom_bar(position = "fill")
```


```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% mutate(n_trans = n(), prop = 1/n()) %>% ungroup() %>% .[order(.$region_id),] %>% 
  group_by(region_id) %>% filter(any(transcript == "0.4 < a < 0.6")) %>%
  ggplot(aes(x = transcript)) +
  geom_bar()
```
nbm
```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% mutate(n_trans = n(), prop = 1/n()) %>% ungroup() %>% .[order(.$region_id),] %>% 
  group_by(region_id) %>% mutate(has_mid = ifelse(any(transcript == "0.4 < a < 0.6"), TRUE, FALSE)) %>%
  ggplot(aes(x = transcript, fill = has_mid)) +
  geom_bar()
  
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% mutate(n_trans = n(), prop = 1/n()) %>% ungroup() %>% .[order(.$region_id),] %>% 
  group_by(region_id) %>% mutate(has_mid = ifelse(any(transcript == "0.4 < a < 0.6"), TRUE, FALSE)) %>%
  ggplot(aes(x = has_mid, fill = transcript)) +
  geom_bar()
```

```{r}
#dtu_tiles_5_exon %>% distinct(region_id, transcript) %>%
#  count(transcript, name = "group_count")
```


```{r}
betas %>%
  ggplot(aes(x = mean_b)) +
  geom_density()

betas %>% filter(mean_b == 0.5) %>% nrow()
```


```{r}
optiles_all_beta <- find_overlaps(as_granges(optiles_all), as_granges(betas)) %>% data.frame()
optiles_all_beta <- optiles_all_beta %>% mutate(alpha_b = bin_delta[match(.$region_id, bin_delta$region_id), "B"],
                                                alpha_cml = bin_delta[match(.$region_id, bin_delta$region_id), "CML"],
                                                alpha_delta_b_minus_cml = bin_delta[match(.$region_id, bin_delta$region_id), "delta_b_minus_cml"],
                                                alpha_b = as.double(alpha_b), alpha_cml = as.double(alpha_cml), alpha_delta_b_minus_cml = as.double(alpha_delta_b_minus_cml))
```

```{r}
interest <- dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>%
  filter(transcript == "0.4 < a < 0.6", overlap_dmr == FALSE) %>% .$region_id
```

```{r}
optiles_all_beta %>% mutate(mid_not_sig_in_dmr = ifelse(region_id %in% interest, TRUE, FALSE)) %>%
  filter(mid_not_sig_in_dmr == TRUE) %>%
  ggplot(aes(x = mean_b, y = mean_cml)) +
  geom_point(alpha = 0.2)

optiles_all_beta %>% mutate(mid_not_sig_in_dmr = ifelse(region_id %in% interest, TRUE, FALSE)) %>%
  filter(mid_not_sig_in_dmr == FALSE) %>%
  ggplot(aes(x = mean_b, y = mean_cml)) +
  geom_point(alpha = 0.2)
```

```{r}
optiles_all_beta %>% mutate(mid_not_sig_in_dmr = ifelse(region_id %in% interest, TRUE, FALSE)) %>%
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(mid_beta = ifelse(mean_b >= 0.4 & mean_b <= 0.6 | mean_cml >= 0.4 & mean_cml <= 0.6, TRUE, FALSE)) %>%
  ggplot(aes(x = mid_not_sig_in_dmr, fill = mid_beta)) +
  geom_bar()

optiles_all_beta %>% mutate(mid_not_sig_in_dmr = ifelse(region_id %in% interest, TRUE, FALSE)) %>%
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(mid_beta = ifelse(mean_b >= 0.4 & mean_b <= 0.6 | mean_cml >= 0.4 & mean_cml <= 0.6, TRUE, FALSE)) %>%
  ggplot(aes(x = mid_not_sig_in_dmr, fill = mid_beta)) +
  geom_bar(position = "fill")
```

```{r}
optiles_all_beta %>% mutate(mid_not_sig_in_dmr = ifelse(region_id %in% interest, TRUE, FALSE)) %>%
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(beta_group_b = case_when(mean_b > 0.4 & mean_b < 0.6 ~ "mid:0.4-0.6", 
                                mean_b < 0.2 ~ "low:<0.2",
                                mean_b > 0.8 ~ "high:>0.8",
                                mean_b >= 0.2 & mean_b <= 0.4 ~ "low-med:0.2-0.4",
                                TRUE ~ "med-high:0.6-0.8")) %>%
  mutate(beta_group_cml = case_when(mean_cml > 0.4 & mean_cml < 0.6 ~ "mid:0.4-0.6", 
                                mean_cml < 0.2 ~ "low:<0.2",
                                mean_cml > 0.8 ~ "high:>0.8",
                                mean_cml >= 0.2 & mean_cml <= 0.4 ~ "low-med:0.2-0.4",
                                TRUE ~ "med-high:0.6-0.8")) %>%
  ggplot(aes(x = beta_group_b, y = beta_group_cml, colour = mid_not_sig_in_dmr)) +
  geom_count()
```

```{r}
optiles_all_beta %>% mutate(mid_not_sig_in_dmr = ifelse(region_id %in% interest, TRUE, FALSE)) %>%
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(beta_group_b = case_when(mean_b > 0.4 & mean_b < 0.6 ~ "mid:0.4-0.6", 
                                mean_b < 0.2 ~ "low:<0.2",
                                mean_b > 0.8 ~ "high:>0.8",
                                mean_b >= 0.2 & mean_b <= 0.4 ~ "low-med:0.2-0.4",
                                TRUE ~ "med-high:0.6-0.8")) %>%
  mutate(beta_group_cml = case_when(mean_cml > 0.4 & mean_cml < 0.6 ~ "mid:0.4-0.6", 
                                mean_cml < 0.2 ~ "low:<0.2",
                                mean_cml > 0.8 ~ "high:>0.8",
                                mean_cml >= 0.2 & mean_cml <= 0.4 ~ "low-med:0.2-0.4",
                                TRUE ~ "med-high:0.6-0.8")) %>%
  pivot_longer(cols = c(beta_group_b, beta_group_cml), names_to = "type", values_to = "beta_group") %>%
  ggplot(aes(x = mid_not_sig_in_dmr, fill = beta_group)) +
  geom_bar()

optiles_all_beta %>% mutate(mid_not_sig_in_dmr = ifelse(region_id %in% interest, TRUE, FALSE)) %>%
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(beta_group_b = case_when(mean_b > 0.4 & mean_b < 0.6 ~ "mid:0.4-0.6", 
                                mean_b < 0.2 ~ "low:<0.2",
                                mean_b > 0.8 ~ "high:>0.8",
                                mean_b >= 0.2 & mean_b <= 0.4 ~ "low-med:0.2-0.4",
                                TRUE ~ "med-high:0.6-0.8")) %>%
  mutate(beta_group_cml = case_when(mean_cml > 0.4 & mean_cml < 0.6 ~ "mid:0.4-0.6", 
                                mean_cml < 0.2 ~ "low:<0.2",
                                mean_cml > 0.8 ~ "high:>0.8",
                                mean_cml >= 0.2 & mean_cml <= 0.4 ~ "low-med:0.2-0.4",
                                TRUE ~ "med-high:0.6-0.8")) %>%
  pivot_longer(cols = c(beta_group_b, beta_group_cml), names_to = "type", values_to = "beta_group") %>%
  ggplot(aes(x = mid_not_sig_in_dmr, fill = beta_group)) +
  geom_bar(position = "fill")
```

```{r}
optiles_all_beta %>% mutate(mid_not_sig_in_dmr = ifelse(region_id %in% interest, TRUE, FALSE)) %>%
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(beta_group_b = case_when(mean_b > 0.4 & mean_b < 0.6 ~ "mid:0.4-0.6", 
                                mean_b < 0.2 ~ "low:<0.2",
                                mean_b > 0.8 ~ "high:>0.8",
                                mean_b >= 0.2 & mean_b <= 0.4 ~ "low-med:0.2-0.4",
                                TRUE ~ "med-high:0.6-0.8")) %>%
  mutate(beta_group_cml = case_when(mean_cml > 0.4 & mean_cml < 0.6 ~ "mid:0.4-0.6", 
                                mean_cml < 0.2 ~ "low:<0.2",
                                mean_cml > 0.8 ~ "high:>0.8",
                                mean_cml >= 0.2 & mean_cml <= 0.4 ~ "low-med:0.2-0.4",
                                TRUE ~ "med-high:0.6-0.8")) %>%
  pivot_longer(cols = c(beta_group_b, beta_group_cml), names_to = "type", values_to = "beta_group") %>%
  ggplot(aes(x = beta_group, fill = mid_not_sig_in_dmr)) +
  geom_bar()

optiles_all_beta %>% mutate(mid_not_sig_in_dmr = ifelse(region_id %in% interest, TRUE, FALSE)) %>%
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(beta_group_b = case_when(mean_b > 0.4 & mean_b < 0.6 ~ "mid:0.4-0.6", 
                                mean_b < 0.2 ~ "low:<0.2",
                                mean_b > 0.8 ~ "high:>0.8",
                                mean_b >= 0.2 & mean_b <= 0.4 ~ "low-med:0.2-0.4",
                                TRUE ~ "med-high:0.6-0.8")) %>%
  mutate(beta_group_cml = case_when(mean_cml > 0.4 & mean_cml < 0.6 ~ "mid:0.4-0.6", 
                                mean_cml < 0.2 ~ "low:<0.2",
                                mean_cml > 0.8 ~ "high:>0.8",
                                mean_cml >= 0.2 & mean_cml <= 0.4 ~ "low-med:0.2-0.4",
                                TRUE ~ "med-high:0.6-0.8")) %>%
  pivot_longer(cols = c(beta_group_b, beta_group_cml), names_to = "type", values_to = "beta_group") %>%
  ggplot(aes(x = beta_group, fill = mid_not_sig_in_dmr)) +
  geom_bar(position = "fill")
```

mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8"))


filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(mid_beta = ifelse(mean_b >= 0.4 & mean_b <= 0.6 | mean_cml >= 0.4 & mean_cml <= 0.6, TRUE, FALSE)) %>%
  group_by(region_id) %>% mutate(has_mid_beta = any(mid_beta == TRUE)) %>% ungroup() %>% distinct(region_id, sig_dmr, sig_dtu, has_mid_beta) %>%
  mutate(sig_dmr = ifelse(sig_dmr == TRUE, "dmr_sig", "dmr_not_sig"), sig_dtu = ifelse(sig_dtu == TRUE, "dtu_sig", "dtu_not_sig")) %>%
  pivot_longer(cols = c(sig_dmr, sig_dtu), names_to = "sig", values_to = "is_sig") %>% .[,c(1,2,4)] %>% distinct() %>%
  ggplot(aes(x = is_sig, fill = has_mid_beta)) +
  geom_bar(position = "fill")


```{r}
overlap2[1:5,]
```


```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>% .[1:5,]
```


## sense check
* Do the regions look how I think they should? - if 2 groups signif, is it each condition has 1 of those inflated?
* Visualise region, and do proportion plot of counts


* want a region that is significant in mid range + significant in an extreme


```{r}
dtu_tiles_5_exon %>% group_by(region_id) %>% filter(setequal(unique(transcript), c("a < 0.2", "0.4 < a < 0.6"))) %>% .[order(.$region_id),] %>% .[1:5,]
```

* this is all of them
```{r}
dtu_tiles_5_exon %>% group_by(region_id) %>% filter(setequal(unique(transcript), c("a < 0.2", "0.4 < a < 0.6"))) %>% .[order(.$region_id),] %>% .[1,]
```

```{r}
reads <- find_overlaps(dtu_tiles_5_exon %>% group_by(region_id) %>% filter(setequal(unique(transcript), c("a < 0.2", "0.4 < a < 0.6"))) %>% .[order(.$region_id),] %>% .[1,] %>% as_granges(), as_granges(all_samples_simp)) %>% data.frame() %>% .$read_id
```


```{r}
source("../code/processData.R")

rrbs_plot4(all_samples, reads)
```
* ok oops need a minimum sample situation

* oh wait are any of these significant in FDR?

```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% filter(setequal(unique(transcript), c("a < 0.2", "0.4 < a < 0.6"))) %>% .[order(.$dtu_rank, decreasing = FALSE),] %>% .[1:5,]
```
# Question

Can I identify a change in signal -> noise, low -> high and visualise it?

Why do I want to do that? 
Because it would add detail to the results, and knowing direction of change is good
And means summarise things a lot
So it would be like delta - but more descriptive

## visualise top 5 ranked regions where mid range and low are the only significant categories

```{r}
source("../code/processData.R")
reads <- find_overlaps(dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% filter(setequal(unique(transcript), c("a < 0.2", "0.4 < a < 0.6"))) %>% .[order(.$dtu_rank, decreasing = FALSE),] %>% .[1,] %>% as_granges(), as_granges(all_samples_simp)) %>% data.frame() %>% .$read_id
rrbs_plot4(all_samples, reads)

reads <- find_overlaps(dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% filter(setequal(unique(transcript), c("a < 0.2", "0.4 < a < 0.6"))) %>% .[order(.$dtu_rank, decreasing = FALSE),] %>% .[2,] %>% as_granges(), as_granges(all_samples_simp)) %>% data.frame() %>% .$read_id
rrbs_plot4(all_samples, reads)

reads <- find_overlaps(dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% filter(setequal(unique(transcript), c("a < 0.2", "0.4 < a < 0.6"))) %>% .[order(.$dtu_rank, decreasing = FALSE),] %>% .[3,] %>% as_granges(), as_granges(all_samples_simp)) %>% data.frame() %>% .$read_id
rrbs_plot4(all_samples, reads)

reads <- find_overlaps(dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% filter(setequal(unique(transcript), c("a < 0.2", "0.4 < a < 0.6"))) %>% .[order(.$dtu_rank, decreasing = FALSE),] %>% .[4,] %>% as_granges(), as_granges(all_samples_simp)) %>% data.frame() %>% .$read_id
rrbs_plot4(all_samples, reads)

reads <- find_overlaps(dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% filter(setequal(unique(transcript), c("a < 0.2", "0.4 < a < 0.6"))) %>% .[order(.$dtu_rank, decreasing = FALSE),] %>% .[5,] %>% as_granges(), as_granges(all_samples_simp)) %>% data.frame() %>% .$read_id
rrbs_plot4(all_samples, reads)
```

* ok so some are visually clearer
* 4 is very visually clear
* but in the others you can see a change in density from a mix to a lower alpha - in most this is lower in cml, in the first its lower in b

```{r}
dtu_tiles_5_exon %>% distinct(transcript) %>% .[1:5,]
```

* may need coverage involved as well

## visualise top 5 where mid range and high are only significant categories
```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% filter(setequal(unique(transcript), c("a > 0.8", "0.4 < a < 0.6"))) %>% .[order(.$dtu_rank, decreasing = FALSE),] %>% .[1:10,]
```

```{r}
source("../code/processData.R")
reads <- find_overlaps(dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% filter(setequal(unique(transcript), c("a > 0.8", "0.4 < a < 0.6"))) %>% .[order(.$dtu_rank, decreasing = FALSE),] %>% .[1,] %>% as_granges(), as_granges(all_samples_simp)) %>% data.frame() %>% .$read_id
rrbs_plot4(all_samples, reads)

reads <- find_overlaps(dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% filter(setequal(unique(transcript), c("a > 0.8", "0.4 < a < 0.6"))) %>% .[order(.$dtu_rank, decreasing = FALSE),] %>% .[2,] %>% as_granges(), as_granges(all_samples_simp)) %>% data.frame() %>% .$read_id
rrbs_plot4(all_samples, reads)

reads <- find_overlaps(dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% filter(setequal(unique(transcript), c("a > 0.8", "0.4 < a < 0.6"))) %>% .[order(.$dtu_rank, decreasing = FALSE),] %>% .[3,] %>% as_granges(), as_granges(all_samples_simp)) %>% data.frame() %>% .$read_id
rrbs_plot4(all_samples, reads)

reads <- find_overlaps(dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% filter(setequal(unique(transcript), c("a > 0.8", "0.4 < a < 0.6"))) %>% .[order(.$dtu_rank, decreasing = FALSE),] %>% .[4,] %>% as_granges(), as_granges(all_samples_simp)) %>% data.frame() %>% .$read_id
rrbs_plot4(all_samples, reads)

reads <- find_overlaps(dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% filter(setequal(unique(transcript), c("a > 0.8", "0.4 < a < 0.6"))) %>% .[order(.$dtu_rank, decreasing = FALSE),] %>% .[5,] %>% as_granges(), as_granges(all_samples_simp)) %>% data.frame() %>% .$read_id
rrbs_plot4(all_samples, reads)
```

* here 4 and 1 are super visually clear, really nice

## visualise top 5 where low and high are only significant categories
```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% filter(setequal(unique(transcript), c("a < 0.2", "a > 0.8"))) %>% .[order(.$dtu_rank, decreasing = FALSE),] %>% .[1:10,]
```

```{r}
source("../code/processData.R")
reads <- find_overlaps(dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% filter(setequal(unique(transcript), c("a < 0.2", "a > 0.8"))) %>% .[order(.$dtu_rank, decreasing = FALSE),] %>% .[1,] %>% as_granges(), as_granges(all_samples_simp)) %>% data.frame() %>% .$read_id
rrbs_plot4(all_samples, reads)

reads <- find_overlaps(dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% filter(setequal(unique(transcript), c("a < 0.2", "a > 0.8"))) %>% .[order(.$dtu_rank, decreasing = FALSE),] %>% .[2,] %>% as_granges(), as_granges(all_samples_simp)) %>% data.frame() %>% .$read_id
rrbs_plot4(all_samples, reads)

reads <- find_overlaps(dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% filter(setequal(unique(transcript), c("a < 0.2", "a > 0.8"))) %>% .[order(.$dtu_rank, decreasing = FALSE),] %>% .[3,] %>% as_granges(), as_granges(all_samples_simp)) %>% data.frame() %>% .$read_id
rrbs_plot4(all_samples, reads)

reads <- find_overlaps(dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% filter(setequal(unique(transcript), c("a < 0.2", "a > 0.8"))) %>% .[order(.$dtu_rank, decreasing = FALSE),] %>% .[4,] %>% as_granges(), as_granges(all_samples_simp)) %>% data.frame() %>% .$read_id
rrbs_plot4(all_samples, reads)

reads <- find_overlaps(dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% filter(setequal(unique(transcript), c("a < 0.2", "a > 0.8"))) %>% .[order(.$dtu_rank, decreasing = FALSE),] %>% .[5,] %>% as_granges(), as_granges(all_samples_simp)) %>% data.frame() %>% .$read_id
rrbs_plot4(all_samples, reads)
```

* these are nice and clear

## summary

Yes, can visualise change in signal from noise -> signal
And signal -> signal
And in a much simpler way than what you could get from beta
So this is a distinct value point for the method
Which I guess is a bit like the proportion plots
Need to do that for these regions
And also will need to filter CpGs, and possibly coverage




