---
title: "dtuResultsWhichGroupSignif"
author: "caitlinpage"
date: "2026-06-10"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

## Introduction
```{r}
library(annotatr)
library(data.table)
library(plyranges)
library(tidyr)
library(stringr)
library(dplyr)
library(ggplot2)
library(edgeR)
library(OpTiles)
library(ggVennDiagram)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
library(missMethyl)
library(UpSetR)
```

```{r}
seqs <- seqnames(BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38)[1:22]

cg_sites <- lapply(seqs, function(x) {
        cbind(data.frame(Biostrings::matchPattern("CG", BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38[[x]])),
            seqnames = x
        )}) %>% bind_rows()
```


```{r}
all_samples <- readRDS("/researchers/caitlin.page/cf_nano/r_output/all_rrbs_samples.rds")
optiles_all <- readRDS("/researchers/caitlin.page/cf_nano/r_output/optiles_all.rds")
dtu.optiles <- readRDS("/researchers/caitlin.page/cf_nano/r_output/ts.gene_optiles.rds")
dmrcate_seq_dmr <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dmr_encode_gm12878_k562.rds")
```

```{r}
all_samples_simp <- all_samples %>% distinct(seqnames, read_start, read_end, strand, sample, read_id, alpha, meth_pattern, num_meth, total)
colnames(all_samples_simp)[2:3] <- c("start", "end")
```

```{r}
dtu_tiles_5_exon <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_5_exon.rds")
```



# analysis


# alpha groups

```{r}
all_samples_simp %>%
  ggplot(aes(x = alpha)) +
  geom_density() +
  geom_vline(xintercept = 0, colour = "red", linetype = "dashed") +
  geom_vline(xintercept = 0.2, colour = "red", linetype = "dashed") +
  geom_vline(xintercept = 0.4, colour = "red", linetype = "dashed") +
  geom_vline(xintercept = 0.6, colour = "red", linetype = "dashed") +
  geom_vline(xintercept = 0.8, colour = "red", linetype = "dashed") +
  geom_vline(xintercept = 1, colour = "red", linetype = "dashed") +
  theme_bw()
```


all_samples_simp %>%
  ggplot(aes(x = alpha)) +
  geom_density() +
  geom_rect(aes(xmin = 0, xmax = 0.2, ymin = 0, ymax = 15), fill = "red", alpha = 0.05) +
  geom_rect(aes(xmin = 0.2, xmax = 0.4, ymin = 0, ymax = 15), fill = "yellowgreen", alpha = 0.05) +
  geom_rect(aes(xmin = 0.4, xmax = 0.6, ymin = 0, ymax = 15), fill = "green", alpha = 0.05) +
  geom_rect(aes(xmin = 0.6, xmax = 0.8, ymin = 0, ymax = 15), fill = "blue", alpha = 0.05) +
  geom_rect(aes(xmin = 0.8, xmax = 1, ymin = 0, ymax = 15), fill = "purple", alpha = 0.05) +
  theme_bw()

* try to run this later



```{r}
dtu_tiles_5_exon$transcript <- factor(dtu_tiles_5_exon$transcript, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))
```

```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% 
  group_by(transcript) %>% summarise(n=n())
```


```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>%
  ggplot(aes(x = transcript, fill = transcript)) +
  geom_bar()
```

* do all the low-med also have significant low? - basically what does it have in common

```{r}
dtu_tiles_5_exon_sig <- dtu_tiles_5_exon %>% filter(FDR < 0.05)
```

levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))

```{r}
upset(fromList(list(low = filter(dtu_tiles_5_exon_sig, transcript == "a < 0.2")$region_id, 
                   low_med = filter(dtu_tiles_5_exon_sig, transcript == "0.2 <= a <= 0.4")$region_id,
                   med = filter(dtu_tiles_5_exon_sig, transcript == "0.4 < a < 0.6")$region_id,
                   med_high = filter(dtu_tiles_5_exon_sig, transcript == "0.6 <= a <= 0.8")$region_id, 
                   high = filter(dtu_tiles_5_exon_sig, transcript == "a > 0.8")$region_id)), order.by = "freq")
```

* yay this is what I want
* and then can redo and add in regions sig on gene level
mn

```{r}
upset(fromList(list(low = filter(dtu_tiles_5_exon_sig, transcript == "a < 0.2")$region_id, 
                   low_med = filter(dtu_tiles_5_exon_sig, transcript == "0.2 <= a <= 0.4")$region_id,
                   med = filter(dtu_tiles_5_exon_sig, transcript == "0.4 < a < 0.6")$region_id,
                   med_high = filter(dtu_tiles_5_exon_sig, transcript == "0.6 <= a <= 0.8")$region_id, 
                   high = filter(dtu_tiles_5_exon_sig, transcript == "a > 0.8")$region_id,
                   region_level = filter(dtu.optiles, FDR < 0.05)$region_id)), order.by = "freq", nsets = 6)
```

```{r}
dmrcate_seq_dmr$rank <- 1:nrow(dmrcate_seq_dmr)
dmr_bin <- find_overlaps(as_granges(dmrcate_seq_dmr), as_granges(optiles_all)) %>% data.frame()
nrow(dmrcate_seq_dmr)
dmr_bin %>% distinct(rank) %>% nrow()
```

```{r}
7849/7914
```

```{r}
upset(fromList(list(low = filter(dtu_tiles_5_exon_sig, transcript == "a < 0.2")$region_id, 
                   low_med = filter(dtu_tiles_5_exon_sig, transcript == "0.2 <= a <= 0.4")$region_id,
                   med = filter(dtu_tiles_5_exon_sig, transcript == "0.4 < a < 0.6")$region_id,
                   med_high = filter(dtu_tiles_5_exon_sig, transcript == "0.6 <= a <= 0.8")$region_id, 
                   high = filter(dtu_tiles_5_exon_sig, transcript == "a > 0.8")$region_id,
                   region_level = filter(dtu.optiles, FDR < 0.05)$region_id,
                   dmr_bin = distinct(dmr_bin, region_id)$region_id)), order.by = "freq", nsets = 7)
```



```{r}
bin_delta <- find_overlaps(as_granges(all_samples), as_granges(optiles_all)) %>% data.frame() %>%
  distinct(sample, read_id, read_start, read_end, alpha, region_id) %>% 
    mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8")) %>%
  mutate(transcript = paste0(region_id, "_", alpha_group)) %>% mutate(sample_type = ifelse(sample %in% c("b1", "b2"), "B", "CML")) %>%
  group_by(region_id, sample_type) %>% mutate(mean_alpha = mean(alpha)) %>% ungroup() %>% distinct(region_id, sample_type, mean_alpha) %>% .[order(.$region_id),]

bin_delta <- bin_delta %>% pivot_wider(names_from = sample_type, values_from = mean_alpha) %>% mutate(delta_b_minus_cml = B - CML)
bin_delta <- data.frame(bin_delta)
```


```{r}
dtu_tiles_5_exon %>% mutate(delta_b_minus_cml = bin_delta[match(.$region_id, bin_delta$region_id), "delta_b_minus_cml"]) %>%
  ggplot(aes(x = logFC, y = delta_b_minus_cml)) +
  geom_point(alpha = 0.2)
```
```{r}
dtu.optiles <- dtu.optiles %>% mutate(delta_b_minus_cml = bin_delta[match(.$region_id, bin_delta$region_id), "delta_b_minus_cml"])
dtu.optiles <- data.frame(dtu.optiles)
```

```{r}
dtu_tiles_5_exon <- dtu_tiles_5_exon %>% mutate(delta_b_minus_cml = bin_delta[match(.$region_id, bin_delta$region_id), "delta_b_minus_cml"])
dtu_tiles_5_exon <- data.frame(dtu_tiles_5_exon)
```

```{r}
dtu_tiles_5_exon %>%
  ggplot(aes(x = logFC, y = -log10(FDR))) +
  geom_point(alpha = 0.2)
```

```{r}
dtu_tiles_5_exon %>% distinct(region_id, delta_b_minus_cml) %>%
  ggplot(aes(x = delta_b_minus_cml)) +
  geom_density()

dtu_tiles_5_exon %>% distinct(region_id, delta_b_minus_cml) %>%
  ggplot(aes(x = delta_b_minus_cml)) +
  geom_histogram(bins = 50)
```


```{r}
dtu_tiles_5_exon %>%
  ggplot(aes(x = logFC)) +
  geom_density()
```

```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>%
  ggplot(aes(x = logFC)) +
  geom_density()
```

# counts per alpha group for the bins

* maybe as boxplots?
* build off the delta_b_minus_cml bin thing


```{r}
alpha_counts <- find_overlaps(as_granges(filter(all_samples_simp)), as_granges(optiles_all)) %>% data.frame() %>%
  distinct(sample, read_id, start, end, alpha, region_id) %>% 
    mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8")) %>%
  mutate(transcript = paste0(region_id, "_", alpha_group)) %>% mutate(sample_type = ifelse(sample %in% c("b1", "b2"), "B", "CML")) %>%
  group_by(sample, transcript) %>%
  mutate(transcript_count = n()) %>% ungroup() %>% data.frame()

alpha_counts$alpha_group <- factor(alpha_counts$alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))
```


* how to plot?
* each bin - counts of each

* 5 density plots - counts of all the group across bins, same as boxplots - this can colour by sample

* but that doesn't relate to indiv bins


```{r}
alpha_counts %>% distinct(transcript, region_id, alpha_group, sample, sample_type, transcript_count) %>% 
  ggplot(aes(x = log2(transcript_count), colour = sample)) +
  geom_density() +
  facet_wrap(~alpha_group)
```

```{r}
alpha_counts %>% distinct(transcript, region_id, alpha_group, sample, sample_type, transcript_count) %>%
  ggplot(aes(x = alpha_group, y = transcript_count, fill = sample)) +
  geom_boxplot()

alpha_counts %>% distinct(transcript, region_id, alpha_group, sample, sample_type, transcript_count) %>%
  ggplot(aes(x = alpha_group, y = log2(transcript_count), fill = sample)) +
  geom_boxplot()
```

```{r}
alpha_counts %>% distinct(region_id) %>% nrow()
```
* I have 165k regions
* that's a lot
* could split by chromosome - and by sample_type

```{r}
alpha_counts[1:5,]
```

```{r}
alpha_counts %>% mutate(FDR = dtu.optiles[match(.$region_id, dtu.optiles$region_id), "FDR"],
                        delta_b_minus_cml = dtu.optiles[match(.$region_id, dtu.optiles$region_id), "delta_b_minus_cml"]) %>% filter(FDR < 0.05) %>% distinct(region_id, alpha_group, sample, sample_type, transcript_count, FDR, delta_b_minus_cml) %>%
  ggplot(aes(x = alpha_group, y = log2(transcript_count), fill = sample)) +
  geom_boxplot()
```

# proportion of alpha group counts in bin

* ooh wait the proportion of the bin is a point - then can make the distribution plots and they're also region informed


```{r}
alpha_counts %>% distinct(sample, sample_type, region_id, alpha_group, transcript_count) %>%
  group_by(sample, region_id) %>% mutate(prop_of_bin = transcript_count/sum(transcript_count)) %>%
  ggplot(aes(x = prop_of_bin, colour = sample)) +
  geom_density() +
  facet_wrap(~alpha_group)
```

```{r}
alpha_counts %>% distinct(sample, sample_type, region_id, alpha_group, transcript_count) %>%
  group_by(sample, region_id) %>% mutate(prop_of_bin = transcript_count/sum(transcript_count)) %>%
  ggplot(aes(x = prop_of_bin, colour = alpha_group)) +
  geom_density() +
  facet_wrap(~sample)
```


```{r}
alpha_counts %>% distinct(sample, sample_type, region_id, alpha_group, transcript_count) %>%
  group_by(sample, region_id) %>% mutate(prop_of_bin = transcript_count/sum(transcript_count)) %>%
  ggplot(aes(x = sample, y = prop_of_bin, fill = alpha_group)) +
  geom_boxplot()
```

```{r}
alpha_counts %>% distinct(sample, sample_type, region_id, alpha_group, transcript_count) %>%
  group_by(sample, region_id) %>% mutate(prop_of_bin = transcript_count/sum(transcript_count)) %>%
  ggplot(aes(x = alpha_group, y = prop_of_bin, fill = sample)) +
  geom_boxplot()
```

* I like this


* does it look different in significant plots?? - facet by signif?

```{r}
alpha_counts %>% distinct(sample, sample_type, region_id, alpha_group, transcript_count) %>% 
  mutate(FDR = dtu.optiles[match(.$region_id, dtu.optiles$region_id), "FDR"], sig_fdr = ifelse(FDR < 0.05, TRUE, FALSE)) %>% filter(!is.na(sig_fdr)) %>%
  group_by(sample, region_id) %>% mutate(prop_of_bin = transcript_count/sum(transcript_count)) %>%
  ggplot(aes(x = alpha_group, y = prop_of_bin, fill = sample)) +
  geom_boxplot() +
  facet_wrap(~sig_fdr)
```

* so they're different
* but not in the way I expected? 

* how do I want to plot this?
* yes reads but I can't be plotting all of them
* which is why i like distrib

```{r}
bin_delta %>%
  ggplot(aes(x = delta_b_minus_cml)) +
  geom_density()
```

* do plot again but say signif id abs(delta_b_minus_cml) > 0.2


```{r}
alpha_counts %>% distinct(sample, sample_type, region_id, alpha_group, transcript_count) %>% 
  mutate(FDR = dtu.optiles[match(.$region_id, dtu.optiles$region_id), "FDR"], sig_fdr = ifelse(FDR < 0.05, TRUE, FALSE),
         delta_b_minus_cml = dtu.optiles[match(.$region_id, dtu.optiles$region_id), "delta_b_minus_cml"], sig_delta = ifelse(abs(delta_b_minus_cml) > 0.2, TRUE, FALSE)) %>% filter(!is.na(sig_delta)) %>%
  group_by(sample, region_id) %>% mutate(prop_of_bin = transcript_count/sum(transcript_count)) %>%
  ggplot(aes(x = alpha_group, y = prop_of_bin, fill = sample)) +
  geom_boxplot() +
  facet_wrap(~sig_delta)
```

* looks pretty similar to using fdr like this



* I think need to just plot regions

# plot regions indiv

* need bin_delta to have something about number of counts overall in the region - because otherwise no good - I want coverage
* also need total cpgs in bin

```{r}
optiles_all_cpg <- find_overlaps(as_granges(optiles_all), as_granges(cg_sites)) %>% data.frame() %>% group_by(region_id, seqnames, start, end) %>% summarise(n_cg = n()) %>%
  ungroup() %>% data.frame()
```


```{r}
coverage <- alpha_counts %>% distinct(region_id, sample, transcript_count) %>% group_by(region_id) %>% summarise(coverage = sum(transcript_count)) %>% ungroup() %>% data.frame()
```

```{r}
bin_delta <- bin_delta %>% mutate(coverage = coverage[match(.$region_id, coverage$region_id), "coverage"])
bin_delta %>%
  ggplot(aes(x = delta_b_minus_cml, y = coverage)) +
  geom_point(alpha = 0.3) +
  scale_y_continuous(limits = c(0, 500))
```

```{r}
bin_delta <- bin_delta %>% mutate(num_cg = optiles_all_cpg[match(.$region_id, optiles_all_cpg$region_id), "n_cg"])
bin_delta %>%
  ggplot(aes(x = num_cg, y = coverage)) +
  geom_point()
```


```{r}
regions <- bin_delta %>% mutate(abs_delta = abs(delta_b_minus_cml)) %>% filter(coverage > 50, num_cg > 5) %>% .[order(.$abs_delta, decreasing = TRUE),] %>% .[1:10,] %>% .$region_id
```

```{r}
source("../code/processData.R")

reads <- alpha_counts %>% filter(region_id == regions[1]) %>% .$read_id
rrbs_plot4(all_samples, reads, cg_sites)
```

```{r}
source("../code/processData.R")

reads <- alpha_counts %>% filter(region_id == regions[2]) %>% .$read_id
rrbs_plot4(all_samples, reads, cg_sites)

reads <- alpha_counts %>% filter(region_id == regions[3]) %>% .$read_id
rrbs_plot4(all_samples, reads, cg_sites)

reads <- alpha_counts %>% filter(region_id == regions[4]) %>% .$read_id
rrbs_plot4(all_samples, reads, cg_sites)

reads <- alpha_counts %>% filter(region_id == regions[5]) %>% .$read_id
rrbs_plot4(all_samples, reads, cg_sites)

reads <- alpha_counts %>% filter(region_id == regions[6]) %>% .$read_id
rrbs_plot4(all_samples, reads, cg_sites)

reads <- alpha_counts %>% filter(region_id == regions[7]) %>% .$read_id
rrbs_plot4(all_samples, reads, cg_sites)

reads <- alpha_counts %>% filter(region_id == regions[8]) %>% .$read_id
rrbs_plot4(all_samples, reads, cg_sites)
```


# need to repeat all of this with the minimum 3 cpgs on read - because these plots aren't that pretty

* boo boring and irritating
```{r}
bin_delta_min3 <- find_overlaps(as_granges(filter(all_samples, total >= 3)), as_granges(optiles_all)) %>% data.frame() %>%
  distinct(sample, read_id, read_start, read_end, alpha, region_id) %>% 
    mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8")) %>%
  mutate(transcript = paste0(region_id, "_", alpha_group)) %>% mutate(sample_type = ifelse(sample %in% c("b1", "b2"), "B", "CML")) %>%
  group_by(region_id, sample_type) %>% mutate(mean_alpha = mean(alpha)) %>% ungroup() %>% distinct(region_id, sample_type, mean_alpha) %>% .[order(.$region_id),]

bin_delta_min3 <- bin_delta_min3 %>% pivot_wider(names_from = sample_type, values_from = mean_alpha) %>% mutate(delta_b_minus_cml = B - CML)
```



```{r}
alpha_counts_filt <- find_overlaps(as_granges(filter(all_samples_simp, total >= 3)), as_granges(optiles_all)) %>% data.frame() %>%
  distinct(sample, read_id, start, end, alpha, region_id) %>% 
    mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8")) %>%
  mutate(transcript = paste0(region_id, "_", alpha_group)) %>% mutate(sample_type = ifelse(sample %in% c("b1", "b2"), "B", "CML")) %>%
  group_by(sample, transcript) %>%
  mutate(transcript_count = n()) %>% ungroup() %>% data.frame()

alpha_counts_filt$alpha_group <- factor(alpha_counts_filt$alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))
```
```{r}
all_samples_min3 <- filter(all_samples, total >= 3) %>% data.frame()
```

```{r}
optiles_all_cpg <- find_overlaps(as_granges(optiles_all), as_granges(cg_sites)) %>% data.frame() %>% group_by(region_id, seqnames, start, end) %>% summarise(n_cg = n()) %>%
  ungroup() %>% data.frame()
```


```{r}
coverage_filt <- alpha_counts_filt %>% distinct(region_id, sample, transcript_count) %>% group_by(region_id) %>% summarise(coverage = sum(transcript_count)) %>% ungroup() %>% data.frame()

bin_delta_min3 <- bin_delta_min3 %>% mutate(coverage = coverage_filt[match(.$region_id, coverage_filt$region_id), "coverage"])

bin_delta_min3 <- bin_delta_min3 %>% mutate(num_cg = optiles_all_cpg[match(.$region_id, optiles_all_cpg$region_id), "n_cg"])
```

```{r}
dtu_function <- function(all_samples, optiles_all, num_groups = 5) {
  dtu_optile <- find_overlaps(as_granges(all_samples), as_granges(optiles_all)) %>% data.frame() %>%
  distinct(sample, read_id, read_start, read_end, alpha, region_id) 
  
  if(num_groups == 5) {
  dtu_optile <- dtu_optile %>% 
    mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8"))
  } else if(num_groups == 3) {
  dtu_optile <- dtu_optile %>% 
    mutate(alpha_group = case_when(alpha < 0.34 ~ "a < 0.34",
                                alpha > 0.65 ~ "a > 0.65",
                                TRUE ~ "0.34 <= a <= 0.65"))
  } else if(num_groups == 4) {
  dtu_optile <- dtu_optile %>% 
    mutate(alpha_group = case_when(alpha <= 0.25 ~ "a <= 0.25",
                                alpha >= 0.8 ~ "a >= 0.75",
                                alpha > 0.25 & alpha <= 0.5 ~ "0.25 <= a <= 0.5",
                                TRUE ~ "0.5 < a < 0.75"))
}

dtu_optile <- dtu_optile %>% 
  mutate(transcript = paste0(region_id, "_", alpha_group)) %>%
  group_by(sample, transcript) %>%
 summarise(transcript_count = n()) %>% ungroup() %>% data.frame()

# build matrix of "transcript" counts - transcript rows, sample as columns
dtu_optile <- dtu_optile %>% pivot_wider(names_from = sample, values_from = transcript_count)
dtu_optile[is.na(dtu_optile)] <- 0
dtu_optile$gene_id <- sub("_.*", "", dtu_optile$transcript)

# DTU analysis with edgeR
## Alicia prefers gene method over Simes method (documentation slightly confusing about difference)
y <- DGEList(as.matrix(dtu_optile[,2:5]), group = c(1,1,2,2), genes = dtu_optile[,c("gene_id", "transcript")])
design <- model.matrix(~ c(1,1,2,2), data = y$samples)
fit <- glmQLFit(y, design, robust=TRUE)
ds <- diffSpliceDGE(fit, coef = 2, geneid = "gene_id", exonid = "transcript")
ts.gene_optiles <- topSpliceDGE(ds, test = "gene", number = Inf)

# add info to results: get proportion counts
gene_cg <- find_overlaps(as_granges(optiles_all), as_granges(cg_sites)) %>% data.frame() %>%
  group_by(region_id) %>%
  summarise(n_cg = n()) %>%
  ungroup() %>% data.frame()
fit$genes <- fit$genes %>% mutate(n_cg = gene_cg[match(.$gene_id, gene_cg$region_id), "n_cg"])
rpk <- y$counts/rep(fit$genes$n_cg/1000, times = ncol(y$counts))
tpm_mat <- sweep(rpk, 2, colSums(rpk)/1e6, "/")
tpm_mat <- tpm_mat %>% data.frame() %>% mutate(id = dtu_optile$transcript, gene_id = dtu_optile$gene_id)

tpm_mat <- tpm_mat %>%
  group_by(gene_id) %>%
  mutate(prop_S1_gm = b1/sum(b1),
         prop_S2_gm = b2/sum(b2),
         prop_S3_k = cml1/sum(cml1),
         prop_S4_k = cml2/sum(cml2)) %>%
  ungroup() %>% data.frame()


# add locations back in so compatible with plyranges stuff
## this bit only applicable to OPTILES because of the way it structures the ID
ts.gene_optiles <- ts.gene_optiles %>%
  mutate(region_id = gene_id) %>%
  separate_wider_delim(gene_id, delim = ":", names = c("seqnames", "gene_id")) %>%
  mutate(seqnames = gsub("chrchr", "chr", seqnames)) %>%
  separate_wider_delim(gene_id, delim = "-", names = c("start", "end"))
ts.gene_optiles$start <- as.double(ts.gene_optiles$start)
ts.gene_optiles$end <- as.double(ts.gene_optiles$end)
ts.gene_optiles$dtu_rank <- 1:nrow(ts.gene_optiles)
ts.gene_optiles
}

dtu_tiles_filt_min3 <- dtu_function(filter(all_samples, total >= 3), optiles_all, 5)
```

```{r}
bin_delta_min3 %>% mutate(abs_delta = abs(delta_b_minus_cml)) %>% filter(coverage > 50, num_cg > 5) %>% .[order(.$abs_delta, decreasing = TRUE),] %>% .[1:5,]
```

```{r}
regions <- bin_delta_min3 %>% mutate(abs_delta = abs(delta_b_minus_cml)) %>% filter(coverage > 50, num_cg > 5) %>% .[order(.$abs_delta, decreasing = TRUE),] %>% .[1:10,] %>% .$region_id
```


```{r}
source("../code/processData.R")
reads <- alpha_counts_filt %>% filter(region_id == regions[1]) %>% .$read_id
rrbs_plot4(all_samples_min3, reads, cg_sites)

reads <- alpha_counts_filt %>% filter(region_id == regions[2]) %>% .$read_id
rrbs_plot4(all_samples_min3, reads, cg_sites)

reads <- alpha_counts_filt %>% filter(region_id == regions[3]) %>% .$read_id
rrbs_plot4(all_samples_min3, reads, cg_sites)

reads <- alpha_counts_filt %>% filter(region_id == regions[4]) %>% .$read_id
rrbs_plot4(all_samples_min3, reads, cg_sites)

reads <- alpha_counts_filt %>% filter(region_id == regions[5]) %>% .$read_id
rrbs_plot4(all_samples_min3, reads, cg_sites)

reads <- alpha_counts_filt %>% filter(region_id == regions[6]) %>% .$read_id
rrbs_plot4(all_samples_min3, reads, cg_sites)

reads <- alpha_counts_filt %>% filter(region_id == regions[7]) %>% .$read_id
rrbs_plot4(all_samples_min3, reads, cg_sites)

reads <- alpha_counts_filt %>% filter(region_id == regions[8]) %>% .$read_id
rrbs_plot4(all_samples_min3, reads, cg_sites)
```

# top fdr regions


```{r}
regions <- dtu_tiles_filt_min3 %>% .[1:10,] %>% .$region_id
```

```{r}
source("../code/processData.R")
reads <- alpha_counts_filt %>% filter(region_id == regions[1]) %>% .$read_id
rrbs_plot4(all_samples_min3, reads, cg_sites)

reads <- alpha_counts_filt %>% filter(region_id == regions[2]) %>% .$read_id
rrbs_plot4(all_samples_min3, reads, cg_sites)

reads <- alpha_counts_filt %>% filter(region_id == regions[3]) %>% .$read_id
rrbs_plot4(all_samples_min3, reads, cg_sites)

reads <- alpha_counts_filt %>% filter(region_id == regions[4]) %>% .$read_id
rrbs_plot4(all_samples_min3, reads, cg_sites)

reads <- alpha_counts_filt %>% filter(region_id == regions[5]) %>% .$read_id
rrbs_plot4(all_samples_min3, reads, cg_sites)

reads <- alpha_counts_filt %>% filter(region_id == regions[6]) %>% .$read_id
rrbs_plot4(all_samples_min3, reads, cg_sites)

reads <- alpha_counts_filt %>% filter(region_id == regions[7]) %>% .$read_id
rrbs_plot4(all_samples_min3, reads, cg_sites)

reads <- alpha_counts_filt %>% filter(region_id == regions[8]) %>% .$read_id
rrbs_plot4(all_samples_min3, reads, cg_sites)
```

```{r}
regions <- dtu_tiles_5_exon_sig %>% .[1:10,] %>% .$region_id
dtu_tiles_5_exon_sig[1:5,]
```

```{r}
source("../code/processData.R")
reads <- alpha_counts %>% filter(region_id == regions[1]) %>% .$read_id
rrbs_plot4(all_samples, reads, cg_sites)

reads <- alpha_counts %>% filter(region_id == regions[2]) %>% .$read_id
rrbs_plot4(all_samples, reads, cg_sites)

reads <- alpha_counts %>% filter(region_id == regions[3]) %>% .$read_id
rrbs_plot4(all_samples, reads, cg_sites)

reads <- alpha_counts %>% filter(region_id == regions[4]) %>% .$read_id
rrbs_plot4(all_samples, reads, cg_sites)

reads <- alpha_counts %>% filter(region_id == regions[5]) %>% .$read_id
rrbs_plot4(all_samples, reads, cg_sites)

reads <- alpha_counts %>% filter(region_id == regions[6]) %>% .$read_id
rrbs_plot4(all_samples, reads, cg_sites)

reads <- alpha_counts %>% filter(region_id == regions[7]) %>% .$read_id
rrbs_plot4(all_samples, reads, cg_sites)

reads <- alpha_counts %>% filter(region_id == regions[8]) %>% .$read_id
rrbs_plot4(all_samples, reads, cg_sites)
```

# logFC instead

```{r}
regions <- dtu_tiles_5_exon %>% mutate(abs_fc = abs(logFC)) %>% .[order(.$abs_fc, decreasing = TRUE),] %>% .[1:10,] %>% .$region_id
dtu_tiles_5_exon_sig[1:5,]
```

```{r}
source("../code/processData.R")
reads <- alpha_counts %>% filter(region_id == regions[1]) %>% .$read_id
rrbs_plot4(all_samples, reads, cg_sites)

reads <- alpha_counts %>% filter(region_id == regions[2]) %>% .$read_id
rrbs_plot4(all_samples, reads, cg_sites)

reads <- alpha_counts %>% filter(region_id == regions[3]) %>% .$read_id
rrbs_plot4(all_samples, reads, cg_sites)

reads <- alpha_counts %>% filter(region_id == regions[4]) %>% .$read_id
rrbs_plot4(all_samples, reads, cg_sites)

reads <- alpha_counts %>% filter(region_id == regions[5]) %>% .$read_id
rrbs_plot4(all_samples, reads, cg_sites)

reads <- alpha_counts %>% filter(region_id == regions[6]) %>% .$read_id
rrbs_plot4(all_samples, reads, cg_sites)

reads <- alpha_counts %>% filter(region_id == regions[7]) %>% .$read_id
rrbs_plot4(all_samples, reads, cg_sites)

reads <- alpha_counts %>% filter(region_id == regions[8]) %>% .$read_id
rrbs_plot4(all_samples, reads, cg_sites)
```
```{r}
dtu_tiles_5_exon %>%
  ggplot(aes(x = logFC, y = log10(FDR), colour = transcript)) +
  geom_point(alpha = 0.4)
```

# overlap of region ids - top 200 delta_b_minus_cml, and top 200 fdr (except low fdr obv)

filter(coverage > 50, num_cg > 5)
```{r}
ggVennDiagram(list(delta_b_minus_cml = bin_delta_min3 %>% mutate(abs_delta = abs(delta_b_minus_cml)) %>% filter(coverage > 50, num_cg > 5) %>% .[order(.$abs_delta, decreasing = TRUE),] %>% .[1:200,] %>% .$region_id,
                   fdr = dtu_tiles_filt_min3[1:200,]$region_id))

ggVennDiagram(list(delta_b_minus_cml = bin_delta_min3 %>% mutate(abs_delta = abs(delta_b_minus_cml)) %>% .[order(.$abs_delta, decreasing = TRUE),] %>% .[1:200,] %>% .$region_id,
                   fdr = dtu_tiles_filt_min3[1:200,]$region_id))
```

```{r}
ggVennDiagram(list(delta_b_minus_cml = bin_delta_min3 %>% mutate(abs_delta = abs(delta_b_minus_cml)) %>% filter(coverage > 50, num_cg > 5) %>% .[order(.$abs_delta, decreasing = TRUE),] %>% .[1:1000,] %>% .$region_id,
                   fdr = dtu_tiles_filt_min3[1:1000,]$region_id))

ggVennDiagram(list(delta_b_minus_cml = bin_delta_min3 %>% mutate(abs_delta = abs(delta_b_minus_cml)) %>% filter(coverage > 50, num_cg > 5) %>% .[order(.$abs_delta, decreasing = TRUE),] %>% .[1:500,] %>% .$region_id,
                   fdr = dtu_tiles_filt_min3[1:500,]$region_id))
```

```{r}
ggVennDiagram(list(delta_b_minus_cml = bin_delta_min3 %>% mutate(abs_delta = abs(delta_b_minus_cml)) %>% filter(coverage > 50, num_cg > 5) %>% .[order(.$abs_delta, decreasing = TRUE),] %>% .[1:5000,] %>% .$region_id,
                   fdr = dtu_tiles_filt_min3[1:5000,]$region_id))

ggVennDiagram(list(delta_b_minus_cml = bin_delta_min3 %>% mutate(abs_delta = abs(delta_b_minus_cml)) %>% filter(coverage > 50, num_cg > 5) %>% .[order(.$abs_delta, decreasing = TRUE),] %>% .[1:10000,] %>% .$region_id,
                   fdr = dtu_tiles_filt_min3[1:10000,]$region_id))
```


dtu_optile <- dtu_optile %>% 
  mutate(transcript = paste0(region_id, "_", alpha_group)) %>%
  group_by(sample, transcript) %>%
 summarise(transcript_count = n()) %>% ungroup() %>% data.frame()

# build matrix of "transcript" counts - transcript rows, sample as columns
dtu_optile <- dtu_optile %>% pivot_wider(names_from = sample, values_from = transcript_count)
dtu_optile[is.na(dtu_optile)] <- 0
dtu_optile$gene_id <- sub("_.*", "", dtu_optile$transcript)




# are there strings of significant tiles???


```{r}
try <- optiles_all %>% mutate(sig_tile = ifelse(region_id %in% filter(dtu.optiles, FDR < 0.05, abs(delta_b_minus_cml) > 0.2)$region_id, TRUE, FALSE)) %>%
  mutate(which_tile_sig = ifelse(sig_tile == TRUE, row_number(), NA)) 
 # mutate(seq_sig = rle(!is.na(which_tile_sig))$values)
#  mutate(string_sig = ifelse(sig_tile == TRUE & !is.na(lag(sig_tile)), lag(which_tile_sig), NA))
```



```{r}
try$grp <- cumsum(!is.na(try$which_tile_sig) & c(TRUE, is.na(head(try$which_tile_sig, -1))))
try$run_len <- ifelse(is.na(try$which_tile_sig), NA, ave(!is.na(try$which_tile_sig), try$grp, FUN = sum))

r <- rle(!is.na(try$which_tile_sig))
```

```{r}
try %>% filter(!is.na(run_len)) %>% distinct(grp, run_len) %>%
  ggplot(aes(x = as.factor(run_len), fill = as.factor(run_len))) +
  geom_bar()
```

```{r}
try %>% filter(!is.na(run_len)) %>% distinct(grp, run_len) %>% group_by(run_len) %>% summarise(n=n()) %>% ungroup() %>% mutate(prop = n/sum(n)) %>% .[1:5,]
```

* ok now it's where are they?

* also how many tiles do dmrs overlap?

```{r}
tiles_dmr <- find_overlaps(as_granges(optiles_all), as_granges(dmrcate_seq_dmr)) %>% data.frame()
```


```{r}
tiles_dmr %>%
  group_by(rank) %>% summarise(n=n()) %>%
  ggplot(aes(x = as.factor(n), fill = as.factor(n))) +
  geom_bar()
```

```{r}
tiles_dmr %>%
  group_by(rank) %>% summarise(n=n()) %>% ungroup() %>% group_by(n) %>% summarise(count = n()) %>% ungroup() %>% mutate(prop = count/sum(count)) %>% .[1:5,]
```

* only about half of dmrs are 1 tile, as opposed to 86% of dtu results


```{r}
txdb_genes <- genes(TxDb.Hsapiens.UCSC.hg19.knownGene)
```

```{r}
find_overlaps(txdb_genes, as_granges(try)) %>% data.frame() %>%
  filter(!is.na(run_len)) %>% distinct(grp, run_len) %>%
  ggplot(aes(x = as.factor(run_len), fill = as.factor(run_len))) +
  geom_bar()
```
* looks about the same - thought might be different
* numbers are different but proportionally similar

* might be different in promoter? - vs other gene features idk


```{r}
#optiles_all %>% mutate(sig_tile = ifelse(region_id %in% filter(dtu.optiles, FDR < 0.05, abs(delta_b_minus_cml) > 0.2)$region_id, TRUE, FALSE)) %>%
#  mutate(which_tile_sig = ifelse(sig_tile == TRUE, row_number(), NA)) %>%
```


* don't want to do that


# betas
* so what about do we get signif when beta is more medium???

```{r}
betas <- all_samples %>% distinct(seqnames, start, sample, beta) %>% pivot_wider(names_from = sample, values_from = beta)
```

```{r}
betas <- betas %>% mutate(mean_b = rowMeans(across(c(b1, b2)), na.rm = TRUE), 
                 mean_cml = rowMeans(across(c(cml1, cml2)), na.rm = TRUE),
                 beta_delta_b_minus_cml = mean_b - mean_cml) %>% data.frame()
betas$end <- betas$start
```

```{r}
dtu.optiles %>% filter(FDR >= 0.05) %>%
  ggplot(aes(x = delta_b_minus_cml)) +
  geom_density()

dtu.optiles %>% filter(FDR >= 0.05) %>%
  ggplot(aes(x = abs(delta_b_minus_cml))) +
  geom_density()
```

```{r}
dtu.optiles %>% mutate(fdr_sig = ifelse(FDR < 0.05, TRUE, FALSE)) %>%
  ggplot(aes(x = delta_b_minus_cml, colour = fdr_sig)) +
  geom_density()

dtu.optiles %>% mutate(fdr_sig = ifelse(FDR < 0.05, TRUE, FALSE)) %>%
  ggplot(aes(x = abs(delta_b_minus_cml), colour = fdr_sig)) +
  geom_density()
```


```{r}
find_overlaps(as_granges(dtu.optiles[1:10,]), as_granges(betas)) %>% data.frame() %>%
  pivot_longer(cols = c(mean_b, mean_cml), names_to = "type", values_to = "beta") %>% group_by(dtu_rank) %>% mutate(pos = 1:n()) %>%
  ggplot(aes(x = pos, y = beta, colour = type)) +
  geom_point() +
  facet_wrap(~ dtu_rank)
```

```{r}
find_overlaps(as_granges(dmrcate_seq_dmr[1:10,]), as_granges(betas)) %>% data.frame() %>%
  pivot_longer(cols = c(mean_b, mean_cml), names_to = "type", values_to = "beta") %>% group_by(rank) %>% mutate(pos = 1:n()) %>%
  ggplot(aes(x = pos, y = beta, colour = type)) +
  geom_point() +
  facet_wrap(~ rank)
```
```{r}
find_overlaps(as_granges(dmrcate_seq_dmr[1:10,]), as_granges(betas)) %>% data.frame() %>%
  pivot_longer(cols = c(mean_b, mean_cml), names_to = "type", values_to = "beta") %>% group_by(rank) %>% mutate(pos = 1:n()) %>% filter(pos <= 80) %>%
  ggplot(aes(x = pos, y = beta, colour = type)) +
  geom_point() +
  facet_wrap(~ rank)
```

* damn again - dmr is getting the mix

* oh important is the ones that don't overlap in signif

```{r}
ggVennDiagram(list(dtu = filter(dtu.optiles, FDR < 0.05, abs(delta_b_minus_cml) > 0.2)$region_id, dmr = tiles_dmr$region_id))
```
* oh - why did I think the overlap was way better???
(maybe it was in gene???)
```{r}
ggVennDiagram(list(dtu = filter(dtu.optiles, FDR < 0.05)$region_id, dmr = tiles_dmr$region_id))
```
```{r}
betas$pos <- paste0(betas$seqnames, "-", betas$start)
```

```{r}
optiles_all_beta <- find_overlaps(as_granges(optiles_all), as_granges(betas)) %>% data.frame()
optiles_all_beta <- optiles_all_beta %>% mutate(alpha_b = bin_delta[match(.$region_id, bin_delta$region_id), "B"],
                                                alpha_cml = bin_delta[match(.$region_id, bin_delta$region_id), "CML"],
                                                alpha_delta_b_minus_cml = bin_delta[match(.$region_id, bin_delta$region_id), "delta_b_minus_cml"],
                                                alpha_b = as.double(alpha_b), alpha_cml = as.double(alpha_cml), alpha_delta_b_minus_cml = as.double(alpha_delta_b_minus_cml))
```


```{r}
optiles_all_beta %>% mutate(sig_dmr = ifelse(region_id %in% tiles_dmr$region_id, TRUE, FALSE), 
                            sig_dtu = ifelse(region_id %in% filter(dtu.optiles, FDR < 0.05, abs(delta_b_minus_cml) > 0.2)$region_id, TRUE, FALSE)) %>% .[1:5,]
# want distrib of betas in tiles that are signif in both, indiv, none - where do we get the medium betas
```

* also the alpha delta_b_minus_cml I have is a mean for the tile
* does that compare at all if I do that for beta? - compare



```{r, eval=FALSE}
optiles_all %>% mutate(sig_tile = ifelse(region_id %in% filter(dtu.optiles, FDR < 0.05, abs(delta_b_minus_cml) > 0.2)$region_id, TRUE, FALSE)) %>%
  mutate(which_tile_sig = ifelse(sig_tile == TRUE, row_number(), NA)) %>%
  mutate(seq_sig = rle(!is.na(which_tile_sig))$lengths[rle(!is.na(which_tile_sig))$values]) %>% .[1:5,]
```

Qs

* does dmr/dtu have a mid range beta? 
* who has more - probs dmrs - ick

```{r}
optiles_all_beta %>% mutate(sig_dmr = ifelse(region_id %in% tiles_dmr$region_id, TRUE, FALSE), 
                            sig_dtu = ifelse(region_id %in% filter(dtu.optiles, FDR < 0.05, abs(delta_b_minus_cml) > 0.2)$region_id, TRUE, FALSE)) %>%
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(mid_beta = ifelse(mean_b >= 0.4 & mean_b <= 0.6 | mean_cml >= 0.4 & mean_cml <= 0.6, TRUE, FALSE)) %>%
  group_by(region_id) %>% mutate(has_mid_beta = any(mid_beta == TRUE)) %>% ungroup() %>% distinct(region_id, sig_dmr, sig_dtu, has_mid_beta) %>%
  ggplot(aes(x = sig_dtu, fill = has_mid_beta)) +
  geom_bar()
```

```{r}
optiles_all_beta %>% mutate(sig_dmr = ifelse(region_id %in% tiles_dmr$region_id, TRUE, FALSE), 
                            sig_dtu = ifelse(region_id %in% filter(dtu.optiles, FDR < 0.05, abs(delta_b_minus_cml) > 0.2)$region_id, TRUE, FALSE)) %>%
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(mid_beta = ifelse(mean_b >= 0.4 & mean_b <= 0.6 | mean_cml >= 0.4 & mean_cml <= 0.6, TRUE, FALSE)) %>%
  group_by(region_id) %>% mutate(has_mid_beta = any(mid_beta == TRUE)) %>% ungroup() %>% distinct(region_id, sig_dmr, sig_dtu, has_mid_beta) %>%
  ggplot(aes(x = sig_dmr, fill = has_mid_beta)) +
  geom_bar()
```


```{r}
optiles_all_beta %>% mutate(sig_dmr = ifelse(region_id %in% tiles_dmr$region_id, TRUE, FALSE), 
                            sig_dtu = ifelse(region_id %in% filter(dtu.optiles, FDR < 0.05, abs(delta_b_minus_cml) > 0.2)$region_id, TRUE, FALSE)) %>%
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(mid_beta = ifelse(mean_b >= 0.4 & mean_b <= 0.6 | mean_cml >= 0.4 & mean_cml <= 0.6, TRUE, FALSE)) %>%
  group_by(region_id) %>% mutate(has_mid_beta = any(mid_beta == TRUE)) %>% ungroup() %>% distinct(region_id, sig_dmr, sig_dtu, has_mid_beta) %>%
  mutate(sig_dmr = ifelse(sig_dmr == TRUE, "dmr_sig", "dmr_not_sig"), sig_dtu = ifelse(sig_dtu == TRUE, "dtu_sig", "dtu_not_sig")) %>%
  pivot_longer(cols = c(sig_dmr, sig_dtu), names_to = "sig", values_to = "is_sig") %>% .[,c(1,2,4)] %>% distinct() %>%
  ggplot(aes(x = is_sig, fill = has_mid_beta)) +
  geom_bar()
```

```{r}
optiles_all_beta %>% mutate(sig_dmr = ifelse(region_id %in% tiles_dmr$region_id, TRUE, FALSE), 
                            sig_dtu = ifelse(region_id %in% filter(dtu.optiles, FDR < 0.05, abs(delta_b_minus_cml) > 0.2)$region_id, TRUE, FALSE)) %>%
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(mid_beta = ifelse(mean_b >= 0.4 & mean_b <= 0.6 | mean_cml >= 0.4 & mean_cml <= 0.6, TRUE, FALSE)) %>%
  group_by(region_id) %>% mutate(has_mid_beta = any(mid_beta == TRUE)) %>% ungroup() %>% distinct(region_id, sig_dmr, sig_dtu, has_mid_beta) %>%
  mutate(sig_dmr = ifelse(sig_dmr == TRUE, "dmr_sig", "dmr_not_sig"), sig_dtu = ifelse(sig_dtu == TRUE, "dtu_sig", "dtu_not_sig")) %>%
  pivot_longer(cols = c(sig_dmr, sig_dtu), names_to = "sig", values_to = "is_sig") %>% .[,c(1,2,4)] %>% distinct() %>%
  ggplot(aes(x = is_sig, fill = has_mid_beta)) +
  geom_bar(position = "fill")
```
* but this for the dmrs - is if the mid beta is in the bin - not the dmr itself

```{r}
find_overlaps(as_granges(dmrcate_seq_dmr), as_granges(betas)) %>% data.frame() %>% 
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(mid_beta = ifelse(mean_b >= 0.4 & mean_b <= 0.6 | mean_cml >= 0.4 & mean_cml <= 0.6, TRUE, FALSE)) %>%
  group_by(rank) %>% mutate(has_mid_beta = any(mid_beta == TRUE)) %>% ungroup() %>% distinct(rank, has_mid_beta) %>%
  ggplot(aes(x = has_mid_beta, fill = has_mid_beta)) +
  geom_bar()

find_overlaps(as_granges(dmrcate_seq_dmr), as_granges(betas)) %>% data.frame() %>% 
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(mid_beta = ifelse(mean_b >= 0.4 & mean_b <= 0.6 | mean_cml >= 0.4 & mean_cml <= 0.6, TRUE, FALSE)) %>%
  group_by(rank) %>% mutate(has_mid_beta = any(mid_beta == TRUE)) %>% ungroup() %>% distinct(rank, has_mid_beta) %>% mutate(a = "a") %>%
  ggplot(aes(x = a, fill = has_mid_beta)) +
  geom_bar(position = "fill")
```
* ok so it is definitely higher in dmrs

```{r, eval=FALSE}
find_overlaps(as_granges(dmrcate_seq_dmr), as_granges(betas)) %>% data.frame() %>% 
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(mid_beta = ifelse(mean_b >= 0.4 & mean_b <= 0.6 | mean_cml >= 0.4 & mean_cml <= 0.6, TRUE, FALSE)) %>%
  group_by(rank) %>% mutate(has_mid_beta = any(mid_beta == TRUE)) %>% ungroup() %>% distinct(rank, has_mid_beta)
```


```{r}
optiles_all_beta %>% mutate(sig_dmr = ifelse(region_id %in% tiles_dmr$region_id, TRUE, FALSE), 
                            sig_dtu = ifelse(region_id %in% filter(dtu.optiles, FDR < 0.05, abs(delta_b_minus_cml) > 0.2)$region_id, TRUE, FALSE)) %>%
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(mid_beta = ifelse(mean_b >= 0.4 & mean_b <= 0.6 | mean_cml >= 0.4 & mean_cml <= 0.6, TRUE, FALSE)) %>%
  group_by(region_id) %>% mutate(has_mid_beta = any(mid_beta == TRUE)) %>% ungroup() %>% distinct(region_id, sig_dmr, sig_dtu, has_mid_beta) %>%
  mutate(sig_dmr = ifelse(sig_dmr == TRUE, "dmr_sig", "dmr_not_sig"), sig_dtu = ifelse(sig_dtu == TRUE, "dtu_sig", "dtu_not_sig")) %>%
  pivot_longer(cols = c(sig_dmr, sig_dtu), names_to = "sig", values_to = "is_sig") %>% .[,c(1,2,4)] %>% distinct() %>%
  ggplot(aes(x = is_sig, fill = has_mid_beta)) +
  geom_bar(position = "fill")
```


```{r, eval=FALSE}
rbind(optiles_all_beta %>% filter(region_id %in% filter(dtu.optiles, FDR < 0.05, abs(delta_b_minus_cml) > 0.2)$region_id) %>%
  filter(!is.na(mean_b), !is.na(mean_cml)) %>% mutate(mid_beta = ifelse(mean_b >= 0.4 & mean_b <= 0.6 | mean_cml >= 0.4 & mean_cml <= 0.6, TRUE, FALSE)) %>%
  group_by(region_id) %>% mutate(has_mid_beta = any(mid_beta == TRUE)) %>% ungroup() %>% 
  mutate(type = "dtu") %>% distinct(region_id, has_mid_beta, type),
  find_overlaps(as_granges(dmrcate_seq_dmr), as_granges(betas)) %>% data.frame() %>% 
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(mid_beta = ifelse(mean_b >= 0.4 & mean_b <= 0.6 | mean_cml >= 0.4 & mean_cml <= 0.6, TRUE, FALSE)) %>%
  group_by(rank) %>% mutate(has_mid_beta = any(mid_beta == TRUE)) %>% ungroup() %>% 
    mutate(region_id = paste0(seqnames, ":", start, "-", end), type = "dmr") %>% distinct(region_id, has_mid_beta, type)) %>%
  ggplot(aes(x = type, fill = has_mid_beta)) +
  geom_bar()

rbind(optiles_all_beta %>% filter(region_id %in% filter(dtu.optiles, FDR < 0.05, abs(delta_b_minus_cml) > 0.2)$region_id) %>%
  filter(!is.na(mean_b), !is.na(mean_cml)) %>% mutate(mid_beta = ifelse(mean_b >= 0.4 & mean_b <= 0.6 | mean_cml >= 0.4 & mean_cml <= 0.6, TRUE, FALSE)) %>%
  group_by(region_id) %>% mutate(has_mid_beta = any(mid_beta == TRUE)) %>% ungroup() %>% 
  mutate(type = "dtu") %>% distinct(region_id, has_mid_beta, type),
  find_overlaps(as_granges(dmrcate_seq_dmr), as_granges(betas)) %>% data.frame() %>% 
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(mid_beta = ifelse(mean_b >= 0.4 & mean_b <= 0.6 | mean_cml >= 0.4 & mean_cml <= 0.6, TRUE, FALSE)) %>%
  group_by(rank) %>% mutate(has_mid_beta = any(mid_beta == TRUE)) %>% ungroup() %>% 
    mutate(region_id = paste0(seqnames, ":", start, "-", end), type = "dmr") %>% distinct(region_id, has_mid_beta, type)) %>%
  ggplot(aes(x = type, fill = has_mid_beta)) +
  geom_bar(position = "fill")
```
* actually - count all the groups - but for beta - low, med, high etc - and see the prop
* it can be for either
* oh wait if doing either - need to separate - coz low in one might be medium in the other

```{r, eval=FALSE}
rbind(optiles_all_beta %>% filter(region_id %in% filter(dtu.optiles, FDR < 0.05, abs(delta_b_minus_cml) > 0.2)$region_id) %>%
  filter(!is.na(mean_b), !is.na(mean_cml)) %>% 
    mutate(beta_b = case_when(mean_b < 0.2 ~ "x < 0.2",
                                mean_b > 0.8 ~ "x > 0.8",
                                mean_b >= 0.2 & mean_b <= 0.4 ~ "0.2 <= x <= 0.4",
                                mean_b > 0.4 & mean_b < 0.6 ~ "0.4 < x < 0.6",
                                TRUE ~ "0.6 <= x <= 0.8")) %>%
    mutate(beta_cml = case_when(mean_cml < 0.2 ~ "x < 0.2",
                                mean_cml > 0.8 ~ "x > 0.8",
                                mean_cml >= 0.2 & mean_cml <= 0.4 ~ "0.2 <= x <= 0.4",
                                mean_cml > 0.4 & mean_cml < 0.6 ~ "0.4 < x < 0.6",
                                TRUE ~ "0.6 <= x <= 0.8")) %>%
  group_by(region_id) %>% mutate(has_mid_beta = any(mid_beta == TRUE)) %>% ungroup() %>% 
  mutate(type = "dtu") %>% distinct(region_id, has_mid_beta, type),
  find_overlaps(as_granges(dmrcate_seq_dmr), as_granges(betas)) %>% data.frame() %>% 
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(mid_beta = ifelse(mean_b >= 0.4 & mean_b <= 0.6 | mean_cml >= 0.4 & mean_cml <= 0.6, TRUE, FALSE)) %>%
  group_by(rank) %>% mutate(has_mid_beta = any(mid_beta == TRUE)) %>% ungroup() %>% 
    mutate(region_id = paste0(seqnames, ":", start, "-", end), type = "dmr") %>% distinct(region_id, has_mid_beta, type)) %>%
  ggplot(aes(x = type, fill = has_mid_beta)) +
  geom_bar()
```

```{r, eval=FALSE}
optiles_all_beta %>% filter(region_id %in% filter(dtu.optiles, FDR < 0.05, abs(delta_b_minus_cml) > 0.2)$region_id) %>%
  filter(!is.na(mean_b), !is.na(mean_cml)) %>% 
    mutate(beta_b = case_when(mean_b < 0.2 ~ "x < 0.2",
                                mean_b > 0.8 ~ "x > 0.8",
                                mean_b >= 0.2 & mean_b <= 0.4 ~ "0.2 <= x <= 0.4",
                                mean_b > 0.4 & mean_b < 0.6 ~ "0.4 < x < 0.6",
                                TRUE ~ "0.6 <= x <= 0.8")) %>%
    mutate(beta_cml = case_when(mean_cml < 0.2 ~ "x < 0.2",
                                mean_cml > 0.8 ~ "x > 0.8",
                                mean_cml >= 0.2 & mean_cml <= 0.4 ~ "0.2 <= x <= 0.4",
                                mean_cml > 0.4 & mean_cml < 0.6 ~ "0.4 < x < 0.6",
                                TRUE ~ "0.6 <= x <= 0.8")) %>% 
  pivot_longer(cols = c(beta_b, beta_cml), names_to = "sample_group", values_to = "mean_beta_group") %>%
  ggplot(aes(x = sample_group, fill = mean_beta_group)) +
    geom_bar()
  
```

* ok this is not the plot I wanted to make
* maybe proportion?? - I'm not sure now
mutate(beta_b = case_when(mean_b < 0.2 ~ "x < 0.2",
                                mean_b > 0.8 ~ "x > 0.8",
                                mean_b >= 0.2 & mean_b <= 0.4 ~ "0.2 <= x <= 0.4",
                                mean_b > 0.4 & mean_b < 0.6 ~ "0.4 < x < 0.6",
                                TRUE ~ "0.6 <= x <= 0.8"))

* what proportion of each dmr / dtu? - like what fraction of the length?

```{r, eval=FALSE}
find_overlaps(as_granges(dmrcate_seq_dmr), as_granges(betas)) %>% data.frame() %>% 
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(mid_beta = ifelse(mean_b >= 0.4 & mean_b <= 0.6 | mean_cml >= 0.4 & mean_cml <= 0.6, TRUE, FALSE)) %>%
  group_by(rank) %>% mutate(total_cpg = n()) %>% group_by(rank, mid_beta) %>% mutate(n = n()) %>% ungroup() %>% group_by(rank) %>% 
  mutate(has_true = any(mid_beta == TRUE)) %>% mutate(n_false = ifelse(has_true == FALSE, 0, NA)) %>% ungroup() %>% distinct(rank, total_cpg, mid_beta, n, has_true, n_false) %>% 
  mutate(prop = ifelse(has_true == TRUE, n/total_cpg, 0/total_cpg))
  distinct(rank, total_cpg, has_true)
  mutate(n_false = ifelse(has_true == TRUE))
  group_by(rank, mid_beta) %>% summarise(n= n()) %>%
  ungroup() %>% group_by(rank) %>% mutate(prop = n/sum(n)) %>% mutate(prop2 = ifelse(any(mid_beta == TRUE)))
  ggplot(aes(x = has_mid_beta, fill = has_mid_beta)) +
  geom_bar()
```

```{r, eval=FALSE}
find_overlaps(as_granges(dmrcate_seq_dmr), as_granges(betas)) %>% data.frame() %>% 
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(mid_beta = ifelse(mean_b >= 0.4 & mean_b <= 0.6 | mean_cml >= 0.4 & mean_cml <= 0.6, TRUE, FALSE)) %>%
  group_by(rank) %>% mutate(total_cpg = n()) %>% ungroup() %>%
  group_by(rank) %>% summarise(prop_true = mean(mid_beta == TRUE), .groups = "drop") %>%
  complete(rank, fill = list(prop_true = 0)) %>%
  ggplot(aes(x = prop_true)) +
  geom_density()

find_overlaps(as_granges(dmrcate_seq_dmr), as_granges(betas)) %>% data.frame() %>% 
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(mid_beta = ifelse(mean_b >= 0.4 & mean_b <= 0.6 | mean_cml >= 0.4 & mean_cml <= 0.6, TRUE, FALSE)) %>%
  group_by(rank) %>% mutate(total_cpg = n()) %>% ungroup() %>%
  group_by(rank) %>% summarise(prop_true = mean(mid_beta == TRUE), .groups = "drop") %>%
  complete(rank, fill = list(prop_true = 0)) %>%
  ggplot(aes(x = prop_true)) +
  geom_histogram(bins = 50)
```
* ok so this is distribution of the proportion that mid range betas make up a dmr
* and mid range beta is if either condition has a mean mid range beta
* so they pop up - but they don't make up much of the dmr
* this could also be because dmrs are so long

* now do for dtu - would be cool if get more in a higher prop 

```{r, eval=FALSE}
find_overlaps(as_granges(filter(dtu.optiles, FDR < 0.05, abs(delta_b_minus_cml) > 0.2)), as_granges(betas)) %>% data.frame() %>% 
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(mid_beta = ifelse(mean_b >= 0.4 & mean_b <= 0.6 | mean_cml >= 0.4 & mean_cml <= 0.6, TRUE, FALSE)) %>%
  group_by(region_id) %>% mutate(total_cpg = n()) %>% ungroup() %>%
  group_by(region_id) %>% summarise(prop_true = mean(mid_beta == TRUE), .groups = "drop") %>%
  complete(region_id, fill = list(prop_true = 0)) %>%
  ggplot(aes(x = prop_true)) +
  geom_density()

find_overlaps(as_granges(filter(dtu.optiles, FDR < 0.05, abs(delta_b_minus_cml) > 0.2)), as_granges(betas)) %>% data.frame() %>% 
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(mid_beta = ifelse(mean_b >= 0.4 & mean_b <= 0.6 | mean_cml >= 0.4 & mean_cml <= 0.6, TRUE, FALSE)) %>%
  group_by(region_id) %>% mutate(total_cpg = n()) %>% ungroup() %>%
  group_by(region_id) %>% summarise(prop_true = mean(mid_beta == TRUE), .groups = "drop") %>%
  complete(region_id, fill = list(prop_true = 0)) %>%
  ggplot(aes(x = prop_true)) +
  geom_histogram(bins = 50)
```

* definitely different distributions here
* there is also the case of just way more significance with dtu - so scales are very much not the same


* do this again but for all the group ranges

```{r, eval=FALSE}
find_overlaps(as_granges(filter(dtu.optiles, FDR < 0.05, abs(delta_b_minus_cml) > 0.2)), as_granges(betas)) %>% data.frame() %>% 
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(mid_beta = ifelse(mean_b >= 0.4 & mean_b <= 0.6 | mean_cml >= 0.4 & mean_cml <= 0.6, TRUE, FALSE)) %>%
  group_by(region_id) %>% mutate(total_cpg = n()) %>% ungroup() %>%
  group_by(region_id) %>% summarise(prop_true = mean(mid_beta == TRUE), .groups = "drop") %>%
  complete(region_id, fill = list(prop_true = 0)) 
```


```{r, eval=FALSE}
optiles_all_beta %>% mutate(sig_dmr = ifelse(region_id %in% tiles_dmr$region_id, TRUE, FALSE), 
                            sig_dtu = ifelse(region_id %in% filter(dtu.optiles, FDR < 0.05, abs(delta_b_minus_cml) > 0.2)$region_id, TRUE, FALSE)) %>%
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(mid_beta = ifelse(mean_b >= 0.4 & mean_b <= 0.6 | mean_cml >= 0.4 & mean_cml <= 0.6, TRUE, FALSE)) %>%
  group_by(region_id) %>% mutate(has_mid_beta = any(mid_beta == TRUE)) %>% ungroup() %>% distinct(region_id, sig_dmr, sig_dtu, has_mid_beta) %>%
  ggplot(aes(x = sig_dtu, fill = has_mid_beta)) +
  geom_bar()
```



* ok big thought - it could be the 5 groups

* 3 groups will probably do it (hopefully)


```{r}
dtu_tiles_5_exon
```


```{r}
overlap2 <- find_overlaps(as_granges(filter(dtu_tiles_5_exon, FDR < 0.05)), as_granges(dmrcate_seq_dmr)) %>% data.frame()
```

```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>%
  ggplot(aes(x = overlap_dmr, fill = transcript)) +
  geom_bar()

dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>%
  ggplot(aes(x = overlap_dmr, fill = transcript)) +
  geom_bar(position = "fill")
```

```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>%
  ggplot(aes(x = transcript, fill = overlap_dmr)) +
  geom_bar()

dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>%
  ggplot(aes(x = transcript, fill = overlap_dmr)) +
  geom_bar(position = "fill")
```

```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>%
  filter(transcript == "0.4 < a < 0.6", overlap_dmr == FALSE)  
```

```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>%
  ggplot(aes(x = dtu_rank, y = transcript, colour = overlap_dmr)) +
  geom_jitter(alpha = 0.3)
```

```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>%
  ggplot(aes(x = dtu_rank, y = transcript, colour = overlap_dmr)) +
  geom_jitter(alpha = 0.3) +
  scale_x_continuous(limits = c(0, 20000))
```

```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>%
  filter(transcript == "0.4 < a < 0.6", overlap_dmr == FALSE, dtu_rank <= 20000) 
```

```{r}

```


* ok what are the betas for these regions

```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>%
  filter(transcript == "0.4 < a < 0.6", overlap_dmr == FALSE)  
```

```{r}
find_overlaps(dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>%
  filter(transcript == "0.4 < a < 0.6", overlap_dmr == FALSE) %>% as_granges(), as_granges(betas)) %>% data.frame() %>%
  ggplot(aes(x = mean_b)) +
  geom_density() +
  geom_density(aes(x = mean_cml), colour = "red")
```

```{r}
find_overlaps(dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>%
  filter(overlap_dmr == FALSE) %>% as_granges(), as_granges(betas)) %>% data.frame() %>%
  ggplot(aes(x = mean_b)) +
  geom_density() +
  geom_density(aes(x = mean_cml), colour = "red")

# beta distrib of tiles that are signif +  overlap dmr 
find_overlaps(dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>%
  filter(overlap_dmr == TRUE) %>% as_granges(), as_granges(betas)) %>% data.frame() %>%
  ggplot(aes(x = mean_b)) +
  geom_density() +
  geom_density(aes(x = mean_cml), colour = "red")
```
* not sure what this says

```{r}
overlap_beta <- find_overlaps(dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>% as_granges(), as_granges(betas)) %>% data.frame()

betas %>% mutate(signif = ifelse(pos %in% overlap_beta$pos, TRUE, FALSE)) %>%
  ggplot(aes(x = mean_b, colour = signif)) +
  geom_density() +
  geom_density(aes(x = mean_cml, colour = signif), linetype = "dashed")
```

```{r}
find_overlaps(dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>% as_granges(), as_granges(betas)) %>% data.frame() %>% mutate(mid_alpha = ifelse(transcript == "0.4 < a < 0.6", TRUE, FALSE)) %>%
  ggplot(aes(x = mean_b, colour = mid_alpha)) +
  geom_density() +
  facet_wrap(~overlap_dmr)
```

```{r}
find_overlaps(dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>% as_granges(), as_granges(betas)) %>% data.frame() %>% mutate(mid_alpha = ifelse(transcript == "0.4 < a < 0.6", TRUE, FALSE)) %>%
  ggplot(aes(x = mean_b, y = mean_cml, colour = transcript)) +
  geom_point(alpha = 0.1)
```

```{r}
find_overlaps(dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>% as_granges(), as_granges(betas)) %>% data.frame() %>% mutate(mid_alpha = ifelse(transcript == "0.4 < a < 0.6", TRUE, FALSE)) %>%
  ggplot(aes(x = mean_b, y = mean_cml, colour = transcript)) +
  geom_point(alpha = 0.1) +
  facet_wrap(~ transcript)
```

```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% mutate(n_trans = n()) %>% ungroup() %>%
  ggplot(aes(x = transcript, y = n_trans)) +
  geom_boxplot() +
  geom_violin()
```

```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% mutate(n_trans = n(), prop = 1/n()) %>% ungroup() %>% .[order(.$region_id),] 
```

* I want how common is each like transcript combo - like if low is sig - what prop of that has the others as sig - in same 
* ok not quite there - but got the proportion in a region of the thing

```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% mutate(n_trans = n(), prop = 1/n()) %>% ungroup() %>% .[order(.$region_id),] %>%
  ggplot(aes(x = transcript, fill = as.factor(prop))) +
  geom_bar()

dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% mutate(n_trans = n(), prop = 1/n()) %>% ungroup() %>% .[order(.$region_id),] %>%
  ggplot(aes(x = transcript, fill = as.factor(prop))) +
  geom_bar(position = "fill")
```

```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% mutate(n_trans = n(), prop = 1/n()) %>% ungroup() %>% .[order(.$region_id),] %>% 
  group_by(region_id) %>% filter(any(transcript == "0.4 < a < 0.6"))
  mutate()
```

```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% mutate(n_trans = n(), prop = 1/n()) %>% ungroup() %>% .[order(.$region_id),] %>% 
  group_by(region_id) %>% filter(any(transcript == "0.4 < a < 0.6")) %>%
  ggplot(aes(x = transcript)) +
  geom_bar()
```
nbm
```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% mutate(n_trans = n(), prop = 1/n()) %>% ungroup() %>% .[order(.$region_id),] %>% 
  group_by(region_id) %>% mutate(has_mid = ifelse(any(transcript == "0.4 < a < 0.6"), TRUE, FALSE)) %>%
  ggplot(aes(x = transcript, fill = has_mid)) +
  geom_bar()
  
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% group_by(region_id) %>% mutate(n_trans = n(), prop = 1/n()) %>% ungroup() %>% .[order(.$region_id),] %>% 
  group_by(region_id) %>% mutate(has_mid = ifelse(any(transcript == "0.4 < a < 0.6"), TRUE, FALSE)) %>%
  ggplot(aes(x = has_mid, fill = transcript)) +
  geom_bar()
```



```{r}
dtu_tiles_5_exon %>%
  distinct(region_id, transcript) %>%
  group_by(region_id) %>%
  summarise(cat_list = list(unique(transcript)), .groups = "drop") %>%
  mutate(pairs = purrr::map(cat_list, ~ {
    if (length(.x) < 2) return(NULL)
    t(combn(sort(.x), 2))
  })) %>%
  select(pairs) %>%
  unnest(pairs) %>%
  count(V1, V2, name = "n")
```

```{r}
dtu_tiles_5_exon %>% distinct(region_id, transcript) %>%
  count(transcript, name = "group_count")
```


```{r}
betas %>%
  ggplot(aes(x = mean_b)) +
  geom_density()

betas %>% filter(mean_b == 0.5)
```


```{r}
betas <- all_samples %>% distinct(seqnames, start, sample, beta) %>% pivot_wider(names_from = sample, values_from = beta)

betas <- betas %>% mutate(mean_b = rowMeans(across(c(b1, b2)), na.rm = TRUE), 
                 mean_cml = rowMeans(across(c(cml1, cml2)), na.rm = TRUE),
                 beta_delta_b_minus_cml = mean_b - mean_cml) %>% data.frame()
betas$end <- betas$start

betas$pos <- paste0(betas$seqnames, "-", betas$start)
```

```{r}
optiles_all_beta <- find_overlaps(as_granges(optiles_all), as_granges(betas)) %>% data.frame()
optiles_all_beta <- optiles_all_beta %>% mutate(alpha_b = bin_delta[match(.$region_id, bin_delta$region_id), "B"],
                                                alpha_cml = bin_delta[match(.$region_id, bin_delta$region_id), "CML"],
                                                alpha_delta_b_minus_cml = bin_delta[match(.$region_id, bin_delta$region_id), "delta_b_minus_cml"],
                                                alpha_b = as.double(alpha_b), alpha_cml = as.double(alpha_cml), alpha_delta_b_minus_cml = as.double(alpha_delta_b_minus_cml))
```

```{r}
interest <- dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE)) %>%
  filter(transcript == "0.4 < a < 0.6", overlap_dmr == FALSE) %>% .$region_id
```

```{r}
optiles_all_beta %>% mutate(mid_not_sig_in_dmr = ifelse(region_id %in% interest, TRUE, FALSE)) %>%
  filter(mid_not_sig_in_dmr == TRUE) %>%
  ggplot(aes(x = mean_b, y = mean_cml)) +
  geom_point(alpha = 0.2)

optiles_all_beta %>% mutate(mid_not_sig_in_dmr = ifelse(region_id %in% interest, TRUE, FALSE)) %>%
  filter(mid_not_sig_in_dmr == FALSE) %>%
  ggplot(aes(x = mean_b, y = mean_cml)) +
  geom_point(alpha = 0.2)
```

```{r}
optiles_all_beta %>% mutate(mid_not_sig_in_dmr = ifelse(region_id %in% interest, TRUE, FALSE)) %>%
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(mid_beta = ifelse(mean_b >= 0.4 & mean_b <= 0.6 | mean_cml >= 0.4 & mean_cml <= 0.6, TRUE, FALSE)) %>%
  ggplot(aes(x = mid_not_sig_in_dmr, fill = mid_beta)) +
  geom_bar()

optiles_all_beta %>% mutate(mid_not_sig_in_dmr = ifelse(region_id %in% interest, TRUE, FALSE)) %>%
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(mid_beta = ifelse(mean_b >= 0.4 & mean_b <= 0.6 | mean_cml >= 0.4 & mean_cml <= 0.6, TRUE, FALSE)) %>%
  ggplot(aes(x = mid_not_sig_in_dmr, fill = mid_beta)) +
  geom_bar(position = "fill")
```

```{r}
optiles_all_beta %>% mutate(mid_not_sig_in_dmr = ifelse(region_id %in% interest, TRUE, FALSE)) %>%
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(beta_group_b = case_when(mean_b > 0.4 & mean_b < 0.6 ~ "mid:0.4-0.6", 
                                mean_b < 0.2 ~ "low:<0.2",
                                mean_b > 0.8 ~ "high:>0.8",
                                mean_b >= 0.2 & mean_b <= 0.4 ~ "low-med:0.2-0.4",
                                TRUE ~ "med-high:0.6-0.8")) %>%
  mutate(beta_group_cml = case_when(mean_cml > 0.4 & mean_cml < 0.6 ~ "mid:0.4-0.6", 
                                mean_cml < 0.2 ~ "low:<0.2",
                                mean_cml > 0.8 ~ "high:>0.8",
                                mean_cml >= 0.2 & mean_cml <= 0.4 ~ "low-med:0.2-0.4",
                                TRUE ~ "med-high:0.6-0.8")) %>%
  ggplot(aes(x = beta_group_b, y = beta_group_cml, colour = mid_not_sig_in_dmr)) +
  geom_count()
```

```{r}
optiles_all_beta %>% mutate(mid_not_sig_in_dmr = ifelse(region_id %in% interest, TRUE, FALSE)) %>%
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(beta_group_b = case_when(mean_b > 0.4 & mean_b < 0.6 ~ "mid:0.4-0.6", 
                                mean_b < 0.2 ~ "low:<0.2",
                                mean_b > 0.8 ~ "high:>0.8",
                                mean_b >= 0.2 & mean_b <= 0.4 ~ "low-med:0.2-0.4",
                                TRUE ~ "med-high:0.6-0.8")) %>%
  mutate(beta_group_cml = case_when(mean_cml > 0.4 & mean_cml < 0.6 ~ "mid:0.4-0.6", 
                                mean_cml < 0.2 ~ "low:<0.2",
                                mean_cml > 0.8 ~ "high:>0.8",
                                mean_cml >= 0.2 & mean_cml <= 0.4 ~ "low-med:0.2-0.4",
                                TRUE ~ "med-high:0.6-0.8")) %>%
  pivot_longer(cols = c(beta_group_b, beta_group_cml), names_to = "type", values_to = "beta_group") %>%
  ggplot(aes(x = mid_not_sig_in_dmr, fill = beta_group)) +
  geom_bar()

optiles_all_beta %>% mutate(mid_not_sig_in_dmr = ifelse(region_id %in% interest, TRUE, FALSE)) %>%
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(beta_group_b = case_when(mean_b > 0.4 & mean_b < 0.6 ~ "mid:0.4-0.6", 
                                mean_b < 0.2 ~ "low:<0.2",
                                mean_b > 0.8 ~ "high:>0.8",
                                mean_b >= 0.2 & mean_b <= 0.4 ~ "low-med:0.2-0.4",
                                TRUE ~ "med-high:0.6-0.8")) %>%
  mutate(beta_group_cml = case_when(mean_cml > 0.4 & mean_cml < 0.6 ~ "mid:0.4-0.6", 
                                mean_cml < 0.2 ~ "low:<0.2",
                                mean_cml > 0.8 ~ "high:>0.8",
                                mean_cml >= 0.2 & mean_cml <= 0.4 ~ "low-med:0.2-0.4",
                                TRUE ~ "med-high:0.6-0.8")) %>%
  pivot_longer(cols = c(beta_group_b, beta_group_cml), names_to = "type", values_to = "beta_group") %>%
  ggplot(aes(x = mid_not_sig_in_dmr, fill = beta_group)) +
  geom_bar(position = "fill")
```

```{r}
optiles_all_beta %>% mutate(mid_not_sig_in_dmr = ifelse(region_id %in% interest, TRUE, FALSE)) %>%
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(beta_group_b = case_when(mean_b > 0.4 & mean_b < 0.6 ~ "mid:0.4-0.6", 
                                mean_b < 0.2 ~ "low:<0.2",
                                mean_b > 0.8 ~ "high:>0.8",
                                mean_b >= 0.2 & mean_b <= 0.4 ~ "low-med:0.2-0.4",
                                TRUE ~ "med-high:0.6-0.8")) %>%
  mutate(beta_group_cml = case_when(mean_cml > 0.4 & mean_cml < 0.6 ~ "mid:0.4-0.6", 
                                mean_cml < 0.2 ~ "low:<0.2",
                                mean_cml > 0.8 ~ "high:>0.8",
                                mean_cml >= 0.2 & mean_cml <= 0.4 ~ "low-med:0.2-0.4",
                                TRUE ~ "med-high:0.6-0.8")) %>%
  pivot_longer(cols = c(beta_group_b, beta_group_cml), names_to = "type", values_to = "beta_group") %>%
  ggplot(aes(x = beta_group, fill = mid_not_sig_in_dmr)) +
  geom_bar()

optiles_all_beta %>% mutate(mid_not_sig_in_dmr = ifelse(region_id %in% interest, TRUE, FALSE)) %>%
  filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(beta_group_b = case_when(mean_b > 0.4 & mean_b < 0.6 ~ "mid:0.4-0.6", 
                                mean_b < 0.2 ~ "low:<0.2",
                                mean_b > 0.8 ~ "high:>0.8",
                                mean_b >= 0.2 & mean_b <= 0.4 ~ "low-med:0.2-0.4",
                                TRUE ~ "med-high:0.6-0.8")) %>%
  mutate(beta_group_cml = case_when(mean_cml > 0.4 & mean_cml < 0.6 ~ "mid:0.4-0.6", 
                                mean_cml < 0.2 ~ "low:<0.2",
                                mean_cml > 0.8 ~ "high:>0.8",
                                mean_cml >= 0.2 & mean_cml <= 0.4 ~ "low-med:0.2-0.4",
                                TRUE ~ "med-high:0.6-0.8")) %>%
  pivot_longer(cols = c(beta_group_b, beta_group_cml), names_to = "type", values_to = "beta_group") %>%
  ggplot(aes(x = beta_group, fill = mid_not_sig_in_dmr)) +
  geom_bar(position = "fill")
```

mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8"))


filter(!is.na(mean_b), !is.na(mean_cml)) %>%
  mutate(mid_beta = ifelse(mean_b >= 0.4 & mean_b <= 0.6 | mean_cml >= 0.4 & mean_cml <= 0.6, TRUE, FALSE)) %>%
  group_by(region_id) %>% mutate(has_mid_beta = any(mid_beta == TRUE)) %>% ungroup() %>% distinct(region_id, sig_dmr, sig_dtu, has_mid_beta) %>%
  mutate(sig_dmr = ifelse(sig_dmr == TRUE, "dmr_sig", "dmr_not_sig"), sig_dtu = ifelse(sig_dtu == TRUE, "dtu_sig", "dtu_not_sig")) %>%
  pivot_longer(cols = c(sig_dmr, sig_dtu), names_to = "sig", values_to = "is_sig") %>% .[,c(1,2,4)] %>% distinct() %>%
  ggplot(aes(x = is_sig, fill = has_mid_beta)) +
  geom_bar(position = "fill")


```{r}
overlap
```


```{r}
dtu_tiles_5_exon %>% filter(FDR < 0.05) %>% mutate(overlap_dmr = ifelse(region_id %in% overlap2$region_id, TRUE, FALSE))
```



