---
title: "emseqBetaAnalysis"
author: "caitlinpage"
date: "2026-07-29"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

## Introduction

```{r}
library(BSgenome.Hsapiens.UCSC.hg38)
library(data.table)
library(plyranges)
library(tidyr)
library(stringr)
library(dplyr)
library(ggplot2)
```

# Obtain beta values

```{r}
seqs <- seqnames(BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38)[1:22]

cg_sites <- lapply(seqs, function(x) {
  cbind(data.frame(Biostrings::matchPattern("CG", BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38[[x]])),
        seqnames = x
  )}) %>% bind_rows()
```

```{r, eval=FALSE}
all_samples <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_all_healthy.rds")
all_samples <- all_samples %>%
  mutate(sample = case_when(sample == "a1" ~ "healthy_785",
                            sample == "a2" ~ "healthy_036",
                            sample == "a3" ~ "healthy_787",
                            sample == "b1" ~ "healthy_786",
                            sample == "b2" ~ "healthy_037",
                            TRUE ~ "healthy_789"))
```


```{r}
betas_for_wgbs <- function(reads) {
  # Split entire columns at once (vectorised, not row-wise)
  meth_split <- strsplit(reads$meth_pattern, "")
  pos_split  <- lapply(strsplit(reads$genom_positions, ","), as.double)
  n          <- lengths(meth_split)   # CpGs per read

  # Unnest with rep/unlist — much faster than tidyr::unnest
  dt <- data.table(
    sample      = rep(reads$sample,   n),
    seqnames    = rep(reads$seqnames, n),
    start       = unlist(pos_split,   use.names = FALSE),
    meth_status = unlist(meth_split,  use.names = FALSE)
  )

  # Count M/U and compute beta in one grouped pass
  result <- dt[, {
    m <- sum(meth_status == "M")
    u <- sum(meth_status == "U")
    list(meth_count_pos = m, unmeth_count_pos = u, beta = m / (m + u))
  }, keyby = .(sample, seqnames, start)]

  # Collapse across samples (matches original behaviour)
  unique(result[, .(sample, seqnames, start, meth_count_pos, unmeth_count_pos, beta)])
}
```


```{r, eval=FALSE}
betas <- betas_for_wgbs(all_samples)
saveRDS(betas, "/researchers/caitlin.page/cf_nano/r_output/emseq_betas_healthy.rds")
```


```{r, eval=FALSE}
all_samples <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.rds")
betas <- betas_for_wgbs(all_samples)
saveRDS(betas, "/researchers/caitlin.page/cf_nano/r_output/emseq_betas_cancer.rds")
```


```{r}
betas <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_betas_healthy.rds")
```

```{r}
b <- betas %>% data.frame()
b <- b %>% filter(seqnames %in% seqs)
```

```{r}
b$cov <- b$meth_count_pos + b$unmeth_count_pos
```

```{r}
b <- b[,c(1:5,7)] %>% pivot_wider(names_from = sample, values_from = c(meth_count_pos, unmeth_count_pos, cov))
b[is.na(b)] <- 0
```

```{r}
samples <- unique(betas$sample)
```


```{r}
ranges <- b[,1:2]
meth <- b[,3:8]
colnames(meth) <- samples
cov <- b[,15:20]
colnames(cov) <- samples
```


## chr1
```{r}
betas <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_betas_healthy.rds")
betas_c <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_betas_cancer.rds")
b <- rbind(betas %>% data.frame(), betas_c %>% data.frame())
b <- b %>% filter(seqnames == "chr1")
b$cov <- b$meth_count_pos + b$unmeth_count_pos
cols <- c("sample", "seqnames", "start", "meth_count_pos", "unmeth_count_pos", "cov")
b <- b[,cols] %>% pivot_wider(names_from = sample, values_from = c(meth_count_pos, unmeth_count_pos, cov))
b[is.na(b)] <- 0
saveRDS(b, "/researchers/caitlin.page/cf_nano/r_output/emseq_betas_chr1.rds")
```

# dmr detection steps

```{r, eval=FALSE}
betas <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_betas_healthy.rds")
betas_c <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_betas_cancer.rds")
b <- rbind(betas %>% data.frame(), betas_c %>% data.frame())
b <- b %>% filter(seqnames == "chr1")
b$cov <- b$meth_count_pos + b$unmeth_count_pos
cols <- c("sample", "seqnames", "start", "meth_count_pos", "unmeth_count_pos", "cov")
b <- b[,cols] %>% pivot_wider(names_from = sample, values_from = c(meth_count_pos, unmeth_count_pos, cov))
b[is.na(b)] <- 0

samples <- c(unique(betas$sample), unique(betas_c$sample))
sample_stuff <- c(Type = c("healthy", "healthy", "healthy", "healthy", "healthy", "healthy", "cancer", "cancer", "cancer", "cancer", "cancer", "cancer")) %>% data.frame()
rownames(sample_stuff) <- samples

ranges <- b[,1:2]
meth <- b[,3:14]
colnames(meth) <- samples
cov <- b[,27:38]
colnames(cov) <- samples

#### bsseq
## make BSseq object
bseq <- bsseq::BSseq(M = as.matrix(meth),
                     Cov = as.matrix(cov), pData = sample_stuff,
                     chr = ranges$seqnames, pos = as.numeric(ranges$start),
                     sampleNames = samples)
bseq <- sort(bseq)
bseq <- bseq[which(rowSums(getCoverage(bseq) >= 10) == ncol(bseq)), ] 
param <- BiocParallel::SnowParam(workers = 8, type = "SOCK")
bs_batch1 <- BSmooth(bseq[, 1:6],  BPPARAM = param)
bs_batch2 <- BSmooth(bseq[, 7:12], BPPARAM = param)

# Combine afterwards
bsmooth_obj <- bsseq::combine(bs_batch1, bs_batch2)

#bsmooth_obj <- BSmooth(BSseq = bseq, verbose = TRUE)

bsmooth_obj <- bsmooth_obj[keep,]
saveRDS(bsmooth_obj, "/researchers/caitlin.page/cf_nano/r_output/bsmooth_obj_emseq_chr1.rds")

### dmrcate
tissue <- factor(c("healthy", "healthy", "healthy", "healthy", "healthy", "healthy", "cancer", "cancer", "cancer", "cancer", "cancer", "cancer"), levels = c("healthy", "cancer"))

#Regular matrix design
design <- model.matrix(~tissue)
methdesign <- edgeR::modelMatrixMeth(design)

#failed here because no significance - good
seq_annot <- sequencing.annotate(bsmooth_obj, methdesign, all.cov = TRUE,
                                 coef = 14, fdr=0.05)

seq_dmrcate.res <- dmrcate(seq_annot, C=2, min.cpgs = 5)

dmrcate_seq_dmr <- extractRanges(seq_dmrcate.res, genome = "hg38")
dmrcate_seq_dmr <- dmrcate_seq_dmr %>% data.frame()
dmrcate_seq_anno <- seq_annot@ranges %>% data.frame()

saveRDS(dmrcate_seq_dmr, "/researchers/caitlin.page/cf_nano/r_output/dmr_emseq_chr1.rds")
saveRDS(dmrcate_seq_anno, "/researchers/caitlin.page/cf_nano/r_output/dm_emseq_chr1.rds")
```


# we have DMRS!!!!!

```{r}
betas_h$cov <- betas_h$meth_count_pos + betas_h$unmeth_count_pos
betas_h <- betas_h %>% filter(seqnames %in% c("chr1", "chr2", "chr3", "chr4", "chr5", "chr6", "chr7", "chr8", "chr9", "chr10",
          "chr11", "chr12", "chr13", "chr14", "chr15", "chr16", "chr17", "chr18", "chr19",
          "chr20", "chr21", "chr22"))

betas_h %>%
  ggplot(aes(x= cov)) +
  geom_histogram(bins = 75)
```

```{r}
betas_h %>%
  ggplot(aes(x= cov)) +
  geom_histogram(bins = 75) +
  scale_x_continuous(limits =c (0, 150))
```

```{r}
optiles_all <- readRDS("/researchers/caitlin.page/cf_nano/r_output/optiles_all_emseq.rds")
```

```{r}
dmr_emseq <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dmr_emseq_strict.rds")

dmr_emseq %>% filter(min_smoothed_fdr < 0.05) %>% nrow()
```

```{r}
dmr_emseq <- dmr_emseq %>% .[order(.$min_smoothed_fdr),] %>% mutate(dmr_rank = 1:n())
```

```{r}
dmr_dtu <- find_overlaps(as_granges(dmr_emseq), as_granges(dtu_emseq)) %>% data.frame()
dmr_tiles <- find_overlaps(as_granges(dmr_emseq), as_granges(optiles_all)) %>% data.frame()

ggVennDiagram::ggVennDiagram(list(dmr = dmr_tiles %>% filter(min_smoothed_fdr < 0.05) %>% .$region_id,
                                  dtu = dtu_emseq %>% filter(FDR < 0.05) %>% .$region_id))
```
```{r}
dmr_dtu
```

```{r}
dmr_dtu %>%
  ggplot(aes(x = dmr_rank, y = dtu_rank)) +
  geom_point(alpha = 0.3)
```

```{r}
dmr_dtu %>% filter(FDR < 0.05, min_smoothed_fdr < 0.05) %>% group_by(dmr_rank) %>% summarise(n_tiles = n()) %>%
  ggplot(aes(x= as.factor(n_tiles))) +
  geom_bar()
```
```{r}
dmr_dtu %>% group_by(dmr_rank) %>% summarise(n_tiles = n()) %>%
  ggplot(aes(x= as.factor(n_tiles))) +
  geom_bar()
```

```{r}
dmr_beta <- find_overlaps(as_granges(dmr_emseq), as_granges(mutate(all_b, seqnames = seqnames.x, start = start.x, width = 1))) %>% data.frame()

```

```{r}
dmr_emseq
tail(dmr_emseq)
```
```{r}
dmr_emseq 
```

```{r}
all_b %>% filter(seqnames.x == dmr_emseq[1,]$seqnames, start.x >= 147077566, start.x <= 147080214) %>%
  tidyr::pivot_longer(cols = c(cancer_0223, cancer_0224, cancer_0225, cancer_0454, cancer_0456, cancer_0457, healthy_036, healthy_037, healthy_785, healthy_786, healthy_787, healthy_789), names_to = "sample", values_to = "beta") %>% mutate(type = substr(sample, 1, 1)) %>%
  ggplot(aes(x = start.x, y = beta, colour = type)) +
  geom_point(alpha = 0.3)
```

```{r}
?substr
```

```{r}
dmr_dtu %>% filter(FDR < 0.05, min_smoothed_fdr < 0.05) %>% group_by(dtu_rank) %>% summarise(n_dmr = n()) %>%
  ggplot(aes(x= as.factor(n_dmr))) +
  geom_bar()
```


```{r}
list_plot <- list()
for(i in 1:28) {
  
  p_data <- dmr_dtu %>% filter(FDR < 0.05, min_smoothed_fdr < 0.05) %>% .[order(.$dtu_rank),] %>% .[i,] %>% 
  tidyr::separate(region_id, into = c("chr", "tile_start", "tile_end"), sep = "[:-]") %>% 
  mutate(tile_start = as.integer(tile_start), tile_end = as.integer(tile_end))
p <- p_data %>%
  ggplot() +
  geom_segment(aes(x = start, xend = end, y = 1, yend = 1), linewidth = 2) +
  geom_segment(aes(x = tile_start, xend = tile_end, y = 2, yend = 2), linewidth = 2) +
  geom_text(aes(x = min(start), y = 1.2, label = dmr_rank),
          hjust = 1,    # right-align so all labels end at the same x
          nudge_x = -5) +
coord_cartesian(clip = "off") +   # allows text to render outside the plot area
theme(plot.margin = margin(5, 5, 5, 60)) +
  geom_text(aes(x = tile_start, y = 1.8, label = dtu_rank)) +
  ggtitle(gsub("^chr", "", unique(p_data$chr)))

list_plot <- list(list_plot, p)
}
list_plot
```
# top 28 dmrs
```{r}
list_plot <- list()
for(i in 1:28) {
  
  p_data <- dmr_dtu %>% filter(FDR < 0.05, min_smoothed_fdr < 0.05) %>% .[order(.$dmr_rank),] %>% .[i,] %>% 
  tidyr::separate(region_id, into = c("chr", "tile_start", "tile_end"), sep = "[:-]") %>% 
  mutate(tile_start = as.integer(tile_start), tile_end = as.integer(tile_end))
p <- p_data %>%
  ggplot() +
  geom_segment(aes(x = start, xend = end, y = 1, yend = 1), linewidth = 2) +
  geom_segment(aes(x = tile_start, xend = tile_end, y = 2, yend = 2), linewidth = 2) +
  geom_text(aes(x = min(start), y = 1.2, label = dmr_rank),
          hjust = 1,    # right-align so all labels end at the same x
          nudge_x = -5) +
coord_cartesian(clip = "off") +   # allows text to render outside the plot area
theme(plot.margin = margin(5, 5, 5, 60)) +
  geom_text(aes(x = tile_start, y = 1.8, label = dtu_rank)) +
  ggtitle(gsub("^chr", "", unique(p_data$chr)))

list_plot <- list(list_plot, p)
}
list_plot
```

```{r}
dmr_dtu <- dmr_dtu %>% mutate(gene_pos = paste0(seqnames, ":", start, "-", end))
dmr_dtu_genes <- find_overlaps(genes_txdb, as_granges(dmr_dtu)) %>% data.frame()
```


```{r}
list_plot <- list()
for(i in 1:18) {
  
  tile_sig <- dtu_gene %>% filter(FDR < 0.05) %>% 
    group_by(gene_id) %>% mutate(n_sig_regions=dplyr::n()) %>% ungroup() %>%
    filter(n_sig_regions > 3) %>% .[order(.$n_sig_regions, decreasing = TRUE),] %>% .[order(.$dtu_rank),] %>% dplyr::distinct(gene_id, dtu_rank)
  
  p_data <- dtu_gene %>% filter(gene_id == tile_sig$gene_id[i]) %>% 
  tidyr::separate(region_id, into = c("chr", "tile_start", "tile_end"), sep = "[:-]") %>% 
    tidyr::separate(gene_pos, into = c("chr2", "dmr_start", "dmr_end"), sep = "[:-]") %>%
  mutate(tile_start = as.integer(tile_start), tile_end = as.integer(tile_end),
         dmr_start = as.integer(dmr_start), dmr_end = as.integer(dmr_end),
         y_pos = dplyr::row_number())
p <- p_data %>%
  ggplot() +
  geom_segment(data = filter(p_data, strand == "-"), aes(x = start, xend = end, y = 0, yend = 0), linewidth = 2,
               arrow = arrow(length = unit(0.2, "cm"), type = "closed", ends = "first")) +
  geom_segment(data = filter(p_data, strand == "+"), aes(x = start, xend = end, y = 0, yend = 0), linewidth = 2,
               arrow = arrow(length = unit(0.2, "cm"), type = "closed", ends = "last")) +
  geom_segment(aes(x = tile_start, xend = tile_end, y = y_pos, yend = y_pos), linewidth = 2) +
  geom_segment(aes(x = tile_start, xend = tile_end, y = y_pos, yend = y_pos), linewidth = 2)
  geom_text(aes(x = min(start), y = 0.2, label = gene_id),
          hjust = 1,    # right-align so all labels end at the same x
          nudge_x = -10000) +
coord_cartesian(clip = "off") +   # allows text to render outside the plot area
theme(plot.margin = margin(5, 5, 5, 60)) +
  geom_text(aes(x = tile_start, y = y_pos, label = dtu_rank)) +
  ggtitle(gsub("^chr", "", unique(p_data$chr)))

list_plot <- list(list_plot, p)
}
list_plot
```

####

```{r}
dmr_emseq[1,]
betas
```

```{r}
betas <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_betas_healthy.rds")
betas_c <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_betas_cancer.rds")
```



* want it with the alpha bands
* which I think is just doing it with alpha?
* do it with beta firstmmm
```{r}
b
```

m
```{r}
betas <- betas %>% data.frame() %>% filter(seqnames == "chr1") %>% .[,c(1:3,6)] %>%
  pivot_wider(names_from = sample, values_from = beta)

betas_c <- betas_c %>% data.frame() %>% filter(seqnames == "chr1") %>% .[,c(1:3,6)] %>%
  pivot_wider(names_from = sample, values_from = beta)
betas$pos <- paste0(betas$seqnames, ":", betas$start)
betas_c$pos <- paste0(betas_c$seqnames, ":", betas_c$start)
betas$type <- "healthy"
betas_c$type <- "cancer"
all_b <- full_join(betas, betas_c, by = "pos")
```
```{r}
dmr_emseq[1,]
```

```{r}
all_b %>% filter(seqnames.x == dmr_emseq[1,]$seqnames, start.x >= 147077566, start.x <= 147080214) %>%
  tidyr::pivot_longer(cols = c(cancer_0223, cancer_0224, cancer_0225, cancer_0454, cancer_0456, cancer_0457, healthy_036, healthy_037, healthy_785, healthy_786, healthy_787, healthy_789), names_to = "sample", values_to = "beta") %>% mutate(type = substr(sample, 1, 1)) %>%
  ggplot(aes(x = start.x, y = beta, colour = type)) +
  geom_point(alpha = 0.3)
```

```{r}
find_overlaps(as_granges(dmr_emseq[1,]), as_granges(dtu_emseq))
```
* overlap 2 signif tiles
em_healthy_785$sample <- "a1"
em_healthy_036$sample <- "a2"
em_healthy_787$sample <- "a3"
em_healthy_786$sample <- "b1"
em_healthy_037$sample <- "b2"
em_healthy_789$sample <- "b3"
```{r}
dmr_emseq <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dmr_emseq_strict.rds")
```

```{r}
emseq_reads_tiles_chr1 <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_reads_tiles_chr1.rds")
emseq_reads_tiles_chr1 %>% distinct(sample)
emseq_reads_tiles_chr1 <- emseq_reads_tiles_chr1 %>% mutate(sample = case_when(sample == "a1" ~ "healthy_785",
                                                     sample == "a2" ~ "healthy_036",
                                                     sample == "a3" ~ "healthy_787",
                                                     sample == "b1" ~ "healthy_786",
                                                     sample == "b2" ~ "healthy_037",
                                                     sample == "b3" ~ "healthy_789", TRUE ~ sample))
emseq_reads_tiles_chr1$type <- sub("_.*", "", emseq_reads_tiles_chr1$sample)
```
```{r}
betas
```

```{r}
b <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_betas_chr1.rds")
```


```{r}
dmr_reads <- find_overlaps(as_granges(dmr_emseq[1,]), as_granges(emseq_reads_tiles_chr1)) %>% data.frame()
dmr_reads %>% distinct(region_id)
```


Plot layers

reads - split by type?

then betas

then dmr? 

then genes? etc

```{r}
rrbs_plot5 <- function(read_data, read_ids, cpgs = cg_sites, jit_amount = 0.1) {
  set.seed(42) # For reproducibility
  # Calculate a single jitter amount per row (segment)
  read_data <- read_data %>% dplyr::filter(read_id %in% read_ids)
  jitter_amount <- runif(nrow(read_data), -jit_amount, jit_amount)
  read_data$alpha_jit <- read_data$alpha + jitter_amount
  read_data <- data.frame(read_data)

  # Add jittered coordinates to new columns in the data frame
  long_data <- read_data %>%
    mutate(meth_status = strsplit(meth_pattern, ""), start = strsplit(genom_positions, ",")) %>%
    unnest(meth_status, start) %>%
    mutate(start = as.double(start))
  
  chr <- unique(read_data$seqnames)
  min_start <- min(read_data$read_start)
  max_end <- max(read_data$read_end)
  
  read_data <- read_data %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8")))
  
  plot <- read_data %>%
    ggplot() +
    geom_segment(aes(x = read_start, xend = read_end, y = alpha_jit, yend = alpha_jit, colour = alpha_group)) +
    geom_point(data = long_data,
               aes(x = start, y = alpha_jit, fill = meth_status, colour = alpha_group), shape = 21) +
    scale_fill_manual(values = c("U" = "white", "M" = "black")) +
    facet_wrap(~ type, nrow=2) +
    labs(title = paste0(chr, ":", min_start, "-", max_end)) 
  plot
}

reads <- distinct(dmr_reads, read_id) %>% .$read_id
rrbs_plot5(dmr_reads, reads)
```

```{r}
dmr_reads
```


```{r}
reads <- distinct(dmr_reads, read_id) %>% .$read_id
```


```{r}
geom_beta <- function(all_b, plot.space=1, plot.height=1) {
    structure(list(all_b = all_b, plot.space = plot.space,
        plot.height = plot.height),
        class = "beta"
    )
}

ggplot_add.beta <- function(object, plot, object_name) {
plot2 <- plot
    while (inherits(plot2, "patchwork")) {
        plot2 <- plot2[[1]]
    }
    plot.data <- plot2$labels$title
    plot.data <- stringr::str_split_1(plot.data, ":")
    # prepare plot range
    # the plot region are not normal, so start is minimum value
    plot.chr <- plot.data[1]
    plot.data <- stringr::str_split_1(plot.data[2], "-")
    plot.region.start <- plot.data[1] %>% as.numeric()
    plot.region.end <- plot.data[2] %>% as.numeric()
    
    all_b <- object$all_b
    plot.space <- object$plot.space
    plot.height <- object$plot.height
    
    all_b <- all_b %>%
  tidyr::pivot_longer(cols = c(cancer_0223, cancer_0224, cancer_0225, cancer_0454, cancer_0456, cancer_0457, healthy_036, healthy_037, healthy_785, healthy_786, healthy_787, healthy_789), names_to = "sample", values_to = "beta") %>% mutate(type = substr(sample, 1, 1)) %>% filter(seqnames.x %in% plot.chr,
                                start.x >= plot.region.start,
                                start.x <= plot.region.end)
    
    b_plot <- all_b %>%
  ggplot(aes(x = start.x, y = beta, colour = type)) +
  geom_point(alpha = 0.3) 
  #    facet_wrap(~type, nrow = 2)
    
    patchwork::wrap_plots(plot + ggplot2::theme(plot.margin = ggplot2::margin(t = plot.space, b = plot.space)),
        b_plot,
        ncol = 1, heights = c(1, plot.height)
    )
}
#rrbs_plot5(dmr_reads, reads) +
#  geom_beta(all_b)
```
```{r}
all_b
```



*dau
```{r}
dau_tiles <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_emseq_cancer_healthy.rds")
```

```{r}
geom_dau <- function(
    dau.df=NULL, dau.label=FALSE, dau.color="black", dau.size=5,
    plot.space=0.1, plot.height=0.05) {
    structure(
        list(
            dau.df = dau.df, dau.label = dau.label, dau.color = dau.color, dau.size = dau.size,
            plot.space = plot.space, plot.height = plot.height
        ),
        class = "dau"
    )
}


#' @export
ggplot_add.dau <- function(object, plot, object_name) {
    if (!is.data.frame(object$dau.df)) {
        stop("data.frame of daus is required")
    }
    plot2 <- plot
    while (inherits(plot2, "patchwork")) {
        plot2 <- plot2[[1]]
    }
    plot.data <- plot2$labels$title
    plot.data <- stringr::str_split_1(plot.data, ":")

    plot.chr <- plot.data[1]
    plot.data <- stringr::str_split_1(plot.data[2], "-")
    plot.region.start <- plot.data[1] %>% as.numeric()
    plot.region.end <- plot.data[2] %>% as.numeric()

    # get parameters
    dau.df <- object$dau.df
    dau.label <- object$dau.label
    dau.color <- object$dau.color
    dau.size <- object$dau.size
    plot.space <- object$plot.space
    plot.height <- object$plot.height

    dau.df <- dau.df %>% filter(seqnames %in% plot.chr,
                                start >= plot.region.start,
                                end <= plot.region.end, 
                                FDR < 0.05) 
  
    dau.plot <- dau.df %>%
            ggplot2::ggplot(ggplot2::aes(x = .data$start, y = 1)) +
            ggplot2::geom_segment(
                mapping = ggplot2::aes(
                    x = .data$start, y = 1,
                    xend = .data$end, yend = 1
                ),
                linewidth = dau.size, color = dau.color
            ) +
      labs(y = "dau") +
      ..peakGatcTheme(margin.len = plot.space, x.range = c(plot.region.start, plot.region.end))

    # assemble plot
    patchwork::wrap_plots(plot + ggplot2::theme(plot.margin = ggplot2::margin(t = plot.space, b = plot.space)),
        dau.plot,
        ncol = 1, heights = c(1, plot.height)
    )
}
#rrbs_plot5(dmr_reads, reads) +
#  geom_beta(all_b) +
#  geom_dau(dau_tiles)
```

*dmr
```{r}
geom_dmr <- function(
    dmr.df=NULL, dmr.label=FALSE, dmr.color="black", dmr.size=5,
    plot.space=0.1, plot.height=0.05) {
    structure(
        list(
            dmr.df = dmr.df, dmr.label = dmr.label, dmr.color = dmr.color, dmr.size = dmr.size,
            plot.space = plot.space, plot.height = plot.height
        ),
        class = "dmr"
    )
}


#' @export
ggplot_add.dmr <- function(object, plot, object_name) {
    if (!is.data.frame(object$dmr.df)) {
        stop("data.frame of dmr is required")
    }
    plot2 <- plot
    while (inherits(plot2, "patchwork")) {
        plot2 <- plot2[[1]]
    }
    plot.data <- plot2$labels$title
    plot.data <- stringr::str_split_1(plot.data, ":")

    plot.chr <- plot.data[1]
    plot.data <- stringr::str_split_1(plot.data[2], "-")
    plot.region.start <- plot.data[1] %>% as.numeric()
    plot.region.end <- plot.data[2] %>% as.numeric()

    # get parameters
    dmr.df <- object$dmr.df
    dmr.label <- object$dmr.label
    dmr.color <- object$dmr.color
    dmr.size <- object$dmr.size
    plot.space <- object$plot.space
    plot.height <- object$plot.height
    
    dmr.df <- dmr.df %>% filter(seqnames %in% plot.chr,
                                start >= plot.region.start,
                                end <= plot.region.end)
    
   # dmr.df <-  dmr.df %>% mutate(score = c(1,1,1)) 
    dmr.plot <- dmr.df %>%
            ggplot2::ggplot(ggplot2::aes(x = .data$start, y = 1)) +
            ggplot2::geom_segment(
                mapping = ggplot2::aes(
                    x = .data$start, y = 1,
                    xend = .data$end, yend = 1
                ),
                linewidth = dmr.size, color = dmr.color
            ) +
      labs(y = "dmr") +
      ..peakGatcTheme(margin.len = plot.space, x.range = c(plot.region.start, plot.region.end))

    # assemble plot
    patchwork::wrap_plots(plot + ggplot2::theme(plot.margin = ggplot2::margin(t = plot.space, b = plot.space)),
        dmr.plot,
        ncol = 1, heights = c(1, plot.height)
    )
}
rrbs_plot5(dmr_reads, reads) +
  geom_beta(all_b) +
  geom_dmr(dmr_emseq)
```


```{r}
plot_test <- rrbs_plot5(dmr_reads, reads) +
  geom_beta(all_b) +
  geom_dau(dau_tiles) +
  geom_dmr(dmr_emseq)
```

```{r}
plot_test2 <- rrbs_plot5(dmr_reads, reads) +
  geom_beta(all_b) +
  geom_dau(dau_tiles) +
  geom_dmr(dmr_emseq)
```

```{r}
plot_test3 <- rrbs_plot5(distinct(dmr_reads), reads) +
  geom_beta(all_b) +
  geom_dau(dau_tiles) +
  geom_dmr(dmr_emseq)
```


```{r}
dmr_reads
```
```{r}
distinct(dmr_reads)
```
```{r}
dau_tiles %>% filter(region_id %in% "chrchr1:147077877-147078376")
```

```{r}
dmr_reads <- distinct(dmr_reads)
```

```{r}
dmr_reads
```


```{r}
tile_dmr_read <- find_overlaps(as_granges(dau_tiles), as_granges(dmr_reads)) %>% data.frame()
tile_dmr_read %>% distinct(region_id.x)
tile_dmr_read %>% distinct(region_id.y)
```
* 3 tiles

* I think there's a missing bit because there's no tile there
* add a tile location layer

* and it's going to be because I only used the healthy samples to make the tiles - and that data is sparse in that spot



```{r}
optiles_all <- readRDS("/researchers/caitlin.page/cf_nano/r_output/optiles_all_emseq.rds")
```


```{r}
geom_tilepos <- function(
    tilepos.df=NULL, tilepos.label=FALSE, tilepos.color="black", tilepos.size=5,
    plot.space=0.1, plot.height=0.05) {
    structure(
        list(
            tilepos.df = tilepos.df, tilepos.label = tilepos.label, tilepos.color = tilepos.color, tilepos.size = tilepos.size,
            plot.space = plot.space, plot.height = plot.height
        ),
        class = "tilepos"
    )
}


#' @export
ggplot_add.tilepos <- function(object, plot, object_name) {
    if (!is.data.frame(object$tilepos.df)) {
        stop("data.frame of tilepos is required")
    }
    plot2 <- plot
    while (inherits(plot2, "patchwork")) {
        plot2 <- plot2[[1]]
    }
    plot.data <- plot2$labels$title
    plot.data <- stringr::str_split_1(plot.data, ":")

    plot.chr <- plot.data[1]
    plot.data <- stringr::str_split_1(plot.data[2], "-")
    plot.region.start <- plot.data[1] %>% as.numeric()
    plot.region.end <- plot.data[2] %>% as.numeric()

    # get parameters
    tilepos.df <- object$tilepos.df
    tilepos.label <- object$tilepos.label
    tilepos.color <- object$tilepos.color
    tilepos.size <- object$tilepos.size
    plot.space <- object$plot.space
    plot.height <- object$plot.height
    
    tilepos.df <- tilepos.df %>% filter(seqnames %in% plot.chr,
                                start >= plot.region.start,
                                end <= plot.region.end)
    
    tilepos.plot <- tilepos.df %>%
            ggplot2::ggplot(ggplot2::aes(x = .data$start, y = 1)) +
            ggplot2::geom_segment(
                mapping = ggplot2::aes(
                    x = .data$start, y = 1,
                    xend = .data$end, yend = 1
                ),
                linewidth = tilepos.size, color = tilepos.color
            ) +
      labs(y = "tilepos") +
      ..peakGatcTheme(margin.len = plot.space, x.range = c(plot.region.start, plot.region.end))

    # assemble plot
    patchwork::wrap_plots(plot + ggplot2::theme(plot.margin = ggplot2::margin(t = plot.space, b = plot.space)),
        tilepos.plot,
        ncol = 1, heights = c(1, plot.height)
    )
}
```

```{r}
plot_test4 <- rrbs_plot5(distinct(dmr_reads), reads) +
  geom_beta(all_b) +
  geom_dau(dau_tiles) +
  geom_dmr(dmr_emseq) +
  geom_tilepos(optiles_all)
```

* ok OP

* duh modifiy the tile
```{r}
tile_dmr_read %>% distinct(region_id.x)
```



```{r}
a <- find_overlaps(as_granges(filter(optiles_all, region_id %in% c("chrchr1:147077877-147078376", "chrchr1:147078743-147079242", "chrchr1:147079643-147080142"))), as_granges(filter(mutate(all_b, seqnames = seqnames.x, start = start.x, width = 1), !is.na)) %>% data.frame()
```

```{r}
optiles_all %>% filter(seqnames == "chr1", start >= 147077877) %>% .[order(.$start),] %>% .[1:5,] %>% mutate(gap_to_next = lead(start) - end)
```

```{r}
all_b %>% filter(seqnames.x == "chr1", start.x >= 147078376, start.x <= 147078743)
```

```{r}
all_b %>% filter(seqnames.x == "chr1", start.x >= 147079242, start.x <= 147079643)
```


```{r}
dmr_reads %>% mutate(read_in_tile)
```


..plotPeak <- function(valid.bed, plot.size, plot.color, peak.label) {
    if (nrow(valid.bed) == 0) {
        message("No data available for this region")
        peak.plot <- ggplot2::ggplot() +
            ggplot2::geom_blank()
    } else {
        peak.plot <- valid.bed %>%
            ggplot2::ggplot(ggplot2::aes(x = .data$start, y = 1)) +
            ggplot2::geom_segment(
                mapping = ggplot2::aes(
                    x = .data$start, y = 1,
                    xend = .data$end, yend = 1
                ),
                linewidth = plot.space, color = plot.color
            ) +
            if (peak.label == TRUE) {
                ggplot2::geom_label(ggplot2::aes(
                    x = (.data$start + .data$end) / 2,
                    label = .data$peak_id
                ), colour = "black", size = 3, vjust = "bottom", nudge_y = 0.02)
            }
    }
    peak.plot
}
```{r}
..peakGatcTheme <- function(margin.len, x.range) {
    list(
        ggplot2::theme_classic(),
        ggplot2::theme(
            axis.line.y = ggplot2::element_blank(),
            axis.text.y = ggplot2::element_blank(),
            axis.title.y.right = ggplot2::element_text(color = "black", angle = 90, vjust = 0.5),
            axis.ticks.y = ggplot2::element_blank(),
            axis.text.x = ggplot2::element_blank(),
            axis.title.x = ggplot2::element_blank(),
            axis.ticks.x = ggplot2::element_blank(),
            panel.border = ggplot2::element_rect(colour = "black", fill = NA, linewidth = 1),
            plot.margin = ggplot2::margin(t = margin.len, b = margin.len)
        ),
        ggplot2::scale_y_continuous(
            limits = c(1 - 0.1, 1 + 0.1),
            expand = c(0, 0), position = "right"
        ),
        ggplot2::scale_x_continuous(expand = c(0, 0)),
        ggplot2::coord_cartesian(xlim = x.range)
    )
}
```


```{r}
dmr_emseq
```


geom_dm <- function(dm_results.df, plot.space=0.1, plot.height=0.1) {
    structure(list(dm_results.df = dm_results.df, plot.space = plot.space,
        plot.height = plot.height),
        class = "dm"
    )
}

#' @export
ggplot_add.dm <- function(object, plot, object_name) {
    if (!is.data.frame(object$dm_results.df)) {
        stop("data.frame of dm results is required")
    }
    plot2 <- plot
    while (inherits(plot2, "patchwork")) {
        plot2 <- plot2[[1]]
    }
    plot.data <- plot2$labels$title
    plot.data <- stringr::str_split_1(plot.data, ":")
    # prepare plot range
    # the plot region are not normal, so start is minimum value
    plot.chr <- plot.data[1]
    plot.data <- stringr::str_split_1(plot.data[2], "-")
    plot.region.start <- plot.data[1] %>% as.numeric()
    plot.region.end <- plot.data[2] %>% as.numeric()

    # get parameters
    dm_results.df <- object$dm_results.df
    plot.space <- object$plot.space
    plot.height <- object$plot.height

    df_regions <- dm_results.df %>% dplyr::filter(
        .data$seqnames == plot.chr,
        .data$start >= plot.region.start,
        .data$end <= plot.region.end
    )
    df_regions <- df_regions %>%
        dplyr::mutate(
            dm = ifelse(.data$dm == -1, 0, .data$dm),
            meth_status = ifelse(.data$meth_status == "Downreg", "No_sig", .data$meth_status)
        )
    df_colour <- ..dmReshape(df_regions)

    df_fc <- ..dmMax(df_regions)

    colours_list <- c("Upreg" = "red", "Not_included" = "grey", "No_sig" = "black")

    dm_res.plot <- df_regions %>%
        ggplot2::ggplot(ggplot2::aes(x = .data$start, y = .data$logFC, colour = factor(.data$meth_status, levels = c("Upreg", "No_sig", "Not_included")))) +
        ggplot2::geom_polygon(data = df_colour, ggplot2::aes(x = .data$Position, y = .data$y_axis_2, fill = factor(.data$meth_status, levels = c("Upreg", "No_sig", "Not_included")))) +
        ggplot2::geom_segment(ggplot2::aes(xend = .data$start, yend = 0)) +
        ggplot2::geom_segment(ggplot2::aes(x = .data$end, xend = .data$end, y = .data$logFC, yend = 0)) +
        ggplot2::geom_segment(ggplot2::aes(x = .data$start, xend = .data$end, y = .data$logFC, yend = .data$logFC)) +
        ggplot2::geom_segment(ggplot2::aes(x = .data$start, xend = .data$end, y = 0, yend = 0)) +
        ..dmTheme(margin.len = plot.space, colours = colours_list, x.range = c(plot.region.start, plot.region.end)) +
        ggplot2::scale_y_continuous(
            limits = c(-(df_fc$abs_fc) - 0.5, df_fc$abs_fc + 0.5),
            expand = c(0, 0), breaks = c(-round(df_fc$abs_fc), 0, round(df_fc$abs_fc)), position = "right"
            )

    # assemble plot
    patchwork::wrap_plots(plot + ggplot2::theme(plot.margin = ggplot2::margin(t = plot.space, b = plot.space)),
        dm_res.plot,
        ncol = 1, heights = c(1, plot.height)
    )
}

```{r}
dmr_reads
```

reads <- find_overlaps(dtu_tiles_healthy_compare %>% filter(FDR < 0.05) %>% .[3,] %>% as_granges(), as_granges(healthy_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(healthy_samples, reads)

