---
title: "emseqBiologyAnalysis"
author: "caitlinpage"
date: "2026-07-28"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

## Introduction

```{r}
library(BSgenome.Hsapiens.UCSC.hg38)
library(data.table)
library(plyranges)
library(tidyr)
library(stringr)
library(dplyr)
library(ggplot2)
library(OpTiles)
library(edgeR)
library(UpSetR)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
```

```{r}
genes_txdb <- genes(TxDb.Hsapiens.UCSC.hg38.knownGene)
```
```{r}
optiles_all <- readRDS("/researchers/caitlin.page/cf_nano/r_output/optiles_all_emseq.rds")
dtu_emseq <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_emseq_cancer_healthy.rds")
dmr_emseq <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dmr_emseq_strict.rds")
dmr_emseq <- dmr_emseq %>% .[order(.$min_smoothed_fdr),] %>% mutate(dmr_rank = 1:n())
```

```{r}
#tile_counts_2
#tile_counts
```

# cpgs and tiles
```{r}
seqs <- seqnames(BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38)[1:22]

cg_sites <- lapply(seqs, function(x) {
  cbind(data.frame(Biostrings::matchPattern("CG", BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38[[x]])),
        seqnames = x
  )}) %>% bind_rows()

optiles_all <- find_overlaps(as_granges(optiles_all), as_granges(cg_sites)) %>% data.frame()
optiles_all <- optiles_all %>% group_by(seqnames, start, end, region_id) %>%
  summarise(n_cg = n()) %>% ungroup()
```

# Understanding genes and tiles

## how many tiles overlap genes?

```{r}
gene_tile <- find_overlaps(genes_txdb, as_granges(optiles_all)) %>% data.frame()
```

```{r}
gene_tile %>%
  group_by(gene_id) %>% summarise(n_tile = n()) %>%
  ggplot(aes(x = n_tile)) +
  geom_histogram(bins = 50) +
  theme_bw()
```

```{r}
gene_tile %>%
  group_by(gene_id) %>% summarise(n_tile = n()) %>%
  ggplot(aes(x = n_tile)) +
  geom_histogram(bins = 50) +
  scale_x_continuous(limits = c(0,500)) +
  theme_bw()

gene_tile %>%
  group_by(gene_id) %>% summarise(n_tile = n()) %>%
  ggplot(aes(x = n_tile)) +
  geom_histogram(bins = 50) +
  scale_x_continuous(limits = c(0,200)) +
  theme_bw()
```


## significant tiles

```{r}
dtu_gene <- find_overlaps(genes_txdb, as_granges(dtu_emseq)) %>% data.frame()
```

```{r}
dtu_gene %>% filter(FDR < 0.05) %>%
  group_by(gene_id) %>% summarise(n=dplyr::n()) %>% ungroup() %>% group_by(n) %>% summarise(n_regions_gene_overlaps = dplyr::n())
```

```{r}
find_overlaps(genes_txdb, as_granges(optiles_all)) %>% data.frame() %>%
  group_by(gene_id) %>% summarise(n_tile = n()) %>%
  ggplot(aes(x = n_tile)) +
  geom_histogram(bins = 50)
```


```{r}
dtu_gene %>% filter(FDR < 0.05) %>%
  group_by(gene_id) %>% summarise(n_sig_regions=dplyr::n()) %>% ungroup() %>%
  ggplot(aes(x = as.factor(n_sig_regions))) +
  geom_bar()
```

## are genes with more tiles more likely to show up as significant?

```{r}
te <- gene_tile %>% group_by(gene_id) %>% summarise(n_tile = n()) %>% mutate(tile_signif = ifelse(gene_id %in% filter(dtu_gene, FDR < 0.05)$gene_id, TRUE, FALSE)) %>%
  ungroup() %>%
  group_by(n_tile, tile_signif) %>%
  summarise(n=n()) %>% ungroup() %>% group_by(n_tile) %>% mutate(prop = n/sum(n)) %>% ungroup()

te_n <- te %>% filter(tile_signif) %>% pull(n_tile)
```

```{r}
te %>% filter(tile_signif == TRUE) %>%
  bind_rows(
    te %>%
      filter(!n_tile %in% te_n) %>%
      distinct(n_tile) %>%
      mutate(tile_signif = TRUE, n = 0L, prop = 0)
  ) %>%
  arrange(n_tile) %>%
  ggplot(aes(x = n_tile, y = prop)) +
  geom_point()
```
* I was expecting this to look like my gene set testing bias plot
* but no it has to be special and difficult
* could be because not binning it 

```{r}
source("../code/runGST.R")
bias_plot <- gene_tile %>% group_by(gene_id) %>% summarise(n_tile = n()) %>% mutate(tile_signif = ifelse(gene_id %in% filter(dtu_gene, FDR < 0.05)$gene_id, 1, 0))
.plotBias(bias_plot$tile_signif, bias_plot$n_tile)
```
* that's better - but yeah it is very low ( also this was number of tiles per gene - not cpgs)

* maybe tiles and cpgs aren't that related to each other?

```{r}
gene_tile %>% group_by(gene_id) %>%
  summarise(n_tile = n(), sum_cg = sum(n_cg)) %>%
  ggplot(aes(x = n_tile, y = sum_cg)) +
  geom_point(alpha = 0.3)
```
* not the case
mmmmmmmmmmmmmmmmm
mmm
* make the bias plot with the cpgs
```{r}
source("../code/runGST.R")
bias_plot <- gene_tile %>% group_by(gene_id) %>% summarise(sum_cg = sum(n_cg)) %>% mutate(tile_signif = ifelse(gene_id %in% filter(dtu_gene, FDR < 0.05)$gene_id, 1, 0))
.plotBias(bias_plot$tile_signif, bias_plot$sum_cg)
```



# 1 gene, multiple tiles
```{r}
dtu_gene %>% filter(FDR < 0.05) %>%
  group_by(gene_id) %>% mutate(n_sig_regions=dplyr::n()) %>% ungroup() %>% filter(n_sig_regions > 3) %>% .[order(.$n_sig_regions, decreasing = TRUE),] %>% .[order(.$dtu_rank),] %>% dplyr::distinct(gene_id, dtu_rank) %>% nrow()

dtu_gene %>% filter(FDR < 0.05) %>%
  group_by(gene_id) %>% summarise(n_sig_regions=dplyr::n()) %>% ungroup() %>% filter(n_sig_regions > 3) %>% .[order(.$n_sig_regions, decreasing = TRUE),] %>% dplyr::distinct(gene_id) %>% nrow()
```

```{r}
dtu_emseq %>% filter(FDR < 0.05) %>% nrow()
```

```{r}
dtu_gene %>% filter(FDR < 0.05) %>% 
    group_by(gene_id) %>% mutate(n_sig_regions=dplyr::n()) %>% ungroup() %>%
    filter(n_sig_regions > 3) %>% .[order(.$n_sig_regions, decreasing = TRUE),] %>% .[order(.$dtu_rank),] %>% dplyr::distinct(gene_id, dtu_rank) %>% .[1,]
```


```{r}
list_plot <- list()
for(i in 1:18) {
  
  tile_sig <- dtu_gene %>% filter(FDR < 0.05) %>% 
    group_by(gene_id) %>% mutate(n_sig_regions=dplyr::n()) %>% ungroup() %>%
    filter(n_sig_regions > 3) %>% .[order(.$n_sig_regions, decreasing = TRUE),] %>% .[order(.$dtu_rank),] %>% dplyr::distinct(gene_id, dtu_rank)
  
  p_data <- dtu_gene %>% filter(gene_id == tile_sig$gene_id[i]) %>% 
  tidyr::separate(region_id, into = c("chr", "tile_start", "tile_end"), sep = "[:-]") %>% 
  mutate(tile_start = as.integer(tile_start), tile_end = as.integer(tile_end),
         y_pos = dplyr::row_number())
p <- p_data %>%
  ggplot() +
  geom_segment(data = filter(p_data, strand == "-"), aes(x = start, xend = end, y = 0, yend = 0), linewidth = 2,
               arrow = arrow(length = unit(0.2, "cm"), type = "closed", ends = "first")) +
  geom_segment(data = filter(p_data, strand == "+"), aes(x = start, xend = end, y = 0, yend = 0), linewidth = 2,
               arrow = arrow(length = unit(0.2, "cm"), type = "closed", ends = "last")) +
  geom_segment(aes(x = tile_start, xend = tile_end, y = y_pos, yend = y_pos), linewidth = 2) +
  geom_text(aes(x = min(start), y = 0.2, label = gene_id),
          hjust = 1,    # right-align so all labels end at the same x
          nudge_x = -10000) +
coord_cartesian(clip = "off") +   # allows text to render outside the plot area
theme(plot.margin = margin(5, 5, 5, 60)) +
  geom_text(aes(x = tile_start, y = y_pos, label = dtu_rank)) +
  ggtitle(gsub("^chr", "", unique(p_data$chr)))

list_plot <- list(list_plot, p)
}
list_plot
```



```{r}
dtu_gene %>% filter(FDR < 0.05) %>% dplyr::distinct(gene_id) %>% nrow()
```


```{r}
library(org.Hs.eg.db)
gene_types <- select(org.Hs.eg.db, keys = dtu_gene %>% filter(FDR < 0.05) %>% dplyr::distinct(gene_id) %>% .$gene_id, 
                     columns = c("SYMBOL", "GENENAME"), 
                     keytype = "ENTREZID")
```


# make some fake promoters

* 2000bp upstream, 200bp downstream
```{r}
promoters_txdb <- promoters(genes_txdb)
```


```{r}
dtu_promoter <- find_overlaps(promoters_txdb, as_granges(dtu_emseq)) %>% data.frame()
```

```{r}
dtu_promoter %>% filter(FDR < 0.05) %>% nrow()
dtu_promoter %>% filter(FDR < 0.05) %>% dplyr::distinct(gene_id) %>% nrow()
```

```{r}
dtu_promoter %>% filter(FDR < 0.05) %>%
  group_by(gene_id) %>% summarise(n_sig_regions=dplyr::n()) %>% ungroup() %>%
  ggplot(aes(x = as.factor(n_sig_regions))) +
  geom_bar()
```

```{r}
gene_types_prom <- select(org.Hs.eg.db, keys = dtu_promoter %>% filter(FDR < 0.05) %>% dplyr::distinct(gene_id) %>% .$gene_id, 
                     columns = c("SYMBOL", "GENENAME"), 
                     keytype = "ENTREZID")
```


```{r}
ggVennDiagram::ggVennDiagram(list(gene = dtu_gene %>% filter(FDR < 0.05) %>% dplyr::distinct(gene_id) %>% .$gene_id, promoter = dtu_promoter %>% filter(FDR < 0.05) %>% dplyr::distinct(gene_id) %>% .$gene_id))
```

# where do overlaps happen in gene

```{r}
dtu_gene %>% filter(FDR < 0.05) %>% .[order(.$FDR),] %>% .[1,]
dtu_gene %>% filter(FDR < 0.05) %>% .[order(.$FDR),] %>% .[1,] %>% 
  tidyr::separate(region_id, into = c("chr", "tile_start", "tile_end"), sep = "[:-]") %>% 
  mutate(tile_start = as.integer(tile_start), tile_end = as.integer(tile_end)) %>% {
    strand_dir <- if (unique(.$strand) == "-") "first" else "last"
  ggplot(.) +
  geom_segment(aes(x = start, xend = end, y = 1, yend = 1), linewidth = 2,
               arrow = arrow(length = unit(0.2, "cm"), type = "closed", ends = strand_dir)) +
  geom_segment(aes(x = tile_start, xend = tile_end, y = 2, yend = 2), linewidth = 2) +
  geom_text(aes(x = start, y = 1.2, label = gene_id))
  }
```

```{r}
list_plot <- list()
for(i in 1:10) {
  
p <- dtu_gene %>% filter(FDR < 0.05) %>% .[order(.$FDR),] %>% .[i,] %>% 
  tidyr::separate(region_id, into = c("chr", "tile_start", "tile_end"), sep = "[:-]") %>% 
  mutate(tile_start = as.integer(tile_start), tile_end = as.integer(tile_end)) %>% {
    strand_dir <- if (unique(.$strand) == "-") "first" else "last"
    chr_label  <- unique(.$chr)
  ggplot(.) +
  geom_segment(aes(x = start, xend = end, y = 1, yend = 1), linewidth = 2,
               arrow = arrow(length = unit(0.2, "cm"), type = "closed", ends = strand_dir)) +
  geom_segment(aes(x = tile_start, xend = tile_end, y = 2, yend = 2), linewidth = 2) +
  geom_text(aes(x = start, y = 1.2, label = gene_id)) +
  geom_text(aes(x = tile_start, y = 1.8, label = dtu_rank)) +
  ggtitle(chr_label)
}
list_plot <- list(list_plot, p)
  }
list_plot
```


# promoter

```{r}
list_plot <- list()
for(i in 1:10) {
  
p <- dtu_promoter %>% filter(FDR < 0.05) %>% .[order(.$FDR),] %>% .[i,] %>% 
  tidyr::separate(region_id, into = c("chr", "tile_start", "tile_end"), sep = "[:-]") %>% 
  mutate(tile_start = as.integer(tile_start), tile_end = as.integer(tile_end)) %>% {
    strand_dir <- if (unique(.$strand) == "-") "first" else "last"
    chr_label <- unique(.$chr)
  ggplot(.) +
  geom_segment(aes(x = start, xend = end, y = 1, yend = 1), linewidth = 2,
               arrow = arrow(length = unit(0.2, "cm"), type = "closed", ends = strand_dir)) +
  geom_segment(aes(x = tile_start, xend = tile_end, y = 2, yend = 2), linewidth = 2) +
  geom_text(aes(x = start, y = 1.2, label = gene_id)) +
  geom_text(aes(x = tile_start, y = 1.8, label = dtu_rank)) +
  ggtitle(chr_label)
  }

list_plot <- list(list_plot, p)
}
list_plot
```


```{r}
dtu_gene %>% filter(FDR < 0.05) %>%
  group_by(region_id) %>% summarise(n_gene_overlap = dplyr::n()) %>%
  ggplot(aes(x = as.factor(n_gene_overlap))) +
  geom_bar()

dtu_promoter %>% filter(FDR < 0.05) %>%
  group_by(region_id) %>% summarise(n_gene_overlap = dplyr::n()) %>%
  ggplot(aes(x = as.factor(n_gene_overlap))) +
  geom_bar()
```
```{r}
dtu_gene %>% filter(FDR < 0.05) %>% 
  group_by(region_id) %>% mutate(n_gene_overlap = dplyr::n()) %>% ungroup() %>%
  filter(n_gene_overlap > 3) %>% .[order(.$n_gene_overlap, decreasing = TRUE),] %>% .[order(.$dtu_rank),] %>% .[1:5,]
```

```{r}
dtu_gene %>% filter(FDR < 0.05) %>% 
  group_by(region_id) %>% mutate(n_gene_overlap = dplyr::n()) %>% ungroup() %>%
  filter(n_gene_overlap > 3) %>% .[order(.$n_gene_overlap, decreasing = TRUE),] %>% .[order(.$dtu_rank),] %>% dplyr::distinct(region_id, dtu_rank)
```


# 1 dtu, multiple genes

```{r}
list_plot <- list()
for(i in 1:18) {
  
  tile_sig <- dtu_gene %>% filter(FDR < 0.05) %>% 
    group_by(region_id) %>% mutate(n_gene_overlap = dplyr::n()) %>% ungroup() %>%
    filter(n_gene_overlap > 3) %>% .[order(.$n_gene_overlap, decreasing = TRUE),] %>% .[order(.$dtu_rank),] %>% dplyr::distinct(region_id, dtu_rank)
  
  p_data <- dtu_gene %>% filter(region_id == tile_sig$region_id[i]) %>% 
  tidyr::separate(region_id, into = c("chr", "tile_start", "tile_end"), sep = "[:-]") %>% 
  mutate(tile_start = as.integer(tile_start), tile_end = as.integer(tile_end),
         y_pos = disjointBins(IRanges(start, end)))
p <- p_data %>%
  ggplot() +
  geom_segment(data = filter(p_data, strand == "-"), aes(x = start, xend = end, y = y_pos, yend = y_pos), linewidth = 2,
               arrow = arrow(length = unit(0.2, "cm"), type = "closed", ends = "first")) +
  geom_segment(data = filter(p_data, strand == "+"), aes(x = start, xend = end, y = y_pos, yend = y_pos), linewidth = 2,
               arrow = arrow(length = unit(0.2, "cm"), type = "closed", ends = "last")) +
  geom_segment(aes(x = tile_start, xend = tile_end, y = 27, yend = 27), linewidth = 2) +
  geom_text(aes(x = min(start), y = y_pos, label = gene_id),
          hjust = 1,    # right-align so all labels end at the same x
          nudge_x = -10000) +
coord_cartesian(clip = "off") +   # allows text to render outside the plot area
theme(plot.margin = margin(5, 5, 5, 60)) +
  geom_text(aes(x = tile_start, y = 26, label = dtu_rank)) +
  ggtitle(gsub("^chr", "", unique(p_data$chr)))

list_plot <- list(list_plot, p)
}
list_plot
```



# where do overlaps happen in promoter?

# PPGL genes (calude)

```{r}
ppgl_drivers <- data.frame(
  symbol = c(
    # Cluster 1 — Pseudohypoxia / TCA cycle
    "SDHA", "SDHB", "SDHC", "SDHD", "SDHAF2",
    "VHL", "FH", "MDH2", "EPAS1", "IDH1", "IDH2",
    "SLC25A11", "DLST", "GOT2",
    # Cluster 2 — Kinase / RAS signalling
    "RET", "NF1", "HRAS", "KRAS", "NRAS",
    "TMEM127", "MAX", "KIF1B", "FGFR1",
    # Cluster 3 — Wnt / cortical
    "MAML3", "CSDE1", "UBTF",
    # Malignancy / progression
    "ATRX", "TERT", "TP53", "CDKN2A",
    "RB1", "SETD2", "PTEN", "MEN1"
  ),
  entrez_id = c(
    6389, 6390, 6391, 6392, 54949,
    7428, 2271, 4191, 2034, 3417, 3418,
    8402, 1743, 2806,
    5979, 4763, 3265, 3845, 4893,
    55654, 4149, 23095, 2260,
    55502, 7812, 7343,
    546, 7015, 7157, 1029,
    5925, 29072, 5728, 4221
  ),
  cluster = c(
    rep("Pseudohypoxia", 14),
    rep("Kinase_RAS", 9),
    rep("Wnt_cortical", 3),
    rep("Malignancy", 8)
  ),
  stringsAsFactors = FALSE
)
```

```{r}
ggVennDiagram::ggVennDiagram(list(ppgl = as.character(ppgl_drivers$entrez_id), dtu = dtu_gene %>% filter(FDR < 0.05) %>% dplyr::distinct(gene_id) %>% .$gene_id))
```

```{r}
ppgl_drivers %>% filter(symbol %in% gene_types$SYMBOL)
ppgl_drivers %>% filter(symbol %in% gene_types_prom$SYMBOL)
```

