---
title: "emseqFalsePositiveCheck"
author: "caitlinpage"
date: "2026-07-28"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

## Introduction

The EMSeq data has 7 healthy samples. This means that we can cross the healthy samples, to check our false positive rate.
Basically, the fewer significant regions we get out, the better.

As there are 7 samples, I'll rotate the groups to run multiple times.

## Packages

```{r}
library(BSgenome.Hsapiens.UCSC.hg38)
library(data.table)
library(plyranges)
library(tidyr)
library(stringr)
library(dplyr)
library(ggplot2)
library(OpTiles)
library(edgeR)
```

```{r}
seqs <- seqnames(BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38)[1:22]

cg_sites <- lapply(seqs, function(x) {
  cbind(data.frame(Biostrings::matchPattern("CG", BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38[[x]])),
        seqnames = x
  )}) %>% bind_rows()
```


```{r}
process_wgbs_short <- function(bismark, bed = NULL) {
  read <- bismark
  names(read) <- c("read_id", "strand", "seqnames", "start", "meth_status")
  read$end <- read$start

  read_overlap <- read
  read_overlap[, meth_status := ifelse(meth_status == "Z", "M", "U")]
  read_overlap[, meth_pattern := paste0(meth_status, collapse = ""), by = read_id]
  read_overlap[, num_meth := str_count(meth_pattern, pattern = "M")]
  read_overlap[, total := str_length(meth_pattern)]
  read_overlap[, alpha := num_meth/total]
  read_overlap[, genom_positions := paste0(start, collapse = ","), by = read_id]

  read_overlap <- read_overlap %>% .[order(.$read_id),]

  read_overlap[, read_start := min(start), by = read_id]
  read_overlap[, read_end := max(start), by = read_id]

  read_overlap
}

bis <- fread("/researchers/caitlin.page/cf_nano/fastq_tothill_emseq_healthy/CpG_OT_PRJ240788_LPRJ240788_S4_L004_R1_001_bismark_bt2_pe.txt")
em_healthy_788 <- process_wgbs_short(bis)
saveRDS(em_healthy_788, "/researchers/caitlin.page/cf_nano/r_output/emseq_healthy_788.rds")
```

CpG_OT_PRJ240788_LPRJ240788_S4_L004_R1_001_bismark_bt2_pe.txt
```{r}
all_samples <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_all_healthy.rds")
```

```{r}
em_healthy_788 <- em_healthy_788 %>% data.frame() %>% filter(seqnames %in% seqs)
em_healthy_788$start <- em_healthy_788$read_start
em_healthy_788$end <- em_healthy_788$read_end
em_healthy_788$sample <- "healthy_788"
```


```{r}
cols <- c("seqnames", "read_start", "read_end", "strand", "read_id", "sample", "meth_pattern", "num_meth", "total", "alpha", "genom_positions")
```


```{r}
all_samples <- all_samples %>%
  mutate(sample = case_when(sample == "a1" ~ "healthy_785",
                            sample == "a2" ~ "healthy_036",
                            sample == "a3" ~ "healthy_787",
                            sample == "b1" ~ "healthy_786",
                            sample == "b2" ~ "healthy_037",
                            TRUE ~ "healthy_789"))

all_samples <- rbind(all_samples, distinct(em_healthy_788[,colnames(all_samples)]))
```

```{r}
dtu_tiles_healthy_compare <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_emseq_healthy.rds")
dtu_tiles_healthy_compare %>% filter(FDR < 0.05) %>% group_by(seqnames) %>% summarise(n=n())
```


```{r}
optiles_all <- readRDS("/researchers/caitlin.page/cf_nano/r_output/optiles_all_emseq.rds")
nrow(optiles_all)
optiles_all %>% group_by(seqnames) %>% summarise(n=n()) %>% .[1:5,]
```

* we'll use chr1

# dtu testing

```{r}
overlap_pos <- find_overlaps(as_granges(filter(optiles_all, seqnames == "chr1")), as_granges(all_samples)) %>% data.frame()
```

```{r}
# dtu
dtu_optile <- overlap_pos %>% filter(total >= 3) %>%
  distinct(sample, read_id, alpha, region_id) %>%
  mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                 alpha > 0.8 ~ "a > 0.8",
                                 alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                 alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                 TRUE ~ "0.6 <= a <= 0.8"))

dtu_optile <- dtu_optile %>%
  mutate(transcript = paste0(region_id, "_", alpha_group)) %>%
  group_by(sample, transcript) %>%
  summarise(transcript_count = n()) %>% ungroup() %>% data.frame()

# build matrix of "transcript" counts - transcript rows, sample as columns
dtu_optile <- dtu_optile %>% pivot_wider(names_from = sample, values_from = transcript_count)
dtu_optile[is.na(dtu_optile)] <- 0
dtu_optile$gene_id <- sub("_.*", "", dtu_optile$transcript)
```

```{r}
unique(all_samples$sample)
```


```{r}
samples <- c(unique(all_samples$sample))  # your 7 sample names

# --- Analysis function ---
run_analysis <- function(combo) {
  cols <- c(combo$g1, combo$g2)
  group <- c(1, 1, 1, 2, 2, 2)

  y <- DGEList(
    as.matrix(dtu_optile[, cols]),
    group = group,
    genes = dtu_optile[, c("gene_id", "transcript")]
  )
  design <- model.matrix(~ group)
  fit <- glmQLFit(y, design, robust = TRUE)
  ds  <- diffSpliceDGE(fit, coef = 2, geneid = "gene_id", exonid = "transcript")

  topSpliceDGE(ds, test = "gene", number = Inf) %>%
    mutate(
      region_id = gene_id,
      excluded  = combo$excluded,
      group1    = paste(combo$g1, collapse = ","),
      group2    = paste(combo$g2, collapse = ",")
    ) %>%
    separate_wider_delim(gene_id, delim = ":", names = c("seqnames", "gene_id")) %>%
    mutate(seqnames = gsub("chrchr", "chr", seqnames)) %>%
    separate_wider_delim(gene_id, delim = "-", names = c("start", "end")) %>%
    mutate(
      start    = as.double(start),
      end      = as.double(end),
      dtu_rank = row_number()
    )
}

# --- Generate 70 combinations ---
all_combos <- unlist(
  lapply(seq_along(samples), function(excl_i) {
    remaining <- samples[-excl_i]
    splits <- combn(remaining, 3, simplify = FALSE)
    splits <- Filter(function(g1) g1[1] < setdiff(remaining, g1)[1], splits)
    lapply(splits, function(g1) {
      list(g1 = g1, g2 = setdiff(remaining, g1), excluded = samples[excl_i])
    })
  }),
  recursive = FALSE
)

# --- Run across all 70 combinations ---
results <- lapply(all_combos, run_analysis)

# --- Combine into one data frame ---
results_df <- bind_rows(results)

saveRDS(results_df, "/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_emseq_healthy_all_fp.rds")
```

# Analysis of tile level results

```{r}
results_df[1:10,]
```
```{r}
results_df %>% filter(FDR < 0.05) %>% nrow()
```

```{r}
results_df %>% distinct(excluded, group1, group2) %>% nrow() # checking correct number of runs
```

```{r}
results_df %>% mutate(fdr_signif = ifelse(FDR < 0.05, TRUE, FALSE)) %>% group_by(excluded, group1, group2, fdr_signif) %>%
  summarise(n = n()) %>%
  ungroup() %>%
  group_by(excluded, group1, group2) %>%
  mutate(total_group = sum(n), prop = n/sum(n)) %>%
  ungroup() %>%
  filter(fdr_signif == TRUE) %>% .[1:5,]
```

* these are some tiny, beautifully low proportions
* crazy how low it is

* also this is only 42 of the 70 groups - so some groups didn't have any significant - no false positives

* this is expected to be related to coverage

```{r}
dtu_optile %>% mutate(reads = rowSums(.[,2:8])) %>%
  group_by(gene_id) %>%
  summarise(coverage = sum(reads)) %>%
  mutate(region_signif = ifelse(gene_id %in% filter(results_df, FDR < 0.05)$region_id, TRUE, FALSE)) %>%
  ggplot(aes(x = region_signif, y = coverage, fill = region_signif)) +
  geom_boxplot() +
  theme(legend.position = "none")

dtu_optile %>% mutate(reads = rowSums(.[,2:8])) %>%
  group_by(gene_id) %>%
  summarise(coverage = sum(reads)) %>%
  mutate(region_signif = ifelse(gene_id %in% filter(results_df, FDR < 0.05)$region_id, TRUE, FALSE)) %>%
  ggplot(aes(x = region_signif, y = log2(coverage), fill = region_signif)) +
  geom_boxplot() +
  theme(legend.position = "none")

dtu_optile %>% mutate(reads = rowSums(.[,2:8])) %>%
  group_by(gene_id) %>%
  summarise(coverage = sum(reads)) %>%
  mutate(region_signif = ifelse(gene_id %in% filter(results_df, FDR < 0.05)$region_id, TRUE, FALSE)) %>%
  ggplot(aes(x = log2(coverage), colour = region_signif)) +
  geom_density()
```
* as expected, higher coverage in significant regions


## visualise the regions
```{r}
rrbs_plot5 <- function(read_data, read_ids, cpgs = cg_sites, jit_amount = 0.1) {
  set.seed(42) # For reproducibility
  # Calculate a single jitter amount per row (segment)
  read_data <- read_data %>% dplyr::filter(read_id %in% read_ids)
  jitter_amount <- runif(nrow(read_data), -jit_amount, jit_amount)
  read_data$alpha_jit <- read_data$alpha + jitter_amount
  read_data <- data.frame(read_data)

  # Add jittered coordinates to new columns in the data frame
  long_data <- read_data %>%
    mutate(meth_status = strsplit(meth_pattern, ""), start = strsplit(genom_positions, ",")) %>%
    unnest(meth_status, start) %>%
    mutate(start = as.double(start))
  
  plot <- read_data %>%
    ggplot() +
    geom_segment(aes(x = read_start, xend = read_end, y = alpha_jit, yend = alpha_jit), alpha = 0.2) +
    geom_point(data = long_data,
               aes(x = start, y = alpha_jit, fill = meth_status), shape = 21, color = "black") +
    scale_fill_manual(values = c("U" = "white", "M" = "black")) +
    facet_wrap(~ sample)
  
  distrib <- read_data %>%
    ggplot(aes(x = alpha, colour = alpha)) +
    geom_density()
  
  list(plot, distrib)
}
```


```{r}
signif_results <- results_df %>% filter(FDR < 0.05) %>% .[order(.$FDR),]
signif_results[1,]$excluded
```

```{r}
reads <- find_overlaps(as_granges(signif_results[1,]), as_granges(filter(all_samples, !sample %in% signif_results[1,]$excluded))) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(as_granges(signif_results[2,]), as_granges(filter(all_samples, !sample %in% signif_results[2,]$excluded))) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(as_granges(signif_results[3,]), as_granges(filter(all_samples, !sample %in% signif_results[3,]$excluded))) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(as_granges(signif_results[4,]), as_granges(filter(all_samples, !sample %in% signif_results[4,]$excluded))) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(as_granges(signif_results[30,]), as_granges(filter(all_samples, !sample %in% signif_results[30,]$excluded))) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(as_granges(signif_results[79,]), as_granges(filter(all_samples, !sample %in% signif_results[79,]$excluded))) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)
```

```{r}
signif_results <- results_df %>% filter(FDR > 0.05) %>% .[order(.$FDR),]

reads <- find_overlaps(as_granges(signif_results[1,]), as_granges(filter(all_samples, !sample %in% signif_results[1,]$excluded))) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(as_granges(signif_results[10,]), as_granges(filter(all_samples, !sample %in% signif_results[10,]$excluded))) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(as_granges(signif_results[100,]), as_granges(filter(all_samples, !sample %in% signif_results[100,]$excluded))) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(as_granges(signif_results[1000,]), as_granges(filter(all_samples, !sample %in% signif_results[1000,]$excluded))) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(as_granges(signif_results[100000,]), as_granges(filter(all_samples, !sample %in% signif_results[100000,]$excluded))) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(as_granges(signif_results[1000000,]), as_granges(filter(all_samples, !sample %in% signif_results[1000000,]$excluded))) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)
```


# repeat but with the alpha band level results alpha band + tile

```{r, eval=FALSE}
samples <- c(unique(all_samples$sample))  # your 7 sample names

# --- Analysis function ---
run_analysis <- function(combo) {
  cols <- c(combo$g1, combo$g2)
  group <- c(1, 1, 1, 2, 2, 2)

  y <- DGEList(
    as.matrix(dtu_optile[, cols]),
    group = group,
    genes = dtu_optile[, c("gene_id", "transcript")]
  )
  design <- model.matrix(~ group)
  fit <- glmQLFit(y, design, robust = TRUE)
  ds  <- diffSpliceDGE(fit, coef = 2, geneid = "gene_id", exonid = "transcript")

  topSpliceDGE(ds, test = "exon", number = Inf) %>%
    mutate(
      region_id = gene_id,
      excluded  = combo$excluded,
      group1    = paste(combo$g1, collapse = ","),
      group2    = paste(combo$g2, collapse = ",")
    ) %>%
    separate_wider_delim(gene_id, delim = ":", names = c("seqnames", "gene_id")) %>%
    mutate(seqnames = gsub("chrchr", "chr", seqnames)) %>%
    separate_wider_delim(gene_id, delim = "-", names = c("start", "end")) %>%
    mutate(
      start    = as.double(start),
      end      = as.double(end),
      dtu_rank = row_number()
    )
}

# --- Generate 70 combinations ---
all_combos <- unlist(
  lapply(seq_along(samples), function(excl_i) {
    remaining <- samples[-excl_i]
    splits <- combn(remaining, 3, simplify = FALSE)
    splits <- Filter(function(g1) g1[1] < setdiff(remaining, g1)[1], splits)
    lapply(splits, function(g1) {
      list(g1 = g1, g2 = setdiff(remaining, g1), excluded = samples[excl_i])
    })
  }),
  recursive = FALSE
)

# --- Run across all 70 combinations ---
results <- lapply(all_combos, run_analysis)

# --- Combine into one data frame ---
alpha_band_results_df <- bind_rows(results)

saveRDS(alpha_band_results_df, "/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_emseq_healthy_all_fp_exon_level.rds")
```

```{r}
alpha_band_results_df <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_emseq_healthy_all_fp_exon_level.rds")
```

```{r}
alpha_band_results_df %>% 
  ggplot(aes(x = FDR)) +
  geom_histogram()
```



```{r}
alpha_band_results_df[1:10,]
```

```{r}
alpha_band_results_df %>% filter(FDR < 0.05) %>% nrow()
```

```{r}
alpha_band_results_df %>% distinct(excluded, group1, group2) %>% nrow() # checking correct number of runs
```

```{r}
alpha_band_results_df %>% mutate(fdr_signif = ifelse(FDR < 0.05, TRUE, FALSE)) %>% group_by(excluded, group1, group2, fdr_signif) %>%
  summarise(n = n()) %>%
  ungroup() %>%
  group_by(excluded, group1, group2) %>%
  mutate(total_group = sum(n), prop = n/sum(n)) %>%
  ungroup() %>%
  filter(fdr_signif == TRUE) %>% .[1:5,]
```

* these are some tiny, beautifully low proportions
* crazy how low it is

* also this is only 42 of the 70 groups - so some groups didn't have any significant - no false positives

* this is expected to be related to coverage

```{r}
dtu_optile %>% mutate(reads = rowSums(.[,2:8])) %>%
  group_by(gene_id) %>%
  summarise(coverage = sum(reads)) %>%
  mutate(region_signif = ifelse(gene_id %in% filter(alpha_band_results_df, FDR < 0.05)$region_id, TRUE, FALSE)) %>%
  ggplot(aes(x = region_signif, y = coverage, fill = region_signif)) +
  geom_boxplot() +
  theme(legend.position = "none")

dtu_optile %>% mutate(reads = rowSums(.[,2:8])) %>%
  group_by(gene_id) %>%
  summarise(coverage = sum(reads)) %>%
  mutate(region_signif = ifelse(gene_id %in% filter(alpha_band_results_df, FDR < 0.05)$region_id, TRUE, FALSE)) %>%
  ggplot(aes(x = region_signif, y = log2(coverage), fill = region_signif)) +
  geom_boxplot() +
  theme(legend.position = "none")

dtu_optile %>% mutate(reads = rowSums(.[,2:8])) %>%
  group_by(gene_id) %>%
  summarise(coverage = sum(reads)) %>%
  mutate(region_signif = ifelse(gene_id %in% filter(alpha_band_results_df, FDR < 0.05)$region_id, TRUE, FALSE)) %>%
  ggplot(aes(x = log2(coverage), colour = region_signif)) +
  geom_density()
```

* want more specific plots using the alpha bands but I'm not quite sure what I want to do that as yet

## visualise reads

```{r}
signif_results <- alpha_band_results_df %>% filter(FDR < 0.05) %>% .[order(.$FDR),]
signif_results[nrow(signif_results),]
```

```{r}
reads <- find_overlaps(as_granges(signif_results[1,]), as_granges(filter(all_samples, !sample %in% signif_results[1,]$excluded))) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(as_granges(signif_results[2,]), as_granges(filter(all_samples, !sample %in% signif_results[2,]$excluded))) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(as_granges(signif_results[3,]), as_granges(filter(all_samples, !sample %in% signif_results[3,]$excluded))) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(as_granges(signif_results[4,]), as_granges(filter(all_samples, !sample %in% signif_results[4,]$excluded))) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(as_granges(signif_results[nrow(signif_results),]), as_granges(filter(all_samples, !sample %in% signif_results[nrow(signif_results),]$excluded))) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)
```

```{r}
signif_results <- alpha_band_results_df %>% filter(FDR > 0.05) %>% .[order(.$FDR),]

reads <- find_overlaps(as_granges(signif_results[1,]), as_granges(filter(all_samples, !sample %in% signif_results[1,]$excluded))) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(as_granges(signif_results[10,]), as_granges(filter(all_samples, !sample %in% signif_results[10,]$excluded))) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(as_granges(signif_results[100,]), as_granges(filter(all_samples, !sample %in% signif_results[100,]$excluded))) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(as_granges(signif_results[1000,]), as_granges(filter(all_samples, !sample %in% signif_results[1000,]$excluded))) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(as_granges(signif_results[nrow(signif_results),]), as_granges(filter(all_samples, !sample %in% signif_results[nrow(signif_results),]$excluded))) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)
```


## compare
```{r}

```

```{r}
results_df %>% filter(FDR < 0.05)
```

```{r}
results_df %>% mutate(sig_fdr = ifelse(FDR < 0.05, TRUE, FALSE), version = paste0(excluded, "_", group1, ":", group2)) %>% group_by(version, sig_fdr) %>% summarise(n=n()) %>% ungroup() %>% group_by(version) %>% mutate(prop = n/sum(n)) %>% filter(sig_fdr == TRUE)
```

```{r}
results_df %>% mutate(sig_fdr = ifelse(FDR < 0.05, TRUE, FALSE), version = paste0(excluded, "_", group1, ":", group2)) %>% group_by(version, sig_fdr) %>% summarise(n=n()) %>% ungroup() %>% group_by(version) %>% mutate(prop = n/sum(n)) %>% filter(sig_fdr == TRUE) %>%
  ggplot(aes(x = as.factor(n))) +
  geom_bar()
```


```{r}
results_df %>% mutate(sig_fdr = ifelse(FDR < 0.05, TRUE, FALSE), version = paste0(excluded, "_", group1, ":", group2)) %>% distinct(region_id)

results_df %>% mutate(sig_fdr = ifelse(FDR < 0.05, TRUE, FALSE), version = paste0(excluded, "_", group1, ":", group2)) %>% group_by(version) %>% summarise(n_region = n()) %>% ungroup() %>%
  ggplot(aes(x = n_region)) +
  geom_histogram()
```

```{r}
6/69968
```

### random plots for right now

```{r}
all_samples_simp <- all_samples %>% distinct(sample, read_id, seqnames, read_start, read_end, meth_pattern, alpha, num_meth, total)
```


```{r}
all_samples_simp %>% mutate(width = read_end - read_start + 1) %>%
  ggplot(aes(x = width)) +
  geom_histogram(bins = 50)

```

```{r}
all_samples_simp %>% mutate(width = read_end - read_start + 1) %>% filter(total >= 3) %>%
  ggplot(aes(x = width)) +
  geom_histogram(bins = 50)
```

```{r}
all_samples_simp$start <- all_samples_simp$read_start
all_samples_simp$end <- all_samples_simp$read_end
```


```{r}
tiles_reads <- find_overlaps(as_granges(optiles_all), as_granges(all_samples_simp)) %>% data.frame()
```


```{r}
tiles_reads %>% filter(total >= 3) %>% group_by(region_id, sample) %>% summarise(n=n()) %>%
  ggplot(aes(x = log2(n), colour = sample)) +
  geom_density()
```
```{r}
tiles_reads %>% filter(total >= 3) %>% group_by(region_id, sample) %>% summarise(n=n()) %>%
  ggplot(aes(x = log2(n), fill = sample)) +
  geom_histogram(bins = 50)
```

# read overlap multiple tiles??

```{r}
tiles_reads %>% group_by(read_id) %>% summarise(n=n()) %>%
  ggplot(aes(x = as.factor(n))) +
  geom_bar()
```

```{r}
tiles_reads %>% group_by(read_id) %>% summarise(n=n())
```

```{r}
tiles_reads %>% group_by(read_id) %>% summarise(n=n()) %>% ungroup() %>%
  group_by(n) %>% summarise(how_many = n())
```

```{r}
tiles_reads %>% group_by(read_id) %>% summarise(n=n()) %>% ungroup() %>%
  ggplot(aes(x = as.factor(n))) +
  geom_bar()
```

```{r}
dtu_emseq_cancer <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_emseq_cancer_healthy.rds")
```

```{r}
dtu_emseq_cancer %>% filter(FDR < 0.05)
```

```{r}
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
genes_txdb <- genes(TxDb.Hsapiens.UCSC.hg38.knownGene)
```

```{r}
library(missMethyl)
library(GO.db)
```
```{r}
genes_5 <- find_overlaps(as_granges(genes_txdb), as_granges(dtu_emseq_cancer)) %>% data.frame()
```


```{r}
txdb_genes <- find_overlaps(genes_txdb, as_granges(cg_sites)) %>% data.frame() %>%
  group_by(gene_id) %>% mutate(num_cg = n()) %>% ungroup() %>% distinct()
```


```{r}
source("../code/runGST.R")
gene_try <- genes_5 %>% filter(FDR < 0.05)
gene_info <- txdb_genes %>% mutate(is_de = ifelse(gene_id %in% gene_try$gene_id, 1, 0)) %>% distinct(gene_id, num_cg, is_de, width)

run_test <- list(
  sig.eg = gene_info %>% filter(is_de == 1) %>% .$gene_id,
  universe = gene_info$gene_id,
  freq = gene_info$num_cg,
  de = gene_info$is_de,
  equiv = gene_info$num_cg
)

gst_5_filt <- run_miss_methyl_b(run_test, method = "Wallenius")

saveRDS(gst_5_filt, "/researchers/caitlin.page/cf_nano/r_output/gst_emseq.rds")
```

```{r}
dtu.optiles <- readRDS("/researchers/caitlin.page/cf_nano/r_output/ts.gene_optiles.rds")
genes_rrbs <- find_overlaps(as_granges(genes_txdb), as_granges(dtu_emseq_cancer)) %>% data.frame()
```


```{r}
source("../code/runGST.R")
gene_try <- genes_rrbs %>% filter(FDR < 0.05)
gene_info <- txdb_genes %>% mutate(is_de = ifelse(gene_id %in% gene_try$gene_id, 1, 0)) %>% distinct(gene_id, num_cg, is_de, width)

run_test <- list(
  sig.eg = gene_info %>% filter(is_de == 1) %>% .$gene_id,
  universe = gene_info$gene_id,
  freq = gene_info$num_cg,
  de = gene_info$is_de,
  equiv = gene_info$num_cg
)

gst_5_filt <- run_miss_methyl_b(run_test, method = "Wallenius")

saveRDS(gst_5_filt, "/researchers/caitlin.page/cf_nano/r_output/gst_rrbs.rds")
```

```{r}
list.files("/researchers/caitlin.page/cf_nano/r_output")
```

```{r}
gst_emseq <- readRDS("/researchers/caitlin.page/cf_nano/r_output/gst_emseq.rds")
gst_rrbs <- readRDS("/researchers/caitlin.page/cf_nano/r_output/gst_rrbs.rds") #this is wrong - accidentally ran it on the cancer data
```

```{r}
gst_rrbs[[1]] %>% filter(FDR < 0.05)
```

```{r}
go_gm12878_vs_k562 <- c(

  ## ── Shared haematopoiesis ──────────────────────────────────────────────────
  "GO:0030097", # hemopoiesis
  "GO:0048534", # hematopoietic or lymphoid organ development
  "GO:0002521", # leukocyte differentiation
  "GO:0002244", # hematopoietic progenitor cell differentiation
  "GO:0060218", # hematopoietic stem cell differentiation

  ## ── B cell / lymphoid (GM12878-enriched) ──────────────────────────────────
  "GO:0030183", # B cell differentiation
  "GO:0042113", # B cell activation
  "GO:0019724", # B cell mediated immunity
  "GO:0050853", # B cell receptor signaling pathway
  "GO:0050851", # antigen receptor-mediated signaling pathway
  "GO:0046649", # lymphocyte activation
  "GO:0002376", # immune system process
  "GO:0006955", # immune response
  "GO:0002250", # adaptive immune response
  "GO:0006959", # humoral immune response
  "GO:0002377", # immunoglobulin production
  "GO:0016064", # immunoglobulin-mediated immune response

  ## ── Antigen presentation (GM12878) ────────────────────────────────────────
  "GO:0019882", # antigen processing and presentation
  "GO:0002474", # antigen processing and presentation of peptide antigen via MHC class I
  "GO:0002495", # antigen processing and presentation of peptide antigen via MHC class II
  "GO:0006954", # inflammatory response

  ## ── Erythroid differentiation (K562-enriched) ────────────────────────────
  "GO:0030218", # erythrocyte differentiation
  "GO:0048821", # erythrocyte development
  "GO:0030099", # myeloid cell differentiation
  "GO:0002043", # blood vessel endothelial cell proliferation involved in sprouting angiogenesis

  ## ── Haemoglobin / oxygen transport (K562) ─────────────────────────────────
  "GO:0015671", # oxygen transport
  "GO:0019825", # oxygen binding
  "GO:0020037", # heme binding
  "GO:0005833", # hemoglobin complex
  "GO:0042744", # hydrogen peroxide catabolic process
  "GO:0006979", # response to oxidative stress

  ## ── Megakaryocyte / platelet (K562) ───────────────────────────────────────
  "GO:0030099", # myeloid cell differentiation
  "GO:0030168", # platelet activation
  "GO:0070527", # platelet aggregation
  "GO:0007596", # blood coagulation
  "GO:0050817", # coagulation
  "GO:0002576", # platelet degranulation

  ## ── BCR-ABL / oncogenic kinase signalling (K562) ──────────────────────────
  "GO:0007169", # transmembrane receptor protein tyrosine kinase signaling pathway
  "GO:0007265", # Ras protein signal transduction
  "GO:0051897", # positive regulation of protein kinase B signaling
  "GO:0038095", # Fc-epsilon receptor signaling pathway
  "GO:0004672", # protein kinase activity
  "GO:0016301", # kinase activity
  "GO:0006468", # protein phosphorylation

  ## ── Cell cycle / proliferation (K562 > GM12878) ───────────────────────────
  "GO:0007049", # cell cycle
  "GO:0051726", # regulation of cell cycle
  "GO:0042127", # regulation of cell population proliferation
  "GO:0008283", # cell population proliferation
  "GO:0000279", # M phase
  "GO:0000087", # mitotic M phase

  ## ── Apoptosis / survival ───────────────────────────────────────────────────
  "GO:0043066", # negative regulation of apoptotic process
  "GO:0006915", # apoptotic process
  "GO:0097190", # apoptotic signaling pathway

  ## ── Metabolism (K562 is highly glycolytic) ────────────────────────────────
  "GO:0006006", # glucose metabolic process
  "GO:0019318", # hexose metabolic process
  "GO:0006096", # glycolytic process
  "GO:0006120", # mitochondrial electron transport, NADH to ubiquinone
  "GO:0045333"  # cellular respiration
)
```

```{r}
gst_rrbs[[1]] %>% mutate(rank = 1:nrow(.)) %>% filter(GOID %in% go_gm12878_vs_k562)
```

```{r}
gst_rrbs[[1]] %>% filter(FDR < 0.05)
gst_emseq[[1]] %>% filter(FDR < 0.05)
```


```{r}
gst_emseq[[1]] %>% mutate(rank = 1:nrow(.), in_set = cumsum(GOID %in% go_adrenal_ppgl)) %>%
  ggplot(aes(x = rank, y = in_set)) +
  geom_line(linewidth = 2) +
  geom_vline(aes(xintercept = 298), colour = "red", linetype = "dashed")
```


```{r}
gst_emseq[[1]] %>% filter(FDR < 0.05)
```



```{r}
gst_emseq[[1]] %>% filter(GOID %in% c("GO:0030325", "GO:0034651", "GO:0031960"))
```

```{r}
go_adrenal_ppgl <- c(
  # Adrenal gland development & differentiation
  "GO:0030325", # adrenal gland development
  "GO:0072047", # adrenal gland morphogenesis
  "GO:0014033", # neural crest cell differentiation
  "GO:0007422", # peripheral nervous system development
  "GO:0021675", # nerve development
  "GO:0048666", # neuron development
  "GO:0007400", # neuroblast fate determination

  # Chromaffin cell biology
  "GO:0060086", # chromaffin cell differentiation
  "GO:0035924", # cellular response to VEGF stimulus
  "GO:0006887", # exocytosis
  "GO:0017156", # calcium ion-regulated exocytosis
  "GO:0030072", # peptide hormone secretion
  "GO:0046903", # secretion

  # Catecholamine biosynthesis and metabolism
  "GO:0006584", # catecholamine metabolic process
  "GO:0006585", # catecholamine biosynthetic process
  "GO:0042415", # norepinephrine metabolic process
  "GO:0042416", # epinephrine metabolic process
  "GO:0042420", # dopamine metabolic process
  "GO:0042423", # catecholamine biosynthetic process from tyrosine
  "GO:0009308", # amine metabolic process
  "GO:0042402", # cellular biogenic amine catabolic process
  "GO:0042398", # cellular modified amino acid biosynthetic process

  # Hypoxia / pseudohypoxia
  "GO:0001666", # response to hypoxia
  "GO:0036293", # response to decreased oxygen levels
  "GO:0071456", # cellular response to hypoxia
  "GO:0061418", # regulation of transcription in response to hypoxia
  "GO:0001935", # endothelial cell proliferation
  "GO:0001525", # angiogenesis
  "GO:0001568", # blood vessel development
  "GO:0045766", # positive regulation of angiogenesis

  # HIF / mTOR / kinase signalling
  "GO:0032007", # negative regulation of TOR signaling
  "GO:0031929", # TOR signaling
  "GO:0043200", # response to amino acid
  "GO:0007189", # adenylate cyclase-activating GPCR signalling
  "GO:0007265", # Ras protein signal transduction
  "GO:0007264", # small GTPase-mediated signal transduction
  "GO:0004672", # protein kinase activity
  "GO:0016301", # kinase activity

  # Steroid hormone synthesis (adrenal cortex)
  "GO:0006700", # C21-steroid hormone biosynthetic process
  "GO:0006704", # glucocorticoid biosynthetic process
  "GO:0006705", # mineralocorticoid biosynthetic process
  "GO:0008207", # C21-steroid hormone metabolic process
  "GO:0016125", # sterol metabolic process
  "GO:0008202", # steroid metabolic process

  # DNA repair / TCA / SDHx-related
  "GO:0006099", # tricarboxylic acid cycle
  "GO:0045333", # cellular respiration
  "GO:0006626", # protein targeting to mitochondrion
  "GO:0006355", # regulation of DNA-templated transcription
  "GO:0006281", # DNA repair
  "GO:0010212", # response to ionizing radiation
  "GO:0000703", # oxidative phosphorylation uncoupler activity
  "GO:0008177", # succinate dehydrogenase (ubiquinone) activity

  # Tumour suppressor / oncogenesis
  "GO:0051726", # regulation of cell cycle
  "GO:0042127", # regulation of cell population proliferation
  "GO:0043066", # negative regulation of apoptotic process
  "GO:0007049", # cell cycle
  "GO:0016055", # Wnt signalling pathway
  "GO:0030307", # positive regulation of cell growth
  "GO:0045087"  # innate immune response
)
```

```{r}
gst_emseq[[1]] %>% mutate(rank = 1:nrow(.)) %>% filter(GOID %in% go_adrenal_ppgl)
```

```{r}
gst_emseq[[1]] %>% mutate(rank = 1:nrow(.)) %>% filter(FDR < 0.05)
```


```{r}
gst_emseq[[1]] %>% mutate(rank = 1:nrow(.), in_set = cumsum(GOID %in% go_adrenal_ppgl)) %>%
  ggplot(aes(x = rank, y = in_set)) +
  geom_line(linewidth = 2) +
  geom_vline(aes(xintercept = 298), colour = "red", linetype = "dashed")
```


truth_gst_100 <- rbind(
  gst_res_dmr1[1:100,] %>% mutate(in_imm = cumsum(GOID %in% immuneGO$GOID), method = "Wall_1") %>% .[names(.) %in% c("rank_fdr", "in_imm", "method")],
  gst_res_dmr1_hyper[1:100,] %>% mutate(in_imm = cumsum(GOID %in% immuneGO$GOID), method = "Hyper_1") %>% .[names(.) %in% c("rank_fdr", "in_imm", "method")])

truth_gst_100 %>%
  ggplot(aes(x = rank_fdr, y = in_imm, colour = method)) +
  geom_line(size = 2) +
  labs(x = "Rank", y = "Cumulative no. in immune sets") +
  theme_bw()

```{r}
gst_emseq[[1]] %>% mutate(in_set = ifelse(GOID %in% go_adrenal_ppgl, TRUE, FALSE))
```


## rrbs tiles

```{r}
tiles_cg <- find_overlaps(as_granges(optiles_all), as_granges(cg_sites)) %>% data.frame()
tiles_cg %>%
  group_by(region_id) %>%
  summarise(n_cg = n()) %>%
  ggplot(aes(x = n_cg)) +
  geom_histogram(bins = 70)
```

```{r}
emseq_cancer <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.rds")
```


```{r}
all_samples %>% filter(total >= 3) %>% filter(seqnames %in% c("chr1", "chr2", "chr3", "chr4", "chr5")) %>%
  ggplot(aes(x = seqnames, fill = seqnames)) +
  geom_bar()
```

```{r}
rrbs_samples <- readRDS("/researchers/caitlin.page/cf_nano/r_output/all_rrbs_samples.rds")

rrbs_samples <- rrbs_samples %>% distinct(seqnames, read_start, read_end, sample, total, alpha) %>% filter(total >= 3)
rrbs_samples %>%
  filter(seqnames %in% c("chr1", "chr2", "chr3", "chr4", "chr5")) %>%
  ggplot(aes(x = seqnames, fill = seqnames)) +
  geom_bar()
```
