---
title: "emseqAnalysis"
author: "caitlinpage"
date: "2026-07-27"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

## Introduction

```{r}
library(BSgenome.Hsapiens.UCSC.hg38)
library(data.table)
library(plyranges)
library(tidyr)
library(stringr)
library(dplyr)
library(ggplot2)
library(OpTiles)
library(edgeR)
library(UpSetR)
```

```{r}
# cpgs
seqs <- seqnames(BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38)[1:22]

cg_sites <- lapply(seqs, function(x) {
  cbind(data.frame(Biostrings::matchPattern("CG", BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38[[x]])),
        seqnames = x
  )}) %>% bind_rows()
```


```{r}
healthy_samples <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_all_healthy.rds")
#healthy_samples <- healthy_samples %>% filter(seqnames %in% seqs)
#healthy_samples <- distinct(healthy_samples)

optiles_all <- readRDS("/researchers/caitlin.page/cf_nano/r_output/optiles_all_emseq.rds") # built from healthy

cancer_samples <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.rds")

healthy_samples$type <- "healthy"
cancer_samples$type <- "cancer"
```



```{r}
dtu_tiles <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_emseq_cancer_healthy.rds")
dtu_tiles_healthy_compare <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_emseq_healthy.rds")
```

```{r}
1+1
```

# Overview of samples

```{r}
healthy_samples %>%
  ggplot(aes(x = total)) +
  geom_density()

healthy_samples %>%
  ggplot(aes(x = alpha)) +
  geom_density()
```

```{r}
cancer_samples %>%
  ggplot(aes(x = total)) +
  geom_density()

cancer_samples %>%
  ggplot(aes(x = alpha)) +
  geom_density()
```


# Analyse the healthy x healthy results


```{r}
dtu_tiles_healthy_compare %>% mutate(false_pos = ifelse(FDR < 0.05, TRUE, FALSE), tile = "tile") %>%
  ggplot(aes(x = tile, fill = false_pos)) +
  geom_bar(position = "fill")
```

```{r}
dtu_tiles_healthy_compare %>% mutate(false_pos = ifelse(FDR < 0.05, TRUE, FALSE), tile = "tile") %>%
  ggplot(aes(x = tile, fill = false_pos)) +
  geom_bar() +
  coord_polar("y", start = 0)
```

```{r}
dtu_tiles_healthy_compare %>% mutate(false_pos = ifelse(FDR < 0.05, TRUE, FALSE), tile = "tile") %>%
  group_by(false_pos) %>%
  summarise(n = n()) %>%
  ungroup() %>%
  mutate(ratio = n/sum(n))
```

```{r}
dtu_tiles_healthy_compare %>%
  ggplot(aes(x = FDR)) +
  geom_density()
```

```{r}
dtu_tiles_healthy_compare %>% filter(FDR < 0.05) %>% nrow()
```

11 false positives
I'll take it

* visualise those regions
```{r}
healthy_samples$read_start <- healthy_samples$start
healthy_samples$read_end <- healthy_samples$end
```


* cool cool got a strand thing going on with these reads - same read_id, oppo strand
```{r}
rrbs_plot5 <- function(read_data, read_ids, cpgs = cg_sites, jit_amount = 0.1) {
  set.seed(42) # For reproducibility
  # Calculate a single jitter amount per row (segment)
  read_data <- read_data %>% dplyr::filter(read_id %in% read_ids)
  jitter_amount <- runif(nrow(read_data), -jit_amount, jit_amount)
  read_data$alpha_jit <- read_data$alpha + jitter_amount
  read_data <- data.frame(read_data)

  # Add jittered coordinates to new columns in the data frame
  long_data <- read_data %>%
    mutate(meth_status = strsplit(meth_pattern, ""), start = strsplit(genom_positions, ",")) %>%
    unnest(meth_status, start) %>%
    mutate(start = as.double(start))
  
  plot <- read_data %>%
    ggplot() +
    geom_segment(aes(x = read_start, xend = read_end, y = alpha_jit, yend = alpha_jit), alpha = 0.2) +
    geom_point(data = long_data,
               aes(x = start, y = alpha_jit, fill = meth_status), shape = 21, color = "black") +
    scale_fill_manual(values = c("U" = "white", "M" = "black")) +
    facet_wrap(~ sample)
  plot
}
```

```{r}
reads <- find_overlaps(dtu_tiles_healthy_compare %>% filter(FDR < 0.05) %>% .[3,] %>% as_granges(), as_granges(healthy_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(healthy_samples, reads)

reads <- find_overlaps(dtu_tiles_healthy_compare %>% filter(FDR < 0.05) %>% .[4,] %>% as_granges(), as_granges(healthy_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(healthy_samples, reads)

reads <- find_overlaps(dtu_tiles_healthy_compare %>% filter(FDR < 0.05) %>% .[5,] %>% as_granges(), as_granges(healthy_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(healthy_samples, reads)

reads <- find_overlaps(dtu_tiles_healthy_compare %>% filter(FDR < 0.05) %>% .[6,] %>% as_granges(), as_granges(healthy_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(healthy_samples, reads)

reads <- find_overlaps(dtu_tiles_healthy_compare %>% filter(FDR < 0.05) %>% .[7,] %>% as_granges(), as_granges(healthy_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(healthy_samples, reads)

reads <- find_overlaps(dtu_tiles_healthy_compare %>% filter(FDR < 0.05) %>% .[8,] %>% as_granges(), as_granges(healthy_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(healthy_samples, reads)

reads <- find_overlaps(dtu_tiles_healthy_compare %>% filter(FDR < 0.05) %>% .[9,] %>% as_granges(), as_granges(healthy_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(healthy_samples, reads)

reads <- find_overlaps(dtu_tiles_healthy_compare %>% filter(FDR < 0.05) %>% .[10,] %>% as_granges(), as_granges(healthy_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(healthy_samples, reads)

reads <- find_overlaps(dtu_tiles_healthy_compare %>% filter(FDR < 0.05) %>% .[11,] %>% as_granges(), as_granges(healthy_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(healthy_samples, reads)
```

* ok so there is a change in proportion for some of these
* but also it's only 11 regions


* check some of the non-significant ones
```{r}
reads <- find_overlaps(dtu_tiles_healthy_compare %>% filter(FDR > 0.05) %>% .[3,] %>% as_granges(), as_granges(healthy_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(healthy_samples, reads)

reads <- find_overlaps(dtu_tiles_healthy_compare %>% filter(FDR > 0.05) %>% .[5,] %>% as_granges(), as_granges(healthy_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(healthy_samples, reads)

reads <- find_overlaps(dtu_tiles_healthy_compare %>% filter(FDR > 0.05) %>% .[10,] %>% as_granges(), as_granges(healthy_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(healthy_samples, reads)

reads <- find_overlaps(dtu_tiles_healthy_compare %>% filter(FDR > 0.05) %>% .[20,] %>% as_granges(), as_granges(healthy_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(healthy_samples, reads)

reads <- find_overlaps(dtu_tiles_healthy_compare %>% filter(FDR > 0.05) %>% .[100,] %>% as_granges(), as_granges(healthy_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(healthy_samples, reads)
```


# Analyse the cancer x healthy results

```{r}
dtu_tiles %>% filter(FDR < 0.05) %>% nrow()
```


```{r}
dtu_tiles %>% mutate(signif = ifelse(FDR < 0.05, TRUE, FALSE), tile = "tile") %>%
  ggplot(aes(x = tile, fill = signif)) +
  geom_bar(position = "fill")
```

```{r}
all_samples <- rbind(healthy_samples, cancer_samples) %>% data.frame()
```


```{r}
reads <- find_overlaps(dtu_tiles %>% filter(FDR < 0.05) %>% .[3,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(dtu_tiles %>% filter(FDR < 0.05) %>% .[4,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(dtu_tiles %>% filter(FDR < 0.05) %>% .[5,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(dtu_tiles %>% filter(FDR < 0.05) %>% .[6,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(dtu_tiles %>% filter(FDR < 0.05) %>% .[7,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(dtu_tiles %>% filter(FDR < 0.05) %>% .[8,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(dtu_tiles %>% filter(FDR < 0.05) %>% .[9,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(dtu_tiles %>% filter(FDR < 0.05) %>% .[10,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(dtu_tiles %>% filter(FDR < 0.05) %>% .[11,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)
```

* non significant tiles
```{r}
reads <- find_overlaps(dtu_tiles %>% filter(FDR > 0.05) %>% .[10,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(dtu_tiles %>% filter(FDR > 0.05) %>% .[110,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)
```


# new

```{r}
o <- find_overlaps(dtu_tiles %>% filter(FDR < 0.05) %>% .[1:10,] %>% as_granges(), as_granges(all_samples)) %>% data.frame()
```


```{r}
o <- o %>% distinct(sample, read_id, type, alpha, meth_pattern, num_meth, total) %>%
  mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                     alpha > 0.8 ~ "a > 0.8",
                                     alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                     alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                     TRUE ~ "0.6 <= a <= 0.8"))
  group_by(type, alpha_group) %>% summarise(n=n()) %>%
  ungroup() %>%
  group_by(type) %>% mutate(prop_in_sample = n/sum(n)) %>%
  ungroup() %>%
  group_by(alpha_group) %>%
  mutate(delta_alpha = prop_in_sample - lead(prop_in_sample))
```


```{r}
chr1 <- reads[seqnames == "chr1"]
saveRDS(chr1, "/researchers/caitlin.page/cf_nano/emseq_cancer_chr1.rds")
```


```{r}
reads %>% data.frame() %>% filter(seqnames %in% c("chr1", "chr2", "chr3", "chr4", "chr5")) %>%
  ggplot(aes(x = seqnames, fill = seqnames)) +
  geom_bar()
```

```{r}
betas <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_betas_cancer.rds")
betas_h <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_betas_healthy.rds")
```
```{r}
cols <- c("sample", "seqnames", "start", "meth_count_pos", "unmeth_count_pos", "beta") "cov")
betas_h[,c(1:3,5)]
```

```{r}
betas_h[,cols]
```

# pca time
```{r}
saveRDS(betas, "/researchers/caitlin.page/cf_nano/r_output/emseq_cancer_beta_only.rds")
```

```{r}
betas <- betas[,c(1:2,4:9)]
betas <- betas %>% group_by(seqnames, start) %>% fill(c(cancer_0223, cancer_0224, cancer_0225, cancer_0454, cancer_0456, cancer_0457), .direction = "downup") %>% distinct() %>% ungroup()
```

```{r}
unique(betas$seqnames)
```

```{r}
seqs <- c("chr1", "chr2", "chr3", "chr4", "chr5", "chr6", "chr7", "chr8", "chr9", "chr10",
          "chr11", "chr12", "chr13", "chr14", "chr15", "chr16", "chr17", "chr18", "chr19",
          "chr20", "chr21", "chr22")
```

* want it with the alpha bands
* which I think is just doing it with alpha?
* do it with beta firstmmm

```{r}
betas_h_mod <- betas_h %>% filter(seqnames %in% seqs) %>% .[,c(1:3,6)] %>%
  pivot_wider(names_from = sample, values_from = beta)
betas_h_mod
```

```{r}
betas$pos <- paste0(betas$seqnames, ":", betas$start)
betas_h_mod$pos <- paste0(betas_h_mod$seqnames, ":", betas_h_mod$start)
```

```{r}
all_b <- full_join(betas_h_mod, betas, by = "pos")
```

```{r}
all_b_filt <- na.omit(all_b)
all_b_filt <- data.frame(all_b_filt)
rownames(all_b_filt) <- all_b_filt$pos
```

```{r}
t(all_b_filt[1:5,1:5])
```

```{r}
rot <- data.frame(t(all_b_filt[,c(3:8,12:17)]))
a <- as.matrix(rot[1:5,1:5])
```


```{r}
pca <- prcomp(rot, center = TRUE, scale. = FALSE)
```

```{r}
pca$x
?prcomp
```


```{r}
pca_df <- as.data.frame(pca$x)
pca_df$Sample <- rownames(pca_df)

# Add any metadata you have (group, tissue, batch, etc.)
pca_df$Group <- c("healthy", "healthy", "healthy", "healthy", "healthy", "healthy",
                  "cancer", "cancer", "cancer", "cancer", "cancer", "cancer")  # match sample order

# Variance explained
var_explained <- round(pca$sdev^2 / sum(pca$sdev^2) * 100, 1)

ggplot(pca_df, aes(x = PC1, y = PC2, colour = Group, label = Sample)) +
  geom_point(size = 4) 
```

```{r}
ggplot(pca_df, aes(x = PC1, y = PC2, colour = Group, label = Sample)) +
  geom_point(size = 4) +
  geom_text() +
#  ggrepel::geom_text_repel() +
  labs(
    x = paste0("PC1 (", var_explained[1], "%)"),
    y = paste0("PC2 (", var_explained[2], "%)"),
    title = "PCA of CpG Methylation"
  ) +
  theme_bw()
```

# alpha categories
(let's just go for chr1)


```{r}
cancer_chr1 <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_cancer_chr1.rds")
cancer_chr1 <- data.frame(cancer_chr1)
healthy <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_all_healthy.rds")
```

```{r}
healthy <- healthy %>% filter(seqnames == "chr1")
```


```{r}
emseq_tiles <- readRDS("/researchers/caitlin.page/cf_nano/r_output/optiles_all_emseq.rds")
```

```{r}
all_chr1 <- rbind(healthy, cancer_chr1[,colnames(healthy)]) %>% data.frame()
```

```{r}
overlap_tiles <- find_overlaps(as_granges(emseq_tiles), as_granges(all_chr1)) %>% data.frame()
```

```{r}
overlap_tiles <- overlap_tiles %>%
  mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                 alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                 alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                 alpha >= 0.6 & alpha <= 0.8 ~ "0.6 <= a <= 0.8",
                                 TRUE > 0.8 ~ "a > 0.8")) 
```

```{r}
tile_counts <- overlap_tiles %>% 
  mutate(transcript = paste0(region_id, "_", alpha_group)) %>%
  group_by(sample, transcript) %>%
  summarise(transcript_count = n()) %>% ungroup() %>% data.frame()
```
```{r}
tile_counts2 <- tile_counts %>% pivot_wider(names_from = transcript, values_from = transcript_count)
tile_counts2 <- data.frame(tile_counts2)
rownames(tile_counts2) <- tile_counts2$sample
```
```{r}
tile_counts_2 <- tile_counts2[,2:ncol(tile_counts2)]
tile_counts_2[is.na(tile_counts_2)] <- 0
```
```{r}
mat <- as.matrix(tile_counts_2)
mat[is.na(mat)] <- 0
tile_counts_2 <- as.data.frame(mat)
```

mm
```{r}
pca_a <- prcomp(tile_counts_2, center = TRUE, scale. = FALSE)
pca_df_a <- as.data.frame(pca_a$x)
pca_df_a$Sample <- rownames(pca_df_a)

# Add any metadata you have (group, tissue, batch, etc.)
pca_df_a$Group <- c("healthy", "healthy", "healthy", "healthy", "healthy", "healthy",
                  "cancer", "cancer", "cancer", "cancer", "cancer", "cancer")  # match sample order

# Variance explained
var_explained_a <- round(pca_a$sdev^2 / sum(pca_a$sdev^2) * 100, 1)

ggplot(pca_df_a, aes(x = PC1, y = PC2, colour = Group, label = Sample)) +
  geom_point(size = 4) 
```

```{r}
ggplot(pca_df_a, aes(x = PC1, y = PC2, colour = Group, label = Sample)) +
  geom_point(size = 4) +
  geom_text() +
#  ggrepel::geom_text_repel() +
  labs(
    x = paste0("PC1 (", var_explained[1], "%)"),
    y = paste0("PC2 (", var_explained[2], "%)"),
    title = "PCA of CpG Methylation"
  ) +
  theme_bw()
```
```{r}
ggplot(pca_df_a, aes(x = PC1, y = PC2, colour = Group, label = Sample)) +
  geom_point(size = 4) +
  geom_text(aes(label = Sample), 
          nudge_x = 200,   # shift right — adjust this to your axis scale
          nudge_y = 300,   # shift up
          hjust = 0,       # left-align text (so it extends away from the point)
          size = 3) +
  labs(
    x = paste0("PC1 (", var_explained[1], "%)"),
    y = paste0("PC2 (", var_explained[2], "%)"),
    title = "PCA of CpG Methylation"
  ) +
  theme_bw()

```

```{r}
rot
```

```{r}
saveRDS(overlap_tiles, "/researchers/caitlin.page/cf_nano/r_output/emseq_reads_tiles_chr1.rds")
```

```{r}
list.files("/researchers/caitlin.page/cf_nano/r_output")
```


```{r}
dtu_optile <- find_overlaps(as_granges(all_samples), as_granges(regions_definition)) %>% data.frame() %>%
  distinct(sample, read_id, read_start, read_end, alpha, region_id) %>%
  mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                 alpha > 0.8 ~ "a > 0.8",
                                 alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                 alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                 TRUE ~ "0.6 <= a <= 0.8"))
# count the occurence of the region+alpha ("transcript")
dtu_optile <- dtu_optile %>%
  mutate(transcript = paste0(region_id, "_", alpha_group)) %>%
  group_by(sample, transcript) %>%
  summarise(transcript_count = n()) %>% ungroup() %>% data.frame()

# build matrix of "transcript" counts - transcript rows, sample as columns
dtu_optile <- dtu_optile %>% pivot_wider(names_from = sample, values_from = transcript_count)
dtu_optile[is.na(dtu_optile)] <- 0
dtu_optile$gene_id <- sub("_.*", "", dtu_optile$transcript)
```

```{r}

```


```{r}
list.files("/researchers/caitlin.page/cf_nano/r_output")
```





```{r}
betas_h <- 
b <- b %>% filter(seqnames %in% seqs)
b$cov <- b$meth_count_pos + b$unmeth_count_pos
cols <- c("sample", "seqnames", "start", "meth_count_pos", "unmeth_count_pos", "cov")
b <- b[,cols] %>% pivot_wider(names_from = sample, values_from = c(meth_count_pos, unmeth_count_pos, cov))
b[is.na(b)] <- 0
```

