---
title: "emseqBiologyAnalysis"
author: "caitlinpage"
date: "2026-07-28"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

## Introduction

```{r}
library(BSgenome.Hsapiens.UCSC.hg38)
library(data.table)
library(plyranges)
#library(tidyr)
library(stringr)
library(dplyr)
library(ggplot2)
library(OpTiles)
library(edgeR)
library(UpSetR)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
```

```{r}
genes_txdb <- genes(TxDb.Hsapiens.UCSC.hg38.knownGene)
```

```{r}
tile_counts_2
tile_counts
```

```{r}
dtu_emseq <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_emseq_cancer_healthy.rds")
```

```{r}
dtu_gene <- find_overlaps(genes_txdb, as_granges(dtu_emseq)) %>% data.frame()
```

```{r}
dtu_gene %>% filter(FDR < 0.05) %>%
  group_by(gene_id) %>% summarise(n=dplyr::n()) %>% ungroup() %>% group_by(n) %>% summarise(n_regions_gene_overlaps = dplyr::n())
```

```{r}
find_overlaps(genes_txdb, as_granges(optiles_all)) %>% data.frame() %>%
  group_by(gene_id) %>% summarise(n_tile = n()) %>%
  ggplot(aes(x = n_tile)) +
  geom_histogram(bins = 50)
```


```{r}
dtu_gene %>% filter(FDR < 0.05) %>%
  group_by(gene_id) %>% summarise(n_sig_regions=dplyr::n()) %>% ungroup() %>%
  ggplot(aes(x = as.factor(n_sig_regions))) +
  geom_bar()
```

# 1 gene, multiple tiles
```{r}
dtu_gene %>% filter(FDR < 0.05) %>%
  group_by(gene_id) %>% mutate(n_sig_regions=dplyr::n()) %>% ungroup() %>% filter(n_sig_regions > 3) %>% .[order(.$n_sig_regions, decreasing = TRUE),] %>% .[order(.$dtu_rank),] %>% dplyr::distinct(gene_id, dtu_rank)

dtu_gene %>% filter(FDR < 0.05) %>%
  group_by(gene_id) %>% summarise(n_sig_regions=dplyr::n()) %>% ungroup() %>% filter(n_sig_regions > 3) %>% .[order(.$n_sig_regions, decreasing = TRUE),] dplyr::distinct(gene_id)
```

```{r}
dtu_emseq %>% filter(FDR < 0.05)
```

```{r}
dtu_gene %>% filter(FDR < 0.05) %>% 
    group_by(gene_id) %>% mutate(n_sig_regions=dplyr::n()) %>% ungroup() %>%
    filter(n_sig_regions > 3) %>% .[order(.$n_sig_regions, decreasing = TRUE),] %>% .[order(.$dtu_rank),] %>% dplyr::distinct(gene_id, dtu_rank) %>% .[1,]
```


```{r}
list_plot <- list()
for(i in 1:18) {
  
  tile_sig <- dtu_gene %>% filter(FDR < 0.05) %>% 
    group_by(gene_id) %>% mutate(n_sig_regions=dplyr::n()) %>% ungroup() %>%
    filter(n_sig_regions > 3) %>% .[order(.$n_sig_regions, decreasing = TRUE),] %>% .[order(.$dtu_rank),] %>% dplyr::distinct(gene_id, dtu_rank)
  
  p_data <- dtu_gene %>% filter(gene_id == tile_sig$gene_id[i]) %>% 
  tidyr::separate(region_id, into = c("chr", "tile_start", "tile_end"), sep = "[:-]") %>% 
  mutate(tile_start = as.integer(tile_start), tile_end = as.integer(tile_end),
         y_pos = dplyr::row_number())
p <- p_data %>%
  ggplot() +
  geom_segment(data = filter(p_data, strand == "-"), aes(x = start, xend = end, y = 0, yend = 0), linewidth = 2,
               arrow = arrow(length = unit(0.2, "cm"), type = "closed", ends = "first")) +
  geom_segment(data = filter(p_data, strand == "+"), aes(x = start, xend = end, y = 0, yend = 0), linewidth = 2,
               arrow = arrow(length = unit(0.2, "cm"), type = "closed", ends = "last")) +
  geom_segment(aes(x = tile_start, xend = tile_end, y = y_pos, yend = y_pos), linewidth = 2) +
  geom_text(aes(x = min(start), y = 0.2, label = gene_id),
          hjust = 1,    # right-align so all labels end at the same x
          nudge_x = -10000) +
coord_cartesian(clip = "off") +   # allows text to render outside the plot area
theme(plot.margin = margin(5, 5, 5, 60)) +
  geom_text(aes(x = tile_start, y = y_pos, label = dtu_rank)) +
  ggtitle(gsub("^chr", "", unique(p_data$chr)))

list_plot <- list(list_plot, p)
}
list_plot
```


tile_sig <- dtu_gene %>% filter(FDR < 0.05) %>% 
    group_by(region_id) %>% mutate(n_gene_overlap = dplyr::n()) %>% ungroup() %>%
    filter(n_gene_overlap > 3) %>% .[order(.$n_gene_overlap, decreasing = TRUE),] %>% .[order(.$dtu_rank),] %>% dplyr::distinct(region_id, dtu_rank)
  
  p_data <- dtu_gene %>% filter(region_id == tile_sig$region_id[i]) %>% 
  tidyr::separate(region_id, into = c("chr", "tile_start", "tile_end"), sep = "[:-]") %>% 
  mutate(tile_start = as.integer(tile_start), tile_end = as.integer(tile_end),
         y_pos = disjointBins(IRanges(start, end)))
p <- p_data %>%
  ggplot() +
  geom_segment(data = filter(p_data, strand == "-"), aes(x = start, xend = end, y = y_pos, yend = y_pos), linewidth = 2,
               arrow = arrow(length = unit(0.2, "cm"), type = "closed", ends = "first")) +
  geom_segment(data = filter(p_data, strand == "+"), aes(x = start, xend = end, y = y_pos, yend = y_pos), linewidth = 2,
               arrow = arrow(length = unit(0.2, "cm"), type = "closed", ends = "last")) +
  geom_segment(aes(x = tile_start, xend = tile_end, y = 27, yend = 27), linewidth = 2) +
  geom_text(aes(x = min(start), y = y_pos, label = gene_id),
          hjust = 1,    # right-align so all labels end at the same x
          nudge_x = -10000) +
coord_cartesian(clip = "off") +   # allows text to render outside the plot area
theme(plot.margin = margin(5, 5, 5, 60)) +
  geom_text(aes(x = tile_start, y = 26, label = dtu_rank)) +
  ggtitle(gsub("^chr", "", unique(p_data$chr)))


```{r}
dtu_gene %>% filter(FDR < 0.05) %>% dplyr::distinct(gene_id)
```

```{r}
dtu_gene %>% filter(FDR < 0.05) %>% dplyr::distinct(gene_id) %>% .$gene_id %>% .[2001:2538]
```

```{r}
dtu_gene %>% filter(FDR < 0.05) %>% dplyr::distinct(gene_id) %>% .$gene_id
```

```{r}
library(org.Hs.eg.db)
gene_types <- select(org.Hs.eg.db, keys = dtu_gene %>% filter(FDR < 0.05) %>% dplyr::distinct(gene_id) %>% .$gene_id, 
                     columns = c("SYMBOL", "GENENAME"), 
                     keytype = "ENTREZID")
```


# make some fake promoters

```{r}
genes_txdb
```
* 2000bp upstream, 200bp downstream
```{r}
promoters_txdb <- promoters(genes_txdb)
```


```{r}
dtu_promoter <- find_overlaps(promoters_txdb, as_granges(dtu_emseq)) %>% data.frame()
```

```{r}
dtu_promoter %>% filter(FDR < 0.05)
dtu_promoter %>% filter(FDR < 0.05) %>% dplyr::distinct(gene_id)
```

```{r}
dtu_promoter %>% filter(FDR < 0.05) %>%
  group_by(gene_id) %>% summarise(n_sig_regions=dplyr::n()) %>% ungroup() %>%
  ggplot(aes(x = as.factor(n_sig_regions))) +
  geom_bar()
```

```{r}
gene_types_prom <- select(org.Hs.eg.db, keys = dtu_promoter %>% filter(FDR < 0.05) %>% dplyr::distinct(gene_id) %>% .$gene_id, 
                     columns = c("SYMBOL", "GENENAME"), 
                     keytype = "ENTREZID")
```


```{r}
ggVennDiagram::ggVennDiagram(list(gene = dtu_gene %>% filter(FDR < 0.05) %>% dplyr::distinct(gene_id) %>% .$gene_id, promoter = dtu_promoter %>% filter(FDR < 0.05) %>% dplyr::distinct(gene_id) %>% .$gene_id))
```

# where do overlaps happen in gene

```{r}
dtu_gene %>% filter(FDR < 0.05) %>% .[order(.$FDR),]
```

```{r}
dtu_gene %>% filter(FDR < 0.05) %>% .[order(.$FDR),] %>% .[1,]
dtu_gene %>% filter(FDR < 0.05) %>% .[order(.$FDR),] %>% .[1,] %>% 
  tidyr::separate(region_id, into = c("chr", "tile_start", "tile_end"), sep = "[:-]") %>% 
  mutate(tile_start = as.integer(tile_start), tile_end = as.integer(tile_end)) %>% {
    strand_dir <- if (unique(.$strand) == "-") "first" else "last"
  ggplot(.) +
  geom_segment(aes(x = start, xend = end, y = 1, yend = 1), linewidth = 2,
               arrow = arrow(length = unit(0.2, "cm"), type = "closed", ends = strand_dir)) +
  geom_segment(aes(x = tile_start, xend = tile_end, y = 2, yend = 2), linewidth = 2) +
  geom_text(aes(x = start, y = 1.2, label = gene_id))
  }
```

```{r}
list_plot <- list()
for(i in 1:10) {
  
p <- dtu_gene %>% filter(FDR < 0.05) %>% .[order(.$FDR),] %>% .[i,] %>% 
  tidyr::separate(region_id, into = c("chr", "tile_start", "tile_end"), sep = "[:-]") %>% 
  mutate(tile_start = as.integer(tile_start), tile_end = as.integer(tile_end)) %>% {
    strand_dir <- if (unique(.$strand) == "-") "first" else "last"
    chr_label  <- unique(.$chr)
  ggplot(.) +
  geom_segment(aes(x = start, xend = end, y = 1, yend = 1), linewidth = 2,
               arrow = arrow(length = unit(0.2, "cm"), type = "closed", ends = strand_dir)) +
  geom_segment(aes(x = tile_start, xend = tile_end, y = 2, yend = 2), linewidth = 2) +
  geom_text(aes(x = start, y = 1.2, label = gene_id)) +
  geom_text(aes(x = tile_start, y = 1.8, label = dtu_rank)) +
  ggtitle(chr_label)
}
list_plot <- list(list_plot, p)
  }
list_plot
```

```{r}
dtu_gene %>% filter(FDR < 0.05) %>% .[order(.$FDR),]
```


# promoter

```{r}
list_plot <- list()
for(i in 1:10) {
  
p <- dtu_promoter %>% filter(FDR < 0.05) %>% .[order(.$FDR),] %>% .[i,] %>% 
  tidyr::separate(region_id, into = c("chr", "tile_start", "tile_end"), sep = "[:-]") %>% 
  mutate(tile_start = as.integer(tile_start), tile_end = as.integer(tile_end)) %>% {
    strand_dir <- if (unique(.$strand) == "-") "first" else "last"
    chr_label <- unique(.$chr)
  ggplot(.) +
  geom_segment(aes(x = start, xend = end, y = 1, yend = 1), linewidth = 2,
               arrow = arrow(length = unit(0.2, "cm"), type = "closed", ends = strand_dir)) +
  geom_segment(aes(x = tile_start, xend = tile_end, y = 2, yend = 2), linewidth = 2) +
  geom_text(aes(x = start, y = 1.2, label = gene_id)) +
  geom_text(aes(x = tile_start, y = 1.8, label = dtu_rank)) +
  ggtitle(chr_label)
  }

list_plot <- list(list_plot, p)
}
list_plot
```


```{r}
dtu_gene %>% filter(FDR < 0.05) %>%
  group_by(region_id) %>% summarise(n_gene_overlap = dplyr::n()) %>%
  ggplot(aes(x = as.factor(n_gene_overlap))) +
  geom_bar()

dtu_promoter %>% filter(FDR < 0.05) %>%
  group_by(region_id) %>% summarise(n_gene_overlap = dplyr::n()) %>%
  ggplot(aes(x = as.factor(n_gene_overlap))) +
  geom_bar()
```
```{r}
dtu_gene %>% filter(FDR < 0.05) %>% 
  group_by(region_id) %>% mutate(n_gene_overlap = dplyr::n()) %>% ungroup() %>%
  filter(n_gene_overlap > 3) %>% .[order(.$n_gene_overlap, decreasing = TRUE),] %>% .[order(.$dtu_rank),] 
```

```{r}
dtu_gene %>% filter(FDR < 0.05) %>% 
  group_by(region_id) %>% mutate(n_gene_overlap = dplyr::n()) %>% ungroup() %>%
  filter(n_gene_overlap > 3) %>% .[order(.$n_gene_overlap, decreasing = TRUE),] %>% .[order(.$dtu_rank),] %>% dplyr::distinct(region_id, dtu_rank)
```

```{r}
dtu_gene$region_id[5]
```

# 1 dtu, multiple genes

```{r}
list_plot <- list()
for(i in 1:18) {
  
  tile_sig <- dtu_gene %>% filter(FDR < 0.05) %>% 
    group_by(region_id) %>% mutate(n_gene_overlap = dplyr::n()) %>% ungroup() %>%
    filter(n_gene_overlap > 3) %>% .[order(.$n_gene_overlap, decreasing = TRUE),] %>% .[order(.$dtu_rank),] %>% dplyr::distinct(region_id, dtu_rank)
  
  p_data <- dtu_gene %>% filter(region_id == tile_sig$region_id[i]) %>% 
  tidyr::separate(region_id, into = c("chr", "tile_start", "tile_end"), sep = "[:-]") %>% 
  mutate(tile_start = as.integer(tile_start), tile_end = as.integer(tile_end),
         y_pos = disjointBins(IRanges(start, end)))
p <- p_data %>%
  ggplot() +
  geom_segment(data = filter(p_data, strand == "-"), aes(x = start, xend = end, y = y_pos, yend = y_pos), linewidth = 2,
               arrow = arrow(length = unit(0.2, "cm"), type = "closed", ends = "first")) +
  geom_segment(data = filter(p_data, strand == "+"), aes(x = start, xend = end, y = y_pos, yend = y_pos), linewidth = 2,
               arrow = arrow(length = unit(0.2, "cm"), type = "closed", ends = "last")) +
  geom_segment(aes(x = tile_start, xend = tile_end, y = 27, yend = 27), linewidth = 2) +
  geom_text(aes(x = min(start), y = y_pos, label = gene_id),
          hjust = 1,    # right-align so all labels end at the same x
          nudge_x = -10000) +
coord_cartesian(clip = "off") +   # allows text to render outside the plot area
theme(plot.margin = margin(5, 5, 5, 60)) +
  geom_text(aes(x = tile_start, y = 26, label = dtu_rank)) +
  ggtitle(gsub("^chr", "", unique(p_data$chr)))

list_plot <- list(list_plot, p)
}
list_plot
```



```{r}
tile_counts
```


# where do overlaps happen in promoter?

# PPGL genes (calude)

```{r}
ppgl_drivers <- data.frame(
  symbol = c(
    # Cluster 1 — Pseudohypoxia / TCA cycle
    "SDHA", "SDHB", "SDHC", "SDHD", "SDHAF2",
    "VHL", "FH", "MDH2", "EPAS1", "IDH1", "IDH2",
    "SLC25A11", "DLST", "GOT2",
    # Cluster 2 — Kinase / RAS signalling
    "RET", "NF1", "HRAS", "KRAS", "NRAS",
    "TMEM127", "MAX", "KIF1B", "FGFR1",
    # Cluster 3 — Wnt / cortical
    "MAML3", "CSDE1", "UBTF",
    # Malignancy / progression
    "ATRX", "TERT", "TP53", "CDKN2A",
    "RB1", "SETD2", "PTEN", "MEN1"
  ),
  entrez_id = c(
    6389, 6390, 6391, 6392, 54949,
    7428, 2271, 4191, 2034, 3417, 3418,
    8402, 1743, 2806,
    5979, 4763, 3265, 3845, 4893,
    55654, 4149, 23095, 2260,
    55502, 7812, 7343,
    546, 7015, 7157, 1029,
    5925, 29072, 5728, 4221
  ),
  cluster = c(
    rep("Pseudohypoxia", 14),
    rep("Kinase_RAS", 9),
    rep("Wnt_cortical", 3),
    rep("Malignancy", 8)
  ),
  stringsAsFactors = FALSE
)
```

```{r}
ggVennDiagram::ggVennDiagram(list(ppgl = as.character(ppgl_drivers$entrez_id), dtu = dtu_gene %>% filter(FDR < 0.05) %>% dplyr::distinct(gene_id) %>% .$gene_id))
```

```{r}
ppgl_drivers %>% filter(symbol %in% gene_types$SYMBOL)
ppgl_drivers %>% filter(symbol %in% gene_types_prom$SYMBOL)
```

