---
title: "emseqAnalysis"
author: "caitlinpage"
date: "2026-07-27"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

## Introduction

```{r}
library(BSgenome.Hsapiens.UCSC.hg38)
library(data.table)
library(plyranges)
library(tidyr)
library(stringr)
library(dplyr)
library(ggplot2)
library(OpTiles)
library(edgeR)
library(UpSetR)
```

```{r}
# cpgs
seqs <- seqnames(BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38)[1:22]

cg_sites <- lapply(seqs, function(x) {
  cbind(data.frame(Biostrings::matchPattern("CG", BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38[[x]])),
        seqnames = x
  )}) %>% bind_rows()
```


```{r}
healthy_samples <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_all_healthy.rds")
healthy_samples <- healthy_samples %>% filter(seqnames %in% seqs)
healthy_samples <- distinct(healthy_samples)

optiles_all <- readRDS("/researchers/caitlin.page/cf_nano/r_output/optiles_all_emseq.rds") # built from healthy

cancer_samples <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.rds")

healthy_samples$type <- "healthy"
cancer_samples$type <- "cancer"
```

```{r}
dtu_tiles <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_emseq_cancer_healthy.rds")
dtu_tiles_healthy_compare <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_emseq_healthy.rds")
```


# Overview of samples

```{r}
healthy_samples %>%
  ggplot(aes(x = total)) +
  geom_density()

healthy_samples %>%
  ggplot(aes(x = alpha)) +
  geom_density()
```

```{r}
cancer_samples %>%
  ggplot(aes(x = total)) +
  geom_density()

cancer_samples %>%
  ggplot(aes(x = alpha)) +
  geom_density()
```


# Analyse the healthy x healthy results


```{r}
dtu_tiles_healthy_compare %>% mutate(false_pos = ifelse(FDR < 0.05, TRUE, FALSE), tile = "tile") %>%
  ggplot(aes(x = tile, fill = false_pos)) +
  geom_bar(position = "fill")
```

```{r}
dtu_tiles_healthy_compare %>% mutate(false_pos = ifelse(FDR < 0.05, TRUE, FALSE), tile = "tile") %>%
  ggplot(aes(x = tile, fill = false_pos)) +
  geom_bar() +
  coord_polar("y", start = 0)
```

```{r}
dtu_tiles_healthy_compare %>% mutate(false_pos = ifelse(FDR < 0.05, TRUE, FALSE), tile = "tile") %>%
  group_by(false_pos) %>%
  summarise(n = n()) %>%
  ungroup() %>%
  mutate(ratio = n/sum(n))
```

```{r}
dtu_tiles_healthy_compare %>%
  ggplot(aes(x = FDR)) +
  geom_density()
```

```{r}
dtu_tiles_healthy_compare %>% filter(FDR < 0.05) %>% nrow()
```

11 false positives
I'll take it

* visualise those regions
```{r}
healthy_samples$read_start <- healthy_samples$start
healthy_samples$read_end <- healthy_samples$end
```


* cool cool got a strand thing going on with these reads - same read_id, oppo strand
```{r}
rrbs_plot5 <- function(read_data, read_ids, cpgs = cg_sites, jit_amount = 0.1) {
  set.seed(42) # For reproducibility
  # Calculate a single jitter amount per row (segment)
  read_data <- read_data %>% dplyr::filter(read_id %in% read_ids)
  jitter_amount <- runif(nrow(read_data), -jit_amount, jit_amount)
  read_data$alpha_jit <- read_data$alpha + jitter_amount
  read_data <- data.frame(read_data)

  # Add jittered coordinates to new columns in the data frame
  long_data <- read_data %>%
    mutate(meth_status = strsplit(meth_pattern, ""), start = strsplit(genom_positions, ",")) %>%
    unnest(meth_status, start) %>%
    mutate(start = as.double(start))
  
  plot <- read_data %>%
    ggplot() +
    geom_segment(aes(x = read_start, xend = read_end, y = alpha_jit, yend = alpha_jit), alpha = 0.2) +
    geom_point(data = long_data,
               aes(x = start, y = alpha_jit, fill = meth_status), shape = 21, color = "black") +
    scale_fill_manual(values = c("U" = "white", "M" = "black")) +
    facet_wrap(~ sample)
  plot
}
```

```{r}
reads <- find_overlaps(dtu_tiles_healthy_compare %>% filter(FDR < 0.05) %>% .[3,] %>% as_granges(), as_granges(healthy_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(healthy_samples, reads)

reads <- find_overlaps(dtu_tiles_healthy_compare %>% filter(FDR < 0.05) %>% .[4,] %>% as_granges(), as_granges(healthy_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(healthy_samples, reads)

reads <- find_overlaps(dtu_tiles_healthy_compare %>% filter(FDR < 0.05) %>% .[5,] %>% as_granges(), as_granges(healthy_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(healthy_samples, reads)

reads <- find_overlaps(dtu_tiles_healthy_compare %>% filter(FDR < 0.05) %>% .[6,] %>% as_granges(), as_granges(healthy_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(healthy_samples, reads)

reads <- find_overlaps(dtu_tiles_healthy_compare %>% filter(FDR < 0.05) %>% .[7,] %>% as_granges(), as_granges(healthy_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(healthy_samples, reads)

reads <- find_overlaps(dtu_tiles_healthy_compare %>% filter(FDR < 0.05) %>% .[8,] %>% as_granges(), as_granges(healthy_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(healthy_samples, reads)

reads <- find_overlaps(dtu_tiles_healthy_compare %>% filter(FDR < 0.05) %>% .[9,] %>% as_granges(), as_granges(healthy_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(healthy_samples, reads)

reads <- find_overlaps(dtu_tiles_healthy_compare %>% filter(FDR < 0.05) %>% .[10,] %>% as_granges(), as_granges(healthy_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(healthy_samples, reads)

reads <- find_overlaps(dtu_tiles_healthy_compare %>% filter(FDR < 0.05) %>% .[11,] %>% as_granges(), as_granges(healthy_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(healthy_samples, reads)
```

* ok so there is a change in proportion for some of these
* but also it's only 11 regions


* check some of the non-significant ones
```{r}
reads <- find_overlaps(dtu_tiles_healthy_compare %>% filter(FDR > 0.05) %>% .[3,] %>% as_granges(), as_granges(healthy_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(healthy_samples, reads)

reads <- find_overlaps(dtu_tiles_healthy_compare %>% filter(FDR > 0.05) %>% .[5,] %>% as_granges(), as_granges(healthy_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(healthy_samples, reads)

reads <- find_overlaps(dtu_tiles_healthy_compare %>% filter(FDR > 0.05) %>% .[10,] %>% as_granges(), as_granges(healthy_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(healthy_samples, reads)

reads <- find_overlaps(dtu_tiles_healthy_compare %>% filter(FDR > 0.05) %>% .[20,] %>% as_granges(), as_granges(healthy_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(healthy_samples, reads)

reads <- find_overlaps(dtu_tiles_healthy_compare %>% filter(FDR > 0.05) %>% .[100,] %>% as_granges(), as_granges(healthy_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(healthy_samples, reads)
```


# Analyse the cancer x healthy results

```{r}
dtu_tiles %>% filter(FDR < 0.05) %>% nrow()
```


```{r}
dtu_tiles %>% mutate(signif = ifelse(FDR < 0.05, TRUE, FALSE), tile = "tile") %>%
  ggplot(aes(x = tile, fill = signif)) +
  geom_bar(position = "fill")
```

```{r}
all_samples <- rbind(healthy_samples, cancer_samples) %>% data.frame()
```


```{r}
reads <- find_overlaps(dtu_tiles %>% filter(FDR < 0.05) %>% .[3,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(dtu_tiles %>% filter(FDR < 0.05) %>% .[4,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(dtu_tiles %>% filter(FDR < 0.05) %>% .[5,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(dtu_tiles %>% filter(FDR < 0.05) %>% .[6,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(dtu_tiles %>% filter(FDR < 0.05) %>% .[7,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(dtu_tiles %>% filter(FDR < 0.05) %>% .[8,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(dtu_tiles %>% filter(FDR < 0.05) %>% .[9,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(dtu_tiles %>% filter(FDR < 0.05) %>% .[10,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(dtu_tiles %>% filter(FDR < 0.05) %>% .[11,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)
```

* non significant tiles
```{r}
reads <- find_overlaps(dtu_tiles %>% filter(FDR > 0.05) %>% .[10,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)

reads <- find_overlaps(dtu_tiles %>% filter(FDR > 0.05) %>% .[110,] %>% as_granges(), as_granges(all_samples)) %>% data.frame() %>% .$read_id
rrbs_plot5(all_samples, reads)
```

