---
title: "epiaIdeas"
author: "caitlinpage"
date: "2026-09-18"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

## Introduction
"/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet",
```{r}
library(BSgenome.Hsapiens.UCSC.hg38)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(plyranges)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(data.table)
library(duckdb)
library(arrow)
```

```{r}
rrbs_plot5 <- function(read_data_path_a, read_data_path_b, chrom, region_start, region_end, cpgs = cg_sites,
                       jit_amount = 0.1) {
  set.seed(42)

  # ── 1. Read only the matching rows — never loads unmatched data into RAM ───
  con <- dbConnect(duckdb())
  on.exit(dbDisconnect(con))
  
  read_data_a <- dbGetQuery(con, glue::glue("
    SELECT *
    FROM read_parquet('{read_data_path_a}')
    WHERE seqnames = '{chrom}'
      AND \"start\" >= {region_start}
      AND \"end\"   <= {region_end}
  "))
  
  read_data_b <- dbGetQuery(con, glue::glue("
    SELECT *
    FROM read_parquet('{read_data_path_b}')
    WHERE seqnames = '{chrom}'
      AND \"start\" >= {region_start}
      AND \"end\"   <= {region_end}
  "))
  
  read_data_a$type <- "healthy"
  read_data_b$type <- "cancer"
  
  read_data <- rbind(read_data_a, read_data_b)
  
  jitter_amount <- runif(nrow(read_data), -jit_amount, jit_amount)
  read_data$alpha_jit <- read_data$alpha + jitter_amount
  read_data <- data.frame(read_data)
  
  read_data <- read_data %>% mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                 alpha > 0.8 ~ "a > 0.8",
                                 alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                 alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                 TRUE ~ "0.6 <= a <= 0.8"))
  read_data <- read_data %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8")))
  
  # Add jittered coordinates to new columns in the data frame
  long_data <- read_data %>%
    mutate(meth_status = strsplit(meth_pattern, ""), start = strsplit(genom_positions, ",")) %>%
    unnest(meth_status, start) %>%
    mutate(start = as.double(start))

  chr <- unique(read_data$seqnames)
  min_start <- min(read_data$read_start)
  max_end <- max(read_data$read_end)
  
  plot <- read_data %>%
    ggplot() +
    geom_segment(aes(x = read_start, xend = read_end,
                     y = alpha_jit, yend = alpha_jit, colour = alpha_group)) +
    geom_point(data = long_data,
               aes(x = start, y = alpha_jit,
                   fill = meth_status, colour = alpha_group), shape = 21) +
    scale_fill_manual(values = c("U" = "white", "M" = "black")) +
    facet_wrap(~ type, nrow=2) +
    labs(title = paste0(chr, ":", min_start, "-", max_end)) +
    theme_bw() +
    labs(x = "genomic position")
  plot
}
```



# what does X chromosome alpha distribution look like?
* someone thought it might look different because of X inactivation
```{r}
cg_sites <- cbind(data.frame(Biostrings::matchPattern("CG", BSgenome.Hsapiens.UCSC.hg38[["chr22"]])), seqnames = "chr22") %>% 
  mutate(pos = paste0(seqnames, "-", start)) %>%
  relocate(pos, seqnames)
cg_sites$index <- 1:nrow(cg_sites)
c(number_of_sites = nrow(cg_sites))
head(cg_sites)
```

```{r}
betas_c %>% distinct(seqnames)
```

"/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet",
```{r}
con <- dbConnect(duckdb())
read_healthy <- dbGetQuery(con, glue::glue("
    SELECT *
    FROM read_parquet('/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet')
    WHERE seqnames IN  (
    'chr1','chr2','chr3','chr4','chr5','chr6','chr7','chr8','chr9','chr10',
    'chr11','chr12','chr13','chr14','chr15','chr16','chr17','chr18','chr19',
    'chr20','chr21','chr22','chrX','chrY'
  )
  "))
```
```{r}
read_healthy %>%
  filter(total > 2) %>%
  ggplot(aes(x = alpha)) +
  geom_density() +
  facet_wrap(~seqnames)
```

```{r}
read_healthy[1:1000,] %>%
  filter(total > 2) %>%
  ggplot(aes(x = alpha)) +
  geom_density() +
  facet_wrap(~seqnames)
```


  on.exit(dbDisconnect(con))
```{r}
read_cancer <- dbGetQuery(con, glue::glue("
    SELECT *
    FROM read_parquet('/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet')
    WHERE seqnames IN  (
    'chr1','chr2','chr3','chr4','chr5','chr6','chr7','chr8','chr9','chr10',
    'chr11','chr12','chr13','chr14','chr15','chr16','chr17','chr18','chr19',
    'chr20','chr21','chr22','chrX','chrY'
  )
  "))
```


```{r}
read_cancer %>%
  filter(total > 2) %>%
  ggplot(aes(x = alpha)) +
  geom_density() +
  facet_wrap(~seqnames)
```

* oh well don't have the x chromosome here to actually check
* but it doesn't happen in the healthy, it all looks about the same

```{r}
read_healthy[1:1000,] %>%
  filter(total > 2) %>%
  
```


```{r}
read_healthy$type <- "healthy"
read_cancer$type <- "cancer"

read_data <- rbind(read_data_a, read_data_b)
```

########

# new

* other questions I thought of up there

** what's the minimum coverage I need to trust my results?
* calculate coverage in each tile, add that to the delta
* and look at the tiles
* this is going to be harder to determine
* adding coverage one thing - doing something with it another


* do I have dau when betas are about the same for conditions?
** basically the tiles that don't overlap - what do the betas look like?
** could do an average beta and sd per tile
** then that's the one stat - ooh I like that
* this should be simpler - or at least more direct

```{r}
setDT(dau_basictiles)
dau_basictiles <- dau_basictiles[, .N, by = names(dau_basictiles)]
```

```{r}
find_overlaps(as_granges(cg_sites), dau_basictiles %>% filter(region_id == "chr1:154602501-154603000") %>% .[1,] %>% as_granges()) %>% data.frame()
```

* yep correct
* also oops that means I need to filter the tiles based on the number of CpGs they cover
* because getting high significance in ones with not many cpgs

```{r}
dau_basictiles %>%
  ggplot(aes(x = N, y = dtu_rank)) +
  geom_point(alpha = 0.3)
```

```{r}
betas_c %>% filter(meth_count_pos < 3 | unmeth_count_pos < 3)
betas_c %>% filter(meth_count_pos < 3 , unmeth_count_pos < 3)
```
```{r}
betas_c %>% mutate(coverage = meth_count_pos + unmeth_count_pos) %>% filter(coverage < 10)
```

```{r}
dau_basictiles %>%
  ggplot(aes(x = log2(N))) +
  geom_histogram(bins = 100) +
  labs(x = "Number CpGs on tile")
```


```{r}
dau_basictiles
```

```{r}
tile_beta <- find_overlaps(as_granges(dau_basictiles), betas_c %>% mutate(coverage = meth_count_pos + unmeth_count_pos) %>% filter(coverage > 10) %>% group_by(seqnames, start) %>% summarise(beta_cancer = mean(beta)) %>% mutate(width = 1) %>% as_granges()) %>% data.frame()
```
```{r}
tile_beta
```

dau_basictiles <- dau_basictiles[, .N, by = names(dau_basictiles)]
```{r}
setDT(tile_beta)
tile_beta <- tile_beta[, .(mean_beta_cancer = mean(beta_cancer)), by = eval(names(tile_beta)[1:12])]
```

```{r}
tile_beta <- find_overlaps(as_granges(tile_beta), betas_h %>% mutate(coverage = meth_count_pos + unmeth_count_pos) %>% filter(coverage > 10) %>% group_by(seqnames, start) %>% summarise(beta_healthy = mean(beta)) %>% mutate(width = 1) %>% as_granges()) %>% data.frame()

setDT(tile_beta)
tile_beta <- tile_beta[, .(mean_beta_healthy = mean(beta_healthy)), by = eval(names(tile_beta)[1:13])]
```
* how to see if the beta value for the tile is similar?
* difference between them? - yeah, then that's delta
* cancer - normal

```{r}
tile_beta <- tile_beta %>% mutate(delta_beta = mean_beta_cancer - mean_beta_healthy)
```

* I am hoping to see significance in a small delta situation
* and for that tile to look cool
* let's check top 100 tiles: order by small delta
```{r}
tile_beta[1:100,] %>% .[order(abs(.$delta_beta)),]
```

* ok some promising stuff
* 78 that are abs(delta) < 0.2

* visualise ranked 12 (has 52 cpgs) 	chr2:232419501-232420000
```{r}
betas_c$type <- "cancer"
betas_h$type <- "healthy"
betas_all <- rbind(betas_c, betas_h)
```

```{r}
rrbs_plot5("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", "chr2", 232419500, 232420001) +
  geom_beta(betas_all, plot.height = 0.5, style = "smooth") 
```
* not sure how convincing this is
* need the proportions
```{r}
props_from_test %>% filter(dtu_rank == 12)
```

```{r}
props_from_test %>% filter(dtu_rank == 12) %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))) %>% pivot_longer(cols = c(cancer, healthy), names_to = "type", values_to = "percent") %>%
  ggplot(aes(x = alpha_group, y = percent, fill = type)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("cancer" = "#C0622A", "healthy" = "#1B3A6B")) +
  scale_x_discrete(labels = c("Low", "Low-med", "Med", "Med-high", "High")) +
  labs(x = "alpha band") +
  facet_wrap(~region_id) +
  theme_bw()
```
* there's a change, but could be more convincing
* repeat with other tiles

```{r}
ranks <- tile_beta[1:100,] %>% .[order(abs(.$delta_beta)),] %>% filter(N > 10, abs(delta_beta) < 0.2) %>% .[order(.$dtu_rank),] %>% .$dtu_rank
```


```{r}
props <- tile_prop("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", 
          dau_basictiles[ranks,])
props <- props[,c(1:14,16:24)]
props <- props %>%
  mutate(
    alpha_group = case_when(
      alpha < 0.2  ~ "a < 0.2",
      alpha > 0.8  ~ "a > 0.8",
      alpha <= 0.4 ~ "0.2 <= a <= 0.4",
      alpha < 0.6  ~ "0.4 < a < 0.6",
      TRUE         ~ "0.6 <= a <= 0.8"
    )
  ) %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))) %>%
  count(type, dtu_rank, region_id, alpha_group, name = "tile_band_count") %>%
  group_by(region_id, dtu_rank, type) %>%
  mutate(prop_alpha_in_tile = tile_band_count / sum(tile_band_count)) %>%
  ungroup() %>%
  select(region_id, dtu_rank, alpha_group, type, prop_alpha_in_tile) %>%
  pivot_wider(names_from = type, values_from = prop_alpha_in_tile) %>%
  mutate(prop_change = cancer - healthy)
```

```{r}
props %>% .[order(abs(.$prop_change), decreasing = TRUE),]
props %>% filter(dtu_rank == 97)
```

```{r}
props %>% filter(abs(prop_change) > 0.15) %>% group_by(region_id, dtu_rank) %>% summarise(n=n()) %>% filter(n > 1)
```
```{r}
props %>% filter(dtu_rank %in% c(65, 76, 41, 88, 21))
```
* these 5 tiles might be slightly stronger
* also wait do any of these overlap dmrs? (would expect no because of my mean beta thing)
```{r}
find_overlaps(as_granges(dau_basictiles[c(21,41,65,76,88)]), as_granges(dmr_emseq)) %>% data.frame()
```
* one overlap

```{r}
# tile 21, chr5:159105001-159105500
rrbs_plot5("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", "chr5", 159105001, 159105500) +
  geom_beta(betas_all, plot.height = 0.5, style = "smooth") +
  geom_dmr(dmr_emseq)

props %>% filter(dtu_rank == 21) %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))) %>% pivot_longer(cols = c(cancer, healthy), names_to = "type", values_to = "percent") %>%
  ggplot(aes(x = alpha_group, y = percent, fill = type)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("cancer" = "#C0622A", "healthy" = "#1B3A6B")) +
  scale_x_discrete(labels = c("Low", "Low-med", "Med", "Med-high", "High")) +
  labs(x = "alpha band") +
  facet_wrap(~region_id) +
  theme_bw()
```


```{r}
# tile 41, chr4:4033001-4033500
rrbs_plot5("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", "chr4", 4033001, 4033500) +
  geom_beta(betas_all, plot.height = 0.5, style = "smooth") 

props %>% filter(dtu_rank == 41) %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))) %>% pivot_longer(cols = c(cancer, healthy), names_to = "type", values_to = "percent") %>%
  ggplot(aes(x = alpha_group, y = percent, fill = type)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("cancer" = "#C0622A", "healthy" = "#1B3A6B")) +
  scale_x_discrete(labels = c("Low", "Low-med", "Med", "Med-high", "High")) +
  labs(x = "alpha band") +
  facet_wrap(~region_id) +
  theme_bw()
```


```{r}
# tile 65, chr18:74458001-74458500
rrbs_plot5("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", "chr18", 74458001, 74458500) +
  geom_beta(betas_all, plot.height = 0.5, style = "smooth") 

props %>% filter(dtu_rank == 65) %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))) %>% pivot_longer(cols = c(cancer, healthy), names_to = "type", values_to = "percent") %>%
  ggplot(aes(x = alpha_group, y = percent, fill = type)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("cancer" = "#C0622A", "healthy" = "#1B3A6B")) +
  scale_x_discrete(labels = c("Low", "Low-med", "Med", "Med-high", "High")) +
  labs(x = "alpha band") +
  facet_wrap(~region_id) +
  theme_bw()
```


```{r}
# tile 76 chr2:237371501-237372000
rrbs_plot5("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", "chr2", 237371501, 237372000) +
  geom_beta(betas_all, plot.height = 0.5, style = "smooth") 

props %>% filter(dtu_rank == 76) %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))) %>% pivot_longer(cols = c(cancer, healthy), names_to = "type", values_to = "percent") %>%
  ggplot(aes(x = alpha_group, y = percent, fill = type)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("cancer" = "#C0622A", "healthy" = "#1B3A6B")) +
  scale_x_discrete(labels = c("Low", "Low-med", "Med", "Med-high", "High")) +
  labs(x = "alpha band") +
  facet_wrap(~region_id) +
  theme_bw()
```


```{r}
# tile 88, chr5:1548501-1549000
rrbs_plot5("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", "chr5", 1548501, 1549000) +
  geom_beta(betas_all, plot.height = 0.5, style = "smooth") 

props %>% filter(dtu_rank == 88) %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))) %>% pivot_longer(cols = c(cancer, healthy), names_to = "type", values_to = "percent") %>%
  ggplot(aes(x = alpha_group, y = percent, fill = type)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("cancer" = "#C0622A", "healthy" = "#1B3A6B")) +
  scale_x_discrete(labels = c("Low", "Low-med", "Med", "Med-high", "High")) +
  labs(x = "alpha band") +
  facet_wrap(~region_id) +
  theme_bw()
```
* none of these are super convincing

```{r}
tile_beta[1:1000,] %>%
  ggplot(aes(x = dtu_rank, y = delta_beta)) +
  geom_point(alpha = 0.2) +
  geom_hline(yintercept = 0)
```
* so this doesn't look like much
*BUT - what about what it's overlap to dmr is like? - because if dmr overlap is in the tiles that have high delta - but not in low delta - I'll take it
* then just back to figuring out if they're interesting
* also can do an average beta across the dmrs (because I don't understand maxdiff/meandiff)

```{r}
dmr_emseq
tile_dmr <- find_overlaps(as_granges(dmr_emseq), as_granges(dau_basictiles)) %>% data.frame()
```

```{r}
tile_beta %>% mutate(overlap_dmr = ifelse(dtu_rank %in% tile_dmr$dtu_rank, TRUE, FALSE)) %>%
  .[1:1000,] %>%
  ggplot(aes(x = dtu_rank, y = delta_beta, colour = overlap_dmr)) +
  geom_point(alpha = 0.2) +
  geom_hline(yintercept = 0)

tile_beta %>% mutate(overlap_dmr = ifelse(dtu_rank %in% tile_dmr$dtu_rank, TRUE, FALSE)) %>%
  ggplot(aes(x = dtu_rank, y = delta_beta, colour = overlap_dmr)) +
  geom_point(alpha = 0.2) +
  geom_hline(yintercept = 0)
```
```{r}
tile_beta %>% mutate(overlap_dmr = ifelse(dtu_rank %in% tile_dmr$dtu_rank, TRUE, FALSE)) %>%
  ggplot(aes(x = dtu_rank, y = delta_beta, colour = overlap_dmr)) +
  geom_point(alpha = 0.2) +
  geom_hline(yintercept = 0) +
  facet_wrap(~overlap_dmr)
```
* so that doesn't tell me much
* but I guess it's interesting that there's regions of big delta change that don't overlap a dmr

```{r}
tile_beta <- tile_beta %>% mutate(overlap_dmr = ifelse(dtu_rank %in% tile_dmr$dtu_rank, TRUE, FALSE))
```

```{r}
tile_beta %>%
  ggplot(aes(x = delta_beta, colour = overlap_dmr)) +
  geom_density()

tile_beta %>%
  ggplot(aes(x = overlap_dmr, y = delta_beta, fill = overlap_dmr)) +
  geom_boxplot()
```

```{r}
tile_beta %>% filter(FDR < 0.05) %>%
  ggplot(aes(x = delta_beta, colour = overlap_dmr)) +
  geom_density()

tile_beta %>% filter(FDR < 0.05) %>%
  ggplot(aes(x = overlap_dmr, y = delta_beta, fill = overlap_dmr)) +
  geom_boxplot()
```
* this distribution is interesting...
* has a little swapsies to it
```{r}
tile_beta %>% filter(FDR < 0.05) %>%
  ggplot(aes(x = abs(delta_beta), colour = overlap_dmr)) +
  geom_density()
```
* doing abs makes them samesiesish
* the little bumps are in the dmr/ tiles - oh wait those are the outliers (looked at the boxplot)
* ok that means the dau is pulling out kinda like the opposite stuff
* i did cancer - healthy, meaning +ve delta cancer hypermeth, -ve delta cancer hypometh (more meth in healthy)

* so the red (which is the novel ones) are more -ve, higher meth in healthy for beta


# look at dau when beta mid range in conditions

## both



```{r}
dau_basictiles
```


# medium beta analysis

```{r}
ranks <- tile_beta %>% filter(abs(mean_beta_cancer) >= 0.4, abs(mean_beta_cancer) <= 0.6,
                     abs(mean_beta_healthy) >= 0.4, abs(mean_beta_healthy) <= 0.6) %>%
  filter(FDR < 0.05) %>% filter(N > 10) %>% .$dtu_rank
```
```{r}
tile_beta %>% filter(abs(mean_beta_cancer) >= 0.4, abs(mean_beta_cancer) <= 0.6,
                     abs(mean_beta_healthy) >= 0.4, abs(mean_beta_healthy) <= 0.6) %>%
  filter(FDR < 0.05) %>% filter(N > 10) %>%
  ggplot(aes(x = delta_beta)) +
  geom_density()
```

```{r}
props_med <- tile_prop("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", 
          dau_basictiles[ranks,])
props_med <- props_med[,c(1:14,16:24)]
props_med <- props_med %>%
  mutate(
    alpha_group = case_when(
      alpha < 0.2  ~ "a < 0.2",
      alpha > 0.8  ~ "a > 0.8",
      alpha <= 0.4 ~ "0.2 <= a <= 0.4",
      alpha < 0.6  ~ "0.4 < a < 0.6",
      TRUE         ~ "0.6 <= a <= 0.8"
    )
  ) %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))) %>%
  count(type, dtu_rank, region_id, alpha_group, name = "tile_band_count") %>%
  group_by(region_id, dtu_rank, type) %>%
  mutate(prop_alpha_in_tile = tile_band_count / sum(tile_band_count)) %>%
  ungroup() %>%
  select(region_id, dtu_rank, alpha_group, type, prop_alpha_in_tile) %>%
  pivot_wider(names_from = type, values_from = prop_alpha_in_tile) %>%
  mutate(prop_change = cancer - healthy)
```

* plot all the dsitributions first
```{r}
ranks[2]
```

```{r}
for (i in 1:length(ranks)) {
  p <- props_med %>% filter(dtu_rank == ranks[i]) %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))) %>%
    pivot_longer(cols = c(cancer, healthy), names_to = "type", values_to = "percent") %>%
  ggplot(aes(x = alpha_group, y = percent, fill = type)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("cancer" = "#C0622A", "healthy" = "#1B3A6B")) +
  scale_x_discrete(labels = c("Low", "Low-med", "Med", "Med-high", "High")) +
  labs(x = "alpha band") +
  facet_wrap(~region_id) +
  theme_bw()
  print(p)
}
```

```{r}
ranks[13] # this one has a cute distribution
tile_beta %>% filter(dtu_rank == ranks[13])


rrbs_plot5("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", "chr2", 94871501, 94872000) +
  geom_beta(betas_all, plot.height = 0.5, style = "smooth") 

props_med %>% filter(dtu_rank == ranks[13]) %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))) %>% pivot_longer(cols = c(cancer, healthy), names_to = "type", values_to = "percent") %>%
  ggplot(aes(x = alpha_group, y = percent, fill = type)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("cancer" = "#C0622A", "healthy" = "#1B3A6B")) +
  scale_x_discrete(labels = c("Low", "Low-med", "Med", "Med-high", "High")) +
  labs(x = "alpha band") +
  facet_wrap(~region_id) +
  theme_bw()
```
* I'm not sure if the ~15% change is enough to be of interest

```{r}
find_overlaps(genes_txdb, tile_beta %>% filter(dtu_rank == ranks[13]) %>% as_granges()) %>% data.frame()
```
* ooh overlaps 2 gene - a big gene
* but it's overlapping in the middle, not as like a promoter in front of the gene
```{r}
find_overlaps(genes_txdb, as_granges(dmr_emseq)) %>% data.frame() %>% filter(gene_id %in% c("150483", "442028"))
```
* new genes (claude is providing info)
* claude: 150483 = TEKT4
** it is connected to PPGL - if hypermethylated, may be relevant
** except it's hypomethylated - and in the middle of the gene
** unless of course I have a significant tile in front of the gene?
** no I don't
```{r}
tile_beta %>% filter(seqnames == "chr2", start >= 94870430) %>% .[order(.$start),] %>% .[1:5,]
```

# gene promoters instead of gene body

```{r}
promoters_txdb <- promoters(genes_txdb, upstream = 1000, downstream = 500)
```

```{r}
find_overlaps(promoters_txdb, tile_beta %>% filter(dtu_rank == ranks[13]) %>% as_granges()) %>% data.frame()
find_overlaps(promoters_txdb, tile_beta %>% filter(dtu_rank %in% ranks) %>% as_granges()) %>% data.frame() %>% distinct()
```
* hey 150483 again
* maybe because I included some downstream in it

* plot just these 5 tiles
```{r}
ranks_ <- find_overlaps(promoters_txdb, tile_beta %>% filter(dtu_rank %in% ranks) %>% as_granges()) %>% data.frame() %>% .$dtu_rank
```

```{r}
#chr19:30668501-30669000
rrbs_plot5("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", "chr19", 30668501, 30669000) +
  geom_beta(betas_all, plot.height = 0.5, style = "smooth") 

props_med %>% filter(dtu_rank == ranks_[1]) %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))) %>% pivot_longer(cols = c(cancer, healthy), names_to = "type", values_to = "percent") %>%
  ggplot(aes(x = alpha_group, y = percent, fill = type)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("cancer" = "#C0622A", "healthy" = "#1B3A6B")) +
  scale_x_discrete(labels = c("Low", "Low-med", "Med", "Med-high", "High")) +
  labs(x = "alpha band") +
  facet_wrap(~region_id) +
  theme_bw()
```

```{r}
#chr2:94871501-94872000
rrbs_plot5("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", "chr2", 94871501, 94872000) +
  geom_beta(betas_all, plot.height = 0.5, style = "smooth") 

props_med %>% filter(dtu_rank == ranks_[2]) %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))) %>% pivot_longer(cols = c(cancer, healthy), names_to = "type", values_to = "percent") %>%
  ggplot(aes(x = alpha_group, y = percent, fill = type)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("cancer" = "#C0622A", "healthy" = "#1B3A6B")) +
  scale_x_discrete(labels = c("Low", "Low-med", "Med", "Med-high", "High")) +
  labs(x = "alpha band") +
  facet_wrap(~region_id) +
  theme_bw()
```

```{r}
#chr16:1226501-1227000
rrbs_plot5("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", "chr16", 1226501, 1227000) +
  geom_beta(betas_all, plot.height = 0.5, style = "smooth") 

props_med %>% filter(dtu_rank == ranks_[3]) %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))) %>% pivot_longer(cols = c(cancer, healthy), names_to = "type", values_to = "percent") %>%
  ggplot(aes(x = alpha_group, y = percent, fill = type)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("cancer" = "#C0622A", "healthy" = "#1B3A6B")) +
  scale_x_discrete(labels = c("Low", "Low-med", "Med", "Med-high", "High")) +
  labs(x = "alpha band") +
  facet_wrap(~region_id) +
  theme_bw()
```

```{r}
#chr5:1201001-1201500
rrbs_plot5("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", "chr5", 1201001, 1201500) +
  geom_beta(betas_all, plot.height = 0.5, style = "smooth") 

props_med %>% filter(dtu_rank == ranks_[4]) %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))) %>% pivot_longer(cols = c(cancer, healthy), names_to = "type", values_to = "percent") %>%
  ggplot(aes(x = alpha_group, y = percent, fill = type)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("cancer" = "#C0622A", "healthy" = "#1B3A6B")) +
  scale_x_discrete(labels = c("Low", "Low-med", "Med", "Med-high", "High")) +
  labs(x = "alpha band") +
  facet_wrap(~region_id) +
  theme_bw()
```

```{r}
#chr16:3012001-3012500
rrbs_plot5("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", "chr16", 3012001, 3012500) +
  geom_beta(betas_all, plot.height = 0.5, style = "smooth") 

props_med %>% filter(dtu_rank == ranks_[5]) %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))) %>% pivot_longer(cols = c(cancer, healthy), names_to = "type", values_to = "percent") %>%
  ggplot(aes(x = alpha_group, y = percent, fill = type)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("cancer" = "#C0622A", "healthy" = "#1B3A6B")) +
  scale_x_discrete(labels = c("Low", "Low-med", "Med", "Med-high", "High")) +
  labs(x = "alpha band") +
  facet_wrap(~region_id) +
  theme_bw()
```



```{r}
for (i in 1:length(ranks)) {
  p <- props_med %>% filter(dtu_rank == ranks[i]) %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))) %>%
    pivot_longer(cols = c(cancer, healthy), names_to = "type", values_to = "percent") %>%
  ggplot(aes(x = alpha_group, y = percent, fill = type)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("cancer" = "#C0622A", "healthy" = "#1B3A6B")) +
  scale_x_discrete(labels = c("Low", "Low-med", "Med", "Med-high", "High")) +
  labs(x = "alpha band") +
  facet_wrap(~region_id) +
  theme_bw()
  print(p)
}
```



```{r}
unique(ranks)
```


# gene bodies should be hypometh in cancer

```{r}
tile_gene <- find_overlaps(genes_txdb, tile_beta %>% as_granges()) %>% data.frame()
```
* significant tiles in gene body would be good - as long as it's hypometh to cancer
* problem though - gene locations and tile positions are not going to line up neatly
* but that's ok - I don't need to do the dau analysis to just get the count proportions

```{r}
rrbs_plot5("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", "chr19", 58345178, 58362751) +
  geom_beta(betas_all, plot.height = 0.5, style = "smooth") 
```
* beta bit funky looking there


* get counts for alpha in gene bodies and for promoters (1000bp upstream)
* and then it's just making sure reads overlap genes

* but then how am I visualising this? - because I'm not about to look at 20k bar plots
* all I can think of is the like per base plot where it gets closer to TSS
* but I don't want a position based thing that's too beta
* a density/boxplot of all the counts across genes for the alpha band
* and facet for cancer/normal
* maybe
```{r}
genes_txdb
```

```{r}
reads_region <- function(read_data_path_a, read_data_path_b, genes, cpgs = cg_sites,
                       jit_amount = 0.1) {
  set.seed(42)

  # ── 1. Read only the matching rows — never loads unmatched data into RAM ───
  con <- dbConnect(duckdb())
  on.exit(dbDisconnect(con))
  
  genes_df <- as.data.frame(genes)[, c("seqnames", "start", "end", "gene_id")]
  genes_df$seqnames <- as.character(genes_df$seqnames)
  duckdb_register(con, "genes", genes_df)
  
  overlap_query <- function(path) {
    dbGetQuery(con, glue::glue("
      SELECT DISTINCT r.*, g.gene_id
      FROM read_parquet('{path}') r
      JOIN genes g
        ON r.seqnames = g.seqnames
       AND r.start >= g.start
       AND r.end   <= g.end
    "))
  }
  
  read_data_a <- overlap_query(read_data_path_a)
  
  read_data_b <- overlap_query(read_data_path_b)
  
  read_data_a$type <- "healthy"
  read_data_b$type <- "cancer"
  
  read_data <- rbind(read_data_a, read_data_b)
  
  jitter_amount <- runif(nrow(read_data), -jit_amount, jit_amount)
  read_data$alpha_jit <- read_data$alpha + jitter_amount
  read_data <- data.frame(read_data)
  
  read_data <- read_data %>% mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                 alpha > 0.8 ~ "a > 0.8",
                                 alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                 alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                 TRUE ~ "0.6 <= a <= 0.8"))
  read_data <- read_data %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))) %>%
    count(type, gene_id, alpha_group, name = "tile_band_count") %>% 
  group_by(gene_id, type) %>% mutate(percent = tile_band_count/sum(tile_band_count)) %>% ungroup()
  read_data
}
reads_genebody <- reads_region("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", genes_txdb)
```

```{r}
saveRDS(reads_genebody, "/researchers/caitlin.page/cf_nano/r_output/emseq_readcounts_genebody.rds")
```

```{r}
reads_genebody <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_readcounts_genebody.rds")
```


```{r}
reads_genebody %>%
  ggplot(aes(x = percent, colour = alpha_group)) +
  geom_density() +
  facet_wrap(~type)
```

```{r}
reads_genebody %>%
  ggplot(aes(x = alpha_group, y=percent, fill = type)) +
  geom_boxplot()
```
* technically the low alpha does have higher average in cancer than healthy - so could call that the hypo
* maybe this is something that is cleaner with betas

```{r}
betas_all %>% mutate(coverage = meth_count_pos + unmeth_count_pos) %>% filter(coverage > 10) %>%
  select(sample, seqnames, start, beta) %>%
  pivot_wider(
    id_cols = c(seqnames, start),
    names_from = sample,
    values_from = beta
  )
```

```{r}
betas_all %>% filter(sample == "cancer_0454", seqnames == "chr1", start == 13528)
```
```{r}
genes_betas <- find_overlaps(genes_txdb, betas_all %>% mutate(coverage = meth_count_pos + unmeth_count_pos) %>% filter(coverage > 10) %>%
  select(sample, seqnames, start, beta) %>%
  pivot_wider(
    id_cols = c(seqnames, start),
    names_from = sample,
    values_from = beta
  ) %>% mutate(width = 1) %>% as_granges()) %>% data.frame()
```

```{r}
setDT(genes_betas)

sample_cols <- grep("^cancer_|^healthy_", names(genes_betas), value = TRUE)

genes_betas2 <- melt(
    genes_betas,
    id.vars = "gene_id",          # only keep what you need downstream
    measure.vars = sample_cols,
    variable.name = "sample",
    value.name = "beta",
    na.rm = TRUE                  # drops NAs during melt, not after
  )[,
    type := sub("_[0-9]+$", "", sample)
  ][,
    alpha_group := cut(beta,
      breaks = c(-Inf, 0.2, 0.4, 0.6, 0.8, Inf),
      labels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"),
      right = FALSE)
  ][,
    .(tile_band_count = .N), by = .(type, gene_id, alpha_group)
  ]

genes_betas2 <- genes_betas2 %>% group_by(type, gene_id) %>% mutate(percent = tile_band_count/sum(tile_band_count)) %>% ungroup() %>% 
  mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8")))
```

```{r}
genes_betas2 %>%
  ggplot(aes(x = alpha_group, y=percent, fill = type)) +
  geom_boxplot() +
  labs(x = "beta_group")
```
* oops mislabelled
* also huh looks different
* but this does say hypomethylation in cancer and normal - whereas alpha was leaning hypermeth
```{r}
reads_genebody
```


```{r}
betas_all
```



```{r}
reads_genebody %>%
  ggplot(aes(x = alpha_group, y=percent, fill = type)) +
  geom_violin()
```

# promoters and tiles

```{r}
promoters_tiles <- find_overlaps(promoters_txdb, tile_beta %>% as_granges()) %>% data.frame()
```

```{r}
tile_beta %>% mutate(overlap_promoter = ifelse(region_id %in% promoters_tiles$region_id, TRUE, FALSE)) %>%
  ggplot(aes(x = overlap_promoter, y = delta_beta)) +
  geom_boxplot()

tile_beta %>% mutate(overlap_promoter = ifelse(region_id %in% promoters_tiles$region_id, TRUE, FALSE)) %>%
  ggplot(aes(x = delta_beta, colour = overlap_promoter)) +
  geom_density()
```
```{r}
tile_beta %>% mutate(overlap_promoter = ifelse(region_id %in% promoters_tiles$region_id, TRUE, FALSE), tile_sig = ifelse(FDR < 0.05, TRUE, FALSE)) %>%
  ggplot(aes(x = tile_sig, fill = overlap_promoter)) +
  geom_bar(position = "dodge")

tile_beta %>% mutate(overlap_promoter = ifelse(region_id %in% promoters_tiles$region_id, TRUE, FALSE), tile_sig = ifelse(FDR < 0.05, TRUE, FALSE)) %>% 
  filter(overlap_promoter == TRUE) %>%
  ggplot(aes(x = tile_sig, fill = overlap_promoter)) +
  geom_bar(position = "dodge")
```
```{r}
tile_beta %>% mutate(overlap_promoter = ifelse(region_id %in% promoters_tiles$region_id, TRUE, FALSE), tile_sig = ifelse(FDR < 0.05, TRUE, FALSE)) %>% 
  filter(overlap_promoter == TRUE, FDR < 0.05)
```


```{r}
tile_beta %>%
  ggplot(aes(x = dtu_rank, y = delta_beta)) +
  geom_point(alpha = 0.2)
```


```{r}
tile_prop <- function(read_data_path_a, read_data_path_b, tiles, cols_from_tiles = NULL) {

  # ── 1. Read only the matching rows — never loads unmatched data into RAM ───
  con <- dbConnect(duckdb())
  duckdb::dbWriteTable(con, "tiles_temp", tiles, overwrite = TRUE)
  on.exit(duckdb::dbRemoveTable(con, "tiles_temp"), add = TRUE)
  
  
  # Build SELECT clause for tile columns
  if (is.null(cols_from_tiles)) {
    # Exclude coordinate cols to avoid clashes with parquet
    tile_cols <- setdiff(colnames(tiles), c("seqnames", "start", "end"))
  } else {
    tile_cols <- cols_from_tiles
  }
  
  tile_select <- if (length(tile_cols) > 0) {
    paste(glue::glue(', t."{tile_cols}"'), collapse = "")
  } else ""
  
  read_data_a <- dbGetQuery(con, glue::glue("
    SELECT b.*{tile_select}
    FROM read_parquet('{read_data_path_a}') b
    JOIN tiles_temp t
      ON b.seqnames = t.seqnames
     AND b.start <= t.end
     AND b.end   >= t.start
  "))
  
  read_data_b <- dbGetQuery(con, glue::glue("
    SELECT b.*{tile_select}
    FROM read_parquet('{read_data_path_b}') b
    JOIN tiles_temp t
      ON b.seqnames = t.seqnames
     AND b.start <= t.end
     AND b.end   >= t.start
  "))
  
  read_data_a$type <- "healthy"
  read_data_b$type <- "cancer"
    read_data <- rbind(read_data_a, read_data_b)
  read_data
}
```


```{r}
props <- tile_prop("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", 
          dau_basictiles[1:20,])
```



```{r}
tile_props %>% filter(region_id %in% c("chr1:147077001-147077500", "chr1:147077501-147078000")) %>%
  mutate(
    alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                 alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                 alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                 alpha <= 0.6 & alpha <= 0.8 ~ "0.6 <= a <= 0.8",
                                 TRUE ~ "a > 0.8")
  ) %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))) %>%
  count(type, region_id, alpha_group, name = "tile_band_count") %>% 
  group_by(region_id, type) %>% mutate(percent = tile_band_count/sum(tile_band_count)) %>%
  ggplot(aes(x = alpha_group, y = percent, fill = type)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("cancer" = "#C0622A", "healthy" = "#1B3A6B")) +
  scale_x_discrete(labels = c("Low", "Low-med", "Med", "Med-high", "High")) +
  labs(x = "alpha band") +
  facet_wrap(~region_id) +
  theme_bw()
```


```{r}
props_from_test <- tile_props %>%
  mutate(
    alpha_group = case_when(
      alpha < 0.2  ~ "a < 0.2",
      alpha > 0.8  ~ "a > 0.8",
      alpha <= 0.4 ~ "0.2 <= a <= 0.4",
      alpha < 0.6  ~ "0.4 < a < 0.6",
      TRUE         ~ "0.6 <= a <= 0.8"
    )
  ) %>%
  count(type, dtu_rank, region_id, alpha_group, name = "tile_band_count") %>%
  group_by(region_id, dtu_rank, type) %>%
  mutate(prop_alpha_in_tile = tile_band_count / sum(tile_band_count)) %>%
  ungroup() %>%
  select(region_id, dtu_rank, alpha_group, type, prop_alpha_in_tile) %>%
  pivot_wider(names_from = type, values_from = prop_alpha_in_tile) %>%
  mutate(prop_change = cancer - healthy)
```



# tiles when overlap dmr vs not overlap dmr

```{r}
overlap_tile_dmr <- find_overlaps(as_granges(dmr_emseq), as_granges(filter(tile_beta, FDR < 0.05))) %>% data.frame()
```

```{r}
tile_beta <- tile_beta %>% mutate(overlap_dmr = ifelse(region_id %in% overlap_tile_dmr$region_id, TRUE, FALSE))
```

```{r}
tile_beta <- distinct(tile_beta)
tile_beta %>% filter(FDR < 0.05) %>% filter(overlap_dmr == TRUE)
```

```{r}
range_dmr <- tile_beta %>% filter(FDR < 0.05) %>% filter(overlap_dmr == TRUE)
```

```{r}
reads_region <- function(read_data_path_a, read_data_path_b, genes,
                       jit_amount = 0.1) {
  set.seed(42)

  # ── 1. Read only the matching rows — never loads unmatched data into RAM ───
  con <- dbConnect(duckdb())
  on.exit(dbDisconnect(con))
  
  genes_df <- as.data.frame(genes)[, c("seqnames", "start", "end", "region_id")]
  genes_df$seqnames <- as.character(genes_df$seqnames)
  duckdb_register(con, "genes", genes_df)
  
  overlap_query <- function(path) {
    dbGetQuery(con, glue::glue("
      SELECT DISTINCT r.*, g.region_id
      FROM read_parquet('{path}') r
      JOIN genes g
        ON r.seqnames = g.seqnames
       AND r.start >= g.start
       AND r.end   <= g.end
    "))
  }
  
  read_data_a <- overlap_query(read_data_path_a)
  
  read_data_b <- overlap_query(read_data_path_b)
  
  read_data_a$type <- "healthy"
  read_data_b$type <- "cancer"
  
  read_data <- rbind(read_data_a, read_data_b)
  
  
  read_data <- read_data %>% mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                 alpha > 0.8 ~ "a > 0.8",
                                 alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                 alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                 TRUE ~ "0.6 <= a <= 0.8"))
  read_data <- read_data %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))) %>%
    count(type, region_id, alpha_group, name = "tile_band_count") %>% 
  group_by(region_id, type) %>% mutate(percent = tile_band_count/sum(tile_band_count)) %>% ungroup()
  read_data
}
reads_dmr <- reads_region("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", range_dmr)
```

```{r}
range_dmr_not <- tile_beta %>% filter(FDR < 0.05) %>% filter(overlap_dmr != TRUE)
reads_dmr_not <- reads_region("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", range_dmr_not[1:100,])
```

```{r}
reads_dmr %>% group_by(region_id, alpha_group) %>% mutate(delta_cancer_normal = diff(percent))
```

```{r}
filter(reads_dmr, region_id == "chr10:129968001-129968500")
```

```{r}
reads_dmr %>% complete(region_id, alpha_group, type, fill = list(percent = 0)) %>% 
  group_by(region_id, alpha_group) %>%
  mutate(delta_cancer_normal = percent[type == "cancer"] - percent[type == "healthy"]) %>%
  ggplot(aes(x = percent, colour = alpha_group)) +
  geom_density() +
  facet_wrap(~type)
```

```{r}
reads_dmr_not %>% complete(region_id, alpha_group, type, fill = list(percent = 0)) %>% 
  group_by(region_id, alpha_group) %>%
  mutate(delta_cancer_normal = percent[type == "cancer"] - percent[type == "healthy"]) %>%
  ggplot(aes(x = percent, colour = alpha_group)) +
  geom_density() +
  facet_wrap(~type)
```

```{r}
reads_dmr %>% complete(region_id, alpha_group, type, fill = list(percent = 0)) %>% 
  group_by(region_id, alpha_group) %>%
  summarise(delta_cancer_normal = percent[type == "cancer"] - percent[type == "healthy"]) %>%
  ggplot(aes(x = delta_cancer_normal, colour = alpha_group)) +
  geom_density() 

reads_dmr_not %>% complete(region_id, alpha_group, type, fill = list(percent = 0)) %>% 
  group_by(region_id, alpha_group) %>%
  summarise(delta_cancer_normal = percent[type == "cancer"] - percent[type == "healthy"]) %>%
  ggplot(aes(x = delta_cancer_normal, colour = alpha_group)) +
  geom_density() 
```

```{r}
reads_dmr_not %>% complete(region_id, alpha_group, type, fill = list(percent = 0)) %>% 
  group_by(region_id, alpha_group) %>%
  summarise(delta_cancer_normal = percent[type == "cancer"] - percent[type == "healthy"]) %>% 
  ungroup() %>% group_by(region_id) %>%
  mutate(max_delta_group = alpha_group[which.max(abs(delta_cancer_normal))],
         max_delta_group2 = alpha_group[order(abs(delta_cancer_normal), decreasing = TRUE)[2]]) %>%
  ungroup() %>% distinct(region_id, max_delta_group, max_delta_group2) %>% mutate(combo = paste0("G1:", max_delta_group, ", G2:", max_delta_group2))
```

```{r}
upset_data <- reads_dmr_not %>% complete(region_id, alpha_group, type, fill = list(percent = 0)) %>% 
  group_by(region_id, alpha_group) %>%
  summarise(delta_cancer_normal = percent[type == "cancer"] - percent[type == "healthy"]) %>% 
  ungroup() %>% group_by(region_id) %>%
  mutate(max_delta_group = alpha_group[which.max(abs(delta_cancer_normal))],
         max_delta_group2 = alpha_group[order(abs(delta_cancer_normal), decreasing = TRUE)[2]]) %>%
  ungroup() %>% distinct(region_id, max_delta_group, max_delta_group2) %>% mutate(combo = paste0("G1:", max_delta_group, ", G2:", max_delta_group2)) %>%
  select(region_id, max_delta_group, max_delta_group2) %>%
  pivot_longer(c(max_delta_group, max_delta_group2), values_to = "alpha_group") %>%
  select(region_id, alpha_group) %>%
  mutate(value = 1) %>%
  pivot_wider(names_from = alpha_group, values_from = value, values_fill = 0) %>%
  select(-region_id) %>%
  as.data.frame()

#library(ComplexUpset)

upset(upset_data, intersect = names(upset_data))

upset(
  upset_data,
  intersect = rev(alpha_levels),
  sort_sets = FALSE,
  queries = lapply(alpha_levels, function(grp) {
    upset_query(set = grp, fill = alpha_colors[grp], color = alpha_colors[grp])
  }),
  set_sizes = upset_set_size() + scale_fill_manual(values = alpha_colors)
)
```


```{r}
upset_data <- reads_dmr %>% complete(region_id, alpha_group, type, fill = list(percent = 0)) %>% 
  group_by(region_id, alpha_group) %>%
  summarise(delta_cancer_normal = percent[type == "cancer"] - percent[type == "healthy"]) %>% 
  ungroup() %>% group_by(region_id) %>%
  mutate(max_delta_group = alpha_group[which.max(abs(delta_cancer_normal))],
         max_delta_group2 = alpha_group[order(abs(delta_cancer_normal), decreasing = TRUE)[2]]) %>%
  ungroup() %>% distinct(region_id, max_delta_group, max_delta_group2) %>% mutate(combo = paste0("G1:", max_delta_group, ", G2:", max_delta_group2)) %>%
  select(region_id, max_delta_group, max_delta_group2) %>%
  pivot_longer(c(max_delta_group, max_delta_group2), values_to = "alpha_group") %>%
  select(region_id, alpha_group) %>%
  mutate(value = 1) %>%
  pivot_wider(names_from = alpha_group, values_from = value, values_fill = 0) %>%
  select(-region_id) %>%
  as.data.frame()

#library(ComplexUpset)

upset(upset_data, intersect = names(upset_data))
```

```{r}
alpha_levels <- c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8")

# Reorder columns to match factor levels
upset_data <- upset_data[, alpha_levels]

# Assign a color to each alpha group
alpha_colors <- c(
  "a < 0.2"          = "#F8766D",
  "0.2 <= a <= 0.4"  = "#A3A500",
  "0.4 < a < 0.6"    = "#00BF7D",
  "0.6 <= a <= 0.8"  = "#00B0F6",
  "a > 0.8"          = "#E76BF3"
)

upset(
  upset_data,
  intersect = rev(alpha_levels),
  sort_sets = FALSE,
  queries = lapply(alpha_levels, function(grp) {
    upset_query(set = grp, fill = alpha_colors[grp], color = alpha_colors[grp])
  }),
  set_sizes = upset_set_size() + scale_fill_manual(values = alpha_colors)
)
```


```{r}
reads_dmr_not %>% complete(region_id, alpha_group, type, fill = list(percent = 0)) %>% 
  group_by(region_id, alpha_group) %>%
  summarise(delta_cancer_normal = percent[type == "cancer"] - percent[type == "healthy"]) %>% 
  ungroup() %>% group_by(region_id) %>%
  mutate(max_delta_group = alpha_group[which.max(abs(delta_cancer_normal))],
         max_delta_group2 = alpha_group[order(abs(delta_cancer_normal), decreasing = TRUE)[2]]) %>%
  ungroup() %>% distinct(region_id, max_delta_group, max_delta_group2) %>%
  ggplot(aes(x = max_delta_group, fill = max_delta_group)) +
  geom_bar() +
  theme_bw()
```

```{r}
reads_dmr %>% complete(region_id, alpha_group, type, fill = list(percent = 0)) %>% 
  group_by(region_id, alpha_group) %>%
  summarise(delta_cancer_normal = percent[type == "cancer"] - percent[type == "healthy"]) %>% 
  ungroup() %>% group_by(region_id) %>%
  mutate(max_delta_group = alpha_group[which.max(abs(delta_cancer_normal))],
         max_delta_group2 = alpha_group[order(abs(delta_cancer_normal), decreasing = TRUE)[2]]) %>%
  ungroup() %>% distinct(region_id, max_delta_group, max_delta_group2) %>% 
  mutate(max_delta_group = factor(max_delta_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))) %>% 
  ggplot(aes(x = max_delta_group, fill = max_delta_group)) +
  geom_bar() +
  theme_bw() +
  scale_fill_manual(values = alpha_colors, drop = FALSE) +
  scale_x_discrete(drop = FALSE)
```

* plot a tile not overlapping a dmr
```{r}
reads_dmr_not %>% complete(region_id, alpha_group, type, fill = list(percent = 0)) %>% 
  group_by(region_id, alpha_group) %>%
  summarise(delta_cancer_normal = percent[type == "cancer"] - percent[type == "healthy"]) %>% 
  ungroup() %>% group_by(region_id) %>%
  mutate(max_delta_group = alpha_group[which.max(abs(delta_cancer_normal))],
         max_delta_group2 = alpha_group[order(abs(delta_cancer_normal), decreasing = TRUE)[2]]) %>%
  ungroup() %>% distinct(region_id, max_delta_group, max_delta_group2) %>% mutate(combo = paste0("G1:", max_delta_group, ", G2:", max_delta_group2)) %>%
  filter(max_delta_group == "0.4 < a < 0.6" | max_delta_group2 == "0.4 < a < 0.6")
```

```{r}
rrbs_plot5("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", "chr11", 126851501, 126852000) +
  geom_beta(betas_all, plot.height = 0.5, style = "smooth") 
```


```{r}
ranks <- tile_beta %>% filter(region_id == "chr11:126851501-126852000") %>% .$dtu_rank

props_med <- tile_prop("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", 
          dau_basictiles[ranks,])
#props_med <- props_med[,c(1:14,16:24)]
props_med <- props_med %>%
  mutate(
    alpha_group = case_when(
      alpha < 0.2  ~ "a < 0.2",
      alpha > 0.8  ~ "a > 0.8",
      alpha <= 0.4 ~ "0.2 <= a <= 0.4",
      alpha < 0.6  ~ "0.4 < a < 0.6",
      TRUE         ~ "0.6 <= a <= 0.8"
    )
  ) %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))) %>%
  count(type, dtu_rank, region_id, alpha_group, name = "tile_band_count") %>%
  group_by(region_id, dtu_rank, type) %>%
  mutate(prop_alpha_in_tile = tile_band_count / sum(tile_band_count)) %>%
  ungroup() %>%
  select(region_id, dtu_rank, alpha_group, type, prop_alpha_in_tile) %>%
  pivot_wider(names_from = type, values_from = prop_alpha_in_tile) %>%
  mutate(prop_change = cancer - healthy)

props_med %>% filter(dtu_rank == ranks) %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))) %>% pivot_longer(cols = c(cancer, healthy), names_to = "type", values_to = "percent") %>%
  ggplot(aes(x = alpha_group, y = percent, fill = type)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("cancer" = "#C0622A", "healthy" = "#1B3A6B")) +
  scale_x_discrete(labels = c("Low", "Low-med", "Med", "Med-high", "High")) +
  labs(x = "alpha band") +
  facet_wrap(~region_id) +
  theme_bw()
```

* plot all the dsitributions first
```{r}
ranks[2]
```

```{r}
for (i in 1:length(ranks)) {
  p <- props_med %>% filter(dtu_rank == ranks[i]) %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))) %>%
    pivot_longer(cols = c(cancer, healthy), names_to = "type", values_to = "percent") %>%
  ggplot(aes(x = alpha_group, y = percent, fill = type)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("cancer" = "#C0622A", "healthy" = "#1B3A6B")) +
  scale_x_discrete(labels = c("Low", "Low-med", "Med", "Med-high", "High")) +
  labs(x = "alpha band") +
  facet_wrap(~region_id) +
  theme_bw()
  print(p)
}
```



```{r}
healthy_fp <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_emseq_healthy_all_fp_exon_level.rds")
healthy_fp %>% filter(FDR < 0.05) %>% .[order(.$FDR),]
knitr::kable(healthy_fp %>% filter(FDR < 0.05) %>% .[order(.$FDR),] %>% .[1:10,c(9,1:8,13)], format = "html") |>
  writeLines("table.html")
```
```{r}
healthy_fp %>% filter(FDR < 0.05) %>% .[order(.$FDR),] %>% .[1:10,c(9,1:8,13)]
```

```{r}
tile_beta
```


```{r}

dau_basictiles <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dtu_basictiles_emseq_cancer_healthy.rds")

dau_optiles <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_emseq_cancer_healthy.rds")
```

```{r}
dau_basictiles %>% filter(FDR < 0.05) %>% nrow()
dau_optiles %>% filter(FDR < 0.05) %>% nrow()
```



```{r}
tile_overlap <- find_overlaps(as_granges(mutate(dau_basictiles, tiling = "basic")), as_granges(mutate(dau_optiles, tiling = "optiles"))) %>% data.frame()
tile_overlap$both <- paste0(tile_overlap$region_id.x, tile_overlap$region_id.y)
```
ggVennDiagram::ggVennDiagram(list(basic = basic_tiles_genes$gene_id, optiles = op_tiles_genes$gene_id))

```{r}
dau_basictiles <- dau_basictiles %>% mutate(both = tile_overlap[match(.$region_id, tile_overlap$region_id.x), "both"])
dau_optiles <- dau_optiles %>% mutate(both = tile_overlap[match(.$region_id, tile_overlap$region_id.y), "both"])
```

```{r}
ggVennDiagram::ggVennDiagram(list(basic = dau_basictiles$both, optiles = dau_optiles$both))
```

```{r}
ggVennDiagram::ggVennDiagram(list(basic = filter(dau_basictiles, FDR < 0.05)$both, optiles = filter(dau_optiles, FDR < 0.05)$both))
```

