---
title: "filteringReads"
author: "caitlinpage"
date: "2025-09-17"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r}
library(annotatr)
library(data.table)
library(plyranges)
library(tidyr)
library(stringr)
library(dplyr)
library(ggplot2)
library(edgeR)
library(OpTiles)
library(ggVennDiagram)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
library(missMethyl)
```

```{r}
seqs <- seqnames(BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38)[1:22]

cg_sites <- lapply(seqs, function(x) {
        cbind(data.frame(Biostrings::matchPattern("CG", BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38[[x]])),
            seqnames = x
        )}) %>% bind_rows()
```


```{r}
all_samples <- readRDS("/researchers/caitlin.page/cf_nano/r_output/all_rrbs_samples.rds")
optiles_all <- readRDS("/researchers/caitlin.page/cf_nano/r_output/optiles_all.rds")
dtu.optiles <- readRDS("/researchers/caitlin.page/cf_nano/r_output/ts.gene_optiles.rds")
dmrcate_seq_dmr <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dmr_encode_gm12878_k562.rds")

all_samples_simp <- all_samples %>% distinct(seqnames, read_start, read_end, strand, sample, read_id, alpha, meth_pattern, num_meth, total)
colnames(all_samples_simp)[2:3] <- c("start", "end")

dtu_tiles_5_exon <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_5_exon.rds")
```

##
NEW ANALYSIS

```{r}
all_samples_simp %>%
  ggplot(aes(x = as.factor(total), fill = as.factor(total))) +
  geom_bar() +
  theme(legend.position = "none") +
  theme_bw()
```

```{r}
all_samples_simp %>%
  ggplot(aes(x = total)) +
  geom_histogram()
```


```{r}
all_samples_simp %>%
  ggplot(aes(x = alpha)) +
  geom_density()
```

```{r}
all_samples_simp %>% filter(total >= 3) %>%
  ggplot(aes(x = alpha)) +
  geom_density() +
  theme_bw()
```


```{r}
all_samples_simp %>% mutate(min_3 = ifelse(total >= 3, TRUE, FALSE)) %>%
  ggplot(aes(x = alpha, colour = min_3)) +
  geom_density() +
  theme_bw()
```

```{r}
all_samples_simp %>% mutate(min_3 = ifelse(total >= 3, TRUE, FALSE)) %>%
  ggplot(aes(x = alpha, fill = min_3)) +
  geom_histogram(alpha = 0.2) +
  theme_bw()
```
* don't like plot

```{r}
all_samples_simp %>% mutate(min_3 = ifelse(total >= 3, TRUE, FALSE)) %>% filter(alpha %in% c(0, 0.5, 1)) %>%
  group_by(min_3, alpha) %>% summarise(n=n()) %>% ungroup() %>% group_by(alpha) %>% mutate(prop_alpha_filt = n/sum(n)) %>% .[1:5,]
```


```{r}
all_samples_simp %>% mutate(min_3 = ifelse(total >= 3, TRUE, FALSE)) %>%
  ggplot(aes(x = as.factor(total), fill = min_3)) +
  geom_bar() +
  theme_bw()
```

```{r}
all_samples_simp %>% mutate(min_3 = ifelse(total >= 3, TRUE, FALSE)) %>%
  ggplot(aes(x = as.factor(alpha), fill = min_3)) +
  geom_bar() +
  theme_bw()
```


# DTU on filtered to compare

* also add to function to have the delta

```{r}
optiles_all[1:5,]
```

```{r}
all_samples_simp %>% distinct(sample)
```

```{r}
bin_delta <- find_overlaps(as_granges(all_samples), as_granges(optiles_all)) %>% data.frame() %>%
  distinct(sample, read_id, read_start, read_end, alpha, region_id) %>% 
    mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8")) %>%
  mutate(transcript = paste0(region_id, "_", alpha_group)) %>% mutate(sample_type = ifelse(sample %in% c("b1", "b2"), "B", "CML")) %>%
  group_by(region_id, sample_type) %>% mutate(mean_alpha = mean(alpha)) %>% ungroup() %>% distinct(region_id, sample_type, mean_alpha) %>% .[order(.$region_id),]

bin_delta <- bin_delta %>% pivot_wider(names_from = sample_type, values_from = mean_alpha) %>% mutate(delta_b_minus_cml = B - CML)
```

* wait there are bins that don't have data in both?? - I thought I fixed my bins??

(later problem)


```{r}
bin_delta_min3 <- find_overlaps(as_granges(filter(all_samples, total >= 3)), as_granges(optiles_all)) %>% data.frame() %>%
  distinct(sample, read_id, read_start, read_end, alpha, region_id) %>% 
    mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8")) %>%
  mutate(transcript = paste0(region_id, "_", alpha_group)) %>% mutate(sample_type = ifelse(sample %in% c("b1", "b2"), "B", "CML")) %>%
  group_by(region_id, sample_type) %>% mutate(mean_alpha = mean(alpha)) %>% ungroup() %>% distinct(region_id, sample_type, mean_alpha) %>% .[order(.$region_id),]

bin_delta_min3 <- bin_delta_min3 %>% pivot_wider(names_from = sample_type, values_from = mean_alpha) %>% mutate(delta_b_minus_cml = B - CML)
```


```{r}
dtu_function <- function(all_samples, optiles_all, num_groups = 5) {
  dtu_optile <- find_overlaps(as_granges(all_samples), as_granges(optiles_all)) %>% data.frame() %>%
  distinct(sample, read_id, read_start, read_end, alpha, region_id) 
  
  if(num_groups == 5) {
  dtu_optile <- dtu_optile %>% 
    mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8"))
  } else if(num_groups == 3) {
  dtu_optile <- dtu_optile %>% 
    mutate(alpha_group = case_when(alpha < 0.34 ~ "a < 0.34",
                                alpha > 0.65 ~ "a > 0.65",
                                TRUE ~ "0.34 <= a <= 0.65"))
  } else if(num_groups == 4) {
  dtu_optile <- dtu_optile %>% 
    mutate(alpha_group = case_when(alpha <= 0.25 ~ "a <= 0.25",
                                alpha >= 0.8 ~ "a >= 0.75",
                                alpha > 0.25 & alpha <= 0.5 ~ "0.25 <= a <= 0.5",
                                TRUE ~ "0.5 < a < 0.75"))
}

dtu_optile <- dtu_optile %>% 
  mutate(transcript = paste0(region_id, "_", alpha_group)) %>%
  group_by(sample, transcript) %>%
 summarise(transcript_count = n()) %>% ungroup() %>% data.frame()

# build matrix of "transcript" counts - transcript rows, sample as columns
dtu_optile <- dtu_optile %>% pivot_wider(names_from = sample, values_from = transcript_count)
dtu_optile[is.na(dtu_optile)] <- 0
dtu_optile$gene_id <- sub("_.*", "", dtu_optile$transcript)

# DTU analysis with edgeR
## Alicia prefers gene method over Simes method (documentation slightly confusing about difference)
y <- DGEList(as.matrix(dtu_optile[,2:5]), group = c(1,1,2,2), genes = dtu_optile[,c("gene_id", "transcript")])
design <- model.matrix(~ c(1,1,2,2), data = y$samples)
fit <- glmQLFit(y, design, robust=TRUE)
ds <- diffSpliceDGE(fit, coef = 2, geneid = "gene_id", exonid = "transcript")
ts.gene_optiles <- topSpliceDGE(ds, test = "gene", number = Inf)

# add info to results: get proportion counts
gene_cg <- find_overlaps(as_granges(optiles_all), as_granges(cg_sites)) %>% data.frame() %>%
  group_by(region_id) %>%
  summarise(n_cg = n()) %>%
  ungroup() %>% data.frame()
fit$genes <- fit$genes %>% mutate(n_cg = gene_cg[match(.$gene_id, gene_cg$region_id), "n_cg"])
rpk <- y$counts/rep(fit$genes$n_cg/1000, times = ncol(y$counts))
tpm_mat <- sweep(rpk, 2, colSums(rpk)/1e6, "/")
tpm_mat <- tpm_mat %>% data.frame() %>% mutate(id = dtu_optile$transcript, gene_id = dtu_optile$gene_id)

tpm_mat <- tpm_mat %>%
  group_by(gene_id) %>%
  mutate(prop_S1_gm = b1/sum(b1),
         prop_S2_gm = b2/sum(b2),
         prop_S3_k = cml1/sum(cml1),
         prop_S4_k = cml2/sum(cml2)) %>%
  ungroup() %>% data.frame()


# add locations back in so compatible with plyranges stuff
## this bit only applicable to OPTILES because of the way it structures the ID
ts.gene_optiles <- ts.gene_optiles %>%
  mutate(region_id = gene_id) %>%
  separate_wider_delim(gene_id, delim = ":", names = c("seqnames", "gene_id")) %>%
  mutate(seqnames = gsub("chrchr", "chr", seqnames)) %>%
  separate_wider_delim(gene_id, delim = "-", names = c("start", "end"))
ts.gene_optiles$start <- as.double(ts.gene_optiles$start)
ts.gene_optiles$end <- as.double(ts.gene_optiles$end)
ts.gene_optiles$dtu_rank <- 1:nrow(ts.gene_optiles)
ts.gene_optiles
}

dtu_tiles_filt_min3 <- dtu_function(filter(all_samples, total >= 3), optiles_all, 5)
```

```{r}
dtu_tiles_filt_min3[1:5,]
```

```{r}
ggVennDiagram(list(all = filter(dtu.optiles, FDR < 0.05)$region_id, 
                   min_3 = filter(dtu_tiles_filt_min3, FDR < 0.05)$region_id))
```

```{r}
rbind(filter(dtu.optiles, FDR < 0.05) %>% mutate(type = "all"), filter(dtu_tiles_filt_min3, FDR < 0.05) %>% mutate(type = "min_3")) %>% distinct(region_id, type, dtu_rank) %>%
  pivot_wider(names_from = type, values_from = dtu_rank) %>%
  ggplot(aes(x = all, y = min_3)) +
  geom_point(alpha = 0.3)
```
* I'm happy to say that those are correlated, which is good it's not pulling out crazy stuff



```{r}
txdb_genes <- genes(TxDb.Hsapiens.UCSC.hg19.knownGene)
```
```{r}
bin_delta <- data.frame(bin_delta)
bin_delta_min3 <- data.frame(bin_delta_min3)
```



```{r}
dtu_filt_and_all <- rbind(dtu.optiles %>% mutate(type = "all", delta_b_minus_cml = bin_delta[match(.$region_id, bin_delta$region_id), "delta_b_minus_cml"]), 
                          dtu_tiles_filt_min3 %>% 
                            filter(FDR < 0.05) %>% mutate(type = "min_3", delta_b_minus_cml = bin_delta_min3[match(.$region_id, bin_delta_min3$region_id), "delta_b_minus_cml"]))
```


```{r}
dtu_filt_and_all %>%
  ggplot(aes(x = delta_b_minus_cml, y = -log10(FDR), colour = type)) +
  geom_point(alpha = 0.3)
```

```{r}
dtu_filt_and_all %>% filter(type == "all") %>%
  ggplot(aes(x = delta_b_minus_cml, y = -log10(FDR))) +
  geom_point(alpha = 0.3)
```

```{r}
dtu_filt_and_all %>% filter(type == "min_3") %>%
  ggplot(aes(x = delta_b_minus_cml, y = -log10(FDR))) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 2, colour = "red", linetype = "dashed")
```


```{r}
dtu_filt_and_all %>% filter(type == "all") %>%
  ggplot(aes(x = delta_b_minus_cml)) +
  geom_density()

dtu_filt_and_all %>% filter(type == "all") %>%
  ggplot(aes(x = delta_b_minus_cml)) +
  geom_histogram(bins = 50)
```

```{r}
dtu_filt_and_all %>% filter(type == "all") %>% filter(FDR < 0.05) %>%
  ggplot(aes(x = delta_b_minus_cml)) +
  geom_density()

dtu_filt_and_all %>% filter(type == "all") %>% filter(FDR < 0.05) %>%
  ggplot(aes(x = delta_b_minus_cml)) +
  geom_histogram(bins = 50)
```



```{r}
dtu_filt_and_all %>% filter(type == "min_3") %>% 
  ggplot(aes(x = delta_b_minus_cml)) +
  geom_density()

dtu_filt_and_all %>% filter(type == "min_3") %>%
  ggplot(aes(x = delta_b_minus_cml)) +
  geom_histogram(bins = 50)
```

```{r}
dtu_filt_and_all %>% filter(type == "min_3") %>% filter(FDR < 0.05) %>%
  ggplot(aes(x = delta_b_minus_cml)) +
  geom_density()

dtu_filt_and_all %>% filter(type == "min_3") %>% filter(FDR < 0.05) %>%
  ggplot(aes(x = delta_b_minus_cml)) +
  geom_histogram(bins = 50)
```


#######

# OLD ANALYSIS

* this was originally done on the cell free data
* hence extra code about read length and healthy/cancer

*rest can rerun with rrbs data

## Introduction



```{r}
all_samples_simp %>% filter(alpha == 0) %>% .$total %>% summary()
all_samples_simp %>% filter(alpha == 1) %>% .$total %>% summary()
```
```{r}
all_samples_simp %>% filter(alpha == 0, total >= 3) %>% .$total %>% summary()
all_samples_simp %>% filter(alpha == 1, total >= 3) %>% .$total %>% summary()

all_samples_simp %>% filter(alpha <= 0.2) %>% .$total %>% summary()
all_samples_simp %>% filter(alpha >= 0.8) %>% .$total %>% summary()


all_samples_simp %>% filter(alpha <= 0.2, total >= 3) %>% .$total %>% summary()
all_samples_simp %>% filter(alpha >= 0.8, total >= 3) %>% .$total %>% summary()
```

```{r}
all_samples_simp %>% filter(alpha %in% c(0,1)) %>%
  ggplot(aes(x = log2(total), colour = as.factor(alpha))) +
  geom_density()

all_samples_simp %>% filter(alpha %in% c(0,1)) %>%
  ggplot(aes(x = as.factor(alpha), y= total, colour = as.factor(alpha))) +
  geom_boxplot()

all_samples_simp %>% filter(alpha %in% c(0,1)) %>%
  ggplot(aes(x = as.factor(alpha), y= log2(total), colour = as.factor(alpha))) +
  geom_boxplot()
```

```{r}
all_samples_simp %>% filter(alpha %in% c(0,1), total >= 3) %>%
  ggplot(aes(x = log2(total), colour = as.factor(alpha))) +
  geom_density()

all_samples_simp %>% filter(alpha %in% c(0,1), total >= 3) %>%
  ggplot(aes(x = as.factor(alpha), y= total, colour = as.factor(alpha))) +
  geom_boxplot()

all_samples_simp %>% filter(alpha %in% c(0,1), total >= 3) %>%
  ggplot(aes(x = as.factor(alpha), y= log2(total), colour = as.factor(alpha))) +
  geom_boxplot()
```
* filtering to reads with at least 3 cpgs evens out the median in the log2 boxplot for alpha groups

```{r}
all_samples_simp %>% filter(alpha >= 0.8 | alpha <= 0.2) %>% filter(total >= 3) %>% mutate(alpha_group = ifelse(alpha >= 0.8, ">=0.8", "<=0.2")) %>%
  ggplot(aes(x = log2(total), colour = as.factor(alpha_group))) +
  geom_density()

all_samples_simp %>% filter(alpha >= 0.8 | alpha <= 0.2) %>% filter(total >= 3) %>% mutate(alpha_group = ifelse(alpha >= 0.8, ">=0.8", "<=0.2")) %>%
  ggplot(aes(x = as.factor(alpha_group), y= total, colour = as.factor(alpha_group))) +
  geom_boxplot()

all_samples_simp %>% filter(alpha >= 0.8 | alpha <= 0.2) %>% filter(total >= 3) %>% mutate(alpha_group = ifelse(alpha >= 0.8, ">=0.8", "<=0.2")) %>%
  ggplot(aes(x = as.factor(alpha_group), y= log2(total), colour = as.factor(alpha_group))) +
  geom_boxplot()
```

```{r}
all_samples_simp %>% filter(alpha >= 0.8 | alpha <= 0.2) %>% mutate(alpha_group = ifelse(alpha >= 0.8, ">=0.8", "<=0.2")) %>%
  ggplot(aes(x = log2(total), colour = as.factor(alpha_group))) +
  geom_density()

all_samples_simp %>% filter(alpha >= 0.8 | alpha <= 0.2) %>% mutate(alpha_group = ifelse(alpha >= 0.8, ">=0.8", "<=0.2")) %>%
  ggplot(aes(x = as.factor(alpha_group), y= total, colour = as.factor(alpha_group))) +
  geom_boxplot()

all_samples_simp %>% filter(alpha >= 0.8 | alpha <= 0.2) %>% mutate(alpha_group = ifelse(alpha >= 0.8, ">=0.8", "<=0.2")) %>%
  ggplot(aes(x = as.factor(alpha_group), y= log2(total), colour = as.factor(alpha_group))) +
  geom_boxplot()
```
* same thing with lower cutoff alpha: median num cpgs higher in 0.8 - is not present when you filter to min cpg = 3


* anyway this isn't really saying anything - except I guess reads that are alpha = 0 are more likely to have less cpgs
```{r}
t.test(filter(all_samples_simp, alpha == 0)$total, filter(all_samples_simp, alpha == 1)$total)
```
* and to be shorter in length
```{r}
#t.test(filter(all_samples_simp, alpha == 0)$read_length, filter(all_samples_simp, alpha == 1)$read_length)
```
* however the differences are small

```{r}
t.test(filter(all_samples_simp, alpha == 0, total >= 3)$total, filter(all_samples_simp, alpha == 1, total >= 3)$total)

#t.test(filter(all_samples_simp, alpha == 0, total >= 3)$read_length, filter(all_samples_simp, alpha == 1, total >= 3)$read_length)
```
* but because there's so many reads - it's significant even when filtering and the differences are even smaller

# add in cancer info


all_samples %>% filter(alpha %in% c(0,1)) %>%
  ggplot(aes(x = as.factor(alpha), y= total, colour = type)) +
  geom_boxplot()

all_samples %>% filter(alpha %in% c(0,1)) %>%
  ggplot(aes(x = as.factor(alpha), y= log2(total), colour = type)) +
  geom_boxplot()

all_samples %>% filter(alpha %in% c(0,1), total >= 3) %>%
  ggplot(aes(x = as.factor(alpha), y= total, colour = type)) +
  geom_boxplot()

all_samples %>% filter(alpha %in% c(0,1), total >= 3) %>%
  ggplot(aes(x = as.factor(alpha), y= log2(total), colour = type)) +
  geom_boxplot()

* with all the data, alpha = 1 has a bigger distrib but that goes away when min cpg = 3
* healthy bigger distrib than luad - but median is basically the same (still might be stat sig)


t.test(filter(all_samples, alpha == 0, type == "healthy")$total, filter(all_samples, alpha == 0, type == "luad")$total)

t.test(filter(all_samples, alpha == 1, type == "healthy")$total, filter(all_samples, alpha == 1, type == "luad")$total)


