---
title: "methWithinTile"
author: "caitlinpage"
date: "2026-09-24"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

## Introduction

* can I distinguish between interesting vs like uniform methylation in a tile
* so within the group (because means more summarised than cancer)
* like with the counts of the tiles
* wait is that just the proportion test but within itself
* dude
```{r}
library(BSgenome.Hsapiens.UCSC.hg38)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(edgeR)
library(plyranges)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(data.table)
library(duckdb)
library(arrow)
```

```{r}
dtu_emseq <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_emseq_cancer_healthy.rds")
dmr_emseq <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dmr_emseq_strict.rds")
dmr_emseq <- dmr_emseq %>% .[order(.$min_smoothed_fdr),] %>% mutate(dmr_rank = 1:n())
```


```{r}
dau_basictiles
```

```{r}
tile_beta <- find_overlaps(as_granges(dau_basictiles), betas_c %>% mutate(coverage = meth_count_pos + unmeth_count_pos) %>% filter(coverage > 10) %>% group_by(seqnames, start) %>% summarise(beta_cancer = mean(beta)) %>% mutate(width = 1) %>% as_granges()) %>% data.frame()

setDT(tile_beta)
tile_beta <- tile_beta[, .(mean_beta_cancer = mean(beta_cancer)), by = eval(names(tile_beta)[1:13])]

tile_beta <- find_overlaps(as_granges(tile_beta), betas_h %>% mutate(coverage = meth_count_pos + unmeth_count_pos) %>% filter(coverage > 10) %>% group_by(seqnames, start) %>% summarise(beta_healthy = mean(beta)) %>% mutate(width = 1) %>% as_granges()) %>% data.frame()

setDT(tile_beta)
tile_beta <- tile_beta[, .(mean_beta_healthy = mean(beta_healthy)), by = eval(names(tile_beta)[1:14])]

tile_beta <- tile_beta %>% mutate(delta_beta = mean_beta_cancer - mean_beta_healthy)
```

```{r}
betas_c$type <- "cancer"
betas_h$type <- "healthy"
betas_all <- rbind(betas_c, betas_h)
```


```{r}
tile_beta
```

```{r}
props_from_test
```


```{r}
reads_region <- function(read_data_path_a, read_data_path_b, genes, cpgs = cg_sites,
                       jit_amount = 0.1) {
  set.seed(42)

  # ── 1. Read only the matching rows — never loads unmatched data into RAM ───
  con <- dbConnect(duckdb())
  on.exit(dbDisconnect(con))
  
  genes_df <- as.data.frame(genes)[, c("seqnames", "start", "end", "gene_id")]
  genes_df$seqnames <- as.character(genes_df$seqnames)
  duckdb_register(con, "genes", genes_df)
  
  overlap_query <- function(path) {
    dbGetQuery(con, glue::glue("
      SELECT DISTINCT r.*, g.gene_id
      FROM read_parquet('{path}') r
      JOIN genes g
        ON r.seqnames = g.seqnames
       AND r.start >= g.start
       AND r.end   <= g.end
    "))
  }
  
  read_data_a <- overlap_query(read_data_path_a)
  
  read_data_b <- overlap_query(read_data_path_b)
  
  read_data_a$type <- "healthy"
  read_data_b$type <- "cancer"
  
  read_data <- rbind(read_data_a, read_data_b)
  
  jitter_amount <- runif(nrow(read_data), -jit_amount, jit_amount)
  read_data$alpha_jit <- read_data$alpha + jitter_amount
  read_data <- data.frame(read_data)
  
  read_data <- read_data %>% mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                 alpha > 0.8 ~ "a > 0.8",
                                 alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                 alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                 TRUE ~ "0.6 <= a <= 0.8"))
  read_data <- read_data %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))) %>%
    count(type, gene_id, alpha_group, name = "tile_band_count") %>% 
  group_by(gene_id, type) %>% mutate(percent = tile_band_count/sum(tile_band_count)) %>% ungroup()
  read_data
}
reads_genebody <- reads_region("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", genes_txdb)
```

```{r}
reads_genebody
```


* find me a noise to signal region

* ie medium alpha to not

```{r}
tile_beta %>% filter(abs(mean_beta_cancer) >= 0.4, abs(mean_beta_cancer) <= 0.6 | 
                     abs(mean_beta_healthy) >= 0.4, abs(mean_beta_healthy) <= 0.6) %>%
  filter(FDR < 0.05) %>% filter(N > 10) %>% .[order(abs(.$delta_beta), decreasing = TRUE),]
```

tile_beta %>% filter(abs(mean_beta_cancer) >= 0.4, abs(mean_beta_cancer) <= 0.6,
                     abs(mean_beta_healthy) >= 0.4, abs(mean_beta_healthy) <= 0.6) %>%
  filter(FDR < 0.05) %>% filter(N > 10)

```{r}
ranks <- tile_beta %>% filter(abs(mean_beta_cancer) >= 0.4, abs(mean_beta_cancer) <= 0.6 | 
                     abs(mean_beta_healthy) >= 0.4, abs(mean_beta_healthy) <= 0.6) %>%
  filter(FDR < 0.05) %>% filter(N > 10) %>% .[order(abs(.$delta_beta), decreasing = TRUE),] %>% .$dtu_rank
```


```{r}
props_a_med <- tile_prop("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", 
          dau_basictiles[ranks,])
```


```{r}
#chr10:132232001-132232500
dau_basictiles[251,]

rrbs_plot5("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", "chr10", 132232001, 132232500) +
  geom_beta(betas_all, plot.height = 0.5, style = "smooth") 

props_a_med %>% filter(dtu_rank == 251) %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))) %>% pivot_longer(cols = c(cancer, healthy), names_to = "type", values_to = "percent") %>%
  ggplot(aes(x = alpha_group, y = percent, fill = type)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("cancer" = "#C0622A", "healthy" = "#1B3A6B")) +
  scale_x_discrete(labels = c("Low", "Low-med", "Med", "Med-high", "High")) +
  labs(x = "alpha band") +
  facet_wrap(~region_id) +
  theme_bw()
```

* this doesn't look like much at all
* how did I find it in the rrbs...

* oh I looked at strings of betas - then did some other stuff
* no it was from the alpha band level dau results
* looking at tiles where medium to low/high was significant
* I haven't done that version with the emseq

```{r}
optiles_all <- readRDS("/researchers/caitlin.page/cf_nano/r_output/optiles_all_emseq.rds")
```

```{r}
tiles_500bp_cg
```


## Tile genome

```{r}
# Get chromosome lengths for standard chromosomes only
seqinfo_hg38 <- seqlengths(BSgenome.Hsapiens.UCSC.hg38)
std_chroms <- grepl("^chr([1-9]|1[0-9]|2[0-2]|X|Y|M)$", names(seqinfo_hg38))
seqinfo_hg38 <- seqinfo_hg38[std_chroms]

# Tile into 500 bp non-overlapping windows
tiles_500bp <- GenomicRanges::tileGenome(seqinfo_hg38, tilewidth = 500, cut.last.tile.in.chrom = TRUE)

# Add a tile number column
tiles_500bp <- tiles_500bp %>%
  data.frame() %>%
  mutate(tile_num = 1:n()) %>%
  as_granges()


# cpgs
seqs <- seqnames(BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38)[1:22]

cg_sites <- lapply(seqs, function(x) {
  cbind(data.frame(Biostrings::matchPattern("CG", BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38[[x]])),
        seqnames = x
  )}) %>% bind_rows()

tiles_500bp_cg <- find_overlaps(tiles_500bp, as_granges(cg_sites)) %>% data.frame() %>%
  group_by(tile_num) %>% mutate(n_cg = n())
tiles_500bp_cg <- tiles_500bp_cg %>% ungroup() %>% distinct()
tiles_500bp_cg$tile_id <- paste0(tiles_500bp_cg$seqnames, ":", tiles_500bp_cg$start, "-", tiles_500bp_cg$end)
```

```{r}
load_counts <- function(read_data_path_a, read_data_path_b, tiles) {
#overlap_pos <- find_overlaps(as_granges(optiles_all), as_granges(all_samples)) %>% data.frame()
  con <- dbConnect(duckdb())
  on.exit(dbDisconnect(con))
  
  tiles_df <- as.data.frame(tiles)[, c("seqnames", "start", "end", "tile_id")]
  tiles_df$seqnames <- as.character(tiles_df$seqnames)
  duckdb_register(con, "tiles", tiles_df)
  
  overlap_query <- function(path) {
    dbGetQuery(con, glue::glue("
      SELECT DISTINCT r.*, t.tile_id
      FROM read_parquet('{path}') r
      JOIN tiles t
        ON r.seqnames = t.seqnames
       AND r.start >= t.start
       AND r.end   <= t.end
    "))
  }
  
  read_data_a <- overlap_query(read_data_path_a)
  
  read_data_b <- overlap_query(read_data_path_b)
  
  read_data_a$type <- "healthy"
  read_data_b$type <- "cancer"
  
  read_data <- rbind(read_data_a, read_data_b)
# dtu
  read_counts <- read_data %>% mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                 alpha > 0.8 ~ "a > 0.8",
                                 alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                 alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                 TRUE ~ "0.6 <= a <= 0.8"))
  read_counts <- read_counts %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))) %>%
    count(type, tile_id, alpha_group, name = "tile_band_count") %>% 
  group_by(tile_id, type) %>% mutate(percent = tile_band_count/sum(tile_band_count)) %>% ungroup()
  
dtu_optile <- read_data %>%
  distinct(sample, read_id, read_start, read_end, alpha, tile_id, type) %>%
  mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                 alpha > 0.8 ~ "a > 0.8",
                                 alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                 alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                 TRUE ~ "0.6 <= a <= 0.8"), 
         transcript = paste0(tile_id, "_", alpha_group)) %>%
  group_by(sample, transcript) %>%
  summarise(transcript_count = n()) %>% ungroup() %>% data.frame()

# build matrix of "transcript" counts - transcript rows, sample as columns
dtu_optile <- dtu_optile %>% pivot_wider(names_from = sample, values_from = transcript_count, values_fill = 0)
dtu_optile$gene_id <- sub("_.*", "", dtu_optile$transcript)



# DTU analysis with edgeR
## Alicia prefers gene method over Simes method (documentation slightly confusing about difference)
y <- DGEList(as.matrix(dtu_optile[,2:13]), group = c(1,1,1,1,1,1,2,2,2,2,2,2), genes = dtu_optile[,c("gene_id", "transcript")])
design <- model.matrix(~ c(1,1,1,1,1,1,2,2,2,2,2,2), data = y$samples)
fit <- glmQLFit(y, design, robust=TRUE)
ds <- diffSpliceDGE(fit, coef = 2, geneid = "gene_id", exonid = "transcript")
dau_simes <- topSpliceDGE(ds, test = "Simes", number = Inf)
dau_exon <- topSpliceDGE(ds, test = "exon", number = Inf)
list(dau_simes, dau_exon, read_counts)
}

dau_band_level <- load_counts("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", tiles_500bp_cg)
```


```{r}
saveRDS(dau_band_level, "/researchers/caitlin.page/cf_nano/r_output/emseq_dau_band_level.rds")
```


```{r}
ts.gene_optiles <- ts.gene_optiles %>%
  mutate(region_id = gene_id) %>%
  separate_wider_delim(gene_id, delim = ":", names = c("seqnames", "gene_id")) %>%
  mutate(seqnames = gsub("chrchr", "chr", seqnames)) %>%
  separate_wider_delim(gene_id, delim = "-", names = c("start", "end"))
ts.gene_optiles$start <- as.double(ts.gene_optiles$start)
ts.gene_optiles$end <- as.double(ts.gene_optiles$end)
ts.gene_optiles$dtu_rank <- 1:nrow(ts.gene_optiles)
saveRDS(ts.gene_optiles, "/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_emseq_cancer_healthy.rds")

```


```{r}
tile_prop <- function(read_data_path_a, read_data_path_b, tiles, cols_from_tiles = NULL) {

  # ── 1. Read only the matching rows — never loads unmatched data into RAM ───
  con <- dbConnect(duckdb())
  duckdb::dbWriteTable(con, "tiles_temp", tiles, overwrite = TRUE)
  on.exit(duckdb::dbRemoveTable(con, "tiles_temp"), add = TRUE)
  
  
  # Build SELECT clause for tile columns
  if (is.null(cols_from_tiles)) {
    # Exclude coordinate cols to avoid clashes with parquet
    tile_cols <- setdiff(colnames(tiles), c("seqnames", "start", "end"))
  } else {
    tile_cols <- cols_from_tiles
  }
  
  tile_select <- if (length(tile_cols) > 0) {
    paste(glue::glue(', t."{tile_cols}"'), collapse = "")
  } else ""
  
  read_data_a <- dbGetQuery(con, glue::glue("
    SELECT b.*{tile_select}
    FROM read_parquet('{read_data_path_a}') b
    JOIN tiles_temp t
      ON b.seqnames = t.seqnames
     AND b.start <= t.end
     AND b.end   >= t.start
  "))
  
  read_data_b <- dbGetQuery(con, glue::glue("
    SELECT b.*{tile_select}
    FROM read_parquet('{read_data_path_b}') b
    JOIN tiles_temp t
      ON b.seqnames = t.seqnames
     AND b.start <= t.end
     AND b.end   >= t.start
  "))
  
  read_data_a$type <- "healthy"
  read_data_b$type <- "cancer"
  props_med <- rbind(read_data_a, read_data_b)

  props_med <- props_med[,c(1:14,16:24)]
  props_med <- props_med %>%
  mutate(
    alpha_group = case_when(
      alpha < 0.2  ~ "a < 0.2",
      alpha > 0.8  ~ "a > 0.8",
      alpha <= 0.4 ~ "0.2 <= a <= 0.4",
      alpha < 0.6  ~ "0.4 < a < 0.6",
      TRUE         ~ "0.6 <= a <= 0.8"
    )
  ) %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8"))) %>%
  count(type, dtu_rank, region_id, alpha_group, name = "tile_band_count") %>%
  group_by(region_id, dtu_rank, type) %>%
  mutate(prop_alpha_in_tile = tile_band_count / sum(tile_band_count)) %>%
  ungroup() %>%
  select(region_id, dtu_rank, alpha_group, type, prop_alpha_in_tile) %>%
  pivot_wider(names_from = type, values_from = prop_alpha_in_tile) %>%
  mutate(prop_change = cancer - healthy)
  props_med
}
```


```{r}
reads_eg <- function(read_data_path_a, read_data_path_b, chrom, region_start, region_end,
                       jit_amount = 0.1) {
  set.seed(42)

  # ── 1. Read only the matching rows — never loads unmatched data into RAM ───
  con <- dbConnect(duckdb())
  on.exit(dbDisconnect(con))
  
  read_data_a <- dbGetQuery(con, glue::glue("
    SELECT *
    FROM read_parquet('{read_data_path_a}')
    WHERE seqnames = '{chrom}'
      AND \"start\" >= {region_start}
      AND \"end\"   <= {region_end}
  "))
  
  read_data_b <- dbGetQuery(con, glue::glue("
    SELECT *
    FROM read_parquet('{read_data_path_b}')
    WHERE seqnames = '{chrom}'
      AND \"start\" >= {region_start}
      AND \"end\"   <= {region_end}
  "))
  
  read_data_a$type <- "healthy"
  read_data_b$type <- "cancer"
  
  read_data <- rbind(read_data_a, read_data_b)
  
  read_data <- data.frame(read_data)
  
  read_data <- read_data %>% mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                 alpha > 0.8 ~ "a > 0.8",
                                 alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                 alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                 TRUE ~ "0.6 <= a <= 0.8"))
  read_data <- read_data %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8")))
  
  # Add jittered coordinates to new columns in the data frame
  long_data <- read_data %>%
    mutate(meth_status = strsplit(meth_pattern, ""), start = strsplit(genom_positions, ",")) %>%
    unnest(meth_status, start) %>%
    mutate(start = as.double(start))
  
  list(read_data, long_data)
}

```

rrbs_plot5("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", "chr10", 132232001, 132232500) +
  geom_beta(betas_all, plot.height = 0.5, style = "smooth")
```{r}
nrow(arrow::open_dataset('/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet'))
nrow(arrow::open_dataset('/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet'))
```

```{r}
123043398 + 718729496
```


```{r}
read_example <- reads_eg("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", "chr1", 100000, 110000)
```

```{r}
read_example[[1]] %>% filter(total > 2) %>% mutate(strand = "*") %>% distinct()
```

```{r}
read_example[[2]] %>% filter(total > 2)
```

```{r}
knitr::kable(read_example[[1]] %>% filter(total > 2) %>% mutate(strand = "*") %>% distinct(), format = "html") |>
  writeLines("table.html")

knitr::kable(read_example[[2]] %>% filter(total > 2) %>% mutate(strand = "*") %>% distinct(), format = "html") |>
  writeLines("table.html")
```

```{r}
dau_basictiles <- distinct(dau_basictiles)
```

```{r}
knitr::kable(dau_basictiles[1:10,], format = "html") |>
  writeLines("table.html")
```

```{r}
dau_basictiles %>% filter(seqnames == "chr1") %>% .[order(.$start),]
```

```{r}
dtu_emseq %>% filter(seqnames == "chr1") %>% .[order(.$start),]
```

