---
title: "mp3_betaGroupsHighBeta"
author: "caitlinpage"
date: "2025-11-21"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

## Introduction

```{r}
library(data.table)
library(plyranges)
library(tidyr)
library(stringr)
library(dplyr)
library(ggplot2)
```

## Data files
```{r}
cg_sites <- cbind(data.frame(Biostrings::matchPattern("CG", BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38[["chr22"]])), seqnames = "chr22") %>% 
  mutate(pos = paste0(seqnames, "-", start)) %>%
  relocate(pos, seqnames)
cg_sites$index <- 1:nrow(cg_sites)
```



* rrbs cell line
```{r}
source("../code/processData.R")

for_beta <- readRDS("/researchers/caitlin.page/cf_nano/r_output/rrbs_alpha_beta.rds")
for_beta <- for_beta %>% filter(seqnames == "chr22")
rrbs_for_beta_pat <- pat_for_wgbs(for_beta)
```


```{r}
for_beta <- for_beta %>% group_by(start) %>% mutate(coverage = n()) %>% ungroup() %>% data.frame()
just_beta <- for_beta %>% distinct(start, beta, coverage) %>% .[order(.$start),]
```

# just look at data when beta > 0.9, beta < 1

```{r}
just_beta <- just_beta %>% mutate(seqnames = "chr22") %>%
  mutate(row_num = 1:nrow(.)) %>%
  mutate(beta_high = ifelse(beta >= 0.9 & beta < 1, "high", "other"))
just_beta[1:10,]
```

```{r}
just_beta %>% 
  group_by(beta_high) %>%
  mutate(consec = n()) %>%
  mutate(count = 1:n()) %>%
  ungroup() %>%
  mutate(length = sequence(rle(.$beta_high)$lengths)) %>%
  group_by(beta_high) %>%
  filter(beta_high == "high", length >= 3) %>% nrow()
```

```{r}
grouped_betas <- just_beta %>% 
  group_by(beta_high) %>%
  mutate(consec = n()) %>%
  mutate(count = 1:n()) %>%
  ungroup() %>%
  mutate(length = sequence(rle(.$beta_high)$lengths)) %>% 
  mutate(group = cumsum(c(TRUE, diff(length) <= 0))) %>%
  group_by(group) %>%
  mutate(length = n(), av_coverage = mean(coverage), av_beta = mean(beta)) %>%
  ungroup() %>% filter(beta_high == "high", length > 2) %>% data.frame() 
grouped_betas[1:10,]
```

```{r}
for_beta %>% mutate(beta_range_group_num = grouped_betas[match(.$start, grouped_betas$start), "group"],
                    beta_range_group_length = grouped_betas[match(.$start, grouped_betas$start), "length"]) %>% 
  filter(!is.na(beta_range_group_num)) %>% 
  ggplot(aes(x = as.factor(beta_range_group_num), y = alpha)) +
  geom_boxplot()

for_beta %>% mutate(beta_range_group_num = grouped_betas[match(.$start, grouped_betas$start), "group"],
                    beta_range_group_length = grouped_betas[match(.$start, grouped_betas$start), "length"]) %>% 
  filter(!is.na(beta_range_group_num)) %>% 
  ggplot(aes(colour = as.factor(beta_range_group_num), x = alpha)) +
  geom_density()
```

```{r}
poten_groups <- for_beta %>% mutate(beta_range_group_num = grouped_betas[match(.$start, grouped_betas$start), "group"],
                    beta_range_group_length = grouped_betas[match(.$start, grouped_betas$start), "length"]) %>% 
  filter(!is.na(beta_range_group_num)) %>% distinct(read_id, alpha, beta_range_group_num, beta_range_group_length) %>%
  mutate(beta_range_group_num = as.factor(beta_range_group_num)) %>%
  group_by(beta_range_group_num) %>%
  summarise(bimodal_p = list(diptest::dip.test(alpha)$p.value)
  ) %>%
  unnest(cols = c(bimodal_p)) %>% filter(bimodal_p < 0.01) %>% .$beta_range_group_num
```

```{r}
source("../code/processData.R")
plot_list <- list()
for (i in 1:length(poten_groups)) {
  row_pos <- just_beta %>%
  group_by(beta_high) %>%
  mutate(consec = n()) %>%
  mutate(count = 1:n()) %>%
  ungroup() %>%
  mutate(length = sequence(rle(.$beta_high)$lengths)) %>% 
  mutate(group = cumsum(c(TRUE, diff(length) <= 0))) %>% filter(group == i)
  min_row <- min(row_pos$row_num)
  max_row <- max(row_pos$row_num)
  plot_list[[i]] <- rrbs_plot(for_beta, for_beta %>% filter(start %in% just_beta[min_row:max_row,]$start) %>% .$read_id %>% unique(), count_instead = TRUE)
}
plot_list
```

## nanopore analysis
```{r}
alpha_and_beta2 <- readRDS("/researchers/caitlin.page/cf_nano/r_output/nano_alpha_beta.rds")
```

```{r}
alpha_and_beta2 <- alpha_and_beta2 %>% group_by(start) %>% mutate(coverage = n()) %>% ungroup() %>% data.frame()
nano_beta <- alpha_and_beta2 %>% distinct(seqnames, start, beta, coverage) %>% .[order(.$start),]
```

```{r}
nano_beta <- nano_beta %>% 
  mutate(row_num = 1:nrow(.)) %>%
  mutate(beta_high = ifelse(beta == 0.5, "high", "other"))
nano_beta[1:10,]
```

```{r}
nano_beta %>% group_by(beta) %>% summarise(n())
```


```{r}
grouped_betas <- nano_beta %>% 
  group_by(beta_high) %>%
  mutate(consec = n()) %>%
  mutate(count = 1:n()) %>%
  ungroup() %>%
  mutate(length = sequence(rle(.$beta_high)$lengths)) %>% 
  mutate(group = cumsum(c(TRUE, diff(length) <= 0))) %>%
  group_by(group) %>%
  mutate(length = n(), av_coverage = mean(coverage), av_beta = mean(beta)) %>%
  ungroup() %>% filter(beta_high == "high", length > 2) %>% data.frame() 
grouped_betas[1:10,]
```

```{r}
alpha_and_beta2 %>% filter(start == 10729016)
reads <- alpha_and_beta2 %>% filter(start == 10729016) %>% .$read_id
```

```{r}
rrbs_plot(alpha_and_beta2, reads)
```


```{r}
alpha_and_beta2 %>% mutate(beta_range_group_num = grouped_betas[match(.$start, grouped_betas$start), "group"],
                    beta_range_group_length = grouped_betas[match(.$start, grouped_betas$start), "length"]) %>% 
  filter(!is.na(beta_range_group_num)) %>% 
  ggplot(aes(x = as.factor(beta_range_group_num), y = alpha)) +
  geom_boxplot()

alpha_and_beta2 %>% mutate(beta_range_group_num = grouped_betas[match(.$start, grouped_betas$start), "group"],
                    beta_range_group_length = grouped_betas[match(.$start, grouped_betas$start), "length"]) %>% 
  filter(!is.na(beta_range_group_num)) %>% 
  ggplot(aes(colour = as.factor(beta_range_group_num), x = alpha)) +
  geom_density()
```

```{r}
poten_groups <- alpha_and_beta2 %>% mutate(beta_range_group_num = grouped_betas[match(.$start, grouped_betas$start), "group"],
                    beta_range_group_length = grouped_betas[match(.$start, grouped_betas$start), "length"]) %>% 
  filter(!is.na(beta_range_group_num)) %>% distinct(read_id, alpha, beta_range_group_num, beta_range_group_length) %>%
  mutate(beta_range_group_num = as.factor(beta_range_group_num)) %>%
  group_by(beta_range_group_num) %>%
  summarise(bimodal_p = list(diptest::dip.test(alpha)$p.value)
  ) %>%
  unnest(cols = c(bimodal_p)) %>% filter(bimodal_p < 0.01) %>% .$beta_range_group_num
```

```{r}
source("../code/processData.R")
plot_list <- list()
for (i in 1:length(poten_groups)) {
  row_pos <- nano_beta %>%
  group_by(beta_high) %>%
  mutate(consec = n()) %>%
  mutate(count = 1:n()) %>%
  ungroup() %>%
  mutate(length = sequence(rle(.$beta_high)$lengths)) %>% 
  mutate(group = cumsum(c(TRUE, diff(length) <= 0))) %>% filter(group == i)
  min_row <- min(row_pos$row_num)
  max_row <- max(row_pos$row_num)
  plot_list[[i]] <- rrbs_plot(for_beta, for_beta %>% filter(start %in% just_beta[min_row:max_row,]$start) %>% .$read_id %>% unique(), count_instead = TRUE)
}
plot_list
```






######

```{r}
for_beta
```

```{r}
rrbs_for_beta_pat
```

