---
title: "otherReadStyleMethods"
author: "caitlinpage"
date: "2026-04-01"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

## Introduction
Description of methods
Analysis can be found [here](otherReadStyleMethodsAnalysis.html)

```{r}

```

* run metheor on cluster


conda create -n run_metheor
conda activate run_metheor
** deactivate active enviro: conda deactivate


conda install -c dohlee metheor

* NOTE: requires sorted and indexed bams
** MUST RUN IN OUTPUT DIRECTORY - FAILS IF PROVIDE PATH TO OUTPUT FILE INSTEAD OF NAME OF OUTPUT FILE

* everything is 0 based coordinates
* my alpha/beta is 1 based coordinate - so need to + 1 to start for each row

* folder for outputs: /researchers/caitlin.page/cf_nano/run_metheor
```{r}
list.files("/researchers/caitlin.page/cf_nano/run_metheor")
```

* folder with rrbs bams: /researchers/caitlin.page/rrbs_encode_cell_line

```{r}
list.files("/researchers/caitlin.page/rrbs_encode_cell_line")[grep("bam",list.files("/researchers/caitlin.page/rrbs_encode_cell_line") )]
```

# PDR

DEFAULTS
-d, --min-depth: Minimum depth of CpG stretches to consider. [default: 10]
-p, --min-cpgs: Minimum number of consecutive CpGs in a CpG stretch to consider. [default: 10]
-q, --min-qual: Minimum quality for a read to be considered [default: 10]
-c, --cpg-set: (Optional) Specify a predefined set of CpGs (in BED file) to be analyzed.

OUTPUT
Produces a tab-separated table with the following six columns. Note that the coordinate system follows that of BED/BedGraph format (0-based, half-open), without header names.

chrom: Chromosome where the CpG exists.
start: 0-based position of the cytosine (C) in CpG
end: 0-based position + 1 of the guanine (G) in CpG
pdr: Value of PDR
n_c: Number of concordant reads (i.e., all the CpGs covered by the read have identical methylation states) supporting the CpG
n_d: Number of discordant reads (i.e., CpGs covered by the read have more than one methylation states)

```{r}
example <- read.table("/researchers/caitlin.page/cf_nano/run_metheor/pdr_ENCFF000LYO.tsv")
colnames(example) <- c("seqnames", "start", "end", "PDR", "n_c", "n_d")
example[1:10,]
```


# LPMD
DEFAULTS
-m, --min-distance: Minimum distance between CpG pairs to consider. [default: 2]
-M, --max-distance: Maximum distance between CpG pairs to consider. [default: 16]
-q, --min-qual: Minimum quality for a read to be considered. [default: 10]

metheor lpmd --input /researchers/caitlin.page/rrbs_encode_cell_line/ENCFF000LYO_bismark_bt2.bam --output /researchers/caitlin.page/phd/run_metheor/lpmd_ENCFF000LYO.tsv

OUTPUT
By default, it produces a simple tab-delimited table containing genomewide average lpmd values. It has two columns:

name: Filename
lpmd: LPMD values averaged for all CpG pairs participating in the analysis
Optionally, when --pairs flag is turned on, you can have a more detailed table with CpG pair-wise statistics for LPMD computation. Note that the size of the file can be large. There are six columns:

chrom: Chromosome where the CpG pair exists
cpg1: 0-based position of the cytosine (C) of the first CpG
cpg2: 0-based position of the cytosine (C) of the second CpG
lpmd: LPMD value. Computed as n_discordant / (n_concordant + n_discordant)
n_concordant: Number of sequencing reads supporting CpG pairs having the same methylation state
n_discordant: Number of sequencing reads supporting CpG pairs having different methylation states


```{r}
example <- read.table("/researchers/caitlin.page/cf_nano/run_metheor/lpmd_ENCFF000LYO.tsv")
example
```
* ah this is what you get without the pairs flag
(redo this one later)
* pairs requires a value and I don't understand - so we skip it for now

# MHL
DEFAULTS
-d, --min-depth: Minimum depth of reads covering epialleles to consider. [default: 10]
-p, --min-cpgs: Minimum number of consecutive CpGs in a CpG stretch to consider.
-q, --min-qual: Minimum quality for a read to be considered. [default: 10]

OUTPUT
Produces a (BedGraph-compatible) tab-separated table with the following four columns.

chrom: Chromosome where the CpG exists.
start: 0-based position of the cytosine (C) in CpG
end: 0-based position + 1 of the guanine (G) in CpG
mhl: Value of MHL

```{r}
example <- read.table("/researchers/caitlin.page/cf_nano/run_metheor/mhl_ENCFF000LYO.tsv")
colnames(example) <- c("seqnames", "start", "end", "MHL")
example[1:10,]
```


# epipolymorph
DEFAULTS
-d, --min-depth: Minimum depth of reads covering epialleles to consider. [default: 10]
-q, --min-qual: Minimum quality for a read to be considered. [default: 10]

OUTPUT
chrom: Chromosome where the CpG quartet exists.
cpg1: 0-based position of the cytosine (C) in the first CpG
cpg2: 0-based position of the cytosine (C) in the second CpG
cpg3: 0-based position of the cytosine (C) in the third CpG
cpg4: 0-based position of the cytosine (C) in the fourth CpG
pm: Value of PM

```{r}
example <- read.table("/researchers/caitlin.page/cf_nano/run_metheor/pm_ENCFF000LYO.tsv")
colnames(example) <- c("seqnames", "start", "start_2", "start_3", "start_4", "PM")
example[1:10,]
```


# entropy
DEFAULTS
-d, --min-depth: Minimum depth of reads covering epialleles to consider. [default: 10]
-q, --min-qual: Minimum quality for a read to be considered. [default: 10]

OUTPUT
chrom: Chromosome where the CpG quartet exists.
cpg1: 0-based position of the cytosine (C) in the first CpG
cpg2: 0-based position of the cytosine (C) in the second CpG
cpg3: 0-based position of the cytosine (C) in the third CpG
cpg4: 0-based position of the cytosine (C) in the fourth CpG
me: Value of ME
NOTE: The order of CpG quartets in the output file is not sorted.

```{r}
example <- read.table("/researchers/caitlin.page/cf_nano/run_metheor/me_ENCFF000LYO.tsv")
colnames(example) <- c("seqnames", "start", "start_2", "start_3", "start_4", "ME")
example[1:10,]
```


# fraction discordant read pairs
DEFAULTS
-q, --min-qual: Minimum quality for a read to be considered. [default: 10]
-d, --min-depth: Minimum depth of reads covering epialleles to consider. [default: 10]
-D, --max-depth: Maximum number of reads to consider. [default: 40]
-l, --min-overlap: Minimum overlap between two reads to consider in basepairs. [default: 35]

OUTPUT
Produces a (BedGraph-compatible) tab-separated table with the following four columns.

chrom: Chromosome where the CpG exists.
start: 0-based position of the cytosine (C) in CpG
end: 0-based position + 1 of the guanine (G) in CpG
fdrp: Value of FDRP

```{r}
example <- read.table("/researchers/caitlin.page/cf_nano/run_metheor/fdrp_ENCFF000LYO.tsv")
colnames(example) <- c("seqnames", "start", "end", "FDRP")
example[1:10,]
```


# quantitative fraction discordant reads

qFDRP is a soft-version of FDRP. Instead of deciding whether the read pair is discordant, qFDRP computes the normalized hamming distance. Note that when the methylation states of the common CpGs are completely same, normalized hamming distance will be 0, and when the methylation states are completely different, it will be 1, which makes qFDRP calculation equivalent to FDRP.

DEFAULTS
-i, --input: Path to input BAM file.
-o, --output: Path to output table file summarizing the result of epipolymorphism calculation.
-q, --min-qual: Minimum quality for a read to be considered. [default: 10]
-d, --min-depth: Minimum depth of reads covering epialleles to consider. [default: 10]
-D, --max-depth: Maximum number of reads to consider. [default: 40]
-l, --min-overlap: Minimum overlap between two reads to consider in basepairs. [default: 35]
-c, --cpg-set: (Optional) Specify a predefined set of CpGs (in BED file) to be analyzed.

OUTPUT
Produces a (BedGraph-compatible) tab-separated table with the following four columns.

chrom: Chromosome where the CpG exists.
start: 0-based position of the cytosine (C) in CpG
end: 0-based position + 1 of the guanine (G) in CpG
qfdrp: Value of qFDRP
```{r}
example <- read.table("/researchers/caitlin.page/cf_nano/run_metheor/qfdrp_ENCFF000LYO.tsv")
colnames(example) <- c("seqnames", "start", "end", "QFDRP")
example[1:10,]
```

