---
title: "otherReadStyleMethodsAnalysis"
author: "caitlinpage"
date: "2026-04-08"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

## Introduction

```{r}
library(annotatr)
library(data.table)
library(plyranges)
library(tidyr)
library(stringr)
library(dplyr)
library(ggplot2)
library(edgeR)
```


Description of methods [here](otherReadStyleMethods.html)

## Analysis

```{r}
for_beta_all <- readRDS("/researchers/caitlin.page/cf_nano/r_output/rrbs_alpha_beta.rds")
#for_beta <- readRDS("../output/cluster_files/rrbs_alpha_beta.rds")

#gm12878_2 <- readRDS("/researchers/caitlin.page/cf_nano/r_output/rrbs_alpha_beta_rep2.rds")

#k562 <- readRDS("/researchers/caitlin.page/cf_nano/r_output/k562_rep1.rds")

#k562_2 <- readRDS("/researchers/caitlin.page/cf_nano/r_output/k562_rep2.rds")
```


```{r}
files <- list.files("/researchers/caitlin.page/cf_nano/run_metheor")
files
files <- paste0("/researchers/caitlin.page/cf_nano/run_metheor/", files)
```


# MHL
```{r}
files[grep("mhl", files)]
```

* MHL 0 based coordinates, my alpha in 1-based coordinates
* so +1 to the start
```{r}
mhl_gm1 <- read.table(files[grep("mhl", files)][1])
colnames(mhl_gm1) <- c("seqnames", "start", "end", "MHL")
mhl_gm1$start <- mhl_gm1$start + 1

#mhl_gm2 <- read.table(files[grep("mhl", files)][2])
#colnames(mhl_gm2) <- c("seqnames", "start", "end", "MHL")

#mhl_k562_1 <- read.table(files[grep("mhl", files)][3])
#colnames(mhl_k562_1) <- c("seqnames", "start", "end", "MHL")

#mhl_k562_2 <- read.table(files[grep("mhl", files)][4])
#colnames(mhl_k562_2) <- c("seqnames", "start", "end", "MHL")
```

```{r}
mhl_gm1[1:10,]
nrow(mhl_gm1)
```

# THEORY: MHL is like a summarised region alpha

LOGIC
* MHL: how well meth haplotypes (consec CpGs) conserved throughout cell pop (across reads) for given region
* calculates a per cpg value
* 0 - all unmethylated, 1 - all methylated, in between - a mix
** the only metric that does this (through metheor - I think someone else modified one of these to distinguish)
* in this it's like alpha - differentiates between states
* and it's a population metric and a read metric 


## Correlation?

```{r}
for_beta_all %>% filter(seqnames == "chr1") %>% distinct(seqnames, start) %>% nrow()
mhl_gm1 %>% filter(seqnames == "chr1") %>% distinct(seqnames, start) %>% nrow()
```

```{r}
mhl_gm1 %>% filter(seqnames == "chr1") %>% distinct(seqnames, start) %>% .$start %in% distinct(filter(for_beta_all, seqnames == "chr1"), seqnames, start)$start %>% summary()
```

```{r}
cg_sites <- cbind(data.frame(Biostrings::matchPattern("CG", BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38[["chr1"]])), seqnames = "chr1") %>% 
  mutate(pos = paste0(seqnames, "-", start)) %>%
  relocate(pos, seqnames)
cg_sites$index <- 1:nrow(cg_sites)

mhl_gm1 %>% filter(seqnames == "chr1") %>% distinct(seqnames, start) %>% .$start %in% cg_sites$start %>% summary()
```

```{r}
for_beta_all %>% filter(seqnames == "chr1") %>% distinct(seqnames, start) %>% .$start %in% cg_sites$start %>% summary()
```

```{r}
mhl_gm1_subset <- mhl_gm1 %>% filter(seqnames == "chr1")
```


```{r}
for_beta_all %>% filter(seqnames == "chr1") %>% 
  group_by(seqnames, start) %>%
  summarise(mean_alpha = mean(alpha)) %>%
  ungroup() %>%
  mutate(mhl = mhl_gm1_subset[match(.$start, mhl_gm1_subset$start), "MHL"]) %>%
  ggplot(aes(x = mean_alpha, y = mhl)) +
  geom_point(alpha = 0.3)
```
```{r}
for_beta_all %>% filter(seqnames == "chr1") %>% 
  distinct(seqnames, start, beta) %>%
  mutate(mhl = mhl_gm1_subset[match(.$start, mhl_gm1_subset$start), "MHL"]) %>%
  ggplot(aes(x = beta, y = mhl)) +
  geom_point(alpha = 0.3)
```

```{r}
for_beta_all %>% filter(seqnames == "chr1") %>% 
  group_by(seqnames, start, beta) %>%
  summarise(mean_alpha = mean(alpha)) %>%
  ungroup() %>%
  mutate(mhl = mhl_gm1_subset[match(.$start, mhl_gm1_subset$start), "MHL"]) %>%
  ggplot(aes(x = mean_alpha, y = mhl, colour = beta)) +
  geom_point(alpha = 0.3)
```

```{r}
test_mhl <- for_beta_all %>% filter(seqnames == "chr1") %>% 
  group_by(seqnames, start, beta) %>%
  summarise(mean_alpha = mean(alpha)) %>%
  ungroup() %>%
  mutate(mhl = mhl_gm1_subset[match(.$start, mhl_gm1_subset$start), "MHL"])

cor.test(test_mhl$mean_alpha, test_mhl$beta)
cor.test(test_mhl$mean_alpha, test_mhl$mhl)
cor.test(test_mhl$beta, test_mhl$mhl)
```

## Q: MHL compared with my previous plots of grouped alphas

```{r}
mhl_gm1_subset[100:1000,] %>%
  ggplot(aes(x = start, y = MHL)) +
  geom_line()
```
```{r}
for_beta_all_subset <- for_beta_all %>% filter(seqnames == "chr1")
```


```{r}
just_beta <- for_beta_all %>% filter(seqnames == "chr1") %>% distinct(seqnames, start, beta) %>% group_by(start) %>% mutate(coverage = n()) %>% ungroup() %>% data.frame() %>% .[order(.$start),] %>%
  mutate(row_num = 1:nrow(.)) %>%
  mutate(beta_high = ifelse(beta >= 0.8 & beta < 1, "high", "other"))

grouped_betas <- just_beta %>% 
  group_by(beta_high) %>%
  mutate(consec = n()) %>%
  mutate(count = 1:n()) %>%
  ungroup() %>%
  mutate(length = sequence(rle(.$beta_high)$lengths)) %>% 
  mutate(group = cumsum(c(TRUE, diff(length) <= 0))) %>%
  group_by(group) %>%
  mutate(length = n(), av_coverage = mean(coverage), av_beta = mean(beta)) %>%
  ungroup() %>% filter(beta_high == "high", length > 2) %>% data.frame() 

poten_groups <- for_beta_all %>% filter(seqnames == "chr1") %>% mutate(beta_range_group_num = grouped_betas[match(.$start, grouped_betas$start), "group"],
                    beta_range_group_length = grouped_betas[match(.$start, grouped_betas$start), "length"]) %>% 
  filter(!is.na(beta_range_group_num)) %>% distinct(read_id, alpha, beta_range_group_num, beta_range_group_length) %>%
  mutate(beta_range_group_num = as.factor(beta_range_group_num)) %>%
  group_by(beta_range_group_num) %>%
  summarise(bimodal_p = list(diptest::dip.test(alpha)$p.value)
  ) %>%
  unnest(cols = c(bimodal_p)) %>% filter(bimodal_p < 0.01) %>% .$beta_range_group_num
```

```{r}
source("../code/processData.R")
plot_list <- list()
for (i in 1:5) {
  row_pos <- just_beta %>%
  group_by(beta_high) %>%
  mutate(consec = n()) %>%
  mutate(count = 1:n()) %>%
  ungroup() %>%
  mutate(length = sequence(rle(.$beta_high)$lengths)) %>% 
  mutate(group = cumsum(c(TRUE, diff(length) <= 0))) %>% filter(group == i)
  min_row <- min(row_pos$row_num)
  max_row <- max(row_pos$row_num)
  plot_list[[i]] <- rrbs_plot(for_beta_all_subset, for_beta_all_subset %>% filter(start %in% just_beta[min_row:max_row,]$start) %>% .$read_id %>% unique(), count_instead = TRUE)
}
plot_list
```
* arbitarirly find one because section with multiple reads
```{r}
mhl_gm1_subset %>% filter(start >= 136870) %>% .[1:10,]
```

```{r}
rrbs_plot(for_beta_all_subset, for_beta_all_subset %>% filter(start >= 136870, start <= 136920) %>% .$read_id %>% unique(), count_instead = TRUE) +
  geom_line(data = filter(mhl_gm1_subset, start >= 181000, start <= 181650), aes(x = start, y = MHL), alpha = 0.3)
```


```{r}
rrbs_plot(for_beta_all_subset, for_beta_all_subset %>% filter(start >= 181000, start <= 181650) %>% .$read_id %>% unique(), count_instead = TRUE) +
  geom_line(data = filter(mhl_gm1_subset, start >= 181000, start <= 181650), aes(x = start, y = MHL), alpha = 0.3)
```

```{r}
mhl_gm1 %>%
  ggplot(aes(x = log2(MHL + 0.1))) +
  geom_density()

mhl_gm1 %>%
  ggplot(aes(x = MHL)) +
  geom_histogram(bins = 75)
```

```{r}
for_beta_all %>% distinct(read_id, alpha) %>%
  ggplot(aes(x = alpha)) +
  geom_histogram(bins = 75)
```


```{r}
for_beta_all_subset %>% 
  group_by(seqnames, start, beta) %>%
  summarise(mean_alpha = mean(alpha)) %>%
  ungroup() %>%
  mutate(mhl = mhl_gm1_subset[match(.$start, mhl_gm1_subset$start), "MHL"],
         row_num = 1:n()) %>% .[1:10,]
```
```{r}
827532 - 827413	
```

```{r}
rrbs_plot(for_beta_all_subset, for_beta_all_subset %>% filter(start >= 827413, start <= 827532) %>% .$read_id %>% unique(), count_instead = TRUE) +
  geom_line(data = filter(mhl_gm1_subset, start >= 827413, start <= 827532), aes(x = start, y = MHL), alpha = 0.3, colour = "purple")
```

```{r}
for_beta_all_subset %>% 
  group_by(seqnames, start, beta) %>%
  summarise(mean_alpha = mean(alpha)) %>%
  ungroup() %>%
  mutate(mhl = mhl_gm1_subset[match(.$start, mhl_gm1_subset$start), "MHL"],
         row_num = 1:n()) %>% 
  filter(!is.na(mhl)) %>%
  filter(mhl > 0.2) %>% .[1:10,]
```

```{r}
for_beta_all_subset %>% 
  group_by(seqnames, start, beta) %>%
  summarise(mean_alpha = mean(alpha)) %>%
  ungroup() %>%
  mutate(mhl = mhl_gm1_subset[match(.$start, mhl_gm1_subset$start), "MHL"],
         row_num = 1:n()) %>% .[145:155,]
```

```{r}
for_beta_all_subset_sum <- for_beta_all_subset %>% 
  group_by(seqnames, start, beta) %>%
  summarise(mean_alpha = mean(alpha)) %>%
  ungroup() %>%
  mutate(mhl = mhl_gm1_subset[match(.$start, mhl_gm1_subset$start), "MHL"],
         row_num = 1:n())
```


```{r}
rrbs_plot(for_beta_all_subset, for_beta_all_subset %>% filter(start >= 909867, start <= 909959) %>% .$read_id %>% unique(), count_instead = TRUE) +
  geom_line(data = filter(mhl_gm1_subset, start >= 909867, start <= 909959), aes(x = start, y = MHL), alpha = 0.3, colour = "purple") +
  geom_line(data = filter(for_beta_all_subset_sum, start >= 909867, start <= 909959), aes(x = start, y = mean_alpha), alpha = 0.3, colour = "orange")
```
* row 494-498
* high beta + mean alpha, lower mhl
```{r}
for_beta_all_subset_sum[494:498,]
rrbs_plot(for_beta_all_subset, for_beta_all_subset %>% filter(start >= 959748, start <= 959777) %>% .$read_id %>% unique(), count_instead = TRUE) +
  geom_line(data = filter(mhl_gm1_subset, start >= 959748, start <= 959777), aes(x = start, y = MHL), alpha = 0.3, colour = "purple") +
  geom_line(data = filter(for_beta_all_subset_sum, start >= 959748, start <= 959777), aes(x = start, y = mean_alpha), alpha = 0.3, colour = "orange")
```

* when mean_alpha = 1
** seems to be a bigger distribution of mhl

```{r}
for_beta_all_subset_sum %>% filter(mean_alpha == 1, !is.na(mhl)) %>%
  ggplot(aes(x = as.factor(mean_alpha), y = mhl)) +
  geom_violin()

for_beta_all_subset_sum %>% filter(mean_alpha == 0, !is.na(mhl)) %>%
  ggplot(aes(x = as.factor(mean_alpha), y = mhl)) +
  geom_violin()
```



```{r}
mhl_gm1_subset[1:10,]
```




# FDRP

```{r}
fdrp_gm1 <- read.table("/researchers/caitlin.page/cf_nano/run_metheor/fdrp_ENCFF000LYO.tsv")
colnames(fdrp_gm1) <- c("seqnames", "start", "end", "FDRP")
fdrp_gm1 <- fdrp_gm1 %>% filter(seqnames == "chr1")
fdrp_gm1$start <- fdrp_gm1$start + 1
```

```{r}
fdrp_gm1 %>% filter(seqnames == "chr1") %>%
  ggplot(aes(x = FDRP)) +
  geom_density()
```
```{r}
fdrp_gm1[1:10,]
```

```{r}
for_beta_all_subset_sum %>% mutate(fdrp = fdrp_gm1[match(.$start, fdrp_gm1$start), "FDRP"]) %>%
  ggplot(aes(x = mean_alpha, y = fdrp)) +
  geom_point(alpha = 0.3)
```

```{r}
for_beta_all_subset_sum %>% mutate(fdrp = fdrp_gm1[match(.$start, fdrp_gm1$start), "FDRP"]) %>%
  ggplot(aes(x = beta, y = fdrp)) +
  geom_point(alpha = 0.3)
```

```{r}
for_beta_all_subset_sum %>% mutate(fdrp = fdrp_gm1[match(.$start, fdrp_gm1$start), "FDRP"]) %>%
  ggplot(aes(x = mhl, y = fdrp)) +
  geom_point(alpha = 0.3)
```

* well these plots mean nothing
* because fdrp measures something different


# entropy

```{r}
entropy_gm1 <- read.table("/researchers/caitlin.page/cf_nano/run_metheor/me_ENCFF000LYO.tsv")
colnames(entropy_gm1) <- c("seqnames", "start", "start_2", "start_3", "start_4", "ME")
entropy_gm1$start <- entropy_gm1$start + 1
entropy_gm1$start_2 <- entropy_gm1$start_2 + 1
entropy_gm1$start_3 <- entropy_gm1$start_3 + 1
entropy_gm1$start_4 <- entropy_gm1$start_4 + 1

entropy_gm1 <- entropy_gm1 %>% filter(seqnames == "chr1") %>% .[order(.$start),]
```

```{r}
entropy_gm1 %>%
  ggplot(aes(x = ME)) +
  geom_density()
```

```{r}
for_beta_all_subset_sum %>% mutate(entropy = entropy_gm1[match(.$start, entropy_gm1$start), "ME"]) %>%
  ggplot(aes(x = mean_alpha, y = entropy)) +
  geom_point(alpha = 0.3)
```

```{r}
for_beta_all_subset_sum %>% mutate(fdrp = fdrp_gm1[match(.$start, fdrp_gm1$start), "FDRP"]) %>% 
  mutate(entropy = entropy_gm1[match(.$start, entropy_gm1$start), "ME"]) %>%
  ggplot(aes(x = fdrp, y = entropy)) +
  geom_point(alpha = 0.3)
```
```{r}
for_beta_all_subset_sum <- for_beta_all_subset_sum %>%
  mutate(var_alpha = for_beta_all_subset %>% group_by(start) %>% summarise(var_alpha = var(alpha)) %>% .$var_alpha) %>%
  mutate(fdrp = fdrp_gm1[match(.$start, fdrp_gm1$start), "FDRP"]) %>% 
  mutate(entropy = entropy_gm1[match(.$start, entropy_gm1$start), "ME"])
```

```{r}
for_beta_all_subset_sum %>% 
  ggplot(aes(x = var_alpha, y = entropy)) +
  geom_point(alpha = 0.3)

for_beta_all_subset_sum %>% 
  ggplot(aes(x = var_alpha, y = fdrp)) +
  geom_point(alpha = 0.3)
```

```{r}
cor.test(for_beta_all_subset_sum$var_alpha, for_beta_all_subset_sum$entropy)
cor.test(for_beta_all_subset_sum$var_alpha, for_beta_all_subset_sum$fdrp)
cor.test(for_beta_all_subset_sum$entropy, for_beta_all_subset_sum$fdrp)
```

# compare variation to DTU??
* entropy and fdrp are both variation measurements
* so what does it look like in dtu?

```{r}
ts.gene.gene <- readRDS("/researchers/caitlin.page/cf_nano/r_output/ts.gene.gene.rds")
```

```{r}
annots <- c("hg38_basicgenes", "hg38_genes_intergenic")

annotations <- annotatr::build_annotations(genome = 'hg38', annotations = annots)
seqs <- annotations %>% data.frame() %>% distinct(seqnames) %>% .[1:22,]
annotations <- annotations %>% data.frame() %>% filter(seqnames %in% seqs)

for_beta_all_genes <- find_overlaps(as_granges(for_beta_all), as_granges(annotations)) %>% data.frame()
```

* luckily 1 of the top 10 genes is on chr1
```{r}
dtu_gene <- for_beta_all_genes %>% filter(symbol %in% ts.gene.gene[1:10,]$gene_id) %>% filter(seqnames == "chr1") %>% distinct(seqnames, start, beta, symbol)
```

```{r}
for_beta_all_subset_sum %>% filter(start %in% dtu_gene$start) %>%
  ggplot(aes(x = start, y = entropy)) +
  geom_point()
```


```{r}
rrbs_plot(for_beta_all_subset, for_beta_all_subset %>% filter(start >= 29124612, start <= 29125168) %>% .$read_id %>% unique(), count_instead = TRUE) +
  geom_line(data = filter(mhl_gm1_subset, start >= 29124612, start <= 29125168), aes(x = start, y = MHL), alpha = 0.3, colour = "purple") +
  geom_line(data = filter(for_beta_all_subset_sum, start >= 29124612, start <= 29125168), aes(x = start, y = mean_alpha), alpha = 0.3, colour = "orange") +
  geom_line(data = filter(for_beta_all_subset_sum, start >= 29124612, start <= 29125168), aes(x = start, y = entropy), alpha = 0.3, colour = "red") +
  geom_line(data = filter(for_beta_all_subset_sum, start >= 29124612, start <= 29125168), aes(x = start, y = fdrp), alpha = 0.3, colour = "blue")
```




```{r}
rrbs_plot(for_beta_all_subset, for_beta_all_subset %>% filter(start >= 29124612, start <= 29125168) %>% .$read_id %>% unique(), count_instead = TRUE) +
  geom_line(data = filter(mhl_gm1_subset, start >= 29124612, start <= 29125168), aes(x = start, y = MHL), alpha = 0.3, colour = "purple") +
  geom_line(data = filter(for_beta_all_subset_sum, start >= 29124612, start <= 29125168), aes(x = start, y = mean_alpha), alpha = 0.3, colour = "orange") +
  geom_line(data = filter(for_beta_all_subset_sum, start >= 29124612, start <= 29125168), aes(x = start, y = entropy), alpha = 0.3, colour = "red") +
  geom_line(data = filter(for_beta_all_subset_sum, start >= 29124612, start <= 29125168), aes(x = start, y = fdrp), alpha = 0.3, colour = "blue") +
  geom_line(data = filter(for_beta_all_subset_sum, start >= 29124612, start <= 29125168), aes(x = start, y = var_alpha), alpha = 0.3, colour = "yellow") +
  scale_x_continuous(limits = c(29124600, 29125000)) +
  scale_y_continuous(limits = c(-0.1, 0.4))
```

