---
title: "patToRead"
author: "caitlinpage"
date: "2026-05-04"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r}
library(annotatr)
library(data.table)
library(plyranges)
library(tidyr)
library(stringr)
library(dplyr)
library(ggplot2)
```

# try to reverse the read info from the pat file
* wgbs_tools has the reference available in hg19
* and an ilmntoCpG file for hg38 - I think that means illumina file to cpg 
* I think I just need the annotation file

* so hg19

## load in files
```{r}
anno <- fread("/researchers/caitlin.page/cf_nano/wgbs_tools_ref/hg19.annotations.bed.gz")
colnames(anno) <- c("seqnames", "start", "end", "feature", "gene")
```
```{r}
anno_cg <- fread("/researchers/caitlin.page/cf_nano/wgbs_tools_ref/hg19.ilmn2CpG.tsv.gz")

```
* damn - that's the microarray stuff - that's not what I want

* and I'm really struggling to build the reference in the terminal with the program
* also conda run_wgbs_tools
* and I do have wgbs_tools installed
and ./wgbstools so it will run
mkm

```{r}
pat <- fread("/researchers/caitlin.page/atlas/GSM5652302_Blood-Monocytes-Z000000TP.pat.gz")
colnames(pat) <- c("seqnames", "index", "pattern", "count")
```

```{r}
pat_1 <- pat %>% filter(seqnames == "chr1")
```
IDEA
* pattern is sequence - get those indexes
and turn it into like my cg_sites
and see if numbers vaguely line up
asked perplexity for help with code


```{r}
pat_1 <- pat_1 %>% mutate(pattern_length = nchar(pattern))
```

abcde
```{r}
pat_1$pat_index <- mapply(
  function(index, pattern_length) seq(from = index, length.out = pattern_length),
  pat_1$index,
  pat_1$pattern_length,
  SIMPLIFY = FALSE
)
```

```{r}
pat_1 <- pat_1 %>% mutate(pat_index2 = Map(function(index, pattern_length) seq(index, length.out = pattern_length), index, pattern_length)) %>% unnest(pat_index2)
```

```{r}
pat_1_cg <- pat_1 %>% group_by(seqnames, pat_index2) %>% summarise(coverage = n())
```

```{r}
cg_sites %>% filter(seqnames == "chr1")
```


* so it worked (eventually)
* but only up to a point
* because again I need the reference for the positions - because I don't know when things are supposed to not line up

```{r}
pat_1_cg %>% mutate(gap = lead(pat_index2 - pat_index2)) %>% distinct(gap)
```
* and my last idea is bust
* I thought if it had gaps - then that could work around the mismatches if I treated the refernce cpgs same way


* other workaround attempt - download the bed files to get genomic coordinates - only for positions with beta vals anyway
