---
title: "patToRead"
author: "caitlinpage"
date: "2026-05-04"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r}
library(annotatr)
library(data.table)
library(plyranges)
library(tidyr)
library(stringr)
library(dplyr)
library(ggplot2)
```

# try to reverse the read info from the pat file
* wgbs_tools has the reference available in hg19
* and an ilmntoCpG file for hg38 - I think that means illumina file to cpg 
* I think I just need the annotation file

* so hg19

## load in files
```{r}
anno <- fread("/researchers/caitlin.page/cf_nano/wgbs_tools_ref/hg19.annotations.bed.gz")
colnames(anno) <- c("seqnames", "start", "end", "feature", "gene")
```
```{r}
anno_cg <- fread("/researchers/caitlin.page/cf_nano/wgbs_tools_ref/hg19.ilmn2CpG.tsv.gz")

```
* damn - that's the microarray stuff - that's not what I want

* and I'm really struggling to build the reference in the terminal with the program
* also conda run_wgbs_tools
* and I do have wgbs_tools installed
and ./wgbstools so it will run
mkm

```{r}
pat <- fread("/researchers/caitlin.page/atlas/GSM5652302_Blood-Monocytes-Z000000TP.pat.gz")
colnames(pat) <- c("seqnames", "index", "pattern", "count")
```

