---
title: "plotDmrDAU"
author: "caitlinpage"
date: "2026-09-03"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r}
library(plyranges)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
```


## compare the tile positions
```{r}
dau_basictiles <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dtu_basictiles_emseq_cancer_healthy.rds")

dau_optiles <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_emseq_cancer_healthy.rds")

dmr_emseq <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dmr_emseq_strict.rds")
dmr_emseq <- dmr_emseq %>% .[order(.$min_smoothed_fdr),] %>% mutate(dmr_rank = 1:n())
```



```{r}
genes_txdb <- genes(TxDb.Hsapiens.UCSC.hg38.knownGene)
```

```{r}
basic_tiles_genes <- find_overlaps(as_granges(dau_basictiles), genes_txdb) %>% data.frame()
op_tiles_genes <- find_overlaps(as_granges(dau_optiles), genes_txdb) %>% data.frame()
dmr_genes <- find_overlaps(as_granges(dmr_emseq), genes_txdb) %>% data.frame()
```


```{r}
ggVennDiagram::ggVennDiagram(list(basic = basic_tiles_genes$gene_id, optiles = op_tiles_genes$gene_id, dmr = dmr_genes$gene_id))
```

```{r}
ggVennDiagram::ggVennDiagram(list(basic = filter(basic_tiles_genes, FDR < 0.05)$gene_id, optiles = filter(op_tiles_genes, FDR < 0.05)$gene_id, dmr = dmr_genes$gene_id))
```


# repeat the compareTilingMethods genes analysis coz why not
```{r libraries}
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(org.Hs.eg.db)
library(GenomicFeatures)
library(AnnotationDbi)
library(clusterProfiler)
library(enrichplot)
library(plyranges)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(DT)
library(pheatmap)
library(ggrepel)
library(ComplexUpset)
library(RColorBrewer)
 
txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene
```
 
---
 
# Gene lists  ← edit here
 
Add or remove lists freely. Names become labels throughout the report.
All IDs are **Entrez** (character strings).
 
```{r gene-lists}
gene_lists <- list(
  op_genes = filter(op_tiles_genes, FDR < 0.05) %>% distinct(gene_id) %>% .$gene_id,  
  basic_genes = filter(basic_tiles_genes, FDR < 0.05) %>% distinct(gene_id) %>% .$gene_id,
  dmr_genes = dmr_genes %>% distinct(gene_id) %>% .$gene_id)
 
# Palette — one colour per list (extend if you have more than 8 lists)
list_colours <- setNames(
  brewer.pal(max(4, length(gene_lists)), "Set2")[seq_along(gene_lists)],
  names(gene_lists)
)
gene_lists
```
 
 
 
```{r helpers, include=FALSE}
# Retrieve gene symbols for a vector of Entrez IDs
entrez_to_symbol <- function(ids) {
  AnnotationDbi::select(
    org.Hs.eg.db,
    keys    = ids,
    columns = c("SYMBOL", "GENENAME"),
    keytype = "ENTREZID"
  )
}
 
# All unique IDs across lists
all_ids <- unique(unlist(gene_lists))
all_anno <- entrez_to_symbol(all_ids)
```
 
---
 
# 1  List overview
 
```{r list-overview}
overview <- purrr::imap_dfr(gene_lists, function(ids, name) {
  tibble(
    List        = name,
    n_genes     = length(ids),
    genes       = paste(
      all_anno$SYMBOL[match(ids, all_anno$ENTREZID)],
      collapse = ", "
    )
  )
})
 
knitr::kable(overview, caption = "Gene list contents")
```
 
---
 
# 2  Pairwise similarity
 
Two complementary metrics are shown:
 
- **Jaccard index** — intersection / union (penalises size differences)
- **Overlap coefficient** — intersection / min(|A|, |B|) (asks "is one list a subset of the other?")
 
```{r similarity}
list_names <- names(gene_lists)
n <- length(list_names)
 
jaccard_mat  <- matrix(NA_real_, n, n, dimnames = list(list_names, list_names))
overlap_mat  <- matrix(NA_real_, n, n, dimnames = list(list_names, list_names))
 
for (i in seq_len(n)) {
  for (j in seq_len(n)) {
    a <- gene_lists[[i]]
    b <- gene_lists[[j]]
    isect <- length(intersect(a, b))
    jaccard_mat[i, j] <- isect / length(union(a, b))
    overlap_mat[i, j] <- isect / min(length(a), length(b))
  }
}
 
# Long form for ggplot
to_long <- function(mat, metric) {
  as.data.frame(mat) %>%
    tibble::rownames_to_column("List_A") %>%
    pivot_longer(-List_A, names_to = "List_B", values_to = metric)
}
 
sim_long <- left_join(
  to_long(jaccard_mat,  "Jaccard"),
  to_long(overlap_mat,  "Overlap"),
  by = c("List_A", "List_B")
)
 
datatable(
  sim_long %>% mutate(across(c(Jaccard, Overlap), \(x) round(x, 3))),
  caption  = "Pairwise similarity metrics",
  rownames = FALSE,
  filter   = "top",
  options  = list(pageLength = 20)
)
```
 
## Jaccard heatmap
 
```{r jaccard-heatmap, fig.height=4}
pheatmap(
  jaccard_mat,
  display_numbers = TRUE,
  number_format   = "%.2f",
  color           = colorRampPalette(c("white", "#2166ac"))(50),
  main            = "Jaccard similarity",
  fontsize         = 12,
  angle_col        = 45
)
```
 
## Overlap coefficient heatmap
 
```{r overlap-heatmap, fig.height=4}
pheatmap(
  overlap_mat,
  display_numbers = TRUE,
  number_format   = "%.2f",
  color           = colorRampPalette(c("white", "#b2182b"))(50),
  main            = "Overlap coefficient (intersection / min set size)",
  fontsize         = 12,
  angle_col        = 45
)
```
 
---
 
# 3  Gene membership (UpSet plot)
 
UpSet plots scale well beyond what Venn diagrams can show cleanly.
Each bar represents a specific combination of lists.
 
```{r upset, fig.height=5}
# Binary membership matrix
membership <- purrr::map_dfc(gene_lists, function(ids) {
  as.integer(all_ids %in% ids)
}) %>%
  mutate(gene = all_ids) %>%
  left_join(all_anno %>% select(ENTREZID, SYMBOL),
            by = c("gene" = "ENTREZID")) %>%
  select(SYMBOL, everything(), -gene)
 
upset_mat <- membership %>%
  select(all_of(list_names)) %>%
  as.data.frame()
 
ComplexUpset::upset(
  upset_mat,
  intersect        = list_names,
  name             = "Gene list combination",
  min_size         = 1,
  width_ratio      = 0.2,
  base_annotations = list(
    "Intersection size" = intersection_size(text = list(size = 3.5))
  ),
  themes = upset_modify_themes(list("default" = theme_minimal(base_size = 11)))
)
```
 
## Genes per intersection
 
```{r intersection-table}
membership %>%
  mutate(
    combination = apply(select(., all_of(list_names)), 1,
                        function(r) paste(list_names[as.logical(r)], collapse = " ∩ "))
  ) %>%
  group_by(combination) %>%
  summarise(
    n_genes = dplyr::n(),
    genes   = paste(SYMBOL, collapse = ", "),
    .groups = "drop"
  ) %>%
  arrange(desc(n_genes)) %>%
  datatable(
    caption  = "Genes in each list combination",
    rownames = FALSE,
    filter   = "top",
    options  = list(pageLength = 20, scrollX = TRUE)
  )
```
 
---
 
# 4  Comparative enrichment
 
`clusterProfiler::compareCluster()` runs ORA on each list simultaneously,
enabling direct visual comparison.
 
## GO Biological Process
 
```{r compare-go-bp, fig.height=8}
# Build a data frame of gene + group for compareCluster
gene_group_df <- purrr::imap_dfr(gene_lists, function(ids, name) {
  tibble(ENTREZID = ids, group = name)
})
 
cmp_go <- compareCluster(
  ENTREZID ~ group,
  data          = gene_group_df,
  fun           = "enrichGO",
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENTREZID",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.20,
  readable      = TRUE
)
 
if (!is.null(cmp_go) && nrow(as.data.frame(cmp_go)) > 0) {
  dotplot(cmp_go, showCategory = 15, title = "GO BP — comparison across lists") +
    theme(axis.text.y = element_text(size = 9),
          legend.position = "right")
} else {
  message("No significant GO BP terms found across any list.")
}
```
 
```{r compare-go-table}
if (!is.null(cmp_go) && nrow(as.data.frame(cmp_go)) > 0) {
  as.data.frame(cmp_go) %>%
    select(Cluster, Description, GeneRatio, BgRatio, p.adjust, geneID) %>%
    datatable(
      caption  = "GO BP enrichment per list",
      rownames = FALSE,
      filter   = "top",
      options  = list(pageLength = 20, scrollX = TRUE)
    )
}
```
 
## KEGG pathways
 
```{r compare-kegg, fig.height=7}
cmp_kegg <- compareCluster(
  ENTREZID ~ group,
  data          = gene_group_df,
  fun           = "enrichKEGG",
  organism      = "hsa",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.20
)
 
if (!is.null(cmp_kegg) && nrow(as.data.frame(cmp_kegg)) > 0) {
  cmp_kegg <- setReadable(cmp_kegg, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")
  dotplot(cmp_kegg, showCategory = 15, title = "KEGG — comparison across lists") +
    theme(axis.text.y = element_text(size = 9),
          legend.position = "right")
} else {
  message("No significant KEGG pathways found across any list.")
}
```
 
## Enrichment heatmap (shared terms)
 
Shows which GO BP terms are significant in which lists.
 
```{r enrich-heatmap, fig.height=7}
if (!is.null(cmp_go) && nrow(as.data.frame(cmp_go)) > 0) {
  go_wide <- as.data.frame(cmp_go) %>%
    select(Cluster, Description, p.adjust) %>%
    mutate(neg_log10_padj = -log10(p.adjust)) %>%
    pivot_wider(
      id_cols     = Description,
      names_from  = Cluster,
      values_from = neg_log10_padj,
      values_fill = 0
    ) %>%
    tibble::column_to_rownames("Description")
 
  # Keep top 30 terms by max score across lists
  top_terms <- names(sort(apply(go_wide, 1, max), decreasing = TRUE))[1:min(30, nrow(go_wide))]
  go_wide   <- go_wide[top_terms, , drop = FALSE]
 
  pheatmap(
    go_wide,
    color        = colorRampPalette(c("#f7f7f7", "#4393c3", "#053061"))(50),
    main         = "GO BP enrichment across lists\n(-log10 adjusted p-value)",
    angle_col    = 45,
    fontsize_row = 8,
    fontsize_col = 11,
    border_color = NA,
    na_col       = "grey95"
  )
}
```
 
---
 
# 5  Chromosomal distribution
 
```{r genomic-locs}
# Pull locations for all genes
all_ranges <- genes(txdb, filter = list(gene_id = all_ids))
 
loc_all <- as.data.frame(all_ranges) %>%
  rename(entrezid = gene_id, chr = seqnames,
         start_bp = start, end_bp = end) %>%
  left_join(all_anno, by = c("entrezid" = "ENTREZID")) %>%
  left_join(
    purrr::imap_dfr(gene_lists, function(ids, nm) tibble(entrezid = ids, list = nm)),
    by = "entrezid"
  ) %>%
  filter(!is.na(list))
 
chr_order <- paste0("chr", c(1:22, "X", "Y"))
loc_all   <- loc_all %>%
  filter(chr %in% chr_order) %>%
  mutate(chr = factor(chr, levels = rev(chr_order)),
         list = factor(list, levels = names(gene_lists)))
```
 
## Strip plot — all lists together
 
```{r strip-plot, fig.height=7}
ggplot(loc_all, aes(x = start_bp / 1e6, y = chr,
                    colour = list, shape = list)) +
  geom_point(size = 3, alpha = 0.8,
             position = position_jitter(height = 0.15, seed = 42)) +
  geom_text_repel(aes(label = SYMBOL), size = 2.5, max.overlaps = 15,
                  show.legend = FALSE) +
  scale_colour_manual(values = list_colours) +
  scale_shape_manual(values = c(16, 17, 15, 18, 8, 9)[seq_along(gene_lists)]) +
  labs(title    = "Gene positions across chromosomes (hg38)",
       x        = "Position (Mb)",
       y        = NULL,
       colour   = "List",
       shape    = "List") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```
 
## Per-list faceted view
 
```{r facet-strip, fig.height=7}
ggplot(loc_all, aes(x = start_bp / 1e6, y = chr, colour = list)) +
  geom_point(size = 2.5, alpha = 0.9) +
  geom_text_repel(aes(label = SYMBOL), size = 2.2, max.overlaps = 10,
                  show.legend = FALSE) +
  scale_colour_manual(values = list_colours) +
  facet_wrap(~list, ncol = 1) +
  labs(title  = "Gene positions — faceted by list",
       x      = "Position (Mb)",
       y      = NULL,
       colour = "List") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none",
        strip.text      = element_text(face = "bold", size = 11))
```
 
## Chromosome enrichment — which chromosomes are over-represented?
 
```{r chr-bar, fig.height=5}
loc_all %>%
  dplyr::count(list, chr) %>%
  ggplot(aes(x = chr, y = n, fill = list)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = list_colours) +
  labs(title = "Gene count per chromosome by list",
       x = NULL, y = "Gene count", fill = "List") +
  coord_flip() +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```



```{r}
cmp_go@compareClusterResult
```

