---
title: "processEmseq"
author: "caitlinpage"
date: "2026-07-09"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

## Introduction

```{r}
library(BSgenome.Hsapiens.UCSC.hg38)
#library(data.table)
library(plyranges)
library(tidyr)
library(stringr)
library(dplyr)
library(ggplot2)
library(OpTiles)
library(edgeR)
```

folder: /researchers/caitlin.page/cf_nano/fastq_tothill_emseq_healthy

CpG_OT_PRJ240785_LPRJ240785_S1_L004_R1_001_bismark_bt2_pe.txt

```{r}
bis <- fread("/researchers/caitlin.page/cf_nano/fastq_tothill_emseq_healthy/CpG_OT_PRJ240785_LPRJ240785_S1_L004_R1_001_bismark_bt2_pe.txt")
```

* good it has same format

```{r}
process_wgbs_short <- function(bismark, bed = NULL) {
  read <- bismark
  names(read) <- c("read_id", "strand", "seqnames", "start", "meth_status")
  read$end <- read$start

  read_overlap <- read
  read_overlap[, meth_status := ifelse(meth_status == "Z", "M", "U")]
  read_overlap[, meth_pattern := paste0(meth_status, collapse = ""), by = read_id]
  read_overlap[, num_meth := str_count(meth_pattern, pattern = "M")]
  read_overlap[, total := str_length(meth_pattern)]
  read_overlap[, alpha := num_meth/total]
  read_overlap[, genom_positions := paste0(start, collapse = ","), by = read_id]

  read_overlap <- read_overlap %>% .[order(.$read_id),]

  read_overlap[, read_start := min(start), by = read_id]
  read_overlap[, read_end := max(start), by = read_id]

  read_overlap
}
```
nmn
```{r}
em_healthy_785 <- process_wgbs_short(bis)
saveRDS(em_healthy_785, "/researchers/caitlin.page/cf_nano/r_output/emseq_healthy_785.rds")
```


```{r}
em_healthy_785
```

```{r}
just_read <- em_healthy_785 %>% distinct(read_id, seqnames, read_start, read_end, meth_pattern, num_meth, total, alpha, genom_positions)
```

```{r}
just_read %>%
  ggplot(aes(x = alpha)) +
  geom_density()
```

```{r}
just_read %>%
  ggplot(aes(x = total)) +
  geom_histogram()
```


```{r}
just_read %>% mutate(width = read_end - read_start) %>%
  ggplot(aes(x = width)) +
  geom_density()
```

```{r}
just_read %>% mutate(width = read_end - read_start) %>%
  ggplot(aes(x = log2(width))) +
  geom_density()
```


```{r}
1+1
```


```{r}
source("../processData.R")

bis
em_healthy_785 <- process_wgbs(bis)
saveRDS(em_healthy_785, "/researchers/caitlin.page/cf_nano/r_output/emseq_healthy_785.rds")
```

CpG_OT_PRJ240786_LPRJ240786_S2_L004_R1_001_bismark_bt2_pe.txt

CpG_OT_PRJ241037_LPRJ241037_S20_L004_R1_001_bismark_bt2_pe.txt
mnmnmn
* others are still going


```{r}
source("../processData.R")

bis <- fread("/researchers/caitlin.page/cf_nano/fastq_tothill_emseq_healthy/CpG_OT_PRJ240786_LPRJ240786_S2_L004_R1_001_bismark_bt2_pe.txt")
em_healthy_786 <- process_wgbs(bis)
saveRDS(em_healthy_786, "/researchers/caitlin.page/cf_nano/r_output/emseq_healthy_786.rds")
```


```{r}
source("../processData.R")

bis <- fread("/researchers/caitlin.page/cf_nano/fastq_tothill_emseq_healthy/CpG_OT_PRJ241037_LPRJ241037_S20_L004_R1_001_bismark_bt2_pe.txt")
em_healthy_037 <- process_wgbs(bis)
saveRDS(em_healthy_037, "/researchers/caitlin.page/cf_nano/r_output/emseq_healthy_037.rds")
```





## run this 

```{r}
process_wgbs_short <- function(bismark, bed = NULL) {
  read <- bismark
  names(read) <- c("read_id", "strand", "seqnames", "start", "meth_status")
  read$end <- read$start

  read_overlap <- read
  read_overlap[, meth_status := ifelse(meth_status == "Z", "M", "U")]
  read_overlap[, meth_pattern := paste0(meth_status, collapse = ""), by = read_id]
  read_overlap[, num_meth := str_count(meth_pattern, pattern = "M")]
  read_overlap[, total := str_length(meth_pattern)]
  read_overlap[, alpha := num_meth/total]
  read_overlap[, genom_positions := paste0(start, collapse = ","), by = read_id]

  read_overlap <- read_overlap %>% .[order(.$read_id),]

  read_overlap[, read_start := min(start), by = read_id]
  read_overlap[, read_end := max(start), by = read_id]

  read_overlap
}
```


```{r}
# process files

em_healthy_785 <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy_785.rds")
em_healthy_786 <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy_786.rds")
em_healthy_787 <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy_787.rds")
em_healthy_789 <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy_789.rds")
em_healthy_036 <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy_036.rds")
em_healthy_037 <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy_037.rds")
```


nmn
```{r}
# cpgs
seqs <- seqnames(BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38)[1:22]

cg_sites <- lapply(seqs, function(x) {
  cbind(data.frame(Biostrings::matchPattern("CG", BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38[[x]])),
        seqnames = x
  )}) %>% bind_rows()
```

```{r}
data.frame(em_healthy_036)[,cols]
```

```{r}
em_healthy_036 <- data.frame(em_healthy_036)
em_healthy_037 <- data.frame(em_healthy_037)
em_healthy_785 <- data.frame(em_healthy_785)
em_healthy_786 <- data.frame(em_healthy_786)
em_healthy_787 <- data.frame(em_healthy_787)
em_healthy_789 <- data.frame(em_healthy_789)
```

```{r}
em_healthy_785$sample <- "a1"
em_healthy_036$sample <- "a2"
em_healthy_787$sample <- "a3"
em_healthy_786$sample <- "b1"
em_healthy_037$sample <- "b2"
em_healthy_789$sample <- "b3"
```


```{r}
# combine samples

cols <- c("seqnames", "read_start", "read_end", "strand", "read_id", "sample", "meth_pattern", "num_meth", "total", "alpha", "genom_positions")

all_samples <- rbind(em_healthy_785[,cols],
                     em_healthy_036[,cols],
                     em_healthy_787[,cols],
                     em_healthy_786[,cols],
                     em_healthy_037[,cols],
                     em_healthy_789[,cols])
colnames(all_samples)[2:3] <- c("start", "end")

saveRDS(all_samples, "/researchers/caitlin.page/cf_nano/r_output/emseq_all_healthy.rds")
```
```{r}
all_samples <- readRDS("/researchers/caitlin.page/cf_nano/r_output/emseq_all_healthy.rds")
```

```{r}
all_samples %>% distinct(seqnames)
```

```{r}
all_samples <- all_samples %>% filter(seqnames %in% seqs)
all_samples <- distinct(all_samples)
```


```{r}
# make tiles
cols <- c("seqnames", "start")
for_op <- all_samples[,cols]
colnames(for_op)[1] <- "chr"
for_op$end <- for_op$start
for_op$strand <- "*"
for_op <- for_op %>% group_by(chr, start, end, strand) %>% summarise(coverage = n()) %>% ungroup()
for_op$numCs <- for_op$coverage - 1
for_op$numTs <- for_op$coverage - for_op$numCs
```



```{r}
all_samples %>% group_by(seqnames, start)
```


```{r}
mr <- new(
  "methylRaw",
  for_op,
  sample.id = "sample1",
  assembly = "hg38",
  context = "CpG",
  resolution = "base"
)

length_region <- 500
```
nnmnnmn
crazy that it's actually still running

```{r}
tiles <- methylKit::tileMethylCounts(mr,win.size=length_region,step.size=length_region,cov.bases = 0)

tilesdf_all <- methylKit::getData(tiles)
```

```{r}
mr
```

```{r}
tiles
```

```{r}
unique(as.character(tiles$chr))
```


```{r}
library(parallel)
```

```{r}
rownames(for_op)
```

* fix from claude
** basically run in // for diff chromosomes

```{r}
detectCores()
```


```{r}

chr_vector <- unique(as.character(tiles$chr))

# Build per-chromosome subsets
idx_chr_cpgs <- list()
meth_subset <- list()
idx_chr_tiles <- list()
tiles_subset <- list()

for(chrnum in chr_vector){
  idx_chr_cpgs[[chrnum]] <- grep(paste0("chr", chrnum, ":"), rownames(for_op))
  meth_subset[[chrnum]] <- for_op[idx_chr_cpgs[[chrnum]], ]
  idx_chr_tiles[[chrnum]] <- grep(paste0("chr", chrnum, ":"), rownames(tilesdf_all))
  tiles_subset[[chrnum]] <- tiles[idx_chr_tiles[[chrnum]], ]
}

# Run in parallel — adjust mc.cores to what you have available
results <- mclapply(chr_vector, function(chr) {
  merging_consecutive_regions(meth_subset[[chr]], tiles_subset[[chr]],
                              da_thr = length_region, cb_thr = length_region)
}, mc.cores = 25)

names(results) <- chr_vector
regions_plus_merged_regions <- do.call(rbind, results)
```
```{r}
chr_vector
```

```{r}
chr_vector <- unique(as.character(tiles$chr))
chr <- chr_vector[1]
results <- merging_consecutive_regions(mr[mr$chr == chr,], tiles[tiles$chr == chr,],
                              da_thr = length_region, cb_thr = length_region)
```

```{r}
res <- results
for (i in 2:length(chr_vector)) {
  r <- merging_consecutive_regions(mr[mr$chr == chr_vector[i],], tiles[tiles$chr == chr_vector[i],],
                              da_thr = length_region, cb_thr = length_region)
  res <- rbind(res, r)
}
```

```{r}
saveRDS(res, "/researchers/caitlin.page/cf_nano/r_output/optiles_all_emseq.rds")
```

```{r}
res
```

```{r}
distinct(res, chr)
```

```{r}
optiles_all <- res
optiles_all <- optiles_all %>% mutate(width = end - start + 1)
colnames(optiles_all)[1] <- "seqnames"
optiles_all$index <- 1:nrow(optiles_all)
```
```{r}
saveRDS(optiles_all, "/researchers/caitlin.page/cf_nano/r_output/optiles_all_emseq.rds")
```

```{r}
optiles_all <- readRDS("/researchers/caitlin.page/cf_nano/r_output/optiles_all_emseq.rds")
```


```{r}

tilesdf_all <- tilesdf_all[,1:4] %>% mutate(width = end - start + 1)
colnames(tilesdf_all)[1] <- "seqnames"
tilesdf_all$id <- 1:nrow(tilesdf_all)

```


```{r}
# overlap tiles and reads
overlap_pos <- find_overlaps(as_granges(optiles_all), as_granges(all_samples)) %>% data.frame()
```

```{r}
all_samples %>%
  ggplot(aes(x = as.factor(total))) +
  geom_bar()
```


```{r}
overlap_pos %>% group_by(region_id) %>% summarise(n_reads = n()) %>%
  ggplot(aes(x = n_reads)) +
  geom_histogram()
```


```{r}
# dtu

dtu_optile <- overlap_pos %>% filter(total >= 3) %>%
  distinct(sample, read_id, alpha, region_id) %>%
  mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                 alpha > 0.8 ~ "a > 0.8",
                                 alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                 alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                 TRUE ~ "0.6 <= a <= 0.8"))

dtu_optile <- dtu_optile %>%
  mutate(transcript = paste0(region_id, "_", alpha_group)) %>%
  group_by(sample, transcript) %>%
  summarise(transcript_count = n()) %>% ungroup() %>% data.frame()

# build matrix of "transcript" counts - transcript rows, sample as columns
dtu_optile <- dtu_optile %>% pivot_wider(names_from = sample, values_from = transcript_count)
dtu_optile[is.na(dtu_optile)] <- 0
dtu_optile$gene_id <- sub("_.*", "", dtu_optile$transcript)

# DTU analysis with edgeR
## Alicia prefers gene method over Simes method (documentation slightly confusing about difference)
y <- DGEList(as.matrix(dtu_optile[,2:7]), group = c(1,1,1,2,2,2), genes = dtu_optile[,c("gene_id", "transcript")])
design <- model.matrix(~ c(1,1,1,2,2,2), data = y$samples)
fit <- glmQLFit(y, design, robust=TRUE)
ds <- diffSpliceDGE(fit, coef = 2, geneid = "gene_id", exonid = "transcript")
ts.gene_optiles <- topSpliceDGE(ds, test = "gene", number = Inf)
```


```{r}
ts.gene_optiles <- ts.gene_optiles %>%
  mutate(region_id = gene_id) %>%
  separate_wider_delim(gene_id, delim = ":", names = c("seqnames", "gene_id")) %>%
  mutate(seqnames = gsub("chrchr", "chr", seqnames)) %>%
  separate_wider_delim(gene_id, delim = "-", names = c("start", "end"))
ts.gene_optiles$start <- as.double(ts.gene_optiles$start)
ts.gene_optiles$end <- as.double(ts.gene_optiles$end)
ts.gene_optiles$dtu_rank <- 1:nrow(ts.gene_optiles)
```


```{r}
saveRDS(ts.gene_optiles, "/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_emseq_healthy.rds")
```

nmn
ṉ
nnmnmnmnmnmnmmnmnmmnmnmnnmnmnm

```{r}
emseq_tiles <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_emseq_healthy.rds")
```


```{r}

```


```{r}
emseq_tiles %>% mutate(false_pos = ifelse(FDR < 0.05, TRUE, FALSE), tile = "tile") %>%
  ggplot(aes(x = tile, fill = false_pos)) +
  geom_bar(position = "fill")
```



```{r}
ts.gene_optiles %>% filter(FDR < 0.05)
```

11 false positives
I'll take it
mmmmmmmmmmmmmmmmmmm
