---
title: "reviewPlotsLocal"
author: "Caitlin Page"
date: "2026-03-16"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

## Introduction

```{r}
library(annotatr)
library(data.table)
library(plyranges)
library(tidyr)
library(stringr)
library(dplyr)
library(ggplot2)
```

```{r}
cg_sites <- cbind(data.frame(Biostrings::matchPattern("CG", BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38[["chr22"]])), seqnames = "chr22") %>% 
  mutate(pos = paste0(seqnames, "-", start)) %>%
  relocate(pos, seqnames)
cg_sites$index <- 1:nrow(cg_sites)
```

```{r}

source("../code/processData.R")
#for_beta_all <- readRDS("/researchers/caitlin.page/cf_nano/r_output/rrbs_alpha_beta.rds")
for_beta_all <- readRDS("../output/cluster_files/rrbs_alpha_beta_all.rds")
for_beta <- for_beta_all %>% filter(seqnames == "chr22")
rrbs_for_beta_pat <- pat_for_wgbs(for_beta)
for_beta <- for_beta %>% group_by(start) %>% mutate(coverage = n(), mean_alpha = mean(alpha)) %>% ungroup() %>% data.frame()
just_beta <- for_beta %>% distinct(start, beta, coverage, mean_alpha) %>% .[order(.$start),]

```


```{r}
annots <- c("hg38_cpgs")

annotations <- annotatr::build_annotations(genome = 'hg38', annotations = annots)
seqs <- annotations %>% data.frame() %>% distinct(seqnames) %>% .[1:22,]
annotations <- annotations %>% data.frame() %>% filter(seqnames %in% seqs)
```

```{r}
for_beta <- find_overlaps(as_granges(for_beta), as_granges(annotations)) %>% data.frame() %>%
  mutate(cgi = ifelse(type == "hg38_cpg_islands", "island", "not_island"))
```

```{r}
annots <- c("hg38_basicgenes", "hg38_genes_intergenic")

annotations <- annotatr::build_annotations(genome = 'hg38', annotations = annots)
seqs <- annotations %>% data.frame() %>% distinct(seqnames) %>% .[1:22,]
annotations <- annotations %>% data.frame() %>% filter(seqnames %in% seqs)
annotations %>% group_by(type) %>% summarise(n=n())
```

```{r}
for_beta_genes <- find_overlaps(as_granges(for_beta), as_granges(annotations)) %>% data.frame()
for_beta_genes[1:10,]
```

```{r}
rbind(
      for_beta %>% mutate(variable = start, value = beta) %>% distinct(variable, value) %>% mutate(technology = "RRBS", method = "beta"),
      for_beta %>% mutate(variable = read_id, value = alpha) %>% distinct(variable, value) %>% mutate(technology = "RRBS", method = "alpha")) %>%
  ggplot(aes(x = method, y = value, fill = method)) +
  geom_violin() +
  theme_bw() +
  theme(legend.position = "none")
```

```{r}
rbind(
      for_beta %>% filter(total > 4) %>% mutate(variable = start, value = beta) %>% distinct(variable, value) %>% mutate(technology = "RRBS", method = "beta"),
      for_beta %>% filter(total > 4) %>% mutate(variable = read_id, value = alpha) %>% distinct(variable, value) %>% mutate(technology = "RRBS", method = "alpha")) %>%
  ggplot(aes(x = method, y = value, fill = method)) +
  geom_violin() +
  theme_bw() +
  theme(legend.position = "none")
```


```{r}
for_beta %>% distinct(start, beta) %>% .[order(.$start),] %>% mutate(dist_to_next = lead(start) - start) %>%
  filter(beta %in% c(0,0.5,1)) %>%
  ggplot(aes(x = as.factor(beta), y = log2(dist_to_next), fill = as.factor(beta))) +
  geom_boxplot() +
  theme_bw() +
  theme(legend.position = "none")
```

```{r}
for_beta %>% group_by(read_id) %>% mutate(dist_to_next = lead(start) - start) %>%
  filter(alpha %in% c(0,0.5,1)) %>%
  ggplot(aes(x = as.factor(alpha), y = log2(dist_to_next), fill = as.factor(alpha))) +
  geom_boxplot() +
  theme_bw() +
  theme(legend.position = "none")
```

* mantel haesnz
```{r}
mantel_haen_dist_next_cpg <- function(long_data, num_cpg_read = NULL) {
  if(!is.null(num_cpg_read)) {
    format_data <- long_data %>%
        filter(total == num_cpg_read) 
  } else {format_data <- long_data}
  format_data <- format_data %>%
      group_by(read_id) %>%
      mutate(dist_to_next = lead(start) - start,
            meth_pattern = paste0(meth_status, lead(meth_status))) %>%
      ungroup() %>%
      fill(dist_to_next, .direction = "down") %>%
      distinct(read_id, meth_pattern, dist_to_next) %>%
      filter(!grepl("A", .$meth_pattern)) %>%
      group_by(dist_to_next, meth_pattern) %>%
      summarise(count = n()) %>%
      ungroup()
  
  chisq_table <- format_data %>%
      mutate(cpg_1 = substr(meth_pattern, 1, 1),
             cpg_2 = substr(meth_pattern, 2, 2)) 
  # check for missing patterns (it screws up the chi test) and replace with 0
  missing_mm <- chisq_table %>% group_by(dist_to_next) %>% filter(!any(grepl("MM", meth_pattern))) %>% distinct(dist_to_next) %>% ungroup() %>% .$dist_to_next
  for(i in 1:length(missing_mm)) {
    chisq_table <- rbind(chisq_table, c(missing_mm[i], "MM", 0, "M", "M"))
  }
  missing_mu <- chisq_table %>% group_by(dist_to_next) %>% filter(!any(grepl("MU", meth_pattern))) %>% distinct(dist_to_next) %>% ungroup() %>% .$dist_to_next
  for(i in 1:length(missing_mu)) {
    chisq_table <- rbind(chisq_table, c(missing_mu[i], "MU", 0, "M", "U"))
  }
  missing_um <- chisq_table %>% group_by(dist_to_next) %>% filter(!any(grepl("UM", meth_pattern))) %>% distinct(dist_to_next) %>% ungroup() %>% .$dist_to_next
  for(i in 1:length(missing_um)) {
    chisq_table <- rbind(chisq_table, c(missing_um[i], "UM", 0, "U", "M"))
  }
  missing_uu <- chisq_table %>% group_by(dist_to_next) %>% filter(!any(grepl("UU", meth_pattern))) %>% distinct(dist_to_next) %>% ungroup() %>% .$dist_to_next
  for(i in 1:length(missing_uu)) {
    chisq_table <- rbind(chisq_table, c(missing_uu[i], "UU", 0, "U", "U"))
  }

  chisq_table$count <- as.integer(chisq_table$count)
  chisq_table$dist_to_next <- as.integer(chisq_table$dist_to_next)
  chisq_table <- chisq_table %>% filter(!is.na(dist_to_next))
  
  chisq_table <- chisq_table %>% 
      group_by(dist_to_next) %>%
      reframe(tab = list(xtabs(count ~ cpg_1 + cpg_2, data = cur_data())))
  chisq_table <- chisq_table$tab
  
  names(chisq_table) <- format_data %>%
      mutate(dist_to_next = paste0("dist_next_", dist_to_next)) %>%
      .$dist_to_next %>% 
    unique()
  distinct_rows <- format_data$dist_to_next %>% 
    unique() %>%
    length()
  chisq_table <- simplify2array(chisq_table)
  
  # the testing
  try <- capture.output(epiDisplay::mhor(mhtable = chisq_table, graph = FALSE))

  try <- try %>% data.frame() %>% .[4:(distinct_rows + 4),] %>% data.frame()
  colnames(try) <- "all"

  try <- try %>% separate_wider_delim(cols = all, delim = " ", names = c("type", "dist", "numbers"), too_few = "align_start", too_many = "merge")

  numeric_lines <- try$numbers
  # get Inf and d separate
  pattern <- "Inf|-Inf|[-+]?\\d+\\.?\\d*(?:[eE][-+]?\\d+)?"
  vals <- regmatches(
    numeric_lines,
    gregexpr(pattern, numeric_lines))

  m <- do.call(rbind, vals)
  m <- as.vector(m)
  names(m) <- rep(c("OR", "lower_lim", "upper_lim", "p_value"), times = (distinct_rows + 1))
  split_vals <- split(m, names(m))
  df <- cbind.data.frame(split_vals)
  df <- lapply(df, as.double) %>% data.frame()

  # merge
  try <- cbind(try[,1:2], df[,c(2,1,4,3)])
  try <- try %>% mutate(IPD = as.numeric(sub(".*_", "", dist)))
  try
}
```

```{r}
mantel_alpha <- mantel_haen_dist_next_cpg(for_beta)
mantel_alpha %>%
  ggplot(aes(x = IPD, y = OR)) +
  geom_point() +
  theme_bw()
```
* this does not look like my old other plot that looks nice
```{r}
mantel_alpha <- mantel_haen_dist_next_cpg(for_beta, 2)
mantel_alpha %>%
  ggplot(aes(x = IPD, y = log(OR + 0.1))) +
  geom_point() +
  theme_bw()
```

```{r}
for_beta %>% mutate(read_length = read_end - read_start) %>% distinct(read_id, alpha, read_length) %>%
  ggplot(aes(x = as.factor(read_length), y = alpha)) +
  geom_boxplot()
```

# binning 
```{r}
seqinfo <- seqlengths(BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38)
seqinfo <- seqinfo[grepl("^chr22+$", names(seqinfo))]
seqinfo
bins_5kb <- tileGenome(seqinfo, tilewidth=5000, cut.last.tile.in.chrom=TRUE)

bins_5kb <- bins_5kb %>% data.frame() %>% mutate(bin_num = 1:n()) %>% as_granges()
bins_5kb

bins_5b <- tileGenome(seqinfo, tilewidth=500, cut.last.tile.in.chrom=TRUE)
bins_5b <- bins_5b %>% data.frame() %>% mutate(bin_num = 1:n()) %>% as_granges()
bins_5b
```


```{r}
overlap_bins_reads <- find_overlaps(bins_5kb, as_granges(for_beta)) %>% data.frame()
overlap_bins_reads_5b <- find_overlaps(bins_5b, as_granges(for_beta)) %>% data.frame()
```
```{r}
overlap_bins_reads %>% group_by(bin_num) %>% summarise(var_alpha = var(alpha)) %>% mutate(sample = "GM12878") %>%
  ggplot(aes(x = sample, y = var_alpha)) +
  geom_boxplot() +
  theme_bw() 
```


```{r}
overlap_bins_reads_5b %>% group_by(bin_num) %>% summarise(var_alpha = var(alpha)) %>% mutate(sample = "GM12878") %>%
  ggplot(aes(x = sample, y = var_alpha)) +
  geom_boxplot() +
  theme_bw() 
```

```{r}
overlap_bins_reads %>% group_by(bin_num) %>% summarise(var_alpha = var(alpha)) %>% .[order(.$var_alpha, decreasing = TRUE),]
```

```{r}
overlap_bins_reads %>% filter(total > 2) %>% group_by(bin_num) %>% summarise(var_alpha = var(alpha), num_reads = n()) %>% .[order(.$var_alpha, decreasing = TRUE),] %>% filter(num_reads > 3)
```
```{r}
overlap_bins_reads_5b %>% filter(total > 2) %>% group_by(bin_num) %>% summarise(var_alpha = var(alpha), num_reads = n()) %>% .[order(.$var_alpha, decreasing = TRUE),] %>% filter(num_reads > 3)
```

```{r}
overlap_bins_reads_5b %>% dplyr::filter(bin_num == 95265) %>%
  ggplot(aes(x = read_start, y = alpha, colour = read_id)) +
  geom_segment(aes(x = read_start, xend = read_end, y = alpha, yend = alpha)) +
  theme(legend.position = "none")

overlap_bins_reads_5b %>% dplyr::filter(bin_num == 87993) %>%
  ggplot(aes(x = read_start, y = alpha, colour = read_id)) +
  geom_segment(aes(x = read_start, xend = read_end, y = alpha, yend = alpha)) +
  theme(legend.position = "none")

overlap_bins_reads_5b %>% dplyr::filter(bin_num == 46292) %>%
  ggplot(aes(x = read_start, y = alpha, colour = read_id)) +
  geom_segment(aes(x = read_start, xend = read_end, y = alpha, yend = alpha)) +
  theme(legend.position = "none")

overlap_bins_reads_5b %>% dplyr::filter(bin_num == 23198) %>%
  ggplot(aes(x = read_start, y = alpha, colour = read_id)) +
  geom_segment(aes(x = read_start, xend = read_end, y = alpha, yend = alpha)) +
  theme(legend.position = "none")

overlap_bins_reads_5b %>% dplyr::filter(bin_num == 97770) %>%
  ggplot(aes(x = read_start, y = alpha, colour = read_id)) +
  geom_segment(aes(x = read_start, xend = read_end, y = alpha, yend = alpha)) +
  theme(legend.position = "none")
```

```{r}
set.seed(42) # For reproducibility
# Calculate a single jitter amount per row (segment)
jitter_amount <- runif(nrow(distinct(overlap_bins_reads_5b, read_id)), -0.01, 0.01) 
read_jitter <- cbind(distinct(overlap_bins_reads_5b, read_id, alpha), jitter_amount) %>% data.frame()
read_jitter$alpha_jit <- read_jitter$alpha + read_jitter$jitter_amount

# Add jittered coordinates to new columns in the data frame
overlap_bins_reads_5b$alpha_jit <- overlap_bins_reads_5b[match(overlap_bins_reads_5b$read_id, read_jitter$read_id), "alpha_jit"]
```


* high var in alpha
```{r}
overlap_bins_reads_5b %>% dplyr::filter(bin_num == 97770) %>%
  ggplot(aes(x = read_start, y = alpha, colour = read_id)) +
  geom_segment(aes(x = read_start, xend = read_end, y = alpha_jit, yend = alpha_jit)) +
  theme(legend.position = "none")
```
```{r}
overlap_bins_reads_5b %>% dplyr::filter(bin_num == 97770) %>% group_by(read_start, alpha) %>% mutate(count = n()) %>%
  ggplot(aes(x = read_start, y = alpha, colour = read_id)) +
  geom_segment(aes(x = read_start, xend = read_end, y = alpha, yend = alpha, linewidth = count)) +
  guides(colour = "none") +
  theme_bw()
 # theme(legend.position = "none")
```

* high density
```{r}
overlap_bins_reads_5b %>% filter(total > 2) %>% group_by(bin_num) %>% summarise(var_alpha = var(alpha), num_reads = n()) %>% .[order(.$num_reads, decreasing = TRUE),]
```
```{r}
overlap_bins_reads_5b %>% dplyr::filter(bin_num == 92143) %>% group_by(read_start, alpha) %>% mutate(count = n()) %>%
  ggplot(aes(x = read_start, y = alpha, colour = read_id)) +
  geom_segment(aes(x = read_start, xend = read_end, y = alpha, yend = alpha, linewidth = count)) +
  guides(colour = "none") +
  theme_bw()
```
```{r}
overlap_bins_reads_5b %>% dplyr::filter(bin_num == 92143) %>% distinct(read_id) %>% .$read_id

overlap_bins_reads_5b %>% distinct(read_id, bin_num) %>% group_by(bin_num) %>% summarise(n=n()) %>% .[order(.$n, decreasing = TRUE),] 
```
```{r}
rrbs_plot(for_beta, read_ids = overlap_bins_reads_5b %>% dplyr::filter(bin_num == 25388) %>% distinct(read_id) %>% .$read_id, count_instead = TRUE) + theme_bw()
rrbs_plot2(for_beta, read_ids = overlap_bins_reads_5b %>% dplyr::filter(bin_num == 25388) %>% distinct(read_id) %>% .$read_id, count_instead = TRUE) + theme_bw()
```


```{r}
rrbs_plot(for_beta, read_ids = overlap_bins_reads_5b %>% dplyr::filter(bin_num == 92143) %>% distinct(read_id) %>% .$read_id, count_instead = TRUE) + theme_bw()
rrbs_plot(for_beta, read_ids = overlap_bins_reads_5b %>% dplyr::filter(bin_num == 21144) %>% distinct(read_id) %>% .$read_id, count_instead = TRUE) + theme_bw()
rrbs_plot(for_beta, read_ids = overlap_bins_reads_5b %>% dplyr::filter(bin_num == 25388) %>% distinct(read_id) %>% .$read_id, count_instead = TRUE) + theme_bw()
```


```{r}
overlap_bins_reads_5b %>% dplyr::filter(bin_num == 92143) %>%
  ggplot(aes(x = read_start, y = alpha, colour = read_id)) +
  geom_segment(aes(x = read_start, xend = read_end, y = alpha, yend = alpha)) +
  theme(legend.position = "none")

overlap_bins_reads_5b %>% dplyr::filter(bin_num == 25388) %>%
  ggplot(aes(x = read_start, y = alpha, colour = read_id)) +
  geom_segment(aes(x = read_start, xend = read_end, y = alpha, yend = alpha)) +
  theme(legend.position = "none")
```


```{r}
overlap_bins_reads %>% dplyr::filter(bin_num == 6040) %>%
  ggplot(aes(x = read_start, y = alpha, colour = read_id)) +
  geom_segment(aes(x = read_start, xend = read_end, y = alpha, yend = alpha)) +
#  labs(title = "bin with highest variance") +
 # theme(legend.position = "none") +
  scale_x_continuous(limits = c(30196400, 30196500))
```


```{r}
overlap_bins_reads %>% dplyr::filter(bin_num == 9527) %>%
  ggplot(aes(x = read_start, y = alpha, colour = read_id)) +
  geom_segment(aes(x = read_start, xend = read_end, y = alpha, yend = alpha)) +
#  labs(title = "bin with highest variance") +
  theme(legend.position = "none") + 
  scale_x_continuous(limits = c(47631750,47632250))
```


```{r}
overlap_bins_reads %>% dplyr::filter(bin_num == 8653) %>%
  ggplot(aes(x = read_start, y = alpha, colour = read_id)) +
  geom_segment(aes(x = read_start, xend = read_end, y = alpha, yend = alpha)) +
#  labs(title = "bin with highest variance") +
  theme(legend.position = "none") +
  scale_x_continuous(limits = c(43263000, 43264500))
```




# string of beta

```{r}
string_same_beta <- function(betas_only, range_type = "medium") {
  df <- betas_only %>% 
   # mutate(seqnames = "chr22") %>% # will need seqnames later for scale
    mutate(row_num = 1:nrow(.)) %>%
    mutate(beta_range = case_when(beta >= 0 & beta <= 0.2 ~ "low", 
                                beta >= 0.4 & beta <= 0.6 ~ "medium",
                                beta >= 0.8 & beta <= 1 ~ "high",
                                beta > 0.2 & beta < 0.4 ~ "low_med",
                                TRUE ~ "med_high")) %>%
  mutate(length = sequence(rle(.$beta_range)$lengths)) %>% # when I have all chroms this will need to change a bit
  group_by(beta_range) %>%
  filter(beta_range == range_type, 
         length == max(length))
  num_cpgs <- df$length
  end_row <- df$row_num
  start_row <- end_row - num_cpgs + 1
  
  c(betas_only[start_row,]$start, betas_only[end_row,]$start) # at some point this will need a chromosome
}
beta_string <- string_same_beta(just_beta)
```

```{r}
just_beta %>% 
   # mutate(seqnames = "chr22") %>% # will need seqnames later for scale
    mutate(row_num = 1:nrow(.)) %>%
    mutate(beta_range = case_when(beta >= 0 & beta <= 0.2 ~ "low", 
                                beta >= 0.4 & beta <= 0.6 ~ "medium",
                                beta >= 0.8 & beta <= 1 ~ "high",
                                beta > 0.2 & beta < 0.4 ~ "low_med",
                                TRUE ~ "med_high")) %>%
  mutate(length = sequence(rle(.$beta_range)$lengths)) %>% filter(beta_range == "medium") %>%
  mutate(id = ifelse(row_num == lag(row_num) + 1, ))
  group_by(beta_range) %>%
  filter(beta_range == "medium", 
         length == max(length))
```

```{r}
just_beta %>% 
  mutate(row_num = 1:nrow(.)) %>%
    mutate(beta_range = case_when(beta >= 0 & beta <= 0.2 ~ "low", 
                                beta >= 0.4 & beta <= 0.6 ~ "medium",
                                beta >= 0.8 & beta <= 1 ~ "high",
                                beta > 0.2 & beta < 0.4 ~ "low_med",
                                TRUE ~ "med_high")) %>%
  group_by(beta_range) %>%
  mutate(consec = n()) %>%
  mutate(count = 1:n()) %>%
  ungroup() %>%
  mutate(length = sequence(rle(.$beta_range)$lengths)) %>% 
  mutate(group = cumsum(c(TRUE, diff(length) <= 0))) %>%
  group_by(group) %>%
  mutate(length = n(), av_coverage = mean(coverage), av_beta = mean(beta)) %>%
  ungroup() %>%
  distinct(group, beta_range, length, av_coverage, av_beta) %>% filter(beta_range == "medium", length > 2) %>% .[order(.$length, decreasing = TRUE),] 
```

```{r}
just_beta <- just_beta %>% mutate(seqnames = "chr22") %>%
  mutate(row_num = 1:nrow(.)) %>%
  mutate(beta_range = case_when(beta >= 0 & beta <= 0.2 ~ "low", 
                                beta >= 0.4 & beta <= 0.6 ~ "medium",
                                beta >= 0.8 & beta <= 1 ~ "high",
                                beta > 0.2 & beta < 0.4 ~ "low_med",
                                TRUE ~ "med_high"))
just_beta[1:5,]

grouped_betas <- just_beta %>% 
  group_by(beta_range) %>%
  mutate(consec = n()) %>%
  mutate(count = 1:n()) %>%
  ungroup() %>%
  mutate(length = sequence(rle(.$beta_range)$lengths)) %>% 
  mutate(group = cumsum(c(TRUE, diff(length) <= 0))) %>%
  group_by(group) %>%
  mutate(length = n(), av_coverage = mean(coverage), av_beta = mean(beta)) %>%
  ungroup() %>% filter(beta_range == "medium", length > 2) %>% data.frame() 
grouped_betas[1:10,]
```

```{r}
source("../code/processData.R")
rrbs_plot(for_beta, for_beta %>% filter(start %in% just_beta[1:6,]$start) %>% .$read_id %>% unique()) + theme_bw()
```
```{r}
rrbs_plot(for_beta, for_beta %>% filter(start %in% just_beta[1:6,]$start) %>% .$read_id %>% unique(), count_instead = TRUE) + theme_bw()
for_beta %>% filter(start %in% just_beta[1:6,]$start) %>% distinct(read_id, alpha) %>%
  ggplot(aes(x = alpha)) +
  geom_density() +
  theme_bw()

for_beta %>% filter(start %in% just_beta[1:6,]$start) %>% distinct(read_id, alpha) %>% mutate(region = "region") %>%
  ggplot(aes(x = region, y = alpha)) +
  geom_violin() +
  theme_bw() +
  theme(legend.position = "none")

rbind(
      for_beta %>% filter(start %in% just_beta[1:6,]$start) %>% mutate(variable = start, value = beta) %>% distinct(variable, value) %>% mutate(technology = "RRBS", method = "beta"),
      for_beta %>% filter(start %in% just_beta[1:6,]$start) %>% mutate(variable = read_id, value = alpha) %>% distinct(variable, value) %>% mutate(technology = "RRBS", method = "alpha")) %>%
  ggplot(aes(x = method, y = value, fill = method)) +
  geom_violin() +
  theme_bw() +
  theme(legend.position = "none")

rbind(
      for_beta %>% filter(start %in% just_beta[1:6,]$start) %>% mutate(variable = start, value = beta) %>% distinct(variable, value) %>% mutate(technology = "RRBS", method = "beta"),
      for_beta %>% filter(start %in% just_beta[1:6,]$start) %>% mutate(variable = read_id, value = alpha) %>% distinct(variable, value) %>% mutate(technology = "RRBS", method = "alpha")) %>%
  ggplot(aes(x = method, y = value, fill = method)) +
  geom_boxplot() +
  geom_violin() +
  theme_bw() +
  theme(legend.position = "none")

rbind(
      for_beta %>% filter(start %in% just_beta[1:6,]$start) %>% mutate(variable = start, value = beta) %>% distinct(variable, value) %>% mutate(technology = "RRBS", method = "beta"),
      for_beta %>% filter(start %in% just_beta[1:6,]$start) %>% mutate(variable = read_id, value = alpha) %>% distinct(variable, value) %>% mutate(technology = "RRBS", method = "alpha")) %>%
  ggplot(aes(x = value, colour = method)) +
  geom_density() +
  theme_bw()
```

```{r}
just_beta[1:6,]
for_beta %>% filter(start %in% just_beta[1:6,]$start) %>% distinct(read_id, alpha) %>% summarise(var(alpha))
```

# data structures

```{r}
for_beta_all %>% distinct(read_id)
for_beta_all
```

```{r}
for_beta_all %>% distinct(seqnames, start)
```

```{r}
a <- just_beta[,c(5,1,2,3)] 
just_beta
a[1:5,]
rrbs_for_beta_pat[1:5,c(1:3,7)]
```

```{r}
for_beta[1:10,c(6,1:5,7:17)]
```

# proportions

```{r}
for_beta %>% filter(seqnames == "chr22") %>% distinct(read_id, alpha, type.y) %>%
  mutate(alpha_group = case_when(alpha < 0.3 ~ "a < 0.3",
                                alpha > 0.7 ~ "a > 0.7",
                                TRUE ~ " 0.3 < a < 0.7")) %>%
  ggplot(aes(x = type.y, fill = alpha_group)) +
  geom_bar(position = "fill") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

```{r}
rbind(
      for_beta %>% filter(seqnames == "chr22") %>% mutate(variable = start, value = beta) %>% distinct(variable, value, cgi) %>% mutate(technology = "RRBS", method = paste0(cgi, ": beta")),
      for_beta %>% filter(seqnames == "chr22") %>% mutate(variable = read_id, value = alpha) %>% distinct(variable, value, cgi) %>% mutate(technology = "RRBS", method = paste0(cgi, ": alpha"))) %>%
  mutate(value_group = case_when(value < 0.3 ~ "x < 0.3",
                                value > 0.7 ~ "x > 0.7",
                                TRUE ~ " 0.3 < x < 0.7")) %>%
  ggplot(aes(x = method, fill = value_group)) +
  geom_bar(position = "fill") +
  theme_bw()
```

```{r}
rbind(
      for_beta_genes %>% filter(seqnames == "chr22") %>% 
        mutate(type = case_when(type.y == "hg38_genes_promoters" | type.y == "hg38_genes_1to5kb" ~ "promoter",
                                type.y == "hg38_genes_introns" | type.y == "hg38_genes_exons" ~ "gene_body", TRUE ~ "UTR")) %>% 
        mutate(variable = start, value = beta) %>% distinct(variable, value, type) %>% mutate(technology = "RRBS", method = paste0(type, ": beta")),
      for_beta_genes %>% filter(seqnames == "chr22") %>%
        mutate(type = case_when(type.y == "hg38_genes_promoters" | type.y == "hg38_genes_1to5kb" ~ "promoter",
                                type.y == "hg38_genes_introns" | type.y == "hg38_genes_exons" ~ "gene_body", TRUE ~ "UTR")) %>% 
        mutate(variable = read_id, value = alpha) %>% distinct(variable, value, type) %>% mutate(technology = "RRBS", method = paste0(type, ": alpha"))) %>%
  mutate(value_group = case_when(value < 0.3 ~ "x < 0.3",
                                value > 0.7 ~ "x > 0.7",
                                TRUE ~ " 0.3 < x < 0.7")) %>%
  ggplot(aes(x = method, fill = value_group)) +
  geom_bar(position = "fill") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) 
```

# dmr analysis

```{r}
for_beta_2 <- readRDS("../output/cluster_files/rrbs_alpha_beta_rep2.rds")

k562_1 <- readRDS("../output/cluster_files/k562_rep1.rds")
k562_2 <- readRDS("../output/cluster_files/k562_rep2.rds")
```

```{r}
dmrs <- readRDS("../output/cluster_files/dmr_encode_gm12878_k562.rds")
dmrs$id <- 1:nrow(dmrs)
```

```{r}
find_overlaps(as_granges(mutate(just_beta, width = 1)), as_granges(dmrs)) %>% data.frame() %>% group_by(id, beta_range) %>% summarise(n())
```
```{r}
find_overlaps(as_granges(mutate(just_beta, width = 1)), as_granges(dmrs)) %>% data.frame() %>% group_by(id) %>% summarise(mean_beta = mean(beta)) %>% ungroup() %>%
  ggplot(aes(x = mean_beta)) +
  geom_density() + theme_bw()

find_overlaps(as_granges(mutate(just_beta, width = 1)), as_granges(dmrs)) %>% data.frame() %>% group_by(id) %>% summarise(mean_beta = mean(beta)) %>% ungroup() %>%
  ggplot(aes(x = mean_beta)) +
  geom_histogram(bins = 50) + theme_bw()
```

```{r}
find_overlaps(as_granges(mutate(just_beta, width = 1)), as_granges(dmrs)) %>% data.frame() %>% 
  ggplot(aes(x = beta)) +
  geom_density() +
  theme_bw()

find_overlaps(as_granges(mutate(just_beta, width = 1)), as_granges(dmrs)) %>% data.frame() %>% 
  ggplot(aes(x = beta)) +
  geom_histogram(bins = 50) +
  theme_bw()
```

```{r}
in_dmrs <- find_overlaps(as_granges(mutate(just_beta, width = 1)), as_granges(dmrs)) %>% data.frame()
just_beta %>% mutate(in_dmr = ifelse(start %in% in_dmrs$start, "DMR", "No_DMR")) %>%
  ggplot(aes(x = beta, colour = in_dmr)) +
  geom_density(linewidth = 1) +
  theme_bw()
```

```{r}
just_beta %>% mutate(in_dmr = ifelse(start %in% in_dmrs$start, "DMR", "No_DMR")) %>%
  ggplot(aes(x = beta, fill = in_dmr)) +
  geom_histogram(alpha = 0.4)
```


```{r}
just_beta %>% mutate(in_dmr = ifelse(start %in% in_dmrs$start, "DMR", "No_DMR")) %>%
  ggplot(aes(x = beta, fill = in_dmr)) +
  geom_histogram() +
  theme_bw() +
  facet_wrap(~ in_dmr)
```


# when beta = 0.5

# find genes with mean beta = 0.5

```{r}
for_beta_genes[,c(1:16,22:23,26:28)] %>% distinct() %>%
  filter(!is.na(symbol.y)) %>%
  distinct(seqnames, start, beta, gene_id.y, symbol.y) %>%
  group_by(gene_id.y) %>%
  mutate(mean_beta = mean(beta), n_cgs= n()) %>% 
  ungroup() %>% distinct(gene_id.y, symbol.y, mean_beta, n_cgs) %>%
  ggplot(aes(x = n_cgs, y = mean_beta)) +
  geom_point()
```

```{r}
for_beta_genes[,c(1:16,22:23,26:28)] %>% distinct() %>%
  filter(!is.na(symbol.y)) %>%
  distinct(seqnames, start, beta, gene_id.y, symbol.y) %>%
  group_by(gene_id.y) %>%
  mutate(mean_beta = mean(beta), n_cgs = n()) %>% 
  ungroup() %>% distinct(gene_id.y, symbol.y, mean_beta, n_cgs) %>%
  ggplot(aes(x = mean_beta)) +
  geom_density()

for_beta_genes[,c(1:16,22:23,26:28)] %>% distinct() %>%
  filter(!is.na(symbol.y)) %>%
  distinct(seqnames, start, beta, gene_id.y, symbol.y) %>%
  group_by(gene_id.y) %>%
  mutate(mean_beta = mean(beta), n_cgs = n()) %>% 
  ungroup() %>% distinct(gene_id.y, symbol.y, mean_beta, n_cgs) %>%
  ggplot(aes(x = mean_beta)) +
  geom_histogram()
```

```{r}
for_beta_genes[,c(1:16,22:23,26:28)] %>% distinct() %>%
  filter(!is.na(symbol.y)) %>%
  distinct(seqnames, start, beta, gene_id.y, symbol.y) %>%
  group_by(gene_id.y) %>%
  mutate(mean_beta = mean(beta), n_cgs = n()) %>% 
  ungroup() %>% distinct(gene_id.y, symbol.y, mean_beta, n_cgs) %>%
  ggplot(aes(x = n_cgs)) +
  geom_density()

for_beta_genes[,c(1:16,22:23,26:28)] %>% distinct() %>%
  filter(!is.na(symbol.y)) %>%
  distinct(seqnames, start, beta, gene_id.y, symbol.y) %>%
  group_by(gene_id.y) %>%
  mutate(mean_beta = mean(beta), n_cgs = n()) %>% 
  ungroup() %>% distinct(gene_id.y, symbol.y, mean_beta, n_cgs) %>%
  ggplot(aes(x = n_cgs)) +
  geom_histogram()
```

```{r}
for_beta_genes[,c(1:16,22:23,26:28)] %>% distinct() %>%
  filter(!is.na(symbol.y)) %>%
  distinct(read_id, alpha, gene_id.y, symbol.y) %>% distinct(symbol.y)
```


```{r}
for_beta_genes[,c(1:16,22:23,26:28)] %>% distinct() %>%
  filter(!is.na(symbol.y)) %>%
  distinct(seqnames, start, beta, gene_id.y, symbol.y) %>%
  group_by(gene_id.y) %>%
  mutate(mean_beta = mean(beta), n_cgs= n()) %>% 
  ungroup() %>% distinct(gene_id.y, symbol.y, mean_beta, n_cgs) %>%
  filter(mean_beta > 0.4 & mean_beta < 0.6) %>% .[order(.$mean_beta),] %>% nrow()

for_beta_genes[,c(1:16,22:23,26:28)] %>% distinct() %>%
  filter(!is.na(symbol.y)) %>%
  distinct(seqnames, start, beta, gene_id.y, symbol.y) %>%
  group_by(gene_id.y) %>%
  mutate(mean_beta = mean(beta), n_cgs= n()) %>% 
  ungroup() %>% distinct(gene_id.y, symbol.y, mean_beta, n_cgs) %>%
  filter(mean_beta > 0.4 & mean_beta < 0.6)

for_beta_genes[,c(1:16,22:23,26:28)] %>% distinct() %>%
  filter(!is.na(symbol.y)) %>%
  distinct(seqnames, start, beta, gene_id.y, symbol.y) %>%
  group_by(gene_id.y) %>%
  mutate(mean_beta = mean(beta), n_cgs= n()) %>% 
  ungroup() %>% distinct(gene_id.y, symbol.y, mean_beta, n_cgs) %>%
  filter(mean_beta == 0.5) 

# mean alpha
for_beta_genes[,c(1:16,22:23,26:28)] %>% distinct() %>%
  filter(!is.na(symbol.y)) %>%
  distinct(read_id, alpha, gene_id.y, symbol.y) %>%
  group_by(gene_id.y) %>%
  mutate(mean_alpha = mean(alpha), n_reads= n()) %>% 
  ungroup() %>% distinct(gene_id.y, symbol.y, mean_alpha, n_reads) %>%
  filter(mean_alpha == 0.5) 

for_beta_genes[,c(1:16,22:23,26:28)] %>% distinct() %>%
  filter(!is.na(symbol.y)) %>%
  distinct(read_id, alpha, gene_id.y, symbol.y) %>%
  group_by(gene_id.y) %>%
  mutate(mean_alpha = mean(alpha), n_reads= n()) %>% 
  ungroup() %>% distinct(gene_id.y, symbol.y, mean_alpha, n_reads) %>%
  filter(mean_alpha > 0.4, mean_alpha < 0.6) 
```
* it is a lot of the same genes that have mean_alpha=mean_beta = 0.5

```{r}
rbind(for_beta_genes[,c(1:16,22:23,26:28)] %>% distinct() %>%
  filter(!is.na(symbol.y)) %>%
  distinct(seqnames, start, beta, gene_id.y, symbol.y) %>%
  group_by(gene_id.y) %>%
  mutate(mean_value = mean(beta), n= n()) %>% 
  ungroup() %>% distinct(gene_id.y, symbol.y, mean_value, n) %>%
  filter(mean_value == 0.5) %>% mutate(value = "beta"),
for_beta_genes[,c(1:16,22:23,26:28)] %>% distinct() %>%
  filter(!is.na(symbol.y)) %>%
  distinct(read_id, alpha, gene_id.y, symbol.y) %>%
  group_by(gene_id.y) %>%
  mutate(mean_value = mean(alpha), n= n()) %>% 
  ungroup() %>% distinct(gene_id.y, symbol.y, mean_value, n) %>%
  filter(mean_value == 0.5) %>% mutate(value = "alpha")) %>% group_by(symbol.y) %>% mutate(n=n()) 
```


```{r}
for_beta_genes[,c(1:16,22:23,26:28)] %>% distinct() %>%
  filter(!is.na(symbol.y)) %>%
  distinct(seqnames, start, beta, gene_id.y, symbol.y) %>%
  group_by(gene_id.y) %>%
  mutate(mean_beta = mean(beta), n_cgs= n()) %>% 
  ungroup() %>% distinct(gene_id.y, symbol.y, mean_beta, n_cgs) %>%
  ggplot(aes(x = mean_beta)) +
  geom_bar()
```


```{r}
medium_genes <- for_beta_genes[,c(1:16,22:23,26:28)] %>% distinct() %>%
  filter(!is.na(symbol.y)) %>%
  distinct(seqnames, start, beta, gene_id.y, symbol.y) %>%
  group_by(gene_id.y) %>%
  mutate(mean_beta = mean(beta), n_cgs= n()) %>% 
  ungroup() %>% distinct(gene_id.y, symbol.y, mean_beta, n_cgs) %>%
  filter(mean_beta >= 0.4 & mean_beta <= 0.6) %>% .[order(.$mean_beta),]

for_beta_genes %>% mutate(gene_status = ifelse(symbol.y %in% medium_genes$symbol.y, "medium_beta", "not")) %>%
  distinct(read_id, alpha, symbol.y, gene_status) %>%
  ggplot(aes(x = gene_status, y = alpha)) +
  geom_violin()

for_beta_genes %>% mutate(gene_status = ifelse(symbol.y %in% medium_genes$symbol.y, "medium_beta", "not")) %>%
  distinct(read_id, alpha, symbol.y, gene_status) %>%
  ggplot(aes(x = alpha, colour = gene_status)) +
  geom_density()
```


* 127 genes
* 8 with mean = 0.5

* SLC5A1 has most cpgs = 18

```{r}
for_beta_genes[,c(1:16,22:23,26:28)] %>% distinct() %>%
  filter(symbol.y == "SLC5A1") %>% distinct(read_id, meth_pattern)
```

```{r}
#source("../code/processData.R")
test_reads <- for_beta_genes[,c(1:16,22:23,26:28)] %>% distinct() %>%
  filter(symbol.y == "SLC5A1") %>% distinct(read_id) %>% .$read_id

rrbs_plot(for_beta, read_ids = test_reads) + theme_bw()
```
```{r}
for_beta_genes[,c(1:16,22:23,26:28)] %>% distinct() %>%
  filter(symbol.y == "SLC5A1") %>%
  distinct(symbol.y, read_id, alpha) %>%
  ggplot(aes(x = symbol.y, fill = as.factor(alpha))) +
  geom_bar(position = "fill")
```
```{r}
for_beta_genes[,c(1:16,22:23,26:28)] %>% distinct() %>%
  filter(symbol.y == "SLC5A1") %>% distinct(read_id, alpha) %>% summarise(mean(alpha))
```

* and an alpha gene
```{r}
for_beta_genes[,c(1:16,22:23,26:28)] %>% distinct() %>%
  filter(symbol.y == "FRG1FP") %>% distinct(start, beta) %>% summarise(mean(beta))
```

```{r}
#source("../code/processData.R")
test_reads <- for_beta_genes[,c(1:16,22:23,26:28)] %>% distinct() %>%
  filter(symbol.y == "FRG1FP") %>% distinct(read_id) %>% .$read_id

rrbs_plot(for_beta, read_ids = test_reads) + theme_bw()
rrbs_plot(for_beta, read_ids = test_reads, count_instead = TRUE) + theme_bw()
```
```{r}
for_beta_genes[,c(1:16,22:23,26:28)] %>% distinct() %>%
  filter(symbol.y == "FRG1FP") %>%
  distinct(symbol.y, read_id, alpha) %>%
  ggplot(aes(x = symbol.y, fill = as.factor(alpha))) +
  geom_bar(position = "fill")
```




```{r}
for_beta_genes[,c(1:16,22:23,26:28)] %>% distinct() %>%
  filter(!is.na(symbol.y)) %>%
  distinct(seqnames, start, beta, gene_id.y, symbol.y) %>%
  group_by(gene_id.y) %>%
  mutate(mean_beta = mean(beta), n_cgs= n()) %>% 
  ungroup() %>% distinct(gene_id.y, symbol.y, mean_beta, n_cgs) %>%
  mutate(beta_range = case_when(mean_beta < 0.3 ~ "x < 0.3",
                                mean_beta > 0.7 ~ "x > 0.7",
                                TRUE ~ " 0.3 < x < 0.7"), x = "gene") %>%
  ggplot(aes(x = x, fill = beta_range)) +
  geom_bar(position = "fill")
```


```{r}
for_beta_genes[,c(1:16,22:23,26:28)] %>% distinct() %>%
  filter(!is.na(symbol.y)) %>%
  distinct(seqnames, start, beta, gene_id.y, symbol.y) %>%
  group_by(gene_id.y) %>%
  mutate(mean_beta = mean(beta), n_cgs= n()) %>% 
  ungroup() %>% distinct(gene_id.y, symbol.y, mean_beta, n_cgs) %>%
  mutate(beta_range = case_when(mean_beta >= 0 & mean_beta <= 0.2 ~ "low", 
                                mean_beta >= 0.4 & mean_beta <= 0.6 ~ "medium",
                                mean_beta >= 0.8 & mean_beta <= 1 ~ "high",
                                mean_beta > 0.2 & mean_beta < 0.4 ~ "low_med",
                                TRUE ~ "med_high"), x = "gene") %>%
  filter(beta_range %in% c("low", "medium", "high")) %>%
  ggplot(aes(x = x, fill = beta_range)) +
  geom_bar(position = "fill")
```

# dmr


####

## DMR analysis

```{r}
#gm12878_2 <- readRDS("/researchers/caitlin.page/cf_nano/r_output/rrbs_alpha_beta_rep2.rds")
gm12878_2 <- readRDS("../output/cluster_files/rrbs_alpha_beta_rep2.rds")
#k562 <- readRDS("/researchers/caitlin.page/cf_nano/r_output/k562_rep1.rds")
k562 <- readRDS("../output/cluster_files/k562_rep1.rds")
#k562_2 <- readRDS("/researchers/caitlin.page/cf_nano/r_output/k562_rep2.rds")
k562_2 <- readRDS("../output/cluster_files/k562_rep2.rds")
```

```{r}
dmrcate_seq_dmr
for_beta_all
gm12878_2
```

```{r}
k562 <- k562[,colnames(for_beta_all)]
k562_2 <- k562_2[,colnames(for_beta_all)]
```


```{r}
all_samples <- rbind(for_beta_all %>% mutate(sample = "b1"), gm12878_2 %>% mutate(sample = "b2"), 
                     k562 %>% mutate(sample = "cml1"), k562_2 %>% mutate(sample = "cml2"))
```

```{r}
rrbs_alpha_beta
```

```{r}
find_overlaps(as_granges(all_samples), as_granges(dmrcate_seq_dmr[1,])) %>% data.frame()
```

```{r}
find_overlaps(as_granges(all_samples), as_granges(dmrcate_seq_dmr[1,])) %>% data.frame()
```




```{r}
find_overlaps(as_granges(all_samples), as_granges(dmrcate_seq_dmr[1,])) %>% data.frame() %>%
  distinct(sample, read_id, read_start, read_end, alpha) %>%
  ggplot() +
  geom_segment(aes(x = read_start, xend = read_end, y = alpha, colour = read_id)) +
  theme(legend.position = "none") +
  facet_wrap(~ sample)
```

```{r}
find_overlaps(as_granges(all_samples), as_granges(dmrcate_seq_dmr[1,])) %>% data.frame() %>%
  group_by(sample, read_id, read_start, read_end, alpha) %>%
  summarise(count = n()) %>% ungroup() %>% filter(count > 1)
```


```{r}
find_overlaps(as_granges(all_samples), as_granges(dmrcate_seq_dmr[1,])) %>% data.frame() %>%
  group_by(sample, read_id, read_start, read_end, alpha) %>%
  summarise(count = n()) %>% ungroup() %>%
  ggplot() +
  geom_segment(aes(x = read_start, xend = read_end, y = alpha, colour = read_id, linewidth = count)) +
  theme(legend.position = "none") +
  facet_wrap(~ sample)
```

```{r}
find_overlaps(as_granges(all_samples), as_granges(dmrcate_seq_dmr[1,])) %>% data.frame() %>%
  distinct(sample, read_id, read_start, read_end, alpha) %>%
  ggplot(aes(x = sample, y = alpha, fill = sample)) +
  geom_violin() +
  theme(legend.position = "none")
```


```{r}
find_overlaps(as_granges(all_samples), as_granges(dmrcate_seq_dmr[2,])) %>% data.frame() %>%
  group_by(sample, read_id, read_start, read_end, alpha) %>%
  summarise(count = n()) %>% ungroup() %>%
  ggplot() +
  geom_segment(aes(x = read_start, xend = read_end, y = alpha, colour = read_id, linewidth = count)) +
  theme(legend.position = "none") +
  facet_wrap(~ sample)
```

```{r}
find_overlaps(as_granges(all_samples), as_granges(dmrcate_seq_dmr[2,])) %>% data.frame() %>%
  distinct(sample, read_id, read_start, read_end, alpha) %>%
  ggplot(aes(x = sample, y = alpha, fill = sample)) +
  geom_violin() +
  theme(legend.position = "none")
```


## proportions in dmrs

```{r}
find_overlaps(as_granges(all_samples), as_granges(dmrcate_seq_dmr[1,])) %>% data.frame() %>%
  distinct(sample, read_id, read_start, read_end, alpha) %>% 
  mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8")) %>%
  ggplot(aes(x = sample, fill = alpha_group)) +
  geom_bar(position = "fill")
```

```{r}
find_overlaps(as_granges(all_samples), as_granges(dmrcate_seq_dmr[2,])) %>% data.frame() %>%
  distinct(sample, read_id, read_start, read_end, alpha) %>% 
  mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8")) %>%
  ggplot(aes(x = sample, fill = alpha_group)) +
  geom_bar(position = "fill")
```

# least significant dmr
```{r}
dmrcate_seq_dmr %>% tail()
find_overlaps(as_granges(all_samples), as_granges(dmrcate_seq_dmr[7909,])) %>% data.frame() %>%
  group_by(sample, read_id, read_start, read_end, alpha) %>%
  summarise(count = n()) %>% ungroup() %>%
  ggplot() +
  geom_segment(aes(x = read_start, xend = read_end, y = alpha, colour = read_id, linewidth = count)) +
  theme(legend.position = "none") +
  facet_wrap(~ sample)

find_overlaps(as_granges(all_samples), as_granges(dmrcate_seq_dmr[7909,])) %>% data.frame() %>%
  distinct(sample, read_id, read_start, read_end, alpha) %>%
  ggplot(aes(x = sample, y = alpha, fill = sample)) +
  geom_violin() +
  theme(legend.position = "none")

find_overlaps(as_granges(all_samples), as_granges(dmrcate_seq_dmr[7909,])) %>% data.frame() %>%
  distinct(sample, read_id, read_start, read_end, alpha) %>% 
  mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8")) %>%
  ggplot(aes(x = sample, fill = alpha_group)) +
  geom_bar(position = "fill")
```




## dtu on top dmr
```{r}
dmrs
```

```{r}
dtu_dmr <- find_overlaps(as_granges(all_samples), as_granges(mutate(dmrcate_seq_dmr, rank = paste0("DMR:", 1:nrow(dmrcate_seq_dmr))))) %>% data.frame() %>%
  distinct(sample, read_id, read_start, read_end, alpha, rank) %>% 
  mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8"))

dtu_dmr <- dtu_dmr %>% 
  mutate(transcript = paste0(rank, "_", alpha_group)) %>%
  group_by(sample, transcript) %>%
  summarise(transcript_count = n()) %>% data.frame()
```


```{r}
edger_counts <- c(dtu_for_beta %>% 
  mutate(transcript = paste0(symbol.y, "_", alpha_group)) %>% .$transcript,
  dtu_k562 %>% 
  mutate(transcript = paste0(symbol.y, "_", alpha_group)) %>% .$transcript,
  dtu_beta_2 %>% 
  mutate(transcript = paste0(symbol.y, "_", alpha_group)) %>% .$transcript,
  dtu_k562_2 %>% 
  mutate(transcript = paste0(symbol.y, "_", alpha_group)) %>% .$transcript) %>% data.frame() %>% distinct()
names(edger_counts) <- "transcript"
edger_counts <- edger_counts %>% mutate(Sample1_gm = trans_count_beta[match(.$transcript, trans_count_beta$transcript), "transcript_count"],
                                        Sample2_gm = trans_count_beta2[match(.$transcript, trans_count_beta2$transcript), "transcript_count"],
                        Sample3_k = trans_count_k[match(.$transcript, trans_count_k$transcript), "transcript_count"],
                        Sample4_k = trans_count_k_2[match(.$transcript, trans_count_k_2$transcript), "transcript_count"])
edger_counts[is.na(edger_counts)] <- 0
edger_counts$gene_id <- sub("_.*", "", edger_counts$transcript)
edger_counts <- edger_counts %>% filter(!gene_id == "NA") 
rownames(edger_counts) <- edger_counts$transcript
edger_counts[1:10,]
```

```{r}
y <- DGEList(as.matrix(edger_counts[,2:5]), group = c(1,1,2,2), genes = edger_counts[,c("gene_id", "transcript")])
y

design <- model.matrix(~ c(1,1,2,2), data = y$samples)

fit <- glmQLFit(y, design, robust=TRUE)

ds <- diffSpliceDGE(fit, coef = 2, geneid = "gene_id", exonid = "transcript")
```

```{r}
ts.gene.simes <- topSpliceDGE(ds, test = "Simes", number = Inf)
ts.gene.simes[1:10,]

ts.gene.gene <- topSpliceDGE(ds, test = "gene", number = Inf)
ts.gene.gene[1:10,]
```

```{r}
gene_cg <- find_overlaps(as_granges(annotations), as_granges(cg_sites)) %>% data.frame() %>%
  group_by(gene_id, symbol) %>%
  summarise(n_cg = n()) %>%
  ungroup() %>% data.frame()

fit$genes <- fit$genes %>% mutate(n_cg = gene_cg[match(.$gene_id, gene_cg$symbol), "n_cg"])
```


```{r}
rpk <- y$counts/rep(fit$genes$n_cg/1000, times = ncol(y$counts))
tpm_mat <- sweep(rpk, 2, colSums(rpk)/1e6, "/")

tpm_mat <- tpm_mat %>% data.frame() %>% mutate(id = edger_counts$transcript, gene_id = edger_counts$gene_id)
tpm_mat[1:10,]
```

* now add in proportions
```{r}
tpm_mat <- tpm_mat %>% 
  group_by(gene_id) %>% 
  mutate(prop_S1_gm = Sample1_gm/sum(Sample1_gm),
         prop_S2_gm = Sample2_gm/sum(Sample2_gm),
         prop_S3_k = Sample3_k/sum(Sample3_k),
         prop_S4_k = Sample4_k/sum(Sample4_k)) %>%
  ungroup() %>% data.frame()
tpm_mat %>% .[order(.$gene_id),] %>% .[1:10,]
```
* nice can see even in these 2 genes, difference in proportions between the groups

```{r}
ts.gene.simes[1:10,]
tail(ts.gene.simes)
c(signif = ts.gene.simes %>% filter(FDR < 0.05) %>% nrow(),
  total = nrow(ts.gene.simes))
```
```{r}
tpm_mat %>% filter(gene_id == ts.gene.simes$gene_id[1])
```

```{r}
tpm_mat %>% filter(gene_id == ts.gene.simes$gene_id[643])
```


```{r}
tpm_mat %>% mutate(alpha_group = sub(".*_", "", id)) %>% .[,6:11] %>%
  filter(gene_id %in% ts.gene.simes$gene_id[1:10]) %>%
  reshape2::melt(id = c("alpha_group", "gene_id")) %>%
  ggplot(aes(x = variable, y = value, fill = alpha_group)) +
  geom_col() +
  scale_x_discrete(labels = c("B_cell_1", "B_cell_2", "CML_1", "CML_2")) +
  facet_wrap(~ gene_id)
```

##
# presentation plots

# dist boxplots but combine them




```{r}
rbind(
      for_beta %>% distinct(start, beta) %>% .[order(.$start),] %>% mutate(dist_to_next = lead(start) - start) %>% mutate(variable = start, value = beta) %>% distinct(variable, value, dist_to_next) %>% mutate(technology = "RRBS", method = "beta"),
      for_beta %>% group_by(read_id) %>% mutate(dist_to_next = lead(start) - start) %>% ungroup() %>% mutate(variable = read_id, value = alpha) %>% distinct(variable, value, dist_to_next) %>% mutate(technology = "RRBS", method = "alpha")) %>% 
  filter(value %in% c(0,0.5,1)) %>%
  ggplot(aes(x = as.factor(value), y = log2(dist_to_next), fill = method)) +
  geom_boxplot() +
  theme_bw() 
  theme(legend.position = "none")
```


```{r}
for_beta %>% distinct(start, beta) %>% .[order(.$start),] %>% mutate(dist_to_next = lead(start) - start) %>%
  filter(beta %in% c(0,0.5,1)) %>%
  ggplot(aes(x = as.factor(beta), y = log2(dist_to_next), fill = as.factor(beta))) +
  geom_boxplot() +
  theme_bw() +
  theme(legend.position = "none")
```

```{r}
for_beta %>% filter(beta == 0.5) %>% distinct(read_id, alpha, beta) %>%
  ggplot(aes(x = alpha)) +
  geom_density() +
  theme_bw() +
  theme(legend.position = "none")

for_beta %>% filter(beta > 0.4, beta < 0.6) %>% distinct(read_id, alpha, beta) %>%
  ggplot(aes(x = alpha)) +
  geom_density() +
  theme_bw() +
  theme(legend.position = "none")
```
```{r}
for_beta %>% distinct(read_id, alpha, total) %>%
  ggplot(aes(x = total)) +
  geom_histogram()


for_beta %>% distinct(read_id, alpha, total) %>% filter(total > 1) %>%
  ggplot(aes(x = as.factor(total), fill = as.factor(total))) +
  geom_bar() +
  theme_bw() +
  theme(legend.position = "none")
```


```{r}
for_beta %>% filter(beta == 0.5) %>% distinct(read_id, alpha, beta) %>%
  ggplot(aes(x = as.factor(beta), y = alpha)) +
  geom_violin(fill = "lavender") +
  theme_bw() +
  theme(legend.position = "none")

for_beta %>% filter(beta != 0.5) %>% distinct(read_id, alpha, beta) %>% mutate(a = "not_0.5") %>%
  ggplot(aes(x = a, y = alpha)) +
  geom_violin(fill = "lavender") +
  theme_bw() +
  theme(legend.position = "none")
```
- default red group: #F8766D
- default blue: #619CFF - actually it's not the colour I want
-well it's the blue in the boxplot but i want the blue from the violin plot
- default green group: #00BA38


```{r}
for_beta %>% distinct(read_id, alpha, beta) %>% mutate(beta_0.5 = ifelse(beta == 0.5, TRUE, FALSE)) %>%
  ggplot(aes(x = beta_0.5, y = alpha, fill = beta_0.5)) +
  geom_violin() +
  theme_bw() +
  theme(legend.position = "none")
```

```{r}
for_beta %>% distinct(read_id, alpha, beta) %>% mutate(beta_0.4_0.6 = ifelse(beta > 0.4 & beta < 0.6, TRUE, FALSE)) %>%
  ggplot(aes(x = beta_0.4_0.6, y = alpha, fill = beta_0.4_0.6)) +
  geom_violin() +
  theme_bw() +
  theme(legend.position = "none")
```

