---
title: "rrbsData"
author: "Caitlin Page"
date: "2025-10-08"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

## Introduction

# cancer cell line encyclopedia
https://sites.broadinstitute.org/ccle/datasets
https://depmap.org/portal/data_page/?tab=allData
promoter data 
```{r}
read.table("../data/rrbs/dep_map_cancer_cell_line/CCLE_RRBS_TSS_1kb_20180614.txt", header = TRUE)
```
no I don't like this I don't understand it and it's too processed

[raw data](https://www.ncbi.nlm.nih.gov/sra/?term=PRJNA523380+rrbs)
except have to go through sra and it's fastq
I would prefer bam

# breast cancer
https://www.nature.com/articles/s41467-021-25661-w#data-availability

just has a link to download processed methylation values so we'll see
also has fastq upon request

# healthy indiv neutrophils rrbs
Genome-scale DNA methylome and transcriptome profiling of human neutrophils
https://www.nature.com/articles/sdata201619#Sec12
https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE59163
* potentially
13 samples
* GSE table files
```{r}
read.table("../data/rrbs/neutrophil_paper_2016/GSE59163_RAW/GSM1429638_X9006_ad3tr65.fastq_bismark.sam_methpcNF2t10.txt.gz")
```
* damn no column names
* guess: v4 = num meth, v5 = num unmeth, v7 = beta as %
```{r}
(92)/(92+8) * 100
52/(52+3)*100
```
* nope they don't match

# rrbsdata (bioconductor package)
https://bioconductor.org/packages/release/data/experiment/html/RRBSdata.html
has 12 samples
this is the paper they got the data from: https://ashpublications.org/blood/article/121/1/178/31062/DNA-methylation-changes-are-a-late-event-in-acute

```{r}
renv::install("bioc::RRBSdata")
```
```{r}
library(BiSeq)
```
```{r}
data(rrbs)
rrbs
```
```{r}
colData(rrbs)
```
APL - acute promyelocytic leukaemia

```{r}
rrbs@rowRanges
```
* it's not read level info though - not bams, just cpgs




# paper - BSseq (pancreatic cancer)
[GEO](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE63123)
has 19 samples
* 4 individuals, diff replicates and situations
* processed with hg19

with bams available to download through SRA toolkit

so it's not actually RRBS data - but it's data and it exists so it's good enough

downloading 4 samples via tmux and file transfer server
all from sample A_38
Lg is lung metastasis
Per is peritoneal metastasis

* SRR1646792: Lg rep 1
* SRR1646793: Lg rep 2
* SRR1646794: Per rep 1
* SRR1646795: Per rep 2

* oops there's 2 more for this sample: 2 liver metastasis: SRR3427332, SRR3427334

* also all of these are massive files
2 fastqs each, 93 is 51.8gb per fastq - thank goodness for gz compression

Steps to download fastqs
SRA run explorer
* Save SRR accession numbers
* FastQ downloads
* Aspera commands - Copy

Steps to process FastQs
(may as well follow the paper the data is from)
[paper](https://www.nature.com/articles/ng.3753#Sec8)
* 100 base pair seq
* BSmooth for alignment
* Bowtie2 for alignment (says this immediately after saying using BSmooth for alignment)
** BSmooth offers 2 options for alignment - 1 of them uses bowtie (so that's what the paper used)
* and the lambda phage genome (hold on this is all supposed to be human)
* got meth measurement for each CpG
* filtering: mapping qual < 20, nuc base qual at cyt position < 10
* filtering: removed meas from 5' most 10nt of both paired reads
* DMRs: bsseq
* DMRs: >= 3 reads per CpG
* DMRs: small: smooth window 20 CpGs, 1kb used, t stat 4.6, meth diff > 20%
* DMRs: big: smooth window 200 CpGs, 10kbp, t stat 2, meth diff > 10%, length dmr > 5kb


WHAT I DID:
I had to make a conda environment which was annoying
oh yeah i ran bwa-meth
to make sams to bams
and then align the bams
samtools sort
samtools index
modkit extract calls

```{r}
list.files("/researchers/caitlin.page/pancreatic_cancer_wgbs")
```


```{r}
fread(paste0("/researchers/caitlin.page/pancreatic_cancer_wgbs/", list.files("/researchers/caitlin.page/pancreatic_cancer_wgbs")[4]))
```

I missed a step
b/c this is BS not nanopore
I need a tool like bismark to process the bams
and modkit is built for nanopore
cancerdetector used bismark to process
have to hope this will get me read level info

it wants the bam file but outputs are currently 0bytes
i think i need to realign just using bismark

ok can't use sorted file (by pos)
can be sorted by read (samtools sort -n)

using unsorted bam got some some for m bias and splitting report, but still nothing for cpg report

try to align with bismark
[Bismark steps](https://felixkrueger.github.io/Bismark/quick_reference/)
Bismark genome preparation:
* made a genome index folder in /researchers/caitlin.page called Bisulfite_Genome
Bismark
* aligned the fastq file
* resulting bam stored in /researchers/caitlin.page/pancreatic_cancer_wgbs/bismark_res
bismark_methylation_extractor
* running now on bam file
