---
title: "rrbsData"
author: "Caitlin Page"
date: "2025-10-08"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

## Introduction

# cancer cell line encyclopedia
https://sites.broadinstitute.org/ccle/datasets
https://depmap.org/portal/data_page/?tab=allData
promoter data 
```{r}
read.table("../data/rrbs/dep_map_cancer_cell_line/CCLE_RRBS_TSS_1kb_20180614.txt", header = TRUE)
```
no I don't like this I don't understand it and it's too processed

[raw data](https://www.ncbi.nlm.nih.gov/sra/?term=PRJNA523380+rrbs)
except have to go through sra and it's fastq
I would prefer bam

# breast cancer
https://www.nature.com/articles/s41467-021-25661-w#data-availability

just has a link to download processed methylation values so we'll see
also has fastq upon request

# healthy indiv neutrophils rrbs
Genome-scale DNA methylome and transcriptome profiling of human neutrophils
https://www.nature.com/articles/sdata201619#Sec12
https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE59163
* potentially
13 samples
* GSE table files
```{r}
read.table("../data/rrbs/neutrophil_paper_2016/GSE59163_RAW/GSM1429638_X9006_ad3tr65.fastq_bismark.sam_methpcNF2t10.txt.gz")
```
* damn no column names
* guess: v4 = num meth, v5 = num unmeth, v7 = beta as %
```{r}
(92)/(92+8) * 100
52/(52+3)*100
```
* nope they don't match

# rrbsdata (bioconductor package)
https://bioconductor.org/packages/release/data/experiment/html/RRBSdata.html
has 12 samples
this is the paper they got the data from: https://ashpublications.org/blood/article/121/1/178/31062/DNA-methylation-changes-are-a-late-event-in-acute

```{r}
renv::install("bioc::RRBSdata")
```
```{r}
library(BiSeq)
```
```{r}
data(rrbs)
rrbs
```
```{r}
colData(rrbs)
```
APL - acute promyelocytic leukaemia

```{r}
rrbs@rowRanges
```
* it's not read level info though - not bams, just cpgs
