---
title: "runOpTiles"
author: "caitlinpage"
date: "2026-04-10"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

## Introduction


```{r}
seqs <- seqnames(BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38)[1:22]

cg_sites <- lapply(seqs, function(x) {
        cbind(data.frame(Biostrings::matchPattern("CG", BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38[[x]])),
            seqnames = x
        )}) %>% bind_rows()
cg_sites$index <- 1:nrow(cg_sites)
```
```{r}
list.files("/researchers/caitlin.page/phd/cf_nano/r_output")
```

devtools::install_github("kassambara/easyGgplot2") - dependency
renv::install("bioc::genomation", "bioc::methylKit") - other dependency
* already installed dependencies: "GenomicRanges", "IRanges", "biomaRt", "limma"
devtools::install_github("fhaive/OpTiles", build = F, type = "source")

* but installed with renv


```{r}
library(annotatr)
library(data.table)
library(plyranges)
library(tidyr)
library(stringr)
library(dplyr)
library(ggplot2)

library(OpTiles)
```

## Data Objects
```{r}
seqs <- seqnames(BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38)[1:22]

cg_sites <- lapply(seqs, function(x) {
        cbind(data.frame(Biostrings::matchPattern("CG", BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38[[x]])),
            seqnames = x
        )}) %>% bind_rows()
cg_sites$index <- 1:nrow(cg_sites)
```

```{r}
seqs <- seqnames(BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38)[1:22]
```

```{r}
for_beta_all <- readRDS("/researchers/caitlin.page/cf_nano/r_output/rrbs_alpha_beta.rds")
#for_beta <- readRDS("../output/cluster_files/rrbs_alpha_beta.rds")
for_beta_all <- for_beta_all %>% filter(seqnames %in% seqs)
for_beta_all <- for_beta_all %>% group_by(seqnames, start) %>% mutate(coverage = n(), mean_alpha = mean(alpha)) %>% ungroup() %>% data.frame()
just_beta <- for_beta_all %>% distinct(seqnames, start, beta, coverage, mean_alpha)

```


```{r}
analyze_methylation_stats(
  meth = just_beta,
  variable_to_investigate = "coverage", 
  position_text = 100)

just_beta %>%
  ggplot(aes(x = coverage)) +
  geom_histogram(bins = 75)
```
```{r}
betavalue <- just_beta$beta
betavalue_df <- as.data.frame(betavalue)
rownames(betavalue_df) <- paste0(just_beta$seqnames, ":", just_beta$start, "-", just_beta$start + 1)
```
```{r}
metadata <- "sample_gm12878"
metadata_df <- betavalue_df %>% mutate(sample = "gm12878_1")
```

```{r}
analyze_methylation_stats(
  meth = betavalue_df,
  variable_to_investigate = "betavalue", 
  position_text = 60,
  metadata = metadata_df,
  sample_name_variable = "sample")
```


```{r}
length_region <- 300
tiles <- methylKit::tileMethylCounts(just_beta,win.size=length_region,step.size=length_region,cov.bases = 0)
```

```{r}
for_op <- just_beta %>% mutate(end = start, strand = "*", numCs = beta * coverage, numTs = coverage - numCs, chr = seqnames) %>% .[,c(10,2,6,7,4,8,9)]
mr <- new(
  "methylRaw",
  for_op,
  sample.id = "sample1",
  assembly = "hg38",
  context = "CpG",
  resolution = "base"
)
```

```{r}
length_region <- 300
tiles <- methylKit::tileMethylCounts(mr,win.size=length_region,step.size=length_region,cov.bases = 0)
```
```{r}
?methylKit::tileMethylCounts
```


```{r}
tilesdf <- methylKit::getData(tiles)
```

```{r}
betavalue_tiles <- percMethylation(tiles)
betavalue_tilesdf <- as.data.frame(betavalue_tiles)
location_all_regions <- paste0("chr", tilesdf$chr, ":", tilesdf$start, "-", tilesdf$end)
rownames(betavalue_tilesdf) <- location_all_regions
```

# this is the optimisation step
```{r}
regions_plus_merged_regions <- merging_consecutive_regions(mr, 
                                                           tiles,
                                                           da_thr = length_region,
                                                           cb_thr = length_region)
```

```{r}
tilesdf <- tilesdf[,1:4] %>% mutate(width = end - start + 1) 
colnames(tilesdf)[1] <- "seqnames"
tilesdf$id <- 1:nrow(tilesdf)
```


```{r}
regions_plus_merged_regions <- regions_plus_merged_regions %>% mutate(width = end - start + 1)
colnames(regions_plus_merged_regions)[1] <- "seqnames"
```

* ok so regions have a set length but the positioning is changed
* supposedly it's better at distribution of cpgs - so maybe compare num cpgs in regions + dist between them

```{r}
find_overlaps(as_granges(tilesdf), as_granges(mutate(just_beta, end = start))) %>% data.frame() %>% 
  group_by(seqnames, start, end, width, id) %>%
  summarise(n_cg = n()) %>%
  ggplot(aes(x = n_cg)) +
  geom_histogram()
```

```{r}
find_overlaps(as_granges(regions_plus_merged_regions), as_granges(mutate(just_beta, end = start))) %>% data.frame() %>% 
  group_by(seqnames, start, end, width, region_id) %>%
  summarise(n_cg = n()) %>%
  ggplot(aes(x = n_cg)) +
  geom_histogram()
```

```{r}
just_beta <- just_beta %>% group_by(seqnames) %>% .[order(.$start),] %>% mutate(dist_to_next = lead(start) - start) %>% ungroup()
```


```{r}
find_overlaps(as_granges(tilesdf), as_granges(mutate(just_beta, end = start))) %>% data.frame() %>%
  group_by(seqnames, start, end, width, id) %>%
  summarise(n_cg = n(), mean_dist = mean(dist_to_next), median_dist = median(dist_to_next), var_dist = var(dist_to_next), sd_dist = sd(dist_to_next)) %>%
  ggplot(aes(x = log2(mean_dist))) +
  geom_density()
```


```{r}
find_overlaps(as_granges(regions_plus_merged_regions), as_granges(mutate(just_beta, end = start))) %>% data.frame() %>%
  group_by(seqnames, start, end, width, region_id) %>%
  summarise(n_cg = n(), mean_dist = mean(dist_to_next), median_dist = median(dist_to_next), var_dist = var(dist_to_next), sd_dist = sd(dist_to_next)) %>%
  ggplot(aes(x = log2(mean_dist))) +
  geom_density()
```


```{r}
annots <- c("hg38_cpgs")

annotations <- annotatr::build_annotations(genome = 'hg38', annotations = annots)
seqs <- annotations %>% data.frame() %>% distinct(seqnames) %>% .[1:22,]
annotations <- annotations %>% data.frame() %>% filter(seqnames %in% seqs)

for_beta_all <- find_overlaps(as_granges(for_beta_all), as_granges(annotations)) %>% data.frame() %>%
  mutate(cgi = ifelse(type == "hg38_cpg_islands", "island", "not_island"))
```

```{r}
annotations %>% filter(type == "hg38_cpg_islands")
```
*26,866 islands
```{r}
nrow(tilesdf)
nrow(regions_plus_merged_regions)
```

```{r}
find_overlaps(as_granges(tilesdf), as_granges(annotations)) %>% data.frame() %>%
  group_by(type) %>% summarise(n=n()) %>% ungroup() %>% mutate(prop = n/sum(n))
```

```{r}
find_overlaps(as_granges(regions_plus_merged_regions), as_granges(annotations)) %>% data.frame() %>%
  group_by(type) %>% summarise(n=n()) %>% ungroup() %>% mutate(prop = n/sum(n))
```

```{r}
rbind(find_overlaps(as_granges(regions_plus_merged_regions), as_granges(annotations)) %>% data.frame() %>%
  group_by(type) %>% summarise(n=n()) %>% ungroup() %>% mutate(prop = n/sum(n)) %>% mutate(version = "optiles"),
  find_overlaps(as_granges(tilesdf), as_granges(annotations)) %>% data.frame() %>%
  group_by(type) %>% summarise(n=n()) %>% ungroup() %>% mutate(prop = n/sum(n)) %>% mutate(version = "original")) %>% group_by(type) %>% summarise(diff = diff(n))
```


```{r}
find_overlaps(as_granges(tilesdf), as_granges(annotations)) %>% data.frame() %>% distinct(id.x) %>% nrow()/nrow(tilesdf)
find_overlaps(as_granges(regions_plus_merged_regions), as_granges(annotations)) %>% data.frame() %>% distinct(region_id) %>% nrow()/nrow(regions_plus_merged_regions)
```


```{r}
for_op %>% filter(chr == "chr1") %>% .[order(.$start),]
```

# DTU on TILE REGIONS
```{r}
library(edgeR)
```

```{r}
rrbs_alpha_beta <- readRDS("/researchers/caitlin.page/cf_nano/r_output/rrbs_alpha_beta.rds")

gm12878_2 <- readRDS("/researchers/caitlin.page/cf_nano/r_output/rrbs_alpha_beta_rep2.rds")

k562 <- readRDS("/researchers/caitlin.page/cf_nano/r_output/k562_rep1.rds")

k562_2 <- readRDS("/researchers/caitlin.page/cf_nano/r_output/k562_rep2.rds")
```

```{r}
k562 <- k562[,colnames(rrbs_alpha_beta)]
k562_2 <- k562_2[,colnames(rrbs_alpha_beta)]
```

```{r}
cols <- c("read_id", "seqnames", "start", "end", "strand", "meth_pattern", "num_meth", "total", "alpha", "read_start", "read_end", "beta", "sample")
```


```{r}
all_samples <- rbind(rrbs_alpha_beta %>% mutate(sample = "b1") %>% .[,cols], gm12878_2 %>% mutate(sample = "b2") %>% .[,cols], 
                     k562 %>% mutate(sample = "cml1") %>% .[,cols], k562_2 %>% mutate(sample = "cml2") %>% .[,cols])
```

```{r}
dtu_optile <- find_overlaps(as_granges(all_samples), as_granges(regions_plus_merged_regions)) %>% data.frame() %>%
  distinct(sample, read_id, read_start, read_end, alpha, region_id) %>% 
  mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                alpha > 0.8 ~ "a > 0.8",
                                alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                TRUE ~ "0.6 <= a <= 0.8"))

dtu_optile <- dtu_optile %>% 
  mutate(transcript = paste0(region_id, "_", alpha_group)) %>%
  group_by(sample, transcript) %>%
 summarise(transcript_count = n()) %>% ungroup() %>% data.frame()
```

```{r}
a <- dtu_optile %>% filter(sample == "b1") %>% .[,2:3]
aa <- dtu_optile %>% filter(sample == "b2") %>% .[,2:3]
aaa <- dtu_optile %>% filter(sample == "cml1") %>% .[,2:3]
aaaa <- dtu_optile %>% filter(sample == "cml2") %>% .[,2:3]


dtu_optile <- distinct(data.frame(transcript = c(a$transcript, aa$transcript, aaa$transcript, aaaa$transcript))) %>%
  mutate(Sample_b1 = a[match(.$transcript, a$transcript), "transcript_count"],
         Sample_b2 = aa[match(.$transcript, aa$transcript), "transcript_count"],
         Sample_cml1 = aaa[match(.$transcript, aaa$transcript), "transcript_count"],
         Sample_cml2 = aaaa[match(.$transcript, aaaa$transcript), "transcript_count"]) 
dtu_optile[is.na(dtu_optile)] <- 0
dtu_optile$gene_id <- sub("_.*", "", dtu_optile$transcript)
```

```{r}
y <- DGEList(as.matrix(dtu_optile[,2:5]), group = c(1,1,2,2), genes = dtu_optile[,c("gene_id", "transcript")])
y

design <- model.matrix(~ c(1,1,2,2), data = y$samples)

fit <- glmQLFit(y, design, robust=TRUE)

ds <- diffSpliceDGE(fit, coef = 2, geneid = "gene_id", exonid = "transcript")
```

```{r}


ts.gene_optile <- topSpliceDGE(ds, test = "gene", number = Inf)
ts.gene_optile[1:10,]
```

```{r}
tpm_mat
```


```{r}
seqs <- seqnames(BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38)[1:22]

df <- lapply(seqs, function(x) {
        cbind(data.frame(Biostrings::matchPattern("CG", BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38[[x]])),
            seqnames = x
        )}) %>% bind_rows()
cg_sites <- df
```

```{r}
gene_cg <- find_overlaps(as_granges(regions_plus_merged_regions), as_granges(cg_sites)) %>% data.frame() %>%
  group_by(region_id) %>%
  summarise(n_cg = n()) %>%
  ungroup() %>% data.frame()

fit$genes <- fit$genes %>% mutate(n_cg = gene_cg[match(.$gene_id, gene_cg$region_id), "n_cg"])
```
```{r}
fit$genes %>% filter(!is.na(n_cg))
```

```{r}
gene_cg$region_id %in% fit$genes$gene_id %>% summary()
```


```{r}
rpk <- y$counts/rep(fit$genes$n_cg/1000, times = ncol(y$counts))
tpm_mat <- sweep(rpk, 2, colSums(rpk)/1e6, "/")

tpm_mat <- tpm_mat %>% data.frame() %>% mutate(gene_id = ts.gene_optile$gene_id)
tpm_mat[1:10,]
```

* now add in proportions
```{r}
tpm_mat <- tpm_mat %>% 
  group_by(gene_id) %>% 
  mutate(prop_S1_gm = Sample_b1/sum(Sample_b1),
         prop_S2_gm = Sample_b2/sum(Sample_b2),
         prop_S3_k = Sample_cml1/sum(Sample_cml1),
         prop_S4_k = Sample_cml2/sum(Sample_cml2)) %>%
  ungroup() %>% data.frame()
tpm_mat %>% .[order(.$gene_id),] %>% .[1:10,]
```
* nice can see even in these 2 genes, difference in proportions between the groups

```{r}
ts.gene.gene %>% filter(FDR < 0.05)
```




```{r}
tpm_mat %>% mutate(alpha_group = sub(".*_", "", id)) %>% .[,6:11] %>%
  filter(gene_id %in% ts.gene_optile$gene_id[1:10]) %>%
  reshape2::melt(id = c("alpha_group", "gene_id")) %>%
  ggplot(aes(x = variable, y = value, fill = alpha_group)) +
  geom_col() +
  scale_x_discrete(labels = c("B_cell_1", "B_cell_2", "CML_1", "CML_2")) +
  facet_wrap(~ gene_id)
```

* overlap dtu regions

```{r}
ts.gene_optile <- ts.gene_optile %>% 
  separate_wider_delim(gene_id, delim = ":", names = c("seqnames", "gene_id")) %>%
  mutate(seqnames = gsub("chrchr", "chr", seqnames)) %>%
  separate_wider_delim(gene_id, delim = "-", names = c("start", "end"))
ts.gene_optile$start <- as.double(ts.gene_optile$start)
ts.gene_optile$end <- as.double(ts.gene_optile$end)
ts.gene_optile$dtu_rank <- 1:nrow(ts.gene_optile)
```

```{r}
find_overlaps(as_granges(ts.gene.gene), as_granges(annotations)) %>% data.frame() %>% filter(FDR < 0.05) %>%
  ggplot(aes(x = type)) +
  geom_bar()
```

```{r}
annots <- c("hg38_basicgenes", "hg38_genes_intergenic")

annotations_genes <- annotatr::build_annotations(genome = 'hg38', annotations = annots)
seqs <- annotations %>% data.frame() %>% distinct(seqnames) %>% .[1:22,]
annotations_genes <- annotations_genes %>% data.frame() %>% filter(seqnames %in% seqs)
```

```{r}
find_overlaps(as_granges(ts.gene.gene), as_granges(annotations_genes)) %>% data.frame() %>% filter(FDR < 0.05) %>%
  ggplot(aes(x = type)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

* compare back to dmrs

```{r}
dmrcate_seq_dmr <- readRDS("/researchers/caitlin.page/cf_nano/r_output/dmr_encode_gm12878_k562.rds")
dmrcate_seq_dmr$dmr_rank <- 1:nrow(dmrcate_seq_dmr)
find_overlaps(as_granges(dmrcate_seq_dmr), as_granges(annotations_genes)) %>% data.frame() %>%
  ggplot(aes(x = type)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
* this is all artific inflated because only 7k dmrs, but genomic features smaller - so same dmr - multiple feat - I think

```{r}
find_overlaps(as_granges(dmrcate_seq_dmr), as_granges(ts.gene.gene)) %>% data.frame() %>% 
  mutate(dtu_signif = ifelse(FDR < 0.05, TRUE, FALSE)) %>%
  ggplot(aes(x = dmr_rank, y = dtu_rank, colour = dtu_signif)) +
  geom_point(alpha = 0.3)


find_overlaps(as_granges(dmrcate_seq_dmr), as_granges(ts.gene.gene)) %>% data.frame() %>% filter(FDR < 0.05) %>%
  ggplot(aes(x = dmr_rank, y = dtu_rank)) +
  geom_point(alpha = 0.3)
```

# visually compare old and new tiles

```{r}
rbind(regions_plus_merged_regions %>% mutate(version = "optiles"),
      tilesdf %>% mutate(region_id = paste0("chr", seqnames, ":", start, "-", end), version = "original") %>% .[,c(1:3,7,8)]) %>% data.frame() %>%
  filter(seqnames == "chr1", end < 5000000) %>%
  mutate(version_num = ifelse(version == "optiles", 1, 0.5), 
         dx = runif(n(), -0.1, 0.1),
         version_num = version_num + dx) %>%
  ggplot() +
  geom_segment(aes(x = start, xend = end, y = version_num, yend = version_num, colour = version), linewidth = 2) +
  scale_x_continuous(limits = c(1050000, 1060000))
```

```{r}
rbind(regions_plus_merged_regions %>% mutate(version = "optiles"),
      tilesdf %>% mutate(region_id = paste0("chr", seqnames, ":", start, "-", end), version = "original") %>% .[,c(1:3,7,8)]) %>% data.frame() %>%
  filter(seqnames == "chr1") %>% group_by(start) %>% mutate(count = n()) %>% filter(count == 2)
```

```{r}
tiles_combined <- rbind(regions_plus_merged_regions %>% mutate(version = "optiles"),
      tilesdf %>% mutate(region_id = paste0("chr", seqnames, ":", start, "-", end), version = "original") %>% .[,c(1:3,7,8)]) %>% data.frame() %>%
  filter(seqnames == "chr1") %>% group_by(start) %>% mutate(count = n()) %>% .[order(.$start),] %>% ungroup() %>% 
  mutate(version = ifelse(count == 2, "both", version)) %>% distinct()
```

```{r}
find_overlaps(as_granges(tiles_combined), as_granges(annotations_genes[,c(1:4,9:10)])) %>% data.frame() %>%
  ggplot(aes(x = version, fill = type)) +
  geom_bar()

find_overlaps(as_granges(tiles_combined), as_granges(annotations_genes[,c(1:4,9:10)])) %>% data.frame() %>%
  ggplot(aes(x = version, fill = type)) +
  geom_bar(position = "fill")
```

* no real significant difference

```{r}
find_overlaps(as_granges(tiles_combined), as_granges(annotations_genes[,c(1:4,9:10)])) %>% data.frame() %>%
  ggplot(aes(fill = version, x = type)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

find_overlaps(as_granges(tiles_combined), as_granges(annotations_genes[,c(1:4,9:10)])) %>% data.frame() %>%
  ggplot(aes(x = type, fill = version)) +
  geom_bar(position = "fill") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
for_beta_all %>% distinct(read_id, alpha, symbol) %>% filter(symbol %in% ts.gene.gene$gene_id) %>% 
  group_by(symbol) %>% summarise(mean_alpha = mean(alpha)) %>% data.frame()
```

```{r}
find_overlaps(as_granges(regions_plus_merged_regions), as_granges(distinct(all_samples, sample, read_id, read_start, read_end, alpha))) %>% data.frame() %>%
  group_by(region_id, sample) %>% summarise(mean_alpha = mean(alpha)) %>%
  pivot_wider(names_from = sample, values_from = mean_alpha)
```
```{r}
tpm_mat
```

```{r}
ts.gene.gene %>% mutate(change = rowMeans(.[,c("start", "end")]) - rowMeans(.[,c("NExons", "gene.F")]))
```


```{r}
all_samples
```

# compare density of tiling measures
```{r}
just_beta$cpg_id <- 1:nrow(just_beta)
just_beta$end <- just_beta$start + 1
```

```{r}
find_overlaps(as_granges(tilesdf), as_granges(just_beta)) %>% data.frame() %>% group_by(id) %>% summarise(count_cpg = n()) %>% 
  ggplot(aes(x = count_cpg)) +
  geom_histogram()
regions_plus_merged_regions
```

```{r}
find_overlaps(as_granges(regions_plus_merged_regions), as_granges(just_beta)) %>% data.frame() %>% group_by(region_id) %>% summarise(count_cpg = n()) %>% 
  ggplot(aes(x = count_cpg)) +
  geom_histogram()
```

```{r}
rbind(find_overlaps(as_granges(mutate(regions_plus_merged_regions[,1:3], id = 1:nrow(regions_plus_merged_regions))), as_granges(just_beta)) %>% data.frame() %>% group_by(id) %>% summarise(count_cpg = n()) %>% ungroup() %>% mutate(type = "optimised"),
      find_overlaps(as_granges(tilesdf[,c(1:3,6)]), as_granges(just_beta)) %>% data.frame() %>% group_by(id) %>% summarise(count_cpg = n()) %>% ungroup() %>% mutate(type = "original_tiles")) %>% 
  ggplot(aes(x = log2(count_cpg), fill = type)) +
  geom_histogram(alpha = 0.5)
```

```{r}
regions_plus_merged_regions
ts.gene_optile %>% filter(FDR < 0.05)
```

```{r}
find_overlaps(as_granges(regions_plus_merged_regions), as_granges(all_samples_simp)) %>% data.frame() %>% group_by(region_id) %>% summarise(n_reads = n()) %>%
  ggplot(aes(x = log2(n_reads))) +
  geom_density()
```

# more basic analysis of tiles

```{r}
regions_plus_merged_regions
reads_simp <- rrbs_alpha_beta %>% distinct(read_id, seqnames, read_start, read_end, num_meth, total, alpha)
colnames(reads_simp)[3:4] <- c("start", "end")
```

```{r}
tile_overlap <- find_overlaps(as_granges(reads_simp), as_granges(regions_plus_merged_regions)) %>% data.frame()
tile_overlap %>% group_by(read_id) %>% summarise(n_tile = n()) %>%
  ggplot(aes(x = n_tile)) +
  geom_histogram()
```

```{r}
tile_overlap %>% group_by(read_id) %>% summarise(n_tile = n()) %>% filter(n_tile != 1) %>% distinct(read_id)
```

```{r}
105517/nrow(just_reads) * 100
```



# how do tiles compare when using different sample as input??

```{r}
seqs <- seqnames(BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38)[1:22]

cg_sites <- lapply(seqs, function(x) {
        cbind(data.frame(Biostrings::matchPattern("CG", BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38[[x]])),
            seqnames = x
        )}) %>% bind_rows()
cg_sites$index <- 1:nrow(cg_sites)
```

```{r}
seqs <- seqnames(BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38)[1:22]
```

```{r}
for_beta_all <- readRDS("/researchers/caitlin.page/cf_nano/r_output/rrbs_alpha_beta.rds")
#for_beta <- readRDS("../output/cluster_files/rrbs_alpha_beta.rds")
for_beta_all <- for_beta_all %>% filter(seqnames %in% seqs)
for_beta_all <- for_beta_all %>% group_by(seqnames, start) %>% mutate(coverage = n(), mean_alpha = mean(alpha)) %>% ungroup() %>% data.frame()
just_beta <- for_beta_all %>% distinct(seqnames, start, beta, coverage, mean_alpha)

```

```{r}
#rrbs_alpha_beta <- readRDS("/researchers/caitlin.page/cf_nano/r_output/rrbs_alpha_beta.rds")

gm12878_2 <- readRDS("/researchers/caitlin.page/cf_nano/r_output/rrbs_alpha_beta_rep2.rds")
gm12878_2 <- gm12878_2 %>% filter(seqnames %in% seqs)
gm12878_2 <- gm12878_2 %>% group_by(seqnames, start) %>% mutate(coverage = n(), mean_alpha = mean(alpha)) %>% ungroup() %>% data.frame()
just_beta_gm2 <- gm12878_2 %>% distinct(seqnames, start, beta, coverage, mean_alpha)

k562 <- readRDS("/researchers/caitlin.page/cf_nano/r_output/k562_rep1.rds")
k562 <- k562 %>% filter(seqnames %in% seqs)
k562 <- k562 %>% group_by(seqnames, start) %>% mutate(coverage = n(), mean_alpha = mean(alpha)) %>% ungroup() %>% data.frame()
just_beta_k1 <- k562 %>% distinct(seqnames, start, beta, coverage, mean_alpha)

k562_2 <- readRDS("/researchers/caitlin.page/cf_nano/r_output/k562_rep2.rds")
k562_2 <- k562_2 %>% filter(seqnames %in% seqs)
k562_2 <- k562_2 %>% group_by(seqnames, start) %>% mutate(coverage = n(), mean_alpha = mean(alpha)) %>% ungroup() %>% data.frame()
just_beta_k2 <- k562_2 %>% distinct(seqnames, start, beta, coverage, mean_alpha)
```


```{r}
for_op <- just_beta %>% mutate(end = start, strand = "*", numCs = beta * coverage, numTs = coverage - numCs, chr = seqnames) %>% .[,c(10,2,6,7,4,8,9)]
mr <- new(
  "methylRaw",
  for_op,
  sample.id = "sample1",
  assembly = "hg38",
  context = "CpG",
  resolution = "base"
)

length_region <- 300
tiles <- methylKit::tileMethylCounts(mr,win.size=length_region,step.size=length_region,cov.bases = 0)

tilesdf <- methylKit::getData(tiles)

regions_plus_merged_regions <- merging_consecutive_regions(mr, 
                                                           tiles,
                                                           da_thr = length_region,
                                                           cb_thr = length_region)

tilesdf <- tilesdf[,1:4] %>% mutate(width = end - start + 1) 
colnames(tilesdf)[1] <- "seqnames"
tilesdf$id <- 1:nrow(tilesdf)

regions_plus_merged_regions <- regions_plus_merged_regions %>% mutate(width = end - start + 1)
colnames(regions_plus_merged_regions)[1] <- "seqnames"
```

```{r}
for_op <- just_beta_gm2 %>% mutate(end = start, strand = "*", numCs = beta * coverage, numTs = coverage - numCs, chr = seqnames) %>% .[,c(10,2,6,7,4,8,9)]
mr <- new(
  "methylRaw",
  for_op,
  sample.id = "sample1",
  assembly = "hg38",
  context = "CpG",
  resolution = "base"
)

length_region <- 300
tiles <- methylKit::tileMethylCounts(mr,win.size=length_region,step.size=length_region,cov.bases = 0)

tilesdf_gm2 <- methylKit::getData(tiles)

optiles_gm2 <- merging_consecutive_regions(mr, tiles, da_thr = length_region, cb_thr = length_region)

tilesdf_gm2 <- tilesdf_gm2[,1:4] %>% mutate(width = end - start + 1) 
colnames(tilesdf_gm2)[1] <- "seqnames"
tilesdf_gm2$id <- 1:nrow(tilesdf_gm2)

optiles_gm2 <- optiles_gm2 %>% mutate(width = end - start + 1)
colnames(optiles_gm2)[1] <- "seqnames"
```

# compare the tiles from 2 diff samples
```{r}
regions_plus_merged_regions
optiles_gm2
```



# run tiles on combined all samples

```{r}
for_op <- rbind(just_beta, just_beta_gm2, just_beta_k1, just_beta_k2) %>% distinct()
for_op <- for_op %>% mutate(end = start, strand = "*", numCs = beta * coverage, numTs = coverage - numCs, chr = seqnames) %>% .[,c(10,2,6,7,4,8,9)]
mr <- new(
  "methylRaw",
  for_op,
  sample.id = "sample1",
  assembly = "hg38",
  context = "CpG",
  resolution = "base"
)

length_region <- 300
tiles <- methylKit::tileMethylCounts(mr,win.size=length_region,step.size=length_region,cov.bases = 0)

tilesdf_gm2 <- methylKit::getData(tiles)

optiles_gm2 <- merging_consecutive_regions(mr, tiles, da_thr = length_region, cb_thr = length_region)

tilesdf_gm2 <- tilesdf_gm2[,1:4] %>% mutate(width = end - start + 1) 
colnames(tilesdf_gm2)[1] <- "seqnames"
tilesdf_gm2$id <- 1:nrow(tilesdf_gm2)

optiles_gm2 <- optiles_gm2 %>% mutate(width = end - start + 1)
colnames(optiles_gm2)[1] <- "seqnames"
optiles_all <- optiles_gm2
```

# compare with all to just with one sample
```{r}
regions_plus_merged_regions
optiles_all
```
* way more when you use all sites - which makes sense
* also means that if you use the cpg positions (my ref based thing) - would get even more
* so yeah basically should be doing it like this

```{r}
find_overlaps(as_granges(regions_plus_merged_regions), as_granges(optiles_all)) %>% data.frame() %>% distinct(region_id.x) %>% nrow()/nrow(regions_plus_merged_regions)
find_overlaps(as_granges(regions_plus_merged_regions), as_granges(optiles_all)) %>% data.frame() %>% distinct(region_id.y) %>% nrow()/nrow(optiles_all)
```
* yeah 99% of the regions from sample 1 overlap with regions from the all

```{r}
for_op
```

