---
title: "sexBasedHealthyAnalysis"
author: "caitlinpage"
date: "2026-09-30"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

## Introduction

```{r}
library(BSgenome.Hsapiens.UCSC.hg38)
library(plyranges)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(edgeR)
library(duckdb)
library(arrow)
```
```{r}
# Get chromosome lengths for standard chromosomes only
seqinfo_hg38 <- seqlengths(BSgenome.Hsapiens.UCSC.hg38)
std_chroms <- grepl("^chr([1-9]|1[0-9]|2[0-2]|X|Y|M)$", names(seqinfo_hg38))
seqinfo_hg38 <- seqinfo_hg38[std_chroms]

# Tile into 500 bp non-overlapping windows
tiles_500bp <- GenomicRanges::tileGenome(seqinfo_hg38, tilewidth = 500, cut.last.tile.in.chrom = TRUE)

# Add a tile number column
tiles_500bp <- tiles_500bp %>%
  data.frame() %>%
  mutate(tile_num = 1:n()) %>%
  as_granges()

# cpgs
seqs <- c(seqnames(BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38)[1:22], "chrX", "chrY")

cg_sites <- lapply(seqs, function(x) {
  cbind(data.frame(Biostrings::matchPattern("CG", BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38[[x]])),
        seqnames = x
  )}) %>% bind_rows()

tiles_500bp_cg <- find_overlaps(tiles_500bp, as_granges(cg_sites)) %>% data.frame() %>%
  group_by(tile_num) %>% mutate(n_cg = n())
tiles_500bp_cg <- tiles_500bp_cg %>% ungroup() %>% distinct()

tiles_500bp_cg$tile_id <- paste0(tiles_500bp_cg$seqnames, ":", tiles_500bp_cg$start, "-", tiles_500bp_cg$end)

tiles_500bp_cg
```

```{r}
reads_chr <- function(read_data_path_a, tiles,
                       jit_amount = 0.1) {
  set.seed(42)

  # ── 1. Read only the matching rows — never loads unmatched data into RAM ───
  con <- dbConnect(duckdb())
  on.exit(dbDisconnect(con))
  
  tiles_df <- as.data.frame(tiles)[, c("seqnames", "start", "end", "tile_id")]
  tiles_df$seqnames <- as.character(tiles_df$seqnames)
  duckdb_register(con, "tiles", tiles_df)
  
  overlap_query <- function(path) {
    dbGetQuery(con, glue::glue("
      SELECT DISTINCT r.*, t.tile_id
      FROM read_parquet('{path}') r
      JOIN tiles t
        ON r.seqnames = t.seqnames
       AND r.start >= t.start
       AND r.end   <= t.end
    "))
  }
  
  
  read_data <- overlap_query(read_data_path_a)

  
  read_data$type <- "healthy"

  read_data <- read_data %>% mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                 alpha > 0.8 ~ "a > 0.8",
                                 alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                 alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                 TRUE ~ "0.6 <= a <= 0.8"))
  read_data <- read_data %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8")))
#  read_data <- read_data %>% count(type, region_id, alpha_group, name = "tile_band_count") %>% group_by(region_id, type) %>% mutate(percent = tile_band_count/sum(tile_band_count)) %>% ungroup()
  sample_names <- unique(read_data$sample)
  read_data <- read_data %>% filter(total >= 3) %>% 
    mutate(transcript = paste0(region_id, "_", alpha_group)) %>%
  group_by(sample, transcript) %>%
  summarise(transcript_count = n()) %>% ungroup() %>% data.frame()
  
  read_data <- read_data %>% pivot_wider(names_from = sample, values_from = transcript_count)
  read_data[is.na(read_data)] <- 0
  read_data$gene_id <- sub("_.*", "", read_data$transcript)
  
  c(read_data, sample_names)
}
reads_sex <- reads_chr("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", c("chrX", "chrY"))
```
* ok so in my processed I left out the sex chromosomes - whoopsies that's bad science
```{r}
reads_sex <- reads_chr("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", "/researchers/caitlin.page/cf_nano/r_output/emseq_all_cancer.parquet", c("chr22"))
```

```{r}
reads_chr <- function(read_data_path_a, tiles) {

  # ── 1. Read only the matching rows — never loads unmatched data into RAM ───
  con <- dbConnect(duckdb())
  on.exit(dbDisconnect(con))
  
  tiles_df <- as.data.frame(tiles)[, c("seqnames", "start", "end", "tile_id")]
  tiles_df$seqnames <- as.character(tiles_df$seqnames)
  duckdb_register(con, "tiles", tiles_df)
  
  overlap_query <- function(path) {
    dbGetQuery(con, glue::glue("
      SELECT DISTINCT r.*, t.tile_id
      FROM read_parquet('{path}') r
      JOIN tiles t
        ON r.seqnames = t.seqnames
       AND r.start >= t.start
       AND r.end   <= t.end
    "))
  }
  
  
  read_data <- overlap_query(read_data_path_a)

  
  read_data$type <- "healthy"

  read_data <- read_data %>% mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                 alpha > 0.8 ~ "a > 0.8",
                                 alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                 alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                 TRUE ~ "0.6 <= a <= 0.8"))
  read_data <- read_data %>% mutate(alpha_group = factor(alpha_group, levels = c("a < 0.2", "0.2 <= a <= 0.4", "0.4 < a < 0.6", "0.6 <= a <= 0.8", "a > 0.8")))
#  read_data <- read_data %>% count(type, region_id, alpha_group, name = "tile_band_count") %>% group_by(region_id, type) %>% mutate(percent = tile_band_count/sum(tile_band_count)) %>% ungroup()
  
  read_data
}
reads_sex <- reads_chr("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", sex_tiles)
```

```{r}
reads_sex %>% group_by(sample, seqnames) %>% summarise(n_reads = n()) 
```
* so read counts in X should be double in females than males
* plot
mm
```{r}
reads_sex %>% 
  ggplot(aes(x = sample, fill = seqnames)) +
  geom_bar(position = "dodge")
```
* and prop of sex from total reads
* because yes from this a2 and b2 kinda look male - but also maybe they just have less reads overall
* because it's about the proportion of sex to autosome


```{r}
reads_healthy <- reads_chr("/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet", tiles_500bp_cg)
```

```{r}
con <- dbConnect(duckdb())

counts_df <- dbGetQuery(con, "
  SELECT seqnames, sample, COUNT(*) AS n
  FROM read_parquet('/researchers/caitlin.page/cf_nano/r_output/emseq_healthy.parquet')
  GROUP BY seqnames, sample
  ORDER BY seqnames, sample
")
dbDisconnect(con)

```

```{r}
counts_df %>% filter(seqnames %in% seqs) %>% group_by(sample) %>% mutate(prop = n/sum(n)) %>% 
  ggplot(aes(x = sample, y = n, fill = seqnames)) +
  geom_bar(stat = "identity")
```
* yeah so a2 and b2 do have the lowest counts

```{r}
counts_df %>% filter(seqnames %in% seqs) %>% group_by(sample) %>% mutate(prop = n/sum(n)) %>%
  ggplot(aes(x = sample, y = prop, fill = seqnames)) +
  geom_bar(stat = "identity")

counts_df %>% filter(seqnames %in% seqs) %>% group_by(sample) %>% mutate(prop = n/sum(n)) %>%
  ggplot(aes(x = sample, y = prop, fill = seqnames)) +
  geom_bar(stat = "identity") +
  facet_wrap(~seqnames)
```
```{r}
counts_df %>% filter(seqnames %in% seqs) %>% group_by(sample) %>% mutate(prop = n/sum(n)) %>% filter(seqnames %in% c("chr7", "chrX", "chrY")) %>%
  ggplot(aes(x = sample, y = prop, fill = seqnames)) +
  geom_bar(stat = "identity") +
  facet_wrap(~seqnames)
```
* claude keeps changing its mind about which samples are male/female
* before thought a split
* now thinks all 6 are male
```{r}
reads_sex %>% 
  ggplot(aes(x = alpha, colour = sample)) +
  geom_density() +
  facet_wrap(~seqnames)
```
of the significant regions, what proportion of them are on X
- it's a true positive test
* compare to dmrs again -see if better sensitivity / coverage of x dmr vs dau
* look if pulling up known cometh regions
```{r}
sex_tiles <- tiles_500bp_cg %>% filter(seqnames %in% c("chrX", "chrY"))
```

```{r}
overlap_pos <- find_overlaps(as_granges(filter(optiles_all, seqnames == "chr1")), as_granges(all_samples)) %>% data.frame()
```

```{r}
# dtu
dtu_optile <- overlap_pos %>% filter(total >= 3) %>%
  distinct(sample, read_id, alpha, region_id) %>%
  mutate(alpha_group = case_when(alpha < 0.2 ~ "a < 0.2",
                                 alpha > 0.8 ~ "a > 0.8",
                                 alpha >= 0.2 & alpha <= 0.4 ~ "0.2 <= a <= 0.4",
                                 alpha > 0.4 & alpha < 0.6 ~ "0.4 < a < 0.6",
                                 TRUE ~ "0.6 <= a <= 0.8"))

dtu_optile <- dtu_optile %>%
  mutate(transcript = paste0(region_id, "_", alpha_group)) %>%
  group_by(sample, transcript) %>%
  summarise(transcript_count = n()) %>% ungroup() %>% data.frame()

# build matrix of "transcript" counts - transcript rows, sample as columns
dtu_optile <- dtu_optile %>% pivot_wider(names_from = sample, values_from = transcript_count)
dtu_optile[is.na(dtu_optile)] <- 0
dtu_optile$gene_id <- sub("_.*", "", dtu_optile$transcript)
```

```{r}
unique(all_samples$sample)
```


```{r}
samples <- c(unique(all_samples$sample))  # your 7 sample names

# --- Analysis function ---
run_analysis <- function(combo) {
  cols <- c(combo$g1, combo$g2)
  group <- c(1, 1, 1, 2, 2, 2)

  y <- DGEList(
    as.matrix(dtu_optile[, cols]),
    group = group,
    genes = dtu_optile[, c("gene_id", "transcript")]
  )
  design <- model.matrix(~ group)
  fit <- glmQLFit(y, design, robust = TRUE)
  ds  <- diffSpliceDGE(fit, coef = 2, geneid = "gene_id", exonid = "transcript")

  topSpliceDGE(ds, test = "gene", number = Inf) %>%
    mutate(
      region_id = gene_id,
      excluded  = combo$excluded,
      group1    = paste(combo$g1, collapse = ","),
      group2    = paste(combo$g2, collapse = ",")
    ) %>%
    separate_wider_delim(gene_id, delim = ":", names = c("seqnames", "gene_id")) %>%
    mutate(seqnames = gsub("chrchr", "chr", seqnames)) %>%
    separate_wider_delim(gene_id, delim = "-", names = c("start", "end")) %>%
    mutate(
      start    = as.double(start),
      end      = as.double(end),
      dtu_rank = row_number()
    )
}

# --- Generate 70 combinations ---
all_combos <- unlist(
  lapply(seq_along(samples), function(excl_i) {
    remaining <- samples[-excl_i]
    splits <- combn(remaining, 3, simplify = FALSE)
    splits <- Filter(function(g1) g1[1] < setdiff(remaining, g1)[1], splits)
    lapply(splits, function(g1) {
      list(g1 = g1, g2 = setdiff(remaining, g1), excluded = samples[excl_i])
    })
  }),
  recursive = FALSE
)

# --- Run across all 70 combinations ---
results <- lapply(all_combos, run_analysis)

# --- Combine into one data frame ---
results_df <- bind_rows(results)

saveRDS(results_df, "/researchers/caitlin.page/cf_nano/r_output/dtu_tiles_emseq_healthy_all_fp.rds")
```


