---
title: "slidingWindowAlpha"
author: "caitlinpage"
date: "2026-02-23"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

## Concept
Also sliding windows anchored at a cpg and make fake alphas, only cpgs in that region, and have window same as read length
More controllable to work with
Be able to inform cpgs

```{r}
library(annotatr)
library(data.table)
library(plyranges)
library(tidyr)
library(stringr)
library(dplyr)
library(ggplot2)
```

## Data Objects
```{r}
cg_sites <- cbind(data.frame(Biostrings::matchPattern("CG", BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38[["chr22"]])), seqnames = "chr22") %>% 
  mutate(pos = paste0(seqnames, "-", start)) %>%
  relocate(pos, seqnames)
cg_sites$index <- 1:nrow(cg_sites)
```


```{r}
source("../code/processData.R")
for_beta_all <- readRDS("/researchers/caitlin.page/cf_nano/r_output/rrbs_alpha_beta.rds")
#for_beta <- readRDS("../output/cluster_files/rrbs_alpha_beta.rds")
for_beta <- for_beta_all %>% filter(seqnames == "chr22")
rrbs_for_beta_pat <- pat_for_wgbs(for_beta)

for_beta <- for_beta %>% group_by(start) %>% mutate(coverage = n(), mean_alpha = mean(alpha)) %>% ungroup() %>% data.frame()
just_beta <- for_beta %>% distinct(start, beta, coverage, mean_alpha) %>% .[order(.$start),]

```


```{r}
annots <- c("hg38_cpgs")

annotations <- annotatr::build_annotations(genome = 'hg38', annotations = annots)
seqs <- annotations %>% data.frame() %>% distinct(seqnames) %>% .[1:22,]
annotations <- annotations %>% data.frame() %>% filter(seqnames %in% seqs)
```

```{r}
for_beta <- find_overlaps(as_granges(for_beta), as_granges(annotations)) %>% data.frame() %>%
  mutate(cgi = ifelse(type == "hg38_cpg_islands", "island", "not_island"))
```

```{r}
annots <- c("hg38_basicgenes", "hg38_genes_intergenic")

annotations <- annotatr::build_annotations(genome = 'hg38', annotations = annots)
seqs <- annotations %>% data.frame() %>% distinct(seqnames) %>% .[1:22,]
annotations <- annotations %>% data.frame() %>% filter(seqnames %in% seqs)
annotations %>% group_by(type) %>% summarise(n=n())
```

```{r}
for_beta_genes <- find_overlaps(as_granges(for_beta), as_granges(annotations)) %>% data.frame()
for_beta_genes[1:10,]
```

## How to sliding window

* anchor at a CpG
* window extends +/- from it
* new alpha for each read is calculated based on the number of CpGs within the window
* make the window length the read length


* this sample the reads are 36bp (I know it's crazy)
* make a granges of windows? - and then overlap and recalculate?
```{r}
windows <- for_beta %>% distinct(seqnames, start) %>%
  mutate(index = start, start = index - 18, end = index + 18) %>% as_granges() 
```

```{r}
for_beta_window <- find_overlaps(windows, as_granges(mutate(for_beta, cpg_pos = start))) %>% data.frame() %>%
  group_by(index, read_id) %>%
  mutate(window_total = n(),
         window_pattern = paste0(meth_status, collapse = ""),
         window_meth = str_count(window_pattern, pattern = "M"),
         window_unmeth = str_count(window_pattern, pattern = "U"),
         window_alpha = window_meth/window_total) %>%
  ungroup()
for_beta_window %>% distinct(index, read_id, alpha, window_alpha, meth_pattern, window_pattern) %>% .[1:10,]
```

# how many windows do reads fit into?

```{r}
for_beta_window %>% distinct(index, read_id, alpha, window_alpha, meth_pattern, window_pattern) %>%
  group_by(read_id) %>%
  summarise(n_window = n()) %>%
  ggplot(aes(x = n_window)) +
  geom_histogram() +
  theme_bw()
```
* this reflects CpG density? - I think?

```{r}
for_beta_window %>% distinct(index, read_id, alpha, window_alpha, meth_pattern, window_pattern, total) %>%
  group_by(read_id, total) %>%
  summarise(n_window = n()) %>%
  ggplot(aes(x = total, y = n_window)) +
  geom_count() +
  theme_bw()
```

```{r}
for_beta_window %>% distinct(index, read_id, alpha, window_alpha, meth_pattern, window_pattern) %>%
  group_by(read_id) %>%
  mutate(n_window = n()) %>%
  ungroup() %>%
  filter(n_window == 12) %>% distinct(read_id) %>% nrow()

for_beta_window %>% distinct(index, read_id, alpha, window_alpha, meth_pattern, window_pattern) %>%
  group_by(read_id) %>%
  mutate(n_window = n()) %>%
  ungroup() %>%
  filter(n_window == 12) %>% .[1:5,]
```

```{r}
for_beta_window %>% filter(read_id == "HWI-EAS149_7:1:100:1168:337:0:1:1") %>%
  ggplot(aes(x = index, y = window_total)) +
  geom_point()
```

# alpha correlations

* each read will now have multiple alphas

* also we now have an alpha for the window, and the original alpha
* are the new alphas more correlated to each other? - and how do they compare to the original?

```{r}
for_beta_window %>% distinct(index, read_id, alpha, window_alpha, meth_pattern, window_pattern, total, window_total) %>%
  ggplot(aes(x = alpha, y = window_alpha)) +
  geom_count() +
  theme_bw()

for_beta_window %>% distinct(index, read_id, alpha, window_alpha, meth_pattern, window_pattern, total, window_total) %>%
  ggplot(aes(x = alpha, y = window_alpha)) +
  geom_point(alpha = 0.4) +
  theme_bw()
```

* it's pretty but I have no clue what it means
```{r}
for_beta_window %>% distinct(index, read_id, alpha, window_alpha, meth_pattern, window_pattern, total, window_total) %>%
  group_by(index) %>%
  summarise(n_read = n()) %>%
  ggplot(aes(x = n_read)) +
  geom_histogram(bins = 100) +
  theme_bw()

for_beta_window %>% distinct(index, read_id, alpha, window_alpha, meth_pattern, window_pattern, total, window_total) %>%
  group_by(index) %>%
  mutate(n_read = n()) %>%
  filter(n_read > 1) %>%
  ggplot(aes(x = n_read)) +
  geom_histogram(bins = 50) +
  theme_bw()
```
```{r}
for_beta_window %>% distinct(index, read_id, alpha, window_alpha, meth_pattern, window_pattern, total, window_total) %>%
  group_by(index) %>%
  mutate(n_read = n()) %>%
  ungroup() %>% summarise(mean(n_read), sd(n_read), median(n_read))
```


* compare new alphas within their windowsm

```{r}
for_beta_window %>% distinct(index, read_id, alpha, window_alpha, meth_pattern, window_pattern, total, window_total) %>%
  group_by(index) %>%
  mutate(n_read = n()) %>%
  ungroup() %>% .[order(.$index),] %>% filter(index <= 10571837) %>%
  ggplot(aes(x = window_alpha, colour = as.factor(index))) +
  geom_density() +
  theme_bw()

for_beta_window %>% distinct(index, read_id, alpha, window_alpha, meth_pattern, window_pattern, total, window_total) %>%
  group_by(index) %>%
  mutate(n_read = n()) %>%
  ungroup() %>% .[order(.$index),] %>% filter(index <= 10571837) %>%
  ggplot(aes(x = window_alpha, colour = as.factor(index)), fill = as.factor(index)) +
  geom_histogram(alpha = 0.4) +
  theme_bw()
```

```{r}
for_beta_window %>% distinct(index, read_id, alpha, window_alpha, meth_pattern, window_pattern, total, window_total) %>%
  group_by(index) %>%
  mutate(n_read = n()) %>%
  ungroup() %>% .[order(.$index),] %>% filter(index <= 10571837) %>%
  ggplot(aes(x = alpha, colour = as.factor(index))) +
  geom_density() +
  theme_bw()
```

* so the distributions in these bins for the alpha is different between window alpha and alpha
* but what does it mean?

```{r}
for_beta_window %>% distinct(index, read_id, alpha, window_alpha, meth_pattern, window_pattern, total, window_total) %>%
  group_by(index) %>%
  mutate(n_read = n(),
         mean_alpha = mean(alpha), 
         mean_window_alpha = mean(window_alpha),
         var_alpha = var(alpha),
         var_window_alpha = var(window_alpha)) %>%
  ungroup() %>%
  .[,c(1,10:13)] %>%
  reshape2::melt(id = "index") %>% distinct() %>% .[order(.$index),] %>% .[1:10,]

```

```{r}
for_beta_window %>% distinct(index, read_id, alpha, window_alpha, meth_pattern, window_pattern, total, window_total) %>%
  group_by(index) %>%
  mutate(n_read = n(),
         mean_alpha = mean(alpha), 
         mean_window_alpha = mean(window_alpha),
         var_alpha = var(alpha),
         var_window_alpha = var(window_alpha)) %>%
  ungroup() %>%
  ggplot(aes(x = mean_alpha, y = mean_window_alpha)) +
  geom_point(alpha = 0.4) +
  theme_bw()
```


```{r}
for_beta_window %>% distinct(index, read_id, alpha, window_alpha, meth_pattern, window_pattern, total, window_total) %>%
  group_by(index) %>%
  mutate(n_read = n(),
         mean_alpha = mean(alpha), 
         mean_window_alpha = mean(window_alpha),
         var_alpha = var(alpha),
         var_window_alpha = var(window_alpha)) %>%
  ungroup() %>%
  ggplot(aes(x = var_alpha, y = var_window_alpha)) +
  geom_point(alpha = 0.4) +
  theme_bw()
```

```{r}
for_beta_window %>% distinct(index, read_id, alpha, window_alpha, meth_pattern, window_pattern, total, window_total) %>%
  group_by(index) %>%
  mutate(n_read = n(),
         mean_alpha = mean(alpha), 
         mean_window_alpha = mean(window_alpha),
         var_alpha = var(alpha),
         var_window_alpha = var(window_alpha)) %>%
  ungroup() %>%
  .[,c(1,12:13)] %>%
  reshape2::melt(id = "index") %>% distinct() %>% .[order(.$index),] %>%
  ggplot(aes(x = value, colour = variable)) +
  geom_density() +
  theme_bw()

for_beta_window %>% distinct(index, read_id, alpha, window_alpha, meth_pattern, window_pattern, total, window_total) %>%
  group_by(index) %>%
  mutate(n_read = n(),
         mean_alpha = mean(alpha), 
         mean_window_alpha = mean(window_alpha),
         var_alpha = var(alpha),
         var_window_alpha = var(window_alpha)) %>%
  ungroup() %>%
  .[,c(1,12:13)] %>%
  reshape2::melt(id = "index") %>% distinct() %>% .[order(.$index),] %>%
  ggplot(aes(x = log2(value), colour = variable)) +
  geom_density() +
  theme_bw()
```
* actually looks like variance is lower in original alpha - interesting

```{r}
for_beta_window %>% distinct(index, read_id, alpha, window_alpha, meth_pattern, window_pattern, total, window_total) %>%
  group_by(index) %>%
  mutate(n_read = n(),
         mean_alpha = mean(alpha), 
         mean_window_alpha = mean(window_alpha),
         var_alpha = var(alpha),
         var_window_alpha = var(window_alpha)) %>%
  ungroup() %>%
  .[,c(1,12:13)] %>%
  reshape2::melt(id = "index") %>% distinct() %>% .[order(.$index),] %>%
  ggplot(aes(x = variable, y = value, fill = variable)) +
  geom_boxplot() +
  theme(legend.position = "none") +
  theme_bw()

for_beta_window %>% distinct(index, read_id, alpha, window_alpha, meth_pattern, window_pattern, total, window_total) %>%
  group_by(index) %>%
  mutate(n_read = n(),
         mean_alpha = mean(alpha), 
         mean_window_alpha = mean(window_alpha),
         var_alpha = var(alpha),
         var_window_alpha = var(window_alpha)) %>%
  ungroup() %>%
  .[,c(1,12:13)] %>%
  reshape2::melt(id = "index") %>% distinct() %>% .[order(.$index),] %>%
  ggplot(aes(x = variable, y = log2(value), fill = variable)) +
  geom_boxplot() +
  theme(legend.position = "none") +
  theme_bw()
```


# levene test

[tutorial](https://www.geeksforgeeks.org/r-language/levenes-test-in-r-programming/)

* compares variations between 2 groups
* group with lower variance is more similar to itself

Package `car`

* oh except I have index groups, and then the alpha/window group
```{r}
lev <- for_beta_window %>% distinct(index, read_id, alpha, window_alpha) %>%
  reshape2::melt(id = c("index", "read_id"))
lev$index <- as.factor(lev$index)
lev[1:10,]
```
```{r}
distinct(lev, index) %>% .[1:10,]
```

```{r}
car::leveneTest(value ~ variable, lev)
```
* fail to reject null
* null is that all population variances are =

```{r}
#lev %>%
 # group_by(index) %>%
#  mutate(car::leveneTest(value ~ variable, .))
```


